### 2022-03-29 GB_Simulation_3Species.Rmd
New repository with C++ changes

---
  title: "Modeling 3 species on George's Bank (Yellowtail Flounder, Cod, Haddock)"
editor_options: 
  chunk_output_type: console
---

  ## Habitat setup

First read in habitat created using point data and covariates (sediment and depth)
and create other required info

It returns a list of suitable habitat for each species (hab), the habitat as
adjusted during the spawning period (spwn_hab) and the binary location of
spawning areas (spwn_loc). It also returns the locations as x1,x2,y1,y2 and the
multiplier of attractiveness to the spawning area during spawning periods
(spwn_mult).

If plot.dist = TRUE, it returns the plots to a file.

```{r habitat}

#LOAD HABITAT PREVIOUSLY CREATED
hab <- readRDS(file="hab_justYT.RDS")

#plot strata
par(mar=c(1,1,1,1))
fields::image.plot(hab$stratas)

#plot habitat
par(mar=c(1,1,1,1))
fields::image.plot(hab[["hab"]][["spp1"]])

#plot spawn habitat
par(mar=c(1,1,1,1))
fields::image.plot(hab[["spwn_hab"]][["spp1"]])

```

# Initialise the simulation

This vignette is a paired down example of how to construct a simulation using
MixFishSim. We include only a basic example and encourage users to explore the
other features of the package. \\

## Base parameters

First we specify the basic parameters of the simulation. This includes the
dimensions of the spatial domain, the number of years to simulate, the number
of fleets and vessels per fleet and the number of species and how often (in
                                                                         weeks) the fish move.

The object returned is used internally by MixFishSim a list with two levels:

  * sim$idx : The different units of different processes 
* sim$brk.idx: breaks for each of the key processes in units of a timestep

```{r basic}

#NEW VERSION that has week breaks for entire simulation
# source("R/init_sim_Bens.R")
# sim <- init_sim_Bens(nrows = 100, ncols = 100, n_years = 20, n_tows_day = 1,
#                 n_days_wk_fished = 1, n_fleets = 1, n_vessels = 0, n_species = 2,
#                 move_freq = 1)

#NEW VERSION that has week breaks for entire simulation and allows fishing on just 1 day per week
source("R/init_sim_Bens_nofish.R")
sim <- init_sim_Bens_nofish(nrows = nrow(hab$hab$spp1), ncols = ncol(hab$hab$spp1), n_years = 22, n_tows_day = 1,n_days_wk_fished = 1, n_fleets = 1, n_vessels = 1, n_species = 1,
                            move_freq = 1)

class(sim)
sim$idx
names(sim$brk.idx)

```

## Population models

Now we need to set up the population models for the simulations. We do this
with the init_pop function. We set the initial population biomasses, movement
rates, recruitment parameter and growth and natural mortality rates.

The object created stores all the starting conditions and containers for
recording the changes in the populations during the simulations.

We can plot the starting distributions for each population as a check.

```{r pop_init}

#load rcpp exports
Rcpp::sourceCpp(file= "src/Movement.cpp")
Rcpp::sourceCpp(file= "src/RcppExports.cpp")
Rcpp::compileAttributes() #this updates RcppExports.R file, which contains function definitions

#CALULATE INDICES OF NONZERO VALUES IN HAB TO PASS TO MOVE_POPULAITON DURING MOVEMENT
nonzero_idx <- lapply(paste0("spp", seq_len(sim$idx[["n.spp"]])), function(s) {

  which(hab[["hab"]][[s]] >0 , arr.ind=T)

})

names(nonzero_idx) <- paste("spp",seq_len(sim$idx[["n.spp"]]), sep ="")

 source("R/init_pop_Bens.R")
# profvis({

#decreasing population settings
Pop <- init_pop_Bens(sim_init = sim, Bio = c("spp1" = 3194), #CHANGE
                     hab = hab[["hab"]], start_cell = c(42,155),
                     lambda = c("spp1" = 0), #set at 0 first, change after initial population set
                     init_move_steps = 20,
                     rec_params = list("spp1" = c("model" = "BH", "a" = 2, "b" = 4, "cv" = 0)), #CHANGE
                     rec_wk = list("spp1" = 9:12),
                     spwn_wk = list("spp1" = 9:12), #CORRECT FOR YTF
                     M = c("spp1" = 0.275), #CHANGE
                     K = c("spp1" = 0.3, "spp2" = 0.3), #CHANGE
                     nz = nonzero_idx
)

#})

#set lambda at 0 first, change after initial population set
Pop$dem_params$spp1$lambda <- .1

#I PUT THIS BACK INSIDE RUN_SIM
# #Calculate movement probabilities (used to be in run_sim)
# Move_Prob  <- lapply(paste0("spp", seq_len(sim[["idx"]][["n.spp"]])), function(s) { move_prob_Lst(lambda = Pop[["dem_params"]][[s]][["lambda"]], hab = hab[["hab"]][[s]])})
# 
# 
# Move_Prob_spwn <- lapply(paste0("spp", seq_len(sim[["idx"]][["n.spp"]])), function(s) { move_prob_Lst(lambda = Pop[["dem_params"]][[s]][["lambda"]], hab = hab[["spwn_hab"]][[s]])})
# 
# 
# names(Move_Prob)      <- paste0("spp", seq_len(sim[["idx"]][["n.spp"]]))
# names(Move_Prob_spwn) <- paste0("spp", seq_len(sim[["idx"]][["n.spp"]]))

fields::image.plot(Pop$Start_pop[[1]], main = "spp1 starting biomass")

```

## Population movement

Now we set up the population tolerance to different temperatures which
determines how the populations move during the course of a year. We can then
plot the combined spatiotemporal suitable habitat to examine how these
interact.

```{r temp}
#load temp gradient

moveCov <- readRDS(file="20 year moveCov matrices/GeorgesBank/GB_22yr_ConstTemp_HaddockStrata")

#set temperature preferences manually. 
#The following assumes moveCov has been created and already has an empty spp_tol sublist
moveCov[["spp_tol"]] <- list() #just in case
moveCov[["spp_tol"]] <- list("spp1" = list("mu" = 9, "va" = 4),  #Haddock
     "spp2" = list("mu" = 8.75, "va" = 4.25),  #Cod
     "spp3" = list("mu" = 9, "va" = 4) )    #Yellowtail

```

## Fleet models

Here we initialise the fleet with fish landings price per tonne, catchability
coefficients per population, fuel cost, the coefficients for the step function
and fleet behaviour.

We can plot the behaviour of the step function to check its suitable for our
simulations. This determines the relationship between the monetary value gained
from a fishing tow and the next move by the vessel when using the correlated
random walk function.

```{r fleets}

library(MixFishSim)

#no fishing
fleets <- init_fleet(sim_init = sim, VPT = list("spp1" = 0, "spp2" = 0), #VPT = value per ton
                     Qs = list("fleet 1" = c("spp1" = 0, "spp2" = 0)   #Q = catchability
                     ),
                     fuelC = list("fleet1" = 3),
                     step_params = list("fleet 1" = c("rate" = 3, "B1" = 1, "B2" = 2, "B3" = 3)
                     ),				
                     past_knowledge = FALSE,  #dont use past knowledge
                     past_year_month = TRUE,
                     past_trip = TRUE,
                     threshold = 0.7
)

```

## Run simulation

Finally we run the simulation. The output is a list of objects containing all
the information on fisheries catches, the population dynamics and population
distributions. These can be examined with some inbuilt plotting functions.

```{r sim}
run_simulation <- function(x){

  source("R/run_sim.R")

  #to source a new go_fish where I edited to skips most things:
  #1: load file
  source("R/go_fish_Bens.R") #my edited version that skips most things
  #2: allow the function to call other hidden functions from mixfishsim 
  environment(go_fish_Bens) <- asNamespace('MixFishSim')
  #3: replace go_fish with go_fish_Bens in the MixFishSim package
  assignInNamespace("go_fish", go_fish_Bens, ns = "MixFishSim")

  library(MixFishSim) #each core needs to load library

  #load data from CPU
  #  load("C:/Users/benjamin.levy/Desktop/Github/READ-PDB-blevy2-toy/20 year moveCov matrices/Final/Final_moveCov_12_9_Bensmethod.RData")

  #load constant temp data from Mars
  #load("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-toy/20 year moveCov matrices/Final/Final_constanttemp_20yr.RData")

  #load increase temp data from Mars
  load("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-toy/20 year moveCov matrices/Final/Increase_temp_20yr_LargerTempVA7.RData")

  #this starts the profiling
library(profvis)

profvis({

start_time <- Sys.time() # record start time

  res<- run_sim(sim_init = sim,
                pop_init = Pop,
                move_cov = moveCov,
                fleets_init = fleets,
                hab_init = hab,
                save_pop_bio = TRUE,
                survey = NULL, #surv_random, will try to survey after simulation
                closure = NULL,
                InParallel = TRUE,
                nz = nonzero_idx)  #does it runin parallel? Doesnt seem like it

end_time <- Sys.time() # record end time
tot_time <- end_time - start_time # calculate duration of lrren() example

})

}

#run in parallel

library(parallel)

nCoresToUse <- detectCores() - 1   #this show 16 cores but I think I have 6??

nCoresToUse <- 6

cl <- parallel::makeCluster(nCoresToUse,revtunnel = TRUE, outfile = "", verbose = TRUE, master=nsl(Sys.info()['nodename'])) #options from https://stackoverflow.com/questions/41639929/r-cannot-makecluster-multinode-due-to-cannot-open-the-connection-error

result <- list()

result <- parallel::parLapply(cl,1:6,run_simulation)
parallel::stopCluster(cl)

# 
# #below throws an error
# #result <- foreach(i=1:3) %dopar% run_simulation(i)
# 
# #below throws a similar error: 3 nodes produced errors; first error: missing value where TRUE/FALSE needed
#  parLapply(cl,1:3,run_simulation)
# stopCluster(cl)

```

## Summary plots

There are a series of input plotting functions to visualise the results of the
simulation. For example, we can explore:

  * the population dynamics for each species
* Seasonal patterns in exploitation
* the location choice of a vessel
* the realised step function for a vessel

Users will wish to define their own plots, depending on the issues of interest
and all the results are saved in the output from the run_sim function.

```{r plots}

## Biological
source("R/plot_pop_summary.R")
p1 <- plot_pop_summary(results = res,
                       timestep = "annual",
                       save = FALSE, 
                       save.location = NULL
)

plot_pop_summary(results = res, timestep = "daily", save = FALSE, save.location = "C:/Users/benjamin.levy/Desktop/Github/READ-PDB-blevy2-toy/testfolder")
p1

p2 <- plot_daily_fdyn(res)
p2

## Fishery

logs <- combine_logs(res[["fleets_catches"]])

p3 <- plot_vessel_move(sim_init = sim, logs = logs, fleet_no = 1, vessel_no = 5,
                       year_trip = 5, trip_no = 10)
p3

p4 <- plot_realised_stepF(logs = logs, fleet_no = 1, vessel_no = 1)
p4

#try some new ones

#plot survey 2 ways
p5 <- plot_survey(survey = res$survey, type = "spatial")
p5

p6 <- plot_survey(survey = res$survey, type = "index")
p6

#spatio temporal population plots
source("R/Bens_plot_pop_spatiotemp.R")
p7 <- Bens_plot_pop_spatiotemp(results = res,
                               save = TRUE, 
                               save.location = "testfolder/spatialplots"
)
p7

```

Note in our example how the fishing mortality rate for species 2 changes
following the spatial closure, which was set to cover some of the core
distribution of the population.

### 2022-03-29 Generic Species.Rmd
New repository with C++ changes

--- 2022-03-29 GB_Simulation_3Species.Rmd
@@
 ---
-  title: "Modeling 3 species on George's Bank (Yellowtail Flounder, Cod, Haddock)"
+  title: "Simple MixFishSim Example"
+output: rmarkdown::html_vignette
+vignette: >
+  %\VignetteIndexEntry{Simple MixFishSim Example}
+%\VignetteEngine{knitr::rmarkdown}
 editor_options: 
@@
 
-  ## Habitat setup
+  This is a simple example of how to use \textbf{'MixFishSim'} to generate
+simulations of the dynamics in a mixed fishery. We describe how to calibrate
+the habitat fields, the population models, the fishery model and implement a
+simple fixed spatial closure. \\
 
-First read in habitat created using point data and covariates (sediment and depth)
-and create other required info
+First, load the packages and set a seed for reproducibility.
 
-It returns a list of suitable habitat for each species (hab), the habitat as
-adjusted during the spawning period (spwn_hab) and the binary location of
-spawning areas (spwn_loc). It also returns the locations as x1,x2,y1,y2 and the
-multiplier of attractiveness to the spawning area during spawning periods
-(spwn_mult).
+# Load MixFishSim
 
-If plot.dist = TRUE, it returns the plots to a file.
+```{r packages}
+#install.packages('devtools')
+#install.packages('githubinstall')
 
-```{r habitat}
+# library(devtools)
+# library(githubinstall)
+# install_github("pdolder/MixFishSim")
+# 
+# 
+# githubinstall("MixFishSim")
 
-#LOAD HABITAT PREVIOUSLY CREATED
-hab <- readRDS(file="hab_justYT.RDS")
+#install.packages('Rcpp')
+#install.packages('Rtools')
+#install.packages('MixFishSim')
+library(MixFishSim)
+#library(knitr)
+#opts_chunk$set(tidy = TRUE)
 
-#plot strata
-par(mar=c(1,1,1,1))
-fields::image.plot(hab$stratas)
+#library(reshape2)
 
-#plot habitat
-par(mar=c(1,1,1,1))
-fields::image.plot(hab[["hab"]][["spp1"]])
-
-#plot spawn habitat
-par(mar=c(1,1,1,1))
-fields::image.plot(hab[["spwn_hab"]][["spp1"]])
+#set.seed(123)
 
@@
 source("R/init_sim_Bens_nofish.R")
-sim <- init_sim_Bens_nofish(nrows = nrow(hab$hab$spp1), ncols = ncol(hab$hab$spp1), n_years = 22, n_tows_day = 1,n_days_wk_fished = 1, n_fleets = 1, n_vessels = 1, n_species = 1,
+sim <- init_sim_Bens_nofish(nrows = 100, ncols = 100, n_years = 1, n_tows_day = 1,
+                            n_days_wk_fished = 1, n_fleets = 1, n_vessels = 1, n_species = 2,
                             move_freq = 1)
@@
 names(sim$brk.idx)
+
+```
+
+## Habitat setup
+
+This function creates the spatial fields which support the fish populations and
+determine their spatial distributions. You define the parameters for the
+matern covariance function for each population and optionally the location of
+any spawning closure areas.
+
+It returns a list of suitable habitat for each species (hab), the habitat as
+adjusted during the spawning period (spwn_hab) and the binary location of
+spawning areas (spwn_loc). It also returns the locations as x1,x2,y1,y2 and the
+multiplier of attractiveness to the spawning area during spawning periods
+(spwn_mult).
+
+If plot.dist = TRUE, it returns the plots to a file.
+
+```{r habitat}
+
+source("R/BENS_plot_habitat.R")
+
+#values settled on from anisotropy and habtest scripts
+
+spp.ctrl = list(
+  "spp.1" = list('nu' = 1/0.05, 
+                 'var' = 1,
+                 'scale' = 5, 
+                 'Aniso' = matrix(nc = 2,  c(1.5, -3, 3, 4) )),
+  "spp.2" = list('nu' = 1/0.015, 
+                 'var'  = 1,
+                 'scale' = 25, 
+                 'Aniso' = matrix(nc = 2,c(1, -2, 1, 2))),
+  plot.dist = TRUE, 
+  plot.file = "testfolder"
+)
+
+#DEFINING UNIFORM NXN GRID
+#defining strata coordinates
+#number of rows and columns to define (assume want n equal sized strata)
+nrows <- 4
+ncols <- 4
+
+#of the form  c(x1, x2, y1, y2) THESE ARE BOUNDARIES OF STRATA VALUES
+# xs are rows ys columns
+strata_num <- 1
+stratas <- list()
+
+ for(r in seq(nrows)){
+
+   for(c in seq(ncols)){
+
+   stratas[[paste("strata",strata_num,sep="")]] = c( (r-1)*(sim$idx[["nrows"]]/nrows)+1, (r)*(sim$idx[["nrows"]]/nrows) , (c-1)*(sim$idx[["ncols"]]/ncols)+1  , (c)*(sim$idx[["ncols"]]/ncols) )
+
+    strata_num <- strata_num +1
+
+  }
+
+}
+
+#DEFINING RANDOM NXN GRID
+#size of our domain
+totalrows <- 100
+totalcols <- 100
+
+#desired dimentions of strata
+nrows <- 5
+ncols <- 6
+
+#of the form  c(x1, x2, y1, y2) THESE ARE BOUNDARIES OF STRATA VALUES
+# xs are rows ys columns
+strata_num <- 1
+stratas <- list()
+
+#generate randow sequence of rows
+rowind <- sort(sample(seq(totalrows),nrows-1,replace=FALSE))
+rowind <- c(1,rowind,totalrows)
+
+row_idx <- 0
+ for(r in seq(length(rowind)-1)){
+row_idx <- row_idx+1
+
+   #generate random sequence of columns
+    colind <- sort(sample(seq(totalcols),ncols-1,replace=FALSE))
+    colind <- c(1,colind,totalcols)
+
+    col_idx <- 0
+
+   for(c in seq((length(colind)-1))){
+
+     col_idx <- col_idx+1
+
+   stratas[[paste("strata",strata_num,sep="")]] = c( rowind[row_idx] , rowind[row_idx+1] , colind[col_idx] , colind[col_idx+1])
+
+    strata_num <- strata_num +1
+
+  }
+
+}
+
+source("R/BENS_create_hab.R")
+
+hab <- BENS_create_hab(sim_init = sim, 
+                       spp.ctrl = spp.ctrl,
+
+                       spawn_areas = list(
+                         "spp1" = list(
+                           'area1' = c(30,45,55,65),   #of the form  c(x1, x2, y1, y2) THESE ARE BOUNDARIES OF MATRIX VALUES
+                           'area2' = c(70,90,50,60)   #need to revisit closure areas
+                         ),
+                         "spp2" = list(
+                           'area1' = c(30,45,55,65),
+                           'area2' = c(70,90,50,60)
+                         )),
+                       spwn_mult = 10,#THIS DOES NOT CHANGE ANYTHING. MUST CHANGE VALUE AT TOP OF BENS_CREATE_HAB 
+                       plot.dist = F, 
+                       plot.file = "testfolder",
+
+                       #created this new part defining strata
+                      strata = stratas
+
+                       )
+
+#plot strata
+par(mar=c(1,1,1,1))
+fields::image.plot(hab$stratas)
+
+#print(hab)
+source("R/BENS_plot_habitat.R")
+## Plot the unadjusted habitat fields
+BENS_plot_habitat(hab = hab$hab, 	spp.ctrl = spp.ctrl)
+
+#old version
+plot_habitat(hab$hab)
+
+## Plot the adjusted habitat fields
+plot_habitat(hab$spwn_hab)
 
@@
 
+#Week 13 is first week of april
+#half way through april to half way through may would be week 15-18
+
 source("R/init_pop_Bens.R")
-# profvis({
+
+#original settings
+# Pop <- init_pop_Bens(sim_init = sim, Bio = c("spp1" = 2e5, "spp2" = 4e5), #these values from paper : 1e5 and 2e5
+# 		hab = hab[["hab"]], start_cell = c(50,50), 
+# 		lambda = c("spp1" = 0.1, "spp2" = 0.1), #same lambda for all?
+# 		init_move_steps = 20,
+# 		rec_params = list("spp1" = c("model" = "BH", "a" = 6, "b" = 4, "cv" = 0.7),
+# 				  "spp2" = c("model" = "BH", "a" = 27, "b" = 4,"cv" = 0.6)), #these values from paper 
+# 		rec_wk = list("spp1" = 12:15, "spp2" = 12:15),
+# 		spwn_wk = list("spp1" = 15:18, "spp2" = 15:18),
+# 		M = c("spp1" = 0.2, "spp2" = 0.1), #these values from paper: c("spp1" = 0.2, "spp2" = 0.1)
+# 		K = c("spp1" = 0.3, "spp2" = 0.3) #all the same for now
+# 		)
 
 #decreasing population settings
-Pop <- init_pop_Bens(sim_init = sim, Bio = c("spp1" = 3194), #CHANGE
-                     hab = hab[["hab"]], start_cell = c(42,155),
-                     lambda = c("spp1" = 0), #set at 0 first, change after initial population set
+Pop <- init_pop_Bens(sim_init = sim, Bio = c("spp1" = 4e5, "spp2" = 10e5), #these values from paper : 1e5 and 2e5
+                     hab = hab[["hab"]], start_cell = c(50,50),
+                     lambda = c("spp1" = 0.1, "spp2" = 0.1), #same lambda for all?
                      init_move_steps = 20,
-                     rec_params = list("spp1" = c("model" = "BH", "a" = 2, "b" = 4, "cv" = 0)), #CHANGE
-                     rec_wk = list("spp1" = 9:12),
-                     spwn_wk = list("spp1" = 9:12), #CORRECT FOR YTF
-                     M = c("spp1" = 0.275), #CHANGE
-                     K = c("spp1" = 0.3, "spp2" = 0.3), #CHANGE
+                     rec_params = list("spp1" = c("model" = "BH", "a" = 2, "b" = 4, "cv" = 0),
+                                       "spp2" = c("model" = "BH", "a" = 7, "b" = 4,"cv" = 0)), #these values from paper
+                     rec_wk = list("spp1" = 15:18, "spp2" = 15:18),
+                     spwn_wk = list("spp1" = 15:18, "spp2" = 15:18),
+                     M = c("spp1" = 0.275, "spp2" = 0.225), #these values from paper: c("spp1" = 0.2, "spp2" = 0.1)
+                     K = c("spp1" = 0.3, "spp2" = 0.3), #all the same for now
                      nz = nonzero_idx
                      )
-
-#})
-
-#set lambda at 0 first, change after initial population set
-Pop$dem_params$spp1$lambda <- .1
 
@@
 
-fields::image.plot(Pop$Start_pop[[1]], main = "spp1 starting biomass")
+names(Pop)
+
+Pop$dem_params
+
+par(mfrow = c(2,1))
+
+image(Pop$Start_pop[[1]], main = "spp1 starting biomass")
+image(Pop$Start_pop[[2]], main = "spp2 starting biomass")
 
@@
 ```{r temp}
-#load temp gradient
-
-moveCov <- readRDS(file="20 year moveCov matrices/GeorgesBank/GB_22yr_ConstTemp_HaddockStrata")
 
@@
 moveCov[["spp_tol"]] <- list() #just in case
-moveCov[["spp_tol"]] <- list("spp1" = list("mu" = 9, "va" = 4),  #Haddock
-     "spp2" = list("mu" = 8.75, "va" = 4.25),  #Cod
-     "spp3" = list("mu" = 9, "va" = 4) )    #Yellowtail
+moveCov[["spp_tol"]] <- list("spp1" = list("mu" = 7.98, "va" = 3),  #8.13 IF TEMP INCREASES   7.98 if temp constant
+                             "spp2" = list("mu" = 7.98, "va" = 3) )
+
+plot(norm_fun(x = 0:25, mu = 8.3, va = 3)/max(norm_fun(0:25, 8.3, 3)),
+     type = "l", xlab = "Temperature", ylab = "Tolerance", lwd = 2)
+lines(norm_fun(x = 0:25, mu = 15, va = 9)/ max(norm_fun(0:25, 15, 9)),
+      type = "l", col = "blue", lwd = 2)
+lines(norm_fun(x = 0:25, mu = 17, va = 7)/ max(norm_fun(0:25, 17, 7)),
+      type = "l", col = "green", lwd = 2)
+
+legend(x = 2, y = 0.9, legend = c("spp1", "spp2"), lwd = 2, col = c("black", "blue"))
+
+# plot_spatiotemp_hab(hab = hab, moveCov = moveCov, spwn_wk = list("spp1" = 15:18, "spp2" = 15:18), plot.file =  "testfolder")
+# 
+# 
+# 
+# #to plot just the temp preferences over time
+# source("R/BENS_plot_spatiotemp_hab_justtemp.R")
+# 
+# BENS_plot_spatiotemp_hab_justtemp(hab = hab, moveCov = moveCov, spwn_wk = list("spp1" = 16:18, "spp2" = 16:19,"spp3" = 16:18, "spp4" = 18:20), plot.file =  "testfolder")
+#
 
@@
 ```{r fleets}
-
-library(MixFishSim)
+#initial settings
+# fleets <- init_fleet(sim_init = sim, VPT = list("spp1" = 4, "spp2" = 3),
+# 		     Qs = list("fleet 1" = c("spp1" = 1e-5, "spp2" = 3e-5),
+# 			       "fleet 2" = c("spp1" = 5e-5, "spp2" = 1e-5)
+# 			       ),
+# 		     fuelC = list("fleet1" = 3, "fleet 2" = 8),
+# 		     step_params = list("fleet 1" = c("rate" = 3, "B1" = 1, "B2" = 2, "B3" = 3),
+# 					"fleet 2" = c("rate" = 3, "B1" = 2, "B2" = 4, "B3" = 4)
+# 					),				
+# 		     past_knowledge = TRUE,
+# 		     past_year_month = TRUE,
+# 		     past_trip = TRUE,
+# 		     threshold = 0.7
+# 		     )
 
@@
 )
+
+test_step(step_params = fleets$fleet_params[[1]]$step_params, rev.max = 1e2)
+test_step(step_params = fleets$fleet_params[[2]]$step_params, rev.max = 1e2)
+
+```
+
+## Spatial closure
+
+We set up a spatial closure. There are multiple options in defining
+this, but we simply define a static fixed site closure for demonstration
+purposes.
+
+```{r close}
+# #practice creating a larger data.frame than just single points
+# library(tidyr)
+# 
+# #set x and y min/max which are coordinates on the grid
+# xmin <- 6
+# xmax <- 10
+# ymin <- 26
+# ymax <- 40
+# 
+# 
+# x <- xmin:xmax
+# y<- ymin:ymax
+# 
+# #View(crossing(x,y))
+# 
+# closure <- init_closure(input_coords = data.frame(x = x, y = y),
+# 			spp1 = "spp1", year_start = 2)
+
+```
+
+## Survey
+
+Its also possible to define a survey design using the init_survey function, but
+we do not do so for this demonstration. Please refer to the function help file
+if this is required.
+
+```{r survey}
+
+source("R/BENS_init_survey.R")
+
+#CURRENTLY NEED TO MAKE SURE THAT N_STATIONS*#YEARS / #STRATA IS A WHOLE NUMBER OTHERWISE DAY, TOW, YEAR WONT LINEUP WITH NUMBER OF STATIONS
+#ALSO NEED N_STATION TO BE DIVISIBLE BY STATIONS_PER_DAY
+#ALSO NEED N_STATIONS / STATIONS_PER_DAY <= 52 otherwise wont get to all of them in a year results in NA in the matrix
+
+#setup catch log
+surv_random <- BENS_init_survey(sim_init = sim,design = 'random_station', n_stations = 80, #this is total per year (20 in each of 4 strata)
+                                start_day = 1, stations_per_day = 1, Qs = c("spp1" = 0.1, "spp2"= 0.2),
+                                strata_coords = hab$strata, strata_num = hab$stratas )
 
@@
 
+    source("R/RcppExports_Bens.R")
+
+    #to source a new move_population where I edited to skips most things:
+  #1: load file
+  Rcpp::sourceCpp("src/Movement_Bens.cpp") #my edited version that skips most things
+  #2: allow the function to call other hidden functions from mixfishsim 
+  environment(move_population_Bens) <- asNamespace('MixFishSim')
+  #3: replace move_population with go_fish_Bens in the MixFishSim package
+  assignInNamespace("move_population", move_population_Bens, ns = "MixFishSim")
+
   library(MixFishSim) #each core needs to load library
@@
 
-  #this starts the profiling
-library(profvis)
-
-profvis({
-
-start_time <- Sys.time() # record start time
-
   res<- run_sim(sim_init = sim,
@@
                 nz = nonzero_idx)  #does it runin parallel? Doesnt seem like it
-
-end_time <- Sys.time() # record end time
-tot_time <- end_time - start_time # calculate duration of lrren() example
-
-})
 

### 2022-03-29 src/Movement.cpp
d

@@ -1,0 +1,293 @@
+#include <Rcpp.h>
+using namespace Rcpp;
+
+
+////////////////////////////////////////////
+/// normalize matrix function       ////
+////////////////////////////////////////////
+
+//' @title sum matrix calculation
+//'
+//' @description \code{sum_mat} sums all values in a matrix
+//'
+//' @param M is a matrix passed from R 
+//'
+//' @return is the sum of the matrix
+//'
+//' @examples sum_mat(M)
+//'
+//' @export
+//  [[Rcpp::export]]
+
+NumericMatrix norm_mat(NumericMatrix M) {
+  //NumericMatrix MM = M(! is_na(M));
+  NumericMatrix S = M / sum(M);
+  return S;
+}
+
+
+////////////////////////////////////////////
+/// distance calculation function       ////
+////////////////////////////////////////////
+
+//' @title distance calculation
+//'
+//' @description \code{distance_calc} calculates the euclidean distance between
+//' two cell references.
+//'
+//' @param x1 is an integar for the starting x position
+//' @param y1 is an integar for the starting y position
+//' @param x2 is an integar for the end x position
+//' @param y2 is an integar for the end y position
+//'
+//' @return is a distance between the two cells
+//'
+//' @examples distance_calc(2, 3, 5, 7) 
+//'
+//' @export
+//  [[Rcpp::export]]
+
+double distance_calc(int x1, int y1, int x2, int y2) {
+  double dist = sqrt(pow(x1 - x2,2) + pow(y1 - y2,2));
+  return dist;
+}
+
+////////////////////////////////////////////
+/// Movement probability function       ////
+////////////////////////////////////////////
+
+//' @title movement probability function 
+//'
+//' @description \code{move_prob} calculates the movement probability between 
+//' a cell and all other cells based on the distance and \emph{lambda}.
+//'
+//' @param start is a Numeric vector of dim 2 for the starting position c(x,y)
+//' @param lambda is an integar for the value for the exponential decay in probability 
+//' of movement, i.e. \eqn{Pr(B|A) = \exp{-\lambda*dist_{a,b}} / Sum(c=1:c=n)  \exp{-\lambda*dist}}
+//' @param hab is a matrix of the habitat suitability
+//' @param Nzero_vals contains all nonzero indices from the habitat for a given species
+//'
+//' @return is a matrix of the movement probabilities from a cell
+//'
[…]

### 2022-03-29 src/RcppExports.cpp

@@ -1,0 +1,91 @@
+// Generated by using Rcpp::compileAttributes() -> do not edit by hand
+// Generator token: 10BE3573-1514-4C36-9D1C-5A225CD40393
+
+#include <Rcpp.h>
+
+using namespace Rcpp;
+
+#ifdef RCPP_USE_GLOBAL_ROSTREAM
+Rcpp::Rostream<true>&  Rcpp::Rcout = Rcpp::Rcpp_cout_get();
+Rcpp::Rostream<false>& Rcpp::Rcerr = Rcpp::Rcpp_cerr_get();
+#endif
+
+// norm_mat
+NumericMatrix norm_mat(NumericMatrix M);
+RcppExport SEXP _MixFishSim_norm_mat(SEXP MSEXP) {
+BEGIN_RCPP
+    Rcpp::RObject rcpp_result_gen;
+    Rcpp::RNGScope rcpp_rngScope_gen;
+    Rcpp::traits::input_parameter< NumericMatrix >::type M(MSEXP);
+    rcpp_result_gen = Rcpp::wrap(norm_mat(M));
+    return rcpp_result_gen;
+END_RCPP
+}
+// distance_calc
+double distance_calc(int x1, int y1, int x2, int y2);
+RcppExport SEXP _MixFishSim_distance_calc(SEXP x1SEXP, SEXP y1SEXP, SEXP x2SEXP, SEXP y2SEXP) {
+BEGIN_RCPP
+    Rcpp::RObject rcpp_result_gen;
+    Rcpp::RNGScope rcpp_rngScope_gen;
+    Rcpp::traits::input_parameter< int >::type x1(x1SEXP);
+    Rcpp::traits::input_parameter< int >::type y1(y1SEXP);
+    Rcpp::traits::input_parameter< int >::type x2(x2SEXP);
+    Rcpp::traits::input_parameter< int >::type y2(y2SEXP);
+    rcpp_result_gen = Rcpp::wrap(distance_calc(x1, y1, x2, y2));
+    return rcpp_result_gen;
+END_RCPP
+}
+// move_prob
+NumericMatrix move_prob(NumericVector start, double lambda, NumericMatrix hab, NumericMatrix Nzero_vals);
+RcppExport SEXP _MixFishSim_move_prob(SEXP startSEXP, SEXP lambdaSEXP, SEXP habSEXP, SEXP Nzero_valsSEXP) {
+BEGIN_RCPP
[…]

### 2022-03-29 R/BENS(old)_init_survey.R
R files

@@ -1,0 +1,170 @@
+#' @title Initialise survey settings
+#'
+#' @description \code{init_survey} is a function to mimic a
+#' fisheries-independent survey to sample catches from the populations.
+#'
+#' @param sim_init is the general simualtion settings from \link{sim_init}
+#' @param design is the survey design used, at the moment only
+#' \emph{fixed_station}
+#' @param n_stations is a Numeric for the number of stations to be fished each.
+#' Note: If using 'fixed_station' design this will be rounded down to maintain
+#' a grid shape if not divisble.  
+#' @param start_day is a Numeric for the first day of the survey each year
+#' @param stations_per_day is a Numeric for the number of stations surveyed per
+#' day
+#' @param Qs is a named Numeric Vector containing any survey catchabilities, assumed to be time invariant.
+#'
+#' @return is a list consisting of the survey setting and a a matrix for
+#' storing the log of catches from the survey, to be used as an input to
+#' \link{run_sim}.
+
+#' @examples
+#' init_survey(design = 'fixed_station', n_stations = 50, start_day = 90, stations_per_days = 5, Qs = c("spp1" = 0.1, "spp2" = 0.2)
+
+#' @export
+
+BENS_init_survey <- function (sim_init = NULL, design = 'fixed_station', n_stations = 50, 
+			     start_day = 90, stations_per_day = 5, Qs = NULL, strata_coords = NULL) {
+
+	# useful indexes
+	idx       <- sim_init[["idx"]]
+	n_fleets  <- idx[["nf"]]
+	n_vessels <- idx[["nv"]]
+	brk.idx   <- sim_init[["brk.idx"]]
+	
+
+
+	### Checks
+	if(!start_day %in% sim_init[["brk.idx"]][["day.seq"]]) stop("We don't fish this day, choose another start_day!")
+
+	## determine design
+
+	if(design == 'fixed_station') {
+
+	# Create a grid of stations
+	x <- round(seq(1,idx[["ncols"]], length.out = round(sqrt(n_stations),0)),0) # sim_init[["idx"]][["n_rows"]], 1)
+	y <- round(seq(1, idx[["nrows"]], length.out = round(sqrt(n_stations),0)),0) # sim_init[["idx"]][["n_cols"]], 1)
+	grid <- cbind(x = x, y = rep(y, each = length(x)))
+
+	# update no stations 
[…]

### 2022-03-29 R/BENS_create_hab.R

@@ -1,0 +1,171 @@
+#' @title Create habitat distribution fields
+#' 
+#' @description \code{create_hab} parametrises and returns the
+#' spatial fields used for the distribution of suitable habitat for the
+#' populations in the simulation. 
+#'
+#' The spatial fields are generated using
+#' \code{\link[RandomFields]{RFsimulate}} function from the \emph{RandomFields}
+#' package.
+#'
+#' @param sim is the parameter settings for the simulation, made by
+#' \code{init_sim} function.
+#' @param spp.ctrl List of controls to generate suitable habitat for each
+#' species. Must be of the form spp.ctrl = list(spp.1 = c(var = 20, ...),
+#' spp.2 = c(var = 10, ..),..) and contain the following:
+#' \itemize{
+#'	\item \strong{nu} (\emph{>=0})
+#' 	\item \strong{var} (\emph{>=0}) Controls the range in a matern covariance
+#' 	\item \strong{scale} (\emph{>=0})
+#'	\item \strong{Aniso} (\emph{matrix, dim = c(2,2)})
+#'}
+#' @param plot.dist Boolean, whether to plot the distributions to file
+#' @param plot.file path to save the plots of the species distributions
+
+#' @return Silently returns a list of spatial distributions of suitable habitat
+#' with first level of the list being the population (1 -> n.spp). If
+#' \code{plot.dist = TRUE} it produces an image of the spatial distributions at
+#' each time step for each of the populations saved to the working directory
+#' (unless specified otherwise in \code{plot.file})
+
+#' @examples
+#' hab <- create_hab(sim.init = sim.init, spp.ctrl = list(
+#'	      'spp.1' = list('nu' = 1/0.15, var = 1, scale = 10, Aniso =
+#'	      matrix(nc=2, c(1.5, 3, -3, 4)))), spawn_areas = list("spp1" =
+#'	      list("area1" = c(2,4,6,8))), list("spp2" = list("area1" =
+#'	      c(0,10,23,35))), spwn_mult = 10, plot.dist = TRUE, plot.file =   getwd())
+
+#' @export
+
[…]

### 2022-03-29 R/BENS_init_survey.R

@@ -1,0 +1,259 @@
+#' @title Initialise survey settings
+#'
+#' @description \code{init_survey} is a function to mimic a
+#' fisheries-independent survey to sample catches from the populations.
+#'
+#' @param sim_init is the general simualtion settings from \link{sim_init}
+#' @param design is the survey design used, at the moment only
+#' \emph{fixed_station}
+#' @param n_stations is a Numeric for the number of stations to be fished each.
+#' Note: If using 'fixed_station' design this will be rounded down to maintain
+#' a grid shape if not divisble.  
+#' @param start_day is a Numeric for the first day of the survey each year
+#' @param stations_per_day is a Numeric for the number of stations surveyed per
+#' day
+#' @param Qs is a named Numeric Vector containing any survey catchabilities, assumed to be time invariant.
+#'
+#' @param strata_coords is created by init_hab and contains the corner coordinates of each strata (if rectangular)
+#'
+#' @param strata_num is created by init_hab. It contains the entire domain with each of the n strata labeled 1 to n 
+#' 
+#' @param years_cut cuts off year_shift from the start of the survey
+#'
+#' @return is a list consisting of the survey setting and a a matrix for
+#' storing the log of catches from the survey, to be used as an input to
+#' \link{run_sim}.
+
+#' @examples
+#' init_survey(design = 'fixed_station', n_stations = 50, start_day = 90, stations_per_days = 5, Qs = c("spp1" = 0.1, "spp2" = 0.2)
+
+#' @export
+
+BENS_init_survey <- function (sim_init = NULL, design = 'fixed_station', n_stations = 50, 
+			     start_day = 90, stations_per_day = 5, Qs = NULL, strata_coords = NULL, strata_num = NULL, years_cut = 0) {
+
+	# useful indexes
+	idx       <- sim_init[["idx"]]
+	n_fleets  <- idx[["nf"]]
+	n_vessels <- idx[["nv"]]
+	brk.idx   <- sim_init[["brk.idx"]]
+	
+	n_strata <- length(strata_coords)
+
+
+	### Checks
+	if(!start_day %in% sim_init[["brk.idx"]][["day.seq"]]) stop("We don't fish this day, choose another start_day!")
+
+	## determine design
+
[…]

### 2022-03-29 R/BENS_plot_habitat.R

@@ -1,0 +1,53 @@
+#' @title Plot the habitat distribution fields for all populations
+#'
+#' @description \code{plot_habitat} is a function to plot the habitat spatial
+#' fields.
+#'
+#' @param hab is the habitat to plot, either unadjusted or adjusted for
+#' spawning area. 
+
+#' @return Silently plots the habitat fields (can be either the unadjusted
+#' habitat, or including the spawning adjustment).
+
+#' @examples
+#' plot_habitat(hab = hab)
+
+#' @export
+
+BENS_plot_habitat <- function(hab = hab, spp.ctrl = spp.ctl) {
+
+	n.spp <- length(hab)
+	nrows <- dim(hab[[1]])[1]
+	ncols <- dim(hab[[1]])[2]
+
+	
+	
+
+
+	
+	show("plotting inside plot_habitat")
+	
+	par(mfrow = c(ceiling(sqrt(n.spp)), ceiling(n.spp/ceiling(sqrt(n.spp)))), mar = c(2, 2, 2, 2))
+	for(i in seq(n.spp)) {
+	  par(pty="s") #this makes the plots square to compare to publication
+	image(hab[[paste0("spp", i)]], cex.axis = 1.5, cex.main = 2, col = grey(seq(1,0, l = 51)), axes = F)
+	axis(1, at = seq(0, 1, by = 0.2), labels = seq(0, nrows, by = nrows/5))
+	axis(2, at = seq(0, 1, by = 0.2), labels = seq(0, ncols, by = ncols/5))
+	text(0.1, 0.98, labels = paste("habitat spp = ", i), cex = 1)
+	
+	
+	nu <- round( spp.ctrl[[paste0("spp.",i)]][["nu"]], digits = 1)
+	var <- round( spp.ctrl[[paste0("spp.",i)]][["var"]], digits = 1)
+	scale <- round( spp.ctrl[[paste0("spp.",i)]][["scale"]], digits = 1)
+	Aniso <- round( spp.ctrl[[paste0("spp.",i)]][["Aniso"]], digits = 1)
+	
+	
+	text(0.1, 0.88, labels = paste("nu = ", nu), cex = 1)
+	text(0.1, 0.78, labels = paste("var = ", var), cex = 1)
+	text(0.1, 0.68, labels = paste("scale = ", scale), cex = 1)
+	text(0.1, 0.58, labels = paste("Aniso = ", Aniso[1],Aniso[2],Aniso[3],Aniso[4]), cex = 1)
+	}
+	
+}
+
+

### 2022-03-29 R/BENS_plot_spatiotemp_hab_justtemp.R

@@ -1,0 +1,112 @@
+#' @title Plot spatiotemporal habitat suitability
+
+#' @description Function to plot out the habitat suitability, as adjusted by
+#' the spatiotemporal move covariates
+
+#' @param hab is the output from \link{create_hab}
+#' @param moveCov is the output from \link{init_moveCov}
+#' @param plot.file path to save the plots of the spatiotemporal habitats 
+#' @param spwn_wk is a named list of the spawning week for each population
+
+#' @param colrange is the color range to use in the image plots. Best to set it as the range of all possible images 
+
+#' @examples None
+
+#' @export
+
+BENS_plot_spatiotemp_hab_justtemp <- function(hab = NULL, moveCov = NULL, plot.file = NULL, spwn_wk = NULL, plot_wk = NULL, colrange = NULL) {
+  
+  nrows <- nrow(hab[["hab"]][[1]]) 
+  ncols <- ncol(hab[["hab"]][[1]])
+  
+  library(fields)
+  library(lattice)
+  library(RColorBrewer)
+  
+  #plot just the temperature gradient
+  
+  nt <- length(moveCov[["cov.matrix"]])
+  if(!is.null(plot.file)) {
+    png(filename = paste0(plot.file,'/','justtemp_spatiotemp.png'), width = 800, height = 800)
+    #pdf(file=paste0(plot.file,'/','justtemp_spatiotemp.pdf'))	  
+  }
+  par(mfrow = c(ceiling(sqrt(length(plot_wk))), ceiling(length(plot_wk)/ceiling(sqrt(length(plot_wk))))), mar = c(1, 1, 1, 1))
+  
+  for(i in plot_wk) {
+    
+    move_cov_wk <- moveCov[["cov.matrix"]][[i]]
+    
+    #move_cov_wk <- move_cov_wk[,nrow(move_cov_wk):1] # Attempt 1: THIS PART ORIENTS THE IMAGE FOR PLOTTING
+    #  move_cov_wk <- apply(move_cov_wk,1,rev)
+    
+    move_cov_wk <- t(move_cov_wk[nrow(move_cov_wk):1,]) # Attempt 2: THIS PART ORIENTS THE IMAGE FOR PLOTTING
+    
+
+    
+    
+    #move_cov_wk <- matrix(unlist(moveCov[["cov.matrix"]][[i]]), ncol = ncols, nrow= nrows)
+    
[…]

### 2022-03-29 R/Bens_plot_pop_spatiotemp.R

@@ -1,0 +1,321 @@
+#' @title Plot population summary
+
+#' @description \code{plot_pop_summary} plots the four population dynamic
+#' metrics: catches, biomass, fishing mortality and recruitment.
+
+#' It can either operate at a daily timestep or an annual timestep
+
+#' @param results is an output from the function \link{run_sim}.
+#' @param timestep is a character string determining whether the plot is
+#' 'daily' or 'annual'
+#' @param save is a logical whether to save the plot
+#' @param save.location is a location (defaults to current directory)
+
+#' @return is a ggplot of all the species and metrics as a faceted plot
+
+#' examples
+#' plot_pop_summary(results = res, timestep = 'daily', save = TRUE, location = '.') 
+#' ## Not run
+
+#' @export
+
+Bens_plot_pop_spatiotemp <- function(results = res, timestep = 'daily', save = FALSE , plot_monthly = FALSE, save.location = NULL, plot_weekly = FALSE, ColBreaks = NULL) {
+  
+  
+  
+  
+  
+  #creating a plot similar to plot_spatiotemp_hab
+  
+  nrows <- nrow(res$pop_bios[[1]]$spp1) 
+  ncols <- ncol(res$pop_bios[[1]]$spp1)
+  
+  
+  if(plot_weekly == TRUE){
+    
+    for(s in seq_len(length(sim$idx[["n.spp"]]))) {  #number of species
+      
+      nt <- length(results[["pop_bios"]])  #number of weeks to plot
+      if(!is.null(save.location)) {
+        png(filename = paste0(save.location,'/','Population_spatiotemp_spp_',s,'.png'), width = 800, height = 800)
+      }
+      par(mfrow = c(ceiling(sqrt(nt)), ceiling(nt/ceiling(sqrt(nt)))), mar = c(1, 1, 1, 1))
+      
+      for(i in seq_len(nt-1)) {   #PLOTTED 1 LESS THAN END BECAUSE COMING UP NULL
+        
+        pop_wk <- results[["pop_bios"]][[i]][s]   
+        
+        
+        
+        image(results[["pop_bios"]][[i]][[paste0('spp',s)]], cex.axis = 1.5, cex.main = 2, col = grey(seq(1,0,l = 51)), axes = F)
+        
+        
+        
+        axis(1, at = seq(0, 1, by = 0.2), labels = seq(0, nrows, by = nrows/5))
+        axis(2, at = seq(0, 1, by = 0.2), labels = seq(0, ncols, by = ncols/5))
[…]

### 2022-03-29 R/Bens_plot_spatiotemp_hab.R

@@ -1,0 +1,113 @@
+#' @title Plot spatiotemporal habitat suitability
+
+#' @description Function to plot out the habitat suitability, as adjusted by
+#' the spatiotemporal move covariates
+
+#' @param hab is the output from \link{create_hab}
+#' @param moveCov is the output from \link{init_moveCov}
+#' @param plot.file path to save the plots of the spatiotemporal habitats 
+#' @param spwn_wk is a named list of the spawning week for each population
+
+#' @examples None
+
+#' @export
+
+plot_spatiotemp_hab <- function(hab = NULL, moveCov = NULL, plot.file = NULL, spwn_wk = NULL, plot_wk = NULL, plot_monthly = NULL,colrange = NULL) {
+
+	nrows <- nrow(hab[["hab"]][[1]]) 
+	ncols <- ncol(hab[["hab"]][[1]])
+	
+	if(!is.null(plot.file)) {
+	png(filename = paste0(plot.file,'/','habitat_spatiotemp_spp_','.png'), width = 800, height = 800)
+	}
+
+for(s in seq_len(length(hab[["hab"]]))) {
+
+	nt <- length(moveCov[["cov.matrix"]])
+
+	par(mfrow = c(ceiling(sqrt(length(plot_wk))), ceiling(length(plot_wk)/ceiling(sqrt(length(plot_wk))))), mar = c(1, 1, 1, 1))
+
+	for(i in plot_wk) {
+
+	move_cov_wk <- moveCov[["cov.matrix"]][[i]]
+
+	move_cov_wk_spp <- matrix(nc = ncols,
+				 nr = nrows, 
+				 sapply(move_cov_wk, norm_fun, 
+                          mu = moveCov[["spp_tol"]][[s]][["mu"]], 
+		  	  va = moveCov[["spp_tol"]][[s]][["va"]]))
+#col = grey(seq(1,0,l = 51)),
+		if(!i %in% spwn_wk[[s]]) {
+		fields::image.plot(hab[["hab"]][[paste0('spp',s)]] * move_cov_wk_spp, cex.axis = 1.5, cex.main = 2,  axes = F)
+		}
+	# col = grey(seq(1,0,l = 51)),
+		if(i %in% spwn_wk[[s]]) {
[…]

### 2022-03-29 R/RcppExports.R

@@ -1,0 +1,101 @@
+# Generated by using Rcpp::compileAttributes() -> do not edit by hand
+# Generator token: 10BE3573-1514-4C36-9D1C-5A225CD40393
+
+#' @title sum matrix calculation
+#'
+#' @description \code{sum_mat} sums all values in a matrix
+#'
+#' @param M is a matrix passed from R 
+#'
+#' @return is the sum of the matrix
+#'
+#' @examples sum_mat(M)
+#'
+#' @export
+norm_mat <- function(M) {
+    .Call('_MixFishSim_norm_mat', PACKAGE = 'MixFishSim', M)
+}
+
+#' @title distance calculation
+#'
+#' @description \code{distance_calc} calculates the euclidean distance between
+#' two cell references.
+#'
+#' @param x1 is an integar for the starting x position
+#' @param y1 is an integar for the starting y position
+#' @param x2 is an integar for the end x position
+#' @param y2 is an integar for the end y position
+#'
+#' @return is a distance between the two cells
+#'
+#' @examples distance_calc(2, 3, 5, 7) 
+#'
+#' @export
+distance_calc <- function(x1, y1, x2, y2) {
+    .Call('_MixFishSim_distance_calc', PACKAGE = 'MixFishSim', x1, y1, x2, y2)
+}
+
+#' @title movement probability function 
+#'
+#' @description \code{move_prob} calculates the movement probability between 
+#' a cell and all other cells based on the distance and \emph{lambda}.
+#'
+#' @param start is a Numeric vector of dim 2 for the starting position c(x,y)
+#' @param lambda is an integar for the value for the exponential decay in probability 
+#' of movement, i.e. \eqn{Pr(B|A) = \exp{-\lambda*dist_{a,b}} / Sum(c=1:c=n)  \exp{-\lambda*dist}}
+#' @param hab is a matrix of the habitat suitability
+#' @param Nzero_vals contains all nonzero indices from the habitat for a given species
+#'
+#' @return is a matrix of the movement probabilities from a cell
+#'
+#' @examples move_prob(c(2, 5), 0.3, matrix(nc = 3, runif(9))) 
+#' 
+#' @export
+move_prob <- function(start, lambda, hab, Nzero_vals) {
+    .Call('_MixFishSim_move_prob', PACKAGE = 'MixFishSim', start, lambda, hab, Nzero_vals)
+}
+
+#' @title movement probability function as a list 
+#'
[…]

### 2022-03-29 R/create_hab.R

@@ -1,0 +1,132 @@
+#' @title Create habitat distribution fields
+#' 
+#' @description \code{create_hab} parametrises and returns the
+#' spatial fields used for the distribution of suitable habitat for the
+#' populations in the simulation. 
+#'
+#' The spatial fields are generated using
+#' \code{\link[RandomFields]{RFsimulate}} function from the \emph{RandomFields}
+#' package.
+#'
+#' @param sim is the parameter settings for the simulation, made by
+#' \code{init_sim} function.
+#' @param spp.ctrl List of controls to generate suitable habitat for each
+#' species. Must be of the form spp.ctrl = list(spp.1 = c(var = 20, ...),
+#' spp.2 = c(var = 10, ..),..) and contain the following:
+#' \itemize{
+#'	\item \strong{nu} (\emph{>=0})
+#' 	\item \strong{var} (\emph{>=0}) Controls the range in a matern covariance
+#' 	\item \strong{scale} (\emph{>=0})
+#'	\item \strong{Aniso} (\emph{matrix, dim = c(2,2)})
+#'}
+#' @param plot.dist Boolean, whether to plot the distributions to file
+#' @param plot.file path to save the plots of the species distributions
+
+#' @return Silently returns a list of spatial distributions of suitable habitat
+#' with first level of the list being the population (1 -> n.spp). If
+#' \code{plot.dist = TRUE} it produces an image of the spatial distributions at
+#' each time step for each of the populations saved to the working directory
+#' (unless specified otherwise in \code{plot.file})
+
+#' @examples
+#' hab <- create_hab(sim.init = sim.init, spp.ctrl = list(
+#'	      'spp.1' = list('nu' = 1/0.15, var = 1, scale = 10, Aniso =
+#'	      matrix(nc=2, c(1.5, 3, -3, 4)))), spawn_areas = list("spp1" =
+#'	      list("area1" = c(2,4,6,8))), list("spp2" = list("area1" =
+#'	      c(0,10,23,35))), spwn_mult = 10, plot.dist = TRUE, plot.file =   getwd())
+
+#' @export
+
[…]

### 2022-03-29 R/go_fish.R

@@ -1,0 +1,450 @@
+#' @title Go fish
+
+#' @description \code{go_fish} is a function used to apply the fishing
+#' simulation model
+
+#' @param sim_init is the initialised object from \code{init_sim}.
+#' @param fleet_params is the parameter settings initialised from \code{_init_fleets}
+#' @param fleet_catches is the DF initialised from \code{_init_fleets}
+#' @param sp_fleet_catches is a list of spatial catches (as a Numeric matrix)
+#' for the fleet of each population @param closed_areas is a dataframe with the
+#' x,y coordinates are any closed areas, provided internally by
+#' \link{close_areas}
+#' 
+
+#' @return is a list containing i) the fleet catch dataframes , ii) the spatial
+#' catches of each population
+
+#' @examples
+#'
+
+#' @export
+
+go_fish <- function(sim_init = NULL, fleet_params = NULL, fleet_catches = NULL, 
+		    sp_fleet_catches = NULL, pops = NULL, closed_areas = NULL, t = t) {
+  
+  
+
+##### extract the relevant components ###########
+params <- fleet_params      # fleet parameter list
+catch  <- fleet_catches     # fleet catches list
+catch_matrix <- sp_fleet_catches
+VPT <- params[["VPT"]] 			# Value per tonne
+Q <- params[["Qs"]]			# Catchability for vessel
+if(length(VPT) != length(Q)) stop("VPT and Q must be the same length")
+fuelC <- params[["fuelC"]]
+
+######## indexes ##############
+idx <- sim_init$idx
+brk.idx <- sim_init$brk.idx
+###############################
+
+##### past knowledge decisions ####
+PastKnowledge <- params[["past_knowledge"]]  # overall flag if past knowledge used
+UseKnowledge  <- use_past_knowledge(p = logistic(Q = 200, B = 0.02/idx[["ny"]], t = t))    # specific flag for this timestep whether past knowledge being used in transition
+##################################
+
+## If its the first location fished, need to choose a random location
+if(t == 1) {
[…]

### 2022-03-29 R/go_fish_Bens.R

@@ -1,0 +1,455 @@
+#' @title Go fish
+
+#' @description \code{go_fish} is a function used to apply the fishing
+#' simulation model
+
+#' @param sim_init is the initialised object from \code{init_sim}.
+#' @param fleet_params is the parameter settings initialised from \code{_init_fleets}
+#' @param fleet_catches is the DF initialised from \code{_init_fleets}
+#' @param sp_fleet_catches is a list of spatial catches (as a Numeric matrix)
+#' for the fleet of each population @param closed_areas is a dataframe with the
+#' x,y coordinates are any closed areas, provided internally by
+#' \link{close_areas}
+#' 
+
+#' @return is a list containing i) the fleet catch dataframes , ii) the spatial
+#' catches of each population
+
+#' @examples
+#'
+
+#' @export
+
+go_fish_Bens <- function(sim_init = NULL, fleet_params = NULL, fleet_catches = NULL, 
+		    sp_fleet_catches = NULL, pops = NULL, closed_areas = NULL, t = t) {
+
+##### extract the relevant components ###########
+params <- fleet_params      # fleet parameter list
+catch  <- fleet_catches     # fleet catches list
+catch_matrix <- sp_fleet_catches
+VPT <- params[["VPT"]] 			# Value per tonne
+Q <- params[["Qs"]]			# Catchability for vessel
+if(length(VPT) != length(Q)) stop("VPT and Q must be the same length")
+fuelC <- params[["fuelC"]]
+
+######## indexes ##############
+idx <- sim_init$idx
+brk.idx <- sim_init$brk.idx
+###############################
+
+##### past knowledge decisions ####
+PastKnowledge <- params[["past_knowledge"]]  # overall flag if past knowledge used
+#UseKnowledge  <- use_past_knowledge(p = logistic(Q = 200, B = 0.02/idx[["ny"]], t = t))    # specific flag for this timestep whether past knowledge being used in transition
+##################################
+
+## If its the first location fished, need to choose a random location
+#if(t == 1) {
+
[…]

### 2022-03-29 R/init_moveCov.R

@@ -1,0 +1,46 @@
+#' @title Initialise movement covariates
+#'
+#' @description This function creates a list of covariates, to be used 
+
+#' @param sim_init is the output from the function \link{init_sim}.
+#' @param steps is a Numeric with the number of timesteps over which the
+#' covariate changes
+#' @param spp_tol is a named list (each species) with a list of mean (mu) and
+#' variance (va) for the normal distribution for thermal tolerance.
+
+#' @examples None
+
+#' @export
+
+init_moveCov <- function (sim_init = NULL, steps = 52, spp_tol = NULL) {	
+
+	## Create list of matrices to capture spatio-temporal move covariates
+
+	nx <- sim_init[["idx"]][["ncols"]]
+	ny <- sim_init[["idx"]][["nrows"]]
+
+	cov.matrix <- list()
+
+	for(i in 1:steps) {
+
+		if(i == 1) {
+	temp <-matrix(NA,nrow=nx, ncol=ny)
+	temp <-row(temp) + col(temp)  ## From South-West to North-East
+	covariate.trend<- (temp-1)/10 #define cost values
+	covariate.trend <- matrix(covariate.trend, nx , ny) #assign costs 
+	cov.matrix[[i]] <- covariate.trend
+	covariate.trend_base <- covariate.trend
+		}
+
+		if(i > 1) {
+			covariate.trend <- ifelse(row(temp) + col(temp) < ((nx + ny) / 2), #this creates a matrix where the upper half is true and the lower half is false going along the antidiagonal
+						  covariate.trend_base + (10 * (i/steps)), 
+						  covariate.trend_base - (10 * (i/steps)))
+			cov.matrix[[i]] <- covariate.trend
+			}
+}
+
+
+	return(list(cov.matrix = cov.matrix, spp_tol = spp_tol))
+
+}

### 2022-03-29 R/init_pop.R

@@ -1,0 +1,126 @@
+#' @title Initialise populations
+#'
+#' @description \code{init_pop} sets up the populations spatial distribution
+#' based on the habitat preference, starting cell and 'n' numbers of movements
+#' for all populations in the simulation.
+#'
+#' @param Bio is a named Numeric vector of the starting (total) biomass for each of the
+#' populations.
+#' @param hab is the list of Matrices with the habitat preferences created by \code{create_hab}
+#' @param spawn_areas is a list of lists, with the first level the population
+#' ("spp1" etc..) and the second the boundary coordinates (x1, x2, y1, y2) for
+#' the \code{create_spawn_hab} function
+#' @param start_cell is a list of Numeric vectors with the starting cells for
+#' the populations
+#' @param lambda is the strength that the movement distance decays at in the
+#' \code{move_prob} function
+#' @param init_move_steps is a Numeric indicating the number of movements to
+#' initialise for the population distributions
+#' @param rec_params is a list with an element for each population, containing
+#' a vector of the stock recruit parameters which must contain \strong{model},
+#' \strong{a}, \strong{b} and \strong{cv}. See \code{Recr} for details.
+#' @param rec_wk is a list with an element for each population, containing a
+#' vector of the weeks in which recruitment takes place for the population
+#' @param spwn_wk is a list with an element for each population, containing a
+#' vector of the weeks in which spawning takes place for the population
+#' @param M is a named vector, with the annual natural mortality rate for each
+#' population
+#' @param K is a named vector, with the annual growth rate for each population
+#'
+#' @return The function returns the recording vectors at the population level,
+#' the spatial matrices for the starting population densities and the
+#' demographic parameters for each population
+
+#' @examples init_pop(sim_init = sim_init, Bio = c("spp1" = 1e6, "spp2" = 2e5), hab = list(spp1 = matrix(nc = 10,
[…]

### 2022-03-29 R/init_pop_Bens.R

@@ -1,0 +1,169 @@
+#' @title Initialise populations
+#'
+#' @description \code{init_pop} sets up the populations spatial distribution
+#' based on the habitat preference, starting cell and 'n' numbers of movements
+#' for all populations in the simulation.
+#'
+#' @param Bio is a named Numeric vector of the starting (total) biomass for each of the
+#' populations.
+#' @param hab is the list of Matrices with the habitat preferences created by \code{create_hab}
+#' @param spawn_areas is a list of lists, with the first level the population
+#' ("spp1" etc..) and the second the boundary coordinates (x1, x2, y1, y2) for
+#' the \code{create_spawn_hab} function
+#' @param start_cell is a list of Numeric vectors with the starting cells for
+#' the populations
+#' @param lambda is the strength that the movement distance decays at in the
+#' \code{move_prob} function
+#' @param init_move_steps is a Numeric indicating the number of movements to
+#' initialise for the population distributions
+#' @param rec_params is a list with an element for each population, containing
+#' a vector of the stock recruit parameters which must contain \strong{model},
+#' \strong{a}, \strong{b} and \strong{cv}. See \code{Recr} for details.
+#' @param rec_wk is a list with an element for each population, containing a
+#' vector of the weeks in which recruitment takes place for the population
+#' @param spwn_wk is a list with an element for each population, containing a
+#' vector of the weeks in which spawning takes place for the population
+#' @param M is a named vector, with the annual natural mortality rate for each
+#' population
+#' @param K is a named vector, with the annual growth rate for each population
+#'
+#' @return The function returns the recording vectors at the population level,
+#' the spatial matrices for the starting population densities and the
+#' demographic parameters for each population
+
[…]

### 2022-03-29 R/init_sim_Bens_nofish.R

@@ -1,0 +1,97 @@
+#' @title Initialise simulation
+#'
+#' @description \code{init_sim} sets up the general simulation parameters such
+#' as number of tows in a day, number of days fished in a week, how often
+#' species movement occurs and number of years for the simulation. It also
+#' creates some vector and matrix structures which are used in the
+#' \code{init_pop} and \code{init_fleet} functions.
+
+#' @param n_years is an integar defining the number of years for the
+#' simulation
+#' @param n_tow_day is an integar defining the number of tows in a days fishing
+#' @param n_days_wk_fished is an integar defining the number of days in a
+#' calendar week that are fished (e.g. 5 (out of 7))
+#' @param n_fleets is an integar defining the number of fleets in the simulation
+#' @param n_vessels is an integar defining the number of vessels in each fleet
+#' @param n_species is an integar defining the number of species in the
+#' simulation
+#' @param nrows Numeric integer with the y dimension of the field in
+#' \emph{nrow * ncol}
+#' @param ncols Numeric integer with the x dimension of the field in
+#' \emph{nrow * ncol}
+#' @param move_freq is an integar defining the duration (in weeks) between
+#' spatial movements for the populations
+
+#' @return is a list of lists, detailing the indexs and data formats necessary
+#' for the simulation.
+
+#' @examples init_sim(n_years = 1, n_tows_day = 4, n_days_wk_fished = 5,
+#' n_fleets = 1, n_vessels = 1, n_species = 1, move_freq = 2)
+
+#' @export
+
+init_sim_Bens_nofish <- function(n_years = 1, n_tows_day = 4, n_days_wk_fished = 5,
+		     n_fleets = 1, n_vessels = 1, n_species = 1, nrows = nrows,
+		     ncols = ncols, move_freq = 2) {
+
+## Create an index for the simulations
+	# nrows   = number of cells in y direction
+	# ncols   = number of cells in x direction
+	# ntd     = number tows per day
+	# ndf     = number days fished
+	# nw      = number of weeks per year
+	# nwm     = number of weeks per month
[…]

### 2022-03-29 R/plot_habitat.R

@@ -1,0 +1,40 @@
+#' @title Plot the habitat distribution fields for all populations
+#'
+#' @description \code{plot_habitat} is a function to plot the habitat spatial
+#' fields.
+#'
+#' @param hab is the habitat to plot, either unadjusted or adjusted for
+#' spawning area. 
+
+#' @return Silently plots the habitat fields (can be either the unadjusted
+#' habitat, or including the spawning adjustment).
+
+#' @examples
+#' plot_habitat(hab = hab)
+
+#' @export
+
+plot_habitat <- function(hab = hab) {
+
+	n.spp <- length(hab)
+	nrows <- dim(hab[[1]])[1]
+	ncols <- dim(hab[[1]])[2]
+
+
+	
+	print(nrows)
+	show(ncols)
+	
+	show("plotting inside plot_habitat")
+	
+	par(mfrow = c(ceiling(sqrt(n.spp)), ceiling(n.spp/ceiling(sqrt(n.spp)))), mar = c(2, 2, 2, 2))
+	for(i in seq(n.spp)) {
+	image(hab[[paste0("spp", i)]], cex.axis = 1.5, cex.main = 2, col = grey(seq(1,0, l = 51)), axes = F)
+	axis(1, at = seq(0, 1, by = 0.2), labels = seq(0, nrows, by = nrows/5))
+	axis(2, at = seq(0, 1, by = 0.2), labels = seq(0, ncols, by = ncols/5))
+	text(0.5, 0.98, labels = paste("habitat spp = ", i), cex = 1)
+	}
+	
+}
+
+

### 2022-03-29 R/plot_pop_summary.R

@@ -1,0 +1,129 @@
+#' @title Plot population summary
+
+#' @description \code{plot_pop_summary} plots the four population dynamic
+#' metrics: catches, biomass, fishing mortality and recruitment.
+
+#' It can either operate at a daily timestep or an annual timestep
+
+#' @param results is an output from the function \link{run_sim}.
+#' @param timestep is a character string determining whether the plot is
+#' 'daily' or 'annual'
+#' @param save is a logical whether to save the plot
+#' @param save.location is a location (defaults to current directory)
+
+#' @return is a ggplot of all the species and metrics as a faceted plot
+
+#' examples
+#' plot_pop_summary(results = res, timestep = 'daily', save = TRUE, location = '.') 
+#' ## Not run
+
+#' @export
+
+plot_pop_summary <- function(results = res, timestep = 'daily', save = FALSE, save.location = '.') {
+  
+  n_spp <- length(results[["pop_summary"]]) 
+  res_df <- lapply(seq_len(n_spp), function(x) {
+    res_spp <- lapply(names(results[["pop_summary"]][[x]]), function(x1) {
+      x1_res <- tidyr::gather(as.data.frame(t(results[["pop_summary"]][[x]][[x1]])), key = "year", factor_key = T)
+      if(x1 == "Bio.mat" | x1 == "Bio.mat.sd") {	res_out <- data.frame("pop" = rep(paste("spp",x, sep = "_"), length.out = nrow(x1_res)), 
+                                                                       "metric" = x1, 
+                                                                       "year" = x1_res$year, 
+                                                                       "day" = rep(1:358, length.out = nrow(x1_res)),#changed 362 to 358
+                                                                       "julien_day" = seq_len(nrow(x1_res)),
+                                                                       "data" = x1_res$value) 
+      
+      return(res_out)
+      
+      }
+      if(x1 == "Rec.mat" | x1 == "Rec.mat.sd") { res_out <- data.frame("pop" = rep(paste("spp",x, sep = "_"), length.out = nrow(x1_res)), 
+                                                                       "metric" = x1, 
+                                                                       "year" = seq_len(nrow(x1_res)), 
[…]

### 2022-03-29 R/run_sim.R

@@ -1,0 +1,695 @@
+#' @title Run sim
+
+#' @description \code{run_sim} is the overarching simulation function, taking
+#' all the parameterised inputs and returning the results.
+
+#' @param sim_init is the parameterised simulation settings from
+#' \code{init_sim}
+#' @param pop_init is the parameterised populations from \code{init_pop}
+#' @param move_cov is a parameterised movement covariate object, from
+#' \code{init_moveCov}
+#' @param fleets_init is the parameterised fleets from \code{init_fleets}
+#' @param hab_init is the parameterised habitat maps from \code{create_hab}
+#' @param save_pop_bio is a logical flag to indicate if you want to record #' true spatial population at each time step (day)
+#' @param survey is the survey settings from \link{init_survey}, else NULL if no survey is due to be simulated
+#' @param closure is the spatial closure settings from \link{init_closure}
+#' else NULL if no closures are to be implemented
+
+#' @return is the results...
+
+#' @examples Not yet
+#'
+#' @export
+
+run_sim <- function(nz = NULL, MoveProb = NULL, MoveProb_spwn = NULL, sim_init = NULL, pop_init = NULL, move_cov = NULL, fleets_init = NULL, hab_init = NULL, save_pop_bio = FALSE, survey = NULL, closure = NULL,...) {
+  # Overarching function for running the simulations
+  
+  start.time <- Sys.time() # for printing runtime
+  
+  #######################
+  ####### Indices #######
+  #######################
+  
+  
+  # Extract these to make indexing easier 
+  
+  ntow         <- sim_init[["idx"]][["ntow"]] # length of t loop
+  n_fleets     <- sim_init[["idx"]][["nf"]]  
+  n_vess       <- sim_init[["idx"]][["nv"]]  
+  n_spp        <- sim_init[["idx"]][["n.spp"]]  
+  ncols        <- sim_init[["idx"]][["ncols"]]  
+  nrows        <- sim_init[["idx"]][["nrows"]]  
+  n_weeks        <- sim_init[["idx"]][["nw"]]  
+  
+  day.breaks     <- sim_init[["brk.idx"]][["day.breaks"]]  
+  week.breaks    <- sim_init[["brk.idx"]][["week.breaks"]]  
+  trip.breaks    <- sim_init[["brk.idx"]][["trip.breaks"]]  
[…]

### 2022-03-29 TestScripts/Environmental_Interp_Kern_Dens_GB_HabitatCreate.R
d

@@ -1,0 +1,827 @@
+#will create the "hab" object for mixfish sim using previously created habiats for each species
+
+#for working in just R
+setwd("C:/Users/benjamin.levy/Desktop/Github/READ-PDB-blevy2-toy/")
+
+
+
+
+loadedPackages <- c("rgdal", "data.table", "maptools","envi", "raster", "RStoolbox", "spatstat.data", "spatstat.geom", "spatstat.core")
+invisible(lapply(loadedPackages, library, character.only = TRUE))
+
+
+
+#load habitat matrices previously created and store values in hab
+
+#haddock
+Had_mat <- readRDS(file="TestScripts/Habitat_plots/Haddock/Haddock_Weighted_AdaptFalse_MATRIX.RDS")
+fields::image.plot(Had_mat)
+
+#cod
+Cod_mat <- readRDS(file="TestScripts/Habitat_plots/Cod/Cod_Weighted_AdaptFalse_MATRIX.RDS")
+fields::image.plot(Cod_mat)
+
+#yellowtail
+Yell_mat <- readRDS(file="TestScripts/Habitat_plots/YellowtailFlounder/YellowtailFlounder_Weighted_AdaptFalse_MATRIX.RDS")
+fields::image.plot(Yell_mat)
+
+hab<- list()
+hab[["hab"]][["spp1"]] <- Had_mat / sum(Had_mat,na.rm = T) #normalize like MFS does
+hab[["hab"]][["spp2"]] <- Cod_mat / sum(Cod_mat, na.rm=T)
+hab[["hab"]][["spp3"]] <- Yell_mat / sum(Yell_mat, na.rm=T)
+
+
+
+
+#CREATE HADDOCK STRATA
+strata.dir <- "C:\\Users\\benjamin.levy\\Desktop\\NOAA\\GIS_Stuff\\" # strata shape files in this directory
+# get the shapefiles
+strata.areas <- readOGR(paste(strata.dir,"Survey_strata", sep="")) #readShapePoly is deprecated; use rgdal::readOGR or sf::st_read 
+#define georges bank
+GB_Had_strata_num <- c("01130","01140","01150","01160","01170","01180","01190","01200","01210","01220","01230","01240","01250", "01290", "01300")
+#pull out indices corresponding to GB strata
+GB_strata_idx <- match(GB_Had_strata_num,strata.areas@data[["STRATUMA"]])
+#plot them
[…]

### 2022-03-29 TestScripts/Environmental_Interp_Kern_Dens_YellowtailFlounder.R

@@ -1,0 +1,854 @@
+#Following from https://cran.r-project.org/web/packages/envi/vignettes/vignette.html
+
+
+
+
+
+
+####################################
+# looking for data to use
+########################
+# 
+# test.dir <- "C:\\Users\\benjamin.levy\\Desktop\\NOAA\\GIS_Stuff\\From_Alicia\\" # strata shape files in this directory
+# library(rgdal)
+# 
+# #east coast outline
+# test <- readOGR(paste(test.dir,"EastCoast_SmoothLines", sep="")) 
+# 
+# plot(test)
+# 
+# #continents
+# test2 <- readOGR(paste(test.dir,"ne_10m_land", sep=""))
+# plot(test2)
+# 
+# 
+# test3 <- readOGR(paste(test.dir,"nw_10m_bathymetry_L_0", sep=""))
+# plot(test3)
+
+
+
+
+#noaa directory
+noaa.dir <- "C:\\Users\\benjamin.levy\\Desktop\\NOAA\\GIS_Stuff\\NOAA_Maps\\" 
+
+#Bathymetry data (depth)
+Bathy <- readOGR(paste(noaa.dir,"Bathy_Poly_Clip", sep="")) 
+par(mfrow = c(1,1),mar = c(1, 1, 1, 1))
+plot(Bathy)
+
+#floortemp data
+FloorTemp <- readOGR(paste(noaa.dir,"Seafloor_Temp_Poly_Clip", sep="")) 
+par(mfrow = c(1,1),mar = c(1, 1, 1, 1))
+plot(FloorTemp)
+
+#Lithology data
+Sediment <- readOGR(paste(noaa.dir,"Atlantic_seafloor_sediment", sep="")) 
+par(mfrow = c(1,1),mar = c(1, 1, 1, 1))
+plot(Sediment)
+
+
+
+
+###################################
+# following from online
+###################################
+
+# trying to use presence and absence data along with 2 covariates (catch weight and bottom temp)
+loadedPackages <- c("envi", "raster", "RStoolbox", "spatstat.data", "spatstat.geom", "spatstat.core")
+invisible(lapply(loadedPackages, library, character.only = TRUE))
+set.seed(1234) # for reproducibility
+
+
+View(spatstat.data::gorillas.extra)
+
+#two covariate datasets (rasters I think)
+slopeangle <- spatstat.data::gorillas.extra$slopeangle
+waterdist <- spatstat.data::gorillas.extra$waterdist
+
[…]

### 2022-03-29 TestScripts/FVCOM_extract_Bens.R

@@ -1,0 +1,1085 @@
+library(readxl)
+library(sdm)
+library(raster)
+library(sf)
+library(ncdf4)
+library(maptools)
+library(rgdal)
+library(maps)
+library(mapdata)
+library(dplyr)
+library(viridis)
+library(sp)
+library(rgeos)
+library(writexl)
+library(SDMTools)
+library(classInt)
+library(raster)
+library(akima)
+library(chron)
+library(mgcv)
+
+#This code can be used to extract depth, temp & salinity data for any year up through 2015. FVCOM changed format after that so 2016-2017 requires additional code which you will find below the -2015 code. No publicly available FVCOM data exists for years past 2017 at this time.
+
+Catch_Data=read.csv("C:\\Users\\benjamin.levy\\Desktop\\NOAA\\GIS_Stuff\\Plot_survey\\ADIOS_SV_172909_GBK_NONE_survey_dist_map_fixed.csv") #The survey catch file you are using to build your model
+
+
+#only pull out the ones you want and set them as the dates to extract
+gis=Catch_Data
+gis$CatchWt <- gis$CATCH_WT_CAL
+gis$CatchWt[is.na(gis$CatchWt)] <- 0
+gis$Year <- gis$YEAR
+gis$Longitude <- gis$LONGITUDE
+gis$Latitude <- gis$LATITUDE
+
+#replace NA with 0
+gis$CATCH_WT_CAL[is.na(gis$CATCH_WT_CAL)] <- 0
+
+#after gear change in 2008.
+Catch_Data <- gis[(gis$Year>=2009) & (gis$Year<=2015),]  #Also,format changes in 2016-2017 (see chunk below this one for extracting those dates) 
+
+#extract day 
+Catch_Data$Day <- substr(Catch_Data$TOWDATE,1,2) #2digit day in locations 1-2 of TOWDATE
+
+#extract month and change into number
+month <- substr(Catch_Data$TOWDATE,3,5) #3 letter month abbreviation in locations 3-5 of TOWDATE
+month_match <- c("JAN","FEB","MAR","APR","MAY","JUN","JUL","AUG","SEP","OCT","NOV","DEC")
+Catch_Data$Month <- match(month,month_match) #changes character abbreviation into number 1-12
+
+#merge 3 individual date columns into 1
[…]

### 2022-03-29 TestScripts/rand_samp.R

@@ -1,0 +1,75 @@
+# rand_samp.R
+# can small sample sizes and many zeros create large CIs?
+# note: no measurement error, simply examining variability due to sampling
+
+library(tidyverse)
+
+# population and sample sizes
+ntot <- 2500 # 50 by 50 grid, one quarter of general setup
+nsamp <- 10
+
+# how to fill population values
+propzero <- 0.3
+meanlog <- 20
+sdlog <- 0.6
+
+# fill population
+set.seed(141593) # make repeatable
+popzero <- rbinom(ntot, 1, (1-propzero))
+poplog <- round(rlnorm(ntot, meanlog = log(meanlog), sdlog = sdlog), 2)
+pop <- popzero * poplog
+sumpop <- sum(pop) # true population
+
+# sample the population once
+onesamp <- sample(pop, nsamp, replace = FALSE)
+meansamp <- mean(onesamp)
+sdsamp <- sd(onesamp)
+cisamp <- c(meansamp - 1.96 * sdsamp, meansamp + 1.96 * sdsamp)
+expandedsamp <- meansamp * ntot 
+expandedci <- cisamp * ntot 
+checkci <- (sumpop >= expandedci[1] & sumpop <= expandedci[2])
+
+# do a bunch of times
+nreps <- 1000
+mysamp <- rep(NA, nreps)
+mysd <- rep(NA, nreps)
+for (i in 1:nreps){
+  thissamp <- sample(pop, nsamp, replace = FALSE)
+  mysamp[i] <- mean(thissamp)
+  sdsamp[i] <- sd(thissamp)
+}
+
+df <- tibble(sampid = 1:nreps,
+             mysamp = mysamp,
+             esamp = mysamp * ntot,
+             sdsamp = sdsamp,
+             ecilow = ntot * (mysamp - 1.96 * sdsamp),
+             ecihigh = ntot * (mysamp + 1.96 * sdsamp)
+             ) %>%
+  mutate(checkci = ifelse(((sumpop >= ecilow) & (sumpop <= ecihigh)), 1, 0))
+
+# how variable are the expanded sample estimates?
+ggplot(df, aes(x = 1, y = esamp)) +
+  geom_violin() +
+  geom_abline(intercept = sumpop, color="blue") +
+  theme_bw()  
+
+summary(df$esamp)
+
+# look at all estimates and CIs compared to true pop (blue line)
[…]

### 2022-03-29 TestScripts/Environmental_Interp_Kern_Dens_GB_HabitatCreate.R
Resampling rasters

@@ -11,20 +11,110 @@ invisible(lapply(loadedPackages, library, character.only = TRUE))
 
 
 
-#load habitat matrices previously created and store values in hab
+# 
+# #load habitat matrices previously created and store values in hab
+# 
+# #haddock
+# Had_mat <- readRDS(file="TestScripts/Habitat_plots/Haddock/Haddock_Weighted_AdaptFalse_MATRIX.RDS")
+# fields::image.plot(Had_mat)
+# 
+# #cod
+# Cod_mat <- readRDS(file="TestScripts/Habitat_plots/Cod/Cod_Weighted_AdaptFalse_MATRIX.RDS")
+# fields::image.plot(Cod_mat)
+# 
+# #yellowtail
+# Yell_mat <- readRDS(file="TestScripts/Habitat_plots/YellowtailFlounder/YellowtailFlounder_Weighted_AdaptFalse_MATRIX.RDS")
+# fields::image.plot(Yell_mat)
+
+
+
+
+
+#load habitat raster previously created and increase resolution
+library(raster)
 
 #haddock
-Had_mat <- readRDS(file="TestScripts/Habitat_plots/Haddock/Haddock_Weighted_AdaptFalse_MATRIX.RDS")
-fields::image.plot(Had_mat)
+Had_ras <- readRDS(file="TestScripts/Habitat_plots/Haddock/Haddock_Weighted_AdaptFalse_RASTER.RDS")
+plot(Had_ras)
 
 #cod
-Cod_mat <- readRDS(file="TestScripts/Habitat_plots/Cod/Cod_Weighted_AdaptFalse_MATRIX.RDS")
-fields::image.plot(Cod_mat)
+Cod_ras <- readRDS(file="TestScripts/Habitat_plots/Cod/Cod_Weighted_AdaptFalse_RASTER.RDS")
+plot(Cod_ras)
 
 #yellowtail
-Yell_mat <- readRDS(file="TestScripts/Habitat_plots/YellowtailFlounder/YellowtailFlounder_Weighted_AdaptFalse_MATRIX.RDS")
-fields::image.plot(Yell_mat)
+Yell_ras <- readRDS(file="TestScripts/Habitat_plots/YellowtailFlounder/YellowtailFlounder_Weighted_AdaptFalse_RASTER.RDS")
+plot(Yell_ras)
 
+
+#alter resolution. 
[…]

### 2022-03-29 TestScripts/Environmental_Interp_Kern_Dens_GB_HabitatCreate.R
fixing

@@ -1,917 +1,0 @@
-#will create the "hab" object for mixfish sim using previously created habiats for each species
-
-#for working in just R
-setwd("C:/Users/benjamin.levy/Desktop/Github/READ-PDB-blevy2-toy/")
-
-
-
-
-loadedPackages <- c("rgdal", "data.table", "maptools","envi", "raster", "RStoolbox", "spatstat.data", "spatstat.geom", "spatstat.core")
-invisible(lapply(loadedPackages, library, character.only = TRUE))
-
-
-
-# 
-# #load habitat matrices previously created and store values in hab
-# 
-# #haddock
-# Had_mat <- readRDS(file="TestScripts/Habitat_plots/Haddock/Haddock_Weighted_AdaptFalse_MATRIX.RDS")
-# fields::image.plot(Had_mat)
-# 
-# #cod
-# Cod_mat <- readRDS(file="TestScripts/Habitat_plots/Cod/Cod_Weighted_AdaptFalse_MATRIX.RDS")
-# fields::image.plot(Cod_mat)
-# 
-# #yellowtail
-# Yell_mat <- readRDS(file="TestScripts/Habitat_plots/YellowtailFlounder/YellowtailFlounder_Weighted_AdaptFalse_MATRIX.RDS")
-# fields::image.plot(Yell_mat)
-
-
-
-
-
-#load habitat raster previously created and increase resolution
-library(raster)
-
-#haddock
-Had_ras <- readRDS(file="TestScripts/Habitat_plots/Haddock/Haddock_Weighted_AdaptFalse_RASTER.RDS")
-plot(Had_ras)
-
-#cod
-Cod_ras <- readRDS(file="TestScripts/Habitat_plots/Cod/Cod_Weighted_AdaptFalse_RASTER.RDS")
-plot(Cod_ras)
-
-#yellowtail
-Yell_ras <- readRDS(file="TestScripts/Habitat_plots/YellowtailFlounder/YellowtailFlounder_Weighted_AdaptFalse_RASTER.RDS")
-plot(Yell_ras)
-
-
-#alter resolution. 
-#Yell_ras1 <- raster::aggregate(Yell_ras,fact=2) #can only use interger factor
-
-res_factor <- .65  #amount to increase resolution
[…]

### 2022-03-29 TestScripts/GB_habitat_create.R
re-establish habitat create code

@@ -1,0 +1,922 @@
+#will create the "hab" object for mixfish sim using previously created habiats for each species
+
+#for working in just R
+setwd("C:/Users/benjamin.levy/Desktop/Github/READ-PDB-blevy2-toy/")
+
+
+
+
+loadedPackages <- c("rgdal", "data.table", "maptools","envi", "raster", "RStoolbox", "spatstat.data", "spatstat.geom", "spatstat.core")
+invisible(lapply(loadedPackages, library, character.only = TRUE))
+
+
+
+# 
+# #load habitat matrices previously created and store values in hab
+# 
+# #haddock
+# Had_mat <- readRDS(file="TestScripts/Habitat_plots/Haddock/Haddock_Weighted_AdaptFalse_MATRIX.RDS")
+# fields::image.plot(Had_mat)
+# 
+# #cod
+# Cod_mat <- readRDS(file="TestScripts/Habitat_plots/Cod/Cod_Weighted_AdaptFalse_MATRIX.RDS")
+# fields::image.plot(Cod_mat)
+# 
+# #yellowtail
+# Yell_mat <- readRDS(file="TestScripts/Habitat_plots/YellowtailFlounder/YellowtailFlounder_Weighted_AdaptFalse_MATRIX.RDS")
+# fields::image.plot(Yell_mat)
+
+
+
+
+
+#load habitat raster previously created and increase resolution
+library(raster)
+
+#haddock
+Had_ras <- readRDS(file="TestScripts/Habitat_plots/Haddock/Haddock_Weighted_AdaptFalse_RASTER.RDS")
+plot(Had_ras)
+
+#cod
+Cod_ras <- readRDS(file="TestScripts/Habitat_plots/Cod/Cod_Weighted_AdaptFalse_RASTER.RDS")
+plot(Cod_ras)
+
+#yellowtail
+Yell_ras <- readRDS(file="TestScripts/Habitat_plots/YellowtailFlounder/YellowtailFlounder_Weighted_AdaptFalse_RASTER.RDS")
+plot(Yell_ras)
+
+
+#alter resolution. 
+#Yell_ras1 <- raster::aggregate(Yell_ras,fact=2) #can only use interger factor
+
+res_factor <- .65  #amount to increase resolution
[…]

### 2022-03-29 R/create_spawn_hab_Bens.R
Created scripts, again

@@ -1,0 +1,39 @@
+#' @title create spawning habitat 
+#' 
+#' @description \code{create_spawn_hab} modifies the habitat preference maps
+#' created by \code{create_hab} to account for spawning habitat preference -
+#' can be used as a substitute during spawning periods. 
+
+#' @param hab is the habitat preference for the population 
+#' @param spwnareas is a list of Numeric vectors with the West, East, South and
+#' North dimensions of the spawning areas, in the form list(spwn1 = c(x1, x2,
+#' y1, y2)
+#' @param mult is a Numeric with the attractiveness of the spawning area (a
+#' multiplier)
+#'
+#' @return is the new habitat preference, taking account of the spawning area
+
+#' @examples create_spawn_hab(hab = matrix(nc = 100, runif(100 *
+#' 100)), spwnareas = list(spwn1 = c(20, 30, 50, 60)), mult = 10)
+#' 
+#' @export
+
+## The basic premise is to modify the habitat to bias towards the defined
+## spawning areas. Probably best to have as a separate input rather than
+## recalculate each time its needed...
+
+create_spawn_hab_Bens <- function(hab = hab, spwnareas = NULL, mult = 10) {
+
+ncols  <- ncol(hab)
+nrows  <- nrow(hab)
+
+# create a matrix of 1s with right dims
+spwn <- matrix(rep(1, nrows * ncols), nc = ncols)
+
+spwn <- define_spawn_Bens(coord = spwnareas, spwn = spwn, mult = mult)
+
+hab.spwn <- hab * spwn
+hab.spwn <- hab.spwn / sum(hab.spwn,na.rm=TRUE)
+return(hab.spwn)
+
+}

### 2022-03-29 R/define_spawn_Bens.R

@@ -1,0 +1,32 @@
+#' @title define spawning areas 
+#' 
+#' @description \code{define_spawn} is an auxiliary function called by
+#' \code{create_spawn_hab} to create the spawning habitat preferences.
+
+#' @param coord is a List of Numeric vectors of the boundaries of the spawning
+#' areas, i.e. list(spwn1 = c(x1, x2, y1, y2), spwn2 = ...)
+#' @param spwn is a Numeric matrix of 1s fed in by \code{create_spawn_hab}
+#' @param mult is a Numeric of the attractiveness of the spawning areas
+
+#' @return a matrix of spawning preference 
+
+#' @examples
+#' define_spawn(coord = list(spwn1 = c(2,4,2,4)), spwn = matrix(nc = 3, runif(9)), mult = 10)  
+
+#' @export
+
+define_spawn_Bens <- function(coord = NULL, spwn = NULL, mult = 10){
+
+	for(i in seq(length(coord[,1]))) {
+	  
+	  
+	x <- coord[i,1] 
+	y <- coord[i,2]
+	
+	spwn[x, y] <- spwn[x,y] * mult
+	}
+
+	return(spwn)
+
+}
+

### 2022-03-29 TestScripts/GB_habitat_create.R
small changes

@@ -338,8 +338,8 @@ hab[["spwn_hab"]][["spp3"]] <- YT_spwn_hab
 # hab$strata <- temp$strata
 # hab$stratas <- temp$stratas
 # hab$spwn_hab <- temp$spwn_hab$spp3
-# saveRDS(hab, file="hab_justYT.RDS")
-
+# saveRDS(hab, file="hab_justYT2.RDS")
+# 
 
 
 

### 2022-03-29 GB_Simulation_3Species.Rmd
Created scripts, again
small changes
d
changes resolution of temp maps

@@
 #LOAD HABITAT PREVIOUSLY CREATED
-hab <- readRDS(file="hab_justYT.RDS")
+#hab <- readRDS(file="hab_justYT.RDS") #finer resolution
+hab <- readRDS(file="hab_justYT2.RDS") #courser resolution
 
@@
 
+#rebuild package
+devtools::build()
+
 #load rcpp exports
@@
 
-moveCov <- readRDS(file="20 year moveCov matrices/GeorgesBank/GB_22yr_ConstTemp_HaddockStrata")
+#moveCov <- readRDS(file="20 year moveCov matrices/GeorgesBank/GB_22yr_ConstTemp_HaddockStrata")
+moveCov <- readRDS(file="20 year moveCov matrices/GeorgesBank/GB_22yr_ConstTemp_HaddockStrata_res2")
 

### 2022-03-29 TestScripts/GB_habitat_create.R

@@ -347,10 +347,29 @@ hab[["spwn_hab"]][["spp3"]] <- YT_spwn_hab
 
 
 
+
+
+
+
+#resample temp gradients in same way
+r <- raster(extent(Yell_ras), nrow = round(res_factor*nrow(Yell_ras)), ncol = round(res_factor*ncol(Yell_ras)) , crs = crs(Yell_ras))
+nrow(r)
+
+temp <- moveCov$cov.matrix
+
+for(i in seq(length(moveCov$cov.matrix))){
+  
+  moveCov$cov.matrix[[i]] <- resample(x=raster::raster(temp[[i]],ext=extent(Yell_ras),crs=crs(Yell_ras)), y=r, method="ngb")
+
+  
+}
+
+
 #INTEGRATE WITH TEMP GRADIENT
 
 #constant temp gradient
 moveCov <- readRDS(file="20 year moveCov matrices/GeorgesBank/GB_22yr_ConstTemp_HaddockStrata")
+
 
 #add temp tolerances order: had, cod, yellow
 moveCov[["spp_tol"]] <- list() #just in case

### 2022-03-29 TestScripts/moveCov_FVCOM_Data.R

@@ -5,10 +5,24 @@ library(MixFishSim)
 library(raster)
 
 #read into GB temp data
-WeeklyTempData <- readRDS(file="TestScripts/FVCOM_GB/YearlyTemp_GB.RDS")
+WeeklyTempData <- readRDS(file="C:\\Users\\benjamin.levy\\Desktop\\Github\\READ-PDB-blevy2-toy\\TestScripts\\FVCOM_GB\\YearlyTemp_GB.RDS")
 
 
+#resample temp gradients in same way as habitat maps
+res_factor <- .65 
+r <- raster(extent(WeeklyTempData$`05`[[1]]), nrow = round(res_factor*nrow(WeeklyTempData$`05`[[1]])), ncol = round(res_factor*ncol(WeeklyTempData$`05`[[1]])) , crs = crs(WeeklyTempData$`05`[[1]]))
+nrow(r)
 
+temp <- WeeklyTempData
+
+for(i in seq(length(WeeklyTempData))){
+  for(j in seq(length(WeeklyTempData[[i]]))){
+  
+  WeeklyTempData[[i]][[j]] <- resample(x=temp[[i]][[j]],crs=crs(WeeklyTempData$`05`[[1]]), y=r, method="ngb")
+  
+  }
+  
+}
 
 
 #plot results

### 2022-03-31 GB_Simulation_3Species.Rmd
updates

@@
 source("R/init_sim_Bens_nofish.R")
-sim <- init_sim_Bens_nofish(nrows = nrow(hab$hab$spp1), ncols = ncol(hab$hab$spp1), n_years = 22, n_tows_day = 1,n_days_wk_fished = 1, n_fleets = 1, n_vessels = 1, n_species = 1,
+sim <- init_sim_Bens_nofish(nrows = nrow(hab$hab$spp1), ncols = ncol(hab$hab$spp1), n_years = 22, n_tows_day = 1,n_days_wk_fished = 1, n_fleets = 1, n_vessels = 1, n_species = 3,
                             move_freq = 1)
@@
 
+#YELLOWTAIL, COD, HADDOCK
+
+STILL NEED: START CELL, REC PARAMS, K GROWTH PARAMS
+
 #decreasing population settings
-Pop <- init_pop_Bens(sim_init = sim, Bio = c("spp1" = 3194), #CHANGE
+Pop <- init_pop_Bens(sim_init = sim, Bio = c("spp1" = 3194, "spp2" = (13000+30000)/2, "spp3" = 150000*1.2), #CHANGE
                      hab = hab[["hab"]], start_cell = c(42,155),
-                     lambda = c("spp1" = 0), #set at 0 first, change after initial population set
+                     lambda = c("spp1" = 0, "spp2" = 0, "spp3" = 0), #set at 0 first, change after initial population set
                      init_move_steps = 20,
                      rec_params = list("spp1" = c("model" = "BH", "a" = 2, "b" = 4, "cv" = 0)), #CHANGE
-                     rec_wk = list("spp1" = 9:12),
-                     spwn_wk = list("spp1" = 9:12), #CORRECT FOR YTF
-                     M = c("spp1" = 0.275), #CHANGE
-                     K = c("spp1" = 0.3, "spp2" = 0.3), #CHANGE
-                     nz = nonzero_idx
-)
+                     rec_wk = list("spp1" = 9:12, "spp2" = 8:13, "spp3" = 11:14),
+                     spwn_wk = list("spp1" = 9:12, "spp2" = 8:13, "spp3" = 11:14  ), #CORRECT FOR YTF
+                     M = c("spp1" = 0.25+0, "spp2" = 0.35+0, "spp3" = 0.2+.05), #CHANGE
+                     K = c("spp1" = 0.3, "spp2" = 0.3, "spp3" = 0.3), #CHANGE
+                     nz = nonzero_idx)
 

### 2022-03-31 TestScripts/GB_habitat_create.R

@@ -121,10 +121,19 @@ Yell_ras <- Yell_ras1
 
 
 hab<- list()
-hab[["hab"]][["spp1"]] <- Had_mat / sum(Had_mat,na.rm = T) #normalize like MFS does
+hab[["hab"]][["spp1"]] <- Yell_mat / sum(Yell_mat, na.rm=T)
 hab[["hab"]][["spp2"]] <- Cod_mat / sum(Cod_mat, na.rm=T)
-hab[["hab"]][["spp3"]] <- Yell_mat / sum(Yell_mat, na.rm=T)
+hab[["hab"]][["spp3"]] <- Had_mat / sum(Had_mat,na.rm = T) #normalize like MFS does
 
+#save new rasters and matrices
+
+saveRDS(Yell_ras, file="Yell_Weighted_AdaptFalse_RASTER_res2.RDS")
+saveRDS(Cod_ras, file="Cod_Weighted_AdaptFalse_RASTER_res2.RDS")
+saveRDS(Had_ras, file="Had_Weighted_AdaptFalse_RASTER_res2.RDS")
+
+saveRDS(Yell_mat, file="Yell_Weighted_AdaptFalse_MATRIX_res2.RDS")
+saveRDS(Cod_mat, file="Cod_Weighted_AdaptFalse_MATRIX_res2.RDS")
+saveRDS(Had_mat, file="Had_Weighted_AdaptFalse_MATRIX_res2.RDS")
 
 
 
@@ -295,38 +304,94 @@ source("R/define_spawn_Bens.R")
 #yellowtail in May (weeks 9, 10, 11, 12)
 
 
-max(hab$hab$spp3,na.rm=T)  #max is 0.0006653
+max(hab$hab$spp1,na.rm=T)  #max is 0.0006653
 
-YT_spwn_ind <-which(hab$hab$spp3 >= 0 , arr.ind=T) #4,279 total non NA cells
-YT_spwn_ind <-which(hab$hab$spp3 > 0 , arr.ind=T)  #3,110 are >0
-YT_spwn_ind <-which(hab$hab$spp3 >= .0006 , arr.ind=T) #832 are above .0006
+YT_spwn_ind <-which(hab$hab$spp1 >= 0 , arr.ind=T) #4,279 total non NA cells
+YT_spwn_ind <-which(hab$hab$spp1 > 0 , arr.ind=T)  #3,110 are >0
[…]

### 2022-04-01 GB_Simulation_3Species.Rmd
Updates

@@
 
+hab <- readRDS("hab_GB_3species.RDS")
+
 #rebuild package
@@
 
+##where should starting point be? check highest values locations
+# which(hab[["hab"]][["spp1"]]==max(hab[["hab"]][["spp1"]],na.rm=T),arr.ind=T)
+# which(hab[["hab"]][["spp2"]]==max(hab[["hab"]][["spp2"]],na.rm=T),arr.ind=T)
+# which(hab[["hab"]][["spp3"]]==max(hab[["hab"]][["spp3"]],na.rm=T),arr.ind=T)
+# 
+# #make sure each have nonzero value on start cell
+# hab[["hab"]][["spp1"]][46,112]
+# hab[["hab"]][["spp2"]][30,134]
+# hab[["hab"]][["spp3"]][24,91]
+
 #YELLOWTAIL, COD, HADDOCK
 
-STILL NEED: START CELL, REC PARAMS, K GROWTH PARAMS
+STILL NEED: REC PARAMS, K GROWTH PARAMS
+
+Recr(model = "BH", params = list("a" = 1000, "b" = 450),
+   B = max(Pop[["Start_pop"]][["spp1"]],na.rm=T), cv = 0)
 
@@
 Pop <- init_pop_Bens(sim_init = sim, Bio = c("spp1" = 3194, "spp2" = (13000+30000)/2, "spp3" = 150000*1.2), #CHANGE
-                     hab = hab[["hab"]], start_cell = c(42,155),
+                     hab = hab[["hab"]], start_cell = list("spp1" = c(46,112),"spp2" = c(30,134),"spp3" = c(24,91)),
                      lambda = c("spp1" = 0, "spp2" = 0, "spp3" = 0), #set at 0 first, change after initial population set
                      init_move_steps = 20,
-                     rec_params = list("spp1" = c("model" = "BH", "a" = 2, "b" = 4, "cv" = 0)), #CHANGE
+                     rec_params = list("spp1" = c("model" = "BH", "a" = 2, "b" = 4, "cv" = 0),
+                                       "spp2" = c("model" = "BH", "a" = 2, "b" = 4, "cv" = 0),
+                                       "spp3" = c("model" = "BH", "a" = 2, "b" = 4, "cv" = 0)), #CHANGE
                      rec_wk = list("spp1" = 9:12, "spp2" = 8:13, "spp3" = 11:14),
@@
 Pop$dem_params$spp1$lambda <- .1
+Pop$dem_params$spp2$lambda <- .1
+Pop$dem_params$spp3$lambda <- .1
 
@@
 fields::image.plot(Pop$Start_pop[[1]], main = "spp1 starting biomass")
+fields::image.plot(Pop$Start_pop[[2]], main = "spp1 starting biomass")
+fields::image.plot(Pop$Start_pop[[3]], main = "spp1 starting biomass")
 
@@
 #no fishing
-fleets <- init_fleet(sim_init = sim, VPT = list("spp1" = 0, "spp2" = 0), #VPT = value per ton
-                     Qs = list("fleet 1" = c("spp1" = 0, "spp2" = 0)   #Q = catchability
+fleets <- init_fleet(sim_init = sim, VPT = list("spp1" = 0, "spp2" = 0, "spp3" = 0), #VPT = value per ton
+                     Qs = list("fleet 1" = c("spp1" = 0, "spp2" = 0, "spp3" = 0)   #Q = catchability
                      ),
@@
   assignInNamespace("go_fish", go_fish_Bens, ns = "MixFishSim")
+
+#   #rebuild package
+devtools::build()
+
+  #load rcpp exports
+Rcpp::sourceCpp(file= "src/Movement.cpp")
+Rcpp::sourceCpp(file= "src/RcppExports.cpp")
+Rcpp::compileAttributes() #this updates RcppExports.R file, which contains function definitions
 

### 2022-04-01 R/RcppExports.R

@@ -12,8 +12,8 @@
 #' @examples sum_mat(M)
 #'
 #' @export
-norm_mat <- function(M) {
-    .Call('_MixFishSim_norm_mat', PACKAGE = 'MixFishSim', M)
+norm_mat <- function(M, MM) {
+    .Call('_MixFishSim_norm_mat', PACKAGE = 'MixFishSim', M, MM)
 }
 
 #' @title distance calculation

### 2022-04-01 R/init_pop_Bens.R

@@ -81,7 +81,7 @@ Pop <- lapply(names(Bio), function(x) {
 
 		      ## Initial distribution
 		      PopIn <- matrix(nc = ncol(hab[[x]]), nr = nrow(hab[[x]]), 0)
-		      PopIn[start_cell[1], start_cell[2]] <- Bio[[x]] 
+		      PopIn[start_cell[[x]][1], start_cell[[x]][2]] <- Bio[[x]] 
 	      #1b. pull out nonzero indices
 		      NZi <- nz[[x]]
 		      # Move the population around a bit

### 2022-04-01 R/run_sim.R

@@ -55,6 +55,7 @@ run_sim <- function(nz = NULL, MoveProb = NULL, MoveProb_spwn = NULL, sim_init =
   ###################################
   
   Rec  <- vector("list", n_spp) # For storing spatial recruitment, is overwritten
+  names(Rec) <- paste0("spp", seq_len(n_spp))
   for(s in paste0("spp",seq_len(n_spp))) { Rec[[s]] <- 0 }
   B    <- pop_init[["Start_pop"]]  # For storing current biomass, is overwritten
   Bm1  <- pop_init[["Start_pop"]]  # For storing last time-step biomass, is overwritten
@@ -164,12 +165,36 @@ run_sim <- function(nz = NULL, MoveProb = NULL, MoveProb_spwn = NULL, sim_init =
     
     ## first tow in a week where any of the stocks recruit
     
+    #old method (same recruit and spawning weeks for all species)
+    
     #CHANGED THIS SO REC IS TRUE ON FIRST WEEK OF REC/SPAWN ONLY
-    Recruit  <- ifelse(t > 1, ifelse(week.breaks[t] != week.breaks[t-1] &
-                                       #IF IN FIRST SPAWN WEEK OR IN LAST SPAWN WEEK IN FIRST YEAR    
-                                       (( week.breaks[t] %in% unlist(sapply(pop_init$dem_params, function(x) x[["spwn_wk"]][[1]])))
-                                        | ( t %in% unlist(sapply(pop_init$dem_params, function(x) tail(x[["spwn_wk"]],n=1))))),
-                                     TRUE, FALSE), FALSE) 
+    # Recruit  <- ifelse(t > 1, ifelse(week.breaks[t] != week.breaks[t-1] &
+    #                                    #IF IN FIRST SPAWN WEEK OR IN LAST SPAWN WEEK IN FIRST YEAR    
+    #                                    (( week.breaks[t] %in% unlist(sapply(pop_init$dem_params, function(x) x[["spwn_wk"]][[1]])))
+    #                                     | ( t %in% unlist(sapply(pop_init$dem_params, function(x) tail(x[["spwn_wk"]],n=1))))),
+    #                                  TRUE, FALSE), FALSE) 
+    # 
+    
+    #new method (allows for different recruit and spawn weeks for each species)
+    
+    Recruit  <- lapply(paste0("spp", seq_len(n_spp)), function(s) {   
+     
[…]

### 2022-04-01 TestScripts/GB_habitat_create.R

@@ -34,15 +34,15 @@ invisible(lapply(loadedPackages, library, character.only = TRUE))
 library(raster)
 
 #haddock
-Had_ras <- readRDS(file="TestScripts/Habitat_plots/Haddock/Haddock_Weighted_AdaptFalse_RASTER.RDS")
+Had_ras <- readRDS(file="TestScripts/Habitat_plots/Haddock/Had_Weighted_AdaptFalse_RASTER_res2.RDS")
 plot(Had_ras)
 
 #cod
-Cod_ras <- readRDS(file="TestScripts/Habitat_plots/Cod/Cod_Weighted_AdaptFalse_RASTER.RDS")
+Cod_ras <- readRDS(file="TestScripts/Habitat_plots/Cod/Cod_Weighted_AdaptFalse_RASTER_res2.RDS")
 plot(Cod_ras)
 
 #yellowtail
-Yell_ras <- readRDS(file="TestScripts/Habitat_plots/YellowtailFlounder/YellowtailFlounder_Weighted_AdaptFalse_RASTER.RDS")
+Yell_ras <- readRDS(file="TestScripts/Habitat_plots/YellowtailFlounder/Yell_Weighted_AdaptFalse_RASTER_res2.RDS")
 plot(Yell_ras)
 
 

### 2022-04-01 src/Movement.cpp

@@ -19,9 +19,9 @@ using namespace Rcpp;
 //' @export
 //  [[Rcpp::export]]
 
-NumericMatrix norm_mat(NumericMatrix M) {
+NumericMatrix norm_mat(NumericMatrix M, NumericVector MM) {
   //NumericMatrix MM = M(! is_na(M));
-  NumericMatrix S = M / sum(M);
+  NumericMatrix S = M / sum(MM);
   return S;
 }
 

### 2022-04-01 src/RcppExports.cpp

@@ -11,13 +11,14 @@ Rcpp::Rostream<false>& Rcpp::Rcerr = Rcpp::Rcpp_cerr_get();
 #endif
 
 // norm_mat
-NumericMatrix norm_mat(NumericMatrix M);
-RcppExport SEXP _MixFishSim_norm_mat(SEXP MSEXP) {
+NumericMatrix norm_mat(NumericMatrix M, NumericVector MM);
+RcppExport SEXP _MixFishSim_norm_mat(SEXP MSEXP, SEXP MMSEXP) {
 BEGIN_RCPP
     Rcpp::RObject rcpp_result_gen;
     Rcpp::RNGScope rcpp_rngScope_gen;
     Rcpp::traits::input_parameter< NumericMatrix >::type M(MSEXP);
-    rcpp_result_gen = Rcpp::wrap(norm_mat(M));
+    Rcpp::traits::input_parameter< NumericVector >::type MM(MMSEXP);
+    rcpp_result_gen = Rcpp::wrap(norm_mat(M, MM));
     return rcpp_result_gen;
 END_RCPP
 }
@@ -77,7 +78,7 @@ END_RCPP
 }
 
 static const R_CallMethodDef CallEntries[] = {
-    {"_MixFishSim_norm_mat", (DL_FUNC) &_MixFishSim_norm_mat, 1},
+    {"_MixFishSim_norm_mat", (DL_FUNC) &_MixFishSim_norm_mat, 2},
     {"_MixFishSim_distance_calc", (DL_FUNC) &_MixFishSim_distance_calc, 4},
     {"_MixFishSim_move_prob", (DL_FUNC) &_MixFishSim_move_prob, 4},
     {"_MixFishSim_move_prob_Lst", (DL_FUNC) &_MixFishSim_move_prob_Lst, 3},

### 2022-04-08 TestScripts/GB_habitat_create.R
New spawning grounds

@@ -9,6 +9,9 @@ setwd("C:/Users/benjamin.levy/Desktop/Github/READ-PDB-blevy2-toy/")
 loadedPackages <- c("rgdal", "data.table", "maptools","envi", "raster", "RStoolbox", "spatstat.data", "spatstat.geom", "spatstat.core")
 invisible(lapply(loadedPackages, library, character.only = TRUE))
 
+#to rotate matrix before fields::image.plot
+rotate <- function(x) t(apply(x, 2, rev))
+
 
 
 # 
@@ -314,20 +317,20 @@ YT_spwn_ind <-which(hab$hab$spp1 >= .0006 , arr.ind=T) #832 are above .0006
 #northeast between rows 40-80 and columns 155-196 
 #use .0002 in NE corner 
 YT_spwn_ind <-which(hab$hab$spp1 >= .0006 , arr.ind=T) #832 are above .0006
-YT_spwn_NE <- YT_spwn_ind[(YT_spwn_ind[,1]>=res_factor*40) & (YT_spwn_ind[,1]<=res_factor*80) & (YT_spwn_ind[,2]>=res_factor*155) & (YT_spwn_ind[,2]<=res_factor*196), ]
+YT_spwn_NE <- YT_spwn_ind[(YT_spwn_ind[,1]>=34) & (YT_spwn_ind[,1]<=60) & (YT_spwn_ind[,2]>=95) & (YT_spwn_ind[,2]<=128), ]
 
 
 #will use southwest red area and northeast red area for spawning
 #SW between rows 96 to 127 and columns 50 to 82 
 #use .0001 in SW corner 
-YT_spwn_ind <-which(hab$hab$spp1 >= .0004 , arr.ind=T) #3,833 are above .0001
-YT_spwn_SW <- YT_spwn_ind[(YT_spwn_ind[,1]>=res_factor*96) & (YT_spwn_ind[,1]<=res_factor*127) & (YT_spwn_ind[,2]>=res_factor*50) & (YT_spwn_ind[,2]<=res_factor*82), ]
+YT_spwn_ind <-which(hab$hab$spp1 >= .0002 , arr.ind=T) #3,833 are above .0001
[…]

### 2022-04-08 GB_Simulation_3Species.Rmd
final movement and matrix normalization

@@
 ---
+
+```{r}
+library(Rcpp)
+
+#just sum matrix for denominator and then do x1 / sum_mat. 1 year took 4.059 minutes
+cppFunction('
+
+            double sum_mat(NumericVector M){
+
+            double S = sum(M);
+
+            return S;
+            }
+            ')
+
+sum(hab$hab$spp1[nonzero_idx[[1]]])
+
+sum_mat(hab$hab$spp1[nonzero_idx[[1]]])
+
+#normalize entire matrix in C++. 4.95 minutes for 1 year.
+#I think because I must use as.matrix to entier MM. Will change to vector instead below
+
+cppFunction('
+
+NumericMatrix norm_mat(NumericMatrix M, NumericMatrix MM, NumericMatrix Nzero_vals) {
+  //NumericMatrix MM = M(! is_na(M));
+  double sm = sum(MM);
+  int nNonzero = Nzero_vals.nrow();
+
+  int nrow = M.nrow();
+  int ncol = M.ncol();
+
+  NumericMatrix new_val(nrow, ncol);// for storing new values
+
+  for(int p = 0; p < nNonzero; p++)  {
+
+    new_val(Nzero_vals(p,0)-1,Nzero_vals(p,1)-1) = M(Nzero_vals(p,0)-1,Nzero_vals(p,1)-1) / sm ;
+
+  }
+
+  return new_val;
+}
+')
+
+#normalize entire matrix in C++. Send MM as vector instead of using as.matrix()
+#took 4.71 minutes to run 1 year
+#norm_mat(x1, x1[nz[[s]]],Nzero_vals = nz[[s]])
+cppFunction('
+
+NumericMatrix norm_mat(NumericMatrix M, NumericVector MM, NumericMatrix Nzero_vals) {
+
+  double sm = sum(MM);
+  int nNonzero = Nzero_vals.nrow();
+
+  int nrow = M.nrow();
+  int ncol = M.ncol();
+
+  NumericMatrix new_val(nrow, ncol);// for storing new values
+
+  for(int p = 0; p < nNonzero; p++)  {
+
+    new_val(Nzero_vals(p,0)-1,Nzero_vals(p,1)-1) = M(Nzero_vals(p,0)-1,Nzero_vals(p,1)-1) / sm ;
+
+  }
+
+  return new_val;
+}
+')
+
+#integrated the normalization into the movement command. This runs the fastest: 3.82 minutes for 1 year
+
+```
 
@@
 
+profvis::profvis({
 #decreasing population settings
@@
                      nz = nonzero_idx)
+
+})
 

### 2022-04-08 R/RcppExports.R

@@ -12,8 +12,8 @@
 #' @examples sum_mat(M)
 #'
 #' @export
-norm_mat <- function(M, MM) {
-    .Call('_MixFishSim_norm_mat', PACKAGE = 'MixFishSim', M, MM)
+norm_mat <- function(M, MM, Nzero_vals) {
+    .Call('_MixFishSim_norm_mat', PACKAGE = 'MixFishSim', M, MM, Nzero_vals)
 }
 
 #' @title distance calculation
@@ -95,7 +95,7 @@ move_prob_Lst <- function(lambda, hab, Nzero_vals) {
 #' @examples None at the moment
 #'
 #' @export
-move_population <- function(moveProp, StartPop, Nzero_vals) {
-    .Call('_MixFishSim_move_population', PACKAGE = 'MixFishSim', moveProp, StartPop, Nzero_vals)
+move_population <- function(moveProp, StartPop, Nzero_vals, HabTemp) {
+    .Call('_MixFishSim_move_population', PACKAGE = 'MixFishSim', moveProp, StartPop, Nzero_vals, HabTemp)
 }
 

### 2022-04-08 R/init_pop_Bens.R

@@ -106,7 +106,14 @@ Pop <- lapply(names(Bio), function(x) {
 		     # PopIn <- move_population(moveProp = MoveProp, StartPop = PopIn,       
 		      #                         Nzero_vals = NZi)
 		       # print(i)
-		        PopIn <- move_population(moveProp = MoveProp, StartPop = PopIn, Nzero_vals = NZi )  
+		        #PopIn <- move_population(moveProp = MoveProp, StartPop = PopIn, Nzero_vals = NZi )  
+		        
+		        # MP = lapply(lapply(MoveProp, function(x1) x1), function(x) x * move_cov_wk_spp)
+		        # View(MP)
+		        #	  print("spawning movement")
+		        PopIn <- move_population(moveProp = MoveProp, #  norm_mat(fx(as.matrix(x1)), fx(as.matrix(x1[nz[[s]]])),Nzero_vals = nz[[s]])
+		                                  StartPop = PopIn, Nzero_vals = nz[[x]], HabTemp = lapply(MoveProp, function(x1) x1[nz[[x]]]))
+		        
 		        
 		      PopIn <- Reduce("+", PopIn)
 		     # print(class(PopIn))

### 2022-04-08 R/run_sim.R

@@ -54,9 +54,42 @@ run_sim <- function(nz = NULL, MoveProb = NULL, MoveProb_spwn = NULL, sim_init =
   ##### Temporary containers ########
   ###################################
   
+  
+#   require(Rcpp)
+#   require(inline)
+#   
+#   fx <- cxxfunction( signature( x_ = "matrix" ), '
+#     NumericMatrix x(x_) ;
+#     int nr = x.nrow(), nc = x.ncol() ;
+#     std::vector< std::vector<double> > vec( nc ) ;
+#     for( int i=0; i<nc; i++){
+#         NumericMatrix::Column col = x(_,i) ;
+#         vec[i].assign( col.begin() , col.end() ) ;
+#     }
+#     // now do whatever with it
+#     // for show here is how Rcpp::wrap can wrap vector<vector<> >
+#     // back to R as a list of numeric vectors
+#     return wrap( vec ) ;
+# ', plugin = "Rcpp" )
+#   
+  
+  
+  
   Rec  <- vector("list", n_spp) # For storing spatial recruitment, is overwritten
-  names(Rec) <- paste0("spp", seq_len(n_spp))
+ 
+  Rec<- lapply(seq_len(n_spp), function(s) {
+ 
+      
+      rec <- matrix(0, ncol = ncols, nrow = nrows)
+      
+     })
+  
+  
   for(s in paste0("spp",seq_len(n_spp))) { Rec[[s]] <- 0 }
+  
+  
+  
+  
   B    <- pop_init[["Start_pop"]]  # For storing current biomass, is overwritten
   Bm1  <- pop_init[["Start_pop"]]  # For storing last time-step biomass, is overwritten
   
@@ -189,11 +222,11 @@ run_sim <- function(nz = NULL, MoveProb = NULL, MoveProb_spwn = NULL, sim_init =
     })
     
   
-    print(Recruit[[1]])
+   # print(Recruit[[1]])
     
-    print(Recruit[[2]])
+    #print(Recruit[[2]])
     
-    print(Recruit[[3]])
+  #  print(Recruit[[3]])
     
     
     if(t != ntow) {
@@ -243,13 +276,28 @@ run_sim <- function(nz = NULL, MoveProb = NULL, MoveProb_spwn = NULL, sim_init =
     #   if(week.breaks[t] == pop_init[["dem_params"]][[s]][["rec_wk"]][1]) {
  
 
-    lapply(seq_len(n_spp), function(s) {
[…]

### 2022-04-08 src/Movement.cpp

@@ -19,10 +19,27 @@ using namespace Rcpp;
 //' @export
 //  [[Rcpp::export]]
 
-NumericMatrix norm_mat(NumericMatrix M, NumericVector MM) {
+NumericMatrix norm_mat(NumericMatrix M, NumericMatrix MM, NumericMatrix Nzero_vals) {
   //NumericMatrix MM = M(! is_na(M));
-  NumericMatrix S = M / sum(MM);
-  return S;
+  double sm = sum(MM);
+  int nNonzero = Nzero_vals.nrow();
+  
+  int nrow = M.nrow();
+  int ncol = M.ncol();
+  
+ 
+  
+  NumericMatrix new_val(nrow, ncol);// for storing new values
+  
+  for(int p = 0; p < nNonzero; p++)  {
+    
+    new_val(Nzero_vals(p,0)-1,Nzero_vals(p,1)-1) = M(Nzero_vals(p,0)-1,Nzero_vals(p,1)-1) / sm ;
+    
+    
+  }
+  
+  
+  return new_val;
 }
 
 
@@ -206,12 +223,14 @@ List move_prob_Lst(double lambda, NumericMatrix hab, NumericMatrix Nzero_vals) {
 //'
 //' @export
 // [[Rcpp::export]]
-List move_population(List moveProp, NumericMatrix StartPop, NumericMatrix Nzero_vals) {
+List move_population(List moveProp, NumericMatrix StartPop, NumericMatrix Nzero_vals, List HabTemp) {
   
   // Define matrix size
   int nrow = StartPop.nrow();
   int ncol = StartPop.ncol();
   int n = Nzero_vals.nrow();
+  
+
   
   // ONLY LOOP THROUGH CELLS WITH NONZERO VALUES
   int nNonzero = Nzero_vals.nrow(); //number of nonzero indices for given species
@@ -239,6 +258,11 @@ List move_population(List moveProp, NumericMatrix StartPop, NumericMatrix Nzero_
 
         //ONLY ALLOCATE THESE MATRICES IF WILL USE THEM
         NumericMatrix Props = moveProp[counter];  // update probs
+      
+      
+      NumericVector MM = HabTemp[counter];
+      //sum of hab*temp matrix for normalizing
+      double sm = sum(MM);
 
         // Calculate the distribution of the population from the cell 
         
[…]

### 2022-04-08 src/RcppExports.cpp

@@ -11,14 +11,15 @@ Rcpp::Rostream<false>& Rcpp::Rcerr = Rcpp::Rcpp_cerr_get();
 #endif
 
 // norm_mat
-NumericMatrix norm_mat(NumericMatrix M, NumericVector MM);
-RcppExport SEXP _MixFishSim_norm_mat(SEXP MSEXP, SEXP MMSEXP) {
+NumericMatrix norm_mat(NumericMatrix M, NumericMatrix MM, NumericMatrix Nzero_vals);
+RcppExport SEXP _MixFishSim_norm_mat(SEXP MSEXP, SEXP MMSEXP, SEXP Nzero_valsSEXP) {
 BEGIN_RCPP
     Rcpp::RObject rcpp_result_gen;
     Rcpp::RNGScope rcpp_rngScope_gen;
     Rcpp::traits::input_parameter< NumericMatrix >::type M(MSEXP);
-    Rcpp::traits::input_parameter< NumericVector >::type MM(MMSEXP);
-    rcpp_result_gen = Rcpp::wrap(norm_mat(M, MM));
+    Rcpp::traits::input_parameter< NumericMatrix >::type MM(MMSEXP);
+    Rcpp::traits::input_parameter< NumericMatrix >::type Nzero_vals(Nzero_valsSEXP);
+    rcpp_result_gen = Rcpp::wrap(norm_mat(M, MM, Nzero_vals));
     return rcpp_result_gen;
 END_RCPP
 }
@@ -64,25 +65,26 @@ BEGIN_RCPP
 END_RCPP
 }
 // move_population
-List move_population(List moveProp, NumericMatrix StartPop, NumericMatrix Nzero_vals);
-RcppExport SEXP _MixFishSim_move_population(SEXP movePropSEXP, SEXP StartPopSEXP, SEXP Nzero_valsSEXP) {
+List move_population(List moveProp, NumericMatrix StartPop, NumericMatrix Nzero_vals, List HabTemp);
+RcppExport SEXP _MixFishSim_move_population(SEXP movePropSEXP, SEXP StartPopSEXP, SEXP Nzero_valsSEXP, SEXP HabTempSEXP) {
 BEGIN_RCPP
     Rcpp::RObject rcpp_result_gen;
     Rcpp::RNGScope rcpp_rngScope_gen;
     Rcpp::traits::input_parameter< List >::type moveProp(movePropSEXP);
[…]

### 2022-04-14 TestScripts/param_calc.R
calculates various input parameters.

@@ -1,6 +1,10 @@
-#This script reads in weight info (avg weight by age and avg number by age) for each species to calculate weighted mean of adults
-#This value is used to calculate phi0 and in delay_diff for MFS
+#This script calculates various parameters for mixfishsim input
 
+##################################################################
+#first calculate weighted average of prerecruits and adults
+#reads in weight info (avg weight by age and avg number by age) for each species to calculate weighted mean of adults
+#This value is used to calculate phi0 and in delay_diff for MFS
+##################################################################
 
 
 #1- YELLOWTAIL FLOUNDER
@@ -179,6 +183,130 @@ Had_adult_weightmean <- sum(Had_FRW_mean*Had_FRN_mean)/sum(Had_FRN_mean)
 Had_prerec_weightmean <- sum(Had_PRN_mean*Had_PRW_mean)/sum(Had_PRN_mean)
 
 
+
+
+
+
+
+################################################################################
+# calculate weighted fishing mortality F for adults and prerecruits
+################################################################################
+
+
+
+#1- YELLOWTAIL FLOUNDER
+
+YT_all_F <- read.csv("C:\\Users\\benjamin.levy\\Desktop\\NOAA\\Stock Assessment Review Committee (SARC) Docs\\YT Fishing Mort Info.csv",header=TRUE,stringsAsFactors=FALSE)
+
+YT_row_F <- nrow(YT_all_F) #total number of observations (top half for weight bottom half for corresponding numbers)
+
+YT_PreRecAge <- 1 #age before they are recruited (1)
+YT_AdultAge <-  2 #age when they become adults (2)
+
+YT_F <- YT_all_F[1:(YT_row/2),] #weight by age
+YT_N <- YT_all_F[(YT_row/2+1):YT_row,] #number by age
+
[…]

### 2022-04-14 GB_Simulation_3Species.Rmd
Updates

@@
 #hab <- readRDS(file="hab_justYT.RDS") #finer resolution
-hab <- readRDS(file="hab_justYT2.RDS") #courser resolution
+#hab <- readRDS(file="hab_justYT2.RDS") #courser resolution
+
+hab <- readRDS(file="hab_GB_3species.RDS") #courser resolution
 
@@
 
+par(mar=c(1,1,1,1))
+fields::image.plot(hab[["hab"]][["spp2"]])
+
+par(mar=c(1,1,1,1))
+fields::image.plot(hab[["hab"]][["spp3"]])
+
 #plot spawn habitat
@@
 fields::image.plot(hab[["spwn_hab"]][["spp1"]])
+
+par(mar=c(1,1,1,1))
+fields::image.plot(hab[["spwn_hab"]][["spp2"]])
+
+par(mar=c(1,1,1,1))
+fields::image.plot(hab[["spwn_hab"]][["spp3"]])
 
@@
 
-hab <- readRDS("hab_GB_3species.RDS")
+#hab <- readRDS("hab_GB_3species.RDS")
 
@@
 
-STILL NEED: REC PARAMS, K GROWTH PARAMS
-
-Recr(model = "BH", params = list("a" = 1000, "b" = 450),
-   B = max(Pop[["Start_pop"]][["spp1"]],na.rm=T), cv = 0)
-
-profvis::profvis({
+#profvis::profvis({
 #decreasing population settings
@@
                      init_move_steps = 20,
-                     rec_params = list("spp1" = c("model" = "BH", "a" = 2, "b" = 4, "cv" = 0),
-                                       "spp2" = c("model" = "BH", "a" = 2, "b" = 4, "cv" = 0),
-                                       "spp3" = c("model" = "BH", "a" = 2, "b" = 4, "cv" = 0)), #CHANGE
+                     rec_params = list("spp1" = c("model" = "BH", "a" = 108582, "b" = 4301, "cv" = 0),
+                                       "spp2" = c("model" = "BH", "a" = 25922 , "b" = 3491, "cv" = 0),
+                                       "spp3" = c("model" = "BH", "a" = 157783, "b" = 40530, "cv" = 0)), #CHANGE
                      rec_wk = list("spp1" = 9:12, "spp2" = 8:13, "spp3" = 11:14),
-                     spwn_wk = list("spp1" = 9:12, "spp2" = 8:13, "spp3" = 11:14  ), #CORRECT FOR YTF
-                     M = c("spp1" = 0.25+0, "spp2" = 0.35+0, "spp3" = 0.2+.05), #CHANGE
-                     K = c("spp1" = 0.3, "spp2" = 0.3, "spp3" = 0.3), #CHANGE
+                     spwn_wk = list("spp1" = 9:12, "spp2" = 8:13, "spp3" = 11:14  ), 
+                     M = c("spp1" = 1.478+.358, "spp2" = 1.299+.511, "spp3" = 1.097+.7985), #NEED F
+                     K = c("spp1" = 1.417, "spp2" = 1.801, "spp3" = 1.422), 
                      nz = nonzero_idx)
 
-})
+#})
 

### 2022-04-15 GB_Simulation_3Species.Rmd
Updated parameters

@@
                      init_move_steps = 20,
-                     rec_params = list("spp1" = c("model" = "BH", "a" = 108582, "b" = 4301, "cv" = 0),
-                                       "spp2" = c("model" = "BH", "a" = 25922 , "b" = 3491, "cv" = 0),
-                                       "spp3" = c("model" = "BH", "a" = 157783, "b" = 40530, "cv" = 0)), #CHANGE
+                     rec_params = list("spp1" = c("model" = "BH", "a" = 30445, "b" = 4301, "cv" = 0),
+                                       "spp2" = c("model" = "BH", "a" = 27868 , "b" = 3491, "cv" = 0),
+                                       "spp3" = c("model" = "BH", "a" = 73568, "b" = 40530, "cv" = 0)), #CHANGE
                      rec_wk = list("spp1" = 9:12, "spp2" = 8:13, "spp3" = 11:14),
                      spwn_wk = list("spp1" = 9:12, "spp2" = 8:13, "spp3" = 11:14  ), 
-                     M = c("spp1" = 1.478+.358, "spp2" = 1.299+.511, "spp3" = 1.097+.7985), #NEED F
-                     K = c("spp1" = 1.417, "spp2" = 1.801, "spp3" = 1.422), 
+                     M = c("spp1" = .2064+.358, "spp2" = .2728+.511, "spp3" = .334+.45), #NEED F
+                     K = c("spp1" = .2295, "spp2" = .16, "spp3" = .2465), 
                      nz = nonzero_idx)

### 2022-04-15 TestScripts/param_calc.R

@@ -320,15 +320,15 @@ Cod_prerec_Fmean <- sum(Cod_PRN_mean*Cod_PRF_mean)/sum(Cod_PRN_mean)
 
 phi0 <- function(M = NULL, Adult_weight = NULL, PreRecruit_weight = NULL){
   
-  p0 <- exp(M)*(Adult_weight/PreRecruit_weight)
+  p0 <- exp(-M)*(Adult_weight/PreRecruit_weight)
   return(p0)
 }
 
-YT_phi0 <- phi0(M = -1.478, Adult_weight = .39, PreRecruit_weight = .13)
+YT_phi0 <- phi0(M = .2064, Adult_weight = .39, PreRecruit_weight = .13)
 
-Cod_phi0 <- phi0(M = -1.299, Adult_weight = 2.95, PreRecruit_weight = .39)
+Cod_phi0 <- phi0(M = .2728, Adult_weight = 2.95, PreRecruit_weight = .39)
 
-Had_phi0 <- phi0(M = -1.097, Adult_weight = 1.12, PreRecruit_weight = .19)
+Had_phi0 <- phi0(M = .334, Adult_weight = 1.12, PreRecruit_weight = .19)
 
 
 #define R0 for each species (see notes document for details on values)

### 2022-04-17 TestScripts/param_calc.R
updated parameters

@@ -373,13 +373,13 @@ beta(h= YT_h, R0= YT_R0, phi0= YT_phi0)
 
 #Cod
 Cod_h<- .793
-  
+
 alpha(h= Cod_h, R0 = Cod_R0)
 beta(h= Cod_h, R0= Cod_R0, phi0= Cod_phi0)
 
 #Haddock
 Had_h <- .657
-  
+
 alpha(h= Had_h, R0 = Had_R0)
 beta(h= Had_h, R0= Had_R0 , phi0= Had_phi0)
 
@@ -390,19 +390,19 @@ beta(h= Had_h, R0= Had_R0 , phi0= Had_phi0)
 
 #yellowtail biomass range
 YT_range <- seq(0,90000)
-YT_params <- list("a"=108582,"b"=4301)  #alph,beta
+YT_params <- list("a"=30445,"b"=4301)  #alph,beta
 
-Cod_range <- seq(0,200000)
-Cod_params <- list("a"=77766,"b"=10472)  #alph,beta
+Cod_range <- seq(0,170000)
+Cod_params <- list("a"=27868,"b"=10472)  #alph,beta
 
 
 Had_range <- seq(0,300000)
-Had_params <- list("a"=157783,"b"=40530)  #alph,beta
+Had_params <- list("a"=73568,"b"=40530)  #alph,beta
 
 
 #change range, params and abline input
-range <- YT_range
-params_ <- YT_params
+range <- Cod_range
+params_ <- Cod_params
 rec <- vector()
 for(i in range){
   rec[i] <- Recr(model = "BH", params = params_, B = i, cv = 0)
@@ -411,7 +411,7 @@ for(i in range){
 plot(rec)
 
 #add intersection line
-abline(0,1/YT_phi0)
+abline(0,1/Cod_phi0)
 
 
 

### 2022-04-17 GB_Simulation_3Species.Rmd
added weight for adult and prerecruit to population dynamics

@@
                      K = c("spp1" = .2295, "spp2" = .16, "spp3" = .2465),
+                     Weight_PreRecruit = (c("spp1" = .13, "spp2" = 2.95, "spp3" = .19)),
+                     Weight_Adult = (c("spp1" = .39, "spp2" = .39, "spp3" = 1.12)),
                      nz = nonzero_idx)

### 2022-04-17 R/init_pop_Bens.R

@@ -42,7 +42,7 @@
 
 #' @export
 
-init_pop_Bens <- function(nz = NULL,sim_init = sim_init, Bio = NULL, hab = NULL, start_cell = NULL, lambda = NULL, init_move_steps = 10, rec_params = NULL, rec_wk = NULL, spwn_wk = NULL, M = NULL, K = NULL, cores = 3) {
+init_pop_Bens <- function(nz = NULL,sim_init = sim_init, Bio = NULL, hab = NULL, start_cell = NULL, lambda = NULL, init_move_steps = 10, rec_params = NULL, rec_wk = NULL, spwn_wk = NULL, M = NULL, K = NULL, cores = 3, Weight_PreRecruit = NULL, Weight_Adult = NULL) {
 
 # extract the indices
 idx <- sim_init[["idx"]]
@@ -158,7 +158,7 @@ names(Pop_vec) <- paste("spp",seq(idx[["n.spp"]]), sep ="")
 
 dem_params <- lapply(names(Bio), function(x) {
 
-dem_params = list(rec_params = rec_params[[x]], rec_wk = rec_wk[[x]], spwn_wk = spwn_wk[[x]], M = M[[x]], K = K[[x]], lambda = lambda[[x]])
+dem_params = list(rec_params = rec_params[[x]], rec_wk = rec_wk[[x]], spwn_wk = spwn_wk[[x]], M = M[[x]], K = K[[x]], lambda = lambda[[x]], Weight_PreRecruit =  Weight_PreRecruit[[x]], Weight_Adult = Weight_Adult[[x]] )
 
 return(dem_params)
 		  

### 2022-04-17 R/run_sim.R
New delay_diff input

@@ -54,42 +54,8 @@ run_sim <- function(nz = NULL, MoveProb = NULL, MoveProb_spwn = NULL, sim_init =
   ##### Temporary containers ########
   ###################################
   
-  
-#   require(Rcpp)
-#   require(inline)
-#   
-#   fx <- cxxfunction( signature( x_ = "matrix" ), '
-#     NumericMatrix x(x_) ;
-#     int nr = x.nrow(), nc = x.ncol() ;
-#     std::vector< std::vector<double> > vec( nc ) ;
-#     for( int i=0; i<nc; i++){
-#         NumericMatrix::Column col = x(_,i) ;
-#         vec[i].assign( col.begin() , col.end() ) ;
-#     }
-#     // now do whatever with it
-#     // for show here is how Rcpp::wrap can wrap vector<vector<> >
-#     // back to R as a list of numeric vectors
-#     return wrap( vec ) ;
-# ', plugin = "Rcpp" )
-#   
-  
-  
-  
   Rec  <- vector("list", n_spp) # For storing spatial recruitment, is overwritten
- 
-  Rec<- lapply(seq_len(n_spp), function(s) {
- 
-      
-      rec <- matrix(0, ncol = ncols, nrow = nrows)
-      
-     })
-  
-  
   for(s in paste0("spp",seq_len(n_spp))) { Rec[[s]] <- 0 }
-  
-  
-  
-  
   B    <- pop_init[["Start_pop"]]  # For storing current biomass, is overwritten
   Bm1  <- pop_init[["Start_pop"]]  # For storing last time-step biomass, is overwritten
   
@@ -139,18 +105,18 @@ run_sim <- function(nz = NULL, MoveProb = NULL, MoveProb_spwn = NULL, sim_init =
   names(MoveProb_spwn) <- paste0("spp", seq_len(n_spp))
   
   
-  # 
-  # #CALULATE INDICES OF NONZERO VALUES IN HAB TO PASS TO MOVE_POPULAITON DURING MOVEMENT
-  # nonzero_idx <- lapply(paste0("spp", seq_len(n_spp)), function(s) {
-  #  
-  #   which(hab_init[["hab"]][[s]] !=0 , arr.ind=T)
-  #  
-  # })
-  # 
-  # names(nonzero_idx) <- paste("spp",seq(sim_init[["idx"]][["n.spp"]]), sep ="")
   
-# View(nonzero_idx)
-
[…]

### 2022-04-17 R/run_sim.R
new delay_diff input

@@ -54,8 +54,42 @@ run_sim <- function(nz = NULL, MoveProb = NULL, MoveProb_spwn = NULL, sim_init =
   ##### Temporary containers ########
   ###################################
   
+  
+#   require(Rcpp)
+#   require(inline)
+#   
+#   fx <- cxxfunction( signature( x_ = "matrix" ), '
+#     NumericMatrix x(x_) ;
+#     int nr = x.nrow(), nc = x.ncol() ;
+#     std::vector< std::vector<double> > vec( nc ) ;
+#     for( int i=0; i<nc; i++){
+#         NumericMatrix::Column col = x(_,i) ;
+#         vec[i].assign( col.begin() , col.end() ) ;
+#     }
+#     // now do whatever with it
+#     // for show here is how Rcpp::wrap can wrap vector<vector<> >
+#     // back to R as a list of numeric vectors
+#     return wrap( vec ) ;
+# ', plugin = "Rcpp" )
+#   
+  
+  
+  
   Rec  <- vector("list", n_spp) # For storing spatial recruitment, is overwritten
+ 
+  Rec<- lapply(seq_len(n_spp), function(s) {
+ 
+      
+      rec <- matrix(0, ncol = ncols, nrow = nrows)
+      
+     })
+  
+  
   for(s in paste0("spp",seq_len(n_spp))) { Rec[[s]] <- 0 }
+  
+  
+  
+  
   B    <- pop_init[["Start_pop"]]  # For storing current biomass, is overwritten
   Bm1  <- pop_init[["Start_pop"]]  # For storing last time-step biomass, is overwritten
   
@@ -105,18 +139,18 @@ run_sim <- function(nz = NULL, MoveProb = NULL, MoveProb_spwn = NULL, sim_init =
   names(MoveProb_spwn) <- paste0("spp", seq_len(n_spp))
   
   
+  # 
+  # #CALULATE INDICES OF NONZERO VALUES IN HAB TO PASS TO MOVE_POPULAITON DURING MOVEMENT
+  # nonzero_idx <- lapply(paste0("spp", seq_len(n_spp)), function(s) {
+  #  
+  #   which(hab_init[["hab"]][[s]] !=0 , arr.ind=T)
+  #  
+  # })
+  # 
+  # names(nonzero_idx) <- paste("spp",seq(sim_init[["idx"]][["n.spp"]]), sep ="")
   
[…]

### 2022-04-17 GB_Simulation_3Species.Rmd
weight error

--- 2022-04-15 GB_Simulation_3Species.Rmd
@@
                      K = c("spp1" = .2295, "spp2" = .16, "spp3" = .2465),
+                     Weight_PreRecruit = (c("spp1" = .13, "spp2" = .39, "spp3" = .19)),
+                     Weight_Adult = (c("spp1" = .39, "spp2" = 2.95, "spp3" = 1.12)),
                      nz = nonzero_idx)

### 2022-04-18 GB_Simulation_3Species.Rmd
adult weight 1

--- 2022-04-15 GB_Simulation_3Species.Rmd
@@
                      K = c("spp1" = .2295, "spp2" = .16, "spp3" = .2465),
+                     Weight_PreRecruit = (c("spp1" = .13/.39, "spp2" = .39/2.95, "spp3" = .19/1.12)),
+                     Weight_Adult = (c("spp1" = 1, "spp2" = 1, "spp3" = 1)),
                      nz = nonzero_idx)
@@
 #})
+# actual weights
+#          Weight_PreRecruit = (c("spp1" = .13, "spp2" = .39, "spp3" = .19)),
+ #                    Weight_Adult = (c("spp1" = .39, "spp2" = 2.95, "spp3" = 1.12)),
 

### 2022-04-19 GB_Simulation_3Species.Rmd
working on delay_diff

@@
 source("R/init_sim_Bens_nofish.R")
-sim <- init_sim_Bens_nofish(nrows = nrow(hab$hab$spp1), ncols = ncol(hab$hab$spp1), n_years = 22, n_tows_day = 1,n_days_wk_fished = 1, n_fleets = 1, n_vessels = 1, n_species = 3,
+sim <- init_sim_Bens_nofish(nrows = nrow(hab$hab$spp1), ncols = ncol(hab$hab$spp1), n_years = 5, n_tows_day = 1,n_days_wk_fished = 1, n_fleets = 1, n_vessels = 1, n_species = 3,
                             move_freq = 1)
@@
                      M = c("spp1" = .2064+.358, "spp2" = .2728+.511, "spp3" = .334+.45), #NEED F
-                     K = c("spp1" = .2295, "spp2" = .16, "spp3" = .2465),
+                   K = c("spp1" = -log((1+exp(-.2295))^(1/52)-1), "spp2" = -log((1+exp(-.16))^(1/52)-1), "spp3" = -log((1+exp(-.2465))^(1/52)-1)),
+                     #  K = c("spp1" = .2295, "spp2" = .16, "spp3" = .2465),
                      Weight_PreRecruit = (c("spp1" = .13/.39, "spp2" = .39/2.95, "spp3" = .19/1.12)),

### 2022-04-19 R/run_sim.R

@@ -222,11 +222,11 @@ run_sim <- function(nz = NULL, MoveProb = NULL, MoveProb_spwn = NULL, sim_init =
     })
     
   
-   # print(Recruit[[1]])
-    
-    #print(Recruit[[2]])
-    
-  #  print(Recruit[[3]])
+    print(Recruit[[1]])
+
+  print(Recruit[[2]])
+
+   print(Recruit[[3]])
     
     
     if(t != ntow) {
@@ -317,11 +317,22 @@ run_sim <- function(nz = NULL, MoveProb = NULL, MoveProb_spwn = NULL, sim_init =
 
                     #NEED TO ACCOUNT FOR ALPHA IN POP DYNMAICS BY SCALING UP RECRUITMENT FOR 1 WEEK
 
-                    rec <- length(pop_init[["dem_params"]][[s]][["rec_wk"]])*Recr_mat(model = pop_init[["dem_params"]][[s]][["rec_params"]][["model"]],
+                    # rec <- length(pop_init[["dem_params"]][[s]][["rec_wk"]])*Recr_mat(model = pop_init[["dem_params"]][[s]][["rec_params"]][["model"]],
+                    #                                                                   params = c("a" = as.numeric(pop_init[["dem_params"]][[s]][["rec_params"]][["a"]]),
+                    #                                                                              "b" = as.numeric(pop_init[["dem_params"]][[s]][["rec_params"]][["b"]])),
+                    #                                                                   B = B[[s]], #THIS IS NEW. USE CURRENT POP IN FIRST YEAR
+                    #                                                                   cv = as.numeric(pop_init[["dem_params"]][[s]][["rec_params"]][["cv"]])),
+                    
+                    
+                    #TRYING RECR.R INSTEAD OF RECR_MAT.R
+                    {SB <- sum(B[[s]],na.rm=T)
+                    rec1 <- length(pop_init[["dem_params"]][[s]][["rec_wk"]])*Recr(model = pop_init[["dem_params"]][[s]][["rec_params"]][["model"]],
                                                                                       params = c("a" = as.numeric(pop_init[["dem_params"]][[s]][["rec_params"]][["a"]]),
[…]

### 2022-04-19 TestScripts/param_calc.R

@@ -401,8 +401,8 @@ Had_params <- list("a"=73568,"b"=40530)  #alph,beta
 
 
 #change range, params and abline input
-range <- Cod_range
-params_ <- Cod_params
+range <- YT_range
+params_ <- YT_params
 rec <- vector()
 for(i in range){
   rec[i] <- Recr(model = "BH", params = params_, B = i, cv = 0)

### 2022-04-19 GB_Simulation_3Species.Rmd
Added comment about rebuilding package for use on computer vs server

@@
 
-  #load rcpp exports
+  #load rcpp exports ON SERVER ONLY. WONT WORK ON PC
 Rcpp::sourceCpp(file= "src/Movement.cpp")
@@
 
+  #ON PC JUST 1) REBUILD PACKAGE (ABOVE) THEN 2) RELOAD PACKAGE 
   library(MixFishSim) #each core needs to load library

### 2022-04-19 GB_Simulation_3Species.Rmd
COrrected cod BH beta parameter

@@
                      rec_params = list("spp1" = c("model" = "BH", "a" = 30445, "b" = 4301, "cv" = 0),
-                                       "spp2" = c("model" = "BH", "a" = 27868 , "b" = 3491, "cv" = 0),
+                                       "spp2" = c("model" = "BH", "a" = 27868 , "b" = 10472, "cv" = 0),
                                        "spp3" = c("model" = "BH", "a" = 73568, "b" = 40530, "cv" = 0)), #CHANGE

### 2022-04-19 R/run_sim.R
Update recruitment storage

@@ -276,7 +276,7 @@ run_sim <- function(nz = NULL, MoveProb = NULL, MoveProb_spwn = NULL, sim_init =
     #   if(week.breaks[t] == pop_init[["dem_params"]][[s]][["rec_wk"]][1]) {
  
 
-    Rec<- lapply(seq_len(n_spp), function(s) {
+    Rec <- lapply(seq_len(n_spp), function(s) {
    #   print(s)
 
       #if outside recruitment weeks or in first year, set as matrix of zeros
@@ -396,7 +396,6 @@ run_sim <- function(nz = NULL, MoveProb = NULL, MoveProb_spwn = NULL, sim_init =
       print("sum rec")
       print("spp3")
       print(sum(Rec[["spp3"]]))
-
 
   
   
@@ -852,9 +851,21 @@ run_sim <- function(nz = NULL, MoveProb = NULL, MoveProb_spwn = NULL, sim_init =
       #  print( "here7")
         pop_init[["Pop_record"]][[s]][["Bio.mat"]][year.breaks[t], day.breaks[t]] <- sum(Bp1[[s]])
       #  print( "here8")
-        pop_init[["Pop_record"]][[s]][["Rec.mat"]][1, year.breaks[t]] <- sum(Rec[[s]], pop_init[["Pop_record"]][[s]][["Rec.mat"]][1, year.breaks[t]], na.rm = T)
-       # print( "here9")
         
+        #only update in first week of recruitment or last week of first year, which will have total for the year
+        if(( week.breaks[t] == pop_init$dem_params[[s]]$rec_wk[1] )
+          | ( t == tail(pop_init$dem_params[[s]]$rec_wk,n=1))){
+          
+       
+        pop_init[["Pop_record"]][[s]][["Rec.mat"]][1, year.breaks[t]] <- sum(Rec[[s]])
+        
+        #if first year divide by length of rec weeks because it all is sent out in 1 week
+        if(t == tail(pop_init$dem_params[[s]]$rec_wk,n=1)){
+        pop_init[["Pop_record"]][[s]][["Rec.mat"]][1, year.breaks[t]] <- sum(Rec[[s]])/ length(pop_init[["dem_params"]][[s]][["rec_wk"]])
+       }
+        
+       # print( "here9")
+        }
       }
       
       

### 2022-04-19 R/run_sim.R
Comment out printing things

@@ -222,12 +222,12 @@ run_sim <- function(nz = NULL, MoveProb = NULL, MoveProb_spwn = NULL, sim_init =
     })
     
   
-    print(Recruit[[1]])
-
-  print(Recruit[[2]])
-
-   print(Recruit[[3]])
-    
+  #   print(Recruit[[1]])
+  # 
+  # print(Recruit[[2]])
+  # 
+  #  print(Recruit[[3]])
+  #   
     
     if(t != ntow) {
       Pop_dyn  <- ifelse(day.breaks[t] != day.breaks[t+1], TRUE, FALSE) ## weekly delay diff
@@ -387,15 +387,15 @@ run_sim <- function(nz = NULL, MoveProb = NULL, MoveProb_spwn = NULL, sim_init =
     names(Rec) <- paste0("spp", seq_len(n_spp))
    
     #  View(Rec)
-      print("sum rec")
-      print("spp1")
-      print(sum(Rec[["spp1"]]))
-      print("sum rec")
-      print("spp2")
-      print(sum(Rec[["spp2"]]))
-      print("sum rec")
-      print("spp3")
-      print(sum(Rec[["spp3"]]))
+      # print("sum rec")
+      # print("spp1")
+      # print(sum(Rec[["spp1"]]))
+      # print("sum rec")
+      # print("spp2")
+      # print(sum(Rec[["spp2"]]))
+      # print("sum rec")
+      # print("spp3")
+      # print(sum(Rec[["spp3"]]))
 
   
   
@@ -529,13 +529,13 @@ run_sim <- function(nz = NULL, MoveProb = NULL, MoveProb_spwn = NULL, sim_init =
     # Need B-1 and B to calc B+1
     
    # print( "here1")
-      print("pre pop dynamics")
-      print("sum of Spp1")
-      print(sum(B[[1]]))
-      print("sum of Spp2")
-      print(sum(B[[2]]))
-      print("sum of Spp3")
-      print(sum(B[[3]]))
+      # print("pre pop dynamics")
+      # print("sum of Spp1")
+      # print(sum(B[[1]]))
+      # print("sum of Spp2")
+      # print(sum(B[[2]]))
+      # print("sum of Spp3")
+      # print(sum(B[[3]]))
     
     if(Pop_dyn) { #POP_DYN IS SET ON LINE 143
       
[…]

### 2022-04-19 GB_Simulation_3Species.Rmd
s

@@
 source("R/init_sim_Bens_nofish.R")
-sim <- init_sim_Bens_nofish(nrows = nrow(hab$hab$spp1), ncols = ncol(hab$hab$spp1), n_years = 5, n_tows_day = 1,n_days_wk_fished = 1, n_fleets = 1, n_vessels = 1, n_species = 3,
+sim <- init_sim_Bens_nofish(nrows = nrow(hab$hab$spp1), ncols = ncol(hab$hab$spp1), n_years = 22, n_tows_day = 1,n_days_wk_fished = 1, n_fleets = 1, n_vessels = 1, n_species = 3,
                             move_freq = 1)

### 2022-04-29 R/BENS_init_survey.R
Updated for use with real strata

@@ -93,25 +93,32 @@ BENS_init_survey <- function (sim_init = NULL, design = 'fixed_station', n_stati
 	  	  for(i in seq(1:length(strata_num[,1]))){
 	    new_strat <- c(new_strat,unique(strata_num[i,]))
 	     }
-	  unique_numbers <- unique(new_strat)
+	  unique_numbers <- unique(new_strat[(new_strat>0)&(!is.na(new_strat))]) #avoid NA and 0 values
+print(unique_numbers)
 	  
 #setup list that 1 entry for each strata
-	  strata_index_list <- vector(mode = "list", length = max(unique_numbers))
+	  strata_index_list <- vector(mode = "list", length = max(unique_numbers,na.rm = T))
 
-
+#replace NAs in strata_num with -999
+	  strata_num[is.na(strata_num)] <- -999
 	  
 	  #prepare indices for each strata 
 	  index <- vector()  #empty vector
+	  
+	  un_idx <- 0  #unique_numbers index
+	  
 	  for(j in unique_numbers){
-	    index[j]<-0
+	   	   
+	     un_idx <- un_idx+1
+	   	 index[j]<-0
+	    
 
-	
 	  #go through entire strata and number each strata 1 to n where n is the total elements in the strata
 	  strata_index <- matrix(0,nrow=idx[["nrows"]],idx[["ncols"]])
 	  
 	  for(i in seq(1:length(strata_num))){
-	    
-	    if(strata_num[i] == unique_numbers[j]){
+
+	    if(strata_num[i] == unique_numbers[un_idx]){
 	      strata_index[i]<-index[strata_num[i]] +1
 	      index[strata_num[i]]<-index[strata_num[i]]+1
 	      
@@ -123,11 +130,11 @@ BENS_init_survey <- function (sim_init = NULL, design = 'fixed_station', n_stati
 	  }	  
 	  #have created matirx with 1 to n in correct strata and 0 elsewhere
 	  #store each in list
-	  strata_index_list[[j]]<-strata_index
+	  strata_index_list[[un_idx]]<-strata_index
 	  }
 	  
 	  
-	  
+	  print(index)
 	  
 	  #go through each strata_index_list, choose correct number of random stations
 	  #translate back to index for given strata
@@ -136,15 +143,20 @@ BENS_init_survey <- function (sim_init = NULL, design = 'fixed_station', n_stati
 	  #while you do keep track of strata number
 	  str_num <-vector()
[…]

### 2022-04-29 TestScripts/run_survey_BENS_NEW.R

@@ -1,6 +1,72 @@
 # after simulation takes place, go through and do a random survey
 
 
+
+
+#Load tow record to figure out how many tows in each stratum. This is Haddock because it has all the stratas
+tows <- read.csv("C:\\Users\\benjamin.levy\\Desktop\\NOAA\\GIS_Stuff\\From_Liz\\survey_tow_latlon-20220107T192759Z-001\\survey_tow_latlon\\ADIOS_SV_164744_GBK_NONE_survey_dist_map_fixed.csv")
+
+#GB strata
+GB_strata <- c("01130","01140","01150","01160","01170","01180","01190","01200","01210","01220","01230","01240","01250", "01290", "01300")
+
+#remove the first 0 from above list to match the data in the tow csv
+GB_strata <- sub('.', '', GB_strata)
+
+#count number of samples in each strata in each year
+num <- matrix(0,nrow=length(seq(2000,2019)),ncol=length(GB_strata))
+idxr <-1
+for(k in seq(2000,2019)){
+idxc<-1
+for(i in GB_strata){
+  
+  num[idxr,idxc] <- sum((tows$YEAR==k)&(tows$STRATUM==GB_strata[idxc]))
+  idxc<-idxc+1
+}
+idxr <- idxr+1
+}
+
+#after looking at above output I have decided to sample the following amounts from each stratum PER SEASON
+#"01130","01140","01150","01160","01170","01180","01190","01200","01210","01220","01230","01240","01250", "01290", "01300"
+#THESE ARE PER SEASON. WILL BE TWICE AS MANY PER YEAR TOTAL
+strata_samples <- c(10,4,3,14,4,4,8,6,4,4,6,7,3,10,3)
+
+
+
+source("R/BENS_init_survey.R")
+
+
+#CURRENTLY NEED TO MAKE SURE THAT N_STATIONS*#YEARS / #STRATA IS A WHOLE NUMBER OTHERWISE DAY, TOW, YEAR WONT LINEUP WITH NUMBER OF STATIONS
+#ALSO NEED N_STATION TO BE DIVISIBLE BY STATIONS_PER_DAY
+#ALSO NEED N_STATIONS / STATIONS_PER_DAY <= 52 otherwise wont get to all of them in a year results in NA in the matrix
+
+nstat <- 2*strata_samples #this is total samples per year per strata 
+surv_random <- BENS_init_survey(sim_init = sim,design = 'random_station', n_stations = nstat, 
[…]

### 2022-04-29 GB_Simulation_3Species.Rmd
small updates

@@
 #profvis::profvis({
-#decreasing population settings
+#original settings
 Pop <- init_pop_Bens(sim_init = sim, Bio = c("spp1" = 3194, "spp2" = (13000+30000)/2, "spp3" = 150000*1.2), #CHANGE
@@
                      M = c("spp1" = .2064+.358, "spp2" = .2728+.511, "spp3" = .334+.45), #NEED F
+                   K = c("spp1" = -log((1+exp(-.2295))^(1/52)-1), "spp2" = -log((1+exp(-.16))^(1/52)-1), "spp3" = -log((1+exp(-.2465))^(1/52)-1)),
+                     #  K = c("spp1" = .2295, "spp2" = .16, "spp3" = .2465),
+                     Weight_PreRecruit = (c("spp1" = .13/.39, "spp2" = .39/2.95, "spp3" = .19/1.12)),
+                     Weight_Adult = (c("spp1" = 1, "spp2" = 1, "spp3" = 1)),
+                     nz = nonzero_idx)
+
+#increasing population
+Pop <- init_pop_Bens(sim_init = sim, Bio = c("spp1" = 3194, "spp2" = (13000+30000)/2, "spp3" = 150000*1.2), #CHANGE
+                     hab = hab[["hab"]], start_cell = list("spp1" = c(46,112),"spp2" = c(30,134),"spp3" = c(24,91)),
+                     lambda = c("spp1" = 0, "spp2" = 0, "spp3" = 0), #set at 0 first, change after initial population set
+                     init_move_steps = 20,
+                     rec_params = list("spp1" = c("model" = "BH", "a" = 40000, "b" = 4301*10, "cv" = 0),
+                                       "spp2" = c("model" = "BH", "a" = 45000 , "b" = 10472*6, "cv" = 0),
+                                       "spp3" = c("model" = "BH", "a" = 100000, "b" = 40530*10, "cv" = 0)), #CHANGE
+                     rec_wk = list("spp1" = 9:12, "spp2" = 8:13, "spp3" = 11:14),
+                     spwn_wk = list("spp1" = 9:12, "spp2" = 8:13, "spp3" = 11:14  ), 
+                     M = c("spp1" = .2064+.358, "spp2" = .2728+.1, "spp3" = .334+.45), #NEED F
                    K = c("spp1" = -log((1+exp(-.2295))^(1/52)-1), "spp2" = -log((1+exp(-.16))^(1/52)-1), "spp3" = -log((1+exp(-.2465))^(1/52)-1)),
@@
 #moveCov <- readRDS(file="20 year moveCov matrices/GeorgesBank/GB_22yr_ConstTemp_HaddockStrata")
+
+#Constant temp
 moveCov <- readRDS(file="20 year moveCov matrices/GeorgesBank/GB_22yr_ConstTemp_HaddockStrata_res2")
+
+#increasing temp
+moveCov <- readRDS(file="20 year moveCov matrices/GeorgesBank/GB_22yr_IncrTemp_HaddockStrata_res2")
 
@@
 
+  #this might track progress. If not check here: https://cran.r-project.org/web/packages/pbapply/pbapply.pdf
+ system(paste("echo 'now processing:",x,"'"))
+
   source("R/run_sim.R")
@@
 #   #rebuild package
-devtools::build()
+#devtools::build()
 
@@
   #load increase temp data from Mars
-  load("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-toy/20 year moveCov matrices/Final/Increase_temp_20yr_LargerTempVA7.RData")
+  load("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-MFS2/Results/GB_3species_IncPop_ConstTemp_environment.RData")
 
   #this starts the profiling
-library(profvis)
-
-profvis({
+# library(profvis)
+# 
+# profvis({
 
@@
 
-})
+#})
+
+ return(res)
 
@@
 
-result <- parallel::parLapply(cl,1:6,run_simulation)
+result <- parallel::parLapply(cl,1:100,run_simulation)
 parallel::stopCluster(cl)

### 2022-04-29 R/run_sim.R

@@ -174,7 +174,7 @@ run_sim <- function(nz = NULL, MoveProb = NULL, MoveProb_spwn = NULL, sim_init =
   ### loop control #
   for (t in seq_len(ntow)) { #for(t in seq(50,ntow,1)){ #THIS DEFINES THE LOOP. NTOW = n_tows_day * n_days_wk_fished * 52 * n_years
     ##################
-    print(t)
+   # print(t)
     ## Loop messages
     
     ## Print when new year

### 2022-05-06 GB_Simulation_3Species.Rmd
Updated survey for GB strata

@@
 ```{r sim}
-run_simulation <- function(x){
+run_simulation <- 
+  function(x){
 
@@
   #load increase temp data from Mars
-  load("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-MFS2/Results/GB_3species_IncPop_ConstTemp_environment.RData")
+  load("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-MFS2/Results/GB_3species_IncPop_IncrTemp_environment.RData")
 
@@
 
-nCoresToUse <- 6
+nCoresToUse <- 5
 
@@
 parallel::stopCluster(cl)
+
+#save results
+save.image("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-MFS2/Results/GB_3species_IncPop_IncrTemp_RESULTS.RData")
 

### 2022-05-06 TestScripts/run_survey_BENS_NEW.R

@@ -1,7 +1,8 @@
 # after simulation takes place, go through and do a random survey
 
 
-
+#to rotate matrix before fields::image.plot
+rotate <- function(x) t(apply(x, 2, rev))
 
 #Load tow record to figure out how many tows in each stratum. This is Haddock because it has all the stratas
 tows <- read.csv("C:\\Users\\benjamin.levy\\Desktop\\NOAA\\GIS_Stuff\\From_Liz\\survey_tow_latlon-20220107T192759Z-001\\survey_tow_latlon\\ADIOS_SV_164744_GBK_NONE_survey_dist_map_fixed.csv")
@@ -133,13 +134,15 @@ surv_random <- BENS_init_survey(sim_init = sim,design = 'random_station', n_stat
 sample_per_sn <- 10   #samples per season per strata
 #10 per season for 2 seasons is 20 samples in each strata over a year, or 80 total per year This would be 1600 over 20 years 
 
-n_spp <- 2     #number of species
+n_spp <- 3     #number of species
 
-nstrata <- 16   #number of strata
+nstrata <- 15   #number of strata
 
 nyears <- 20 #number of simulation years
 
 years_cut <- 2 #number of extra years cut off the front
+
+n_cols <- 7 + n_spp  #number of columns in survey matrix. 7 + number of species
 
 strat_samp_tot <- nstat*nyears #total number of samples to collect in each strata over entire simulation
 #(number of stations in each strata per year) * (number of years)
@@ -149,11 +152,11 @@ strat_samp_tot <- nstat*nyears #total number of samples to collect in each strat
 strata_surv <- list()
 #pull out each strata survey info and store separately  
 
-temp <- matrix(unlist(surv_random$log.mat),ncol=9, nrow=sum(strat_samp_tot))
+temp <- matrix(unlist(surv_random$log.mat),ncol=n_cols, nrow=sum(strat_samp_tot))
 idx <- 1
 for(i in seq(nstrata)){
   
-  strata_surv[[i]] <- temp[idx:(idx+strat_samp_tot[i]-1),1:9]
+  strata_surv[[i]] <- temp[idx:(idx+strat_samp_tot[i]-1),1:n_cols]
   
   idx <- idx + strat_samp_tot[i]
   # print(idx)
[…]

### 2022-05-18 TestScripts/run_survey_BENS_NEW.R
Updates providing more generality to work with arbitrary number of n strata that may not be numbered 1 to n

@@ -400,16 +400,26 @@ surv_noise <- vector("list",length(list_all))
 for(iter in seq(length(list_all))){ #go down each iteration
   temp <- list_all[[iter]]
   
-  spp1_sam <- as.numeric(temp[,8])   #spp1 sample in column 8 
-  spp2_sam <- as.numeric(temp[,9])   #spp1 sample in column 9
+  #pull out individual species samples
+  sam_list <- vector("list",n_spp) 
+  for(s in seq(n_spp)){
+    
+   sam_list[[s]] <- assign(paste0("spp", s, "_sam", sep="") , as.numeric(temp[,7+s]) )
+    
+  }
+  # old method of above with 2 species  
+  # spp1_sam <- as.numeric(temp[,8])   #spp1 sample in column 8 
+  # spp2_sam <- as.numeric(temp[,9])   #spp1 sample in column 9
+  
   for(noise_samp in seq(samp_per_iter)){   #add noise to each survey from each iteration samp_per_iter times
     
-    
-    
     #adding noise to survey
-    temp[,8] <- sapply(spp1_sam,function(x){rlnorm(1,mean=log(x),sdlog=.35)}) #what should sdlog be?
-    temp[,9] <- sapply(spp2_sam,function(x){rlnorm(1,mean=log(x),sdlog=.35)})
+    temp[,8:(7+n_spp)] <-  sapply(seq(n_spp) , function(s) {  sapply(sam_list[[s]] , function(x){rlnorm(1,mean=log(x),sdlog=.35)} ) } )
     
+    # old method of above with 2 species  
+    # temp[,8] <- sapply(spp1_sam,function(x){rlnorm(1,mean=log(x),sdlog=.35)}) #what should sdlog be?
+    # temp[,9] <- sapply(spp2_sam,function(x){rlnorm(1,mean=log(x),sdlog=.35)})
+    # 
     
     surv_noise[[iter]][[noise_samp]]  <- temp
     
@@ -450,9 +460,9 @@ source("TestScripts/Calc_strat_mean/fn_srs_survey_BENS.R")
 
 
 #DEFINE INDIVIDUAL STRATUM AREAS 
-stratum <- seq(nstrata)
+stratum <- sort(unique(surv_random$log.mat[,4]))
 
-STRATUM_AREA <- surv_random$cells_per_strata # old way: rep(10000/nstrata,nstrata) #100x100 grid so each corner has area 2500
[…]

### 2022-05-18 GB_Simulation_3Species.Rmd
Movement testing script, changes to GB simulation and VAST script

@@
 #set lambda at 0 first, change after initial population set
-Pop$dem_params$spp1$lambda <- .1
-Pop$dem_params$spp2$lambda <- .1
-Pop$dem_params$spp3$lambda <- .1
+Pop$dem_params$spp1$lambda <- .7
+Pop$dem_params$spp2$lambda <- .7
+Pop$dem_params$spp3$lambda <- .7
 

### 2022-05-18 TestScripts/movement_check.R

@@ -189,6 +189,8 @@ YT_spring_mnth <-  list() #7 months with survey observations
 cod_spring_mnth <- list() #7 months
 had_spring_mnth <- list() #7 months
 
+Hab_pref <- list(list(),list(),list())
+
 m_idx <- 1
 
 for (m in spring_sample_mnts) {
@@ -211,7 +213,7 @@ for (m in spring_sample_mnts) {
   )
   
   
-  m_idx <- m_idx+1
+
   # 
   # YT_mnth[[idx]] <-
   #   (YT_wk[[w]] + YT_wk[[w + 1]] + YT_wk[[w + 2]] + YT_wk[[w + 3]]) / 4
@@ -220,6 +222,37 @@ for (m in spring_sample_mnts) {
   # had_mnth[[idx]] <-
   #   (had_wk[[w]] + had_wk[[w + 1]] + had_wk[[w + 2]] + had_wk[[w + 3]]) / 4
   # idx <- idx + 1
+  
+  
+  
+  #CREATE MONTHLY AVERAGE OF HABITAT + TEMPERATURE PREFERENCES WITH LAMBDA = 0 
+  #FOR COMPARISON TO TRUE MODEL REALIZATION WITH LAMBDA NONZERO
+  
+  for(s in seq_len(length(hab[["hab"]]))) {
+    
+    move_cov_wk1 <- moveCov[["cov.matrix"]][[wk1]]
+    move_cov_wk2 <- moveCov[["cov.matrix"]][[wk1+1]]   #FALL USES 5 WEEKS AS DEFINED ABOVE
+    move_cov_wk3 <- moveCov[["cov.matrix"]][[wk1+2]]
+    move_cov_wk4 <- moveCov[["cov.matrix"]][[wk1+3]]
+    move_cov_wk5 <- moveCov[["cov.matrix"]][[wk1+4]]
+    
+    move_cov_wk1_spp <- matrix(nc = ncol(move_cov_wk1), nr = nrow(move_cov_wk1), sapply(move_cov_wk1, norm_fun, mu = moveCov[["spp_tol"]][[s]][["mu"]],va = moveCov[["spp_tol"]][[s]][["va"]]))
+    move_cov_wk2_spp <- matrix(nc = ncol(move_cov_wk1), nr = nrow(move_cov_wk1), sapply(move_cov_wk2, norm_fun, mu = moveCov[["spp_tol"]][[s]][["mu"]],va = moveCov[["spp_tol"]][[s]][["va"]]))
[…]

### 2022-05-18 VAST/From Chris/test_simulate.R

@@ -1,7 +1,7 @@
 # test_simulate.R
 # from the Wiki
 
-RootDir = "C:\\Users\\chris.legault\\Documents\\Working\\VAST training 2019\\VASTwd\\simulate example"
+RootDir = "C:\\Users\\benjamin.levy\\Desktop\\Github\\READ-PDB-blevy2-MFS2\\VAST"
 
 library(TMB)
 library(VAST)
@@ -31,7 +31,7 @@ dir.create(DateFile)
 
 # Load data
 data( EBS_pollock_data )
-Data_Geostat = data.frame( "Catch_KG"=EBS_pollock_data[,'catch'], "Year"=EBS_pollock_data[,'year'], "Vessel"="missing", "AreaSwept_km2"=0.01, "Lat"=EBS_pollock_data[,'lat'], "Lon"=EBS_pollock_data[,'long'], "Pass"=0)
+Data_Geostat = data.frame( "Catch_KG"=EBS_pollock_data$sampling_data[,'catch'], "Year"=EBS_pollock_data$covariate_data[,'Year'], "Vessel"="missing", "AreaSwept_km2"=0.01, "Lat"=EBS_pollock_data$sampling_data[,'lat'], "Lon"=EBS_pollock_data$sampling_data[,'long'], "Pass"=0)
 Extrapolation_List = make_extrapolation_info( Region=Region, strata.limits=strata.limits )
 Spatial_List = make_spatial_info( grid_size_km=grid_size_km, n_x=n_x, Method=Method, Lon=Data_Geostat[,'Lon'], Lat=Data_Geostat[,'Lat'], Extrapolation_List=Extrapolation_List, randomseed=Kmeans_Config[["randomseed"]], nstart=Kmeans_Config[["nstart"]], iter.max=Kmeans_Config[["iter.max"]], DirPath=DateFile, Save_Results=TRUE )
 Data_Geostat = cbind( Data_Geostat, "knot_i"=Spatial_List$knot_i )

### 2022-05-20 GB_Simulation_3Species.Rmd
UPdates to survey to be more generalized for real strata

@@
 
+#constant population
+Pop <- init_pop_Bens(sim_init = sim, Bio = c("spp1" = 3194, "spp2" = (13000+30000)/2, "spp3" = 150000*1.2), #CHANGE
+                     hab = hab[["hab"]], start_cell = list("spp1" = c(46,112),"spp2" = c(30,134),"spp3" = c(24,91)),
+                     lambda = c("spp1" = 0, "spp2" = 0, "spp3" = 0), #set at 0 first, change after initial population set
+                     init_move_steps = 20,
+                     rec_params = list("spp1" = c("model" = "BH", "a" = 30445, "b" = 4301, "cv" = 0),
+                                       "spp2" = c("model" = "BH", "a" = 27868 , "b" = 10472, "cv" = 0),
+                                       "spp3" = c("model" = "BH", "a" = 73568, "b" = 40530, "cv" = 0)), #CHANGE
+                     rec_wk = list("spp1" = 9:12, "spp2" = 8:13, "spp3" = 11:14),
+                     spwn_wk = list("spp1" = 9:12, "spp2" = 8:13, "spp3" = 11:14  ), 
+                     M = c("spp1" = .2064+.358+2, "spp2" = .2728+.511+.05, "spp3" = .334-.025), #NEED F
+                   K = c("spp1" = -log((1+exp(-.2295))^(1/52)-1), "spp2" = -log((1+exp(-.16))^(1/52)-1), "spp3" = -log((1+exp(-.2465))^(1/52)-1)),
+                     #  K = c("spp1" = .2295, "spp2" = .16, "spp3" = .2465),
+                     Weight_PreRecruit = (c("spp1" = .13/.39, "spp2" = .39/2.95, "spp3" = .19/1.12)),
+                     Weight_Adult = (c("spp1" = 1, "spp2" = 1, "spp3" = 1)),
+                     nz = nonzero_idx)
+
 #})
@@
   #load increase temp data from Mars
-  load("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-MFS2/Results/GB_3species_IncPop_IncrTemp_environment.RData")
+  load("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-MFS2/Results/GB_3species_ConsPop_ConsTemp_environment.RData")
 
@@
 #save results
-save.image("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-MFS2/Results/GB_3species_IncPop_IncrTemp_RESULTS.RData")
+save.image("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-MFS2/Results/GB_3species_ConsPop_ConsTemp_RESULTS.RData")
 

### 2022-05-20 TestScripts/run_survey_BENS_NEW.R

@@ -650,9 +650,9 @@ for(s in seq(length(sum_survey_iter))){ #2 species
 
 # NEED TO UPDATE BELOW
 # #write csvs
-# write.csv(sum_survey_iter_final[[1]], file="spp1_SRS_16Randomstrata_option1_IncTemp.csv", row.names=F)
-# write.csv(sum_survey_iter_final[[2]], file="spp2_SRS_16Randomstrata_option1_IncTemp.csv", row.names=F)
-# write.csv(sum_survey_iter_final[[1]], file="spp3_SRS_16Randomstrata_option1_IncTemp.csv", row.names=F)
+write.csv(sum_survey_iter_final[[1]], file="YT_SRS_GB_allstrata_IncPop_ConTemp.csv", row.names=F)
+write.csv(sum_survey_iter_final[[2]], file="Cod_SRS_GB_allstrata_IncPop_ConTemp.csv", row.names=F)
+write.csv(sum_survey_iter_final[[3]], file="Had_SRS_GB_allstrata_IncPop_ConTemp.csv", row.names=F)
 
 
 
@@ -859,74 +859,67 @@ log.mat <- rbind(sum_survey_results[[1]],sum_survey_results[[2]],sum_survey_resu
 #1) go through each iteration and put individual results in their own list 
 #2) use Reduce to add all items in each list together, then by total number in list to obtain mean
 #3) use info from https://stackoverflow.com/questions/39351013/standard-deviation-over-a-list-of-matrices-in-r to calculate standard deviation
+                  
+                 #spp1 spp2 spp3
+short_names <- c("YT","Cod","Had") 
 
-pop_sum_s1_biomat <- list()
-pop_sum_s1_recmat <- list()
-pop_sum_s2_biomat <- list()
-pop_sum_s2_recmat <- list()
+
+pop_sum_biomats <- list()
+pop_sum_recmats <- list()
+
+
 
 
 for(i in seq(length(result))){
   
-  #make a list of each category from each iteration
-  pop_sum_s1_biomat[[i]] <- result[[i]][["pop_summary"]][["spp1"]][["Bio.mat"]] 
-  pop_sum_s1_recmat[[i]] <- result[[i]][["pop_summary"]][["spp1"]][["Rec.mat"]] 
-  pop_sum_s2_biomat[[i]] <- result[[i]][["pop_summary"]][["spp2"]][["Bio.mat"]]
[…]

### 2022-05-24 GB_Simulation_3Species.Rmd
Improving generality, especially in plotting at the end

@@
   #load increase temp data from Mars
-  load("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-MFS2/Results/GB_3species_ConsPop_ConsTemp_environment.RData")
+  load("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-MFS2/Results/GB_3species_IncPop_IncTemp_environment.RData")
 
@@
 
-nCoresToUse <- 5
+nCoresToUse <- 4
 
@@
 #save results
-save.image("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-MFS2/Results/GB_3species_ConsPop_ConsTemp_RESULTS.RData")
+save.image("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-MFS2/Results/GB_3species_ConPop_IncTemp_RESULTS.RData")
 

### 2022-05-24 TestScripts/run_survey_BENS_NEW.R

@@ -50,6 +50,22 @@ surv_random <- BENS_init_survey(sim_init = sim,design = 'random_station', n_stat
 
 
 
+#plot individual strata for identification purposes
+for(strata in unique(surv_random$log.mat[,4])){
+  
+  temp <- hab$stratas
+  temp[temp!= strata]= -999
+  fields::image.plot(rotate(temp))
+  text(0.5, 0.98, labels = paste(strata))
+  
+  
+  
+}
+
+
+
+
+
 #make sure samples are in correct strata
 
 fields::image.plot(rotate(hab$stratas))
@@ -445,10 +461,17 @@ for(iter in seq(length(list_all))){ #go down each iteration
 
 #choose some strata to exclude, if desired
 
-#1- exclude northeast 4 corner strata (#3, 4, 6, 7)
-exclude <- c(3,4,7,8)
+#Generic setup- out of 16 strata, exclude northeast 4 corner strata (#3, 4, 6, 7)
+exclude <- list(c(3,4,7,8),c(3,4,7,8)) #2 species
 
-exclude <- c(0)
+
+#George's Bank Setup by species
+                        #YT            Cod          Haddock
+exclude <- list(c(13,14,15,17,18), c(23,24,25), c(23,24,25,29,30))
+
+#exclude none
+exclude <- list(c(0),c(0),c(0)) #3 species
+
 
 #NOW WE NEED TO CREATE A STRATIFIED MEAN FROM EACH OF THESE SAMPLES
 #there are #strata * #iterations * #samp_per_iter total samples
@@ -470,16 +493,6 @@ source("TestScripts/Calc_strat_mean/fn_srs_survey_BENS.R")
 
 
 
-#DEFINE INDIVIDUAL STRATUM AREAS 
-stratum <- sort(unique(surv_random$log.mat[,4]))
-
-STRATUM_AREA <- na.omit(surv_random$cells_per_strata) # old way: rep(10000/nstrata,nstrata) #100x100 grid so each corner has area 2500
-
-sv.area <- as_tibble(data.frame(stratum,STRATUM_AREA))
-
-#remove strata to exclude from stratified mean calculation
-sv.area <- sv.area[!(sv.area$stratum %in% exclude),]#sv.area %>% slice(-exclude)
-
 #setup dimensions for each species- 1 for each strata
 strat_mean_all <- vector("list",length(seq(n_spp)))
 
@@ -500,10 +513,27 @@ for(iter in seq(length(surv_noise))){
[…]

### 2022-05-24 GB_3species.Rmd
New file to aling with server
rec cv =0.55

--- 2022-05-18 GB_Simulation_3Species.Rmd
@@
                      init_move_steps = 20,
-                     rec_params = list("spp1" = c("model" = "BH", "a" = 30445, "b" = 4301, "cv" = 0),
-                                       "spp2" = c("model" = "BH", "a" = 27868 , "b" = 10472, "cv" = 0),
-                                       "spp3" = c("model" = "BH", "a" = 73568, "b" = 40530, "cv" = 0)), #CHANGE
+                     rec_params = list("spp1" = c("model" = "BH", "a" = 30445, "b" = 4301, "cv" = 0.55),
+                                       "spp2" = c("model" = "BH", "a" = 27868 , "b" = 10472, "cv" = 0.55),
+                                       "spp3" = c("model" = "BH", "a" = 73568, "b" = 40530, "cv" = 0.55)), #CHANGE
                      rec_wk = list("spp1" = 9:12, "spp2" = 8:13, "spp3" = 11:14),
@@
                      init_move_steps = 20,
-                     rec_params = list("spp1" = c("model" = "BH", "a" = 40000, "b" = 4301*10, "cv" = 0),
-                                       "spp2" = c("model" = "BH", "a" = 45000 , "b" = 10472*6, "cv" = 0),
-                                       "spp3" = c("model" = "BH", "a" = 100000, "b" = 40530*10, "cv" = 0)), #CHANGE
+                     rec_params = list("spp1" = c("model" = "BH", "a" = 40000, "b" = 4301*10, "cv" = 0.55),
+                                       "spp2" = c("model" = "BH", "a" = 45000 , "b" = 10472*6, "cv" = 0.55),
+                                       "spp3" = c("model" = "BH", "a" = 100000, "b" = 40530*10, "cv" = 0.55)), #CHANGE
                      rec_wk = list("spp1" = 9:12, "spp2" = 8:13, "spp3" = 11:14),
@@
                      M = c("spp1" = .2064+.358, "spp2" = .2728+.1, "spp3" = .334+.45), #NEED F
+                     K = c("spp1" = -log((1+exp(-.2295))^(1/52)-1), "spp2" = -log((1+exp(-.16))^(1/52)-1), "spp3" = -log((1+exp(-.2465))^(1/52)-1)),
+                     #  K = c("spp1" = .2295, "spp2" = .16, "spp3" = .2465),
+                     Weight_PreRecruit = (c("spp1" = .13/.39, "spp2" = .39/2.95, "spp3" = .19/1.12)),
+                     Weight_Adult = (c("spp1" = 1, "spp2" = 1, "spp3" = 1)),
+                     nz = nonzero_idx)
+
+#constant population
+Pop <- init_pop_Bens(sim_init = sim, Bio = c("spp1" = 3194, "spp2" = (13000+30000)/2, "spp3" = 150000*1.2), #CHANGE
+                     hab = hab[["hab"]], start_cell = list("spp1" = c(46,112),"spp2" = c(30,134),"spp3" = c(24,91)),
+                     lambda = c("spp1" = 0, "spp2" = 0, "spp3" = 0), #set at 0 first, change after initial population set
+                     init_move_steps = 20,
+                     rec_params = list("spp1" = c("model" = "BH", "a" = 30445, "b" = 4301, "cv" = 0.55),
+                                       "spp2" = c("model" = "BH", "a" = 27868 , "b" = 10472, "cv" = 0.55),
+                                       "spp3" = c("model" = "BH", "a" = 73568, "b" = 40530, "cv" = 0.55)), #CHANGE
+                     rec_wk = list("spp1" = 9:12, "spp2" = 8:13, "spp3" = 11:14),
+                     spwn_wk = list("spp1" = 9:12, "spp2" = 8:13, "spp3" = 11:14  ), 
+                     M = c("spp1" = .2064+.358+2, "spp2" = .2728+.511+.05, "spp3" = .334-.025), #NEED F
+                     K = c("spp1" = -log((1+exp(-.2295))^(1/52)-1), "spp2" = -log((1+exp(-.16))^(1/52)-1), "spp3" = -log((1+exp(-.2465))^(1/52)-1)),
+                     #  K = c("spp1" = .2295, "spp2" = .16, "spp3" = .2465),
+                     Weight_PreRecruit = (c("spp1" = .13/.39, "spp2" = .39/2.95, "spp3" = .19/1.12)),
+                     Weight_Adult = (c("spp1" = 1, "spp2" = 1, "spp3" = 1)),
+                     nz = nonzero_idx)
+
+#decreasing populations
+Pop <- init_pop_Bens(sim_init = sim, Bio = c("spp1" = 50000, "spp2" = (13000+30000)/2, "spp3" = 150000*1.2), #CHANGE
+                     hab = hab[["hab"]], start_cell = list("spp1" = c(46,112),"spp2" = c(30,134),"spp3" = c(24,91)),
+                     lambda = c("spp1" = 0, "spp2" = 0, "spp3" = 0), #set at 0 first, change after initial population set
+                     init_move_steps = 20,
+                     rec_params = list("spp1" = c("model" = "BH", "a" = 1.071946e+12, "b" = 2.296998e+12, "cv" = 0.55),
+                                       "spp2" = c("model" = "BH", "a" = 389480664 , "b" = 980163673, "cv" = 0.55),
+                                       "spp3" = c("model" = "BH", "a" = 497371062, "b" = 2078612769, "cv" = 0.55)), #CHANGE
+                     rec_wk = list("spp1" = 9:12, "spp2" = 8:13, "spp3" = 11:14),
+                     spwn_wk = list("spp1" = 9:12, "spp2" = 8:13, "spp3" = 11:14  ), 
+                     M = c("spp1" = .2064+.358, "spp2" = .2728+.511, "spp3" = .334+.45), #NEED F
                      K = c("spp1" = -log((1+exp(-.2295))^(1/52)-1), "spp2" = -log((1+exp(-.16))^(1/52)-1), "spp3" = -log((1+exp(-.2465))^(1/52)-1)),
@@
     #load increase temp data from Mars
-  load("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-MFS2/Results/GB_3species_IncPop_IncrTemp_environment.RData")
+    load("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-MFS2/Results/GB_3species_IncPop_IncTemp_environment.RData")
 
@@
     start_time <- Sys.time() # record start time
+    p<-profmem::profmem({
 
@@
                   nz = nonzero_idx)  #does it runin parallel? Doesnt seem like it
-
+    })
     end_time <- Sys.time() # record end time
@@
 #save results
-save.image("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-MFS2/Results/GB_3species_IncPop_IncrTemp_RESULTS.RData")
+save.image("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-MFS2/Results/GB_3species_ConPop_IncTemp_RESULTS.RData")
 

### 2022-05-26 TestScripts/run_survey_BENS_NEW.R
delete to align with server

@@ -1,1446 +1,0 @@
-# after simulation takes place, go through and do a random survey
-
-
-#to rotate matrix before fields::image.plot
-rotate <- function(x) t(apply(x, 2, rev))
-
-#Load tow record to figure out how many tows in each stratum. This is Haddock because it has all the stratas
-tows <- read.csv("C:\\Users\\benjamin.levy\\Desktop\\NOAA\\GIS_Stuff\\From_Liz\\survey_tow_latlon-20220107T192759Z-001\\survey_tow_latlon\\ADIOS_SV_164744_GBK_NONE_survey_dist_map_fixed.csv")
-
-#GB strata
-GB_strata <- c("01130","01140","01150","01160","01170","01180","01190","01200","01210","01220","01230","01240","01250", "01290", "01300")
-
-#remove the first 0 from above list to match the data in the tow csv
-GB_strata <- sub('.', '', GB_strata)
-
-#count number of samples in each strata in each year
-num <- matrix(0,nrow=length(seq(2000,2019)),ncol=length(GB_strata))
-idxr <-1
-for(k in seq(2000,2019)){
-idxc<-1
-for(i in GB_strata){
-  
-  num[idxr,idxc] <- sum((tows$YEAR==k)&(tows$STRATUM==GB_strata[idxc]))
-  idxc<-idxc+1
-}
-idxr <- idxr+1
-}
-
-#after looking at above output I have decided to sample the following amounts from each stratum PER SEASON
-#"01130","01140","01150","01160","01170","01180","01190","01200","01210","01220","01230","01240","01250", "01290", "01300"
-#THESE ARE PER SEASON. WILL BE TWICE AS MANY PER YEAR TOTAL
-strata_samples <- c(10,4,3,14,4,4,8,6,4,4,6,7,3,10,3)
-
-
-
-source("R/BENS_init_survey.R")
-
-
-#CURRENTLY NEED TO MAKE SURE THAT N_STATIONS*#YEARS / #STRATA IS A WHOLE NUMBER OTHERWISE DAY, TOW, YEAR WONT LINEUP WITH NUMBER OF STATIONS
-#ALSO NEED N_STATION TO BE DIVISIBLE BY STATIONS_PER_DAY
-#ALSO NEED N_STATIONS / STATIONS_PER_DAY <= 52 otherwise wont get to all of them in a year results in NA in the matrix
-
-nstat <- 2*strata_samples #this is total samples per year per strata 
[…]

### 2022-05-26 GB_3species.Rmd
Decreasing parameters and aligning survey file with server

@@
                      spwn_wk = list("spp1" = 9:12, "spp2" = 8:13, "spp3" = 11:14  ), 
-                     M = c("spp1" = .2064+.358, "spp2" = .2728+.511, "spp3" = .334+.45), #NEED F
+                     M = c("spp1" = .2064+.358+.2, "spp2" = .2728+.35, "spp3" = .334), #NEED F
                      K = c("spp1" = -log((1+exp(-.2295))^(1/52)-1), "spp2" = -log((1+exp(-.16))^(1/52)-1), "spp3" = -log((1+exp(-.2465))^(1/52)-1)),
@@
     #load increase temp data from Mars
-    load("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-MFS2/Results/GB_3species_IncPop_IncTemp_environment.RData")
+    load("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-MFS2/Results/GB_3species_DecPop_IncTemp_environment.RData")
 
@@
     start_time <- Sys.time() # record start time
-    p<-profmem::profmem({
+   # p<-profmem::profmem({
 
@@
                   nz = nonzero_idx)  #does it runin parallel? Doesnt seem like it
-    })
+   # })
     end_time <- Sys.time() # record end time
@@
 
-nCoresToUse <- 4
+nCoresToUse <- 5
 
@@
 #save results
-save.image("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-MFS2/Results/GB_3species_ConPop_IncTemp_RESULTS.RData")
+save.image("/net/home5/blevy/Bens_R_Projects/READ-PDB-blevy2-MFS2/Results/GB_3species_DecPop_IncTemp_RESULTS.RData")
 

### 2022-05-26 TestScripts/run_survey_BENS_NEW.R

@@ -1,0 +1,1316 @@
+# after simulation takes place, go through and do a random survey
+
+
+#to rotate matrix before fields::image.plot
+rotate <- function(x) t(apply(x, 2, rev))
+
+#Load tow record to figure out how many tows in each stratum. This is Haddock because it has all the stratas
+tows <- read.csv("C:\\Users\\benjamin.levy\\Desktop\\NOAA\\GIS_Stuff\\From_Liz\\survey_tow_latlon-20220107T192759Z-001\\survey_tow_latlon\\ADIOS_SV_164744_GBK_NONE_survey_dist_map_fixed.csv")
+
+#GB strata
+GB_strata <- c("01130","01140","01150","01160","01170","01180","01190","01200","01210","01220","01230","01240","01250", "01290", "01300")
+
+#remove the first 0 from above list to match the data in the tow csv
+GB_strata <- sub('.', '', GB_strata)
+
+#count number of samples in each strata in each year
+num <- matrix(0,nrow=length(seq(2000,2019)),ncol=length(GB_strata))
+idxr <-1
+for(k in seq(2000,2019)){
+  idxc<-1
+  for(i in GB_strata){
+    
+    num[idxr,idxc] <- sum((tows$YEAR==k)&(tows$STRATUM==GB_strata[idxc]))
+    idxc<-idxc+1
+  }
+  idxr <- idxr+1
+}
+
+#after looking at above output I have decided to sample the following amounts from each stratum PER SEASON
+#"01130","01140","01150","01160","01170","01180","01190","01200","01210","01220","01230","01240","01250", "01290", "01300"
+#THESE ARE PER SEASON. WILL BE TWICE AS MANY PER YEAR TOTAL
+strata_samples <- c(10,4,3,14,4,4,8,6,4,4,6,7,3,10,3)
+
+
+
+source("R/BENS_init_survey.R")
+
+
+#CURRENTLY NEED TO MAKE SURE THAT N_STATIONS*#YEARS / #STRATA IS A WHOLE NUMBER OTHERWISE DAY, TOW, YEAR WONT LINEUP WITH NUMBER OF STATIONS
+#ALSO NEED N_STATION TO BE DIVISIBLE BY STATIONS_PER_DAY
+#ALSO NEED N_STATIONS / STATIONS_PER_DAY <= 52 otherwise wont get to all of them in a year results in NA in the matrix
+
+nstat <- 2*strata_samples #this is total samples per year per strata 
[…]

### 2022-12-20 Manuscript/Draft1.Rmd
updating plotting code and manuscript

---
csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
output:
  bookdown::pdf_document2:
    toc: no
    number_sections: no
    fig_caption: yes
    keep_tex: yes
  bibliography: IBMWG.bib
  word_document: default
  html_document: default
  pdf_document:
    toc: no
fontsize: 12pt
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage[belowskip=0pt,aboveskip=0pt]{caption}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if(!require(kableExtra)) {  
  install.packages("kableExtra")
  require(kableExtra)}

#read in results tables
YT_results<-read.csv("Yellowtail_results.csv")
Cod_results<-read.csv("Cod_results.csv")
Had_results<-read.csv("Haddock_results.csv")
```

\newpage

# Estimating Population Trends with Stratified Random Sampling Under the Pressures of Climate Change

Benjamin A. Levy^1^, Christopher M. Legault^2^, Timothy J. Miller^2^, Elizabeth N. Brooks^2^

^1^Ben's Institution, USA  
^2^National Marine Fisheries Service, Northeast Fisheries Science Center, Woods Hole, MA, USA

Corresponding author: Ben Levy (benjamin.levy@noaa.gov)

Competing interests: The authors declare there are no competing interests.

<!-- Table1 <- YT_results -->

<!-- write.table( -->
<!--    format(f, digits=2), "", -->
<!--    sep="|", row.names=F, col.names=F, quote=F, eol="|\n") -->

\newpage

## Abstract

An Abstract

## Keywords

keyword 1, keyword 2

\newpage

```{r}
knitr::kable(YT_results, caption = "Yellowtail error results")

knitr::kable(Cod_results, caption = "Cod error results")

knitr::kable(Had_results, caption = "Haddock error results")
```

\newpage

\section{Introduction}

much of below is from https://apps-nefsc.fisheries.noaa.gov/nefsc/ecosystem-ecology/
or
https://www.fisheries.noaa.gov/data-tools/fisheries-economics-united-states-data-and-visualizations

- The eastern continental shelf is ecologically diverse and economically important

The Northeast United States continental shelf spans from the Outer Banks of North Carolina to the Gulf of Maine. The region covers over 250,000 km$^2$ of ocean, extending over 200 km from shore in the largest areas in New England to just 30 km off shore in the southern regions. This ecologically diverse region contains approximately 18,000 vertebrate marine species. Commercial fisheries have been an important part of local economies for centuries. In 2019, New England fisheries produced $22 billion in sales, which sustained over 200,000 jobs. Maintaining a healthy ecosystem is therefore vital to sustained ecological health and economic prosperity of the region. [@nefmc20]

- Bottom trawl survey is important for monitoring population trends

Fish stocks in this highly productive and economically important region are managed by the National Oceanic and Atmospheric Administration’s (NOAA) Northeast Fisheries Science Center (NEFSC) in Woods Hole, Massachusetts. Federal biologists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (cite survey paper). The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf **(include image from johnson_sosebee?)**. The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits **just listed a few uses of survey change/others?**.

The survey takes place twice each year- once in the spring and again in the fall. Since most spatial analyses and  projections of future distributions typically assume a constant survey catchability and/or availability over time, NOAA's survey design includes sampling during approximately the same 2-3 week time period in each season.

- Climate change is happening
 Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades, resulting in a shift in spatial distributions of many species. Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments. The implication of this is that the bottom trawl survey is actually sampling the population during a different life cycle stage than was originally assumed, which can lead to biased stock assessments. We are therefore interested in analyzing the impact of climate change on the accuracy of future stock assessment models as measured by NOAA’s ongoing bottom-trawl survey along the East coast.

**use more info from initial proposal**

- Fish are changing spatial distribution and have altered life stages (?) because of climate change
NYE paper

- Population indexing methods may be becoming biased as a result

- Briefly describe our study to test this

To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depend on temperature preferences. We can then consider the impact of climate change by simulating scenarios with repeating temperature patterns and those where temperature increases on average over time. In both cases we analyze the ability of stratified random sampling to track population trends.

\section{Methods}

- Describe simulation study

We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on George's Bank, where movement of each species combine static species-specific habitat preferences with dynamic temperature preferences. Model dynamics are driven by dynamic temperature gradients estimated from data to create simulated data sets for each population where the true biomass is known. Using temperature gradients that repeat each year creates data sets with predictable, repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time. We conducting stratified random sampling on our simulation output to mimic the bottom trawl survey and compare the ability of contemporary indexing methods to track population trends.

\subsection{Population Model Formulation}

-- Used MixFishSim. Describe edits made to package

We use the R package *MixFishSim* (MFS) to model our populations [@dolder2020highly]. MFS is a discrete spatiotemporal simulation tool where users can model multiple species under varying environmental conditions. The package uses a delay-difference population model with discrete processes for growth, death, and recruitment of the population. We formulate the following inputs for the MFS package to address our research question:

<font size="5">*Study Area*</font>

We obtained a shapefile for the 15 strata that comprise George's Bank to use as our modeling environment. We discritized the region into a raster with 88 rows and 144 columns. Haddock inhabit all 15 strata in the domain Cod inhabit 13 strata, and yellowtail exist in 9 strata. **include figure?**

<font size="5">*Population Dynamics and Recruitment*</font>

The time step for our models is one week. MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass of in each cell in our study area [@dolder2020highly]. Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits. We chose to represent recruitment in the model using a Beverton-Holt formulation. Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.

--Many parameter inputs were obtained from literature

Each species is assumed to have normally distributed temperature preferences ($~N(\mu,\sigma)$). We assume Yellowtail's preferences are $N(8.75,4.25)$, while Haddock and Cod have preferences $N(9,4)$. We chose these values by combining information in the literature with temperatures recorded in the bottom trawl survey. A full list of parameters used in our model can be seen below in Table ???.

**go through each process in model and discuss edits/removals?**

<font size="5">*Temperature and Habitat Input*</font>

The package was designed to generate theoretical habitat preferences using Gaussian Random Fields that combine with hypothetical temperature gradients to simulate an entire fishery. Since we are modeling real species on the northeast continental shelf w

We obtain estimated temperature data for the region for 2012 from FVCOM **cite**.

--Describe difference between increasing and constant temperature scenarios (images?)

-- Describe each scenario that is considered

We consider three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time. Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees calcius over the duration of the 20 year simulation.

<font size="5">*Simulating Bottom Trawl Survey and Population Indexing*</font>

-Describe post hoc sampling process and how data is used

After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year. The number of the samples taken in each strata reflect true values and we sample from the same weeks that the Spring and Fall surveys take place. Most strata contain enough cells to sample a unique location in each survey. For smaller strata we are forced to repeat locations. We then use the samples in contemporary population indexing methods to track population trends. Since we know the true population values in our simulations we can compare methods by comparing the error calculated from each estimate.

--Stratified mean vs VAST with and without covariates

We compare the stratified mean estimate of abundance to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model. The stratified mean is a typical survey-based approach that scales individual samples to the  strata-level by considering the area of each strata, before scaling to the region-level based on the relative size of each strata. VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurance (presence/absence). If desired, VAST also allows users to include covariate data to better inform the model.

## Discussion

A range of data-limited methods for setting catch advice were evaluated for stocks where assessment models may be rejected due to strong, positive retrospective patterns. A method was considered to perform well if it limited overfishing without resulting in light exploitation rates ($F << F_{MSY}$), thereby allowing depleted stocks to recover to $SSB_{MSY}$ (or for healthy stocks to remain there), and for high and stable catches (close to $MSY$).

Overall, none of the methods evaluated performed best across the scenarios exploring the different sources of the retrospective pattern (unreported catch or increasing $M$) and different levels of historical fishing intensity. A number of methods did perform well in many cases, however, while others performed consistently poorly, resulting in frequent and intense overfishing ($F>>F_{MSY}$). We performed simulations for a couple of scenarios with no source of retrospective patterns and found the expected result that all DLMs and the SCAA performed better ($SSB$, $F$, and catch were all closer to the $MSY$ reference points) than when either source of retrospective patterns was present. Due to the focus of this study, we did not examine the no retrospective source in detail and do not comment on it further.

Currently, in the Northeast U.S., if an assessment model is rejected due to a large rho value in $SSB$, the catch advice from that model is ignored and some data-limited approach is used. However, the rho-adjusted SCAA model performed better than a number of the alternatives explored here. Therefore, there should not necessarily be an expectation that a data-limited method will perform better than the rejected assessment model. The SCAA only resulted in high exploitation rates ($F >> F_{MSY}$) when unreported catch was the source of the retrospective pattern and for the scenario where $F=F_{MSY}$ at the end of the base period that left the stock in relatively good condition ($SSB \sim SSB_{MSY}$). In contrast, this method was particularly effective when the stock was depleted and there was unreported catch. When $M$ was the source of the retrospective pattern, the rho-adjusted SCAA method typically resulted in light exploitation rates, on average. The light exploitation rates in these cases were likely driven by the combination of using a rho-adjustment, but also using the lower $M$ from the beginning of the base period rather than the higher $M$ that occurred during the feedback period. Using an $M$ value that is too low in a stock assessment will typically bias estimates of biomass and reference points too low, resulting in catch advice that is below target levels [@Johnsonetal2014; @Puntetal2021M]. The consequences of using a value for $M$ that is too low versus too high is also asymmetrical [@Johnsonetal2014], with negative consequences being more severe when $M$ is assumed too high than low, and the results here are consistent with these previous conclusions.

The methods that adjusted recent average catches based on trends in the survey (Ismooth and Islope) performed well overall in terms of catch, stock status, and variation in catch. The method using the expanded survey biomass with the recent exploitation rate (ES-Frecent) also performed well and similarly to Ismooth. The performance of these methods was also generally robust among scenarios, with the exception of when there were unreported catches and the stock was depleted (see below). The generally positive performance of these methods was consistent with @Hilbornetal2002 and @CoxKronlund2008, both of which evaluated a variant of a "hold-steady" DLM. In the case of @Hilbornetal2002, the "hold-steady" DLM policy was designed to adjust catches in order to keep rockfish (*Sebastes spp.*) populations at recently observed index levels, and did so by functioning as a constant escapement harvest control rule where target catches were set to zero below some pre-specified index level. In the variant used by @CoxKronlund2008, catches were adjusted to maintain a sablefish (*Anoplopoma fimbria*) population at a pre-specified index level thought to be sustainable and desirable in terms of meeting fishery objectives (e.g., high catch), but never permitted target catches of zero and so functioned as a constant exploitation rate control rule. The "hold-steady" DLM of @CoxKronlund2008 performed similarly in terms of catch, stock depletion, and variation in catch, as a constant exploitation rate policy where target catch was specified as the product of desired exploitation rate and an estimate of biomass from a SCAA model. This result was robust to uncertainty in initial stock status and steepness [@CoxKronlund2008]. The SCAA model was always correctly specified (i.e., expected to produce unbiased estimates on average), however, and no comparison to the results of this research in the presence of retrospective patterns is possible [@CoxKronlund2008]. The "hold-steady" policy of @Hilbornetal2002 performed similarly to or better in terms of catch and stock status than other harvest control rules that relied on assessment estimates of biomass (i.e., 40:10 and constant *F*). The performance of the "hold-steady" DLM was also more robust to uncertainty in steepness and to the presence of unreported catch [@Hilbornetal2002]. The performance of the two harvest policies that relied on assessment estimates of biomass (i.e., constant exploitation rate and a "40:10" biomass-based policy) also degraded when the estimates of biomass were biased, which is an issue that does not effect the "hold-steady" DLM [@Hilbornetal2002]. The bias in the assessment estimates considered in @Hilbornetal2002 were not necessarily induced by a retrospective pattern, however, and no consideration of making a rho-adjustment was possible in that study.

The Ismooth method is currently used to set catches for Georges Bank cod [@nefsc19] and red hake (*Urophycis chuss*; @nefsc20). Variations of the ES-Frecent have been used for witch flounder and Georges Bank yellowtail flounder. While the findings here generally support the continued use of the Ismooth and ES-Frecent methods, they may not be well suited for depleted stocks where unreported catches are believed to be an issue. The Ismooth, Islope, and ES-Frecent DLMs produced high $Fs$ and limited stock recovery with unreported catches and when the stock was depleted. While @Hilbornetal2002 and @CoxKronlund2008 did not reach the same conclusion about the "hold-steady" DLM, those studies did not consider initial levels of depletion as low as in this study. These results highlight the importance of accurate catch reporting, as unreported catch can create a negative feedback loop with perpetually high $Fs$ being produced by a management system that seemingly should result in sustainable catch advice.

Three methods were consistently risk-averse across scenarios, limiting the frequency and magnitude of overfishing and resulting in high stock biomass. These methods were the two catch curve options (CC-FM and CC-FSPR) and DynLin. The catch curve methods produced a wider range of average catches across scenarios, and also had greater interannual variability in catches compared to DynLin. While the lower exploitation rates from these approaches may be undesirable due to forgone yield, there may be circumstances where they are preferred. For example, for stocks that are believed to be heavily depleted, low exploitation rates would allow for a more rapid recovery.

A number of methods performed poorly, particularly when catches were unreported. These methods include three of the expanded survey biomass approaches (ES-Fstable, ES-FM, ES-FSPR), AIM, and Skate. The AIM model has been widely used across stocks in the region [@nefsc02a; @nefsc05; @nefsc08], although there is a decreasing trend in its use across model resistant stocks [@nefsc19]. The findings here suggest that alternative approaches should be considered in cases where AIM is still used and there is concern over unreported catches. The Skate method is used to manage the skate complex in the Northeast U.S. (a group of seven co-managed species). Interestingly, six of the seven species are considered in good condition with high survey biomass indices in recent years [@nefmc20]. That the Skate method performed poorly in our analysis but performs well for the skate complex illustrates how the performance of methods in this analysis may be sensitive to the scenarios and species life history considered. As may be the case for the Skate method, the performance of some methods may depend on the condition of the stock when the method is first applied, and less so on life-history. Therefore, care is needed when trying to generalize these results across stocks that may have different life histories, exploitation histories, and without unreported catches or increases in $M$.

In addition to the analytical differences among the thirteen DLMs, most of the DLMs and control rules had multiple options that could be adjusted to make them more or less risk averse. DynLin had a large number of user defined decision points. Given the large range of options already explored in the study, one suite of options was selected for each DLM-control rule and kept constant for all simulations. Further studies could explore the different options within an individual DLM to understand how they might affect performance.

Many other data-limited methods exist for setting catch advice that were not included in this evaluation, and they vary widely in complexity, data inputs, and assumptions required [e.g., @carruthers2018dlm]. Length based methods were not evaluated to keep the overall number of methods tractable, and due to the availability of age based information in the region. Methods that require only catch data or snap shots of survey data were not considered due to the availability of the relatively long and contiguous Northeast Fisheries Science Center's spring and fall, coastwide bottom trawl surveys, and the fact that "catch only" methods have been shown to perform poorly [e.g., @carruthers2014eval]. Complete catch histories are not available for stocks in the region (i.e., from the inception of fishing). Consequently, methods that required complete catch histories or required assumptions about relative depletion [e.g., DCAC in @maccall2009dca; DB-SRA in @dick2022dsra] were also omitted from consideration. The need for short run-times and the desire for methods that could be reviewed quickly prevented the use of modern state-space production models such as SPiCT [@pedersen2017spict] and JABBA [@winker2018jabba].

The SCAA was confronted with inconsistent data in this study, while the DLMs typically used only a single source of data and thus did not encounter inconsistencies. A recent examination of the data used in assessments in this region similarly found inconsistencies in data streams even before modeling. @wiedenmann2022strange found a negative relationship between relative F (catch/survey) and survey Z for stocks with strong retrospective patterns but the expected positive relationship for stocks without a retrospective pattern. It is exactly this sort of tension that creates retrospective patterns in integrated models, but is not found in DLMs that only use one type of data.

Despite conducting hundreds of thousands of simulations, there are still limitations to our study. We only examined one life history representative of groundfish in the region. We acknowledge that best practice is to select a DLM for a specific life history and fishery condition [e.g., @fischer2020dlm]. As is typically the case with large simulation studies, we were not able to tune any of the DLMs or the SCAA in any given realization, which would occur in practice for an actual stock assessment. We also examined only scenarios that started with Mohn's rho values near 0.5 for spawning stock biomass. This is a strong retrospective pattern, but some stocks in the region have even stronger retrospectives. Performance of the DLMs and SCAA would be expected to degrade with stronger retrospectives, but by how much is still an open area for research. Similarly, sources of retrospective patterns that create different relationships between the true values and estimated values should also be explored [see @deroba2014retro]. To make the results interpretable, we only examined a single source for the retrospective pattern at a time. In reality, there may be more than one factor leading to an observed retrospective pattern. How the multiple sources would interact to influence performance is another topic for future research. Development of harvest control rules specifically for situations where retrospective patterns are found in age-based assessments would also be beneficial. The large number of scenarios examined and the large number of realizations gives us confidence that our results are meaningful in general, but that the performance of any of the DLMs may differ in actual practice.

An interesting finding of this study is the linear versus diffuse patterns between $SSB$ and catch across methods. These patterns have implications for the trade-offs among methods, with linear relationships resulting in more consistent exploitation rates across stock sizes. Therefore, these methods have higher certainty of a given catch at a given stock size. However, they also tended to result in lower stock sizes, on average, across methods. The more diffuse relationships resulted in more variable exploitation rates across stock sizes, with some situations where the population biomass was quite high but the catch was low (relative to MSY), resulting in a very low $F$. The reasons behind these different patterns remain unclear, and future work to explore these patterns is warranted.

One of the reasons for the difference in performance between the catch and natural mortality retrospective sources was how the reference points were calculated. In all cases, the initial conditions, including the natural mortality rate, were used to compute the reference points. This decision was made based on the fact that the increase in natural mortality was assumed to be unknown in the simulations. If the increase in natural mortality was known, the age-structured assessments would have accounted for it, different reference points might have been computed [@legault2016increaseM] and there may not have been a retrospective pattern at all [@legault2020rose], and no need to consider alternative DLMs. The reference points for the increased $M$ scenarios would have been different if they were computed using the values from the final year of the base period, but the overall conclusions regarding the different DLMs would not change as this just results in a rescaling of the axis. These results are not shown to reduce confusion regarding the simulations.

Closed-loop simulation is a common tool for examining performance of catch advice from various stock assessment approaches in a feedback setting. It is often used as part of a full management strategy evaluation when working with stakeholders to develop management regulations that make trade offs between near term and long term catches, risk to the fish population, and mixed-fleet allocations [@carruthers2016simpleMPs; @goethel2019mse; @harlyan2019hcr]. We did not conduct a full management strategy evaluation with stakeholder input [@goethel2019stakeholder], but see that as a fruitful next step that could build on the conclusions from our closed-loop work. Using a generic groundfish life-history and monitoring standard performance metrics related to stock status and catch stability, we were able to cull the herd of potential DLMs and we would not carry the consistent poor performers forward for further study.  The wide range of expertise reflected in the authorship was by design so that the simulation specifications and performance metrics were broadly useful. Before undertaking a full management strategy evaluation and engaging regional stakeholders, we would want to select a specific stock and jointly identify specific management regulations to be tested [@deroba2019dream]. Results of this work have been presented to both local fishery management councils, with generally positive feedback about the utility of the conclusions for identifying appropriate model approaches when an SCAA is rejected. Our work was similar to all other closed-loop simulations in that it was designed to address a specific situation, including much recent work comparing the performance of data-limited and data rich assessment approaches [e.g., @fulton2016datarich; @sagarese2019dlm; @bouch2020datapoor; @li2022dlm].

This study is a first attempt to identify suitable methods for setting catch advice when stock assessment models are rejected due to large, positive retrospective patterns. Although no single method performed best across scenarios, a number of generally suitable and unsuitable methods were identified under specific conditions. The results of this work can help scientists and managers select a subset of possible options for consideration to set catch advice when assessment models are rejected. The approach developed here can, and should be expanded to consider other cases not explored here, as performance of individual methods are very likely case-dependent.

## Acknowledgements
We thank the Index-Based Methods and Control Rules Research Track review panel of Paul Rago (chair), Yong Chen, Robin Cook, and Paul Medley for feedback on preliminary results, three anonymous reviewers and the associate editor for reviewing an earlier version of this work. The scientific results and conclusions, as well as any views or opinions expressed herein, are those of the authors and do not necessarily reflect those of NOAA or the Department of Commerce.

## Data and Code Availability
All data and code used in this work are available at https://github.com/cmlegault/IBMWG.

## References

<div id="refs"></div>

\pagebreak

## Tables
</font>

\pagebreak
Table 1. Parameters used in all population models.
```{r paramsALL, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
|		|	Description | Unit |Yellowtail | Cod | Haddock | Source
|:---	|:------- |:-- |:--- |:-- |:-- |:--
| $\\rho$ | Ford's growth coefficient | wk$^{-1}$  | 4.48 | 4.43 | 4.49 | |
| $M$ | Natural Mortality| wk$^{-1}$  | 0.2064 | 0.2728 | 0.3340 | |
| $W_R$ |Weight of fully recruited fish| kg | 0.39 | 2.95 | 1.12 | |
| $W_{R-1}$ | Weight of pre-recruit fish| kg | 0.13 | 0.39 | 0.19 | |
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

Table 2. Parameters used in models with relatively constant populations.
```{r paramsCON, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
|		|	Description | Yellowtail | Cod | Haddock 
|:---	|:------- |:--- |:-- |:-- |:--
| $M+F$ | Adjusted Mortality (Natural + Fishing)| wk$^{-1}$  | 0.764 | 0.83 | 0.309 |

"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

Table 3. Parameters used in models with increasing populations.
```{r paramsINC, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
|		|	Description | Yellowtail | Cod | Haddock | Source
|:---	|:------- |:--- |:-- |:-- |:--
| $M+F$ | Adjusted Mortality (Natural + Fishing)| wk$^{-1}$  | 0.564 | 0.372 |  0.134 |

"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

Table 4. Parameters used in models with decreasing populations.
```{r paramsDEC, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
|		|	Description | Yellowtail | Cod | Haddock | Source
|:---	|:------- |:--- |:-- |:-- |:--
| $M+F$ | Adjusted Mortality (Natural + Fishing)| wk$^{-1}$  | 0.764 | 0.623 | 0.334 |

"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

Table 1. Maturity-, weight-, and selectivity-at-age of the simulated fish population.
<!---markdown style table. f*cking escape characters-->
```{r lifehisttabl, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
|	Age	|	Maturity | Weight (kg) | Fishery Selectivity (before change if  applicable) | Fishery Selectivity (after change if applicable)
|:---	|:------ |:------ |:------ |:------
|	1	|	0.04 | 0.15 | 0.07 | 0.02
|	2	|	0.25 | 0.5 | 0.17 | 0.05
|	3	|	0.60 | 0.9 | 0.36 | 0.12
| 4 | 0.77 | 1.4 | 0.61 | 0.27
| 5 | 0.85 | 2.0 | 0.81 | 0.50
| 6 | 0.92 | 2.6 | 0.92 | 0.74
| 7 | 1.00 | 3.2 | 0.97 | 0.89
| 8 | 1.00 | 4.1 | 0.99 | 0.96
| 9 | 1.00 | 5.9 | 1.00 | 0.99
| 10+ | 1.00 | 9.0 | 1.00 | 1.00
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

\pagebreak
Table 2. Naming convention and details of the data-limited methods evaluated.
```{r methodtable, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
|	Method	|	Details
|:---	|:------ 
|	Ismooth	|	$C_{targ,y+1:y+2} =\\overline{C}_{3,y}(e^{\\lambda})$ <br> where $\\overline{C}_{3,y}$ is the most recent three year average; <br> $\\overline{C}_{3,y} = \\frac{1}{3}\\sum_{t=1}^{t=3}C_{y-t}$ <br> and $\\lambda$ is the slope of a log linear regression of a LOESS-smoothed average index of abundance (spring and fall) with span = 0.3: <br> $\\hat{I}_y=loess(\\hat{I}_y)$ and $LN(\\widehat{I_y}) = b + \\lambda y$
| Islope	|	$C_{targ,y+1:y+2} = 0.8\\overline{C}_{5,y}(1+0.4e^{\\lambda})$ where $\\overline{C}_{5,y}$ is the most recent five-year average catch through year $y-1$: <br> $\\overline{C}_{5,y} = \\frac{1}{5}\\sum_{t=1}^{t=5}C_{y-t}$ <br> and $\\lambda$ is the slope of a log-linear regression of the most recent five years of the averaged index.
|	Itarget	|	$C_{targ,y+1:y+2}=\\left[0.5C_{ref}\\left(\\frac{\\overline{I}_{5,y}-I_{thresh}}{I_{target}-I_{thresh}}\\right)\\right]$ $\\overline{I}_{5,y} \\geq I_{thresh}$ <br> $C_{targ,y+1:y+2}=\\left[0.5C_{ref}\\left(\\frac{\\overline{I}_{5,y}}{I_{thresh}}\\right)^{2}\\right]$ $\\overline{I}_{5,y} <  I_{thresh}$; <br>  $C_{ref}$ is the average catch over the reference period (years 26 through 50): <br> $C_{ref} = \\frac{1}{25}\\sum_{y=26}^{y=50}C_y$; <br>  $I_{target}$ is 1.5 times the average index over the reference period: <br> $I_{target} = \\frac{1}{25}\\sum_{y=26}^{y=50}\\overline{I}_y$; <br> $I_{thresh}$ = 0.8 $I_{target}$, and is the most recent five year average of the combined spring and fall index: <br> $\\overline{I}_{5,y} = \\frac{1}{5}\\sum_{t=1}^{t=5}\\overline{I}_{y-t+1}$
| Skate | $C_{targ,y+1:y+2}= F_{rel} \\overline{I}_{3,y}$ where <br> $F_{rel} =median \\left(\\frac{\\overline{C}_{3,\\textbf{Y}}}{\\overline{I}_{3,\\textbf{Y}}} \\right)$ <br> is the median relative fishing mortality rate calculated using a 3 year moving average of the catch and average survey index across all available years ($\\textbf{Y}$): <br> $\\overline{C}_{3,y} = \\frac{1}{3} \\sum_{t=1}^{t=3} C_{y-t}$ and <br> $\\overline{I}_{3,y} = \\frac{1}{3} \\sum_{t=1}^{t=3} I_{y-t+1}$
| An Index Method (AIM) | AIM first calculates the annual relative $F$: $F_{rel,y}=\\frac{C_y}{\\frac{1}{3}\\sum_{t=1}^{t=3}\\overline{I}_{y-t+1}}$ <br> and the annual replacement ratio: $\\Psi_y = \\frac{\\overline{I}_y}{\\frac{1}{5}\\sum_{t=1}^{t=5}\\overline{I}_{y-t}}$. <br> These values are used in a regression: $LN(\\Psi_y)= b + \\lambda LN(F_{rel,y})$ to determine $F_{rel,*}$, which is the value of $F_{rel,y}$ where the predicted $\\Psi=1$ or <br> $LN(\\Psi)=0$. $F_{rel,*}$ is called either the “stable” or “replacement” $F$, and is used to calculate the target catch: $C_{targ,y+1:y+2} = \\overline{I}_yF_{rel,*}$.
| Dynamic Linear Model (DynLin) | @Langan2021DLM.
| Expanded survey  biomass method 1 $F_{40\\%}$ (ES-FSPR) | $C_{targ,y+1:y+2}=B_{\\bar{I},y}\\mu_{targ}$ <br> where $B_{\\bar{I}}$ is the average of estimated fully-selected biomass from each survey: <br> $B_{\\bar{I},y}=\\frac{1}{2}\\left(\\frac{I_{spr,y}}{q_{spr}} +\\frac{I_{fall,y-1}}{q_{fall}}\\right)$ <br> and target exploitation fraction, $\\mu_{targ}$ is calculated as: <br> $\\mu_{targ}=\\frac{F_{targ}}{Z_{targ}}\\left(1-e^{-Z_{targ}}\\right)$; <br> $F_{targ}=F_{40\\%}$ and <br> $Z_{targ}=F_{targ}+M$
| Expanded survey biomass method 2 $F=$ AIM replacement (ES-Fstable) | Same as the above expanded survey method, but with $\\mu_{targ}$ equal to the stable exploitation fraction $F_{rel,*}$ calculated using the AIM approach (see above).
| Expanded survey biomass method 3 $F=M$ (ES-FM) | Same as the above expanded survey methods, but with the target exploitation rate set to the assumed $M$: $F_{targ}=M$.
| Expanded survey biomass method 4 $F=$ recent average (ES-Frecent) | Same as the above expanded survey methods, but with the target exploitation fraction set to the most recent three year average exploitation fraction: <br> ${\\mu_{targ}}=\\frac{\\sum_{y-2}^{y}\\mu_y}{3}$ <br> $\\mu_{y}=\\frac{C_{y-1}}{B_{\\bar{I},y}}$
| Catch curve Method 1 $F_{40\\%}$ (CC-FSPR) | $C_{targ,y+1:y+2}= \\frac{F_{targ}}{Z_{avg,y}}B_{cc,y}\\left(1-e^{-Z_{avg,y}}\\right)$ <br>  where $B_{cc}$ is the estimated biomass: <br> $B_{cc,y} = \\frac{C_{y-1}}{\\frac{F_{avg,y}}{Z_{avg,y}}{\\left(1-e^{-Z_{avg,y}}\\right)}}$ with <br> $Z_{avg,y} = \\frac{Z_{spring, y}+Z_{fall,y-1}}{2}$; <br> $F_{avg,y-1} = Z_{avg,y-1} - M$ and, <br> $F_{targ}=F_{40\\%}$. Survey catch at age used in catch curve to estimate $Z$.
| Catch curve Method 2 $M$ (CC-FM) | Same as catch curve method 1 above, but with $F_{targ}=M$.
| Ensemble | Median of catch advice provided by AIM, CC-FSPR, ES-Frecent, ES-FSPR, Islope, Itarget, Ismooth, and Skate methods.
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

\pagebreak
Table 3. Summary of the scenarios evaluated within the study design.
```{r schemetable, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "

|	Factors | Variants
|:------ |:------
|	retrospective source | catch or <br> natural mortality
|	fishing history | $F_{MSY}$ in second half of base period or <br> overfishing throughout base period (2.5x$F_{MSY}$)
|	fishery selectivity blocks | constant selectivity or <br> selectivity changes in second half of base period
| catch advice multiplier | applied as is from DLM (1) or <br> reduced from DLM (0.75)
"
#Table: (\#tab:schemtable) Coding scheme.
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

\pagebreak

## List of Figures

```{r figcaps, echo=FALSE}
figcap1 <- "Inner quartiles and medians for all performance measures across all scenarios and runs for each method. Vertical lines are shown at a value of 1 for the performance measures that are relative to the MSY reference points (A,B,C)."
figcap2 <- "Relationship between long-term average spawning biomass and average catch (relative to MSY levels) for each method. Each point represents the median for a given scenario, separated by the source of the retrospective pattern (catch or M)."
figcap3 <- "Median performance measures for each method, separated by the source of the retrospective error (catch = black, M = gray) and the exploitation history in the base period (always overfishing at $2.5 x F_{MSY}$ (circle), or $F$ reduced to $F_{MSY}$ during base period (triangle)). Vertical lines are shown at a value of 1 for the performance measures that are relative to the MSY reference points (A,B,C)."
figcap4 <- "Median $F / F_{MSY}$ for each method, with results separated by the exploitation history in the base period (always overfishing at $2.5 x F_{MSY}$ (circle), or $F$ reduced to $F_{MSY}$ during base period (triangle)) showing A) short- (gray) versus long-term (black) values, and B) with (black) or without (gray) a buffer applied when setting the catch (catch multiplier = 0.75 or 1)."
figcap5 <- "Relationship between long-term average catch and spawning stock biomass relative to their reference points by method. Each point represents the average for years 21-40 in the feedback period for a single iteration of a scenario. The scenario shown is where catch was the source of the retrospective pattern with $F$ reduced to $F_{MSY}$ in the second half of the base period, there was a single selectivity block, and where no buffer was applied to the catch advice (catch multiplier = 1)."

```

Figure 1. `r figcap1`

Figure 2. `r figcap2`

Figure 3. `r figcap3`

Figure 4. `r figcap4`

Figure 5. `r figcap5`

```{r PMsallscenarios, echo=FALSE, message=FALSE, warnings=FALSE, out.width='7in', fig.cap=figcap1}
#knitr::include_graphics(file.path(getwd(),"MS_tables_figs_rev1/final_ms_figs","Figure_1.pdf"))
```

```{r SSmsyvsCMSY, echo=FALSE, message=FALSE, warnings=FALSE, out.width='7in', fig.cap=figcap2}
#knitr::include_graphics(file.path(getwd(),"MS_tables_figs_rev1/final_ms_figs","Figure_2.pdf"))
```

```{r PMsbyretroF, echo=FALSE, message=FALSE, warnings=FALSE, out.width='6in', fig.cap=figcap3}
#knitr::include_graphics(file.path(getwd(),"MS_tables_figs_rev1/final_ms_figs","Figure_3.pdf"))
```

```{r FFmsybyscenario, echo=FALSE, message=FALSE, warnings=FALSE, out.width='6in', fig.cap=figcap4}
#knitr::include_graphics(file.path(getwd(),"MS_tables_figs_rev1/final_ms_figs","Figure_4.pdf"))
```

```{r linearvsdiffuse, echo=FALSE, message=FALSE, warnings=FALSE, out.width='7in', fig.cap=figcap5}
#knitr::include_graphics(file.path(getwd(),"MS_tables_figs_rev1/final_ms_figs","Figure_5.pdf"))
```

### 2022-12-20 Manuscript/Draft1.tex
updating plotting code and manuscript

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\author{}
\date{\vspace{-2.5em}}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace}
\doublespacing
\usepackage{lineno}
\linenumbers
\usepackage[belowskip=0pt,aboveskip=0pt]{caption}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\begin{document}

\newpage

\hypertarget{estimating-population-trends-with-stratified-random-sampling-under-the-pressures-of-climate-change}{%
\section{Estimating Population Trends with Stratified Random Sampling Under the Pressures of Climate Change}\label{estimating-population-trends-with-stratified-random-sampling-under-the-pressures-of-climate-change}}

Benjamin A. Levy\textsuperscript{1}, Christopher M. Legault\textsuperscript{2}, Timothy J. Miller\textsuperscript{2}, Elizabeth N. Brooks\textsuperscript{2}

\textsuperscript{1}Ben's Institution, USA\\
\textsuperscript{2}National Marine Fisheries Service, Northeast Fisheries Science Center, Woods Hole, MA, USA

Corresponding author: Ben Levy (\href{mailto:benjamin.levy@noaa.gov}{\nolinkurl{benjamin.levy@noaa.gov}})

Competing interests: The authors declare there are no competing interests.

\newpage

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

An Abstract

\hypertarget{keywords}{%
\subsection{Keywords}\label{keywords}}

keyword 1, keyword 2

\newpage

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{knitr}\SpecialCharTok{::}\FunctionTok{kable}\NormalTok{(YT\_results, }\AttributeTok{caption =} \StringTok{"Yellowtail error results"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{table}

\caption{\label{tab:unnamed-chunk-1}Yellowtail error results}
\centering
\begin{tabular}[t]{l|l|l|l|l|l|l|l|l|l|l|l}
\hline
X & X.1 & X.2 & X.3 & X.4 & Constant.Population & X.5 & Increasing.Population & X.6 & Decreasing.Population & X.7 & X.8\\
\hline
Temp & Covariate & Strata & Noise & season & Stratified Mean & VAST Estimate & Stratified Mean & VAST Estimate & Stratified Mean & VAST Estimate & \\
\hline
const & no cov & all & no & spring & 0.21 & 0.11 & 0.16 & 0.13 & 0.23 & 0.08 & \\
\hline
const & no cov & all & yes & spring & 0.25 & 0.16 & 0.22 & 0.21 & 0.27 & 0.11 & \\
\hline
const & w/ cov & all & no & spring & 0.21 & 0.07 & 0.16 & 0.06 & 0.23 & 0.06 & \\
\hline
const & w/ cov & all & yes & spring & 0.25 & 0.08 & 0.22 & 0.07 & 0.27 & 0.07 & \\
\hline
const & no cov & all & no & fall & 0.32 & 0.68 & 0.34 & 0.36 & 0.41 & 0.81 & \\
\hline
const & no cov & all & yes & fall & 0.31 & 0.77 & 0.46 & 0.44 & 0.44 & 1.09 & \\
\hline
const & w/ cov & all & no & fall & 0.32 & 0.08 & 0.34 & 0.08 & 0.41 & 0.08 & \\
\hline
const & w/ cov & all & yes & fall & 0.31 & 0.11 & 0.46 & 0.17 & 0.44 & 0.18 & \\
\hline
const & no cov & reduced & no & spring & 0.27 & 0.19 & 0.2 & 0.15 & 0.25 & 0.19 & \\
\hline
const & no cov & reduced & yes & spring & 0.26 & 0.15 & 0.22 & 0.11 & 0.3 & 0.16 & \\
\hline
const & w/ cov & reduced & no & spring & 0.27 & 0.19 & 0.2 & 0.17 & 0.25 & 0.19 & \\
\hline
const & w/ cov & reduced & yes & spring & 0.26 & 0.14 & 0.22 & 0.13 & 0.3 & 0.16 & \\
\hline
const & no cov & reduced & no & fall & 0.47 & 0.25 & 0.41 & 0.11 & 0.55 & 0.24 & \\
\hline
const & no cov & reduced & yes & fall & 0.49 & 0.36 & 0.46 & 0.21 & 0.53 & 0.36 & adding cov made worse\\
\hline
const & w/ cov & reduced & no & fall & 0.47 & 0.19 & 0.41 & 0.22 & 0.55 & 0.23 & \\
\hline
const & w/ cov & reduced & yes & fall & 0.49 & 0.17 & 0.46 & 0.19 & 0.53 & 0.24 & \\
\hline
increasing & no cov & all & no & spring & 0.28 & 0.11 & 0.32 & 0.13 & 0.22 & 0.15 & \\
\hline
increasing & no cov & all & yes & spring & 0.28 & 0.15 & 0.34 & 0.16 & 0.26 & 0.17 & cov really helped\\
\hline
increasing & w/ cov & all & no & spring & 0.28 & 0.06 & 0.32 & 0.07 & 0.22 & 0.07 & \\
\hline
increasing & w/ cov & all & yes & spring & 0.28 & 0.12 & 0.34 & 0.1 & 0.26 & 0.1 & \\
\hline
increasing & no cov & all & no & fall & 0.51 & 1.26 & 0.3 & 0.66 & 0.28 & 1.06 & \\
\hline
increasing & no cov & all & yes & fall & 0.5 & 1.38 & 0.39 & 0.71 & 0.25 & 1.1 & \\
\hline
increasing & w/ cov & all & no & fall & 0.51 & 0.23 & 0.3 & 0.21 & 0.28 & 0.15 & cov good, why no cov so bad???\\
\hline
increasing & w/ cov & all & yes & fall & 0.5 & 0.28 & 0.3 & 0.37 & 0.25 & 0.2 & \\
\hline
increasing & no cov & reduced & no & spring & 0.31 & 0.17 & 0.4 & 0.33 & 0.32 & 0.14 & \\
\hline
increasing & no cov & reduced & yes & spring & 0.29 & 0.19 & 0.41 & 0.26 & 0.29 & 0.15 & \\
\hline
increasing & w/ cov & reduced & no & spring & 0.31 & 0.17 & 0.4 & 0.3 & 0.32 & 0.22 & \\
\hline
increasing & w/ cov & reduced & yes & spring & 0.29 & 0.15 & 0.41 & 0.25 & 0.29 & 0.21 & \\
\hline
increasing & no cov & reduced & no & fall & 0.64 & 0.75 & 0.7 & 0.49 & 0.54 & 0.6 & \\
\hline
increasing & no cov & reduced & yes & fall & 0.66 & 0.89 & 0.69 & 0.48 & 0.53 & 0.62 & cov didnt do much\\
\hline
increasing & w/ cov & reduced & no & fall & 0.64 & 0.2 & 0.7 & 0.53 & 0.54 & 0.31 & \\
\hline
increasing & w/ cov & reduced & yes & fall & 0.66 & 0.13 & 0.69 & 0.5 & 0.53 & 0.32 & \\
\hline
\end{tabular}
\end{table}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{knitr}\SpecialCharTok{::}\FunctionTok{kable}\NormalTok{(Cod\_results, }\AttributeTok{caption =} \StringTok{"Cod error results"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{table}

\caption{\label{tab:unnamed-chunk-1}Cod error results}
\centering
\begin{tabular}[t]{l|l|l|l|r|r|r|l|l}
\hline
Temp & Strata & Noise & season & VAST.No.Cov & VAST.w..Cov & Stratified.Mean & X & X.1\\
\hline
const & all & no & spring & 0.11 & 0.12 & 0.36 & NA & \\
\hline
const & all & yes & spring & 0.12 & 0.15 & 0.35 & NA & Cod\\
\hline
const & all & no & fall & 0.19 & 0.05 & 0.49 & NA & Decreasing Population\\
\hline
const & all & yes & fall & 0.30 & 0.23 & 0.41 & NA & \\
\hline
const & reduced & no & spring & 0.17 & 0.24 & 0.41 & NA & \\
\hline
const & reduced & yes & spring & 0.20 & 0.23 & 0.46 & NA & \\
\hline
const & reduced & no & fall & 0.21 & 0.33 & 0.60 & NA & \\
\hline
const & reduced & yes & fall & 0.18 & 0.31 & 0.58 & NA & \\
\hline
increasing & all & no & spring & 0.12 & 0.15 & 0.25 & NA & \\
\hline
increasing & all & yes & spring & 0.19 & 0.19 & 0.27 & NA & \\
\hline
increasing & all & no & fall & 0.76 & 0.13 & 0.45 & NA & \\
\hline
increasing & all & yes & fall & 0.89 & 0.33 & 0.44 & NA & \\
\hline
increasing & reduced & no & spring & 0.14 & 0.22 & 0.32 & NA & \\
\hline
increasing & reduced & yes & spring & 0.15 & 0.21 & 0.29 & NA & \\
\hline
increasing & reduced & no & fall & 0.60 & 0.31 & 0.54 & NA & \\
\hline
increasing & reduced & yes & fall & 0.62 & 0.32 & 0.53 & NA & \\
\hline
\end{tabular}
\end{table}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{knitr}\SpecialCharTok{::}\FunctionTok{kable}\NormalTok{(Had\_results, }\AttributeTok{caption =} \StringTok{"Haddock error results"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{table}

\caption{\label{tab:unnamed-chunk-1}Haddock error results}
\centering
\begin{tabular}[t]{l|l|l|l|r|r|r|l|l}
\hline
Temp & Strata & Noise & season & VAST.No.Cov & VAST.w..Cov & Stratified.Mean & X & X.1\\
\hline
const & all & no & spring & 0.49 & 0.18 & 0.18 & NA & \\
\hline
const & all & yes & spring & 0.73 & 0.43 & 0.21 & NA & Haddock\\
\hline
const & all & no & fall & 0.28 & 0.05 & 0.26 & NA & Increasing Population\\
\hline
const & all & yes & fall & 0.41 & 0.06 & 0.27 & NA & \\
\hline
const & reduced & no & spring & 0.34 & 0.35 & 0.45 & NA & \\
\hline
const & reduced & yes & spring & 0.30 & 0.33 & 0.46 & NA & \\
\hline
const & reduced & no & fall & 0.36 & 0.48 & 0.54 & NA & \\
\hline
const & reduced & yes & fall & 0.33 & 0.46 & 0.52 & NA & \\
\hline
increasing & all & no & spring & 0.25 & 0.05 & 0.26 & NA & \\
\hline
increasing & all & yes & spring & 0.30 & 0.06 & 0.31 & NA & \\
\hline
increasing & all & no & fall & 0.89 & 0.23 & 0.40 & NA & \\
\hline
increasing & all & yes & fall & 1.04 & 0.35 & 0.42 & NA & \\
\hline
increasing & reduced & no & spring & 0.32 & 0.40 & 0.44 & NA & \\
\hline
increasing & reduced & yes & spring & 0.38 & 0.37 & 0.37 & NA & \\
\hline
increasing & reduced & no & fall & 0.44 & 0.64 & 0.72 & NA & \\
\hline
increasing & reduced & yes & fall & 0.42 & 0.62 & 0.70 & NA & \\
\hline
\end{tabular}
\end{table}

\newpage

\section{Introduction}

much of below is from \url{https://apps-nefsc.fisheries.noaa.gov/nefsc/ecosystem-ecology/}
or
\url{https://www.fisheries.noaa.gov/data-tools/fisheries-economics-united-states-data-and-visualizations}

\begin{itemize}
\tightlist
\item
  The eastern continental shelf is ecologically diverse and economically important
\end{itemize}

The Northeast United States continental shelf spans from the Outer Banks of North Carolina to the Gulf of Maine. The region covers over 250,000 km\(^2\) of ocean, extending over 200 km from shore in the largest areas in New England to just 30 km off shore in the southern regions. This ecologically diverse region contains approximately 18,000 vertebrate marine species. Commercial fisheries have been an important part of local economies for centuries. In 2019, New England fisheries produced \$22 billion in sales, which sustained over 200,000 jobs. Maintaining a healthy ecosystem is therefore vital to sustained ecological health and economic prosperity of the region. {[}@nefmc20{]}

\begin{itemize}
\tightlist
\item
  Bottom trawl survey is important for monitoring population trends
\end{itemize}

Fish stocks in this highly productive and economically important region are managed by the National Oceanic and Atmospheric Administration's (NOAA) Northeast Fisheries Science Center (NEFSC) in Woods Hole, Massachusetts. Federal biologists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (cite survey paper). The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf \textbf{(include image from johnson\_sosebee?)}. The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits \textbf{just listed a few uses of survey change/others?}.

The survey takes place twice each year- once in the spring and again in the fall. Since most spatial analyses and projections of future distributions typically assume a constant survey catchability and/or availability over time, NOAA's survey design includes sampling during approximately the same 2-3 week time period in each season.

\begin{itemize}
\tightlist
\item
  Climate change is happening
  Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades, resulting in a shift in spatial distributions of many species. Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments. The implication of this is that the bottom trawl survey is actually sampling the population during a different life cycle stage than was originally assumed, which can lead to biased stock assessments. We are therefore interested in analyzing the impact of climate change on the accuracy of future stock assessment models as measured by NOAA's ongoing bottom-trawl survey along the East coast.
\end{itemize}

\textbf{use more info from initial proposal}

\begin{itemize}
\item
  Fish are changing spatial distribution and have altered life stages (?) because of climate change
  NYE paper
\item
  Population indexing methods may be becoming biased as a result
\item
  Briefly describe our study to test this
\end{itemize}

To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depend on temperature preferences. We can then consider the impact of climate change by simulating scenarios with repeating temperature patterns and those where temperature increases on average over time. In both cases we analyze the ability of stratified random sampling to track population trends.

\section{Methods}

\begin{itemize}
\tightlist
\item
  Describe simulation study
\end{itemize}

We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on George's Bank, where movement of each species combine static species-specific habitat preferences with dynamic temperature preferences. Model dynamics are driven by dynamic temperature gradients estimated from data to create simulated data sets for each population where the true biomass is known. Using temperature gradients that repeat each year creates data sets with predictable, repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time. We conducting stratified random sampling on our simulation output to mimic the bottom trawl survey and compare the ability of contemporary indexing methods to track population trends.

\subsection{Population Model Formulation}

-- Used MixFishSim. Describe edits made to package

We use the R package \emph{MixFishSim} (MFS) to model our populations {[}@dolder2020highly{]}. MFS is a discrete spatiotemporal simulation tool where users can model multiple species under varying environmental conditions. The package uses a delay-difference population model with discrete processes for growth, death, and recruitment of the population. We formulate the following inputs for the MFS package to address our research question:

\emph{Study Area}

We obtained a shapefile for the 15 strata that comprise George's Bank to use as our modeling environment. We discritized the region into a raster with 88 rows and 144 columns. Haddock inhabit all 15 strata in the domain Cod inhabit 13 strata, and yellowtail exist in 9 strata. \textbf{include figure?}

\emph{Population Dynamics and Recruitment}

The time step for our models is one week. MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass of in each cell in our study area {[}@dolder2020highly{]}. Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits. We chose to represent recruitment in the model using a Beverton-Holt formulation. Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.

--Many parameter inputs were obtained from literature

Each species is assumed to have normally distributed temperature preferences (\(~N(\mu,\sigma)\)). We assume Yellowtail's preferences are \(N(8.75,4.25)\), while Haddock and Cod have preferences \(N(9,4)\). We chose these values by combining information in the literature with temperatures recorded in the bottom trawl survey. A full list of parameters used in our model can be seen below in Table ???.

\textbf{go through each process in model and discuss edits/removals?}

\emph{Temperature and Habitat Input}

The package was designed to generate theoretical habitat preferences using Gaussian Random Fields that combine with hypothetical temperature gradients to simulate an entire fishery. Since we are modeling real species on the northeast continental shelf w

We obtain estimated temperature data for the region for 2012 from FVCOM \textbf{cite}.

--Describe difference between increasing and constant temperature scenarios (images?)

-- Describe each scenario that is considered

We consider three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time. Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees calcius over the duration of the 20 year simulation.

\emph{Simulating Bottom Trawl Survey and Population Indexing}

-Describe post hoc sampling process and how data is used

After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year. The number of the samples taken in each strata reflect true values and we sample from the same weeks that the Spring and Fall surveys take place. Most strata contain enough cells to sample a unique location in each survey. For smaller strata we are forced to repeat locations. We then use the samples in contemporary population indexing methods to track population trends. Since we know the true population values in our simulations we can compare methods by comparing the error calculated from each estimate.

--Stratified mean vs VAST with and without covariates

We compare the stratified mean estimate of abundance to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model. The stratified mean is a typical survey-based approach that scales individual samples to the strata-level by considering the area of each strata, before scaling to the region-level based on the relative size of each strata. VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurance (presence/absence). If desired, VAST also allows users to include covariate data to better inform the model.

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

A range of data-limited methods for setting catch advice were evaluated for stocks where assessment models may be rejected due to strong, positive retrospective patterns. A method was considered to perform well if it limited overfishing without resulting in light exploitation rates (\(F << F_{MSY}\)), thereby allowing depleted stocks to recover to \(SSB_{MSY}\) (or for healthy stocks to remain there), and for high and stable catches (close to \(MSY\)).

Overall, none of the methods evaluated performed best across the scenarios exploring the different sources of the retrospective pattern (unreported catch or increasing \(M\)) and different levels of historical fishing intensity. A number of methods did perform well in many cases, however, while others performed consistently poorly, resulting in frequent and intense overfishing (\(F>>F_{MSY}\)). We performed simulations for a couple of scenarios with no source of retrospective patterns and found the expected result that all DLMs and the SCAA performed better (\(SSB\), \(F\), and catch were all closer to the \(MSY\) reference points) than when either source of retrospective patterns was present. Due to the focus of this study, we did not examine the no retrospective source in detail and do not comment on it further.

Currently, in the Northeast U.S., if an assessment model is rejected due to a large rho value in \(SSB\), the catch advice from that model is ignored and some data-limited approach is used. However, the rho-adjusted SCAA model performed better than a number of the alternatives explored here. Therefore, there should not necessarily be an expectation that a data-limited method will perform better than the rejected assessment model. The SCAA only resulted in high exploitation rates (\(F >> F_{MSY}\)) when unreported catch was the source of the retrospective pattern and for the scenario where \(F=F_{MSY}\) at the end of the base period that left the stock in relatively good condition (\(SSB \sim SSB_{MSY}\)). In contrast, this method was particularly effective when the stock was depleted and there was unreported catch. When \(M\) was the source of the retrospective pattern, the rho-adjusted SCAA method typically resulted in light exploitation rates, on average. The light exploitation rates in these cases were likely driven by the combination of using a rho-adjustment, but also using the lower \(M\) from the beginning of the base period rather than the higher \(M\) that occurred during the feedback period. Using an \(M\) value that is too low in a stock assessment will typically bias estimates of biomass and reference points too low, resulting in catch advice that is below target levels {[}@Johnsonetal2014; @Puntetal2021M{]}. The consequences of using a value for \(M\) that is too low versus too high is also asymmetrical {[}@Johnsonetal2014{]}, with negative consequences being more severe when \(M\) is assumed too high than low, and the results here are consistent with these previous conclusions.

The methods that adjusted recent average catches based on trends in the survey (Ismooth and Islope) performed well overall in terms of catch, stock status, and variation in catch. The method using the expanded survey biomass with the recent exploitation rate (ES-Frecent) also performed well and similarly to Ismooth. The performance of these methods was also generally robust among scenarios, with the exception of when there were unreported catches and the stock was depleted (see below). The generally positive performance of these methods was consistent with @Hilbornetal2002 and @CoxKronlund2008, both of which evaluated a variant of a ``hold-steady'' DLM. In the case of @Hilbornetal2002, the ``hold-steady'' DLM policy was designed to adjust catches in order to keep rockfish (\emph{Sebastes spp.}) populations at recently observed index levels, and did so by functioning as a constant escapement harvest control rule where target catches were set to zero below some pre-specified index level. In the variant used by @CoxKronlund2008, catches were adjusted to maintain a sablefish (\emph{Anoplopoma fimbria}) population at a pre-specified index level thought to be sustainable and desirable in terms of meeting fishery objectives (e.g., high catch), but never permitted target catches of zero and so functioned as a constant exploitation rate control rule. The ``hold-steady'' DLM of @CoxKronlund2008 performed similarly in terms of catch, stock depletion, and variation in catch, as a constant exploitation rate policy where target catch was specified as the product of desired exploitation rate and an estimate of biomass from a SCAA model. This result was robust to uncertainty in initial stock status and steepness {[}@CoxKronlund2008{]}. The SCAA model was always correctly specified (i.e., expected to produce unbiased estimates on average), however, and no comparison to the results of this research in the presence of retrospective patterns is possible {[}@CoxKronlund2008{]}. The ``hold-steady'' policy of @Hilbornetal2002 performed similarly to or better in terms of catch and stock status than other harvest control rules that relied on assessment estimates of biomass (i.e., 40:10 and constant \emph{F}). The performance of the ``hold-steady'' DLM was also more robust to uncertainty in steepness and to the presence of unreported catch {[}@Hilbornetal2002{]}. The performance of the two harvest policies that relied on assessment estimates of biomass (i.e., constant exploitation rate and a ``40:10'' biomass-based policy) also degraded when the estimates of biomass were biased, which is an issue that does not effect the ``hold-steady'' DLM {[}@Hilbornetal2002{]}. The bias in the assessment estimates considered in @Hilbornetal2002 were not necessarily induced by a retrospective pattern, however, and no consideration of making a rho-adjustment was possible in that study.

The Ismooth method is currently used to set catches for Georges Bank cod {[}@nefsc19{]} and red hake (\emph{Urophycis chuss}; @nefsc20). Variations of the ES-Frecent have been used for witch flounder and Georges Bank yellowtail flounder. While the findings here generally support the continued use of the Ismooth and ES-Frecent methods, they may not be well suited for depleted stocks where unreported catches are believed to be an issue. The Ismooth, Islope, and ES-Frecent DLMs produced high \(Fs\) and limited stock recovery with unreported catches and when the stock was depleted. While @Hilbornetal2002 and @CoxKronlund2008 did not reach the same conclusion about the ``hold-steady'' DLM, those studies did not consider initial levels of depletion as low as in this study. These results highlight the importance of accurate catch reporting, as unreported catch can create a negative feedback loop with perpetually high \(Fs\) being produced by a management system that seemingly should result in sustainable catch advice.

Three methods were consistently risk-averse across scenarios, limiting the frequency and magnitude of overfishing and resulting in high stock biomass. These methods were the two catch curve options (CC-FM and CC-FSPR) and DynLin. The catch curve methods produced a wider range of average catches across scenarios, and also had greater interannual variability in catches compared to DynLin. While the lower exploitation rates from these approaches may be undesirable due to forgone yield, there may be circumstances where they are preferred. For example, for stocks that are believed to be heavily depleted, low exploitation rates would allow for a more rapid recovery.

A number of methods performed poorly, particularly when catches were unreported. These methods include three of the expanded survey biomass approaches (ES-Fstable, ES-FM, ES-FSPR), AIM, and Skate. The AIM model has been widely used across stocks in the region {[}@nefsc02a; @nefsc05; @nefsc08{]}, although there is a decreasing trend in its use across model resistant stocks {[}@nefsc19{]}. The findings here suggest that alternative approaches should be considered in cases where AIM is still used and there is concern over unreported catches. The Skate method is used to manage the skate complex in the Northeast U.S. (a group of seven co-managed species). Interestingly, six of the seven species are considered in good condition with high survey biomass indices in recent years {[}@nefmc20{]}. That the Skate method performed poorly in our analysis but performs well for the skate complex illustrates how the performance of methods in this analysis may be sensitive to the scenarios and species life history considered. As may be the case for the Skate method, the performance of some methods may depend on the condition of the stock when the method is first applied, and less so on life-history. Therefore, care is needed when trying to generalize these results across stocks that may have different life histories, exploitation histories, and without unreported catches or increases in \(M\).

In addition to the analytical differences among the thirteen DLMs, most of the DLMs and control rules had multiple options that could be adjusted to make them more or less risk averse. DynLin had a large number of user defined decision points. Given the large range of options already explored in the study, one suite of options was selected for each DLM-control rule and kept constant for all simulations. Further studies could explore the different options within an individual DLM to understand how they might affect performance.

Many other data-limited methods exist for setting catch advice that were not included in this evaluation, and they vary widely in complexity, data inputs, and assumptions required {[}e.g., @carruthers2018dlm{]}. Length based methods were not evaluated to keep the overall number of methods tractable, and due to the availability of age based information in the region. Methods that require only catch data or snap shots of survey data were not considered due to the availability of the relatively long and contiguous Northeast Fisheries Science Center's spring and fall, coastwide bottom trawl surveys, and the fact that ``catch only'' methods have been shown to perform poorly {[}e.g., @carruthers2014eval{]}. Complete catch histories are not available for stocks in the region (i.e., from the inception of fishing). Consequently, methods that required complete catch histories or required assumptions about relative depletion {[}e.g., DCAC in @maccall2009dca; DB-SRA in @dick2022dsra{]} were also omitted from consideration. The need for short run-times and the desire for methods that could be reviewed quickly prevented the use of modern state-space production models such as SPiCT {[}@pedersen2017spict{]} and JABBA {[}@winker2018jabba{]}.

The SCAA was confronted with inconsistent data in this study, while the DLMs typically used only a single source of data and thus did not encounter inconsistencies. A recent examination of the data used in assessments in this region similarly found inconsistencies in data streams even before modeling. @wiedenmann2022strange found a negative relationship between relative F (catch/survey) and survey Z for stocks with strong retrospective patterns but the expected positive relationship for stocks without a retrospective pattern. It is exactly this sort of tension that creates retrospective patterns in integrated models, but is not found in DLMs that only use one type of data.

Despite conducting hundreds of thousands of simulations, there are still limitations to our study. We only examined one life history representative of groundfish in the region. We acknowledge that best practice is to select a DLM for a specific life history and fishery condition {[}e.g., @fischer2020dlm{]}. As is typically the case with large simulation studies, we were not able to tune any of the DLMs or the SCAA in any given realization, which would occur in practice for an actual stock assessment. We also examined only scenarios that started with Mohn's rho values near 0.5 for spawning stock biomass. This is a strong retrospective pattern, but some stocks in the region have even stronger retrospectives. Performance of the DLMs and SCAA would be expected to degrade with stronger retrospectives, but by how much is still an open area for research. Similarly, sources of retrospective patterns that create different relationships between the true values and estimated values should also be explored {[}see @deroba2014retro{]}. To make the results interpretable, we only examined a single source for the retrospective pattern at a time. In reality, there may be more than one factor leading to an observed retrospective pattern. How the multiple sources would interact to influence performance is another topic for future research. Development of harvest control rules specifically for situations where retrospective patterns are found in age-based assessments would also be beneficial. The large number of scenarios examined and the large number of realizations gives us confidence that our results are meaningful in general, but that the performance of any of the DLMs may differ in actual practice.

An interesting finding of this study is the linear versus diffuse patterns between \(SSB\) and catch across methods. These patterns have implications for the trade-offs among methods, with linear relationships resulting in more consistent exploitation rates across stock sizes. Therefore, these methods have higher certainty of a given catch at a given stock size. However, they also tended to result in lower stock sizes, on average, across methods. The more diffuse relationships resulted in more variable exploitation rates across stock sizes, with some situations where the population biomass was quite high but the catch was low (relative to MSY), resulting in a very low \(F\). The reasons behind these different patterns remain unclear, and future work to explore these patterns is warranted.

One of the reasons for the difference in performance between the catch and natural mortality retrospective sources was how the reference points were calculated. In all cases, the initial conditions, including the natural mortality rate, were used to compute the reference points. This decision was made based on the fact that the increase in natural mortality was assumed to be unknown in the simulations. If the increase in natural mortality was known, the age-structured assessments would have accounted for it, different reference points might have been computed {[}@legault2016increaseM{]} and there may not have been a retrospective pattern at all {[}@legault2020rose{]}, and no need to consider alternative DLMs. The reference points for the increased \(M\) scenarios would have been different if they were computed using the values from the final year of the base period, but the overall conclusions regarding the different DLMs would not change as this just results in a rescaling of the axis. These results are not shown to reduce confusion regarding the simulations.

Closed-loop simulation is a common tool for examining performance of catch advice from various stock assessment approaches in a feedback setting. It is often used as part of a full management strategy evaluation when working with stakeholders to develop management regulations that make trade offs between near term and long term catches, risk to the fish population, and mixed-fleet allocations {[}@carruthers2016simpleMPs; @goethel2019mse; @harlyan2019hcr{]}. We did not conduct a full management strategy evaluation with stakeholder input {[}@goethel2019stakeholder{]}, but see that as a fruitful next step that could build on the conclusions from our closed-loop work. Using a generic groundfish life-history and monitoring standard performance metrics related to stock status and catch stability, we were able to cull the herd of potential DLMs and we would not carry the consistent poor performers forward for further study. The wide range of expertise reflected in the authorship was by design so that the simulation specifications and performance metrics were broadly useful. Before undertaking a full management strategy evaluation and engaging regional stakeholders, we would want to select a specific stock and jointly identify specific management regulations to be tested {[}@deroba2019dream{]}. Results of this work have been presented to both local fishery management councils, with generally positive feedback about the utility of the conclusions for identifying appropriate model approaches when an SCAA is rejected. Our work was similar to all other closed-loop simulations in that it was designed to address a specific situation, including much recent work comparing the performance of data-limited and data rich assessment approaches {[}e.g., @fulton2016datarich; @sagarese2019dlm; @bouch2020datapoor; @li2022dlm{]}.

This study is a first attempt to identify suitable methods for setting catch advice when stock assessment models are rejected due to large, positive retrospective patterns. Although no single method performed best across scenarios, a number of generally suitable and unsuitable methods were identified under specific conditions. The results of this work can help scientists and managers select a subset of possible options for consideration to set catch advice when assessment models are rejected. The approach developed here can, and should be expanded to consider other cases not explored here, as performance of individual methods are very likely case-dependent.

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

We thank the Index-Based Methods and Control Rules Research Track review panel of Paul Rago (chair), Yong Chen, Robin Cook, and Paul Medley for feedback on preliminary results, three anonymous reviewers and the associate editor for reviewing an earlier version of this work. The scientific results and conclusions, as well as any views or opinions expressed herein, are those of the authors and do not necessarily reflect those of NOAA or the Department of Commerce.

\hypertarget{data-and-code-availability}{%
\subsection{Data and Code Availability}\label{data-and-code-availability}}

All data and code used in this work are available at \url{https://github.com/cmlegault/IBMWG}.

\hypertarget{references}{%
\subsection{References}\label{references}}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\end{CSLReferences}

\pagebreak

\hypertarget{tables}{%
\subsection{Tables}\label{tables}}

\pagebreak

Table 1. Parameters used in all population models.

\begin{longtable}[]{@{}lllllll@{}}
\toprule
& Description & Unit & Yellowtail & Cod & Haddock & Source \\
\midrule
\endhead
\(\rho\) & Ford's growth coefficient & wk\(^{-1}\) & 4.48 & 4.43 & 4.49 & \\
\(M\) & Natural Mortality & wk\(^{-1}\) & 0.2064 & 0.2728 & 0.3340 & \\
\(W_R\) & Weight of fully recruited fish & kg & 0.39 & 2.95 & 1.12 & \\
\(W_{R-1}\) & Weight of pre-recruit fish & kg & 0.13 & 0.39 & 0.19 & \\
\bottomrule
\end{longtable}

Table 2. Parameters used in models with relatively constant populations.

\begin{longtable}[]{@{}llllll@{}}
\toprule
& Description & Yellowtail & Cod & Haddock & \\
\midrule
\endhead
\(M+F\) & Adjusted Mortality (Natural + Fishing) & wk\(^{-1}\) & 0.764 & 0.83 & 0.309 \\
\bottomrule
\end{longtable}

Table 3. Parameters used in models with increasing populations.

\begin{longtable}[]{@{}llllll@{}}
\toprule
& Description & Yellowtail & Cod & Haddock & Source \\
\midrule
\endhead
\(M+F\) & Adjusted Mortality (Natural + Fishing) & wk\(^{-1}\) & 0.564 & 0.372 & 0.134 \\
\bottomrule
\end{longtable}

Table 4. Parameters used in models with decreasing populations.

\begin{longtable}[]{@{}llllll@{}}
\toprule
& Description & Yellowtail & Cod & Haddock & Source \\
\midrule
\endhead
\(M+F\) & Adjusted Mortality (Natural + Fishing) & wk\(^{-1}\) & 0.764 & 0.623 & 0.334 \\
\bottomrule
\end{longtable}

Table 1. Maturity-, weight-, and selectivity-at-age of the simulated fish population.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.12}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.22}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.22}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.22}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.22}}@{}}
\toprule
\begin{minipage}[b]{\linewidth}\raggedright
Age
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Maturity
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Weight (kg)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Fishery Selectivity (before change if applicable)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Fishery Selectivity (after change if applicable)
\end{minipage} \\
\midrule
\endhead
1 & 0.04 & 0.15 & 0.07 & 0.02 \\
2 & 0.25 & 0.5 & 0.17 & 0.05 \\
3 & 0.60 & 0.9 & 0.36 & 0.12 \\
4 & 0.77 & 1.4 & 0.61 & 0.27 \\
5 & 0.85 & 2.0 & 0.81 & 0.50 \\
6 & 0.92 & 2.6 & 0.92 & 0.74 \\
7 & 1.00 & 3.2 & 0.97 & 0.89 \\
8 & 1.00 & 4.1 & 0.99 & 0.96 \\
9 & 1.00 & 5.9 & 1.00 & 0.99 \\
10+ & 1.00 & 9.0 & 1.00 & 1.00 \\
\bottomrule
\end{longtable}

\pagebreak

Table 2. Naming convention and details of the data-limited methods evaluated.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.36}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.64}}@{}}
\toprule
\begin{minipage}[b]{\linewidth}\raggedright
Method
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Details
\end{minipage} \\
\midrule
\endhead
Ismooth & \(C_{targ,y+1:y+2} =\overline{C}_{3,y}(e^{\lambda})\) where \(\overline{C}_{3,y}\) is the most recent three year average; \(\overline{C}_{3,y} = \frac{1}{3}\sum_{t=1}^{t=3}C_{y-t}\) and \(\lambda\) is the slope of a log linear regression of a LOESS-smoothed average index of abundance (spring and fall) with span = 0.3: \(\hat{I}_y=loess(\hat{I}_y)\) and \(LN(\widehat{I_y}) = b + \lambda y\) \\
Islope & \(C_{targ,y+1:y+2} = 0.8\overline{C}_{5,y}(1+0.4e^{\lambda})\) where \(\overline{C}_{5,y}\) is the most recent five-year average catch through year \(y-1\): \(\overline{C}_{5,y} = \frac{1}{5}\sum_{t=1}^{t=5}C_{y-t}\) and \(\lambda\) is the slope of a log-linear regression of the most recent five years of the averaged index. \\
Itarget & \(C_{targ,y+1:y+2}=\left[0.5C_{ref}\left(\frac{\overline{I}_{5,y}-I_{thresh}}{I_{target}-I_{thresh}}\right)\right]\) \(\overline{I}_{5,y} \geq I_{thresh}\) \(C_{targ,y+1:y+2}=\left[0.5C_{ref}\left(\frac{\overline{I}_{5,y}}{I_{thresh}}\right)^{2}\right]\) \(\overline{I}_{5,y} < I_{thresh}\); \(C_{ref}\) is the average catch over the reference period (years 26 through 50): \(C_{ref} = \frac{1}{25}\sum_{y=26}^{y=50}C_y\); \(I_{target}\) is 1.5 times the average index over the reference period: \(I_{target} = \frac{1}{25}\sum_{y=26}^{y=50}\overline{I}_y\); \(I_{thresh}\) = 0.8 \(I_{target}\), and is the most recent five year average of the combined spring and fall index: \(\overline{I}_{5,y} = \frac{1}{5}\sum_{t=1}^{t=5}\overline{I}_{y-t+1}\) \\
Skate & \(C_{targ,y+1:y+2}= F_{rel} \overline{I}_{3,y}\) where \(F_{rel} =median \left(\frac{\overline{C}_{3,\textbf{Y}}}{\overline{I}_{3,\textbf{Y}}} \right)\) is the median relative fishing mortality rate calculated using a 3 year moving average of the catch and average survey index across all available years (\(\textbf{Y}\)): \(\overline{C}_{3,y} = \frac{1}{3} \sum_{t=1}^{t=3} C_{y-t}\) and \(\overline{I}_{3,y} = \frac{1}{3} \sum_{t=1}^{t=3} I_{y-t+1}\) \\
An Index Method (AIM) & AIM first calculates the annual relative \(F\): \(F_{rel,y}=\frac{C_y}{\frac{1}{3}\sum_{t=1}^{t=3}\overline{I}_{y-t+1}}\) and the annual replacement ratio: \(\Psi_y = \frac{\overline{I}_y}{\frac{1}{5}\sum_{t=1}^{t=5}\overline{I}_{y-t}}\). These values are used in a regression: \(LN(\Psi_y)= b + \lambda LN(F_{rel,y})\) to determine \(F_{rel,*}\), which is the value of \(F_{rel,y}\) where the predicted \(\Psi=1\) or \(LN(\Psi)=0\). \(F_{rel,*}\) is called either the ``stable'' or ``replacement'' \(F\), and is used to calculate the target catch: \(C_{targ,y+1:y+2} = \overline{I}_yF_{rel,*}\). \\
Dynamic Linear Model (DynLin) & @Langan2021DLM. \\
Expanded survey biomass method 1 \(F_{40\%}\) (ES-FSPR) & \(C_{targ,y+1:y+2}=B_{\bar{I},y}\mu_{targ}\) where \(B_{\bar{I}}\) is the average of estimated fully-selected biomass from each survey: \(B_{\bar{I},y}=\frac{1}{2}\left(\frac{I_{spr,y}}{q_{spr}} +\frac{I_{fall,y-1}}{q_{fall}}\right)\) and target exploitation fraction, \(\mu_{targ}\) is calculated as: \(\mu_{targ}=\frac{F_{targ}}{Z_{targ}}\left(1-e^{-Z_{targ}}\right)\); \(F_{targ}=F_{40\%}\) and \(Z_{targ}=F_{targ}+M\) \\
Expanded survey biomass method 2 \(F=\) AIM replacement (ES-Fstable) & Same as the above expanded survey method, but with \(\mu_{targ}\) equal to the stable exploitation fraction \(F_{rel,*}\) calculated using the AIM approach (see above). \\
Expanded survey biomass method 3 \(F=M\) (ES-FM) & Same as the above expanded survey methods, but with the target exploitation rate set to the assumed \(M\): \(F_{targ}=M\). \\
Expanded survey biomass method 4 \(F=\) recent average (ES-Frecent) & Same as the above expanded survey methods, but with the target exploitation fraction set to the most recent three year average exploitation fraction: \({\mu_{targ}}=\frac{\sum_{y-2}^{y}\mu_y}{3}\) \(\mu_{y}=\frac{C_{y-1}}{B_{\bar{I},y}}\) \\
Catch curve Method 1 \(F_{40\%}\) (CC-FSPR) & \(C_{targ,y+1:y+2}= \frac{F_{targ}}{Z_{avg,y}}B_{cc,y}\left(1-e^{-Z_{avg,y}}\right)\) where \(B_{cc}\) is the estimated biomass: \(B_{cc,y} = \frac{C_{y-1}}{\frac{F_{avg,y}}{Z_{avg,y}}{\left(1-e^{-Z_{avg,y}}\right)}}\) with \(Z_{avg,y} = \frac{Z_{spring, y}+Z_{fall,y-1}}{2}\); \(F_{avg,y-1} = Z_{avg,y-1} - M\) and, \(F_{targ}=F_{40\%}\). Survey catch at age used in catch curve to estimate \(Z\). \\
Catch curve Method 2 \(M\) (CC-FM) & Same as catch curve method 1 above, but with \(F_{targ}=M\). \\
Ensemble & Median of catch advice provided by AIM, CC-FSPR, ES-Frecent, ES-FSPR, Islope, Itarget, Ismooth, and Skate methods. \\
\bottomrule
\end{longtable}

\pagebreak

Table 3. Summary of the scenarios evaluated within the study design.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.50}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.50}}@{}}
\toprule
\begin{minipage}[b]{\linewidth}\raggedright
Factors
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Variants
\end{minipage} \\
\midrule
\endhead
retrospective source & catch or natural mortality \\
fishing history & \(F_{MSY}\) in second half of base period or overfishing throughout base period (2.5x\(F_{MSY}\)) \\
fishery selectivity blocks & constant selectivity or selectivity changes in second half of base period \\
catch advice multiplier & applied as is from DLM (1) or reduced from DLM (0.75) \\
\bottomrule
\end{longtable}

\pagebreak

\hypertarget{list-of-figures}{%
\subsection{List of Figures}\label{list-of-figures}}

Figure 1. Inner quartiles and medians for all performance measures across all scenarios and runs for each method. Vertical lines are shown at a value of 1 for the performance measures that are relative to the MSY reference points (A,B,C).

Figure 2. Relationship between long-term average spawning biomass and average catch (relative to MSY levels) for each method. Each point represents the median for a given scenario, separated by the source of the retrospective pattern (catch or M).

Figure 3. Median performance measures for each method, separated by the source of the retrospective error (catch = black, M = gray) and the exploitation history in the base period (always overfishing at \(2.5 x F_{MSY}\) (circle), or \(F\) reduced to \(F_{MSY}\) during base period (triangle)). Vertical lines are shown at a value of 1 for the performance measures that are relative to the MSY reference points (A,B,C).

Figure 4. Median \(F / F_{MSY}\) for each method, with results separated by the exploitation history in the base period (always overfishing at \(2.5 x F_{MSY}\) (circle), or \(F\) reduced to \(F_{MSY}\) during base period (triangle)) showing A) short- (gray) versus long-term (black) values, and B) with (black) or without (gray) a buffer applied when setting the catch (catch multiplier = 0.75 or 1).

Figure 5. Relationship between long-term average catch and spawning stock biomass relative to their reference points by method. Each point represents the average for years 21-40 in the feedback period for a single iteration of a scenario. The scenario shown is where catch was the source of the retrospective pattern with \(F\) reduced to \(F_{MSY}\) in the second half of the base period, there was a single selectivity block, and where no buffer was applied to the catch advice (catch multiplier = 1).

\end{document}

### 2022-12-20 Manuscript/Example_from_Chris.Rmd
updating plotting code and manuscript

--- 2022-12-20 Manuscript/Draft1.Rmd
@@
 output:
+  pdf_document:
+    toc: no
   bookdown::pdf_document2:
@@
     keep_tex: yes
-  bibliography: IBMWG.bib
   word_document: default
   html_document: default
-  pdf_document:
-    toc: no
 fontsize: 12pt
@@
   require(kableExtra)}
-
-#read in results tables
-YT_results<-read.csv("Yellowtail_results.csv")
-Cod_results<-read.csv("Cod_results.csv")
-Had_results<-read.csv("Haddock_results.csv")
 ```
@@
 
-# Estimating Population Trends with Stratified Random Sampling Under the Pressures of Climate Change
+# Data Rich but Model Resistant: An Evaluation of Data-Limited Methods to Manage Fisheries with Failed Age-based Stock Assessments
 
-Benjamin A. Levy^1^, Christopher M. Legault^2^, Timothy J. Miller^2^, Elizabeth N. Brooks^2^
+Christopher M.
+Legault^1^, John Wiedenmann^2^ Jonathan J.
+Deroba^1^, Gavin Fay^3^, Timothy J.
+Miller^1^, Elizabeth N.
+Brooks^1^, Richard J.
+Bell^4^, Joseph A.
+Langan^5^, Jamie M.
+Cournane^6^, Andrew W.
+Jones^1^, Brandon Muffley^7^
 
-^1^Ben's Institution, USA  
-^2^National Marine Fisheries Service, Northeast Fisheries Science Center, Woods Hole, MA, USA
+^1^National Marine Fisheries Service, Northeast Fisheries Science Center, Woods Hole, MA, USA  
+^2^Department of Ecology, Evolution and Natural Resources, Rutgers, The State University of New Jersey, New Brunswick, NJ, USA  
+^3^University of Massachusetts Dartmouth, School for Marine Science
+and Technology, MA, USA  
+^4^The Nature Conservancy, RI, USA  
+^5^University of Rhode Island, Graduate School of Oceanography, Narragansett, RI, USA  
+^6^New England Fishery Management Council, Newburyport, MA, USA  
+^7^Mid-Atlantic Fishery Management Council, Dover, DE, USA
 
-Corresponding author: Ben Levy (benjamin.levy@noaa.gov)
+Corresponding author: C.M. Legault (chris.legault@noaa.gov)
 
 Competing interests: The authors declare there are no competing interests.
-
-<!-- Table1 <- YT_results -->
-
-<!-- write.table( -->
-<!--    format(f, digits=2), "", -->
-<!--    sep="|", row.names=F, col.names=F, quote=F, eol="|\n") -->
 
@@
 
-An Abstract
+Age-based stock assessments are sometimes rejected by review panels due to large retrospective patterns.
+When this occurs, data-limited approaches are often used to set catch advice, under the assumption that these simpler methods will not be impacted by the problems causing retrospective patterns in the age-based assessment.
+This assumption has never been formally evaluated.
+Closed-loop simulations were conducted where a known source of error caused a retrospective pattern in an age-based assessment.
+Twelve data-limited methods, an ensemble of a subset of these methods, and a statistical catch-at-age model with retrospective adjustment were all evaluated to examine their ability to prevent overfishing and rebuild overfished stocks.
+Overall, none of the methods evaluated performed best across the scenarios.
+A number of methods performed consistently poorly, resulting in frequent and intense overfishing and low stock sizes.
+The retrospective adjusted statistical catch-at-age assessment performed better than a number of the alternatives explored.
+Thus, using a data-limited approach to set catch advice will not necessarily result in better performance than relying on the age-based assessment with a retrospective adjustment.
 
@@
 
-keyword 1, keyword 2
+closed-loop simulation, data-limited methods, retrospective analysis, management advice
 
@@
 
-```{r}
-knitr::kable(YT_results, caption = "Yellowtail error results")
+## Introduction
+In the U.S., age-based, integrated, fisheries stock assessment models are frequently used to estimate annual stock abundance (biomass), fishing mortality rates, and management reference points [@maunder2013review].
+These models must undergo peer review, where an independent panel of experts determines whether or not results from the model are suitable as the basis for determining stock status and for setting catch advice.
+There are a number of model diagnostics that are used to evaluate uncertainty and stability of assessment model results, but one that is commonly used and carries substantial weight during review is the retrospective pattern.
+A retrospective pattern is a systematic inconsistency among a series of sequential assessment estimates of population size (or other related assessment variables), based on increasing time periods of data used in the model fitting [@mohn1999rho].
+These inconsistencies in assessment estimates are indicative of one or more mismatches between model assumptions and patterns in the data used to fit the model.
+Large or persistent retrospective patterns indicate an instability in model results, and may therefore be the basis for a peer review panel to determine that model results are not suitable for management purposes [@punt2020reject].
 
-knitr::kable(Cod_results, caption = "Cod error results")
+Many stock assessments in the Northeast U.S.
+have a history of strong retrospective patterns, whereby estimates of biomass are typically revised downward and estimates of fishing mortality rate are revised upward as new data are added to the model (i.e., implying systematic overestimation of biomass and underestimation of fishing mortality; [@ices2020wkforbias]).
+NOAA Fisheries, the New England Fishery Management Council, the Mid-Atlantic Fishery Management Council, and the Atlantic States Marine Fisheries Commission manage these stocks, and retrospective issues remain a challenge for managers when setting catch advice and tracking stock status.
+This problem has been particularly acute for, but not limited to, stocks in the New England groundfish complex [@nefsc02a; @nefsc05; @nefsc08; @nefsc15a; @nefsc15b; @nefsc17; @nefsc19; @deroba2010mackerel], managed under NOAA Fisheries and the New England Council’s Northeast Multispecies (Groundfish) fishery management plan.
+Stock assessments exhibiting retrospective patterns can be found around the world and can be associated with a wide range of assessment approaches [@ices2020wkforbias].
 
-knitr::kable(Had_results, caption = "Haddock error results")
-```
+The magnitude of the retrospective pattern is typically measured with a statistic called Mohn’s rho [@mohn1999rho].
+Mohn's rho can be used to adjust terminal year estimates of biomass in anticipation that the retrospective pattern will persist, and some accounting for the pattern will provide a more accurate estimate.
+Stock assessments where the so-called rho-adjusted value is outside the 90\% confidence interval of the terminal year estimate of spawning stock biomass ($SSB$) or fishing mortality rate are classified as strong retrospective patterns.
+In these cases, the rho-adjusted values are used for status determination and to modify the starting population for projections used to provide catch advice [@brooks2016retroforecast].
 
-\newpage
+There are many possible causes for retrospective patterns, but typically there is a temporal change in either the data or a model parameter that is not accounted for in the stock assessment model [@legault2020rose; @hurtado2014rearview; @deroba2014retro].
+The strong retrospective patterns seen in the region under study have required very large magnitudes of change in order to remove the retrospective pattern.
+For example, the scale of missing catch needed to be three to five times the reported catch, or natural mortality needed to increase from 0.2 to near 1.0 to reproduce observed retrospective patterns; the scales of these changes have not been deemed believable by review panels.
+Some approaches have been used to estimate missing catch [@vanbeveren2017catch; @perretti2020sim] and increased natural mortality [@cadigan2016ss; @rossi2019inferring].
+However, identifying the correct source of the retrospective pattern is difficult and using the wrong fix can lead to poor management advice [@szuwalski2017retro].
+This is clearly an area where more research is needed, but currently addressing strong retrospective patterns is challenging.
 
-\section{Introduction}
+There is no formal criteria in the region for rejecting an assessment based on Mohn’s rho, but large, positive values of rho for $SSB$, especially those persisting across several assessments, have played an important role in the rejection of recent age-based assessments, including Atlantic mackerel (*Scomber scombrus*), Georges Bank Atlantic cod (*Gadus morhua*), Georges Bank yellowtail flounder (*Limanda ferruginea*), and witch flounder (*Glyptocephalus cynoglossus*) [@deroba2010mackerel; @legault2014tracgbytf; @nefsc15a; @nefsc15b].
+In each of these cases, and another where the assessment rejection was not based on the retrospective pattern [black sea bass, *Centropristis striatus*, @nefsc12], the Councils have relied on a variety of data-limited approaches for setting catch advice for these stocks [@mcnamee2015mafmc; @nefsc15a; @nefsc15b; @wiedenmann2015mackerel].
+These approaches have all been ad-hoc, and a recent analysis suggested that some of the data-limited approaches may not be suitable for stocks in the Northeast U.S.
+with a history of high exploitation rates [@wiedenman2019dlm].
+In addition, large, positive retrospective patterns in $SSB$ persist for a number of other stocks in the region [@nefsc19], raising concerns that additional stocks may rely on data-limited approaches in the future.
 
-much of below is from https://apps-nefsc.fisheries.noaa.gov/nefsc/ecosystem-ecology/
-or
-https://www.fisheries.noaa.gov/data-tools/fisheries-economics-united-states-data-and-visualizations
+Current practice in the region requires identification of a back-up assessment approach for all age-based assessments in case the age-based assessment is rejected during peer review.
+These back-up approaches are required to be simple enough that only minor review is needed so that management advice can continue to be provided for the stock.
+While these DLMs cannot provide stock status determinations in our study because they rely on ad hoc setting of reference points, they all can provide catch advice.
+Therefore, there is an immediate need to identify suitable data-limited approaches for setting catch advice for stocks with age-based assessments that did not pass review.
 
-- The eastern continental shelf is ecologically diverse and economically important
+We developed a closed-loop simulation [e.g., @punt2016mse; @huynh2022retro] to evaluate the suitability of alternative data-limited methods (DLMs) for setting target catches when age-based stock assessments fail.
+In particular, focus was placed on methods that use survey indices of abundance.
+The closed-loop simulation was designed to test the two most common hypothesized sources of retrospective pattern (missing catch or increases in natural mortality), and to evaluate performance of various methods relative to exploitation history and changes in fishery selectivity.
+Results of this factorial simulation study are summarized for quantities of interest that impact fisheries management advice.
+The goal of this work is to examine the hypothesis that catch advice from DLMs is more robust to under-reported catch or changes in natural mortality than from a rho-adjusted statistical catch at age model.
 
-The Northeast United States continental shelf spans from the Outer Banks of North Carolina to the Gulf of Maine.
-The region covers over 250,000 km$^2$ of ocean, extending over 200 km from shore in the largest areas in New England to just 30 km off shore in the southern regions.
-This ecologically diverse region contains approximately 18,000 vertebrate marine species.
-Commercial fisheries have been an important part of local economies for centuries.
-In 2019, New England fisheries produced $22 billion in sales, which sustained over 200,000 jobs.
-Maintaining a healthy ecosystem is therefore vital to sustained ecological health and economic prosperity of the region.
-[@nefmc20]
+## Methods
 
-- Bottom trawl survey is important for monitoring population trends
+<font size="5">*Overview*</font>
 
-Fish stocks in this highly productive and economically important region are managed by the National Oceanic and Atmospheric Administration’s (NOAA) Northeast Fisheries Science Center (NEFSC) in Woods Hole, Massachusetts.
-Federal biologists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (cite survey paper).
-The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf **(include image from johnson_sosebee?)**.
-The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits **just listed a few uses of survey change/others?**.
+A closed-loop simulation was designed to approximate a process where an age-based assessment was rejected due to a retrospective pattern, requiring catch advice to be determined using a DLM.
+As such, the operating model (OM) used to define the "true" underlying biological and fishery dynamics was also age-based.
+The OM was run for an initial 50 year period of time (called the base period) that controls the historical population dynamics and fishing pressure, and allows for sufficient data to be simulated in the observation model to be used in the different DLMs.
+After the base period, a given management approach (i.e., DLM) was applied to set the target catch for the stock, which is then removed from the population.
+This process is repeated at a fixed interval for 40 years in what is called the feedback period.
+Multiple OMs were developed so that the performance of the DLMs could be compared among several sources of uncertainty that are especially common in the northeast U.S., but relevant more broadly.
+The set of OMs featured one of two possible patterns of time varying dynamics in the last 20 years of the base period, that if left misspecified as time invariant, would be sufficient to generate retrospective patterns resulting in the rejection of an age-based stock assessment, requiring transition to a DLM.
+The details of these dynamics, and the suite of factors explored in the closed-loop simulation, are described in sections below.
 
-The survey takes place twice each year- once in the spring and again in the fall.
-Since most spatial analyses and  projections of future distributions typically assume a constant survey catchability and/or availability over time, NOAA's survey design includes sampling during approximately the same 2-3 week time period in each season.
+<font size="5">*Operating and Observation Models*</font>
 
-- Climate change is happening
- Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades, resulting in a shift in spatial distributions of many species.
-Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments.
-The implication of this is that the bottom trawl survey is actually sampling the population during a different life cycle stage than was originally assumed, which can lead to biased stock assessments.
-We are therefore interested in analyzing the impact of climate change on the accuracy of future stock assessment models as measured by NOAA’s ongoing bottom-trawl survey along the East coast.
+The Woods Hole Assessment Model [WHAM, @miller2020wham; @stock2021wham] was used as the basis for the OM in the closed-loop simulations.
+WHAM is an R package and the general model is built using the Template Model Builder package [@kristensenetal2016TMB].
+While WHAM can serve as a stock assessment model used to estimate parameters, it can also simulate the data needed for age-based stock assessments and DLMs given a range of input parameters.
+WHAM was used to simulate data with known properties during the base and feedback periods.
+Catch and index observations upon which the DLMs largely relied were simulated according to user supplied biological and fishery parameters for each scenario (see below).
+Catches during the feedback period were iteratively updated based on a DLM and harvest control rule that used the simulated observations to produce catch advice.
+Catch advice from a given combination of DLM and control rule was specified in two year blocks, a typical catch specification timeframe for New England and Mid-Atlantic Council managed fisheries.
+WHAM used these catches, along with the user supplied biological and fishery inputs, to have the simulated population respond to the DLM, thereby completing the closed-loop simulation aspect.
+A limit was placed on the maximum fishing mortality rate when the fishery attempted to remove the catch advice from the population during the feedback period.
+There was no implementation error in the removal of the catch advice otherwise, except when missing catch was the source of the retrospective pattern as described below.
 
-**use more info from initial proposal**
+The age-structured OM had ten ages, with the oldest age being a plus group.
+Maturity- and weight-at-age were time and simulation invariant and reflected values observed for groundfish in the region (Table 1).
+The OM simulated catch and age composition data for a single fishery with logistic selectivity (Table 1; see below).
+Annual, total catch observations (metric tons) were simulated as lognormal deviations from the underlying "true" catches with a coefficient of variation (CV) equal to 0.1.
+Fishery age composition data were assumed to follow a multinomial distribution with an effective sample size (ESS) equal to 200.
+Two fishery independent surveys were simulated and were intended to represent the spring and fall, coastwide bottom trawl surveys conducted in the region.
+Both surveys were assumed to have time invariant logistic selectivity and constant catchability.
+Annual survey observations were simulated as lognormal deviations from the underlying "true" survey catches with a CV of 0.3 in the spring survey and 0.4 in the fall.
+Survey age composition data were assumed to follow a multinomial distribution with an ESS equal to 100 in both seasons.
 
-- Fish are changing spatial distribution and have altered life stages (?) because of climate change
-NYE paper
+Annual recruitment was simulated as autoregressive, lag-1 (AR-1) deviations from an underlying Beverton-Holt stock-recruitment relationship with steepness equal to 0.74.
+The degree of correlation in the AR-1 process equaled 0.4 with a conditional standard deviation about this relationship equal to 0.5.
+Unfished recruitment was time- and simulation invariant and equaled 10-million age-1 fish.
+These stock-recruitment values were based on an average of groundfish parameters estimated for the region.
 
-- Population indexing methods may be becoming biased as a result
+<font size="5">*Data-Limited Methods Explored*</font>
 
-- Briefly describe our study to test this
+The range of DLMs evaluated was generally constrained to those that have been used or were considered plausible (e.g., based on data requirements) for the Northeast Shelf.
+Ultimately, thirteen DLMs were selected for evaluation.
+Although catch-curve analyses are not currently applied in the region, they were included here since age information is available for most of the stocks, and because @wiedenman2019dlm showed they performed well in application to groundfish stocks.
+Two additional DLMs (Islope and Itarget) not currently used in the region were also evaluated, as these have been tested in other applications and shown promise [@geromont2015complex; @geromont2015datapoor; @carruthers2016simpleMPs; @wiedenman2019dlm].
+An ensemble of models was also considered based on recent findings that improved performance can result from combining the results from multiple models [@anderson2017superensemble; @rosenberg2018ensemble; @spence2018combineecomodels; @stewart2018ensemble].
+The catch advice from the ensemble approach equaled the median of the catch advice resulting from the range of methods included in the ensemble (Table 2).
+This assumes an equal weighting of ensemble members.
+The DynLin approach was excluded from the ensemble due to the relatively long computing time required.
+Other methods were excluded (CC-FM, ES-FM, ES-Fstable) because they were slight variations of a more generic DLM (i.e., CC- and ES-) and including them all may have unduly overweighted the performance of the ensemble towards these methods.
+For the methods with multiple variations, the variant retained in the ensemble had superior performance than the alternatives based on preliminary results, or had already been considered for application in the region.
+The full range of methods included in this analysis were detailed below with equations (Table 2).
+Each method was applied to data that would lead to retrospective patterns in an age-based stock assessment and performance was evaluated using a range of metrics (see below).
 
-To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depend on temperature preferences.
-We can then consider the impact of climate change by simulating scenarios with repeating temperature patterns and those where temperature increases on average over time.
-In both cases we analyze the ability of stratified random sampling to track population trends.
+Each of the methods evaluated produces a single target catch value that was fixed over a two year interval.
+If the methods were being applied in year $y$, then target catches are set for years $y+1$ and $y+2$ (denoted $C_{targ, y+1:y+2}$).
+In practice, the timing of setting target catches in the region generally occurs in late summer or early fall in between the spring and fall surveys, and before complete catch data are available.
+Therefore, in year $y$ complete catch data are available through year $y-1$, and survey data are available for the spring survey through year $y$ and for the fall survey through year $y-1$.
+Applications of DLMs in this region have used an average of the spring index in year $y$ $(I_{spr,y})$ and the fall index in year $y-1$ $(I_{fall,y-1})$ to reflect average abundance at the start of year $y$ $(\bar{I}_{y})$.
+For this study, the same 1 year lag was implemented for methods that use the average of both simulated indices to generate catch advice:
 
-\section{Methods}
+$\bar{I}_{y} = \frac{I_{fall,y-1} + I_{spr,y}}{2}$.
 
-- Describe simulation study
+<font size="5">*Control Rules*</font>
 
-We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on George's Bank, where movement of each species combine static species-specific habitat preferences with dynamic temperature preferences.
-Model dynamics are driven by dynamic temperature gradients estimated from data to create simulated data sets for each population where the true biomass is known.
-Using temperature gradients that repeat each year creates data sets with predictable, repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time.
-We conducting stratified random sampling on our simulation output to mimic the bottom trawl survey and compare the ability of contemporary indexing methods to track population trends.
+Most DLMs do not have the ability to estimate a biomass reference point (e.g., $B_{MSY}$), which made consideration of so called biomass-based harvest control rules that reduce $F$ or catch in response to estimated changes in relative stock status impossible.
+Although reference points can be created for DLMs, they typically rely on local expert judgment [@harford2021harvest] and are geared towards either keeping the stock about where it is or else increasing it towards a relative amount that was thought to be good.
+Neither of these provide a proxy for maximum sustainable yield reference points, but might instead provide pretty good yield [@hilborn2010pgy].
 
-\subsection{Population Model Formulation}
+Lack of clarity exists, however, on whether the catch advice from DLMs should be used directly or reduced to account for uncertainty.
+In the U.S.
+management system, an overfishing limit is the catch that would result from applying $F_{MSY}$, whereas an acceptable biological catch is a catch reduced from the overfishing limit to account for scientific uncertainty.
+Each DLM was evaluated using two harvest control rules: 1) the catch advice from a given DLM was applied directly and assumed to serve as a proxy for the catch associated with $F_{MSY}$ (catch multiplier = 1), and 2) the catch advice from a given DLM was reduced by 25\% to account for unspecified scientific uncertainty (catch multiplier = 0.75).
+The case where catches were reduced by 25\% was intended to reflect a common default control rule in the region that uses 0.75$F_{MSY}$.
 
--- Used MixFishSim. Describe edits made to package
+<font size="5">*Application of a Statistical Catch-at-Age Assessment (SCAA)*</font>
 
-We use the R package *MixFishSim* (MFS) to model our populations [@dolder2020highly].
-MFS is a discrete spatiotemporal simulation tool where users can model multiple species under varying environmental conditions.
-The package uses a delay-difference population model with discrete processes for growth, death, and recruitment of the population.
-We formulate the following inputs for the MFS package to address our research question:
+A SCAA model was also applied to all scenarios to generate catch advice for comparison with the DLMs.
+Although virtual population analysis (VPA) is also used for some age-based assessments in the region, SCAA models are more widely used.
+Applications of the SCAA model assumed that the assessment had the correct underlying structure for selectivity, and CVs and ESS were specified at their true underlying values.
+The SCAA model estimated annual recruitment deviations assuming no underlying stock-recruit relationship, annual fully-selected fishing mortality rates, fishery and survey selectivity parameters (logistic), abundance-at-age in year one of the period being assessed, and survey catchabilies.
+Mohn’s rho was calculated (7 year peels) for abundance at age for all model fits during the feedback period and used to retro-adjust abundance at age for projections (divided by one plus Mohn’s rho; [@brooks2016retroforecast]).
+Catch advice was determined by specifying fully-selected $F = 0.75F_{40\%}$, always assuming M=0.2.
+All life history parameters were fixed at their correct value, except for the natural mortality rate when it was the source of the retrospective pattern.
 
-<font size="5">*Study Area*</font>
+<font size="5">*Study Design*</font>
 
-We obtained a shapefile for the 15 strata that comprise George's Bank to use as our modeling environment.
-We discritized the region into a raster with 88 rows and 144 columns.
-Haddock inhabit all 15 strata in the domain Cod inhabit 13 strata, and yellowtail exist in 9 strata.
-**include figure?**
+In addition to the two control rules applied for each DLM described above, three aspects of the OM were varied in a full factorial study design: fishing history, fishery selectivity, and cause of the retrospective pattern (Table 3).
+Two variants of fishing history were considered, with fully selected fishing mortality during the base period either constant at a level equal to $2.5F_{MSY}$ (always overfishing) or equaling $2.5F_{MSY}$ in the first half of the base period then a knife-edged decline to $F_{MSY}$ for the second half of the base period.
+These patterns in fishing mortality rate were based on observed patterns for Northeast groundfish [@wiedenman2019dlm].
+These two different fishing intensities during the latter half of the base period led to different starting conditions for the feedback period.
 
-<font size="5">*Population Dynamics and Recruitment*</font>
+Two variations of the OM were considered with either time invariant, asymptotic, fishery selectivity in the base and feedback periods, or a change in selectivity after the first half of the base period so that the age at 50\% selectivity increased from approximately 3.7 to 5 (Table 1).
+The asymptotic selectivity pattern was based on Northeast groundfish fishery selectivity patterns.
+The change in the selectivity pattern when selectivity varied through time approximated an increase in mesh size in the fishery to avoid younger fish.
 
-The time step for our models is one week.
-MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass of in each cell in our study area [@dolder2020highly].
-Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits.
-We chose to represent recruitment in the model using a Beverton-Holt formulation.
-Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.
+Two different sources of stock assessment misspecification leading to retrospective patterns were considered, temporal changes in natural mortality and misreported catch.
+The degree to which natural mortality and unreported catch changed through time was determined by attempting to achieve an average Mohn’s rho of approximately 0.5 for $SSB$ when an SCAA model (i.e., configured using WHAM) was used to fit the simulated data.
+We also fit the same SCAA configuration to data without misspecified M or catch to verify that retrospective patterns were not present on average (see Supplemental Materials Figure S1).
+A third source of misspecification was also attempted, time varying survey catchability, but this source of misspecification was unable to produce severe enough retrospective patterns and was abandoned.
 
---Many parameter inputs were obtained from literature
+For the natural mortality misspecification, the true natural mortality changed from 0.2 to 0.32 in scenarios where the fishing history was always overfishing or from 0.2 to 0.36 when the fishing history included a reduction from overfished to $F_{MSY}$, with the differences between fishing histories necessary to produce the desired retrospective pattern severity (see Supplemental Materials Figures S2 and S3).
+In each case, natural mortality trended linearly from 0.2 to the higher value between years 31 and 40 of the base period and held constant at the higher level for years 41-50.
+Natural mortality remained constant at the higher level throughout the feedback period.
+Those DLMs that required natural mortality as an input parameter used the value from before any change in natural mortality (0.2) because the change in natural mortality is meant to be unknown.
 
-Each species is assumed to have normally distributed temperature preferences ($~N(\mu,\sigma)$).
-We assume Yellowtail's preferences are $N(8.75,4.25)$, while Haddock and Cod have preferences $N(9,4)$.
-We chose these values by combining information in the literature with temperatures recorded in the bottom trawl survey.
-A full list of parameters used in our model can be seen below in Table ???.
+For catch misspecification, a scalar multiple of the true catch observation is provided as the observed catch to the DLMs.
+The scalar is 0.2 when fishing intensity was always overfishing and for both selectivity patterns, 0.44 when the fishing history included a reduction to $F_{MSY}$ and with time variant selectivity, or 0.40 when the fishing history included a reduction to $F_{MSY}$ and selectivity was time invariant.
+The shift in scalar trended linearly from 1 to the lower value between years 31 and 40 of the base period and remained at the lower value for years 41-50.
+These scalars were applied only to the aggregate catch so that they affect all catches at age equally.
+When catch misspecification was applied in conjunction with a DLM during the feedback period, the true catch in the OM equaled the catch advice provided by the DLM multiplied by the inverse of the scalar multipliers (i.e., the true catches were higher than the DLM catch advice).
+Thus, when the scalar multipliers were applied to the true catch from the OM in order to provide observed catches at the next application of the DLM, the observed catch equaled the catch advice from the previous application of the DLM, on average.
+In other words, managers and analysts would be given the perception that the DLM catch advice was being caught by the fishery, when in fact the true catches were always higher.
+This meant that the source of the retrospective pattern continued in the feedback period.
+The magnitude of the retrospective pattern in the feedback period varied due to the observation error applied in each realization (See Supplemental Materials Figure S4).
 
-**go through each process in model and discuss edits/removals?**
+Fourteen methods for setting catches were explored (13 DLMs and the SCAA) and were applied to all 16 scenarios, which created 224 factorial combinations in the study design.
+For each element of the full factorial combinations, 1,000 simulations were conducted.
+The simulations used the same random number seeds across all combinations in the study design resulting in the same patterns of recruitment deviations and observation errors.
+Two DLMs (AIM and ES-Fstable) had two failed simulations each, which were caused by relatively high catch advice (i.e., requiring relatively high $F$) that triggered errors in the Newton-Raphson iterations used to determine the $F$ that would produce the desired catch.
+This small number of failures was unlikely to effect results and conclusions, and so were not considered further.
 
-<font size="5">*Temperature and Habitat Input*</font>
+<font size="5">*Performance Metrics*</font>
 
-The package was designed to generate theoretical habitat preferences using Gaussian Random Fields that combine with hypothetical temperature gradients to simulate an entire fishery.
-Since we are modeling real species on the northeast continental shelf w
+Six metrics thought to be of broad interest were reported here, each calculated and reported separately for a short-term (i.e., first six years of the feedback period) and long-term (i.e., last 20 years of the feedback period) period.
+These metrics were selected to represent the tradeoffs in terms of benefits to the fishery and risks to the stock.
+The specific metrics reported were: $\frac{SSB}{SSB_{MSY}}$, $\frac{F}{F_{MSY}}$, catch relative to $MSY$, interannual variation in catch [@amar2010mse], number of years of overfishing ($F > F_{MSY}$), and number of years of the stock being overfished ($SSB < 0.5 SSB_{MSY}$).
 
-We obtain estimated temperature data for the region for 2012 from FVCOM **cite**.
+## Results
 
---Describe difference between increasing and constant temperature scenarios (images?)
+Overall performance varied widely across methods, and the individual performance of a method was sensitive to the different scenarios explored.
+Performance for each method was sensitive to the source of the retrospective pattern (missing catch or M), the exploitation history, when in the feedback period the metric was calculated (short- or long-term), and whether or not a 25% buffer was applied when setting the catch advice from a given method.
+Overall, similar results occurred for the scenarios with one or two selectivity blocks, so the impact of the selectivity scenarios was not discussed further.
 
--- Describe each scenario that is considered
+<font size="5">*Aggregate performance*</font>
 
-We consider three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time.
-Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees calcius over the duration of the 20 year simulation.
+In Figure \ref{fig:PMsallscenarios}, the inner quartiles and medians for all performance measures are shown, calculated across all scenarios combined.
+In general, methods that resulted in high mean $F / F_{MSY}$ (Figure \ref{fig:PMsallscenarios}B) resulted in lower stock biomass (Figure \ref{fig:PMsallscenarios}A), more years of overfishing (Figure \ref{fig:PMsallscenarios}E) and of being overfished (Figure \ref{fig:PMsallscenarios}F), and vice-versa.
+Higher $F$ values were also associated with higher catches (Figure \ref{fig:PMsallscenarios}C), on average, and a greater variability in catch, but there were some methods that produced lower $F$ values that also resulted in high catch variability (CC-FM, CC-FSPR; Figure \ref{fig:PMsallscenarios}D).
 
-<font size="5">*Simulating Bottom Trawl Survey and Population Indexing*</font>
+A number of methods performed poorly overall, resulting in high exploitation rates and low stock size, on average (Figure \ref{fig:PMsallscenarios}).
+These methods include AIM, three of the four expanded survey biomass methods (ES-FM, ES-FSPR, and ES-Fstable), and the Skate method.
+The Itarget and ensemble methods also resulted in $SSB < SSM_{MSY}$ and $F > F_{MSY}$, on average, though departures from the MSY levels were not as severe as the other methods (Figure \ref{fig:PMsallscenarios}).
+The remaining methods (CC-FM, CC-FSPR, DynLin, ES-Frecent, Islope, Ismooth, and SCAA) were able to limit overfishing and keep biomass above $SSB_{MSY}$, on average, although for four of these methods (CC-FM, CC-FSPR, DynLin, and Ismooth) biomass was more than 50% higher than $SSB_{MSY}$ (Figure \ref{fig:PMsallscenarios}).
+Principal components analysis of the median values for all methods and metrics resulted in groupings similar to those noted above (see Supplemental Materials Figure S5).
 
--Describe post hoc sampling process and how data is used
+<font size="5">*Scenario-dependent performance*</font>
 
-After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year.
-The number of the samples taken in each strata reflect true values and we sample from the same weeks that the Spring and Fall surveys take place.
-Most strata contain enough cells to sample a unique location in each survey.
-For smaller strata we are forced to repeat locations.
-We then use the samples in contemporary population indexing methods to track population trends.
-Since we know the true population values in our simulations we can compare methods by comparing the error calculated from each estimate.
+The source of the retrospective pattern had a large impact on results for a given method.
+The relationship between $SSB/SSB_{MSY}$ and $C/MSY$ is shown across scenarios for the different sources of retrospective error.
+Stock size and catch (relative to MSY levels) are clustered for many of the methods with no overlap between $M$ and unreported catch sources (AIM, ES-FM, ES-FSPR, ES-Fstable, Itarget, Skate, Ensemble, and SCAA).
+For all of these methods, $SSB/SSB_{MSY}$ was lower when unreported catch was the source of the retrospective pattern, and $C/MSY$ was also lower except for the Itarget and the SCAA methods compared to the scenarios when increased natural mortality was the source of the retrospective pattern (Figure \ref{fig:SSmsyvsCMSY}).
+The source of the retrospective pattern also had a large impact on the other performance measures (Figure \ref{fig:PMsbyretroF}).
+In general, when unreported catch was the source of the retrospective pattern, interannual variability in catch was higher, overfishing was more frequent and with a larger $F/F_{MSY}$, and the stock had a higher risk of being overfished compared to the scenarios when increased natural mortality was the source of the retrospective pattern (Figure \ref{fig:PMsbyretroF}).
+Six methods (AIM, ES-FM, ES-FSPR, ES-Fstable, Itarget, Skate, Ensemble) resulted in overfishing in nearly every year of the feedback period (often with very high $F/F_{MSY}$) when missing catch was the source of the retrospective pattern (Figure \ref{fig:PMsbyretroF}B, \ref{fig:PMsbyretroF}E).
+In contrast, all methods except Skate, AIM, and ES-Fstable had low $F/F_{MSY}$, high $SSB/SSB_{MSY}$, and few years of being overfished when increased natural mortality was the source of the retrospective pattern (Figure \ref{fig:PMsbyretroF}B, \ref{fig:PMsbyretroF}A, \ref{fig:PMsbyretroF}F).
+The $C/MSY$ when increased natural mortality was the source of the retrospective pattern varied widely with some DLMs well below 1.0 and others well above (Figure \ref{fig:PMsbyretroF}C).
+The SCAA method also resulted in frequent overfishing in the missing catch scenario, but less so when the stock was more depleted at the start of the feedback period (Figure \ref{fig:PMsbyretroF}F).
 
---Stratified mean vs VAST with and without covariates
+Exploitation history also impacted the performance of many of the other methods.
+For four methods (Islope, Ismooth, DynLin and ES-Frecent), exploitation rates were higher when the stock experienced overfishing for the entire base period, but the impact was more dramatic in the short-term.
+Over time as these methods were used, $F$ declined and remained below  $F_{MSY}$ in the long-term (Figure \ref{fig:FFmsybyscenario}A), allowing stock recovery.
+The majority of the other methods also resulted in greater exploitation rates in the short-term, though some methods kept $F/F_{MSY} < 1$ regardless of the time-period (CC-FM, CC-FSPR, and SCAA), while others (AIM, ES-Fstable, Skate, Ensemble) kept $F/F_{MSY} > 1$ over the short- and long-term (Figure \ref{fig:FFmsybyscenario}A).
+For the ES-FM and ES-FSPR methods, there was not a consistent pattern in exploitation rates when comparing the short- and long-term periods (Figure \ref{fig:FFmsybyscenario}A).
 
-We compare the stratified mean estimate of abundance to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
-The stratified mean is a typical survey-based approach that scales individual samples to the  strata-level by considering the area of each strata, before scaling to the region-level based on the relative size of each strata.
-VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurance (presence/absence).
-If desired, VAST also allows users to include covariate data to better inform the model.
+As expected, application of a buffer to the catch advice resulted in lower exploitation rates compared to no buffer across all methods, but the magnitude of the impact differed by method (Figure \ref{fig:FFmsybyscenario}B).
+For poor-performing methods where $F/F_{MSY} >>1$, the use of a buffer tended to result in greater reductions in $F$ than other methods.
+Methods like AIM, ES-FM, ES-FSPR, ES-Fstable and Skate all had large reductions in $F$ when the buffer was applied, but the reduction was insufficient to reduce $F/F_{MSY} < 1$ (Figure \ref{fig:FFmsybyscenario}B).
+For some methods (CC-FM, CC-FSPR, SCAA), the median $F/F_{MSY}$ was always below 1 with or without the buffer, whereas for other methods (DynLin, ES-Frecent, Islope, Ismooth, Itarget, and Ensemble) there were instances where using a buffer pushed $F/F_{MSY}$ below 1 (though it depended on the exploitation history; Figure \ref{fig:FFmsybyscenario}B).
+
+The median and interquartile range performance measures reported thus far do not express the full range of results across individual runs, however.
+When all the simulations are plotted, there is clearly a wide range of possible outcomes for the population, indicating that performance for a particular series of environmental conditions, expressed through recruitment deviations, can vary widely.
+For example, Figure \ref{fig:linearvsdiffuse} shows the long-term average $SSB/SSB_{MSY}$ and $C/MSY$ relationship across runs for a single scenario.
+Different patterns in the relationship between the SSB and catch ratios resulted, with methods falling into two groups.
+In the first group, there is a near linear relationship between $SSB/SSB_{MSY}$ and $C/MSY$ (AIM, ES-Fstable, ES-FSPR, ES-FM, Itarget, Skate, Ensemble, and SCAA; Figure \ref{fig:linearvsdiffuse}).
+In the second group (CC-FSPR, CC-FM, DynLin, ES-Frecent, Ismooth, and Islope) the relationship is more diffuse, with a wide range of $C/MSY$ for a given $SSB/SSB_{MSY}$.
+The linear or diffuse relationships persisted across scenarios, although the upper limit of $C/MSY$ was greatly reduced for the diffuse methods when the buffer was applied to the catch advice.
+(See Supplemental Figures S6-S21 for these plots across all 16 scenarios and Figures S22-S37 for similar plots showing $F/F_{MSY}$ versus $SSB/SSB_{MSY}$).
 
@@
 
-\pagebreak
-Table 1. Parameters used in all population models.
-```{r paramsALL, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-tabl <- "
-|		|	Description | Unit |Yellowtail | Cod | Haddock | Source
-|:---	|:------- |:-- |:--- |:-- |:-- |:--
-| $\\rho$ | Ford's growth coefficient | wk$^{-1}$  | 4.48 | 4.43 | 4.49 | |
-| $M$ | Natural Mortality| wk$^{-1}$  | 0.2064 | 0.2728 | 0.3340 | |
-| $W_R$ |Weight of fully recruited fish| kg | 0.39 | 2.95 | 1.12 | |
-| $W_{R-1}$ | Weight of pre-recruit fish| kg | 0.13 | 0.39 | 0.19 | |
-"
-cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
-```
-
-Table 2. Parameters used in models with relatively constant populations.
-```{r paramsCON, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-tabl <- "
-|		|	Description | Yellowtail | Cod | Haddock 
-|:---	|:------- |:--- |:-- |:-- |:--
-| $M+F$ | Adjusted Mortality (Natural + Fishing)| wk$^{-1}$  | 0.764 | 0.83 | 0.309 |
-
-"
-cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
-```
-
-Table 3. Parameters used in models with increasing populations.
-```{r paramsINC, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-tabl <- "
-|		|	Description | Yellowtail | Cod | Haddock | Source
-|:---	|:------- |:--- |:-- |:-- |:--
-| $M+F$ | Adjusted Mortality (Natural + Fishing)| wk$^{-1}$  | 0.564 | 0.372 |  0.134 |
-
-"
-cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
-```
-
-Table 4. Parameters used in models with decreasing populations.
-```{r paramsDEC, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-tabl <- "
-|		|	Description | Yellowtail | Cod | Haddock | Source
-|:---	|:------- |:--- |:-- |:-- |:--
-| $M+F$ | Adjusted Mortality (Natural + Fishing)| wk$^{-1}$  | 0.764 | 0.623 | 0.334 |
-
-"
-cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
-```
-
 Table 1. Maturity-, weight-, and selectivity-at-age of the simulated fish population.

### 2022-12-20 TestScripts/GB_habitat_create.R

@@ -1406,8 +1406,9 @@ strata_species[["Had"]] <- c(13,14,15,16,17,18,19,20,21,22,23,24,25,29,30)
 
 
 #load results from a given simulation
-scenario <- "DecPop_IncTemp"
+scenario <- "IncPop_ConTemp"
 
+######################################################################################
 #choose which simulation iteration to use (chosen in different file)
 good_iter <- c(1,13,6) #ConPop_ConTemp
 good_iter <- c(1,1,3) #ConPop_IncTemp
@@ -1415,6 +1416,57 @@ good_iter <- c(77,63,98) #IncPop_ConTemp
 good_iter <- c(77,44,100) #IncPop_IncTemp
 good_iter <- c(25,18,6) #DecPop_ConTemp
 good_iter <- c(13,44,9) #DecPop_IncTemp
+######################################################################################
+
+######################################################################################
+# #THESE FOR CONPOP_CONTEMP
+color_max[["YT"]][[4]] <- 205  #spring survey pop max
+color_max[["YT"]][[10]] <- 80 #fall survey
+color_max[["Cod"]][[4]] <- 170  #spring survey pop max
+color_max[["Cod"]][[10]] <- 335 #fall survey
+color_max[["Had"]][[4]] <- 1525  #spring survey pop max
+color_max[["Had"]][[10]] <-1273  #fall survey
+
+#THESE FOR CONPOP_INCTEMP
+color_max[["YT"]][[4]] <- 125  #spring survey pop max
+color_max[["YT"]][[10]] <- 200 #fall survey
+color_max[["Cod"]][[4]] <- 437  #spring survey pop max
+color_max[["Cod"]][[10]] <- 3044 #fall survey
+color_max[["Had"]][[4]] <- 3945  #spring survey pop max
+color_max[["Had"]][[10]] <-3775  #fall survey
+
+#THESE FOR INCPOP_CONTEMP
+color_max[["YT"]][[4]] <- 405  #spring survey pop max
+color_max[["YT"]][[10]] <- 343 #fall survey 
+color_max[["Cod"]][[4]] <- 637  #spring survey pop max
+color_max[["Cod"]][[10]] <- 1550 #fall survey 
+color_max[["Had"]][[4]] <- 2645  #spring survey pop max
[…]

### 2022-12-21 Manuscript/Draft1.Rmd
update manuscript

@@
   bibliography: IBMWG.bib
+
   word_document: default
@@
 - \usepackage[belowskip=0pt,aboveskip=0pt]{caption}
+ \usepackage{array}
+ \usepackage{caption}
+ \usepackage{graphicx}
+ \usepackage{siunitx}
+ \usepackage{colortbl}
+ \usepackage{multirow}
+ \usepackage{hhline}
+ \usepackage{calc}
+ \usepackage{tabularx}
+ \usepackage{tabulary}
+ \usepackage{threeparttable}
+ \usepackage{wrapfig}
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE)
 if(!require(kableExtra)) {  
@@
 ```{r}
+
 knitr::kable(YT_results, caption = "Yellowtail error results")
@@
 Federal biologists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (cite survey paper).
-The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf **(include image from johnson_sosebee?)**.
+The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf.
 The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits **just listed a few uses of survey change/others?**.
@@
 Haddock inhabit all 15 strata in the domain Cod inhabit 13 strata, and yellowtail exist in 9 strata.
-**include figure?**
+Figure \@ref(fig:strata-plot) shows the regions used in our models.
+
+```{r strata-plot, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Strata inhabited by each species in our population models."}
+knitr::include_graphics(c("Images/Strata.png"))
+```
 
@@
 Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.
+Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics.
+A full list of parameters used in our model can be seen below in Tables \@ref(paramsALL) and \@ref(tab:paramsSCENARIOS).
+**missing spawning weeks, recruitment weeks, lambda**
 
---Many parameter inputs were obtained from literature
+<font size="5">*Temperature and Habitat Input*</font>
+
+The package was designed to generate theoretical habitat preferences using Gaussian Random Fields that combine with hypothetical temperature gradients to drive the probability of movement from cell $I$ to cell $J$ using the formulation
+
+\begin{align*}
+Pr(C_{wk+1}=J|C_{wk}=I) = \frac{e^{-\lambda \cdot d_{I,J}}\cdot(Hab^2_{J,p} \cdot Tol_{J,p,wk})}{\sum^C_{c=1}e^{-\lambda \cdot d} \cdot (Hab^2_{c,p} \cdot Tol_{c,p,wk})},
+\end{align*}
+
+ where
+
+   $e^{-\lambda \cdot d_{I,J}}$ accounts for distance between cells $I$ and $J$,
+
+  $Hab^2_{J,p}$ is the static habitat value for species $p$ in cell $J$, and
+
+   $Tol_{c,p,wk}$ is the value from normally distributed temperature tolerance for species $p$ in cell $c$ in week $wk$.
+
+Since we are modeling real species on the northeast continental shelf, we formulate the habitat and temperature components as follows.
 
@@
 We chose these values by combining information in the literature with temperatures recorded in the bottom trawl survey.
-A full list of parameters used in our model can be seen below in Table ???.
+We obtain estimated temperature data for the region for 2012 from FVCOM **cite**.
+We chose 2012 because the data displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values.
+We transform the temperature data to create an oscillating  pattern that increases 5 degrees Celsius over the duration of the simulation.
+**show image of average temperature oscilations?**
 
-**go through each process in model and discuss edits/removals?**
-
-<font size="5">*Temperature and Habitat Input*</font>
-
-The package was designed to generate theoretical habitat preferences using Gaussian Random Fields that combine with hypothetical temperature gradients to simulate an entire fishery.
-Since we are modeling real species on the northeast continental shelf w
-
-We obtain estimated temperature data for the region for 2012 from FVCOM **cite**.
+Species-specific habitat preferences were derived by relating bottom trawl data to covariate predictors to create a niche model for each species.
 
@@
 We consider three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time.
-Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees calcius over the duration of the 20 year simulation.
+Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius over the duration of the 20 year simulation.
 
@@
 After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year.
-The number of the samples taken in each strata reflect true values and we sample from the same weeks that the Spring and Fall surveys take place.
-Most strata contain enough cells to sample a unique location in each survey.
-For smaller strata we are forced to repeat locations.
-We then use the samples in contemporary population indexing methods to track population trends.
-Since we know the true population values in our simulations we can compare methods by comparing the error calculated from each estimate.
+We sample in the same weeks that the Spring and Fall surveys take place and the number of the samples taken in each strata reflect true values.
+Most strata contain enough cells to sample a unique location in each survey over the duration of the simulation.
+For smaller strata we must repeat some sample locations.
+We then use the biomass collected from our samples in contemporary population indexing methods to estimate population trends.
+Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.
 
@@
 The stratified mean is a typical survey-based approach that scales individual samples to the  strata-level by considering the area of each strata, before scaling to the region-level based on the relative size of each strata.
-VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurance (presence/absence).
+VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurrence (presence/absence).
 If desired, VAST also allows users to include covariate data to better inform the model.
@@
 | $W_{R-1}$ | Weight of pre-recruit fish| kg | 0.13 | 0.39 | 0.19 | |
+| $\\sigma^2$ | Variance in recruited fish | kg$^2$ | 0.55 | 0.55 | 0.55 ||
 "
@@
 
-Table 2. Parameters used in models with relatively constant populations.
-```{r paramsCON, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-tabl <- "
-|		|	Description | Yellowtail | Cod | Haddock 
-|:---	|:------- |:--- |:-- |:-- |:--
-| $M+F$ | Adjusted Mortality (Natural + Fishing)| wk$^{-1}$  | 0.764 | 0.83 | 0.309 |
+<!-- Table 2. Parameters used in models with relatively constant populations. -->
+<!-- ```{r paramsCON, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'} -->
+<!-- tabl <- " -->
+<!-- |		|	Description | Yellowtail | Cod | Haddock  -->
+<!-- |:---	|:------- |:--- |:-- |:-- |:-- -->
+<!-- | $M+F$ | Adjusted Mortality (Natural + Fishing)| wk$^{-1}$  | 0.764 | 0.83 | 0.309 |  -->
 
-"
-cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
+<!-- " -->
+<!-- cat(tabl) # output the table in a format good for HTML/PDF/docx conversion -->
+<!-- ``` -->
+
+<!-- Table 3. Parameters used in models with increasing populations. -->
+<!-- ```{r paramsINC, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'} -->
+<!-- tabl <- " -->
+<!-- |		|	Description | Yellowtail | Cod | Haddock | Source -->
+<!-- |:---	|:------- |:--- |:-- |:-- |:-- -->
+<!-- | $M+F$ | Adjusted Mortality (Natural + Fishing)| wk$^{-1}$  | 0.564 | 0.372 |  0.134 |  -->
+
+<!-- " -->
+<!-- cat(tabl) # output the table in a format good for HTML/PDF/docx conversion -->
+<!-- ``` -->
+
+<!-- Table 4. Parameters used in models with decreasing populations. -->
+<!-- ```{r paramsDEC, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'} -->
+<!-- tabl <- " -->
+<!-- |		|	Description | Yellowtail | Cod | Haddock | Source -->
+<!-- |:---	|:------- |:--- |:-- |:-- |:-- -->
+<!-- | $M+F$ | Adjusted Mortality (Natural + Fishing)| wk$^{-1}$  | 0.764 | 0.623 | 0.334 |  -->
+
+<!-- " -->
+<!-- cat(tabl) # output the table in a format good for HTML/PDF/docx conversion -->
+<!-- ``` -->
+
+<!-- DOESNT WORK!!
+<!-- Table 2. Parameters used in population models for each scenario. -->
+<!-- ```{r paramsSCENARIOS, echo=FALSE, message=TRUE, warnings=TRUE, results='asis'} -->
+<!-- library(huxtable) -->
+
+<!-- ht <- tribble_hux( -->
+<!--   ~ "Section/topic", ~ "Item no", ~ "Checklist item", ~ "Reported on Page No", -->
+<!--   "Title"          , ""         , ""                , "", -->
+<!--   "Title"          , "1"        , "Identify..."     , "",  -->
+<!--   "Abstract"       , ""         , ""                , "", -->
+<!--   "Structured summary", "2"     , "Provide..."      , "" -->
+<!--   # et cetera... -->
+<!-- )  -->
+
+<!-- # using the pipe from 4.1.0... -->
+
+<!-- ht |>  -->
+<!--   set_header_rows(c(2, 4), TRUE)                |> -->
+<!--   merge_across(c(2, 4), everywhere)             |> -->
+<!--   style_header_rows(bold = TRUE)                |> -->
+<!--   set_all_borders(brdr(0.4, "solid", "grey70")) |> -->
+<!--   set_background_color("grey97")                |> -->
+<!--   set_background_color(1, 1:3, "grey90")        |> -->
+<!--   set_col_width(c(0.2, 0.05, 0.55, 0.2))        |> -->
+<!--   set_font("cmss")                               -->
+<!-- ``` -->
+
+<!-- ```{r,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'} -->
+<!-- library("knitr") -->
+<!-- library("kableExtra") -->
+<!-- library("dplyr") -->
+<!-- library("readr") -->
+<!-- library("tidyr") -->
+<!-- library("forcats") -->
+
+<!-- #orig_ui_deaths = read_csv("https://www.opendata.nhs.scot/dataset/b0135993-3d8a-4f3b-afcf-e01f4d52137c/resource/89807e07-fc5f-4b5e-a077-e4cf59491139/download/ui_deaths_2020.csv") -->
+
+<!-- summary_tab <- orig_ui_deaths %>% -->
+<!--   # Apply filters in same was as in plots so looking at one year, whole of Scotland, and exlucing "All" entries -->
+<!--   filter(Year == "2018", HBR == "S92000003", AgeGroup != "All" & Sex != "All", InjuryType != "Accidental exposure" & InjuryType != "All") %>% -->
+<!-- # Group the table according to injury type, age, and sex -->
+<!--   group_by(InjuryType, AgeGroup, Sex) %>% -->
+<!-- # Create a summary of total number of deaths for neater appearance to the table and demonstration of the figures -->
+<!--   summarise(total_deaths = sum(NumberofDeaths)) %>% -->
+<!-- # Change the orientation of the table so that age groups become the variable headings -->
+<!--   pivot_wider(names_from = AgeGroup, values_from = total_deaths) %>% -->
+<!--   mutate(Sex = factor(Sex)) %>% -->
+<!--   arrange(Sex) %>% -->
+<!--   select(Sex, everything()) -->
+
+<!-- # Create table to present the data] -->
+<!-- Fig_2 <- kable(summary_tab[, -1],  -->
+<!--   caption = "Deaths from unintentional injuries in Scotland", -->
+<!--     format = "latex", booktabs = TRUE) %>% -->
+<!--   kable_styling(font_size = 10) %>% -->
+<!--   pack_rows(tab_kable, colnum = 1, -->
+<!--     index = table(fct_inorder(summary_tab$Sex), useNA = "no")) -->
+<!-- ``` -->
+<!-- ```{r, results='asis'} -->
+<!-- Fig_2 -->
+<!-- ``` -->
+
+```{r paramsSCENARIOS,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
+library("knitr")
+library("kableExtra")
+library("dplyr")
+library("readr")
+library("tidyr")
+library("forcats")
+
+param_table = read_csv("Parameters.csv")
+
+summary_tab <- param_table %>% 
+  dplyr::mutate_if(is.numeric, funs(as.character(signif(., 3))))%>%
+
+ mutate_if(grepl('alpha',.), ~replace(., grepl('alpha', .), paste("\U03B1"))) %>% #should be replacing alpha with the symbol
+  mutate_if(grepl('beta',.), ~replace(., grepl('beta', .), paste("\U03B2"))) %>% #should be replacing beta with the symbol
+    mutate_if(grepl('wk',.), ~replace(., grepl('wk', .), "1/wk")) %>% #should be replacing beta with the symbol
+
+# Group the table according to Parameter, Scenario
+  group_by(Parameter, Scenario)%>%
+# Change the orientation of the table so that Species become the variable headings
+  pivot_wider(names_from = Species,  values_from = Value) %>%
+  mutate(Scenario = factor(Scenario)) %>%
+  arrange(Scenario) %>%
+  select(Scenario, everything())
+
+# Create table to present the data]
+Fig_3 <- kable(summary_tab[, -1], 
+  caption = "Parameters used in population models for each scenario.",
+    format = "latex", booktabs = TRUE) %>%
+  kable_styling(font_size = 10) %>%
+  pack_rows(tab_kable, colnum = 1,
+    index = table(fct_inorder(summary_tab$Scenario), useNA = "no"))
 ```
-
-Table 3. Parameters used in models with increasing populations.
-```{r paramsINC, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-tabl <- "
-|		|	Description | Yellowtail | Cod | Haddock | Source
-|:---	|:------- |:--- |:-- |:-- |:--
-| $M+F$ | Adjusted Mortality (Natural + Fishing)| wk$^{-1}$  | 0.564 | 0.372 |  0.134 |
-
-"
-cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
-```
-
-Table 4. Parameters used in models with decreasing populations.
-```{r paramsDEC, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-tabl <- "
-|		|	Description | Yellowtail | Cod | Haddock | Source
-|:---	|:------- |:--- |:-- |:-- |:--
-| $M+F$ | Adjusted Mortality (Natural + Fishing)| wk$^{-1}$  | 0.764 | 0.623 | 0.334 |
-
-"
-cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
+```{r, results='asis'}
+Fig_3
 ```

### 2022-12-21 Manuscript/Draft1.tex
update manuscript

@@
 \usepackage[margin=1in]{geometry}
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \usepackage{longtable,booktabs,array}
@@
 \linenumbers
-\usepackage[belowskip=0pt,aboveskip=0pt]{caption}
+\usepackage[belowskip=0pt,aboveskip=0pt]{caption} \usepackage{array} \usepackage{caption} \usepackage{graphicx} \usepackage{siunitx} \usepackage{colortbl} \usepackage{multirow} \usepackage{hhline} \usepackage{calc} \usepackage{tabularx} \usepackage{tabulary} \usepackage{threeparttable} \usepackage{wrapfig}
 \usepackage{booktabs}
@@
 \newpage
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{knitr}\SpecialCharTok{::}\FunctionTok{kable}\NormalTok{(YT\_results, }\AttributeTok{caption =} \StringTok{"Yellowtail error results"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{knitr}\SpecialCharTok{::}\FunctionTok{kable}\NormalTok{(Cod\_results, }\AttributeTok{caption =} \StringTok{"Cod error results"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
 \begin{table}
@@
 \end{table}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{knitr}\SpecialCharTok{::}\FunctionTok{kable}\NormalTok{(Had\_results, }\AttributeTok{caption =} \StringTok{"Haddock error results"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
 
@@
 Federal biologists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (cite survey paper).
-The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf \textbf{(include image from johnson\_sosebee?)}.
+The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf.
 The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits \textbf{just listed a few uses of survey change/others?}.
@@
 Haddock inhabit all 15 strata in the domain Cod inhabit 13 strata, and yellowtail exist in 9 strata.
-\textbf{include figure?}
+Figure \ref{fig:strata-plot} shows the regions used in our models.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{Images/Strata}
+
+}
+
+\caption{Strata inhabited by each species in our population models.}\label{fig:strata-plot}
+\end{figure}
 
@@
 Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.
+Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics.
+A full list of parameters used in our model can be seen below in Tables \ref{paramsALL} and \ref{tab:paramsSCENARIOS}.
+\textbf{missing spawning weeks, recruitment weeks, lambda}
 
---Many parameter inputs were obtained from literature
+\emph{Temperature and Habitat Input}
+
+The package was designed to generate theoretical habitat preferences using Gaussian Random Fields that combine with hypothetical temperature gradients to drive the probability of movement from cell \(I\) to cell \(J\) using the formulation
+
+\begin{align*}
+Pr(C_{wk+1}=J|C_{wk}=I) = \frac{e^{-\lambda \cdot d_{I,J}}\cdot(Hab^2_{J,p} \cdot Tol_{J,p,wk})}{\sum^C_{c=1}e^{-\lambda \cdot d} \cdot (Hab^2_{c,p} \cdot Tol_{c,p,wk})},
+\end{align*}
+
+where
+
+\(e^{-\lambda \cdot d_{I,J}}\) accounts for distance between cells \(I\) and \(J\),
+
+\(Hab^2_{J,p}\) is the static habitat value for species \(p\) in cell \(J\), and
+
+\(Tol_{c,p,wk}\) is the value from normally distributed temperature tolerance for species \(p\) in cell \(c\) in week \(wk\).
+
+Since we are modeling real species on the northeast continental shelf, we formulate the habitat and temperature components as follows.
 
@@
 We chose these values by combining information in the literature with temperatures recorded in the bottom trawl survey.
-A full list of parameters used in our model can be seen below in Table ???.
+We obtain estimated temperature data for the region for 2012 from FVCOM \textbf{cite}.
+We chose 2012 because the data displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values.
+We transform the temperature data to create an oscillating pattern that increases 5 degrees Celsius over the duration of the simulation.
+\textbf{show image of average temperature oscilations?}
 
-\textbf{go through each process in model and discuss edits/removals?}
-
-\emph{Temperature and Habitat Input}
-
-The package was designed to generate theoretical habitat preferences using Gaussian Random Fields that combine with hypothetical temperature gradients to simulate an entire fishery.
-Since we are modeling real species on the northeast continental shelf w
-
-We obtain estimated temperature data for the region for 2012 from FVCOM \textbf{cite}.
+Species-specific habitat preferences were derived by relating bottom trawl data to covariate predictors to create a niche model for each species.
 
@@
 The stratified mean is a typical survey-based approach that scales individual samples to the strata-level by considering the area of each strata, before scaling to the region-level based on the relative size of each strata.
-VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurance (presence/absence).
+VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurrence (presence/absence).
 If desired, VAST also allows users to include covariate data to better inform the model.
@@
 \(W_{R-1}\) & Weight of pre-recruit fish & kg & 0.13 & 0.39 & 0.19 & \\
+\(\sigma^2\) & Variance in recruited fish & kg\(^2\) & 0.55 & 0.55 & 0.55 & \\
 \bottomrule
@@
 
-Table 2. Parameters used in models with relatively constant populations.
+\begin{verbatim}
+## Warning: `funs()` was deprecated in dplyr 0.8.0.
+## Please use a list of either functions or lambdas: 
+## 
+##   # Simple named list: 
+##   list(mean = mean, median = median)
+## 
+##   # Auto named with `tibble::lst()`: 
+##   tibble::lst(mean, median)
+## 
+##   # Using lambdas
+##   list(~ mean(., trim = .2), ~ median(., na.rm = TRUE))
+## This warning is displayed once every 8 hours.
+## Call `lifecycle::last_lifecycle_warnings()` to see where this warning was generated.
+\end{verbatim}
 
-\begin{longtable}[]{@{}llllll@{}}
+\begin{table}
+
+\caption{\label{tab:paramsSCENARIOS}Parameters used in population models for each scenario.}
+\centering
+\fontsize{10}{12}\selectfont
+\begin{tabular}[t]{llllll}
 \toprule
-& Description & Yellowtail & Cod & Haddock & \\
+Parameter & Description & Unit & Yellowtail & Cod & Haddock\\
 \midrule
-\endhead
-\(M+F\) & Adjusted Mortality (Natural + Fishing) & wk\(^{-1}\) & 0.764 & 0.83 & 0.309 \\
+\addlinespace[0.3em]
+\multicolumn{1}{l}{\textbf{Constant Population}}\\
+\hspace{1em}M+F & Adjusted Mortality (Natural + Fishing) & 1/wk & 0.764 & 0.83 & 0.309\\
+\hspace{1em}P0 & Initial Biomass & kg & 3190 & 21500 & \vphantom{1} 180000\\
+\hspace{1em}a & Max recruitment rate & kg & 30400 & 27900 & 73600\\
+\hspace{1em}ß & Recruitment half saturation value & kg & 4300 & 10500 & 40500\\
+\addlinespace[0.3em]
+\multicolumn{1}{l}{\textbf{Decreasing Population}}\\
+\hspace{1em}M+F & Adjusted Mortality (Natural + Fishing) & 1/wk & 0.764 & 0.623 & 0.334\\
+\hspace{1em}P0 & Initial Biomass & kg & 50000 & 21500 & 180000\\
+\hspace{1em}a & Max recruitment rate & kg & 1.07e+12 & 3.89e+08 & 4.97e+08\\
+\hspace{1em}ß & Recruitment half saturation value & kg & 2.3e+12 & 9.8e+08 & 2.08e+09\\
+\addlinespace[0.3em]
+\multicolumn{1}{l}{\textbf{Increasing Population}}\\
+\hspace{1em}M+F & Adjusted Mortality (Natural + Fishing) & 1/wk & 0.564 & 0.372 & 0.134\\
+\hspace{1em}P0 & Initial Biomass & kg & 3190 & 21500 & 180000\\
+\hspace{1em}a & Max recruitment rate & kg & 40000 & 45000 & 1e+05\\
+\hspace{1em}ß & Recruitment half saturation value & kg & 43000 & 62800 & 405000\\
 \bottomrule
-\end{longtable}
-
-Table 3. Parameters used in models with increasing populations.
-
-\begin{longtable}[]{@{}llllll@{}}
-\toprule
-& Description & Yellowtail & Cod & Haddock & Source \\
-\midrule
-\endhead
-\(M+F\) & Adjusted Mortality (Natural + Fishing) & wk\(^{-1}\) & 0.564 & 0.372 & 0.134 \\
-\bottomrule
-\end{longtable}
-
-Table 4. Parameters used in models with decreasing populations.
-
-\begin{longtable}[]{@{}llllll@{}}
-\toprule
-& Description & Yellowtail & Cod & Haddock & Source \\
-\midrule
-\endhead
-\(M+F\) & Adjusted Mortality (Natural + Fishing) & wk\(^{-1}\) & 0.764 & 0.623 & 0.334 \\
-\bottomrule
-\end{longtable}
+\end{tabular}
+\end{table}
 

### 2022-12-21 TestScripts/GB_habitat_create.R

@@ -152,7 +152,7 @@ GB_strata_idx <- match(GB_Had_strata_num,strata.areas@data[["STRATUMA"]])
 #plot(strata.areas[GB_strata_idx,])
 #define GB strata as own object
 GB_had_strata <- strata.areas[GB_strata_idx,]
-plot(GB_had_strata)
+plot(GB_had_strata,main='Haddock Strata')
 
 #CREATE COD STRATA
 #define georges bank
@@ -163,7 +163,7 @@ GB_strata_idx <- match(GB_Cod_strata_num,strata.areas@data[["STRATUMA"]])
 #plot(strata.areas[GB_strata_idx,])
 #define GB strata as own object
 GB_cod_strata <- strata.areas[GB_strata_idx,]
-plot(GB_cod_strata)
+plot(GB_cod_strata,main='Atlantic Cod Strata')
 
 #CREATE YELLOWTAIL STRATA
 #define georges bank
@@ -174,8 +174,7 @@ GB_strata_idx <- match(GB_Yel_strata_num,strata.areas@data[["STRATUMA"]])
 #plot(strata.areas[GB_strata_idx,])
 #define GB strata as own object
 GB_yell_strata <- strata.areas[GB_strata_idx,]
-plot(GB_yell_strata)
-
+plot(GB_yell_strata,main='Yellowtail Flounder Strata')
 
 
 #ADD "STRATA" list to hab which is used to determine how many total strata there are in Bens_init_survey

### 2022-12-21 Manuscript/Draft1.Rmd
updating manuscript

@@
 keyword 1, keyword 2
-
-\newpage
-
-```{r}
-
-knitr::kable(YT_results, caption = "Yellowtail error results")
-
-knitr::kable(Cod_results, caption = "Cod error results")
-
-knitr::kable(Had_results, caption = "Haddock error results")
-```
 
@@
 
-<font size="5">*Temperature and Habitat Input*</font>
+<font size="5">*Movement*</font>
 
@@
 
-\begin{align*}
+\begin{align}
 Pr(C_{wk+1}=J|C_{wk}=I) = \frac{e^{-\lambda \cdot d_{I,J}}\cdot(Hab^2_{J,p} \cdot Tol_{J,p,wk})}{\sum^C_{c=1}e^{-\lambda \cdot d} \cdot (Hab^2_{c,p} \cdot Tol_{c,p,wk})},
-\end{align*}
+\label{moveP}
+\end{align}
 
@@
 
-Each species is assumed to have normally distributed temperature preferences ($~N(\mu,\sigma)$).
-We assume Yellowtail's preferences are $N(8.75,4.25)$, while Haddock and Cod have preferences $N(9,4)$.
+<font size="5">*Habitat Input*</font>
+
+Species-specific habitat preferences were derived using the *lrren* tool from the R package *envi* **cite** to create a niche model for each species.
+The *lrren* tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors.
+We used bottom trawl point data from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow **cite trawl data?**.
+Depth and mean sediment size were used as our covariate predictors **cite**.
+Since the values in $Hab^2_{J,p}$ are required to be between 0 and 1, we transform the spatial estimates from *lrren* to fall between these bounds.
+See Figure ???
+for a visual representation of this process being applied to Cod.
+Figure \@ref(fig:hab-plot) depicts habitat preferences $Hab^2_{J,p}$ for each species.
+
+```{r hab-plot, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Static habitat preferences for each species in our population models (Yellotwtail, Cod, Haddock)."}
+knitr::include_graphics(c("Images/Habitat_3species.png"))
+```
+
+<font size="5">*Temperature Input*</font>
+
+Each species is assumed to have normally distributed temperature preferences ($N(\mu,\sigma)$).
+We assume Yellowtail Flounder's preferences are $N(8.75,4.25)$, while Haddock and Cod have preferences $N(9,4)$.
 We chose these values by combining information in the literature with temperatures recorded in the bottom trawl survey.
-We obtain estimated temperature data for the region for 2012 from FVCOM **cite**.
+Weekly estimated temperature data for the region for 2012 was obtained from FVCOM **cite**.
 We chose 2012 because the data displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values.
-We transform the temperature data to create an oscillating  pattern that increases 5 degrees Celsius over the duration of the simulation.
-**show image of average temperature oscilations?**
-
-Species-specific habitat preferences were derived by relating bottom trawl data to covariate predictors to create a niche model for each species.
+This data was also transform to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation.
+**show images of temperature and/or and/or temp videos???
+and/or average temperature oscillations?**
 
 --Describe difference between increasing and constant temperature scenarios (images?)
+
+In equation (1), $Hab^2_{J,p}$ is constant for the duration of the simulation, while $Tol_{c,p,wk}$ changes each week.
+Using a temperature gradient that repeats each year produces the same spatial preferences in a given week each year, which results in consistent spatial biomass patterns.
+Scenarios where the temperature increases over time creates spatial preferences that evolve as the water warms, which creates spatial biomass patterns that shift in a given week over the duration of the simulation.
 
@@
 If desired, VAST also allows users to include covariate data to better inform the model.
+
+\section{Results}
+
+```{r}
+
+knitr::kable(YT_results, caption = "Yellowtail error results")
+
+knitr::kable(Cod_results, caption = "Cod error results")
+
+knitr::kable(Had_results, caption = "Haddock error results")
+```
 

### 2022-12-21 Manuscript/Draft1.tex
updating manuscript

@@
 \newpage
+
+\section{Introduction}
+
+much of below is from \url{https://apps-nefsc.fisheries.noaa.gov/nefsc/ecosystem-ecology/}
+or
+\url{https://www.fisheries.noaa.gov/data-tools/fisheries-economics-united-states-data-and-visualizations}
+
+\begin{itemize}
+\tightlist
+\item
+  The eastern continental shelf is ecologically diverse and economically important
+\end{itemize}
+
+The Northeast United States continental shelf spans from the Outer Banks of North Carolina to the Gulf of Maine.
+The region covers over 250,000 km\(^2\) of ocean, extending over 200 km from shore in the largest areas in New England to just 30 km off shore in the southern regions.
+This ecologically diverse region contains approximately 18,000 vertebrate marine species.
+Commercial fisheries have been an important part of local economies for centuries.
+In 2019, New England fisheries produced \$22 billion in sales, which sustained over 200,000 jobs.
+Maintaining a healthy ecosystem is therefore vital to sustained ecological health and economic prosperity of the region.
+{[}@nefmc20{]}
+
+\begin{itemize}
+\tightlist
+\item
+  Bottom trawl survey is important for monitoring population trends
+\end{itemize}
+
+Fish stocks in this highly productive and economically important region are managed by the National Oceanic and Atmospheric Administration's (NOAA) Northeast Fisheries Science Center (NEFSC) in Woods Hole, Massachusetts.
+Federal biologists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (cite survey paper).
+The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf.
+The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits \textbf{just listed a few uses of survey change/others?}.
+
+The survey takes place twice each year- once in the spring and again in the fall.
+Since most spatial analyses and projections of future distributions typically assume a constant survey catchability and/or availability over time, NOAA's survey design includes sampling during approximately the same 2-3 week time period in each season.
+
+\begin{itemize}
+\tightlist
+\item
+  Climate change is happening
+  Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades, resulting in a shift in spatial distributions of many species.
+Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments.
+The implication of this is that the bottom trawl survey is actually sampling the population during a different life cycle stage than was originally assumed, which can lead to biased stock assessments.
+We are therefore interested in analyzing the impact of climate change on the accuracy of future stock assessment models as measured by NOAA's ongoing bottom-trawl survey along the East coast.
+\end{itemize}
+
+\textbf{use more info from initial proposal}
+
+\begin{itemize}
+\item
+  Fish are changing spatial distribution and have altered life stages (?) because of climate change
+  NYE paper
+\item
+  Population indexing methods may be becoming biased as a result
+\item
+  Briefly describe our study to test this
+\end{itemize}
+
+To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depend on temperature preferences.
+We can then consider the impact of climate change by simulating scenarios with repeating temperature patterns and those where temperature increases on average over time.
+In both cases we analyze the ability of stratified random sampling to track population trends.
+
+\section{Methods}
+
+\begin{itemize}
+\tightlist
+\item
+  Describe simulation study
+\end{itemize}
+
+We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on George's Bank, where movement of each species combine static species-specific habitat preferences with dynamic temperature preferences.
+Model dynamics are driven by dynamic temperature gradients estimated from data to create simulated data sets for each population where the true biomass is known.
+Using temperature gradients that repeat each year creates data sets with predictable, repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time.
+We conducting stratified random sampling on our simulation output to mimic the bottom trawl survey and compare the ability of contemporary indexing methods to track population trends.
+
+\subsection{Population Model Formulation}
+
+-- Used MixFishSim. Describe edits made to package
+
+We use the R package \emph{MixFishSim} (MFS) to model our populations {[}@dolder2020highly{]}.
+MFS is a discrete spatiotemporal simulation tool where users can model multiple species under varying environmental conditions.
+The package uses a delay-difference population model with discrete processes for growth, death, and recruitment of the population.
+We formulate the following inputs for the MFS package to address our research question:
+
+\emph{Study Area}
+
+We obtained a shapefile for the 15 strata that comprise George's Bank to use as our modeling environment.
+We discritized the region into a raster with 88 rows and 144 columns.
+Haddock inhabit all 15 strata in the domain Cod inhabit 13 strata, and yellowtail exist in 9 strata.
+Figure \ref{fig:strata-plot} shows the regions used in our models.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{Images/Strata}
+
+}
+
+\caption{Strata inhabited by each species in our population models.}\label{fig:strata-plot}
+\end{figure}
+
+\emph{Population Dynamics and Recruitment}
+
+The time step for our models is one week.
+MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass of in each cell in our study area {[}@dolder2020highly{]}.
+Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits.
+We chose to represent recruitment in the model using a Beverton-Holt formulation.
+Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.
+Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics.
+A full list of parameters used in our model can be seen below in Tables \ref{paramsALL} and \ref{tab:paramsSCENARIOS}.
+\textbf{missing spawning weeks, recruitment weeks, lambda}
+
+\emph{Movement}
+
+The package was designed to generate theoretical habitat preferences using Gaussian Random Fields that combine with hypothetical temperature gradients to drive the probability of movement from cell \(I\) to cell \(J\) using the formulation
+
+\begin{align}
+Pr(C_{wk+1}=J|C_{wk}=I) = \frac{e^{-\lambda \cdot d_{I,J}}\cdot(Hab^2_{J,p} \cdot Tol_{J,p,wk})}{\sum^C_{c=1}e^{-\lambda \cdot d} \cdot (Hab^2_{c,p} \cdot Tol_{c,p,wk})},
+\label{moveP}
+\end{align}
+
+where
+
+\(e^{-\lambda \cdot d_{I,J}}\) accounts for distance between cells \(I\) and \(J\),
+
+\(Hab^2_{J,p}\) is the static habitat value for species \(p\) in cell \(J\), and
+
+\(Tol_{c,p,wk}\) is the value from normally distributed temperature tolerance for species \(p\) in cell \(c\) in week \(wk\).
+
+Since we are modeling real species on the northeast continental shelf, we formulate the habitat and temperature components as follows.
+
+\emph{Habitat Input}
+
+Species-specific habitat preferences were derived using the \emph{lrren} tool from the R package \emph{envi} \textbf{cite} to create a niche model for each species.
+The \emph{lrren} tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors.
+We used bottom trawl point data from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow \textbf{cite trawl data?}.
+Depth and mean sediment size were used as our covariate predictors \textbf{cite}.
+Since the values in \(Hab^2_{J,p}\) are required to be between 0 and 1, we transform the spatial estimates from \emph{lrren} to fall between these bounds.
+See Figure ???
+for a visual representation of this process being applied to Cod.
+Figure \ref{fig:hab-plot} depicts habitat preferences \(Hab^2_{J,p}\) for each species.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{Images/Habitat_3species}
+
+}
+
+\caption{Static habitat preferences for each species in our population models (Yellotwtail, Cod, Haddock).}\label{fig:hab-plot}
+\end{figure}
+
+\emph{Temperature Input}
+
+Each species is assumed to have normally distributed temperature preferences (\(N(\mu,\sigma)\)).
+We assume Yellowtail Flounder's preferences are \(N(8.75,4.25)\), while Haddock and Cod have preferences \(N(9,4)\).
+We chose these values by combining information in the literature with temperatures recorded in the bottom trawl survey.
+Weekly estimated temperature data for the region for 2012 was obtained from FVCOM \textbf{cite}.
+We chose 2012 because the data displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values.
+This data was also transform to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation.
+\textbf{show images of temperature and/or and/or temp videos???
+and/or average temperature oscillations?}
+
+--Describe difference between increasing and constant temperature scenarios (images?)
+
+In equation (1), \(Hab^2_{J,p}\) is constant for the duration of the simulation, while \(Tol_{c,p,wk}\) changes each week.
+Using a temperature gradient that repeats each year produces the same spatial preferences in a given week each year, which results in consistent spatial biomass patterns.
+Scenarios where the temperature increases over time creates spatial preferences that evolve as the water warms, which creates spatial biomass patterns that shift in a given week over the duration of the simulation.
+
+-- Describe each scenario that is considered
+
+We consider three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time.
+Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius over the duration of the 20 year simulation.
+
+\emph{Simulating Bottom Trawl Survey and Population Indexing}
+
+-Describe post hoc sampling process and how data is used
+
+After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year.
+We sample in the same weeks that the Spring and Fall surveys take place and the number of the samples taken in each strata reflect true values.
+Most strata contain enough cells to sample a unique location in each survey over the duration of the simulation.
+For smaller strata we must repeat some sample locations.
+We then use the biomass collected from our samples in contemporary population indexing methods to estimate population trends.
+Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.
+
+--Stratified mean vs VAST with and without covariates
+
+We compare the stratified mean estimate of abundance to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
+The stratified mean is a typical survey-based approach that scales individual samples to the strata-level by considering the area of each strata, before scaling to the region-level based on the relative size of each strata.
+VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurrence (presence/absence).
+If desired, VAST also allows users to include covariate data to better inform the model.
+
+\section{Results}
 
@@
 \end{table}
-
-\newpage
-
-\section{Introduction}
-
-much of below is from \url{https://apps-nefsc.fisheries.noaa.gov/nefsc/ecosystem-ecology/}
-or
-\url{https://www.fisheries.noaa.gov/data-tools/fisheries-economics-united-states-data-and-visualizations}
-
-\begin{itemize}
-\tightlist
-\item
-  The eastern continental shelf is ecologically diverse and economically important
-\end{itemize}
-
-The Northeast United States continental shelf spans from the Outer Banks of North Carolina to the Gulf of Maine.
-The region covers over 250,000 km\(^2\) of ocean, extending over 200 km from shore in the largest areas in New England to just 30 km off shore in the southern regions.
-This ecologically diverse region contains approximately 18,000 vertebrate marine species.
-Commercial fisheries have been an important part of local economies for centuries.
-In 2019, New England fisheries produced \$22 billion in sales, which sustained over 200,000 jobs.
-Maintaining a healthy ecosystem is therefore vital to sustained ecological health and economic prosperity of the region.
-{[}@nefmc20{]}
-
-\begin{itemize}
-\tightlist
-\item
-  Bottom trawl survey is important for monitoring population trends
-\end{itemize}
-
-Fish stocks in this highly productive and economically important region are managed by the National Oceanic and Atmospheric Administration's (NOAA) Northeast Fisheries Science Center (NEFSC) in Woods Hole, Massachusetts.
-Federal biologists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (cite survey paper).
-The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf.
-The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits \textbf{just listed a few uses of survey change/others?}.
-
-The survey takes place twice each year- once in the spring and again in the fall.
-Since most spatial analyses and projections of future distributions typically assume a constant survey catchability and/or availability over time, NOAA's survey design includes sampling during approximately the same 2-3 week time period in each season.
-
-\begin{itemize}
-\tightlist
-\item
-  Climate change is happening
-  Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades, resulting in a shift in spatial distributions of many species.
-Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments.
-The implication of this is that the bottom trawl survey is actually sampling the population during a different life cycle stage than was originally assumed, which can lead to biased stock assessments.
-We are therefore interested in analyzing the impact of climate change on the accuracy of future stock assessment models as measured by NOAA's ongoing bottom-trawl survey along the East coast.
-\end{itemize}
-
-\textbf{use more info from initial proposal}
-
-\begin{itemize}
-\item
-  Fish are changing spatial distribution and have altered life stages (?) because of climate change
-  NYE paper
-\item
-  Population indexing methods may be becoming biased as a result
-\item
-  Briefly describe our study to test this
-\end{itemize}
-
-To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depend on temperature preferences.
-We can then consider the impact of climate change by simulating scenarios with repeating temperature patterns and those where temperature increases on average over time.
-In both cases we analyze the ability of stratified random sampling to track population trends.
-
-\section{Methods}
-
-\begin{itemize}
-\tightlist
-\item
-  Describe simulation study
-\end{itemize}
-
-We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on George's Bank, where movement of each species combine static species-specific habitat preferences with dynamic temperature preferences.
-Model dynamics are driven by dynamic temperature gradients estimated from data to create simulated data sets for each population where the true biomass is known.
-Using temperature gradients that repeat each year creates data sets with predictable, repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time.
-We conducting stratified random sampling on our simulation output to mimic the bottom trawl survey and compare the ability of contemporary indexing methods to track population trends.
-
-\subsection{Population Model Formulation}
-
--- Used MixFishSim. Describe edits made to package
-
-We use the R package \emph{MixFishSim} (MFS) to model our populations {[}@dolder2020highly{]}.
-MFS is a discrete spatiotemporal simulation tool where users can model multiple species under varying environmental conditions.
-The package uses a delay-difference population model with discrete processes for growth, death, and recruitment of the population.
-We formulate the following inputs for the MFS package to address our research question:
-
-\emph{Study Area}
-
-We obtained a shapefile for the 15 strata that comprise George's Bank to use as our modeling environment.
-We discritized the region into a raster with 88 rows and 144 columns.
-Haddock inhabit all 15 strata in the domain Cod inhabit 13 strata, and yellowtail exist in 9 strata.
-Figure \ref{fig:strata-plot} shows the regions used in our models.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{Images/Strata}
-
-}
-
-\caption{Strata inhabited by each species in our population models.}\label{fig:strata-plot}
-\end{figure}
-
-\emph{Population Dynamics and Recruitment}
-
-The time step for our models is one week.
-MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass of in each cell in our study area {[}@dolder2020highly{]}.
-Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits.
-We chose to represent recruitment in the model using a Beverton-Holt formulation.
-Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.
-Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics.
-A full list of parameters used in our model can be seen below in Tables \ref{paramsALL} and \ref{tab:paramsSCENARIOS}.
-\textbf{missing spawning weeks, recruitment weeks, lambda}
-
-\emph{Temperature and Habitat Input}
-
-The package was designed to generate theoretical habitat preferences using Gaussian Random Fields that combine with hypothetical temperature gradients to drive the probability of movement from cell \(I\) to cell \(J\) using the formulation
-
-\begin{align*}
-Pr(C_{wk+1}=J|C_{wk}=I) = \frac{e^{-\lambda \cdot d_{I,J}}\cdot(Hab^2_{J,p} \cdot Tol_{J,p,wk})}{\sum^C_{c=1}e^{-\lambda \cdot d} \cdot (Hab^2_{c,p} \cdot Tol_{c,p,wk})},
-\end{align*}
-
-where
-
-\(e^{-\lambda \cdot d_{I,J}}\) accounts for distance between cells \(I\) and \(J\),
-
-\(Hab^2_{J,p}\) is the static habitat value for species \(p\) in cell \(J\), and
-
-\(Tol_{c,p,wk}\) is the value from normally distributed temperature tolerance for species \(p\) in cell \(c\) in week \(wk\).
-
-Since we are modeling real species on the northeast continental shelf, we formulate the habitat and temperature components as follows.
-
-Each species is assumed to have normally distributed temperature preferences (\(~N(\mu,\sigma)\)).
-We assume Yellowtail's preferences are \(N(8.75,4.25)\), while Haddock and Cod have preferences \(N(9,4)\).
-We chose these values by combining information in the literature with temperatures recorded in the bottom trawl survey.
-We obtain estimated temperature data for the region for 2012 from FVCOM \textbf{cite}.
-We chose 2012 because the data displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values.
-We transform the temperature data to create an oscillating pattern that increases 5 degrees Celsius over the duration of the simulation.
-\textbf{show image of average temperature oscilations?}
-
-Species-specific habitat preferences were derived by relating bottom trawl data to covariate predictors to create a niche model for each species.
-
---Describe difference between increasing and constant temperature scenarios (images?)
-
--- Describe each scenario that is considered
-
-We consider three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time.
-Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius over the duration of the 20 year simulation.
-
-\emph{Simulating Bottom Trawl Survey and Population Indexing}
-
--Describe post hoc sampling process and how data is used
-
-After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year.
-We sample in the same weeks that the Spring and Fall surveys take place and the number of the samples taken in each strata reflect true values.
-Most strata contain enough cells to sample a unique location in each survey over the duration of the simulation.
-For smaller strata we must repeat some sample locations.
-We then use the biomass collected from our samples in contemporary population indexing methods to estimate population trends.
-Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.
-
---Stratified mean vs VAST with and without covariates
-
-We compare the stratified mean estimate of abundance to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
-The stratified mean is a typical survey-based approach that scales individual samples to the strata-level by considering the area of each strata, before scaling to the region-level based on the relative size of each strata.
-VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurrence (presence/absence).
-If desired, VAST also allows users to include covariate data to better inform the model.
 

### 2022-12-21 TestScripts/GB_habitat_create.R

@@ -1762,3 +1762,111 @@ max(temp_max[21:40,1])
 
 
 
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+######################################################################################
+
+
+#5) PLOTTING HABITAT FOR PUBLICATION
+
+
+spp_names <- c("Yellowtail Flounder","Cod","Haddock")
+
+spp_names_short <- c("YT","Cod","Had")
+
+
+#function to rotate image before plotting because image.plot rotates it
+rotate <- function(x) t(apply(x, 2, rev))
+
+library(raster)
+#haddock contains all stratas used
+Had_ras <- readRDS(file="TestScripts/Habitat_plots/Haddock/Had_Weighted_AdaptFalse_RASTER_res2.RDS")
+plot(Had_ras)
+#load others to extract covariate values
+#Yellowtail
+YT_ras <- readRDS(file="TestScripts/Habitat_plots/YellowtailFlounder/Yell_Weighted_AdaptFalse_RASTER_res2.RDS")
+plot(YT_ras)
+#Cod
+Cod_ras <- readRDS(file="TestScripts/Habitat_plots/Cod/Cod_Weighted_AdaptFalse_RASTER_res2.RDS")
+plot(Cod_ras)
+
+hab <- readRDS(file="hab_GB_3species.RDS") #courser resolution
+names(hab$hab) <- c("YT","Cod","Had")
+
+library(ggplot2)
+library(gridExtra)
+library(plotly)
+
+
+
+  surv_temp1 <- list() #FOR STORING PLOTS FOR SURVEYS POINTS OVER POPULATION VALUES
+  
+      
+
+for(s in seq(3)) {
+  
+      #STORING PLOTS FOR SURVEYS POINTS OVER HABITAT COVARIATES 
+      #set habitat
+      if(s==1){hab_ras = YT_ras}
+      if(s==2){hab_ras = Cod_ras}
+      if(s==3){hab_ras = Had_ras}
+      
+      temp_ras <- reshape2::melt(as.matrix(hab_ras), c("x", "y"), value.name = "Suitability") 
+      
+      surv_temp1[[s]] <- ggplot() +
+        geom_raster(data=temp_ras,aes(x=y,y=rev(x),fill=Suitability)) + #plot biomass
+        
+        scale_fill_distiller(palette = "Spectral")+  #set the color pallet and color limits
+        theme_void()+
[…]

### 2022-12-22 Manuscript/Draft1.Rmd
updating manuscript

@@
     toc: no
-fontsize: 12pt
 header-includes:
@@
 
-We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on George's Bank, where movement of each species combine static species-specific habitat preferences with dynamic temperature preferences.
-Model dynamics are driven by dynamic temperature gradients estimated from data to create simulated data sets for each population where the true biomass is known.
+We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on George's Bank, where movement of each species combine static species-specific habitat preferences with temperature preferences.
+Model dynamics are driven by a time series of temperature gradients that were estimated from data to create simulated data sets for each population where the true biomass is known.
 Using temperature gradients that repeat each year creates data sets with predictable, repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time.
@@
 The package uses a delay-difference population model with discrete processes for growth, death, and recruitment of the population.
-We formulate the following inputs for the MFS package to address our research question:
+We formulate the following inputs for the MFS package to address our research question.
 
@@
 We discritized the region into a raster with 88 rows and 144 columns.
-Haddock inhabit all 15 strata in the domain Cod inhabit 13 strata, and yellowtail exist in 9 strata.
+Haddock inhabit all 15 strata in the domain, Cod inhabit 13 strata, and Yellowtail exist in 9 strata.
 Figure \@ref(fig:strata-plot) shows the regions used in our models.
@@
 The time step for our models is one week.
-MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass of in each cell in our study area [@dolder2020highly].
+MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass in each cell in our study area [@dolder2020highly].
 Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits.
-We chose to represent recruitment in the model using a Beverton-Holt formulation.
+We chose to represent recruitment in the model using a Beverton-Holt formulation **cite**.
 Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.
@@
 A full list of parameters used in our model can be seen below in Tables \@ref(paramsALL) and \@ref(tab:paramsSCENARIOS).
-**missing spawning weeks, recruitment weeks, lambda**
 
@@
 \begin{align}
-Pr(C_{wk+1}=J|C_{wk}=I) = \frac{e^{-\lambda \cdot d_{I,J}}\cdot(Hab^2_{J,p} \cdot Tol_{J,p,wk})}{\sum^C_{c=1}e^{-\lambda \cdot d} \cdot (Hab^2_{c,p} \cdot Tol_{c,p,wk})},
+Pr(C_{wk+1}=J|C_{wk}=I) = \frac{e^{-\lambda \cdot d_{I,J}}\cdot(Hab^2_{J,s} \cdot Tol_{J,s,wk})}{\sum^C_{c=1}e^{-\lambda \cdot d} \cdot (Hab^2_{c,s} \cdot Tol_{c,s,wk})},
 \label{moveP}
@@
 
-  $Hab^2_{J,p}$ is the static habitat value for species $p$ in cell $J$, and
+  $Hab^2_{J,s}$ is the static habitat value for species $s$ in cell $J$, and
 
-   $Tol_{c,p,wk}$ is the value from normally distributed temperature tolerance for species $p$ in cell $c$ in week $wk$.
+   $Tol_{c,s,wk}$ is the value from normally distributed temperature tolerance for species $s$ in cell $c$ in week $wk$.
 
@@
 Depth and mean sediment size were used as our covariate predictors **cite**.
-Since the values in $Hab^2_{J,p}$ are required to be between 0 and 1, we transform the spatial estimates from *lrren* to fall between these bounds.
+Since the values in $Hab^2_{J,s}$ are required to be between 0 and 1, we transform the spatial estimates from *lrren* to fall between these bounds.
 See Figure ???
 for a visual representation of this process being applied to Cod.
-Figure \@ref(fig:hab-plot) depicts habitat preferences $Hab^2_{J,p}$ for each species.
+Figure \@ref(fig:hab-plot) depicts habitat preferences $Hab^2_{J,s}$ for each species.
 
@@
 
-In equation (1), $Hab^2_{J,p}$ is constant for the duration of the simulation, while $Tol_{c,p,wk}$ changes each week.
+In equation (1), $Hab^2_{J,s}$ is constant for the duration of the simulation, while $Tol_{c,s,wk}$ changes each week.
 Using a temperature gradient that repeats each year produces the same spatial preferences in a given week each year, which results in consistent spatial biomass patterns.
@@
 
-We consider three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time.
+We consider 20 year simulations under three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time.
 Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius over the duration of the 20 year simulation.
+We therefore simulate a total of 6 scenarios.
+**show line plot of population for each?**
 
@@
 
-We compare the stratified mean estimate of abundance to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
+We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
 The stratified mean is a typical survey-based approach that scales individual samples to the  strata-level by considering the area of each strata, before scaling to the region-level based on the relative size of each strata.
@@
 If desired, VAST also allows users to include covariate data to better inform the model.
+Covariates can be static (eg.
+habitat preferences) or dynamics (eg.
+temperature).
+The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and take on the order of hours to complete.
+
+We follow the advice given in **cite Thorson 2019 Guidance...** to build VAST models to estimate biomass on George's Bank using stratified mean samples from our model output.
+In addition to exploring different link functions and assumed distributions, our VAST model-building process included testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
+We also carried out the same process both including covariates in our model as well as running models without covariate information.
+We considered covaraites in the form of dyamic temperature values and/or static habitat values from our population model.
+When using covariates we ultimately decided to provide the most information to the model by including both covariates for both linear predictors.
+Since we know the true population values in our models we calculate the absolute error of each VAST estimate to compare between potential settings.
+Through this process, and in consultation with the VAST package creator, we determined setting that allowed VAST models to converge for all of our scenarios while also providing the lowest absolute error values.
+Settings for our VAST models can be seen in Table ???.
+
+Our goal is to determine indexing approaches and settings that are robust to future environmental conditions and resulting spatial biomass patterns.
+An underlying assumption in all indexing methods that individual random samples combine to accurately represent true abudance by a) sampling all strata in which the population exists and b) low enough noise level in the samples to allow for a discernable pattern.
+This assumption can be questioned given enough noise in the sampling process **cite?** and/or shifting spatial preferences driven by climate change causing a population to move into a previously uninhabited strata.
+To simulate the impact of noise, indexing estimates after adding noise to our samples versus those using the true sampling values.
+To evaluate the effect of populations moving into new habitat, we compare indexing estimates using samples from all strata versus those that only include a subset of the full spatial domain for each species.
+
+When combining population trends for each species, differing temperature scenarios, and sampling possibilities (noise, strata, covariates) there are a number of scenario combinations to consider.
+The columnbs in Table COMBOS show the possible choices that define each scenario.
+
+<div align="center">Table COMBOS Each index estimate chooses one condition from each of the following columns.
+There are $3*3*2*2*2*2=144$ VAST model combinations and $3*3*2*2*2=72$ stratified mean estimates.</div>
+
+```{r scenarios, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
+library("knitr")
+library("kableExtra")
+library("dplyr")
+library("readr")
+library("tidyr")
+library("forcats")
+
+tabl <- "
+|	Species	|	Population Trend | Temperature Scenario | Strata Included | Noise in Data | Covariates (VAST only) 
+|:--	|:-- |:-- |:-- |:-- |:--- 
+|Yellowtail | Increasing | Repeating | All strata | No Noise| No Covariates
+|Cod | Constant | Increasing 5$^{\\circ}$| Subset | Yes Noise| Temp + Habitat
+|Haddock | Decreasing | | | |
+"
+cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
+
+```
 
@@
 \pagebreak
-Table 1. Parameters used in all population models.
+
+<div align="center">Table 1. Parameters used in all population models.</div>
+
 ```{r paramsALL, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
+library("knitr")
+library("kableExtra")
+library("dplyr")
+library("readr")
+library("tidyr")
+library("forcats")
+
 tabl <- "
-|		|	Description | Unit |Yellowtail | Cod | Haddock | Source
+|	Parameter	|	Description | Unit |Yellowtail | Cod | Haddock | Source
 |:---	|:------- |:-- |:--- |:-- |:-- |:--
@@
 | $\\sigma^2$ | Variance in recruited fish | kg$^2$ | 0.55 | 0.55 | 0.55 ||
+| $\\lambda$ | Decay rate for movement | - | 0.7 | 0.7 | 0.7 ||
+| $Spwn_s$ | Spawning weeks for species $s$ | wk | 9-12 | 8-13 | 11-14 ||
+| $Rec_s$ | Recruitment weeks for species $s$ | wk | 9-12 | 8-13 | 11-14 ||
 "
 cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
+
+```
+
+```{r paramsALL2,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
+library("knitr")
+library("kableExtra")
+library("dplyr")
+library("readr")
+library("tidyr")
+library("forcats")
+
+param_table1 = read_csv("Parameters1.csv")
+
+summary_tab1 <- param_table1 %>% 
+ # dplyr::mutate_if(is.numeric, list(~as.character(signif(., 3))))%>%
+
+  mutate_if(grepl('rrr',.), ~replace(., grepl('rrr', .), paste("\U03C1")))#should be replacing r with the rho symbol
+ # mutate_if(grepl('W1',.), ~replace(., grepl('W1', .), "WW")) #should be replacing beta with the symbol
+ #    mutate_if(grepl('wk',.), ~replace(., grepl('wk', .), "1/wk")) %>% #should be replacing beta with the symbol
+ #
+
+# Change the orientation of the table so that Species become the variable headings
+
+# Create table to present the data]
+Fig_1 <- kable(summary_tab1, 
+  caption = "Parameters used in all population models.",
+    format = "latex", booktabs = TRUE) %>%
+  kable_styling(font_size = 10)
+
+```
+
+ ```{r, results='asis'}
+
+ Fig_1
+
  ```
@@
 summary_tab <- param_table %>% 
-  dplyr::mutate_if(is.numeric, funs(as.character(signif(., 3))))%>%
+  dplyr::mutate_if(is.numeric, list(~as.character(signif(., 3))))%>%
 
@@
 
-Table 1. Maturity-, weight-, and selectivity-at-age of the simulated fish population.
-<!---markdown style table. f*cking escape characters-->
-```{r lifehisttabl, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
+<div align="center">Table XX. Parameters used in all VAST models.</div>
+
+```{r VASTsettings, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
+
 tabl <- "
-|	Age	|	Maturity | Weight (kg) | Fishery Selectivity (before change if  applicable) | Fishery Selectivity (after change if applicable)
-|:---	|:------ |:------ |:------ |:------
-|	1	|	0.04 | 0.15 | 0.07 | 0.02
-|	2	|	0.25 | 0.5 | 0.17 | 0.05
-|	3	|	0.60 | 0.9 | 0.36 | 0.12
-| 4 | 0.77 | 1.4 | 0.61 | 0.27
-| 5 | 0.85 | 2.0 | 0.81 | 0.50
-| 6 | 0.92 | 2.6 | 0.92 | 0.74
-| 7 | 1.00 | 3.2 | 0.97 | 0.89
-| 8 | 1.00 | 4.1 | 0.99 | 0.96
-| 9 | 1.00 | 5.9 | 1.00 | 0.99
-| 10+ | 1.00 | 9.0 | 1.00 | 1.00
+|	Parameter	|	Description | Input/Value
+|:-	|:------- |:----
+| *ObsModel* | Link function and assumed distribution | c(10,2)
+|FieldCOnfig | Specified spatial and/or spatio-temporal variation in predictors| c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1)
+| RhoConfig | Specifying whether intercepts or spatio-temporal variation is structured among time intervals | c(Beta1=3, Beta2=3, Epsilon1=0, Epsilon2=4)
+| X1_formula | Right-sided formula affecting the 1st linear predictor  | X1_formula = ~ poly(Temp, degree=2 )
+| X2_formula | Right-sided formula affecting the 2nd linear predictor | X2_formula = ~ poly(Temp, degree=2 ) + poly(Habitat, degree=2 )
+
 "
 cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
-```
-
-\pagebreak
-Table 2. Naming convention and details of the data-limited methods evaluated.
-```{r methodtable, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-tabl <- "
-|	Method	|	Details
-|:---	|:------ 
-|	Ismooth	|	$C_{targ,y+1:y+2} =\\overline{C}_{3,y}(e^{\\lambda})$ <br> where $\\overline{C}_{3,y}$ is the most recent three year average; <br> $\\overline{C}_{3,y} = \\frac{1}{3}\\sum_{t=1}^{t=3}C_{y-t}$ <br> and $\\lambda$ is the slope of a log linear regression of a LOESS-smoothed average index of abundance (spring and fall) with span = 0.3: <br> $\\hat{I}_y=loess(\\hat{I}_y)$ and $LN(\\widehat{I_y}) = b + \\lambda y$
-| Islope	|	$C_{targ,y+1:y+2} = 0.8\\overline{C}_{5,y}(1+0.4e^{\\lambda})$ where $\\overline{C}_{5,y}$ is the most recent five-year average catch through year $y-1$: <br> $\\overline{C}_{5,y} = \\frac{1}{5}\\sum_{t=1}^{t=5}C_{y-t}$ <br> and $\\lambda$ is the slope of a log-linear regression of the most recent five years of the averaged index.
-|	Itarget	|	$C_{targ,y+1:y+2}=\\left[0.5C_{ref}\\left(\\frac{\\overline{I}_{5,y}-I_{thresh}}{I_{target}-I_{thresh}}\\right)\\right]$ $\\overline{I}_{5,y} \\geq I_{thresh}$ <br> $C_{targ,y+1:y+2}=\\left[0.5C_{ref}\\left(\\frac{\\overline{I}_{5,y}}{I_{thresh}}\\right)^{2}\\right]$ $\\overline{I}_{5,y} <  I_{thresh}$; <br>  $C_{ref}$ is the average catch over the reference period (years 26 through 50): <br> $C_{ref} = \\frac{1}{25}\\sum_{y=26}^{y=50}C_y$; <br>  $I_{target}$ is 1.5 times the average index over the reference period: <br> $I_{target} = \\frac{1}{25}\\sum_{y=26}^{y=50}\\overline{I}_y$; <br> $I_{thresh}$ = 0.8 $I_{target}$, and is the most recent five year average of the combined spring and fall index: <br> $\\overline{I}_{5,y} = \\frac{1}{5}\\sum_{t=1}^{t=5}\\overline{I}_{y-t+1}$
-| Skate | $C_{targ,y+1:y+2}= F_{rel} \\overline{I}_{3,y}$ where <br> $F_{rel} =median \\left(\\frac{\\overline{C}_{3,\\textbf{Y}}}{\\overline{I}_{3,\\textbf{Y}}} \\right)$ <br> is the median relative fishing mortality rate calculated using a 3 year moving average of the catch and average survey index across all available years ($\\textbf{Y}$): <br> $\\overline{C}_{3,y} = \\frac{1}{3} \\sum_{t=1}^{t=3} C_{y-t}$ and <br> $\\overline{I}_{3,y} = \\frac{1}{3} \\sum_{t=1}^{t=3} I_{y-t+1}$
-| An Index Method (AIM) | AIM first calculates the annual relative $F$: $F_{rel,y}=\\frac{C_y}{\\frac{1}{3}\\sum_{t=1}^{t=3}\\overline{I}_{y-t+1}}$ <br> and the annual replacement ratio: $\\Psi_y = \\frac{\\overline{I}_y}{\\frac{1}{5}\\sum_{t=1}^{t=5}\\overline{I}_{y-t}}$.
-<br> These values are used in a regression: $LN(\\Psi_y)= b + \\lambda LN(F_{rel,y})$ to determine $F_{rel,*}$, which is the value of $F_{rel,y}$ where the predicted $\\Psi=1$ or <br> $LN(\\Psi)=0$.
-$F_{rel,*}$ is called either the “stable” or “replacement” $F$, and is used to calculate the target catch: $C_{targ,y+1:y+2} = \\overline{I}_yF_{rel,*}$.
-| Dynamic Linear Model (DynLin) | @Langan2021DLM.
-| Expanded survey  biomass method 1 $F_{40\\%}$ (ES-FSPR) | $C_{targ,y+1:y+2}=B_{\\bar{I},y}\\mu_{targ}$ <br> where $B_{\\bar{I}}$ is the average of estimated fully-selected biomass from each survey: <br> $B_{\\bar{I},y}=\\frac{1}{2}\\left(\\frac{I_{spr,y}}{q_{spr}} +\\frac{I_{fall,y-1}}{q_{fall}}\\right)$ <br> and target exploitation fraction, $\\mu_{targ}$ is calculated as: <br> $\\mu_{targ}=\\frac{F_{targ}}{Z_{targ}}\\left(1-e^{-Z_{targ}}\\right)$; <br> $F_{targ}=F_{40\\%}$ and <br> $Z_{targ}=F_{targ}+M$
-| Expanded survey biomass method 2 $F=$ AIM replacement (ES-Fstable) | Same as the above expanded survey method, but with $\\mu_{targ}$ equal to the stable exploitation fraction $F_{rel,*}$ calculated using the AIM approach (see above).
-| Expanded survey biomass method 3 $F=M$ (ES-FM) | Same as the above expanded survey methods, but with the target exploitation rate set to the assumed $M$: $F_{targ}=M$.
-| Expanded survey biomass method 4 $F=$ recent average (ES-Frecent) | Same as the above expanded survey methods, but with the target exploitation fraction set to the most recent three year average exploitation fraction: <br> ${\\mu_{targ}}=\\frac{\\sum_{y-2}^{y}\\mu_y}{3}$ <br> $\\mu_{y}=\\frac{C_{y-1}}{B_{\\bar{I},y}}$
-| Catch curve Method 1 $F_{40\\%}$ (CC-FSPR) | $C_{targ,y+1:y+2}= \\frac{F_{targ}}{Z_{avg,y}}B_{cc,y}\\left(1-e^{-Z_{avg,y}}\\right)$ <br>  where $B_{cc}$ is the estimated biomass: <br> $B_{cc,y} = \\frac{C_{y-1}}{\\frac{F_{avg,y}}{Z_{avg,y}}{\\left(1-e^{-Z_{avg,y}}\\right)}}$ with <br> $Z_{avg,y} = \\frac{Z_{spring, y}+Z_{fall,y-1}}{2}$; <br> $F_{avg,y-1} = Z_{avg,y-1} - M$ and, <br> $F_{targ}=F_{40\\%}$.
-Survey catch at age used in catch curve to estimate $Z$.
-| Catch curve Method 2 $M$ (CC-FM) | Same as catch curve method 1 above, but with $F_{targ}=M$.
-| Ensemble | Median of catch advice provided by AIM, CC-FSPR, ES-Frecent, ES-FSPR, Islope, Itarget, Ismooth, and Skate methods.
-"
-cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
-```
-
-\pagebreak
-Table 3. Summary of the scenarios evaluated within the study design.
-```{r schemetable, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-tabl <- "
-
-|	Factors | Variants
-|:------ |:------
-|	retrospective source | catch or <br> natural mortality
-|	fishing history | $F_{MSY}$ in second half of base period or <br> overfishing throughout base period (2.5x$F_{MSY}$)
-|	fishery selectivity blocks | constant selectivity or <br> selectivity changes in second half of base period
-| catch advice multiplier | applied as is from DLM (1) or <br> reduced from DLM (0.75)
-"
-#Table: (\#tab:schemtable) Coding scheme.
-cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
-```
-
-\pagebreak
-
-## List of Figures
-
-```{r figcaps, echo=FALSE}
-figcap1 <- "Inner quartiles and medians for all performance measures across all scenarios and runs for each method.
-Vertical lines are shown at a value of 1 for the performance measures that are relative to the MSY reference points (A,B,C)."
-figcap2 <- "Relationship between long-term average spawning biomass and average catch (relative to MSY levels) for each method.
-Each point represents the median for a given scenario, separated by the source of the retrospective pattern (catch or M)."
-figcap3 <- "Median performance measures for each method, separated by the source of the retrospective error (catch = black, M = gray) and the exploitation history in the base period (always overfishing at $2.5 x F_{MSY}$ (circle), or $F$ reduced to $F_{MSY}$ during base period (triangle)).
-Vertical lines are shown at a value of 1 for the performance measures that are relative to the MSY reference points (A,B,C)."
-figcap4 <- "Median $F / F_{MSY}$ for each method, with results separated by the exploitation history in the base period (always overfishing at $2.5 x F_{MSY}$ (circle), or $F$ reduced to $F_{MSY}$ during base period (triangle)) showing A) short- (gray) versus long-term (black) values, and B) with (black) or without (gray) a buffer applied when setting the catch (catch multiplier = 0.75 or 1)."
-figcap5 <- "Relationship between long-term average catch and spawning stock biomass relative to their reference points by method.
-Each point represents the average for years 21-40 in the feedback period for a single iteration of a scenario.
-The scenario shown is where catch was the source of the retrospective pattern with $F$ reduced to $F_{MSY}$ in the second half of the base period, there was a single selectivity block, and where no buffer was applied to the catch advice (catch multiplier = 1)."
 
@@
 
-Figure 1. `r figcap1`
+<!-- ```{r PMsallscenarios, echo=FALSE, message=FALSE, warnings=FALSE, out.width='7in', fig.cap=figcap1} -->
+<!-- #knitr::include_graphics(file.path(getwd(),"MS_tables_figs_rev1/final_ms_figs","Figure_1.pdf")) -->
+<!-- ``` -->
 
-Figure 2. `r figcap2`
+<!-- ```{r SSmsyvsCMSY, echo=FALSE, message=FALSE, warnings=FALSE, out.width='7in', fig.cap=figcap2} -->
+<!-- #knitr::include_graphics(file.path(getwd(),"MS_tables_figs_rev1/final_ms_figs","Figure_2.pdf")) -->
+<!-- ``` -->
 
-Figure 3. `r figcap3`
+<!-- ```{r PMsbyretroF, echo=FALSE, message=FALSE, warnings=FALSE, out.width='6in', fig.cap=figcap3} -->
+<!-- #knitr::include_graphics(file.path(getwd(),"MS_tables_figs_rev1/final_ms_figs","Figure_3.pdf")) -->
+<!-- ``` -->
 
-Figure 4. `r figcap4`
+<!-- ```{r FFmsybyscenario, echo=FALSE, message=FALSE, warnings=FALSE, out.width='6in', fig.cap=figcap4} -->
+<!-- #knitr::include_graphics(file.path(getwd(),"MS_tables_figs_rev1/final_ms_figs","Figure_4.pdf")) -->
+<!-- ``` -->
 
-Figure 5. `r figcap5`
-
-```{r PMsallscenarios, echo=FALSE, message=FALSE, warnings=FALSE, out.width='7in', fig.cap=figcap1}
-#knitr::include_graphics(file.path(getwd(),"MS_tables_figs_rev1/final_ms_figs","Figure_1.pdf"))
-```
-
-```{r SSmsyvsCMSY, echo=FALSE, message=FALSE, warnings=FALSE, out.width='7in', fig.cap=figcap2}
-#knitr::include_graphics(file.path(getwd(),"MS_tables_figs_rev1/final_ms_figs","Figure_2.pdf"))
-```
-
-```{r PMsbyretroF, echo=FALSE, message=FALSE, warnings=FALSE, out.width='6in', fig.cap=figcap3}
-#knitr::include_graphics(file.path(getwd(),"MS_tables_figs_rev1/final_ms_figs","Figure_3.pdf"))
-```
-
-```{r FFmsybyscenario, echo=FALSE, message=FALSE, warnings=FALSE, out.width='6in', fig.cap=figcap4}
-#knitr::include_graphics(file.path(getwd(),"MS_tables_figs_rev1/final_ms_figs","Figure_4.pdf"))
-```
-
-```{r linearvsdiffuse, echo=FALSE, message=FALSE, warnings=FALSE, out.width='7in', fig.cap=figcap5}
-#knitr::include_graphics(file.path(getwd(),"MS_tables_figs_rev1/final_ms_figs","Figure_5.pdf"))
-```
+<!-- ```{r linearvsdiffuse, echo=FALSE, message=FALSE, warnings=FALSE, out.width='7in', fig.cap=figcap5} -->
+<!-- #knitr::include_graphics(file.path(getwd(),"MS_tables_figs_rev1/final_ms_figs","Figure_5.pdf")) -->
+<!-- ``` -->

### 2022-12-22 Manuscript/Draft1.tex
updating manuscript

@@
 \documentclass[
-  12pt,
 ]{article}
@@
 We discritized the region into a raster with 88 rows and 144 columns.
-Haddock inhabit all 15 strata in the domain Cod inhabit 13 strata, and yellowtail exist in 9 strata.
+Haddock inhabit all 15 strata in the domain, Cod inhabit 13 strata, and Yellowtail exist in 9 strata.
 Figure \ref{fig:strata-plot} shows the regions used in our models.
@@
 The time step for our models is one week.
-MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass of in each cell in our study area {[}@dolder2020highly{]}.
+MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass in each cell in our study area {[}@dolder2020highly{]}.
 Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits.
-We chose to represent recruitment in the model using a Beverton-Holt formulation.
+We chose to represent recruitment in the model using a Beverton-Holt formulation \textbf{cite}.
 Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.
@@
 A full list of parameters used in our model can be seen below in Tables \ref{paramsALL} and \ref{tab:paramsSCENARIOS}.
-\textbf{missing spawning weeks, recruitment weeks, lambda}
 
@@
 
-\(Hab^2_{J,p}\) is the static habitat value for species \(p\) in cell \(J\), and
+\(Hab^2_{J,s}\) is the static habitat value for species \(s\) in cell \(J\), and
 
-\(Tol_{c,p,wk}\) is the value from normally distributed temperature tolerance for species \(p\) in cell \(c\) in week \(wk\).
+\(Tol_{c,s,wk}\) is the value from normally distributed temperature tolerance for species \(s\) in cell \(c\) in week \(wk\).
 
@@
 Depth and mean sediment size were used as our covariate predictors \textbf{cite}.
-Since the values in \(Hab^2_{J,p}\) are required to be between 0 and 1, we transform the spatial estimates from \emph{lrren} to fall between these bounds.
+Since the values in \(Hab^2_{J,s}\) are required to be between 0 and 1, we transform the spatial estimates from \emph{lrren} to fall between these bounds.
 See Figure ???
 for a visual representation of this process being applied to Cod.
-Figure \ref{fig:hab-plot} depicts habitat preferences \(Hab^2_{J,p}\) for each species.
+Figure \ref{fig:hab-plot} depicts habitat preferences \(Hab^2_{J,s}\) for each species.
 
@@
 
-In equation (1), \(Hab^2_{J,p}\) is constant for the duration of the simulation, while \(Tol_{c,p,wk}\) changes each week.
+In equation (1), \(Hab^2_{J,s}\) is constant for the duration of the simulation, while \(Tol_{c,s,wk}\) changes each week.
 Using a temperature gradient that repeats each year produces the same spatial preferences in a given week each year, which results in consistent spatial biomass patterns.
@@
 
-We consider three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time.
+We consider 20 year simulations under three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time.
 Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius over the duration of the 20 year simulation.
+We therefore simulate a total of 6 scenarios.
+\textbf{show line plot of population for each?}
 
@@
 
-We compare the stratified mean estimate of abundance to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
+We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
 The stratified mean is a typical survey-based approach that scales individual samples to the strata-level by considering the area of each strata, before scaling to the region-level based on the relative size of each strata.
@@
 If desired, VAST also allows users to include covariate data to better inform the model.
+Covariates can be static (eg.
+habitat preferences) or dynamics (eg.
+temperature).
+The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and take on the order of hours to complete.
+
+We follow the advice given in \textbf{cite Thorson 2019 Guidance\ldots{}} to build VAST models to estimate biomass on George's Bank using stratified mean samples from our model output.
+In addition to exploring different link functions and assumed distributions, our VAST model-building process included testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
+We also carried out the same process both including covariates in our model as well as running models without covariate information.
+We considered covaraites in the form of dyamic temperature values and/or static habitat values from our population model.
+When using covariates we ultimately decided to provide the most information to the model by including both covariates for both linear predictors.
+Since we know the true population values in our models we calculate the absolute error of each VAST estimate to compare between potential settings.
+Through this process, and in consultation with the VAST package creator, we determined setting that allowed VAST models to converge for all of our scenarios while also providing the lowest absolute error values.
+Settings for our VAST models can be seen in Table ???.
+
+Our goal is to determine indexing approaches and settings that are robust to future environmental conditions and resulting spatial biomass patterns.
+An underlying assumption in all indexing methods that individual random samples combine to accurately represent true abudance by a) sampling all strata in which the population exists and b) low enough noise level in the samples to allow for a discernable pattern.
+This assumption can be questioned given enough noise in the sampling process \textbf{cite?} and/or shifting spatial preferences driven by climate change causing a population to move into a previously uninhabited strata.
+To simulate the impact of noise, indexing estimates after adding noise to our samples versus those using the true sampling values.
+To evaluate the effect of populations moving into new habitat, we compare indexing estimates using samples from all strata versus those that only include a subset of the full spatial domain for each species.
+
+When combining population trends for each species, differing temperature scenarios, and sampling possibilities (noise, strata, covariates) there are a number of scenario combinations to consider.
+The columnbs in Table COMBOS show the possible choices that define each scenario.
+
+Table COMBOS Each index estimate chooses one condition from each of the following columns. There are \(3*3*2*2*2*2=144\) VAST model combinations and \(3*3*2*2*2=72\) stratified mean estimates.
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.16}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.16}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.16}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.16}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.16}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.21}}@{}}
+\toprule
+\begin{minipage}[b]{\linewidth}\raggedright
+Species
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Population Trend
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Temperature Scenario
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Strata Included
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Noise in Data
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Covariates (VAST only)
+\end{minipage} \\
+\midrule
+\endhead
+Yellowtail & Increasing & Repeating & All strata & No Noise & No Covariates \\
+Cod & Constant & Increasing 5\(^{\circ}\) & Subset & Yes Noise & Temp + Habitat \\
+Haddock & Decreasing & & & & \\
+\bottomrule
+\end{longtable}
 
@@
 \toprule
-& Description & Unit & Yellowtail & Cod & Haddock & Source \\
+Parameter & Description & Unit & Yellowtail & Cod & Haddock & Source \\
 \midrule
@@
 \(\sigma^2\) & Variance in recruited fish & kg\(^2\) & 0.55 & 0.55 & 0.55 & \\
+\(\lambda\) & Decay rate for movement & - & 0.7 & 0.7 & 0.7 & \\
+\(Spwn_s\) & Spawning weeks for species \(s\) & wk & 9-12 & 8-13 & 11-14 & \\
+\(Rec_s\) & Recruitment weeks for species \(s\) & wk & 9-12 & 8-13 & 11-14 & \\
 \bottomrule
@@
 
-\begin{verbatim}
-## Warning: `funs()` was deprecated in dplyr 0.8.0.
-## Please use a list of either functions or lambdas: 
-## 
-##   # Simple named list: 
-##   list(mean = mean, median = median)
-## 
-##   # Auto named with `tibble::lst()`: 
-##   tibble::lst(mean, median)
-## 
-##   # Using lambdas
-##   list(~ mean(., trim = .2), ~ median(., na.rm = TRUE))
-## This warning is displayed once every 8 hours.
-## Call `lifecycle::last_lifecycle_warnings()` to see where this warning was generated.
-\end{verbatim}
+\begin{table}
+
+ \caption{\label{tab:paramsALL2}Parameters used in all population models.}
+ \centering
+ \fontsize{10}{12}\selectfont
+ \begin{tabular}[t]{lllllll}
+ \toprule
+ Parameter & Description & Unit & Yellowtail & Cod & Haddock & Source\\
+ \midrule
+ <U+03C1> & Ford's growth coefficient & 1/wk & 4.48 & 4.43 & 4.49 & NA\\
+ M & Natural mortality & 1/wk & 0.2064 & 0.2728 & 0.334 & NA\\
+ W1 & Weight of fully recruited fish & kg & 0.39 & 2.95 & 1.12 & NA\\
+ W2 & Weight of pre-recruit fish & kg & 0.13 & 0.39 & 0.19 & NA\\
+ sigma & Variance in recruited fish & kg*kg & 0.55 & 0.55 & 0.55 & NA\\
+ \addlinespace
+ lambda & Decay rate for movement & - & 0.7 & 0.7 & 0.7 & NA\\
+ spwn & Spawning weeks for species s & wks & 9-12 & 8-13 & 11-14 & NA\\
+ rec & Recruitment weeks for species s & wks & 9-12 & 8-13 & 11-14 & NA\\
+ \bottomrule
+ \end{tabular}
+ \end{table}
 
@@
 
-Table 1. Maturity-, weight-, and selectivity-at-age of the simulated fish population.
+Table XX. Parameters used in all VAST models.
 
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.12}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.22}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.22}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.22}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.22}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.13}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.53}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.33}}@{}}
 \toprule
 \begin{minipage}[b]{\linewidth}\raggedright
-Age
+Parameter
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Maturity
+Description
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Weight (kg)
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Fishery Selectivity (before change if applicable)
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Fishery Selectivity (after change if applicable)
+Input/Value
 \end{minipage} \\
@@
 \endhead
-1 & 0.04 & 0.15 & 0.07 & 0.02 \\
-2 & 0.25 & 0.5 & 0.17 & 0.05 \\
-3 & 0.60 & 0.9 & 0.36 & 0.12 \\
-4 & 0.77 & 1.4 & 0.61 & 0.27 \\
-5 & 0.85 & 2.0 & 0.81 & 0.50 \\
-6 & 0.92 & 2.6 & 0.92 & 0.74 \\
-7 & 1.00 & 3.2 & 0.97 & 0.89 \\
-8 & 1.00 & 4.1 & 0.99 & 0.96 \\
-9 & 1.00 & 5.9 & 1.00 & 0.99 \\
-10+ & 1.00 & 9.0 & 1.00 & 1.00 \\
+\emph{ObsModel} & Link function and assumed distribution & c(10,2) \\
+FieldCOnfig & Specified spatial and/or spatio-temporal variation in predictors & c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1) \\
+RhoConfig & Specifying whether intercepts or spatio-temporal variation is structured among time intervals & c(Beta1=3, Beta2=3, Epsilon1=0, Epsilon2=4) \\
+X1\_formula & Right-sided formula affecting the 1st linear predictor & X1\_formula = \textasciitilde{} poly(Temp, degree=2 ) \\
+X2\_formula & Right-sided formula affecting the 2nd linear predictor & X2\_formula = \textasciitilde{} poly(Temp, degree=2 ) + poly(Habitat, degree=2 ) \\
 \bottomrule
@@
 
-\pagebreak
-
-Table 2. Naming convention and details of the data-limited methods evaluated.
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.36}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.64}}@{}}
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-Method
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Details
-\end{minipage} \\
-\midrule
-\endhead
-Ismooth & \(C_{targ,y+1:y+2} =\overline{C}_{3,y}(e^{\lambda})\) where \(\overline{C}_{3,y}\) is the most recent three year average; \(\overline{C}_{3,y} = \frac{1}{3}\sum_{t=1}^{t=3}C_{y-t}\) and \(\lambda\) is the slope of a log linear regression of a LOESS-smoothed average index of abundance (spring and fall) with span = 0.3: \(\hat{I}_y=loess(\hat{I}_y)\) and \(LN(\widehat{I_y}) = b + \lambda y\) \\
-Islope & \(C_{targ,y+1:y+2} = 0.8\overline{C}_{5,y}(1+0.4e^{\lambda})\) where \(\overline{C}_{5,y}\) is the most recent five-year average catch through year \(y-1\): \(\overline{C}_{5,y} = \frac{1}{5}\sum_{t=1}^{t=5}C_{y-t}\) and \(\lambda\) is the slope of a log-linear regression of the most recent five years of the averaged index.
-\\
-Itarget & \(C_{targ,y+1:y+2}=\left[0.5C_{ref}\left(\frac{\overline{I}_{5,y}-I_{thresh}}{I_{target}-I_{thresh}}\right)\right]\) \(\overline{I}_{5,y} \geq I_{thresh}\) \(C_{targ,y+1:y+2}=\left[0.5C_{ref}\left(\frac{\overline{I}_{5,y}}{I_{thresh}}\right)^{2}\right]\) \(\overline{I}_{5,y} < I_{thresh}\); \(C_{ref}\) is the average catch over the reference period (years 26 through 50): \(C_{ref} = \frac{1}{25}\sum_{y=26}^{y=50}C_y\); \(I_{target}\) is 1.5 times the average index over the reference period: \(I_{target} = \frac{1}{25}\sum_{y=26}^{y=50}\overline{I}_y\); \(I_{thresh}\) = 0.8 \(I_{target}\), and is the most recent five year average of the combined spring and fall index: \(\overline{I}_{5,y} = \frac{1}{5}\sum_{t=1}^{t=5}\overline{I}_{y-t+1}\) \\
-Skate & \(C_{targ,y+1:y+2}= F_{rel} \overline{I}_{3,y}\) where \(F_{rel} =median \left(\frac{\overline{C}_{3,\textbf{Y}}}{\overline{I}_{3,\textbf{Y}}} \right)\) is the median relative fishing mortality rate calculated using a 3 year moving average of the catch and average survey index across all available years (\(\textbf{Y}\)): \(\overline{C}_{3,y} = \frac{1}{3} \sum_{t=1}^{t=3} C_{y-t}\) and \(\overline{I}_{3,y} = \frac{1}{3} \sum_{t=1}^{t=3} I_{y-t+1}\) \\
-An Index Method (AIM) & AIM first calculates the annual relative \(F\): \(F_{rel,y}=\frac{C_y}{\frac{1}{3}\sum_{t=1}^{t=3}\overline{I}_{y-t+1}}\) and the annual replacement ratio: \(\Psi_y = \frac{\overline{I}_y}{\frac{1}{5}\sum_{t=1}^{t=5}\overline{I}_{y-t}}\).
-These values are used in a regression: \(LN(\Psi_y)= b + \lambda LN(F_{rel,y})\) to determine \(F_{rel,*}\), which is the value of \(F_{rel,y}\) where the predicted \(\Psi=1\) or \(LN(\Psi)=0\).
-\(F_{rel,*}\) is called either the ``stable'' or ``replacement'' \(F\), and is used to calculate the target catch: \(C_{targ,y+1:y+2} = \overline{I}_yF_{rel,*}\).
-\\
-Dynamic Linear Model (DynLin) & @Langan2021DLM. \\
-Expanded survey biomass method 1 \(F_{40\%}\) (ES-FSPR) & \(C_{targ,y+1:y+2}=B_{\bar{I},y}\mu_{targ}\) where \(B_{\bar{I}}\) is the average of estimated fully-selected biomass from each survey: \(B_{\bar{I},y}=\frac{1}{2}\left(\frac{I_{spr,y}}{q_{spr}} +\frac{I_{fall,y-1}}{q_{fall}}\right)\) and target exploitation fraction, \(\mu_{targ}\) is calculated as: \(\mu_{targ}=\frac{F_{targ}}{Z_{targ}}\left(1-e^{-Z_{targ}}\right)\); \(F_{targ}=F_{40\%}\) and \(Z_{targ}=F_{targ}+M\) \\
-Expanded survey biomass method 2 \(F=\) AIM replacement (ES-Fstable) & Same as the above expanded survey method, but with \(\mu_{targ}\) equal to the stable exploitation fraction \(F_{rel,*}\) calculated using the AIM approach (see above).
-\\
-Expanded survey biomass method 3 \(F=M\) (ES-FM) & Same as the above expanded survey methods, but with the target exploitation rate set to the assumed \(M\): \(F_{targ}=M\). \\
-Expanded survey biomass method 4 \(F=\) recent average (ES-Frecent) & Same as the above expanded survey methods, but with the target exploitation fraction set to the most recent three year average exploitation fraction: \({\mu_{targ}}=\frac{\sum_{y-2}^{y}\mu_y}{3}\) \(\mu_{y}=\frac{C_{y-1}}{B_{\bar{I},y}}\) \\
-Catch curve Method 1 \(F_{40\%}\) (CC-FSPR) & \(C_{targ,y+1:y+2}= \frac{F_{targ}}{Z_{avg,y}}B_{cc,y}\left(1-e^{-Z_{avg,y}}\right)\) where \(B_{cc}\) is the estimated biomass: \(B_{cc,y} = \frac{C_{y-1}}{\frac{F_{avg,y}}{Z_{avg,y}}{\left(1-e^{-Z_{avg,y}}\right)}}\) with \(Z_{avg,y} = \frac{Z_{spring, y}+Z_{fall,y-1}}{2}\); \(F_{avg,y-1} = Z_{avg,y-1} - M\) and, \(F_{targ}=F_{40\%}\).
-Survey catch at age used in catch curve to estimate \(Z\).
-\\
-Catch curve Method 2 \(M\) (CC-FM) & Same as catch curve method 1 above, but with \(F_{targ}=M\). \\
-Ensemble & Median of catch advice provided by AIM, CC-FSPR, ES-Frecent, ES-FSPR, Islope, Itarget, Ismooth, and Skate methods. \\
-\bottomrule
-\end{longtable}
-
-\pagebreak
-
-Table 3. Summary of the scenarios evaluated within the study design.
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.50}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.50}}@{}}
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-Factors
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Variants
-\end{minipage} \\
-\midrule
-\endhead
-retrospective source & catch or natural mortality \\
-fishing history & \(F_{MSY}\) in second half of base period or overfishing throughout base period (2.5x\(F_{MSY}\)) \\
-fishery selectivity blocks & constant selectivity or selectivity changes in second half of base period \\
-catch advice multiplier & applied as is from DLM (1) or reduced from DLM (0.75) \\
-\bottomrule
-\end{longtable}
-
-\pagebreak
-
-\hypertarget{list-of-figures}{%
-\subsection{List of Figures}\label{list-of-figures}}
-
-Figure 1.
-Inner quartiles and medians for all performance measures across all scenarios and runs for each method.
-Vertical lines are shown at a value of 1 for the performance measures that are relative to the MSY reference points (A,B,C).
-
-Figure 2.
-Relationship between long-term average spawning biomass and average catch (relative to MSY levels) for each method.
-Each point represents the median for a given scenario, separated by the source of the retrospective pattern (catch or M).
-
-Figure 3.
-Median performance measures for each method, separated by the source of the retrospective error (catch = black, M = gray) and the exploitation history in the base period (always overfishing at \(2.5 x F_{MSY}\) (circle), or \(F\) reduced to \(F_{MSY}\) during base period (triangle)).
-Vertical lines are shown at a value of 1 for the performance measures that are relative to the MSY reference points (A,B,C).
-
-Figure 4.
-Median \(F / F_{MSY}\) for each method, with results separated by the exploitation history in the base period (always overfishing at \(2.5 x F_{MSY}\) (circle), or \(F\) reduced to \(F_{MSY}\) during base period (triangle)) showing A) short- (gray) versus long-term (black) values, and B) with (black) or without (gray) a buffer applied when setting the catch (catch multiplier = 0.75 or 1).
-
-Figure 5.
-Relationship between long-term average catch and spawning stock biomass relative to their reference points by method.
-Each point represents the average for years 21-40 in the feedback period for a single iteration of a scenario.
-The scenario shown is where catch was the source of the retrospective pattern with \(F\) reduced to \(F_{MSY}\) in the second half of the base period, there was a single selectivity block, and where no buffer was applied to the catch advice (catch multiplier = 1).
-
 \end{document}

### 2022-12-23 Manuscript/Draft1.Rmd
update manuscript

@@
 Fish stocks in this highly productive and economically important region are managed by the National Oceanic and Atmospheric Administration’s (NOAA) Northeast Fisheries Science Center (NEFSC) in Woods Hole, Massachusetts.
-Federal biologists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (cite survey paper).
+Federal scientists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (cite survey paper).
 The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf.
-The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits **just listed a few uses of survey change/others?**.
+The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits **just listed a few uses of survey.
+change/add others?**.
 
@@
 We are therefore interested in analyzing the impact of climate change on the accuracy of future stock assessment models as measured by NOAA’s ongoing bottom-trawl survey along the East coast.
+
+- say something about our study species? Past abundance history, biological characteristics, others? Maybe better suited in below section
 
@@
 
-Since we are modeling real species on the northeast continental shelf, we formulate the habitat and temperature components as follows.
+The following sections describe how we formulated the habitat and temperature components to model real species on the northeast continental shelf.
 
@@
 
-Species-specific habitat preferences were derived using the *lrren* tool from the R package *envi* **cite** to create a niche model for each species.
+Species-specific habitat preferences were derived using the *lrren* tool from the R package *envi* [@envi] to create a niche model for each species.
 The *lrren* tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors.
-We used bottom trawl point data from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow **cite trawl data?**.
-Depth and mean sediment size were used as our covariate predictors **cite**.
+We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow **cite trawl data?**.
+We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences.
+Depth and mean sediment size were used as our covariate predictors.
+Estimated depth for the region was obtained from FVCOM [@chen2006unstructured].
+The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method **cite arcmap or more?** using point data collected by the United States Geologic Survey (USGS) [@mcmullen2005usgs].
 Since the values in $Hab^2_{J,s}$ are required to be between 0 and 1, we transform the spatial estimates from *lrren* to fall between these bounds.
-See Figure ???
-for a visual representation of this process being applied to Cod.
+See Figure \@ref(fig:hab-plot1) for a visual representation of this process being applied to Cod.
 Figure \@ref(fig:hab-plot) depicts habitat preferences $Hab^2_{J,s}$ for each species.
+
+```{r hab-plot1, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Visual representation of niche model for Cod."}
+knitr::include_graphics(c("Images/hab_snip.png"))
+```
 
@@
 We chose these values by combining information in the literature with temperatures recorded in the bottom trawl survey.
-Weekly estimated temperature data for the region for 2012 was obtained from FVCOM **cite**.
+Weekly estimated temperature data for the region for 2012 was obtained from FVCOM [@chen2006unstructured].
 We chose 2012 because the data displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values.
@@
 In equation (1), $Hab^2_{J,s}$ is constant for the duration of the simulation, while $Tol_{c,s,wk}$ changes each week.
-Using a temperature gradient that repeats each year produces the same spatial preferences in a given week each year, which results in consistent spatial biomass patterns.
-Scenarios where the temperature increases over time creates spatial preferences that evolve as the water warms, which creates spatial biomass patterns that shift in a given week over the duration of the simulation.
+Using a temperature gradient that repeats every 52 weeks produces the same spatial preferences in a given week each year, resulting in consistent spatial biomass patterns.
+Scenarios where the temperature increases over time creates spatial preferences that evolve as the water warms, producing spatial biomass patterns that shift in a given week over the duration of the simulation.
+Thus, stratified random samples in scenarios with a repeating temperature pattern will have constant survey catchability and availability over time, which may not be true for increasing temperature scenarios due to evolving spatial preferences.
 
@@
 
-We follow the advice given in **cite Thorson 2019 Guidance...** to build VAST models to estimate biomass on George's Bank using stratified mean samples from our model output.
+We follow the advice given in [@thorson2019guidance] to build VAST models to estimate biomass on George's Bank using stratified mean samples from our model output.
 In addition to exploring different link functions and assumed distributions, our VAST model-building process included testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
 We also carried out the same process both including covariates in our model as well as running models without covariate information.
-We considered covaraites in the form of dyamic temperature values and/or static habitat values from our population model.
+We considered covariates in the form of dynamic temperature values and/or static habitat values from our population model.
 When using covariates we ultimately decided to provide the most information to the model by including both covariates for both linear predictors.
@@
 Our goal is to determine indexing approaches and settings that are robust to future environmental conditions and resulting spatial biomass patterns.
-An underlying assumption in all indexing methods that individual random samples combine to accurately represent true abudance by a) sampling all strata in which the population exists and b) low enough noise level in the samples to allow for a discernable pattern.
+An underlying assumption in all indexing methods that individual random samples combine to accurately represent true abundance by a) sampling all strata in which the population exists and b) low enough noise level in the samples to allow for a discernible pattern.
 This assumption can be questioned given enough noise in the sampling process **cite?** and/or shifting spatial preferences driven by climate change causing a population to move into a previously uninhabited strata.
@@
 
-When combining population trends for each species, differing temperature scenarios, and sampling possibilities (noise, strata, covariates) there are a number of scenario combinations to consider.
-The columnbs in Table COMBOS show the possible choices that define each scenario.
+When combining population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, covariates) there are a large number of scenario combinations to consider.
+The columns in Table COMBOS show the choices that define each scenario.
 
 <div align="center">Table COMBOS Each index estimate chooses one condition from each of the following columns.
-There are $3*3*2*2*2*2=144$ VAST model combinations and $3*3*2*2*2=72$ stratified mean estimates.</div>
+There are $3*3*2*2*2*2*2=288$ VAST model combinations and $3*3*2*2*2=144$ stratified mean estimates.</div>
 
@@
 tabl <- "
-|	Species	|	Population Trend | Temperature Scenario | Strata Included | Noise in Data | Covariates (VAST only) 
-|:--	|:-- |:-- |:-- |:-- |:--- 
-|Yellowtail | Increasing | Repeating | All strata | No Noise| No Covariates
-|Cod | Constant | Increasing 5$^{\\circ}$| Subset | Yes Noise| Temp + Habitat
-|Haddock | Decreasing | | | |
+|	Species	|	Population Trend | Temperature Scenario | Strata Included | Noise in Data | Covariates (VAST only) | Season
+|:--	|:-- |:-- |:-- |:-- |:--- |:--
+|Yellowtail | Increasing | Repeating | All strata | No Noise| No Covariates | Spring
+|Cod | Constant | Increasing 5$^{\\circ}$| Subset | Yes Noise| Temp + Habitat | Fall
+|Haddock | Decreasing | | | | |
 "
@@
 \section{Results}
+
+The goal of our project was to analyze how well different contemporary population indexing methods can track population trends under a host of conditions, as depicted in Table COMBOS.
+Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year **cite**.
+For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
+We consider all possible scenario combinations for Yellowtail Flounder to provide a more comprehensive analysis of population indexing methods.
 
@@
 ## Discussion
-
-A range of data-limited methods for setting catch advice were evaluated for stocks where assessment models may be rejected due to strong, positive retrospective patterns.
-A method was considered to perform well if it limited overfishing without resulting in light exploitation rates ($F << F_{MSY}$), thereby allowing depleted stocks to recover to $SSB_{MSY}$ (or for healthy stocks to remain there), and for high and stable catches (close to $MSY$).
-
-Overall, none of the methods evaluated performed best across the scenarios exploring the different sources of the retrospective pattern (unreported catch or increasing $M$) and different levels of historical fishing intensity.
-A number of methods did perform well in many cases, however, while others performed consistently poorly, resulting in frequent and intense overfishing ($F>>F_{MSY}$).
-We performed simulations for a couple of scenarios with no source of retrospective patterns and found the expected result that all DLMs and the SCAA performed better ($SSB$, $F$, and catch were all closer to the $MSY$ reference points) than when either source of retrospective patterns was present.
-Due to the focus of this study, we did not examine the no retrospective source in detail and do not comment on it further.
-
-Currently, in the Northeast U.S., if an assessment model is rejected due to a large rho value in $SSB$, the catch advice from that model is ignored and some data-limited approach is used.
-However, the rho-adjusted SCAA model performed better than a number of the alternatives explored here.
-Therefore, there should not necessarily be an expectation that a data-limited method will perform better than the rejected assessment model.
-The SCAA only resulted in high exploitation rates ($F >> F_{MSY}$) when unreported catch was the source of the retrospective pattern and for the scenario where $F=F_{MSY}$ at the end of the base period that left the stock in relatively good condition ($SSB \sim SSB_{MSY}$).
-In contrast, this method was particularly effective when the stock was depleted and there was unreported catch.
-When $M$ was the source of the retrospective pattern, the rho-adjusted SCAA method typically resulted in light exploitation rates, on average.
-The light exploitation rates in these cases were likely driven by the combination of using a rho-adjustment, but also using the lower $M$ from the beginning of the base period rather than the higher $M$ that occurred during the feedback period.
-Using an $M$ value that is too low in a stock assessment will typically bias estimates of biomass and reference points too low, resulting in catch advice that is below target levels [@Johnsonetal2014; @Puntetal2021M].
-The consequences of using a value for $M$ that is too low versus too high is also asymmetrical [@Johnsonetal2014], with negative consequences being more severe when $M$ is assumed too high than low, and the results here are consistent with these previous conclusions.
-
-The methods that adjusted recent average catches based on trends in the survey (Ismooth and Islope) performed well overall in terms of catch, stock status, and variation in catch.
-The method using the expanded survey biomass with the recent exploitation rate (ES-Frecent) also performed well and similarly to Ismooth.
-The performance of these methods was also generally robust among scenarios, with the exception of when there were unreported catches and the stock was depleted (see below).
-The generally positive performance of these methods was consistent with @Hilbornetal2002 and @CoxKronlund2008, both of which evaluated a variant of a "hold-steady" DLM.
-In the case of @Hilbornetal2002, the "hold-steady" DLM policy was designed to adjust catches in order to keep rockfish (*Sebastes spp.*) populations at recently observed index levels, and did so by functioning as a constant escapement harvest control rule where target catches were set to zero below some pre-specified index level.
-In the variant used by @CoxKronlund2008, catches were adjusted to maintain a sablefish (*Anoplopoma fimbria*) population at a pre-specified index level thought to be sustainable and desirable in terms of meeting fishery objectives (e.g., high catch), but never permitted target catches of zero and so functioned as a constant exploitation rate control rule.
-The "hold-steady" DLM of @CoxKronlund2008 performed similarly in terms of catch, stock depletion, and variation in catch, as a constant exploitation rate policy where target catch was specified as the product of desired exploitation rate and an estimate of biomass from a SCAA model.
-This result was robust to uncertainty in initial stock status and steepness [@CoxKronlund2008].
-The SCAA model was always correctly specified (i.e., expected to produce unbiased estimates on average), however, and no comparison to the results of this research in the presence of retrospective patterns is possible [@CoxKronlund2008].
-The "hold-steady" policy of @Hilbornetal2002 performed similarly to or better in terms of catch and stock status than other harvest control rules that relied on assessment estimates of biomass (i.e., 40:10 and constant *F*).
-The performance of the "hold-steady" DLM was also more robust to uncertainty in steepness and to the presence of unreported catch [@Hilbornetal2002].
-The performance of the two harvest policies that relied on assessment estimates of biomass (i.e., constant exploitation rate and a "40:10" biomass-based policy) also degraded when the estimates of biomass were biased, which is an issue that does not effect the "hold-steady" DLM [@Hilbornetal2002].
-The bias in the assessment estimates considered in @Hilbornetal2002 were not necessarily induced by a retrospective pattern, however, and no consideration of making a rho-adjustment was possible in that study.
-
-The Ismooth method is currently used to set catches for Georges Bank cod [@nefsc19] and red hake (*Urophycis chuss*; @nefsc20).
-Variations of the ES-Frecent have been used for witch flounder and Georges Bank yellowtail flounder.
-While the findings here generally support the continued use of the Ismooth and ES-Frecent methods, they may not be well suited for depleted stocks where unreported catches are believed to be an issue.
-The Ismooth, Islope, and ES-Frecent DLMs produced high $Fs$ and limited stock recovery with unreported catches and when the stock was depleted.
-While @Hilbornetal2002 and @CoxKronlund2008 did not reach the same conclusion about the "hold-steady" DLM, those studies did not consider initial levels of depletion as low as in this study.
-These results highlight the importance of accurate catch reporting, as unreported catch can create a negative feedback loop with perpetually high $Fs$ being produced by a management system that seemingly should result in sustainable catch advice.
-
-Three methods were consistently risk-averse across scenarios, limiting the frequency and magnitude of overfishing and resulting in high stock biomass.
-These methods were the two catch curve options (CC-FM and CC-FSPR) and DynLin.
-The catch curve methods produced a wider range of average catches across scenarios, and also had greater interannual variability in catches compared to DynLin.
-While the lower exploitation rates from these approaches may be undesirable due to forgone yield, there may be circumstances where they are preferred.
-For example, for stocks that are believed to be heavily depleted, low exploitation rates would allow for a more rapid recovery.
-
-A number of methods performed poorly, particularly when catches were unreported.
-These methods include three of the expanded survey biomass approaches (ES-Fstable, ES-FM, ES-FSPR), AIM, and Skate.
-The AIM model has been widely used across stocks in the region [@nefsc02a; @nefsc05; @nefsc08], although there is a decreasing trend in its use across model resistant stocks [@nefsc19].
-The findings here suggest that alternative approaches should be considered in cases where AIM is still used and there is concern over unreported catches.
-The Skate method is used to manage the skate complex in the Northeast U.S.
-(a group of seven co-managed species).
-Interestingly, six of the seven species are considered in good condition with high survey biomass indices in recent years [@nefmc20].
-That the Skate method performed poorly in our analysis but performs well for the skate complex illustrates how the performance of methods in this analysis may be sensitive to the scenarios and species life history considered.
-As may be the case for the Skate method, the performance of some methods may depend on the condition of the stock when the method is first applied, and less so on life-history.
-Therefore, care is needed when trying to generalize these results across stocks that may have different life histories, exploitation histories, and without unreported catches or increases in $M$.
-
-In addition to the analytical differences among the thirteen DLMs, most of the DLMs and control rules had multiple options that could be adjusted to make them more or less risk averse.
-DynLin had a large number of user defined decision points.
-Given the large range of options already explored in the study, one suite of options was selected for each DLM-control rule and kept constant for all simulations.
-Further studies could explore the different options within an individual DLM to understand how they might affect performance.
-
-Many other data-limited methods exist for setting catch advice that were not included in this evaluation, and they vary widely in complexity, data inputs, and assumptions required [e.g., @carruthers2018dlm].
-Length based methods were not evaluated to keep the overall number of methods tractable, and due to the availability of age based information in the region.
-Methods that require only catch data or snap shots of survey data were not considered due to the availability of the relatively long and contiguous Northeast Fisheries Science Center's spring and fall, coastwide bottom trawl surveys, and the fact that "catch only" methods have been shown to perform poorly [e.g., @carruthers2014eval].
-Complete catch histories are not available for stocks in the region (i.e., from the inception of fishing).
-Consequently, methods that required complete catch histories or required assumptions about relative depletion [e.g., DCAC in @maccall2009dca; DB-SRA in @dick2022dsra] were also omitted from consideration.
-The need for short run-times and the desire for methods that could be reviewed quickly prevented the use of modern state-space production models such as SPiCT [@pedersen2017spict] and JABBA [@winker2018jabba].
-
-The SCAA was confronted with inconsistent data in this study, while the DLMs typically used only a single source of data and thus did not encounter inconsistencies.
-A recent examination of the data used in assessments in this region similarly found inconsistencies in data streams even before modeling.
-@wiedenmann2022strange found a negative relationship between relative F (catch/survey) and survey Z for stocks with strong retrospective patterns but the expected positive relationship for stocks without a retrospective pattern.
-It is exactly this sort of tension that creates retrospective patterns in integrated models, but is not found in DLMs that only use one type of data.
-
-Despite conducting hundreds of thousands of simulations, there are still limitations to our study.
-We only examined one life history representative of groundfish in the region.
-We acknowledge that best practice is to select a DLM for a specific life history and fishery condition [e.g., @fischer2020dlm].
-As is typically the case with large simulation studies, we were not able to tune any of the DLMs or the SCAA in any given realization, which would occur in practice for an actual stock assessment.
-We also examined only scenarios that started with Mohn's rho values near 0.5 for spawning stock biomass.
-This is a strong retrospective pattern, but some stocks in the region have even stronger retrospectives.
-Performance of the DLMs and SCAA would be expected to degrade with stronger retrospectives, but by how much is still an open area for research.
-Similarly, sources of retrospective patterns that create different relationships between the true values and estimated values should also be explored [see @deroba2014retro].
-To make the results interpretable, we only examined a single source for the retrospective pattern at a time.
-In reality, there may be more than one factor leading to an observed retrospective pattern.
-How the multiple sources would interact to influence performance is another topic for future research.
-Development of harvest control rules specifically for situations where retrospective patterns are found in age-based assessments would also be beneficial.
-The large number of scenarios examined and the large number of realizations gives us confidence that our results are meaningful in general, but that the performance of any of the DLMs may differ in actual practice.
-
-An interesting finding of this study is the linear versus diffuse patterns between $SSB$ and catch across methods.
-These patterns have implications for the trade-offs among methods, with linear relationships resulting in more consistent exploitation rates across stock sizes.
-Therefore, these methods have higher certainty of a given catch at a given stock size.
-However, they also tended to result in lower stock sizes, on average, across methods.
-The more diffuse relationships resulted in more variable exploitation rates across stock sizes, with some situations where the population biomass was quite high but the catch was low (relative to MSY), resulting in a very low $F$.
-The reasons behind these different patterns remain unclear, and future work to explore these patterns is warranted.
-
-One of the reasons for the difference in performance between the catch and natural mortality retrospective sources was how the reference points were calculated.
-In all cases, the initial conditions, including the natural mortality rate, were used to compute the reference points.
-This decision was made based on the fact that the increase in natural mortality was assumed to be unknown in the simulations.
-If the increase in natural mortality was known, the age-structured assessments would have accounted for it, different reference points might have been computed [@legault2016increaseM] and there may not have been a retrospective pattern at all [@legault2020rose], and no need to consider alternative DLMs.
-The reference points for the increased $M$ scenarios would have been different if they were computed using the values from the final year of the base period, but the overall conclusions regarding the different DLMs would not change as this just results in a rescaling of the axis.
-These results are not shown to reduce confusion regarding the simulations.
-
-Closed-loop simulation is a common tool for examining performance of catch advice from various stock assessment approaches in a feedback setting.
-It is often used as part of a full management strategy evaluation when working with stakeholders to develop management regulations that make trade offs between near term and long term catches, risk to the fish population, and mixed-fleet allocations [@carruthers2016simpleMPs; @goethel2019mse; @harlyan2019hcr].
-We did not conduct a full management strategy evaluation with stakeholder input [@goethel2019stakeholder], but see that as a fruitful next step that could build on the conclusions from our closed-loop work.
-Using a generic groundfish life-history and monitoring standard performance metrics related to stock status and catch stability, we were able to cull the herd of potential DLMs and we would not carry the consistent poor performers forward for further study.
-The wide range of expertise reflected in the authorship was by design so that the simulation specifications and performance metrics were broadly useful.
-Before undertaking a full management strategy evaluation and engaging regional stakeholders, we would want to select a specific stock and jointly identify specific management regulations to be tested [@deroba2019dream].
-Results of this work have been presented to both local fishery management councils, with generally positive feedback about the utility of the conclusions for identifying appropriate model approaches when an SCAA is rejected.
-Our work was similar to all other closed-loop simulations in that it was designed to address a specific situation, including much recent work comparing the performance of data-limited and data rich assessment approaches [e.g., @fulton2016datarich; @sagarese2019dlm; @bouch2020datapoor; @li2022dlm].
-
-This study is a first attempt to identify suitable methods for setting catch advice when stock assessment models are rejected due to large, positive retrospective patterns.
-Although no single method performed best across scenarios, a number of generally suitable and unsuitable methods were identified under specific conditions.
-The results of this work can help scientists and managers select a subset of possible options for consideration to set catch advice when assessment models are rejected.
-The approach developed here can, and should be expanded to consider other cases not explored here, as performance of individual methods are very likely case-dependent.
 
@@
 
+```{r YT_results,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
+library("knitr")
+library("kableExtra")
+library("dplyr")
+library("readr")
+library("tidyr")
+library("forcats")
+
+result_table = read_csv("Yellowtail_results_final.csv")
+
+summary_tab <- result_table %>%
+
+# Group the table according to Parameter, Scenario
+  group_by(Season, Scenario)%>%
+# Change the orientation of the table so that Species become the variable headings
+  mutate(Scenario = factor(Scenario)) %>%
+  arrange(Scenario) %>%
+  select(Scenario, everything())
+
+# Create table to present the data]
+Fig_5 <- kable(summary_tab)
+
+```
+```{r, results='asis'}
+Fig_5
+```
+
 <!-- ```{r PMsallscenarios, echo=FALSE, message=FALSE, warnings=FALSE, out.width='7in', fig.cap=figcap1} -->

### 2022-12-23 Manuscript/Draft1.tex
update manuscript

@@
 Fish stocks in this highly productive and economically important region are managed by the National Oceanic and Atmospheric Administration's (NOAA) Northeast Fisheries Science Center (NEFSC) in Woods Hole, Massachusetts.
-Federal biologists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (cite survey paper).
+Federal scientists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (cite survey paper).
 The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf.
-The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits \textbf{just listed a few uses of survey change/others?}.
+The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits \textbf{just listed a few uses of survey.
+change/add others?}.
 
@@
 \begin{itemize}
-\tightlist
 \item
@@
 We are therefore interested in analyzing the impact of climate change on the accuracy of future stock assessment models as measured by NOAA's ongoing bottom-trawl survey along the East coast.
+\item
+  say something about our study species? Past abundance history, biological characteristics, others? Maybe better suited in below section
 \end{itemize}
@@
 
-Species-specific habitat preferences were derived using the \emph{lrren} tool from the R package \emph{envi} \textbf{cite} to create a niche model for each species.
+Species-specific habitat preferences were derived using the \emph{lrren} tool from the R package \emph{envi} {[}@envi{]} to create a niche model for each species.
 The \emph{lrren} tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors.
-We used bottom trawl point data from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow \textbf{cite trawl data?}.
-Depth and mean sediment size were used as our covariate predictors \textbf{cite}.
+We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow \textbf{cite trawl data?}.
+We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences.
+Depth and mean sediment size were used as our covariate predictors.
+Estimated depth for the region was obtained from FVCOM {[}@chen2006unstructured{]}.
+The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method \textbf{cite arcmap or more?} using point data collected by the United States Geologic Survey (USGS) {[}@mcmullen2005usgs{]}.
 Since the values in \(Hab^2_{J,s}\) are required to be between 0 and 1, we transform the spatial estimates from \emph{lrren} to fall between these bounds.
-See Figure ???
-for a visual representation of this process being applied to Cod.
+See Figure \ref{fig:hab-plot1} for a visual representation of this process being applied to Cod.
 Figure \ref{fig:hab-plot} depicts habitat preferences \(Hab^2_{J,s}\) for each species.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{Images/hab_snip}
+
+}
+
+\caption{Visual representation of niche model for Cod.}\label{fig:hab-plot1}
+\end{figure}
 
@@
 We chose these values by combining information in the literature with temperatures recorded in the bottom trawl survey.
-Weekly estimated temperature data for the region for 2012 was obtained from FVCOM \textbf{cite}.
+Weekly estimated temperature data for the region for 2012 was obtained from FVCOM {[}@chen2006unstructured{]}.
 We chose 2012 because the data displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values.
@@
 In equation (1), \(Hab^2_{J,s}\) is constant for the duration of the simulation, while \(Tol_{c,s,wk}\) changes each week.
-Using a temperature gradient that repeats each year produces the same spatial preferences in a given week each year, which results in consistent spatial biomass patterns.
-Scenarios where the temperature increases over time creates spatial preferences that evolve as the water warms, which creates spatial biomass patterns that shift in a given week over the duration of the simulation.
+Using a temperature gradient that repeats every 52 weeks produces the same spatial preferences in a given week each year, resulting in consistent spatial biomass patterns.
+Scenarios where the temperature increases over time creates spatial preferences that evolve as the water warms, producing spatial biomass patterns that shift in a given week over the duration of the simulation.
+Thus, stratified random samples in scenarios with a repeating temperature pattern will have constant survey catchability and availability over time, which may not be true for increasing temperature scenarios due to evolving spatial preferences.
 
@@
 
-We follow the advice given in \textbf{cite Thorson 2019 Guidance\ldots{}} to build VAST models to estimate biomass on George's Bank using stratified mean samples from our model output.
+We follow the advice given in {[}@thorson2019guidance{]} to build VAST models to estimate biomass on George's Bank using stratified mean samples from our model output.
 In addition to exploring different link functions and assumed distributions, our VAST model-building process included testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
 We also carried out the same process both including covariates in our model as well as running models without covariate information.
-We considered covaraites in the form of dyamic temperature values and/or static habitat values from our population model.
+We considered covariates in the form of dynamic temperature values and/or static habitat values from our population model.
 When using covariates we ultimately decided to provide the most information to the model by including both covariates for both linear predictors.
@@
 Our goal is to determine indexing approaches and settings that are robust to future environmental conditions and resulting spatial biomass patterns.
-An underlying assumption in all indexing methods that individual random samples combine to accurately represent true abudance by a) sampling all strata in which the population exists and b) low enough noise level in the samples to allow for a discernable pattern.
+An underlying assumption in all indexing methods that individual random samples combine to accurately represent true abundance by a) sampling all strata in which the population exists and b) low enough noise level in the samples to allow for a discernible pattern.
 This assumption can be questioned given enough noise in the sampling process \textbf{cite?} and/or shifting spatial preferences driven by climate change causing a population to move into a previously uninhabited strata.
@@
 
-When combining population trends for each species, differing temperature scenarios, and sampling possibilities (noise, strata, covariates) there are a number of scenario combinations to consider.
-The columnbs in Table COMBOS show the possible choices that define each scenario.
+When combining population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, covariates) there are a large number of scenario combinations to consider.
+The columns in Table COMBOS show the choices that define each scenario.
 
-Table COMBOS Each index estimate chooses one condition from each of the following columns. There are \(3*3*2*2*2*2=144\) VAST model combinations and \(3*3*2*2*2=72\) stratified mean estimates.
+Table COMBOS Each index estimate chooses one condition from each of the following columns. There are \(3*3*2*2*2*2*2=288\) VAST model combinations and \(3*3*2*2*2=144\) stratified mean estimates.
 
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.16}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.16}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.16}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.16}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.16}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.21}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.14}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.14}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.14}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.14}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.14}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.18}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.14}}@{}}
 \toprule
@@
 Covariates (VAST only)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Season
 \end{minipage} \\
@@
 \endhead
-Yellowtail & Increasing & Repeating & All strata & No Noise & No Covariates \\
-Cod & Constant & Increasing 5\(^{\circ}\) & Subset & Yes Noise & Temp + Habitat \\
-Haddock & Decreasing & & & & \\
+Yellowtail & Increasing & Repeating & All strata & No Noise & No Covariates & Spring \\
+Cod & Constant & Increasing 5\(^{\circ}\) & Subset & Yes Noise & Temp + Habitat & Fall \\
+Haddock & Decreasing & & & & & \\
 \bottomrule
@@
 
+The goal of our project was to analyze how well different contemporary population indexing methods can track population trends under a host of conditions, as depicted in Table COMBOS.
+Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year \textbf{cite}.
+For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
+We consider all possible scenario combinations for Yellowtail Flounder to provide a more comprehensive analysis of population indexing methods.
+
 \begin{table}
@@
 \centering
-\begin{tabular}[t]{l|l|l|l|l|l|l|l|l|l|l|l}
+\begin{tabular}[t]{l|l|l|l|l|l|l|l|l|l|l}
 \hline
-X & X.1 & X.2 & X.3 & X.4 & Constant.Population & X.5 & Increasing.Population & X.6 & Decreasing.Population & X.7 & X.8\\
+X & X.1 & X.2 & X.3 & X.4 & Constant.Population & X.5 & Increasing.Population & X.6 & Decreasing.Population & X.7\\
 \hline
-Temp & Covariate & Strata & Noise & season & Stratified Mean & VAST Estimate & Stratified Mean & VAST Estimate & Stratified Mean & VAST Estimate & \\
+Temp & Covariate & Strata & Noise & season & Stratified Mean & VAST Estimate & Stratified Mean & VAST Estimate & Stratified Mean & VAST Estimate\\
 \hline
-const & no cov & all & no & spring & 0.21 & 0.11 & 0.16 & 0.13 & 0.23 & 0.08 & \\
+const & no cov & all & no & spring & 0.21 & 0.11 & 0.16 & 0.13 & 0.23 & 0.08\\
 \hline
-const & no cov & all & yes & spring & 0.25 & 0.16 & 0.22 & 0.21 & 0.27 & 0.11 & \\
+const & no cov & all & yes & spring & 0.25 & 0.16 & 0.22 & 0.21 & 0.27 & 0.11\\
 \hline
-const & w/ cov & all & no & spring & 0.21 & 0.07 & 0.16 & 0.06 & 0.23 & 0.06 & \\
+const & w/ cov & all & no & spring & 0.21 & 0.07 & 0.16 & 0.06 & 0.23 & 0.06\\
 \hline
-const & w/ cov & all & yes & spring & 0.25 & 0.08 & 0.22 & 0.07 & 0.27 & 0.07 & \\
+const & w/ cov & all & yes & spring & 0.25 & 0.08 & 0.22 & 0.07 & 0.27 & 0.07\\
 \hline
-const & no cov & all & no & fall & 0.32 & 0.68 & 0.34 & 0.36 & 0.41 & 0.81 & \\
+const & no cov & all & no & fall & 0.32 & 0.68 & 0.34 & 0.36 & 0.41 & 0.81\\
 \hline
-const & no cov & all & yes & fall & 0.31 & 0.77 & 0.46 & 0.44 & 0.44 & 1.09 & \\
+const & no cov & all & yes & fall & 0.31 & 0.77 & 0.46 & 0.44 & 0.44 & 1.09\\
 \hline
-const & w/ cov & all & no & fall & 0.32 & 0.08 & 0.34 & 0.08 & 0.41 & 0.08 & \\
+const & w/ cov & all & no & fall & 0.32 & 0.08 & 0.34 & 0.08 & 0.41 & 0.08\\
 \hline
-const & w/ cov & all & yes & fall & 0.31 & 0.11 & 0.46 & 0.17 & 0.44 & 0.18 & \\
+const & w/ cov & all & yes & fall & 0.31 & 0.11 & 0.46 & 0.17 & 0.44 & 0.18\\
 \hline
-const & no cov & reduced & no & spring & 0.27 & 0.19 & 0.2 & 0.15 & 0.25 & 0.19 & \\
+const & no cov & reduced & no & spring & 0.27 & 0.19 & 0.2 & 0.15 & 0.25 & 0.19\\
 \hline
-const & no cov & reduced & yes & spring & 0.26 & 0.15 & 0.22 & 0.11 & 0.3 & 0.16 & \\
+const & no cov & reduced & yes & spring & 0.26 & 0.15 & 0.22 & 0.11 & 0.3 & 0.16\\
 \hline
-const & w/ cov & reduced & no & spring & 0.27 & 0.19 & 0.2 & 0.17 & 0.25 & 0.19 & \\
+const & w/ cov & reduced & no & spring & 0.27 & 0.19 & 0.2 & 0.17 & 0.25 & 0.19\\
 \hline
-const & w/ cov & reduced & yes & spring & 0.26 & 0.14 & 0.22 & 0.13 & 0.3 & 0.16 & \\
+const & w/ cov & reduced & yes & spring & 0.26 & 0.14 & 0.22 & 0.13 & 0.3 & 0.16\\
 \hline
-const & no cov & reduced & no & fall & 0.47 & 0.25 & 0.41 & 0.11 & 0.55 & 0.24 & \\
+const & no cov & reduced & no & fall & 0.47 & 0.25 & 0.41 & 0.11 & 0.55 & 0.24\\
 \hline
-const & no cov & reduced & yes & fall & 0.49 & 0.36 & 0.46 & 0.21 & 0.53 & 0.36 & adding cov made worse\\
+const & no cov & reduced & yes & fall & 0.49 & 0.36 & 0.46 & 0.21 & 0.53 & 0.36\\
 \hline
-const & w/ cov & reduced & no & fall & 0.47 & 0.19 & 0.41 & 0.22 & 0.55 & 0.23 & \\
+const & w/ cov & reduced & no & fall & 0.47 & 0.19 & 0.41 & 0.22 & 0.55 & 0.23\\
 \hline
-const & w/ cov & reduced & yes & fall & 0.49 & 0.17 & 0.46 & 0.19 & 0.53 & 0.24 & \\
+const & w/ cov & reduced & yes & fall & 0.49 & 0.17 & 0.46 & 0.19 & 0.53 & 0.24\\
 \hline
-increasing & no cov & all & no & spring & 0.28 & 0.11 & 0.32 & 0.13 & 0.22 & 0.15 & \\
+increasing & no cov & all & no & spring & 0.28 & 0.11 & 0.32 & 0.13 & 0.22 & 0.15\\
 \hline
-increasing & no cov & all & yes & spring & 0.28 & 0.15 & 0.34 & 0.16 & 0.26 & 0.17 & cov really helped\\
+increasing & no cov & all & yes & spring & 0.28 & 0.15 & 0.34 & 0.16 & 0.26 & 0.17\\
 \hline
-increasing & w/ cov & all & no & spring & 0.28 & 0.06 & 0.32 & 0.07 & 0.22 & 0.07 & \\
+increasing & w/ cov & all & no & spring & 0.28 & 0.06 & 0.32 & 0.07 & 0.22 & 0.07\\
 \hline
-increasing & w/ cov & all & yes & spring & 0.28 & 0.12 & 0.34 & 0.1 & 0.26 & 0.1 & \\
+increasing & w/ cov & all & yes & spring & 0.28 & 0.12 & 0.34 & 0.1 & 0.26 & 0.1\\
 \hline
-increasing & no cov & all & no & fall & 0.51 & 1.26 & 0.3 & 0.66 & 0.28 & 1.06 & \\
+increasing & no cov & all & no & fall & 0.51 & 1.26 & 0.3 & 0.66 & 0.28 & 1.06\\
 \hline
-increasing & no cov & all & yes & fall & 0.5 & 1.38 & 0.39 & 0.71 & 0.25 & 1.1 & \\
+increasing & no cov & all & yes & fall & 0.5 & 1.38 & 0.39 & 0.71 & 0.25 & 1.1\\
 \hline
-increasing & w/ cov & all & no & fall & 0.51 & 0.23 & 0.3 & 0.21 & 0.28 & 0.15 & cov good, why no cov so bad???\\
+increasing & w/ cov & all & no & fall & 0.51 & 0.23 & 0.3 & 0.21 & 0.28 & 0.15\\
 \hline
-increasing & w/ cov & all & yes & fall & 0.5 & 0.28 & 0.3 & 0.37 & 0.25 & 0.2 & \\
+increasing & w/ cov & all & yes & fall & 0.5 & 0.28 & 0.3 & 0.37 & 0.25 & 0.2\\
 \hline
-increasing & no cov & reduced & no & spring & 0.31 & 0.17 & 0.4 & 0.33 & 0.32 & 0.14 & \\
+increasing & no cov & reduced & no & spring & 0.31 & 0.17 & 0.4 & 0.33 & 0.32 & 0.14\\
 \hline
-increasing & no cov & reduced & yes & spring & 0.29 & 0.19 & 0.41 & 0.26 & 0.29 & 0.15 & \\
+increasing & no cov & reduced & yes & spring & 0.29 & 0.19 & 0.41 & 0.26 & 0.29 & 0.15\\
 \hline
-increasing & w/ cov & reduced & no & spring & 0.31 & 0.17 & 0.4 & 0.3 & 0.32 & 0.22 & \\
+increasing & w/ cov & reduced & no & spring & 0.31 & 0.17 & 0.4 & 0.3 & 0.32 & 0.22\\
 \hline
-increasing & w/ cov & reduced & yes & spring & 0.29 & 0.15 & 0.41 & 0.25 & 0.29 & 0.21 & \\
+increasing & w/ cov & reduced & yes & spring & 0.29 & 0.15 & 0.41 & 0.25 & 0.29 & 0.21\\
 \hline
-increasing & no cov & reduced & no & fall & 0.64 & 0.75 & 0.7 & 0.49 & 0.54 & 0.6 & \\
+increasing & no cov & reduced & no & fall & 0.64 & 0.75 & 0.7 & 0.49 & 0.54 & 0.6\\
 \hline
-increasing & no cov & reduced & yes & fall & 0.66 & 0.89 & 0.69 & 0.48 & 0.53 & 0.62 & cov didnt do much\\
+increasing & no cov & reduced & yes & fall & 0.66 & 0.89 & 0.69 & 0.48 & 0.53 & 0.62\\
 \hline
-increasing & w/ cov & reduced & no & fall & 0.64 & 0.2 & 0.7 & 0.53 & 0.54 & 0.31 & \\
+increasing & w/ cov & reduced & no & fall & 0.64 & 0.2 & 0.7 & 0.53 & 0.54 & 0.31\\
 \hline
-increasing & w/ cov & reduced & yes & fall & 0.66 & 0.13 & 0.69 & 0.5 & 0.53 & 0.32 & \\
+increasing & w/ cov & reduced & yes & fall & 0.66 & 0.13 & 0.69 & 0.5 & 0.53 & 0.32\\
 \hline
@@
 \subsection{Discussion}\label{discussion}}
-
-A range of data-limited methods for setting catch advice were evaluated for stocks where assessment models may be rejected due to strong, positive retrospective patterns.
-A method was considered to perform well if it limited overfishing without resulting in light exploitation rates (\(F << F_{MSY}\)), thereby allowing depleted stocks to recover to \(SSB_{MSY}\) (or for healthy stocks to remain there), and for high and stable catches (close to \(MSY\)).
-
-Overall, none of the methods evaluated performed best across the scenarios exploring the different sources of the retrospective pattern (unreported catch or increasing \(M\)) and different levels of historical fishing intensity.
-A number of methods did perform well in many cases, however, while others performed consistently poorly, resulting in frequent and intense overfishing (\(F>>F_{MSY}\)).
-We performed simulations for a couple of scenarios with no source of retrospective patterns and found the expected result that all DLMs and the SCAA performed better (\(SSB\), \(F\), and catch were all closer to the \(MSY\) reference points) than when either source of retrospective patterns was present.
-Due to the focus of this study, we did not examine the no retrospective source in detail and do not comment on it further.
-
-Currently, in the Northeast U.S., if an assessment model is rejected due to a large rho value in \(SSB\), the catch advice from that model is ignored and some data-limited approach is used.
-However, the rho-adjusted SCAA model performed better than a number of the alternatives explored here.
-Therefore, there should not necessarily be an expectation that a data-limited method will perform better than the rejected assessment model.
-The SCAA only resulted in high exploitation rates (\(F >> F_{MSY}\)) when unreported catch was the source of the retrospective pattern and for the scenario where \(F=F_{MSY}\) at the end of the base period that left the stock in relatively good condition (\(SSB \sim SSB_{MSY}\)).
-In contrast, this method was particularly effective when the stock was depleted and there was unreported catch.
-When \(M\) was the source of the retrospective pattern, the rho-adjusted SCAA method typically resulted in light exploitation rates, on average.
-The light exploitation rates in these cases were likely driven by the combination of using a rho-adjustment, but also using the lower \(M\) from the beginning of the base period rather than the higher \(M\) that occurred during the feedback period.
-Using an \(M\) value that is too low in a stock assessment will typically bias estimates of biomass and reference points too low, resulting in catch advice that is below target levels {[}@Johnsonetal2014; @Puntetal2021M{]}.
-The consequences of using a value for \(M\) that is too low versus too high is also asymmetrical {[}@Johnsonetal2014{]}, with negative consequences being more severe when \(M\) is assumed too high than low, and the results here are consistent with these previous conclusions.
-
-The methods that adjusted recent average catches based on trends in the survey (Ismooth and Islope) performed well overall in terms of catch, stock status, and variation in catch.
-The method using the expanded survey biomass with the recent exploitation rate (ES-Frecent) also performed well and similarly to Ismooth.
-The performance of these methods was also generally robust among scenarios, with the exception of when there were unreported catches and the stock was depleted (see below).
-The generally positive performance of these methods was consistent with @Hilbornetal2002 and @CoxKronlund2008, both of which evaluated a variant of a ``hold-steady'' DLM.
-In the case of @Hilbornetal2002, the ``hold-steady'' DLM policy was designed to adjust catches in order to keep rockfish (\emph{Sebastes spp.}) populations at recently observed index levels, and did so by functioning as a constant escapement harvest control rule where target catches were set to zero below some pre-specified index level.
-In the variant used by @CoxKronlund2008, catches were adjusted to maintain a sablefish (\emph{Anoplopoma fimbria}) population at a pre-specified index level thought to be sustainable and desirable in terms of meeting fishery objectives (e.g., high catch), but never permitted target catches of zero and so functioned as a constant exploitation rate control rule.
-The ``hold-steady'' DLM of @CoxKronlund2008 performed similarly in terms of catch, stock depletion, and variation in catch, as a constant exploitation rate policy where target catch was specified as the product of desired exploitation rate and an estimate of biomass from a SCAA model.
-This result was robust to uncertainty in initial stock status and steepness {[}@CoxKronlund2008{]}.
-The SCAA model was always correctly specified (i.e., expected to produce unbiased estimates on average), however, and no comparison to the results of this research in the presence of retrospective patterns is possible {[}@CoxKronlund2008{]}.
-The ``hold-steady'' policy of @Hilbornetal2002 performed similarly to or better in terms of catch and stock status than other harvest control rules that relied on assessment estimates of biomass (i.e., 40:10 and constant \emph{F}).
-The performance of the ``hold-steady'' DLM was also more robust to uncertainty in steepness and to the presence of unreported catch {[}@Hilbornetal2002{]}.
-The performance of the two harvest policies that relied on assessment estimates of biomass (i.e., constant exploitation rate and a ``40:10'' biomass-based policy) also degraded when the estimates of biomass were biased, which is an issue that does not effect the ``hold-steady'' DLM {[}@Hilbornetal2002{]}.
-The bias in the assessment estimates considered in @Hilbornetal2002 were not necessarily induced by a retrospective pattern, however, and no consideration of making a rho-adjustment was possible in that study.
-
-The Ismooth method is currently used to set catches for Georges Bank cod {[}@nefsc19{]} and red hake (\emph{Urophycis chuss}; @nefsc20).
-Variations of the ES-Frecent have been used for witch flounder and Georges Bank yellowtail flounder.
-While the findings here generally support the continued use of the Ismooth and ES-Frecent methods, they may not be well suited for depleted stocks where unreported catches are believed to be an issue.
-The Ismooth, Islope, and ES-Frecent DLMs produced high \(Fs\) and limited stock recovery with unreported catches and when the stock was depleted.
-While @Hilbornetal2002 and @CoxKronlund2008 did not reach the same conclusion about the ``hold-steady'' DLM, those studies did not consider initial levels of depletion as low as in this study.
-These results highlight the importance of accurate catch reporting, as unreported catch can create a negative feedback loop with perpetually high \(Fs\) being produced by a management system that seemingly should result in sustainable catch advice.
-
-Three methods were consistently risk-averse across scenarios, limiting the frequency and magnitude of overfishing and resulting in high stock biomass.
-These methods were the two catch curve options (CC-FM and CC-FSPR) and DynLin.
-The catch curve methods produced a wider range of average catches across scenarios, and also had greater interannual variability in catches compared to DynLin.
-While the lower exploitation rates from these approaches may be undesirable due to forgone yield, there may be circumstances where they are preferred.
-For example, for stocks that are believed to be heavily depleted, low exploitation rates would allow for a more rapid recovery.
-
-A number of methods performed poorly, particularly when catches were unreported.
-These methods include three of the expanded survey biomass approaches (ES-Fstable, ES-FM, ES-FSPR), AIM, and Skate.
-The AIM model has been widely used across stocks in the region {[}@nefsc02a; @nefsc05; @nefsc08{]}, although there is a decreasing trend in its use across model resistant stocks {[}@nefsc19{]}.
-The findings here suggest that alternative approaches should be considered in cases where AIM is still used and there is concern over unreported catches.
-The Skate method is used to manage the skate complex in the Northeast U.S.
-(a group of seven co-managed species).
-Interestingly, six of the seven species are considered in good condition with high survey biomass indices in recent years {[}@nefmc20{]}.
-That the Skate method performed poorly in our analysis but performs well for the skate complex illustrates how the performance of methods in this analysis may be sensitive to the scenarios and species life history considered.
-As may be the case for the Skate method, the performance of some methods may depend on the condition of the stock when the method is first applied, and less so on life-history.
-Therefore, care is needed when trying to generalize these results across stocks that may have different life histories, exploitation histories, and without unreported catches or increases in \(M\).
-
-In addition to the analytical differences among the thirteen DLMs, most of the DLMs and control rules had multiple options that could be adjusted to make them more or less risk averse.
-DynLin had a large number of user defined decision points.
-Given the large range of options already explored in the study, one suite of options was selected for each DLM-control rule and kept constant for all simulations.
-Further studies could explore the different options within an individual DLM to understand how they might affect performance.
-
-Many other data-limited methods exist for setting catch advice that were not included in this evaluation, and they vary widely in complexity, data inputs, and assumptions required {[}e.g., @carruthers2018dlm{]}.
-Length based methods were not evaluated to keep the overall number of methods tractable, and due to the availability of age based information in the region.
-Methods that require only catch data or snap shots of survey data were not considered due to the availability of the relatively long and contiguous Northeast Fisheries Science Center's spring and fall, coastwide bottom trawl surveys, and the fact that ``catch only'' methods have been shown to perform poorly {[}e.g., @carruthers2014eval{]}.
-Complete catch histories are not available for stocks in the region (i.e., from the inception of fishing).
-Consequently, methods that required complete catch histories or required assumptions about relative depletion {[}e.g., DCAC in @maccall2009dca; DB-SRA in @dick2022dsra{]} were also omitted from consideration.
-The need for short run-times and the desire for methods that could be reviewed quickly prevented the use of modern state-space production models such as SPiCT {[}@pedersen2017spict{]} and JABBA {[}@winker2018jabba{]}.
-
-The SCAA was confronted with inconsistent data in this study, while the DLMs typically used only a single source of data and thus did not encounter inconsistencies.
-A recent examination of the data used in assessments in this region similarly found inconsistencies in data streams even before modeling.
-@wiedenmann2022strange found a negative relationship between relative F (catch/survey) and survey Z for stocks with strong retrospective patterns but the expected positive relationship for stocks without a retrospective pattern.
-It is exactly this sort of tension that creates retrospective patterns in integrated models, but is not found in DLMs that only use one type of data.
-
-Despite conducting hundreds of thousands of simulations, there are still limitations to our study.
-We only examined one life history representative of groundfish in the region.
-We acknowledge that best practice is to select a DLM for a specific life history and fishery condition {[}e.g., @fischer2020dlm{]}.
-As is typically the case with large simulation studies, we were not able to tune any of the DLMs or the SCAA in any given realization, which would occur in practice for an actual stock assessment.
-We also examined only scenarios that started with Mohn's rho values near 0.5 for spawning stock biomass.
-This is a strong retrospective pattern, but some stocks in the region have even stronger retrospectives.
-Performance of the DLMs and SCAA would be expected to degrade with stronger retrospectives, but by how much is still an open area for research.
-Similarly, sources of retrospective patterns that create different relationships between the true values and estimated values should also be explored {[}see @deroba2014retro{]}.
-To make the results interpretable, we only examined a single source for the retrospective pattern at a time.
-In reality, there may be more than one factor leading to an observed retrospective pattern.
-How the multiple sources would interact to influence performance is another topic for future research.
-Development of harvest control rules specifically for situations where retrospective patterns are found in age-based assessments would also be beneficial.
-The large number of scenarios examined and the large number of realizations gives us confidence that our results are meaningful in general, but that the performance of any of the DLMs may differ in actual practice.
-
-An interesting finding of this study is the linear versus diffuse patterns between \(SSB\) and catch across methods.
-These patterns have implications for the trade-offs among methods, with linear relationships resulting in more consistent exploitation rates across stock sizes.
-Therefore, these methods have higher certainty of a given catch at a given stock size.
-However, they also tended to result in lower stock sizes, on average, across methods.
-The more diffuse relationships resulted in more variable exploitation rates across stock sizes, with some situations where the population biomass was quite high but the catch was low (relative to MSY), resulting in a very low \(F\).
-The reasons behind these different patterns remain unclear, and future work to explore these patterns is warranted.
-
-One of the reasons for the difference in performance between the catch and natural mortality retrospective sources was how the reference points were calculated.
-In all cases, the initial conditions, including the natural mortality rate, were used to compute the reference points.
-This decision was made based on the fact that the increase in natural mortality was assumed to be unknown in the simulations.
-If the increase in natural mortality was known, the age-structured assessments would have accounted for it, different reference points might have been computed {[}@legault2016increaseM{]} and there may not have been a retrospective pattern at all {[}@legault2020rose{]}, and no need to consider alternative DLMs.
-The reference points for the increased \(M\) scenarios would have been different if they were computed using the values from the final year of the base period, but the overall conclusions regarding the different DLMs would not change as this just results in a rescaling of the axis.
-These results are not shown to reduce confusion regarding the simulations.
-
-Closed-loop simulation is a common tool for examining performance of catch advice from various stock assessment approaches in a feedback setting.
-It is often used as part of a full management strategy evaluation when working with stakeholders to develop management regulations that make trade offs between near term and long term catches, risk to the fish population, and mixed-fleet allocations {[}@carruthers2016simpleMPs; @goethel2019mse; @harlyan2019hcr{]}.
-We did not conduct a full management strategy evaluation with stakeholder input {[}@goethel2019stakeholder{]}, but see that as a fruitful next step that could build on the conclusions from our closed-loop work.
-Using a generic groundfish life-history and monitoring standard performance metrics related to stock status and catch stability, we were able to cull the herd of potential DLMs and we would not carry the consistent poor performers forward for further study.
-The wide range of expertise reflected in the authorship was by design so that the simulation specifications and performance metrics were broadly useful.
-Before undertaking a full management strategy evaluation and engaging regional stakeholders, we would want to select a specific stock and jointly identify specific management regulations to be tested {[}@deroba2019dream{]}.
-Results of this work have been presented to both local fishery management councils, with generally positive feedback about the utility of the conclusions for identifying appropriate model approaches when an SCAA is rejected.
-Our work was similar to all other closed-loop simulations in that it was designed to address a specific situation, including much recent work comparing the performance of data-limited and data rich assessment approaches {[}e.g., @fulton2016datarich; @sagarese2019dlm; @bouch2020datapoor; @li2022dlm{]}.
-
-This study is a first attempt to identify suitable methods for setting catch advice when stock assessment models are rejected due to large, positive retrospective patterns.
-Although no single method performed best across scenarios, a number of generally suitable and unsuitable methods were identified under specific conditions.
-The results of this work can help scientists and managers select a subset of possible options for consideration to set catch advice when assessment models are rejected.
-The approach developed here can, and should be expanded to consider other cases not explored here, as performance of individual methods are very likely case-dependent.
 
@@
 
+\begin{tabular}{l|l|l|l|l|l|r|r}
+\hline
+Scenario & Temp & Covariate & Strata & Noise & Season & Stratified Mean & VAST Estimate\\
+\hline
+Constant Population & constantant & no cov & all & no & spring & 0.21 & 0.11\\
+\hline
+Constant Population & constantant & no cov & all & yes & spring & 0.25 & 0.16\\
+\hline
+Constant Population & constantant & w/ cov & all & no & spring & 0.21 & 0.07\\
+\hline
+Constant Population & constantant & w/ cov & all & yes & spring & 0.25 & 0.08\\
+\hline
+Constant Population & constantant & no cov & all & no & fall & 0.32 & 0.68\\
+\hline
+Constant Population & constantant & no cov & all & yes & fall & 0.31 & 0.77\\
+\hline
+Constant Population & constantant & w/ cov & all & no & fall & 0.32 & 0.08\\
+\hline
+Constant Population & constant & w/ cov & all & yes & fall & 0.31 & 0.11\\
+\hline
+Constant Population & constant & no cov & reduced & no & spring & 0.27 & 0.19\\
+\hline
+Constant Population & constant & no cov & reduced & yes & spring & 0.26 & 0.15\\
+\hline
+Constant Population & constant & w/ cov & reduced & no & spring & 0.27 & 0.19\\
+\hline
+Constant Population & constant & w/ cov & reduced & yes & spring & 0.26 & 0.14\\
+\hline
+Constant Population & constant & no cov & reduced & no & fall & 0.47 & 0.25\\
+\hline
+Constant Population & constant & no cov & reduced & yes & fall & 0.49 & 0.36\\
+\hline
+Constant Population & constant & w/ cov & reduced & no & fall & 0.47 & 0.19\\
+\hline
+Constant Population & constant & w/ cov & reduced & yes & fall & 0.49 & 0.17\\
+\hline
+Constant Population & increasing & no cov & all & no & spring & 0.28 & 0.11\\
+\hline
+Constant Population & increasing & no cov & all & yes & spring & 0.28 & 0.15\\
+\hline
+Constant Population & increasing & w/ cov & all & no & spring & 0.28 & 0.06\\
+\hline
+Constant Population & increasing & w/ cov & all & yes & spring & 0.28 & 0.12\\
+\hline
+Constant Population & increasing & no cov & all & no & fall & 0.51 & 1.26\\
+\hline
+Constant Population & increasing & no cov & all & yes & fall & 0.50 & 1.38\\
+\hline
+Constant Population & increasing & w/ cov & all & no & fall & 0.51 & 0.23\\
+\hline
+Constant Population & increasing & w/ cov & all & yes & fall & 0.50 & 0.28\\
+\hline
+Constant Population & increasing & no cov & reduced & no & spring & 0.31 & 0.17\\
+\hline
+Constant Population & increasing & no cov & reduced & yes & spring & 0.29 & 0.19\\
+\hline
+Constant Population & increasing & w/ cov & reduced & no & spring & 0.31 & 0.17\\
+\hline
+Constant Population & increasing & w/ cov & reduced & yes & spring & 0.29 & 0.15\\
+\hline
+Constant Population & increasing & no cov & reduced & no & fall & 0.64 & 0.75\\
+\hline
+Constant Population & increasing & no cov & reduced & yes & fall & 0.66 & 0.89\\
+\hline
+Constant Population & increasing & w/ cov & reduced & no & fall & 0.64 & 0.20\\
+\hline
+Constant Population & increasing & w/ cov & reduced & yes & fall & 0.66 & 0.13\\
+\hline
+Decreasing Population & constant & no cov & all & no & spring & 0.23 & 0.08\\
+\hline
+Decreasing Population & constant & no cov & all & yes & spring & 0.27 & 0.11\\
+\hline
+Decreasing Population & constant & w/ cov & all & no & spring & 0.23 & 0.06\\
+\hline
+Decreasing Population & constant & w/ cov & all & yes & spring & 0.27 & 0.07\\
+\hline
+Decreasing Population & constant & no cov & all & no & fall & 0.41 & 0.81\\
+\hline
+Decreasing Population & constant & no cov & all & yes & fall & 0.44 & 1.09\\
+\hline
+Decreasing Population & constant & w/ cov & all & no & fall & 0.41 & 0.08\\
+\hline
+Decreasing Population & constant & w/ cov & all & yes & fall & 0.44 & 0.18\\
+\hline
+Decreasing Population & constant & no cov & reduced & no & spring & 0.25 & 0.19\\
+\hline
+Decreasing Population & constant & no cov & reduced & yes & spring & 0.30 & 0.16\\
+\hline
+Decreasing Population & constant & w/ cov & reduced & no & spring & 0.25 & 0.19\\
+\hline
+Decreasing Population & constant & w/ cov & reduced & yes & spring & 0.30 & 0.16\\
+\hline
+Decreasing Population & constant & no cov & reduced & no & fall & 0.55 & 0.24\\
+\hline
+Decreasing Population & constant & no cov & reduced & yes & fall & 0.53 & 0.36\\
+\hline
+Decreasing Population & constant & w/ cov & reduced & no & fall & 0.55 & 0.23\\
+\hline
+Decreasing Population & constant & w/ cov & reduced & yes & fall & 0.53 & 0.24\\
+\hline
+Decreasing Population & increasing & no cov & all & no & spring & 0.22 & 0.15\\
+\hline
+Decreasing Population & increasing & no cov & all & yes & spring & 0.26 & 0.17\\
+\hline
+Decreasing Population & increasing & w/ cov & all & no & spring & 0.22 & 0.07\\
+\hline
+Decreasing Population & increasing & w/ cov & all & yes & spring & 0.26 & 0.10\\
+\hline
+Decreasing Population & increasing & no cov & all & no & fall & 0.28 & 1.06\\
+\hline
+Decreasing Population & increasing & no cov & all & yes & fall & 0.25 & 1.10\\
+\hline
+Decreasing Population & increasing & w/ cov & all & no & fall & 0.28 & 0.15\\
+\hline
+Decreasing Population & increasing & w/ cov & all & yes & fall & 0.25 & 0.20\\
+\hline
+Decreasing Population & increasing & no cov & reduced & no & spring & 0.32 & 0.14\\
+\hline
+Decreasing Population & increasing & no cov & reduced & yes & spring & 0.29 & 0.15\\
+\hline
+Decreasing Population & increasing & w/ cov & reduced & no & spring & 0.32 & 0.22\\
+\hline
+Decreasing Population & increasing & w/ cov & reduced & yes & spring & 0.29 & 0.21\\
+\hline
+Decreasing Population & increasing & no cov & reduced & no & fall & 0.54 & 0.60\\
+\hline
+Decreasing Population & increasing & no cov & reduced & yes & fall & 0.53 & 0.62\\
+\hline
+Decreasing Population & increasing & w/ cov & reduced & no & fall & 0.54 & 0.31\\
+\hline
+Decreasing Population & increasing & w/ cov & reduced & yes & fall & 0.53 & 0.32\\
+\hline
+Increasing Population & constant & no cov & all & no & spring & 0.16 & 0.13\\
+\hline
+Increasing Population & constant & no cov & all & yes & spring & 0.22 & 0.21\\
+\hline
+Increasing Population & constant & w/ cov & all & no & spring & 0.16 & 0.06\\
+\hline
+Increasing Population & constant & w/ cov & all & yes & spring & 0.22 & 0.07\\
+\hline
+Increasing Population & constant & no cov & all & no & fall & 0.34 & 0.36\\
+\hline
+Increasing Population & constant & no cov & all & yes & fall & 0.46 & 0.44\\
+\hline
+Increasing Population & constant & w/ cov & all & no & fall & 0.34 & 0.08\\
+\hline
+Increasing Population & constant & w/ cov & all & yes & fall & 0.46 & 0.17\\
+\hline
+Increasing Population & constant & no cov & reduced & no & spring & 0.20 & 0.15\\
+\hline
+Increasing Population & constant & no cov & reduced & yes & spring & 0.22 & 0.11\\
+\hline
+Increasing Population & constant & w/ cov & reduced & no & spring & 0.20 & 0.17\\
+\hline
+Increasing Population & constant & w/ cov & reduced & yes & spring & 0.22 & 0.13\\
+\hline
+Increasing Population & constant & no cov & reduced & no & fall & 0.41 & 0.11\\
+\hline
+Increasing Population & constant & no cov & reduced & yes & fall & 0.46 & 0.21\\
+\hline
+Increasing Population & constant & w/ cov & reduced & no & fall & 0.41 & 0.22\\
+\hline
+Increasing Population & constant & w/ cov & reduced & yes & fall & 0.46 & 0.19\\
+\hline
+Increasing Population & increasing & no cov & all & no & spring & 0.32 & 0.13\\
+\hline
+Increasing Population & increasing & no cov & all & yes & spring & 0.34 & 0.16\\
+\hline
+Increasing Population & increasing & w/ cov & all & no & spring & 0.32 & 0.07\\
+\hline
+Increasing Population & increasing & w/ cov & all & yes & spring & 0.34 & 0.10\\
+\hline
+Increasing Population & increasing & no cov & all & no & fall & 0.30 & 0.66\\
+\hline
+Increasing Population & increasing & no cov & all & yes & fall & 0.39 & 0.71\\
+\hline
+Increasing Population & increasing & w/ cov & all & no & fall & 0.30 & 0.21\\
+\hline
+Increasing Population & increasing & w/ cov & all & yes & fall & 0.30 & 0.37\\
+\hline
+Increasing Population & increasing & no cov & reduced & no & spring & 0.40 & 0.33\\
+\hline
+Increasing Population & increasing & no cov & reduced & yes & spring & 0.41 & 0.26\\
+\hline
+Increasing Population & increasing & w/ cov & reduced & no & spring & 0.40 & 0.30\\
+\hline
+Increasing Population & increasing & w/ cov & reduced & yes & spring & 0.41 & 0.25\\
+\hline
+Increasing Population & increasing & no cov & reduced & no & fall & 0.70 & 0.49\\
+\hline
+Increasing Population & increasing & no cov & reduced & yes & fall & 0.69 & 0.48\\
+\hline
+Increasing Population & increasing & w/ cov & reduced & no & fall & 0.70 & 0.53\\
+\hline
+Increasing Population & increasing & w/ cov & reduced & yes & fall & 0.69 & 0.50\\
+\hline
+\end{tabular}
+
 \end{document}

### 2022-12-23 Manuscript/Draft1.Rmd
updating manuscript

@@
 For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
-We consider all possible scenario combinations for Yellowtail Flounder to provide a more comprehensive analysis of population indexing methods.
+We consider all possible scenario combinations for Yellowtail Flounder to provide a comprehensive analysis of population indexing methods.
+
+In the Cod results, when using reduced strata adding covariates produces worst VAST results (though still better than stratified mean).
+VAST without covariates much worse than stratified mean in fall with increasing temperture and all strata, but adding covariates corrects this.
+
+For Haddock, VAST has a particularly hard time in spring regulalry producing larger errors than the stratified mean with added covariates only improving to the level of the stratifie dmean.
+VAST shows imrpoved results in fall relative to the stratified mean with added covaraites producing extremely low errors in some cases.
+
+In considering the Yellowtail Flounder results in Table XXX, we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
+However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in the Table.
+These errors are corrected by including covariate information allowing for an improved VAST estimate that are significantly lower than their stratified mean counterparts.
 
@@
 
-## Discussion
+\section{Discussion}
 
-## Acknowledgements
-We thank the Index-Based Methods and Control Rules Research Track review panel of Paul Rago (chair), Yong Chen, Robin Cook, and Paul Medley for feedback on preliminary results, three anonymous reviewers and the associate editor for reviewing an earlier version of this work.
-The scientific results and conclusions, as well as any views or opinions expressed herein, are those of the authors and do not necessarily reflect those of NOAA or the Department of Commerce.
+\section{Acknowledgements}
 

### 2022-12-23 Manuscript/Draft1.tex
updating manuscript

@@
 
-\hypertarget{discussion}{%
-\subsection{Discussion}\label{discussion}}
+\section{Discussion}
 
-\hypertarget{acknowledgements}{%
-\subsection{Acknowledgements}\label{acknowledgements}}
-
-We thank the Index-Based Methods and Control Rules Research Track review panel of Paul Rago (chair), Yong Chen, Robin Cook, and Paul Medley for feedback on preliminary results, three anonymous reviewers and the associate editor for reviewing an earlier version of this work.
-The scientific results and conclusions, as well as any views or opinions expressed herein, are those of the authors and do not necessarily reflect those of NOAA or the Department of Commerce.
+\section{Acknowledgements}
 

### 2022-12-26 Manuscript/Draft1.Rmd
Update manuscript and readme file

@@
 
-The goal of our project was to analyze how well different contemporary population indexing methods can track population trends under a host of conditions, as depicted in Table COMBOS.
+The goal of our project was to analyze how well contemporary population indexing methods can track population trends under a host of conditions, as depicted in Table COMBOS.
 Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year **cite**.
-For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
-We consider all possible scenario combinations for Yellowtail Flounder to provide a comprehensive analysis of population indexing methods.
+For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod (see Table ???) and increasing population scenarios for Haddock (see Table ???).
+To provide a comprehensive analysis of population indexing methods we consider all possible scenario combinations for Yellowtail Flounder (see Table ???).
+
+General themes that exist in Tables X, Y and Z are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors.
+We also see individual cases where the stratified mean produced the lowest absolut error and instances where including covarites in VAST models actually increase the absolute error.
+When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error, though there are some scenarios where the impact is minimal.
 
 In the Cod results, when using reduced strata adding covariates produces worst VAST results (though still better than stratified mean).
-VAST without covariates much worse than stratified mean in fall with increasing temperture and all strata, but adding covariates corrects this.
+VAST without covariates much worse than stratified mean in fall with increasing temperature and all strata, but adding covariates corrects this.
 
-For Haddock, VAST has a particularly hard time in spring regulalry producing larger errors than the stratified mean with added covariates only improving to the level of the stratifie dmean.
-VAST shows imrpoved results in fall relative to the stratified mean with added covaraites producing extremely low errors in some cases.
+For Haddock, VAST has a particularly hard time in spring regularly producing larger errors than the stratified mean with added covariates only improving to the level of the stratified mean.
+VAST shows improved results in fall relative to the stratified mean with added covaraites producing extremely low errors in some cases.
 
@@
 # Group the table according to Parameter, Scenario
-  group_by(Season, Scenario)%>%
-# Change the orientation of the table so that Species become the variable headings
-  mutate(Scenario = factor(Scenario)) %>%
-  arrange(Scenario) %>%
-  select(Scenario, everything())
+  group_by(Scenario,Strata)
 
 # Create table to present the data]
-Fig_5 <- kable(summary_tab)
+Fig_5 <- kable(summary_tab) %>%
+  kable_styling(font_size = 10)  %>%
+  pack_rows(tab_kable, colnum = c(1,4),
+    index = table(fct_inorder(summary_tab$Strata), useNA = "no")) %>%
+  pack_rows(tab_kable, colnum = 1,
+    index = table(fct_inorder(summary_tab$Scenario), useNA = "no"))
 

### 2022-12-26 Manuscript/Draft1.tex
Update manuscript and readme file

@@
 
-The goal of our project was to analyze how well different contemporary population indexing methods can track population trends under a host of conditions, as depicted in Table COMBOS.
+The goal of our project was to analyze how well contemporary population indexing methods can track population trends under a host of conditions, as depicted in Table COMBOS.
 Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year \textbf{cite}.
-For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
-We consider all possible scenario combinations for Yellowtail Flounder to provide a comprehensive analysis of population indexing methods.
+For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod (see Table ???) and increasing population scenarios for Haddock (see Table ???).
+To provide a comprehensive analysis of population indexing methods we consider all possible scenario combinations for Yellowtail Flounder (see Table ???).
+
+General themes that exist in Tables X, Y and Z are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors.
+We also see individual cases where the stratified mean produced the lowest absolut error and instances where including covarites in VAST models actually increase the absolute error.
+When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error, though there are some scenarios where the impact is minimal.
 
 In the Cod results, when using reduced strata adding covariates produces worst VAST results (though still better than stratified mean).
-VAST without covariates much worse than stratified mean in fall with increasing temperture and all strata, but adding covariates corrects this.
+VAST without covariates much worse than stratified mean in fall with increasing temperature and all strata, but adding covariates corrects this.
 
-For Haddock, VAST has a particularly hard time in spring regulalry producing larger errors than the stratified mean with added covariates only improving to the level of the stratifie dmean.
-VAST shows imrpoved results in fall relative to the stratified mean with added covaraites producing extremely low errors in some cases.
+For Haddock, VAST has a particularly hard time in spring regularly producing larger errors than the stratified mean with added covariates only improving to the level of the stratified mean.
+VAST shows improved results in fall relative to the stratified mean with added covaraites producing extremely low errors in some cases.
 
@@
 
+\begin{table}
+\centering\begingroup\fontsize{10}{12}\selectfont
+
 \begin{tabular}{l|l|l|l|l|l|r|r}
 \hline
-Scenario & Temp & Covariate & Strata & Noise & Season & Stratified Mean & VAST Estimate\\
+Temp & Covariate & Strata & Noise & Season & Scenario & Stratified Mean & VAST Estimate\\
 \hline
-Constant Population & constantant & no cov & all & no & spring & 0.21 & 0.11\\
+\multicolumn{1}{l}{\textbf{Constant Population}}\\
 \hline
-Constant Population & constantant & no cov & all & yes & spring & 0.25 & 0.16\\
+\multicolumn{3}{l}{\textbf{all}}\\
 \hline
-Constant Population & constantant & w/ cov & all & no & spring & 0.21 & 0.07\\
+\hspace{1em}\hspace{1em}constant & no cov & all & no & spring & Constant Population & 0.21 & 0.11\\
 \hline
-Constant Population & constantant & w/ cov & all & yes & spring & 0.25 & 0.08\\
+\hspace{1em}\hspace{1em}constant & no cov & all & yes & spring & Constant Population & 0.25 & 0.16\\
 \hline
-Constant Population & constantant & no cov & all & no & fall & 0.32 & 0.68\\
+\hspace{1em}\hspace{1em}constant & w/ cov & all & no & spring & Constant Population & 0.21 & 0.07\\
 \hline
-Constant Population & constantant & no cov & all & yes & fall & 0.31 & 0.77\\
+\hspace{1em}\hspace{1em}constant & w/ cov & all & yes & spring & Constant Population & 0.25 & 0.08\\
 \hline
-Constant Population & constantant & w/ cov & all & no & fall & 0.32 & 0.08\\
+\hspace{1em}\hspace{1em}constant & no cov & all & no & fall & Constant Population & 0.32 & 0.68\\
 \hline
-Constant Population & constant & w/ cov & all & yes & fall & 0.31 & 0.11\\
+\hspace{1em}\hspace{1em}constant & no cov & all & yes & fall & Constant Population & 0.31 & 0.77\\
 \hline
-Constant Population & constant & no cov & reduced & no & spring & 0.27 & 0.19\\
+\hspace{1em}\hspace{1em}constant & w/ cov & all & no & fall & Constant Population & 0.32 & 0.08\\
 \hline
-Constant Population & constant & no cov & reduced & yes & spring & 0.26 & 0.15\\
+\hspace{1em}\hspace{1em}constant & w/ cov & all & yes & fall & Constant Population & 0.31 & 0.11\\
 \hline
-Constant Population & constant & w/ cov & reduced & no & spring & 0.27 & 0.19\\
+\hspace{1em}\hspace{1em}constant & no cov & reduced & no & spring & Constant Population & 0.27 & 0.19\\
 \hline
-Constant Population & constant & w/ cov & reduced & yes & spring & 0.26 & 0.14\\
+\hspace{1em}\hspace{1em}constant & no cov & reduced & yes & spring & Constant Population & 0.26 & 0.15\\
 \hline
-Constant Population & constant & no cov & reduced & no & fall & 0.47 & 0.25\\
+\hspace{1em}\hspace{1em}constant & w/ cov & reduced & no & spring & Constant Population & 0.27 & 0.19\\
 \hline
-Constant Population & constant & no cov & reduced & yes & fall & 0.49 & 0.36\\
+\hspace{1em}\hspace{1em}constant & w/ cov & reduced & yes & spring & Constant Population & 0.26 & 0.14\\
 \hline
-Constant Population & constant & w/ cov & reduced & no & fall & 0.47 & 0.19\\
+\hspace{1em}\hspace{1em}constant & no cov & reduced & no & fall & Constant Population & 0.47 & 0.25\\
 \hline
-Constant Population & constant & w/ cov & reduced & yes & fall & 0.49 & 0.17\\
+\hspace{1em}\hspace{1em}constant & no cov & reduced & yes & fall & Constant Population & 0.49 & 0.36\\
 \hline
-Constant Population & increasing & no cov & all & no & spring & 0.28 & 0.11\\
+\hspace{1em}\hspace{1em}constant & w/ cov & reduced & no & fall & Constant Population & 0.47 & 0.19\\
 \hline
-Constant Population & increasing & no cov & all & yes & spring & 0.28 & 0.15\\
+\hspace{1em}\hspace{1em}constant & w/ cov & reduced & yes & fall & Constant Population & 0.49 & 0.17\\
 \hline
-Constant Population & increasing & w/ cov & all & no & spring & 0.28 & 0.06\\
+\hspace{1em}\hspace{1em}increasing & no cov & all & no & spring & Constant Population & 0.28 & 0.11\\
 \hline
-Constant Population & increasing & w/ cov & all & yes & spring & 0.28 & 0.12\\
+\hspace{1em}\hspace{1em}increasing & no cov & all & yes & spring & Constant Population & 0.28 & 0.15\\
 \hline
-Constant Population & increasing & no cov & all & no & fall & 0.51 & 1.26\\
+\hspace{1em}\hspace{1em}increasing & w/ cov & all & no & spring & Constant Population & 0.28 & 0.06\\
 \hline
-Constant Population & increasing & no cov & all & yes & fall & 0.50 & 1.38\\
+\hspace{1em}\hspace{1em}increasing & w/ cov & all & yes & spring & Constant Population & 0.28 & 0.12\\
 \hline
-Constant Population & increasing & w/ cov & all & no & fall & 0.51 & 0.23\\
+\hspace{1em}\hspace{1em}increasing & no cov & all & no & fall & Constant Population & 0.51 & 1.26\\
 \hline
-Constant Population & increasing & w/ cov & all & yes & fall & 0.50 & 0.28\\
+\hspace{1em}\hspace{1em}increasing & no cov & all & yes & fall & Constant Population & 0.50 & 1.38\\
 \hline
-Constant Population & increasing & no cov & reduced & no & spring & 0.31 & 0.17\\
+\hspace{1em}\hspace{1em}increasing & w/ cov & all & no & fall & Constant Population & 0.51 & 0.23\\
 \hline
-Constant Population & increasing & no cov & reduced & yes & spring & 0.29 & 0.19\\
+\hspace{1em}\hspace{1em}increasing & w/ cov & all & yes & fall & Constant Population & 0.50 & 0.28\\
 \hline
-Constant Population & increasing & w/ cov & reduced & no & spring & 0.31 & 0.17\\
+\hspace{1em}\hspace{1em}increasing & no cov & reduced & no & spring & Constant Population & 0.31 & 0.17\\
 \hline
-Constant Population & increasing & w/ cov & reduced & yes & spring & 0.29 & 0.15\\
+\hspace{1em}\hspace{1em}increasing & no cov & reduced & yes & spring & Constant Population & 0.29 & 0.19\\
 \hline
-Constant Population & increasing & no cov & reduced & no & fall & 0.64 & 0.75\\
+\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & no & spring & Constant Population & 0.31 & 0.17\\
 \hline
-Constant Population & increasing & no cov & reduced & yes & fall & 0.66 & 0.89\\
+\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & yes & spring & Constant Population & 0.29 & 0.15\\
 \hline
-Constant Population & increasing & w/ cov & reduced & no & fall & 0.64 & 0.20\\
+\hspace{1em}\hspace{1em}increasing & no cov & reduced & no & fall & Constant Population & 0.64 & 0.75\\
 \hline
-Constant Population & increasing & w/ cov & reduced & yes & fall & 0.66 & 0.13\\
+\hspace{1em}\hspace{1em}increasing & no cov & reduced & yes & fall & Constant Population & 0.66 & 0.89\\
 \hline
-Decreasing Population & constant & no cov & all & no & spring & 0.23 & 0.08\\
+\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & no & fall & Constant Population & 0.64 & 0.20\\
 \hline
-Decreasing Population & constant & no cov & all & yes & spring & 0.27 & 0.11\\
+\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & yes & fall & Constant Population & 0.66 & 0.13\\
 \hline
-Decreasing Population & constant & w/ cov & all & no & spring & 0.23 & 0.06\\
+\multicolumn{1}{l}{\textbf{Increasing Population}}\\
 \hline
-Decreasing Population & constant & w/ cov & all & yes & spring & 0.27 & 0.07\\
+\hspace{1em}\hspace{1em}constant & no cov & all & no & spring & Increasing Population & 0.16 & 0.13\\
 \hline
-Decreasing Population & constant & no cov & all & no & fall & 0.41 & 0.81\\
+\hspace{1em}\hspace{1em}constant & no cov & all & yes & spring & Increasing Population & 0.22 & 0.21\\
 \hline
-Decreasing Population & constant & no cov & all & yes & fall & 0.44 & 1.09\\
+\hspace{1em}\hspace{1em}constant & w/ cov & all & no & spring & Increasing Population & 0.16 & 0.06\\
 \hline
-Decreasing Population & constant & w/ cov & all & no & fall & 0.41 & 0.08\\
+\hspace{1em}\hspace{1em}constant & w/ cov & all & yes & spring & Increasing Population & 0.22 & 0.07\\
 \hline
-Decreasing Population & constant & w/ cov & all & yes & fall & 0.44 & 0.18\\
+\hspace{1em}\hspace{1em}constant & no cov & all & no & fall & Increasing Population & 0.34 & 0.36\\
 \hline
-Decreasing Population & constant & no cov & reduced & no & spring & 0.25 & 0.19\\
+\hspace{1em}\hspace{1em}constant & no cov & all & yes & fall & Increasing Population & 0.46 & 0.44\\
 \hline
-Decreasing Population & constant & no cov & reduced & yes & spring & 0.30 & 0.16\\
+\hspace{1em}\hspace{1em}constant & w/ cov & all & no & fall & Increasing Population & 0.34 & 0.08\\
 \hline
-Decreasing Population & constant & w/ cov & reduced & no & spring & 0.25 & 0.19\\
+\hspace{1em}\hspace{1em}constant & w/ cov & all & yes & fall & Increasing Population & 0.46 & 0.17\\
 \hline
-Decreasing Population & constant & w/ cov & reduced & yes & spring & 0.30 & 0.16\\
+\hspace{1em}\hspace{1em}constant & no cov & reduced & no & spring & Increasing Population & 0.20 & 0.15\\
 \hline
-Decreasing Population & constant & no cov & reduced & no & fall & 0.55 & 0.24\\
+\hspace{1em}\hspace{1em}constant & no cov & reduced & yes & spring & Increasing Population & 0.22 & 0.11\\
 \hline
-Decreasing Population & constant & no cov & reduced & yes & fall & 0.53 & 0.36\\
+\hspace{1em}\hspace{1em}constant & w/ cov & reduced & no & spring & Increasing Population & 0.20 & 0.17\\
 \hline
-Decreasing Population & constant & w/ cov & reduced & no & fall & 0.55 & 0.23\\
+\hspace{1em}\hspace{1em}constant & w/ cov & reduced & yes & spring & Increasing Population & 0.22 & 0.13\\
 \hline
-Decreasing Population & constant & w/ cov & reduced & yes & fall & 0.53 & 0.24\\
+\hspace{1em}\hspace{1em}constant & no cov & reduced & no & fall & Increasing Population & 0.41 & 0.11\\
 \hline
-Decreasing Population & increasing & no cov & all & no & spring & 0.22 & 0.15\\
+\hspace{1em}\hspace{1em}constant & no cov & reduced & yes & fall & Increasing Population & 0.46 & 0.21\\
 \hline
-Decreasing Population & increasing & no cov & all & yes & spring & 0.26 & 0.17\\
+\hspace{1em}\hspace{1em}constant & w/ cov & reduced & no & fall & Increasing Population & 0.41 & 0.22\\
 \hline
-Decreasing Population & increasing & w/ cov & all & no & spring & 0.22 & 0.07\\
+\hspace{1em}\hspace{1em}constant & w/ cov & reduced & yes & fall & Increasing Population & 0.46 & 0.19\\
 \hline
-Decreasing Population & increasing & w/ cov & all & yes & spring & 0.26 & 0.10\\
+\multicolumn{3}{l}{\textbf{reduced}}\\
 \hline
-Decreasing Population & increasing & no cov & all & no & fall & 0.28 & 1.06\\
+\hspace{1em}\hspace{1em}increasing & no cov & all & no & spring & Increasing Population & 0.32 & 0.13\\
 \hline
-Decreasing Population & increasing & no cov & all & yes & fall & 0.25 & 1.10\\
+\hspace{1em}\hspace{1em}increasing & no cov & all & yes & spring & Increasing Population & 0.34 & 0.16\\
 \hline
-Decreasing Population & increasing & w/ cov & all & no & fall & 0.28 & 0.15\\
+\hspace{1em}\hspace{1em}increasing & w/ cov & all & no & spring & Increasing Population & 0.32 & 0.07\\
 \hline
-Decreasing Population & increasing & w/ cov & all & yes & fall & 0.25 & 0.20\\
+\hspace{1em}\hspace{1em}increasing & w/ cov & all & yes & spring & Increasing Population & 0.34 & 0.10\\
 \hline
-Decreasing Population & increasing & no cov & reduced & no & spring & 0.32 & 0.14\\
+\hspace{1em}\hspace{1em}increasing & no cov & all & no & fall & Increasing Population & 0.30 & 0.66\\
 \hline
-Decreasing Population & increasing & no cov & reduced & yes & spring & 0.29 & 0.15\\
+\hspace{1em}\hspace{1em}increasing & no cov & all & yes & fall & Increasing Population & 0.39 & 0.71\\
 \hline
-Decreasing Population & increasing & w/ cov & reduced & no & spring & 0.32 & 0.22\\
+\hspace{1em}\hspace{1em}increasing & w/ cov & all & no & fall & Increasing Population & 0.30 & 0.21\\
 \hline
-Decreasing Population & increasing & w/ cov & reduced & yes & spring & 0.29 & 0.21\\
+\hspace{1em}\hspace{1em}increasing & w/ cov & all & yes & fall & Increasing Population & 0.30 & 0.37\\
 \hline
-Decreasing Population & increasing & no cov & reduced & no & fall & 0.54 & 0.60\\
+\hspace{1em}\hspace{1em}increasing & no cov & reduced & no & spring & Increasing Population & 0.40 & 0.33\\
 \hline
-Decreasing Population & increasing & no cov & reduced & yes & fall & 0.53 & 0.62\\
+\hspace{1em}\hspace{1em}increasing & no cov & reduced & yes & spring & Increasing Population & 0.41 & 0.26\\
 \hline
-Decreasing Population & increasing & w/ cov & reduced & no & fall & 0.54 & 0.31\\
+\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & no & spring & Increasing Population & 0.40 & 0.30\\
 \hline
-Decreasing Population & increasing & w/ cov & reduced & yes & fall & 0.53 & 0.32\\
+\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & yes & spring & Increasing Population & 0.41 & 0.25\\
 \hline
-Increasing Population & constant & no cov & all & no & spring & 0.16 & 0.13\\
+\hspace{1em}\hspace{1em}increasing & no cov & reduced & no & fall & Increasing Population & 0.70 & 0.49\\
 \hline
-Increasing Population & constant & no cov & all & yes & spring & 0.22 & 0.21\\
+\hspace{1em}\hspace{1em}increasing & no cov & reduced & yes & fall & Increasing Population & 0.69 & 0.48\\
 \hline
-Increasing Population & constant & w/ cov & all & no & spring & 0.16 & 0.06\\
+\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & no & fall & Increasing Population & 0.70 & 0.53\\
 \hline
-Increasing Population & constant & w/ cov & all & yes & spring & 0.22 & 0.07\\
+\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & yes & fall & Increasing Population & 0.69 & 0.50\\
 \hline
-Increasing Population & constant & no cov & all & no & fall & 0.34 & 0.36\\
+\multicolumn{1}{l}{\textbf{Decreasing Population}}\\
 \hline
-Increasing Population & constant & no cov & all & yes & fall & 0.46 & 0.44\\
+\hspace{1em}\hspace{1em}constant & no cov & all & no & spring & Decreasing Population & 0.23 & 0.08\\
 \hline
-Increasing Population & constant & w/ cov & all & no & fall & 0.34 & 0.08\\
+\hspace{1em}\hspace{1em}constant & no cov & all & yes & spring & Decreasing Population & 0.27 & 0.11\\
 \hline
-Increasing Population & constant & w/ cov & all & yes & fall & 0.46 & 0.17\\
+\hspace{1em}\hspace{1em}constant & w/ cov & all & no & spring & Decreasing Population & 0.23 & 0.06\\
 \hline
-Increasing Population & constant & no cov & reduced & no & spring & 0.20 & 0.15\\
+\hspace{1em}\hspace{1em}constant & w/ cov & all & yes & spring & Decreasing Population & 0.27 & 0.07\\
 \hline
-Increasing Population & constant & no cov & reduced & yes & spring & 0.22 & 0.11\\
+\hspace{1em}\hspace{1em}constant & no cov & all & no & fall & Decreasing Population & 0.41 & 0.81\\
 \hline
-Increasing Population & constant & w/ cov & reduced & no & spring & 0.20 & 0.17\\
+\hspace{1em}\hspace{1em}constant & no cov & all & yes & fall & Decreasing Population & 0.44 & 1.09\\
 \hline
-Increasing Population & constant & w/ cov & reduced & yes & spring & 0.22 & 0.13\\
+\hspace{1em}\hspace{1em}constant & w/ cov & all & no & fall & Decreasing Population & 0.41 & 0.08\\
 \hline
-Increasing Population & constant & no cov & reduced & no & fall & 0.41 & 0.11\\
+\hspace{1em}\hspace{1em}constant & w/ cov & all & yes & fall & Decreasing Population & 0.44 & 0.18\\
 \hline
-Increasing Population & constant & no cov & reduced & yes & fall & 0.46 & 0.21\\
+\hspace{1em}\hspace{1em}constant & no cov & reduced & no & spring & Decreasing Population & 0.25 & 0.19\\
 \hline
-Increasing Population & constant & w/ cov & reduced & no & fall & 0.41 & 0.22\\
+\hspace{1em}\hspace{1em}constant & no cov & reduced & yes & spring & Decreasing Population & 0.30 & 0.16\\
 \hline
-Increasing Population & constant & w/ cov & reduced & yes & fall & 0.46 & 0.19\\
+\hspace{1em}\hspace{1em}constant & w/ cov & reduced & no & spring & Decreasing Population & 0.25 & 0.19\\
 \hline
-Increasing Population & increasing & no cov & all & no & spring & 0.32 & 0.13\\
+\hspace{1em}\hspace{1em}constant & w/ cov & reduced & yes & spring & Decreasing Population & 0.30 & 0.16\\
 \hline
-Increasing Population & increasing & no cov & all & yes & spring & 0.34 & 0.16\\
+\hspace{1em}\hspace{1em}constant & no cov & reduced & no & fall & Decreasing Population & 0.55 & 0.24\\
 \hline
-Increasing Population & increasing & w/ cov & all & no & spring & 0.32 & 0.07\\
+\hspace{1em}\hspace{1em}constant & no cov & reduced & yes & fall & Decreasing Population & 0.53 & 0.36\\
 \hline
-Increasing Population & increasing & w/ cov & all & yes & spring & 0.34 & 0.10\\
+\hspace{1em}\hspace{1em}constant & w/ cov & reduced & no & fall & Decreasing Population & 0.55 & 0.23\\
 \hline
-Increasing Population & increasing & no cov & all & no & fall & 0.30 & 0.66\\
+\hspace{1em}\hspace{1em}constant & w/ cov & reduced & yes & fall & Decreasing Population & 0.53 & 0.24\\
 \hline
-Increasing Population & increasing & no cov & all & yes & fall & 0.39 & 0.71\\
+\hspace{1em}\hspace{1em}increasing & no cov & all & no & spring & Decreasing Population & 0.22 & 0.15\\
 \hline
-Increasing Population & increasing & w/ cov & all & no & fall & 0.30 & 0.21\\
+\hspace{1em}\hspace{1em}increasing & no cov & all & yes & spring & Decreasing Population & 0.26 & 0.17\\
 \hline
-Increasing Population & increasing & w/ cov & all & yes & fall & 0.30 & 0.37\\
+\hspace{1em}\hspace{1em}increasing & w/ cov & all & no & spring & Decreasing Population & 0.22 & 0.07\\
 \hline
-Increasing Population & increasing & no cov & reduced & no & spring & 0.40 & 0.33\\
+\hspace{1em}\hspace{1em}increasing & w/ cov & all & yes & spring & Decreasing Population & 0.26 & 0.10\\
 \hline
-Increasing Population & increasing & no cov & reduced & yes & spring & 0.41 & 0.26\\
+\hspace{1em}\hspace{1em}increasing & no cov & all & no & fall & Decreasing Population & 0.28 & 1.06\\
 \hline
-Increasing Population & increasing & w/ cov & reduced & no & spring & 0.40 & 0.30\\
+\hspace{1em}\hspace{1em}increasing & no cov & all & yes & fall & Decreasing Population & 0.25 & 1.10\\
 \hline
-Increasing Population & increasing & w/ cov & reduced & yes & spring & 0.41 & 0.25\\
+\hspace{1em}\hspace{1em}increasing & w/ cov & all & no & fall & Decreasing Population & 0.28 & 0.15\\
 \hline
-Increasing Population & increasing & no cov & reduced & no & fall & 0.70 & 0.49\\
+\hspace{1em}\hspace{1em}increasing & w/ cov & all & yes & fall & Decreasing Population & 0.25 & 0.20\\
 \hline
-Increasing Population & increasing & no cov & reduced & yes & fall & 0.69 & 0.48\\
+\hspace{1em}\hspace{1em}increasing & no cov & reduced & no & spring & Decreasing Population & 0.32 & 0.14\\
 \hline
-Increasing Population & increasing & w/ cov & reduced & no & fall & 0.70 & 0.53\\
+\hspace{1em}\hspace{1em}increasing & no cov & reduced & yes & spring & Decreasing Population & 0.29 & 0.15\\
 \hline
-Increasing Population & increasing & w/ cov & reduced & yes & fall & 0.69 & 0.50\\
+\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & no & spring & Decreasing Population & 0.32 & 0.22\\
+\hline
+\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & yes & spring & Decreasing Population & 0.29 & 0.21\\
+\hline
+\hspace{1em}\hspace{1em}increasing & no cov & reduced & no & fall & Decreasing Population & 0.54 & 0.60\\
+\hline
+\hspace{1em}\hspace{1em}increasing & no cov & reduced & yes & fall & Decreasing Population & 0.53 & 0.62\\
+\hline
+\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & no & fall & Decreasing Population & 0.54 & 0.31\\
+\hline
+\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & yes & fall & Decreasing Population & 0.53 & 0.32\\
 \hline
 \end{tabular}
+\endgroup{}
+\end{table}
 

### 2022-12-26 VAST/Test_3Species_GeorgesBank_Had_Only.R

@@ -18,13 +18,13 @@ orig.dir <- getwd()
 # surv_random_sample <- readRDS(file="surv_random_sample.RDS")
 # surv_random_sample <- as.matrix(surv_random_sample,ncol= 12)
 
-scenario1 <- "IncPop_IncTemp"
+scenario1 <- "IncPop_ConTemp"
 #survey results without noise
 list_all <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Results\\",scenario1,"\\list_all_",scenario1,".RDS",sep=""))
 
 exclude_strata <- FALSE
 
-with_noise <- FALSE
+with_noise <- TRUE
 
 covariates <- FALSE
 
@@ -37,7 +37,7 @@ cov_used <- "_WithCov"  #"Temp_LinearBasic"
 #DecPop_ConTemp 6 is pretty good
 #DecPop_IncTemp 9 is pretty good
 
-surv_random_sample <- list_all[[100]]
+surv_random_sample <- list_all[[98]]
 
 
 setwd(orig.dir)
@@ -451,7 +451,7 @@ for(j in 1:6){
   
   
   FC1 = c("Omega1" = 0, "Epsilon1" =0, "Omega2" = 1, "Epsilon2" = 1) 
-  RhoConfig = c("Beta1" = 3, "Beta2" = 3, "Epsilon1" = 0, "Epsilon2" = 4)
+  RhoConfig = c("Beta1" = 3, "Beta2" = 3, "Epsilon1" = 0, "Epsilon2" = 2) #was 4 for most runs
   
   #FieldConfig = c("Omega1"=0, "Epsilon1"=0, "Omega2"="IID", "Epsilon2"=0
   
@@ -709,7 +709,7 @@ for(j in 1:6){
   #FC2 = c("Omega1" = 1, "Epsilon1" = 1, "Omega2" = 1, "Epsilon2" = 1) 
   
   FC2 = c("Omega1" = 0, "Epsilon1" =0, "Omega2" = 1, "Epsilon2" = 1) 
-  RhoConfig = c("Beta1" = 3, "Beta2" = 3, "Epsilon1" = 0, "Epsilon2" =4)
+  RhoConfig = c("Beta1" = 3, "Beta2" = 3, "Epsilon1" = 0, "Epsilon2" =2) #used to be 4
   
   settings <- make_settings(n_x = 500,
                             Region=example$Region,

### 2022-12-28 Manuscript/Draft1.Rmd
attemps at citations

@@
     keep_tex: yes
+  citation_package: biblatex
   bibliography: IBMWG.bib

### 2022-12-28 VAST/Test_3Species_GeorgesBank_Had_Only.R

@@ -451,7 +451,7 @@ for(j in 1:6){
   
   
   FC1 = c("Omega1" = 0, "Epsilon1" =0, "Omega2" = 1, "Epsilon2" = 1) 
-  RhoConfig = c("Beta1" = 3, "Beta2" = 3, "Epsilon1" = 0, "Epsilon2" = 2) #was 4 for most runs
+  RhoConfig = c("Beta1" = 3, "Beta2" = 3, "Epsilon1" = 0, "Epsilon2" = 0) #tried 2 #was 4 for most runs
   
   #FieldConfig = c("Omega1"=0, "Epsilon1"=0, "Omega2"="IID", "Epsilon2"=0
   

### 2023-01-03 Manuscript/Draft1.Rmd
updating manuscript

@@
 - Climate change is happening
+
  Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades, resulting in a shift in spatial distributions of many species.
@@
 
+```{r CodERROR,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
+library("knitr")
+library("kableExtra")
+library("dplyr")
+library("readr")
+library("tidyr")
+library("forcats")
+
+param_table1 = read_csv("Cod_results.csv")
+
+summary_tab1 <- param_table1 %>% 
+# Group the table according to Parameter, Scenario
+  group_by(Temp)%>%
+  mutate(Temp = factor(Temp)) %>%
+  arrange(Temp) %>%
+  select(Temp, everything())
+
+# Create table to present the data]
+Fig_3a <-  kable(summary_tab1[, -1], 
+  caption = "Cod error results.",
+    format = "latex", booktabs = TRUE) %>%
+  kable_styling(font_size = 10) %>%
+  pack_rows(tab_kable(), colnum = 1,
+    index = table(fct_inorder(summary_tab1$Temp), useNA = "no"))
+```
+
+```{r, results='asis'}
+Fig_3a
+```
+
 <!-- ```{r PMsallscenarios, echo=FALSE, message=FALSE, warnings=FALSE, out.width='7in', fig.cap=figcap1} -->

### 2023-01-03 Manuscript/Draft1.tex
attemps at citations
updating manuscript

@@
 \begin{itemize}
+\tightlist
 \item
   Climate change is happening
+\end{itemize}
+
 Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades, resulting in a shift in spatial distributions of many species.
@@
 We are therefore interested in analyzing the impact of climate change on the accuracy of future stock assessment models as measured by NOAA's ongoing bottom-trawl survey along the East coast.
+
+\begin{itemize}
+\tightlist
 \item
@@
 \centering
-\begin{tabular}[t]{l|l|l|l|r|r|r|l|l}
+\begin{tabular}[t]{l|l|l|l|r|r|r}
 \hline
-Temp & Strata & Noise & season & VAST.No.Cov & VAST.w..Cov & Stratified.Mean & X & X.1\\
+Temp & Strata & Noise & season & VAST.No.Cov & VAST.w..Cov & Stratified.Mean\\
 \hline
-const & all & no & spring & 0.11 & 0.12 & 0.36 & NA & \\
+constant & all & no & spring & 0.11 & 0.12 & 0.36\\
 \hline
-const & all & yes & spring & 0.12 & 0.15 & 0.35 & NA & Cod\\
+constant & all & yes & spring & 0.12 & 0.15 & 0.35\\
 \hline
-const & all & no & fall & 0.19 & 0.05 & 0.49 & NA & Decreasing Population\\
+constant & all & no & fall & 0.19 & 0.05 & 0.49\\
 \hline
-const & all & yes & fall & 0.30 & 0.23 & 0.41 & NA & \\
+constant & all & yes & fall & 0.30 & 0.23 & 0.41\\
 \hline
-const & reduced & no & spring & 0.17 & 0.24 & 0.41 & NA & \\
+constant & reduced & no & spring & 0.17 & 0.24 & 0.41\\
 \hline
-const & reduced & yes & spring & 0.20 & 0.23 & 0.46 & NA & \\
+constant & reduced & yes & spring & 0.20 & 0.23 & 0.46\\
 \hline
-const & reduced & no & fall & 0.21 & 0.33 & 0.60 & NA & \\
+constant & reduced & no & fall & 0.21 & 0.33 & 0.60\\
 \hline
-const & reduced & yes & fall & 0.18 & 0.31 & 0.58 & NA & \\
+constant & reduced & yes & fall & 0.18 & 0.31 & 0.58\\
 \hline
-increasing & all & no & spring & 0.12 & 0.15 & 0.25 & NA & \\
+increasing & all & no & spring & 0.12 & 0.15 & 0.25\\
 \hline
-increasing & all & yes & spring & 0.19 & 0.19 & 0.27 & NA & \\
+increasing & all & yes & spring & 0.19 & 0.19 & 0.27\\
 \hline
-increasing & all & no & fall & 0.76 & 0.13 & 0.45 & NA & \\
+increasing & all & no & fall & 0.76 & 0.13 & 0.45\\
 \hline
-increasing & all & yes & fall & 0.89 & 0.33 & 0.44 & NA & \\
+increasing & all & yes & fall & 0.89 & 0.33 & 0.44\\
 \hline
-increasing & reduced & no & spring & 0.14 & 0.22 & 0.32 & NA & \\
+increasing & reduced & no & spring & 0.14 & 0.22 & 0.32\\
 \hline
-increasing & reduced & yes & spring & 0.15 & 0.21 & 0.29 & NA & \\
+increasing & reduced & yes & spring & 0.15 & 0.21 & 0.29\\
 \hline
-increasing & reduced & no & fall & 0.60 & 0.31 & 0.54 & NA & \\
+increasing & reduced & no & fall & 0.60 & 0.31 & 0.54\\
 \hline
-increasing & reduced & yes & fall & 0.62 & 0.32 & 0.53 & NA & \\
+increasing & reduced & yes & fall & 0.62 & 0.32 & 0.53\\
 \hline
@@
 \hline
+\multicolumn{1}{l}{\textbf{all}}\\
+\hline
 \multicolumn{1}{l}{\textbf{Constant Population}}\\
-\hline
-\multicolumn{3}{l}{\textbf{all}}\\
 \hline
@@
 \hline
-\multicolumn{3}{l}{\textbf{reduced}}\\
+\multicolumn{1}{l}{\textbf{reduced}}\\
 \hline
@@
 
+\begin{table}
+
+\caption{\label{tab:CodERROR}Cod error results.}
+\centering
+\fontsize{10}{12}\selectfont
+\begin{tabular}[t]{lllrrr}
+\toprule
+Strata & Noise & season & VAST No Cov & VAST w/ Cov & Stratified Mean\\
+\midrule
+\addlinespace[0.3em]
+\multicolumn{1}{l}{\textbf{constant}}\\
+\hspace{1em}all & no & spring & 0.11 & 0.12 & 0.36\\
+\hspace{1em}all & yes & spring & 0.12 & 0.15 & 0.35\\
+\hspace{1em}all & no & fall & 0.19 & 0.05 & 0.49\\
+\hspace{1em}all & yes & fall & 0.30 & 0.23 & 0.41\\
+\hspace{1em}reduced & no & spring & 0.17 & 0.24 & 0.41\\
+\hspace{1em}reduced & yes & spring & 0.20 & 0.23 & 0.46\\
+\hspace{1em}reduced & no & fall & 0.21 & 0.33 & 0.60\\
+\hspace{1em}reduced & yes & fall & 0.18 & 0.31 & 0.58\\
+\addlinespace[0.3em]
+\multicolumn{1}{l}{\textbf{increasing}}\\
+\hspace{1em}all & no & spring & 0.12 & 0.15 & 0.25\\
+\hspace{1em}all & yes & spring & 0.19 & 0.19 & 0.27\\
+\hspace{1em}all & no & fall & 0.76 & 0.13 & 0.45\\
+\hspace{1em}all & yes & fall & 0.89 & 0.33 & 0.44\\
+\hspace{1em}reduced & no & spring & 0.14 & 0.22 & 0.32\\
+\hspace{1em}reduced & yes & spring & 0.15 & 0.21 & 0.29\\
+\hspace{1em}reduced & no & fall & 0.60 & 0.31 & 0.54\\
+\hspace{1em}reduced & yes & fall & 0.62 & 0.32 & 0.53\\
+\bottomrule
+\end{tabular}
+\end{table}
+
 \end{document}

### 2023-01-03 Manuscript/Draft1cml.Rmd
example of working references

--- 2022-12-26 Manuscript/Draft1.Rmd
@@
     keep_tex: yes
-  bibliography: IBMWG.bib
-
+  pdf_document:
+    toc: no
   word_document: default
   html_document: default
-  pdf_document:
-    toc: no
+bibliography: test.bib
+fontsize: 12pt
 header-includes:
@@
 - \usepackage[belowskip=0pt,aboveskip=0pt]{caption}  
- \usepackage{array}
- \usepackage{caption}
- \usepackage{graphicx}
- \usepackage{siunitx}
- \usepackage{colortbl}
- \usepackage{multirow}
- \usepackage{hhline}
- \usepackage{calc}
- \usepackage{tabularx}
- \usepackage{tabulary}
- \usepackage{threeparttable}
- \usepackage{wrapfig}
 ---
@@
 
-much of below is from https://apps-nefsc.fisheries.noaa.gov/nefsc/ecosystem-ecology/
-or
-https://www.fisheries.noaa.gov/data-tools/fisheries-economics-united-states-data-and-visualizations
-
-- The eastern continental shelf is ecologically diverse and economically important
-
-The Northeast United States continental shelf spans from the Outer Banks of North Carolina to the Gulf of Maine.
-The region covers over 250,000 km$^2$ of ocean, extending over 200 km from shore in the largest areas in New England to just 30 km off shore in the southern regions.
-This ecologically diverse region contains approximately 18,000 vertebrate marine species.
-Commercial fisheries have been an important part of local economies for centuries.
-In 2019, New England fisheries produced $22 billion in sales, which sustained over 200,000 jobs.
-Maintaining a healthy ecosystem is therefore vital to sustained ecological health and economic prosperity of the region.
-[@nefmc20]
-
-- Bottom trawl survey is important for monitoring population trends
-
-Fish stocks in this highly productive and economically important region are managed by the National Oceanic and Atmospheric Administration’s (NOAA) Northeast Fisheries Science Center (NEFSC) in Woods Hole, Massachusetts.
-Federal scientists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (cite survey paper).
-The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf.
-The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits **just listed a few uses of survey.
-change/add others?**.
-
-The survey takes place twice each year- once in the spring and again in the fall.
-Since most spatial analyses and  projections of future distributions typically assume a constant survey catchability and/or availability over time, NOAA's survey design includes sampling during approximately the same 2-3 week time period in each season.
-
-- Climate change is happening
- Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades, resulting in a shift in spatial distributions of many species.
-Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments.
-The implication of this is that the bottom trawl survey is actually sampling the population during a different life cycle stage than was originally assumed, which can lead to biased stock assessments.
-We are therefore interested in analyzing the impact of climate change on the accuracy of future stock assessment models as measured by NOAA’s ongoing bottom-trawl survey along the East coast.
-
-- say something about our study species? Past abundance history, biological characteristics, others? Maybe better suited in below section
-
-**use more info from initial proposal**
-
-- Fish are changing spatial distribution and have altered life stages (?) because of climate change
-NYE paper
-
-- Population indexing methods may be becoming biased as a result
-
-- Briefly describe our study to test this
-
-To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depend on temperature preferences.
-We can then consider the impact of climate change by simulating scenarios with repeating temperature patterns and those where temperature increases on average over time.
-In both cases we analyze the ability of stratified random sampling to track population trends.
-
 \section{Methods}
-
-- Describe simulation study
-
-We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on George's Bank, where movement of each species combine static species-specific habitat preferences with temperature preferences.
-Model dynamics are driven by a time series of temperature gradients that were estimated from data to create simulated data sets for each population where the true biomass is known.
-Using temperature gradients that repeat each year creates data sets with predictable, repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time.
-We conducting stratified random sampling on our simulation output to mimic the bottom trawl survey and compare the ability of contemporary indexing methods to track population trends.
-
-\subsection{Population Model Formulation}
 
@@
 We formulate the following inputs for the MFS package to address our research question.
-
-<font size="5">*Study Area*</font>
-
-We obtained a shapefile for the 15 strata that comprise George's Bank to use as our modeling environment.
-We discritized the region into a raster with 88 rows and 144 columns.
-Haddock inhabit all 15 strata in the domain, Cod inhabit 13 strata, and Yellowtail exist in 9 strata.
-Figure \@ref(fig:strata-plot) shows the regions used in our models.
-
-```{r strata-plot, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Strata inhabited by each species in our population models."}
-knitr::include_graphics(c("Images/Strata.png"))
-```
-
-<font size="5">*Population Dynamics and Recruitment*</font>
-
-The time step for our models is one week.
-MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass in each cell in our study area [@dolder2020highly].
-Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits.
-We chose to represent recruitment in the model using a Beverton-Holt formulation **cite**.
-Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.
-Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics.
-A full list of parameters used in our model can be seen below in Tables \@ref(paramsALL) and \@ref(tab:paramsSCENARIOS).
-
-<font size="5">*Movement*</font>
-
-The package was designed to generate theoretical habitat preferences using Gaussian Random Fields that combine with hypothetical temperature gradients to drive the probability of movement from cell $I$ to cell $J$ using the formulation
-
-\begin{align}
-Pr(C_{wk+1}=J|C_{wk}=I) = \frac{e^{-\lambda \cdot d_{I,J}}\cdot(Hab^2_{J,s} \cdot Tol_{J,s,wk})}{\sum^C_{c=1}e^{-\lambda \cdot d} \cdot (Hab^2_{c,s} \cdot Tol_{c,s,wk})},
-\label{moveP}
-\end{align}
-
- where
-
-   $e^{-\lambda \cdot d_{I,J}}$ accounts for distance between cells $I$ and $J$,
-
-  $Hab^2_{J,s}$ is the static habitat value for species $s$ in cell $J$, and
-
-   $Tol_{c,s,wk}$ is the value from normally distributed temperature tolerance for species $s$ in cell $c$ in week $wk$.
-
-The following sections describe how we formulated the habitat and temperature components to model real species on the northeast continental shelf.
-
-<font size="5">*Habitat Input*</font>
-
-Species-specific habitat preferences were derived using the *lrren* tool from the R package *envi* [@envi] to create a niche model for each species.
-The *lrren* tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors.
-We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow **cite trawl data?**.
-We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences.
-Depth and mean sediment size were used as our covariate predictors.
-Estimated depth for the region was obtained from FVCOM [@chen2006unstructured].
-The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method **cite arcmap or more?** using point data collected by the United States Geologic Survey (USGS) [@mcmullen2005usgs].
-Since the values in $Hab^2_{J,s}$ are required to be between 0 and 1, we transform the spatial estimates from *lrren* to fall between these bounds.
-See Figure \@ref(fig:hab-plot1) for a visual representation of this process being applied to Cod.
-Figure \@ref(fig:hab-plot) depicts habitat preferences $Hab^2_{J,s}$ for each species.
-
-```{r hab-plot1, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Visual representation of niche model for Cod."}
-knitr::include_graphics(c("Images/hab_snip.png"))
-```
-
-```{r hab-plot, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Static habitat preferences for each species in our population models (Yellotwtail, Cod, Haddock)."}
-knitr::include_graphics(c("Images/Habitat_3species.png"))
-```
-
-<font size="5">*Temperature Input*</font>
-
-Each species is assumed to have normally distributed temperature preferences ($N(\mu,\sigma)$).
-We assume Yellowtail Flounder's preferences are $N(8.75,4.25)$, while Haddock and Cod have preferences $N(9,4)$.
-We chose these values by combining information in the literature with temperatures recorded in the bottom trawl survey.
-Weekly estimated temperature data for the region for 2012 was obtained from FVCOM [@chen2006unstructured].
-We chose 2012 because the data displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values.
-This data was also transform to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation.
-**show images of temperature and/or and/or temp videos???
-and/or average temperature oscillations?**
-
---Describe difference between increasing and constant temperature scenarios (images?)
-
-In equation (1), $Hab^2_{J,s}$ is constant for the duration of the simulation, while $Tol_{c,s,wk}$ changes each week.
-Using a temperature gradient that repeats every 52 weeks produces the same spatial preferences in a given week each year, resulting in consistent spatial biomass patterns.
-Scenarios where the temperature increases over time creates spatial preferences that evolve as the water warms, producing spatial biomass patterns that shift in a given week over the duration of the simulation.
-Thus, stratified random samples in scenarios with a repeating temperature pattern will have constant survey catchability and availability over time, which may not be true for increasing temperature scenarios due to evolving spatial preferences.
-
--- Describe each scenario that is considered
-
-We consider 20 year simulations under three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time.
-Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius over the duration of the 20 year simulation.
-We therefore simulate a total of 6 scenarios.
-**show line plot of population for each?**
-
-<font size="5">*Simulating Bottom Trawl Survey and Population Indexing*</font>
-
--Describe post hoc sampling process and how data is used
-
-After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year.
-We sample in the same weeks that the Spring and Fall surveys take place and the number of the samples taken in each strata reflect true values.
-Most strata contain enough cells to sample a unique location in each survey over the duration of the simulation.
-For smaller strata we must repeat some sample locations.
-We then use the biomass collected from our samples in contemporary population indexing methods to estimate population trends.
-Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.
-
---Stratified mean vs VAST with and without covariates
-
-We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
-The stratified mean is a typical survey-based approach that scales individual samples to the  strata-level by considering the area of each strata, before scaling to the region-level based on the relative size of each strata.
-VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurrence (presence/absence).
-If desired, VAST also allows users to include covariate data to better inform the model.
-Covariates can be static (eg.
-habitat preferences) or dynamics (eg.
-temperature).
-The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and take on the order of hours to complete.
-
-We follow the advice given in [@thorson2019guidance] to build VAST models to estimate biomass on George's Bank using stratified mean samples from our model output.
-In addition to exploring different link functions and assumed distributions, our VAST model-building process included testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
-We also carried out the same process both including covariates in our model as well as running models without covariate information.
-We considered covariates in the form of dynamic temperature values and/or static habitat values from our population model.
-When using covariates we ultimately decided to provide the most information to the model by including both covariates for both linear predictors.
-Since we know the true population values in our models we calculate the absolute error of each VAST estimate to compare between potential settings.
-Through this process, and in consultation with the VAST package creator, we determined setting that allowed VAST models to converge for all of our scenarios while also providing the lowest absolute error values.
-Settings for our VAST models can be seen in Table ???.
-
-Our goal is to determine indexing approaches and settings that are robust to future environmental conditions and resulting spatial biomass patterns.
-An underlying assumption in all indexing methods that individual random samples combine to accurately represent true abundance by a) sampling all strata in which the population exists and b) low enough noise level in the samples to allow for a discernible pattern.
-This assumption can be questioned given enough noise in the sampling process **cite?** and/or shifting spatial preferences driven by climate change causing a population to move into a previously uninhabited strata.
-To simulate the impact of noise, indexing estimates after adding noise to our samples versus those using the true sampling values.
-To evaluate the effect of populations moving into new habitat, we compare indexing estimates using samples from all strata versus those that only include a subset of the full spatial domain for each species.
-
-When combining population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, covariates) there are a large number of scenario combinations to consider.
-The columns in Table COMBOS show the choices that define each scenario.
-
-<div align="center">Table COMBOS Each index estimate chooses one condition from each of the following columns.
-There are $3*3*2*2*2*2*2=288$ VAST model combinations and $3*3*2*2*2=144$ stratified mean estimates.</div>
-
-```{r scenarios, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-library("knitr")
-library("kableExtra")
-library("dplyr")
-library("readr")
-library("tidyr")
-library("forcats")
-
-tabl <- "
-|	Species	|	Population Trend | Temperature Scenario | Strata Included | Noise in Data | Covariates (VAST only) | Season
-|:--	|:-- |:-- |:-- |:-- |:--- |:--
-|Yellowtail | Increasing | Repeating | All strata | No Noise| No Covariates | Spring
-|Cod | Constant | Increasing 5$^{\\circ}$| Subset | Yes Noise| Temp + Habitat | Fall
-|Haddock | Decreasing | | | | |
-"
-cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
-
-```
-
-\section{Results}
-
-The goal of our project was to analyze how well contemporary population indexing methods can track population trends under a host of conditions, as depicted in Table COMBOS.
-Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year **cite**.
-For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod (see Table ???) and increasing population scenarios for Haddock (see Table ???).
-To provide a comprehensive analysis of population indexing methods we consider all possible scenario combinations for Yellowtail Flounder (see Table ???).
-
-General themes that exist in Tables X, Y and Z are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors.
-We also see individual cases where the stratified mean produced the lowest absolut error and instances where including covarites in VAST models actually increase the absolute error.
-When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error, though there are some scenarios where the impact is minimal.
-
-In the Cod results, when using reduced strata adding covariates produces worst VAST results (though still better than stratified mean).
-VAST without covariates much worse than stratified mean in fall with increasing temperature and all strata, but adding covariates corrects this.
-
-For Haddock, VAST has a particularly hard time in spring regularly producing larger errors than the stratified mean with added covariates only improving to the level of the stratified mean.
-VAST shows improved results in fall relative to the stratified mean with added covaraites producing extremely low errors in some cases.
-
-In considering the Yellowtail Flounder results in Table XXX, we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
-However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in the Table.
-These errors are corrected by including covariate information allowing for an improved VAST estimate that are significantly lower than their stratified mean counterparts.
-
-```{r}
-
-knitr::kable(YT_results, caption = "Yellowtail error results")
-
-knitr::kable(Cod_results, caption = "Cod error results")
-
-knitr::kable(Had_results, caption = "Haddock error results")
-```
-
-\section{Discussion}
-
-\section{Acknowledgements}
-
-## Data and Code Availability
-All data and code used in this work are available at https://github.com/cmlegault/IBMWG.
 
@@
 </font>
-
-\pagebreak
-
-<div align="center">Table 1. Parameters used in all population models.</div>
-
-```{r paramsALL, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-library("knitr")
-library("kableExtra")
-library("dplyr")
-library("readr")
-library("tidyr")
-library("forcats")
-
-tabl <- "
-|	Parameter	|	Description | Unit |Yellowtail | Cod | Haddock | Source
-|:---	|:------- |:-- |:--- |:-- |:-- |:--
-| $\\rho$ | Ford's growth coefficient | wk$^{-1}$  | 4.48 | 4.43 | 4.49 | |
-| $M$ | Natural Mortality| wk$^{-1}$  | 0.2064 | 0.2728 | 0.3340 | |
-| $W_R$ |Weight of fully recruited fish| kg | 0.39 | 2.95 | 1.12 | |
-| $W_{R-1}$ | Weight of pre-recruit fish| kg | 0.13 | 0.39 | 0.19 | |
-| $\\sigma^2$ | Variance in recruited fish | kg$^2$ | 0.55 | 0.55 | 0.55 ||
-| $\\lambda$ | Decay rate for movement | - | 0.7 | 0.7 | 0.7 ||
-| $Spwn_s$ | Spawning weeks for species $s$ | wk | 9-12 | 8-13 | 11-14 ||
-| $Rec_s$ | Recruitment weeks for species $s$ | wk | 9-12 | 8-13 | 11-14 ||
-"
-cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
-
-```
-
-```{r paramsALL2,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-library("knitr")
-library("kableExtra")
-library("dplyr")
-library("readr")
-library("tidyr")
-library("forcats")
-
-param_table1 = read_csv("Parameters1.csv")
-
-summary_tab1 <- param_table1 %>% 
- # dplyr::mutate_if(is.numeric, list(~as.character(signif(., 3))))%>%
-
-  mutate_if(grepl('rrr',.), ~replace(., grepl('rrr', .), paste("\U03C1")))#should be replacing r with the rho symbol
- # mutate_if(grepl('W1',.), ~replace(., grepl('W1', .), "WW")) #should be replacing beta with the symbol
- #    mutate_if(grepl('wk',.), ~replace(., grepl('wk', .), "1/wk")) %>% #should be replacing beta with the symbol
- #
-
-# Change the orientation of the table so that Species become the variable headings
-
-# Create table to present the data]
-Fig_1 <- kable(summary_tab1, 
-  caption = "Parameters used in all population models.",
-    format = "latex", booktabs = TRUE) %>%
-  kable_styling(font_size = 10)
-
-```
-
- ```{r, results='asis'}
-
- Fig_1
-
- ```
-
-<!-- Table 2. Parameters used in models with relatively constant populations. -->
-<!-- ```{r paramsCON, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'} -->
-<!-- tabl <- " -->
-<!-- |		|	Description | Yellowtail | Cod | Haddock  -->
-<!-- |:---	|:------- |:--- |:-- |:-- |:-- -->
-<!-- | $M+F$ | Adjusted Mortality (Natural + Fishing)| wk$^{-1}$  | 0.764 | 0.83 | 0.309 |  -->
-
-<!-- " -->
-<!-- cat(tabl) # output the table in a format good for HTML/PDF/docx conversion -->
-<!-- ``` -->
-
-<!-- Table 3. Parameters used in models with increasing populations. -->
-<!-- ```{r paramsINC, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'} -->
-<!-- tabl <- " -->
-<!-- |		|	Description | Yellowtail | Cod | Haddock | Source -->
-<!-- |:---	|:------- |:--- |:-- |:-- |:-- -->
-<!-- | $M+F$ | Adjusted Mortality (Natural + Fishing)| wk$^{-1}$  | 0.564 | 0.372 |  0.134 |  -->
-
-<!-- " -->
-<!-- cat(tabl) # output the table in a format good for HTML/PDF/docx conversion -->
-<!-- ``` -->
-
-<!-- Table 4. Parameters used in models with decreasing populations. -->
-<!-- ```{r paramsDEC, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'} -->
-<!-- tabl <- " -->
-<!-- |		|	Description | Yellowtail | Cod | Haddock | Source -->
-<!-- |:---	|:------- |:--- |:-- |:-- |:-- -->
-<!-- | $M+F$ | Adjusted Mortality (Natural + Fishing)| wk$^{-1}$  | 0.764 | 0.623 | 0.334 |  -->
-
-<!-- " -->
-<!-- cat(tabl) # output the table in a format good for HTML/PDF/docx conversion -->
-<!-- ``` -->
-
-<!-- DOESNT WORK!!
-<!-- Table 2. Parameters used in population models for each scenario. -->
-<!-- ```{r paramsSCENARIOS, echo=FALSE, message=TRUE, warnings=TRUE, results='asis'} -->
-<!-- library(huxtable) -->
-
-<!-- ht <- tribble_hux( -->
-<!--   ~ "Section/topic", ~ "Item no", ~ "Checklist item", ~ "Reported on Page No", -->
-<!--   "Title"          , ""         , ""                , "", -->
-<!--   "Title"          , "1"        , "Identify..."     , "",  -->
-<!--   "Abstract"       , ""         , ""                , "", -->
-<!--   "Structured summary", "2"     , "Provide..."      , "" -->
-<!--   # et cetera... -->
-<!-- )  -->
-
-<!-- # using the pipe from 4.1.0... -->
-
-<!-- ht |>  -->
-<!--   set_header_rows(c(2, 4), TRUE)                |> -->
-<!--   merge_across(c(2, 4), everywhere)             |> -->
-<!--   style_header_rows(bold = TRUE)                |> -->
-<!--   set_all_borders(brdr(0.4, "solid", "grey70")) |> -->
-<!--   set_background_color("grey97")                |> -->
-<!--   set_background_color(1, 1:3, "grey90")        |> -->
-<!--   set_col_width(c(0.2, 0.05, 0.55, 0.2))        |> -->
-<!--   set_font("cmss")                               -->
-<!-- ``` -->
-
-<!-- ```{r,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'} -->
-<!-- library("knitr") -->
-<!-- library("kableExtra") -->
-<!-- library("dplyr") -->
-<!-- library("readr") -->
-<!-- library("tidyr") -->
-<!-- library("forcats") -->
-
-<!-- #orig_ui_deaths = read_csv("https://www.opendata.nhs.scot/dataset/b0135993-3d8a-4f3b-afcf-e01f4d52137c/resource/89807e07-fc5f-4b5e-a077-e4cf59491139/download/ui_deaths_2020.csv") -->
-
-<!-- summary_tab <- orig_ui_deaths %>% -->
-<!--   # Apply filters in same was as in plots so looking at one year, whole of Scotland, and exlucing "All" entries -->
-<!--   filter(Year == "2018", HBR == "S92000003", AgeGroup != "All" & Sex != "All", InjuryType != "Accidental exposure" & InjuryType != "All") %>% -->
-<!-- # Group the table according to injury type, age, and sex -->
-<!--   group_by(InjuryType, AgeGroup, Sex) %>% -->
-<!-- # Create a summary of total number of deaths for neater appearance to the table and demonstration of the figures -->
-<!--   summarise(total_deaths = sum(NumberofDeaths)) %>% -->
-<!-- # Change the orientation of the table so that age groups become the variable headings -->
-<!--   pivot_wider(names_from = AgeGroup, values_from = total_deaths) %>% -->
-<!--   mutate(Sex = factor(Sex)) %>% -->
-<!--   arrange(Sex) %>% -->
-<!--   select(Sex, everything()) -->
-
-<!-- # Create table to present the data] -->
-<!-- Fig_2 <- kable(summary_tab[, -1],  -->
-<!--   caption = "Deaths from unintentional injuries in Scotland", -->
-<!--     format = "latex", booktabs = TRUE) %>% -->
-<!--   kable_styling(font_size = 10) %>% -->
-<!--   pack_rows(tab_kable, colnum = 1, -->
-<!--     index = table(fct_inorder(summary_tab$Sex), useNA = "no")) -->
-<!-- ``` -->
-<!-- ```{r, results='asis'} -->
-<!-- Fig_2 -->
-<!-- ``` -->
-
-```{r paramsSCENARIOS,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-library("knitr")
-library("kableExtra")
-library("dplyr")
-library("readr")
-library("tidyr")
-library("forcats")
-
-param_table = read_csv("Parameters.csv")
-
-summary_tab <- param_table %>% 
-  dplyr::mutate_if(is.numeric, list(~as.character(signif(., 3))))%>%
-
- mutate_if(grepl('alpha',.), ~replace(., grepl('alpha', .), paste("\U03B1"))) %>% #should be replacing alpha with the symbol
-  mutate_if(grepl('beta',.), ~replace(., grepl('beta', .), paste("\U03B2"))) %>% #should be replacing beta with the symbol
-    mutate_if(grepl('wk',.), ~replace(., grepl('wk', .), "1/wk")) %>% #should be replacing beta with the symbol
-
-# Group the table according to Parameter, Scenario
-  group_by(Parameter, Scenario)%>%
-# Change the orientation of the table so that Species become the variable headings
-  pivot_wider(names_from = Species,  values_from = Value) %>%
-  mutate(Scenario = factor(Scenario)) %>%
-  arrange(Scenario) %>%
-  select(Scenario, everything())
-
-# Create table to present the data]
-Fig_3 <- kable(summary_tab[, -1], 
-  caption = "Parameters used in population models for each scenario.",
-    format = "latex", booktabs = TRUE) %>%
-  kable_styling(font_size = 10) %>%
-  pack_rows(tab_kable, colnum = 1,
-    index = table(fct_inorder(summary_tab$Scenario), useNA = "no"))
-```
-```{r, results='asis'}
-Fig_3
-```
-
-<div align="center">Table XX. Parameters used in all VAST models.</div>
-
-```{r VASTsettings, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-
-tabl <- "
-|	Parameter	|	Description | Input/Value
-|:-	|:------- |:----
-| *ObsModel* | Link function and assumed distribution | c(10,2)
-|FieldCOnfig | Specified spatial and/or spatio-temporal variation in predictors| c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1)
-| RhoConfig | Specifying whether intercepts or spatio-temporal variation is structured among time intervals | c(Beta1=3, Beta2=3, Epsilon1=0, Epsilon2=4)
-| X1_formula | Right-sided formula affecting the 1st linear predictor  | X1_formula = ~ poly(Temp, degree=2 )
-| X2_formula | Right-sided formula affecting the 2nd linear predictor | X2_formula = ~ poly(Temp, degree=2 ) + poly(Habitat, degree=2 )
-
-"
-cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
-
-```
-
-```{r YT_results,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-library("knitr")
-library("kableExtra")
-library("dplyr")
-library("readr")
-library("tidyr")
-library("forcats")
-
-result_table = read_csv("Yellowtail_results_final.csv")
-
-summary_tab <- result_table %>%
-
-# Group the table according to Parameter, Scenario
-  group_by(Scenario,Strata)
-
-# Create table to present the data]
-Fig_5 <- kable(summary_tab) %>%
-  kable_styling(font_size = 10)  %>%
-  pack_rows(tab_kable, colnum = c(1,4),
-    index = table(fct_inorder(summary_tab$Strata), useNA = "no")) %>%
-  pack_rows(tab_kable, colnum = 1,
-    index = table(fct_inorder(summary_tab$Scenario), useNA = "no"))
-
-```
-```{r, results='asis'}
-Fig_5
-```
-
-<!-- ```{r PMsallscenarios, echo=FALSE, message=FALSE, warnings=FALSE, out.width='7in', fig.cap=figcap1} -->
-<!-- #knitr::include_graphics(file.path(getwd(),"MS_tables_figs_rev1/final_ms_figs","Figure_1.pdf")) -->
-<!-- ``` -->
-
-<!-- ```{r SSmsyvsCMSY, echo=FALSE, message=FALSE, warnings=FALSE, out.width='7in', fig.cap=figcap2} -->
-<!-- #knitr::include_graphics(file.path(getwd(),"MS_tables_figs_rev1/final_ms_figs","Figure_2.pdf")) -->
-<!-- ``` -->
-
-<!-- ```{r PMsbyretroF, echo=FALSE, message=FALSE, warnings=FALSE, out.width='6in', fig.cap=figcap3} -->
-<!-- #knitr::include_graphics(file.path(getwd(),"MS_tables_figs_rev1/final_ms_figs","Figure_3.pdf")) -->
-<!-- ``` -->
-
-<!-- ```{r FFmsybyscenario, echo=FALSE, message=FALSE, warnings=FALSE, out.width='6in', fig.cap=figcap4} -->
-<!-- #knitr::include_graphics(file.path(getwd(),"MS_tables_figs_rev1/final_ms_figs","Figure_4.pdf")) -->
-<!-- ``` -->
-
-<!-- ```{r linearvsdiffuse, echo=FALSE, message=FALSE, warnings=FALSE, out.width='7in', fig.cap=figcap5} -->
-<!-- #knitr::include_graphics(file.path(getwd(),"MS_tables_figs_rev1/final_ms_figs","Figure_5.pdf")) -->
-<!-- ``` -->

### 2023-01-04 Manuscript/Draft1.Rmd
Chris initial comments

@@
 
+[[Chris comments in double square brackets, search for them to see comments]]
+
 <!-- Table1 <- YT_results -->
@@
 Fish stocks in this highly productive and economically important region are managed by the National Oceanic and Atmospheric Administration’s (NOAA) Northeast Fisheries Science Center (NEFSC) in Woods Hole, Massachusetts.
-Federal scientists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (cite survey paper).
+Federal scientists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (cite survey paper[[, some possible citations are https://repository.library.noaa.gov/view/noaa/4825/noaa_4825_DS1.pdf and T.S.
+Azarovitz 1981 A brief historical review of the Woods Hole Laboratory trawl survey time series.
+W.G.
+Doubleday, D.
+Rivard (Eds.), Bottom Trawl Surveys, Can.
+Spec.
+Pub.
+Fish.
+Aqua.
+Sci.
+No.
+58 (1981), pp.
+62-67]]).
 The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf.
@@
 Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments.
-The implication of this is that the bottom trawl survey is actually sampling the population during a different life cycle stage than was originally assumed, which can lead to biased stock assessments.
+The implication of this is that the bottom trawl survey is actually sampling the population during a different life cycle stage than was originally assumed [[I don't think the concern is about life cycle stage, but rather that the fish catchability or availability in the survey will change, thus changing the relationship between the index and the true population]], which can lead to biased stock assessments.
 We are therefore interested in analyzing the impact of climate change on the accuracy of future stock assessment models as measured by NOAA’s ongoing bottom-trawl survey along the East coast.
 
-- say something about our study species? Past abundance history, biological characteristics, others? Maybe better suited in below section
+- say something about our study species?
+Past abundance history, biological characteristics, others?
+Maybe better suited in below section [[I think the specific species can wait for the methods section, just laying the groundwork here in the intro]]
 
@@
 
-- Fish are changing spatial distribution and have altered life stages (?) because of climate change
+- Fish are changing spatial distribution and have altered life stages (?) because of climate change [[see note above about catchability and availability]]
 NYE paper
@@
 
-We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on George's Bank, where movement of each species combine static species-specific habitat preferences with temperature preferences.
+We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on [[Georges]] George's Bank, where movement of each species combine static species-specific habitat preferences with temperature preferences.
 Model dynamics are driven by a time series of temperature gradients that were estimated from data to create simulated data sets for each population where the true biomass is known.
-Using temperature gradients that repeat each year creates data sets with predictable, repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time.
+Using temperature gradients that repeat each year creates data sets with predictable, repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time [[and mimics the possible future in the region under climate change?]].
 We conducting stratified random sampling on our simulation output to mimic the bottom trawl survey and compare the ability of contemporary indexing methods to track population trends.
@@
 We formulate the following inputs for the MFS package to address our research question.
+[[perhaps for discussion, but we should mention somewhere that MFS only tracks biomass not age structure, the latter is important for stock assessments but not considered in this paper]]
 
@@
 
-We obtained a shapefile for the 15 strata that comprise George's Bank to use as our modeling environment.
+We obtained a shapefile for the 15 strata that comprise [[Georges Bank does not have an apostrophe, I made the same mistake when I started here]] George's Bank to use as our modeling environment.
 We discritized the region into a raster with 88 rows and 144 columns.
@@
 Figure \@ref(fig:strata-plot) shows the regions used in our models.
+[[can mention that the different strata are based on biological differences among the species, possibly add that there are multiple stocks for all three species and the delineations among the stocks differ by species]]
 
@@
 Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits.
-We chose to represent recruitment in the model using a Beverton-Holt formulation **cite**.
+[[not sure we need to highlight the recruitment model, might be able to just use the table because Bev-Holt formulation is well known in fisheries]] We chose to represent recruitment in the model using a Beverton-Holt formulation **cite**.
 Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.
@@
 The *lrren* tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors.
-We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow **cite trawl data?**.
+We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow [[I don't think we need to cite the trawl data because we said we were using trawl data above]] **cite trawl data?**.
 We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences.
@@
 Estimated depth for the region was obtained from FVCOM [@chen2006unstructured].
-The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method **cite arcmap or more?** using point data collected by the United States Geologic Survey (USGS) [@mcmullen2005usgs].
+The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method [[citing ArcMap should suffice]] **cite arcmap or more?** using point data collected by the United States Geologic Survey (USGS) [@mcmullen2005usgs].
 Since the values in $Hab^2_{J,s}$ are required to be between 0 and 1, we transform the spatial estimates from *lrren* to fall between these bounds.
@@
 **show images of temperature and/or and/or temp videos???
-and/or average temperature oscillations?**
+and/or average temperature oscillations?**[[as we discussed, one plot here comparing average temperatures for constant and increasing scenarios in the paper, one or more animations in supplemental materials]]
 
@@
 Thus, stratified random samples in scenarios with a repeating temperature pattern will have constant survey catchability and availability over time, which may not be true for increasing temperature scenarios due to evolving spatial preferences.
+[[may want to mention here or in the discussion that we used repeating 2012 to simplify the simulations and reduce the number of factors explored that would have occurred with different annual cycles]]
 
@@
 We consider 20 year simulations under three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time.
-Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius over the duration of the 20 year simulation.
+Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius [[need some justification for this large change, wanted to be extreme because if the extreme doesn't make a difference, then more reasonable increases wouldn't make a difference]] over the duration of the 20 year simulation.
 We therefore simulate a total of 6 scenarios.
@@
 We sample in the same weeks that the Spring and Fall surveys take place and the number of the samples taken in each strata reflect true values.
-Most strata contain enough cells to sample a unique location in each survey over the duration of the simulation.
+[[need to make sure that we don't lose sight of random samlping with the following sentences]] Most strata contain enough cells to sample a unique location in each survey over the duration of the simulation.
 For smaller strata we must repeat some sample locations.
 We then use the biomass collected from our samples in contemporary population indexing methods to estimate population trends.
-Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.
+Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.[[should we mention here or later that we did both with and without measurement error analyses?]]
 
@@
 
-We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
+[[could use an intro sentence here to distinguish design-based (stratified mean) and model-based (VAST) estimators]] We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
 The stratified mean is a typical survey-based approach that scales individual samples to the  strata-level by considering the area of each strata, before scaling to the region-level based on the relative size of each strata.
@@
 temperature).
-The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and take on the order of hours to complete.
+The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and take on the order of hours to complete.[[may want to highlight the potential of VAST with covariates to address the climate change issue here as the reason we are doing these simulations]]
 
-We follow the advice given in [@thorson2019guidance] to build VAST models to estimate biomass on George's Bank using stratified mean samples from our model output.
+We follow the advice given in [@thorson2019guidance] to build VAST models to estimate biomass on [[Georges]] George's Bank using stratified mean samples from our model output.
 In addition to exploring different link functions and assumed distributions, our VAST model-building process included testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
@@
 To evaluate the effect of populations moving into new habitat, we compare indexing estimates using samples from all strata versus those that only include a subset of the full spatial domain for each species.
+[[may need to spell the aspect of reduced spatial domain a bit more because it might be confusing to readers who are expected us to just add areas around the current ones instead of reducing the strata]]
 
@@
 
+#[[note: the library calls can all be made in the first block of R code instead, it sometimes makes it easier to find them if they are all in one place, this approach of defining them within separate blocks works too, just noting a possible different approach]]
+
 tabl <- "
@@
 The goal of our project was to analyze how well contemporary population indexing methods can track population trends under a host of conditions, as depicted in Table COMBOS.
-Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year **cite**.
+Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year [[can cite the 2022 management track assessments, see https://apps-nefsc.fisheries.noaa.gov/saw/sari.php for when the document becomes available]] **cite**.
 For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod (see Table ???) and increasing population scenarios for Haddock (see Table ???).
@@
 We also see individual cases where the stratified mean produced the lowest absolut error and instances where including covarites in VAST models actually increase the absolute error.
-When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error, though there are some scenarios where the impact is minimal.
+When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error [[as expected]], though there are some scenarios where the impact is minimal.
 
@@
 
+[[We'll want to expand the results section to focus on the impact of each factor one at at time within each species.
+This will be a bit dull, but it is important to walk through the results in words to ensure that readers get the message.
+They can of course examine the tables in detail and draw their own conclusions, but we should put our interpretation down on paper.]]
+
 ```{r}
@@
 \section{Discussion}
+[[ some things that I think we'll need to address, in addition to those mentioned above are (in no particular order):
+number of simulations done for each scenario - time required to do them and analyses
+pros and cons of using VAST with covariates in a changing environment
+implications for current practice of using stratified mean (not horrible, but might be able to do better with model-based using covariates if enviro changing, note we don't want to say stratified mean should not be used because that will draw a lot of tomato throwing)
+future work - some of the ideas we've discussed already could be mentioned
+limitations to our study
+]]
 
 \section{Acknowledgements}
+[[we should thank Jim (obviously), but also those who helped you with the habitat data and others?
+also should note that funding provided by Climate-Groundfish source (I'll dig up the official name of the funding source)]]
 
 ## Data and Code Availability
-All data and code used in this work are available at https://github.com/cmlegault/IBMWG.
+All data and code used in this work are available at https://github.com/cmlegault/IBMWG. [[need to update this]]
 

### 2023-01-05 Manuscript/Draft1.Rmd
test

--- 2023-01-03 Manuscript/Draft1.Rmd
@@
     keep_tex: yes
-  citation_package: biblatex
-  bibliography: IBMWG.bib
-
+  pdf_document:
+    toc: no
   word_document: default
   html_document: default
-  pdf_document:
-    toc: no
+bibliography: IBMWG.bib
+fontsize: 12pt
 header-includes:
@@
 ```{r hab-plot1, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Visual representation of niche model for Cod."}
-knitr::include_graphics(c("Images/hab_snip.png"))
+knitr::include_graphics(c("Images/hab_snip3.png"))
 ```

### 2023-01-05 Manuscript/Draft1.Rmd

--- 2023-01-04 Manuscript/Draft1.Rmd
@@
 <!-- ``` -->
+>>>>>>> fac774b6b2cab2991e9875992ad8ca7eb3de1326

### 2023-01-05 Manuscript/Draft1.tex

@@
 \documentclass[
+  12pt,
 ]{article}
@@
 
+{[}{[}Chris comments in double square brackets, search for them to see comments{]}{]}
+
 \newpage
@@
 Maintaining a healthy ecosystem is therefore vital to sustained ecological health and economic prosperity of the region.
-{[}@nefmc20{]}
+(NEFMC 2020)
 
@@
 Fish stocks in this highly productive and economically important region are managed by the National Oceanic and Atmospheric Administration's (NOAA) Northeast Fisheries Science Center (NEFSC) in Woods Hole, Massachusetts.
-Federal scientists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (cite survey paper).
+Federal scientists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (cite survey paper{[}{[}, some possible citations are \url{https://repository.library.noaa.gov/view/noaa/4825/noaa_4825_DS1.pdf} and T.S.
+Azarovitz 1981 A brief historical review of the Woods Hole Laboratory trawl survey time series.
+W.G.
+Doubleday, D.
+Rivard (Eds.), Bottom Trawl Surveys, Can.
+Spec.
+Pub.
+Fish.
+Aqua.
+Sci.
+No.~58 (1981), pp.~62-67{]}{]}).
 The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf.
@@
 Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments.
-The implication of this is that the bottom trawl survey is actually sampling the population during a different life cycle stage than was originally assumed, which can lead to biased stock assessments.
+The implication of this is that the bottom trawl survey is actually sampling the population during a different life cycle stage than was originally assumed {[}{[}I don't think the concern is about life cycle stage, but rather that the fish catchability or availability in the survey will change, thus changing the relationship between the index and the true population{]}{]}, which can lead to biased stock assessments.
 We are therefore interested in analyzing the impact of climate change on the accuracy of future stock assessment models as measured by NOAA's ongoing bottom-trawl survey along the East coast.
@@
 \item
-  say something about our study species? Past abundance history, biological characteristics, others? Maybe better suited in below section
+  say something about our study species?
+Past abundance history, biological characteristics, others?
+Maybe better suited in below section {[}{[}I think the specific species can wait for the methods section, just laying the groundwork here in the intro{]}{]}
 \end{itemize}
@@
 \item
-  Fish are changing spatial distribution and have altered life stages (?) because of climate change
+  Fish are changing spatial distribution and have altered life stages (?) because of climate change {[}{[}see note above about catchability and availability{]}{]}
   NYE paper
@@
 
-We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on George's Bank, where movement of each species combine static species-specific habitat preferences with temperature preferences.
+We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on {[}{[}Georges{]}{]} George's Bank, where movement of each species combine static species-specific habitat preferences with temperature preferences.
 Model dynamics are driven by a time series of temperature gradients that were estimated from data to create simulated data sets for each population where the true biomass is known.
-Using temperature gradients that repeat each year creates data sets with predictable, repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time.
+Using temperature gradients that repeat each year creates data sets with predictable, repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time {[}{[}and mimics the possible future in the region under climate change?{]}{]}.
 We conducting stratified random sampling on our simulation output to mimic the bottom trawl survey and compare the ability of contemporary indexing methods to track population trends.
@@
 
-We use the R package \emph{MixFishSim} (MFS) to model our populations {[}@dolder2020highly{]}.
+We use the R package \emph{MixFishSim} (MFS) to model our populations (Dolder et al.
+2020).
 MFS is a discrete spatiotemporal simulation tool where users can model multiple species under varying environmental conditions.
@@
 We formulate the following inputs for the MFS package to address our research question.
+{[}{[}perhaps for discussion, but we should mention somewhere that MFS only tracks biomass not age structure, the latter is important for stock assessments but not considered in this paper{]}{]}
 
@@
 
-We obtained a shapefile for the 15 strata that comprise George's Bank to use as our modeling environment.
+We obtained a shapefile for the 15 strata that comprise {[}{[}Georges Bank does not have an apostrophe, I made the same mistake when I started here{]}{]} George's Bank to use as our modeling environment.
 We discritized the region into a raster with 88 rows and 144 columns.
@@
 Figure \ref{fig:strata-plot} shows the regions used in our models.
+{[}{[}can mention that the different strata are based on biological differences among the species, possibly add that there are multiple stocks for all three species and the delineations among the stocks differ by species{]}{]}
 
@@
 The time step for our models is one week.
-MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass in each cell in our study area {[}@dolder2020highly{]}.
+MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass in each cell in our study area (Dolder et al.
+2020).
 Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits.
-We chose to represent recruitment in the model using a Beverton-Holt formulation \textbf{cite}.
+{[}{[}not sure we need to highlight the recruitment model, might be able to just use the table because Bev-Holt formulation is well known in fisheries{]}{]} We chose to represent recruitment in the model using a Beverton-Holt formulation \textbf{cite}.
 Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.
@@
 
-Species-specific habitat preferences were derived using the \emph{lrren} tool from the R package \emph{envi} {[}@envi{]} to create a niche model for each species.
+Species-specific habitat preferences were derived using the \emph{lrren} tool from the R package \emph{envi} (Buller 2022) to create a niche model for each species.
 The \emph{lrren} tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors.
-We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow \textbf{cite trawl data?}.
+We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow {[}{[}I don't think we need to cite the trawl data because we said we were using trawl data above{]}{]} \textbf{cite trawl data?}.
 We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences.
 Depth and mean sediment size were used as our covariate predictors.
-Estimated depth for the region was obtained from FVCOM {[}@chen2006unstructured{]}.
-The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method \textbf{cite arcmap or more?} using point data collected by the United States Geologic Survey (USGS) {[}@mcmullen2005usgs{]}.
+Estimated depth for the region was obtained from FVCOM (Chen et al.
+2006).
+The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method {[}{[}citing ArcMap should suffice{]}{]} \textbf{cite arcmap or more?} using point data collected by the United States Geologic Survey (USGS) (McMullen et al.
+n.d.).
 Since the values in \(Hab^2_{J,s}\) are required to be between 0 and 1, we transform the spatial estimates from \emph{lrren} to fall between these bounds.
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{Images/hab_snip}
+{\centering \includegraphics[width=0.95\linewidth]{Images/hab_snip3}
 
@@
 We chose these values by combining information in the literature with temperatures recorded in the bottom trawl survey.
-Weekly estimated temperature data for the region for 2012 was obtained from FVCOM {[}@chen2006unstructured{]}.
+Weekly estimated temperature data for the region for 2012 was obtained from FVCOM (Chen et al.
+2006).
 We chose 2012 because the data displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values.
@@
 \textbf{show images of temperature and/or and/or temp videos???
-and/or average temperature oscillations?}
+and/or average temperature oscillations?}{[}{[}as we discussed, one plot here comparing average temperatures for constant and increasing scenarios in the paper, one or more animations in supplemental materials{]}{]}
 
@@
 Thus, stratified random samples in scenarios with a repeating temperature pattern will have constant survey catchability and availability over time, which may not be true for increasing temperature scenarios due to evolving spatial preferences.
+{[}{[}may want to mention here or in the discussion that we used repeating 2012 to simplify the simulations and reduce the number of factors explored that would have occurred with different annual cycles{]}{]}
 
@@
 We consider 20 year simulations under three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time.
-Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius over the duration of the 20 year simulation.
+Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius {[}{[}need some justification for this large change, wanted to be extreme because if the extreme doesn't make a difference, then more reasonable increases wouldn't make a difference{]}{]} over the duration of the 20 year simulation.
 We therefore simulate a total of 6 scenarios.
@@
 We sample in the same weeks that the Spring and Fall surveys take place and the number of the samples taken in each strata reflect true values.
-Most strata contain enough cells to sample a unique location in each survey over the duration of the simulation.
+{[}{[}need to make sure that we don't lose sight of random samlping with the following sentences{]}{]} Most strata contain enough cells to sample a unique location in each survey over the duration of the simulation.
 For smaller strata we must repeat some sample locations.
 We then use the biomass collected from our samples in contemporary population indexing methods to estimate population trends.
-Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.
+Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.{[}{[}should we mention here or later that we did both with and without measurement error analyses?{]}{]}
 
@@
 
-We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
+{[}{[}could use an intro sentence here to distinguish design-based (stratified mean) and model-based (VAST) estimators{]}{]} We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
 The stratified mean is a typical survey-based approach that scales individual samples to the strata-level by considering the area of each strata, before scaling to the region-level based on the relative size of each strata.
@@
 temperature).
-The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and take on the order of hours to complete.
-
-We follow the advice given in {[}@thorson2019guidance{]} to build VAST models to estimate biomass on George's Bank using stratified mean samples from our model output.
+The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and take on the order of hours to complete.{[}{[}may want to highlight the potential of VAST with covariates to address the climate change issue here as the reason we are doing these simulations{]}{]}
+
+We follow the advice given in (Thorson 2019) to build VAST models to estimate biomass on {[}{[}Georges{]}{]} George's Bank using stratified mean samples from our model output.
 In addition to exploring different link functions and assumed distributions, our VAST model-building process included testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
@@
 To evaluate the effect of populations moving into new habitat, we compare indexing estimates using samples from all strata versus those that only include a subset of the full spatial domain for each species.
+{[}{[}may need to spell the aspect of reduced spatial domain a bit more because it might be confusing to readers who are expected us to just add areas around the current ones instead of reducing the strata{]}{]}
 
@@
 The goal of our project was to analyze how well contemporary population indexing methods can track population trends under a host of conditions, as depicted in Table COMBOS.
-Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year \textbf{cite}.
+Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year {[}{[}can cite the 2022 management track assessments, see \url{https://apps-nefsc.fisheries.noaa.gov/saw/sari.php} for when the document becomes available{]}{]} \textbf{cite}.
 For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod (see Table ???) and increasing population scenarios for Haddock (see Table ???).
@@
 We also see individual cases where the stratified mean produced the lowest absolut error and instances where including covarites in VAST models actually increase the absolute error.
-When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error, though there are some scenarios where the impact is minimal.
+When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error {[}{[}as expected{]}{]}, though there are some scenarios where the impact is minimal.
 
@@
 
+{[}{[}We'll want to expand the results section to focus on the impact of each factor one at at time within each species.
+This will be a bit dull, but it is important to walk through the results in words to ensure that readers get the message.
+They can of course examine the tables in detail and draw their own conclusions, but we should put our interpretation down on paper.{]}{]}
+
 \begin{table}
@@
 
+{[}{[} some things that I think we'll need to address, in addition to those mentioned above are (in no particular order):
+number of simulations done for each scenario - time required to do them and analyses
+pros and cons of using VAST with covariates in a changing environment
+implications for current practice of using stratified mean (not horrible, but might be able to do better with model-based using covariates if enviro changing, note we don't want to say stratified mean should not be used because that will draw a lot of tomato throwing)
+future work - some of the ideas we've discussed already could be mentioned
+limitations to our study{]}{]}
+
 \section{Acknowledgements}
+
+{[}{[}we should thank Jim (obviously), but also those who helped you with the habitat data and others?
+also should note that funding provided by Climate-Groundfish source (I'll dig up the official name of the funding source){]}{]}
 
@@
 
-All data and code used in this work are available at \url{https://github.com/cmlegault/IBMWG}.
+All data and code used in this work are available at \url{https://github.com/cmlegault/IBMWG}. {[}{[}need to update this{]}{]}
 
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-envi}{}}%
+Buller, I.D.
+2022.
+Envi: Environmental interpolation using spatial kernel density estimation.
+The Comprehensive R Archive Network.
+doi:\href{https://doi.org/10.5281/zenodo.5347826}{10.5281/zenodo.5347826}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-chen2006unstructured}{}}%
+Chen, C., Beardsley, R.C., Cowles, G., Qi, J., Lai, Z., Gao, G., and others. 2006. An unstructured grid, finite-volume coastal ocean model: FVCOM user manual. SMAST/UMASSD: 6--8.
+
+\leavevmode\vadjust pre{\hypertarget{ref-dolder2020highly}{}}%
+Dolder, P.J., Minto, C., Guarini, J.-M., and Poos, J.J. 2020. Highly resolved spatiotemporal simulations for exploring mixed fishery dynamics. Ecological Modelling \textbf{424}: 109000. Elsevier.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mcmullen2005usgs}{}}%
+McMullen, K., Paskevich, V., and Poppe, L. (n.d.). USGS east-coast sediment analysis: Procedures, database, and GIS data. US Geological Survey Open-File Report \textbf{2005}: 1001.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nefmc20}{}}%
+NEFMC. 2020. {Northeast skate complex fishery management plan. Amendment 5 discussion document}. {Available online at https://www.nefmc.org/library/amendment-5-3}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-thorson2019guidance}{}}%
+Thorson, J.T.
+2019.
+Guidance for decisions using the vector autoregressive spatio-temporal (VAST) package in stock, ecosystem, habitat and climate assessments.
+Fisheries Research \textbf{210}: 143--161.
+Elsevier.
+
 \end{CSLReferences}
@@
 
+\begin{quote}
+\begin{quote}
+\begin{quote}
+\begin{quote}
+\begin{quote}
+\begin{quote}
+\begin{quote}
+fac774b6b2cab2991e9875992ad8ca7eb3de1326
+\end{quote}
+\end{quote}
+\end{quote}
+\end{quote}
+\end{quote}
+\end{quote}
+\end{quote}
+
 \end{document}

### 2023-01-05 Manuscript/Draft1.Rmd
update manuscript and plotting file

@@
 <!-- ``` -->
->>>>>>> fac774b6b2cab2991e9875992ad8ca7eb3de1326

### 2023-01-05 TestScripts/GB_habitat_create.R

@@ -1869,4 +1869,316 @@ for(s in seq(3)) {
                            nrow=1),
                mylegend, nrow=2,heights=c(10, 1))
 
+  
+  
+  
+  #plot just one habitat
+  
+  ggplot() +
+    geom_raster(data=temp_ras,aes(x=y,y=rev(x),fill=Suitability)) + #plot biomass
+    
+    scale_fill_distiller(palette = "Spectral")+  #set the color pallet and color limits
+    theme_void()+
+    theme(legend.position = "none")
+    # ggtitle("Cod Habitat")+  
+    # theme(plot.title = element_text(hjust = 0.5,size=50),legend.position="none" )
 
+  
+  
+  
+  
+  
+  
+  
+  
+  
+  
+  
+  
+  ######################################################################################
+  
+  
+  #6) PLOTTING POPULATION SCENARIOS USED FOR PUBLICATION
+  
+  #LOAD ALL SCENARIOS USED
+  
+  
+  ######################################################################################
+  
+  #I AM COPYING FROM CALC_SRS_INDEX_SURVEY_BENS, which was adapted from Liz's code to create below
+  library(tibble)
+  library(ggplot2)
+  library(plyr)
+  library(dplyr)
+  library(tidyr)
+  library(readr)
+  library(here)
+  
+  
+  
+  orig.dir <- getwd()
+  
+  n_spp <- 3
+  
+  years_sim <- 22
+  
+  years_cut <- 2
+  
+  spp_names_short <- c("YT","Cod","Had")
+  
+  #choose which simulation iteration to use (chosen in different file)
+  good_iter <- list()
+  good_iter[["ConPop"]][["ConTemp"]] <- c(1) #yellowtail only c(1,13,6) #ConPop_ConTemp
+  good_iter[["ConPop"]][["IncTemp"]] <- c(1) #yellowtail only c(1,1,3) #ConPop_IncTemp
+  good_iter[["IncPop"]][["ConTemp"]] <- c(77,98) #YT and Had c(77,63,98) #IncPop_ConTemp
+  good_iter[["IncPop"]][["IncTemp"]] <- c(77,100) #YT and Had c(77,44,100) #IncPop_IncTemp
+  good_iter[["DecPop"]][["ConTemp"]] <- c(25,18) #YT and Cod c(25,18,6) #DecPop_ConTemp
[…]

### 2023-01-06 Manuscript/Draft1.Rmd
manuscript and plotting file update

--- 2023-01-05 Manuscript/Draft1.Rmd
@@
 Maintaining a healthy ecosystem is therefore vital to sustained ecological health and economic prosperity of the region.
-[@nefmc20]
 
@@
 Fish stocks in this highly productive and economically important region are managed by the National Oceanic and Atmospheric Administration’s (NOAA) Northeast Fisheries Science Center (NEFSC) in Woods Hole, Massachusetts.
-Federal scientists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (cite survey paper).
+Federal scientists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 [@azarovitz1981brief; @politis2014northeast].
 The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf.
@@
 Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments.
-The implication of this is that the bottom trawl survey is actually sampling the population during a different life cycle stage than was originally assumed, which can lead to biased stock assessments.
+More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey will change, altering the relationship between the index and the true population.
 We are therefore interested in analyzing the impact of climate change on the accuracy of future stock assessment models as measured by NOAA’s ongoing bottom-trawl survey along the East coast.
 
-- say something about our study species? Past abundance history, biological characteristics, others? Maybe better suited in below section
-
 **use more info from initial proposal**
-
-- Fish are changing spatial distribution and have altered life stages (?) because of climate change
-NYE paper
-
-- Population indexing methods may be becoming biased as a result
 
@@
 
-We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on George's Bank, where movement of each species combine static species-specific habitat preferences with temperature preferences.
+We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on Georges Bank, where movement of each species combine static species-specific habitat preferences with temperature preferences.
 Model dynamics are driven by a time series of temperature gradients that were estimated from data to create simulated data sets for each population where the true biomass is known.
-Using temperature gradients that repeat each year creates data sets with predictable, repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time.
+Using temperature gradients that repeat each year creates data sets with predictable, repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time, approximating the future distribution in the region under climate change.
 We conducting stratified random sampling on our simulation output to mimic the bottom trawl survey and compare the ability of contemporary indexing methods to track population trends.
@@
 
-We obtained a shapefile for the 15 strata that comprise George's Bank to use as our modeling environment.
-We discritized the region into a raster with 88 rows and 144 columns.
+We obtained a shapefile for the 15 strata that comprise Georges Bank, where strata were partitioned based primarily on depth and secondarily by latitude [@politis2014northeast].
+The region was discritized into a raster with 88 rows and 144 columns to use as our modeling environment.
+A fish stock is considered to be a subpopulation of a species that has similar intrinsic parameters.
+Each of the species being modeled has multiple distinct stocks along the Atlantic coast resulting from local environmental conditions.
+Biological differences between species results in each stock inhabiting a different number of strata on Georges Bank.
 Haddock inhabit all 15 strata in the domain, Cod inhabit 13 strata, and Yellowtail exist in 9 strata.
@@
 Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits.
-We chose to represent recruitment in the model using a Beverton-Holt formulation **cite**.
+[[not sure we need to highlight the recruitment model, might be able to just use the table because Bev-Holt formulation is well known in fisheries]]**BL: MFS has 2 recruitment options and I edited the package to use previous years biomass** We chose to represent recruitment in the model using a Beverton-Holt formulation **cite**.
 Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.
@@
 The *lrren* tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors.
-We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow **cite trawl data?**.
+We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow [[I don't think we need to cite the trawl data because we said we were using trawl data above]] **BL: Does trawl data ever get cited?
+It was not cited above.
+cite trawl data?**.
 We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences.
@@
 Estimated depth for the region was obtained from FVCOM [@chen2006unstructured].
-The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method **cite arcmap or more?** using point data collected by the United States Geologic Survey (USGS) [@mcmullen2005usgs].
+The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method **cite arcmap** using point data collected by the United States Geologic Survey (USGS) [@mcmullen2005usgs].
 Since the values in $Hab^2_{J,s}$ are required to be between 0 and 1, we transform the spatial estimates from *lrren* to fall between these bounds.
@@
 Weekly estimated temperature data for the region for 2012 was obtained from FVCOM [@chen2006unstructured].
-We chose 2012 because the data displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values.
-This data was also transform to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation.
-**show images of temperature and/or and/or temp videos???
-and/or average temperature oscillations?**
+The 2012 data was repeated yearly in the constant temperature scenario because displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values.
+We chose to repeat estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data.
+The 2012 temperature data was also transform to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation.
+We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation.
+Figure \@ref(fig:temp-scenarios) depicts mean trends for the temperature scenarios used in our models.
+**dont forget to include gif in final submission**
+
+```{r temp-scenarios, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Mean trends of temperature data used in our model."}
+knitr::include_graphics(c("Images/TempScenarios.png"))
+```
 
@@
 Thus, stratified random samples in scenarios with a repeating temperature pattern will have constant survey catchability and availability over time, which may not be true for increasing temperature scenarios due to evolving spatial preferences.
+[[**BL: included in above paragraph** may want to mention here or in the discussion that we used repeating 2012 to simplify the simulations and reduce the number of factors explored that would have occurred with different annual cycles]]
 
@@
 We consider 20 year simulations under three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time.
-Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius over the duration of the 20 year simulation.
-We therefore simulate a total of 6 scenarios.
-**show line plot of population for each?**
+Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius [[**BL: included in above paragraph** need some justification for this large change, wanted to be extreme because if the extreme doesn't make a difference, then more reasonable increases wouldn't make a difference]] over the duration of the 20 year simulation.
+We therefore simulate 6 scenarios for each population.
 
@@
 After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year.
-We sample in the same weeks that the Spring and Fall surveys take place and the number of the samples taken in each strata reflect true values.
-Most strata contain enough cells to sample a unique location in each survey over the duration of the simulation.
-For smaller strata we must repeat some sample locations.
+We sample each strata in the same weeks in which the Spring and Fall surveys take place and the number of the samples taken reflect true target values for each strata.
+Most strata contain enough cells that we were able to generate a list of sampling locations without replacement, allowing us to sample a unique random cell in each survey over the duration of the simulation.
+For strata that contain small numbers of cells, we generate a random sampling locations with replacement of cells previously sampled.
 We then use the biomass collected from our samples in contemporary population indexing methods to estimate population trends.
 Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.
+**BL: added both to above paragraph**
+[[should we mention here or later that we did both with and without measurement error analyses?]]
+[[need to make sure that we don't lose sight of random sampling with the following sentences]]
 
@@
 temperature).
-The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and take on the order of hours to complete.
+The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and take on the order of hours to complete.[[may want to highlight the potential of VAST with covariates to address the climate change issue here as the reason we are doing these simulations]]
 
-We follow the advice given in [@thorson2019guidance] to build VAST models to estimate biomass on George's Bank using stratified mean samples from our model output.
+We follow the advice given in [@thorson2019guidance] to build VAST models to estimate biomass on Georges Bank using stratified mean samples from our model output.
 In addition to exploring different link functions and assumed distributions, our VAST model-building process included testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
@@
 
+#[[note: the library calls can all be made in the first block of R code instead, it sometimes makes it easier to find them if they are all in one place, this approach of defining them within separate blocks works too, just noting a possible different approach]]
+
 tabl <- "
-|	Species	|	Population Trend | Temperature Scenario | Strata Included | Noise in Data | Covariates (VAST only) | Season
-|:--	|:-- |:-- |:-- |:-- |:--- |:--
-|Yellowtail | Increasing | Repeating | All strata | No Noise| No Covariates | Spring
-|Cod | Constant | Increasing 5$^{\\circ}$| Subset | Yes Noise| Temp + Habitat | Fall
+|	Species	|	Population Trend | Temperature Scenario | Strata Included | Noise Added | Covariates (VAST) | Season
+|:--	|:-- |:--- |:-- |:-- |:-- |:-
+|Yellowtail | Increasing | Repeating | All strata | No| No Covariates | Spring
+|Cod | Constant | Increasing 5$^{\\circ}$| Subset | Yes| Temp + Habitat | Fall
 |Haddock | Decreasing | | | | |
 "
+
+#knitr::kable(tabl, vline = "|")
+
 cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
@@
 The goal of our project was to analyze how well contemporary population indexing methods can track population trends under a host of conditions, as depicted in Table COMBOS.
-Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year **cite**.
+Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year [[can cite the 2022 management track assessments, see https://apps-nefsc.fisheries.noaa.gov/saw/sari.php for when the document becomes available]] **cite**.
 For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod (see Table ???) and increasing population scenarios for Haddock (see Table ???).
 To provide a comprehensive analysis of population indexing methods we consider all possible scenario combinations for Yellowtail Flounder (see Table ???).
+The specific population trends used in our analyses can be see in Figure \@ref(fig:pop-scenarios)
+
+```{r pop-scenarios, echo=FALSE,out.width="99%",fig.show='hold', fig.align='center', fig.cap="True population trends used in indexing analyses.
+Spring biomass plots are shown with fall values being very similar."}
+knitr::include_graphics(c("Images/Population_scenarios.png"))
+```
 
@@
 \section{Discussion}
+[[ some things that I think we'll need to address, in addition to those mentioned above are (in no particular order):
+number of simulations done for each scenario - time required to do them and analyses
+pros and cons of using VAST with covariates in a changing environment
+implications for current practice of using stratified mean (not horrible, but might be able to do better with model-based using covariates if enviro changing, note we don't want to say stratified mean should not be used because that will draw a lot of tomato throwing)
+future work - some of the ideas we've discussed already could be mentioned
+limitations to our study
+perhaps for discussion, but we should mention somewhere that MFS only tracks biomass not age structure, the latter is important for stock assessments but not considered in this paper
+]]
 
 \section{Acknowledgements}
+[[we should thank Jim (obviously), but also those who helped you with the habitat data (robyn, david Chev.) and others?
+also should note that funding provided by Climate-Groundfish source (I'll dig up the official name of the funding source)]]
 
 ## Data and Code Availability
-All data and code used in this work are available at https://github.com/cmlegault/IBMWG.
+All data and code used in this work are available at https://github.com/cmlegault/IBMWG. [[need to update this]]
 

### 2023-01-06 Manuscript/Draft1.tex
update manuscript and plotting file
manuscript and plotting file update

@@
 Maintaining a healthy ecosystem is therefore vital to sustained ecological health and economic prosperity of the region.
-(NEFMC 2020)
 
@@
 Fish stocks in this highly productive and economically important region are managed by the National Oceanic and Atmospheric Administration's (NOAA) Northeast Fisheries Science Center (NEFSC) in Woods Hole, Massachusetts.
-Federal scientists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (cite survey paper{[}{[}, some possible citations are \url{https://repository.library.noaa.gov/view/noaa/4825/noaa_4825_DS1.pdf} and T.S.
-Azarovitz 1981 A brief historical review of the Woods Hole Laboratory trawl survey time series.
-W.G.
-Doubleday, D.
-Rivard (Eds.), Bottom Trawl Surveys, Can.
-Spec.
-Pub.
-Fish.
-Aqua.
-Sci.
-No.~58 (1981), pp.~62-67{]}{]}).
+Federal scientists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (Azarovitz 1981; Politis et al.
+2014).
 The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf.
@@
 Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments.
-The implication of this is that the bottom trawl survey is actually sampling the population during a different life cycle stage than was originally assumed {[}{[}I don't think the concern is about life cycle stage, but rather that the fish catchability or availability in the survey will change, thus changing the relationship between the index and the true population{]}{]}, which can lead to biased stock assessments.
+More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey will change, altering the relationship between the index and the true population.
 We are therefore interested in analyzing the impact of climate change on the accuracy of future stock assessment models as measured by NOAA's ongoing bottom-trawl survey along the East coast.
+
+\textbf{use more info from initial proposal}
 
@@
 \tightlist
-\item
-  say something about our study species?
-Past abundance history, biological characteristics, others?
-Maybe better suited in below section {[}{[}I think the specific species can wait for the methods section, just laying the groundwork here in the intro{]}{]}
-\end{itemize}
-
-\textbf{use more info from initial proposal}
-
-\begin{itemize}
-\item
-  Fish are changing spatial distribution and have altered life stages (?) because of climate change {[}{[}see note above about catchability and availability{]}{]}
-  NYE paper
-\item
-  Population indexing methods may be becoming biased as a result
 \item
@@
 
-We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on {[}{[}Georges{]}{]} George's Bank, where movement of each species combine static species-specific habitat preferences with temperature preferences.
+We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on Georges Bank, where movement of each species combine static species-specific habitat preferences with temperature preferences.
 Model dynamics are driven by a time series of temperature gradients that were estimated from data to create simulated data sets for each population where the true biomass is known.
-Using temperature gradients that repeat each year creates data sets with predictable, repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time {[}{[}and mimics the possible future in the region under climate change?{]}{]}.
+Using temperature gradients that repeat each year creates data sets with predictable, repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time, approximating the future distribution in the region under climate change.
 We conducting stratified random sampling on our simulation output to mimic the bottom trawl survey and compare the ability of contemporary indexing methods to track population trends.
@@
 
-We obtained a shapefile for the 15 strata that comprise {[}{[}Georges Bank does not have an apostrophe, I made the same mistake when I started here{]}{]} George's Bank to use as our modeling environment.
-We discritized the region into a raster with 88 rows and 144 columns.
+We obtained a shapefile for the 15 strata that comprise Georges Bank, where strata were partitioned based primarily on depth and secondarily by latitude (Politis et al.
+2014).
+The region was discritized into a raster with 88 rows and 144 columns to use as our modeling environment.
+A fish stock is considered to be a subpopulation of a species that has similar intrinsic parameters.
+Each of the species being modeled has multiple distinct stocks along the Atlantic coast resulting from local environmental conditions.
+Biological differences between species results in each stock inhabiting a different number of strata on Georges Bank.
 Haddock inhabit all 15 strata in the domain, Cod inhabit 13 strata, and Yellowtail exist in 9 strata.
 Figure \ref{fig:strata-plot} shows the regions used in our models.
-{[}{[}can mention that the different strata are based on biological differences among the species, possibly add that there are multiple stocks for all three species and the delineations among the stocks differ by species{]}{]}
 
@@
 Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits.
-{[}{[}not sure we need to highlight the recruitment model, might be able to just use the table because Bev-Holt formulation is well known in fisheries{]}{]} We chose to represent recruitment in the model using a Beverton-Holt formulation \textbf{cite}.
+{[}{[}not sure we need to highlight the recruitment model, might be able to just use the table because Bev-Holt formulation is well known in fisheries{]}{]}\textbf{BL: MFS has 2 recruitment options and I edited the package to use previous years biomass} We chose to represent recruitment in the model using a Beverton-Holt formulation \textbf{cite}.
 Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.
@@
 The \emph{lrren} tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors.
-We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow {[}{[}I don't think we need to cite the trawl data because we said we were using trawl data above{]}{]} \textbf{cite trawl data?}.
+We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow {[}{[}I don't think we need to cite the trawl data because we said we were using trawl data above{]}{]} \textbf{BL: Does trawl data ever get cited?
+It was not cited above.
+cite trawl data?}.
 We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences.
@@
 2006).
-The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method {[}{[}citing ArcMap should suffice{]}{]} \textbf{cite arcmap or more?} using point data collected by the United States Geologic Survey (USGS) (McMullen et al.
+The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method \textbf{cite arcmap} using point data collected by the United States Geologic Survey (USGS) (McMullen et al.
 n.d.).
@@
 2006).
-We chose 2012 because the data displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values.
-This data was also transform to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation.
-\textbf{show images of temperature and/or and/or temp videos???
-and/or average temperature oscillations?}{[}{[}as we discussed, one plot here comparing average temperatures for constant and increasing scenarios in the paper, one or more animations in supplemental materials{]}{]}
+The 2012 data was repeated yearly in the constant temperature scenario because displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values.
+We chose to repeat estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data.
+The 2012 temperature data was also transform to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation.
+We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation.
+Figure \ref{fig:temp-scenarios} depicts mean trends for the temperature scenarios used in our models.
+\textbf{dont forget to include gif in final submission}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{Images/TempScenarios}
+
+}
+
+\caption{Mean trends of temperature data used in our model.}\label{fig:temp-scenarios}
+\end{figure}
 
@@
 Thus, stratified random samples in scenarios with a repeating temperature pattern will have constant survey catchability and availability over time, which may not be true for increasing temperature scenarios due to evolving spatial preferences.
-{[}{[}may want to mention here or in the discussion that we used repeating 2012 to simplify the simulations and reduce the number of factors explored that would have occurred with different annual cycles{]}{]}
+{[}{[}\textbf{BL: included in above paragraph} may want to mention here or in the discussion that we used repeating 2012 to simplify the simulations and reduce the number of factors explored that would have occurred with different annual cycles{]}{]}
 
@@
 We consider 20 year simulations under three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time.
-Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius {[}{[}need some justification for this large change, wanted to be extreme because if the extreme doesn't make a difference, then more reasonable increases wouldn't make a difference{]}{]} over the duration of the 20 year simulation.
-We therefore simulate a total of 6 scenarios.
-\textbf{show line plot of population for each?}
+Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius {[}{[}\textbf{BL: included in above paragraph} need some justification for this large change, wanted to be extreme because if the extreme doesn't make a difference, then more reasonable increases wouldn't make a difference{]}{]} over the duration of the 20 year simulation.
+We therefore simulate 6 scenarios for each population.
 
@@
 After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year.
-We sample in the same weeks that the Spring and Fall surveys take place and the number of the samples taken in each strata reflect true values.
-{[}{[}need to make sure that we don't lose sight of random samlping with the following sentences{]}{]} Most strata contain enough cells to sample a unique location in each survey over the duration of the simulation.
-For smaller strata we must repeat some sample locations.
+We sample each strata in the same weeks in which the Spring and Fall surveys take place and the number of the samples taken reflect true target values for each strata.
+Most strata contain enough cells that we were able to generate a list of sampling locations without replacement, allowing us to sample a unique random cell in each survey over the duration of the simulation.
+For strata that contain small numbers of cells, we generate a random sampling locations with replacement of cells previously sampled.
 We then use the biomass collected from our samples in contemporary population indexing methods to estimate population trends.
-Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.{[}{[}should we mention here or later that we did both with and without measurement error analyses?{]}{]}
+Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.
+\textbf{BL: added both to above paragraph}
+{[}{[}should we mention here or later that we did both with and without measurement error analyses?{]}{]}
+{[}{[}need to make sure that we don't lose sight of random sampling with the following sentences{]}{]}
 
@@
 
-We follow the advice given in (Thorson 2019) to build VAST models to estimate biomass on {[}{[}Georges{]}{]} George's Bank using stratified mean samples from our model output.
+We follow the advice given in (Thorson 2019) to build VAST models to estimate biomass on Georges Bank using stratified mean samples from our model output.
 In addition to exploring different link functions and assumed distributions, our VAST model-building process included testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.14}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.19}}
   >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.14}}
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.14}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.18}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.14}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.10}}@{}}
 \toprule
@@
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Noise in Data
+Noise Added
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Covariates (VAST only)
+Covariates (VAST)
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
@@
 \endhead
-Yellowtail & Increasing & Repeating & All strata & No Noise & No Covariates & Spring \\
-Cod & Constant & Increasing 5\(^{\circ}\) & Subset & Yes Noise & Temp + Habitat & Fall \\
+Yellowtail & Increasing & Repeating & All strata & No & No Covariates & Spring \\
+Cod & Constant & Increasing 5\(^{\circ}\) & Subset & Yes & Temp + Habitat & Fall \\
 Haddock & Decreasing & & & & & \\
@@
 To provide a comprehensive analysis of population indexing methods we consider all possible scenario combinations for Yellowtail Flounder (see Table ???).
+The specific population trends used in our analyses can be see in Figure \ref{fig:pop-scenarios}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.99\linewidth]{Images/Population_scenarios}
+
+}
+
+\caption{True population trends used in indexing analyses. Spring biomass plots are shown with fall values being very similar.}\label{fig:pop-scenarios}
+\end{figure}
 
@@
 future work - some of the ideas we've discussed already could be mentioned
-limitations to our study{]}{]}
+limitations to our study
+perhaps for discussion, but we should mention somewhere that MFS only tracks biomass not age structure, the latter is important for stock assessments but not considered in this paper{]}{]}
 
@@
 
-{[}{[}we should thank Jim (obviously), but also those who helped you with the habitat data and others?
+{[}{[}we should thank Jim (obviously), but also those who helped you with the habitat data (robyn, david Chev.) and others?
 also should note that funding provided by Climate-Groundfish source (I'll dig up the official name of the funding source){]}{]}
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-azarovitz1981brief}{}}%
+Azarovitz, T. 1981. A brief historical review of the woods hole laboratory trawl survey time series. Bottom trawl surveys.
+
 \leavevmode\vadjust pre{\hypertarget{ref-envi}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-nefmc20}{}}%
-NEFMC. 2020. {Northeast skate complex fishery management plan. Amendment 5 discussion document}. {Available online at https://www.nefmc.org/library/amendment-5-3}.
+\leavevmode\vadjust pre{\hypertarget{ref-politis2014northeast}{}}%
+Politis, P.J., Galbraith, J.K., Kostovick, P., and Brown, R.W. 2014. Northeast fisheries science center bottom trawl survey protocols for the NOAA ship henry b. bigelow.
 
@@
 
-\begin{quote}
-\begin{quote}
-\begin{quote}
-\begin{quote}
-\begin{quote}
-\begin{quote}
-\begin{quote}
-fac774b6b2cab2991e9875992ad8ca7eb3de1326
-\end{quote}
-\end{quote}
-\end{quote}
-\end{quote}
-\end{quote}
-\end{quote}
-\end{quote}
-
 \end{document}

### 2023-01-06 TestScripts/GB_habitat_create.R

@@ -1905,7 +1905,6 @@ for(s in seq(3)) {
   
   ######################################################################################
   
-  #I AM COPYING FROM CALC_SRS_INDEX_SURVEY_BENS, which was adapted from Liz's code to create below
   library(tibble)
   library(ggplot2)
   library(plyr)
@@ -2086,25 +2085,28 @@ for(s in seq(3)) {
 
     for(scenario in c("ConTemp","IncTemp")){
       
-  
-    
+  ifelse(scenario == "ConTemp",ss <- "Constant Temperature",ss <- "Increasing Temperature")
+
       PopPlot[[s]][[scenario]] <- ggplot() +
         
         #this way plots data by season
-        geom_point(data =  subset(as.data.frame(pop_by_season[[s]][[paste("ConPop_",scenario,sep="")]]),season==1), aes(x=as.numeric(year),y=as.numeric(biomass), group = season, color = "Constant Pop."),size=3) +
-        geom_line(data =  subset(as.data.frame(pop_by_season[[s]][[paste("ConPop_",scenario,sep="")]]),season==1), aes(x=as.numeric(year),y=as.numeric(biomass), group = season, color = "Constant Pop."),size=1) +
+        geom_point(data =  subset(as.data.frame(pop_by_season[[s]][[paste("ConPop_",scenario,sep="")]]),season==1), aes(x=as.numeric(year),y=as.numeric(biomass), group = season, color = "Constant Population"),size=3) +
+        geom_line(data =  subset(as.data.frame(pop_by_season[[s]][[paste("ConPop_",scenario,sep="")]]),season==1), aes(x=as.numeric(year),y=as.numeric(biomass), group = season, color = "Constant Population"),size=1) +
         
-        geom_point(data =  subset(as.data.frame(pop_by_season[[s]][[paste("IncPop_",scenario,sep="")]]),season==1), aes(x=as.numeric(year),y=as.numeric(biomass), group = season, color = "Increasing Pop."),size=3) +
-        geom_line(data =  subset(as.data.frame(pop_by_season[[s]][[paste("IncPop_",scenario,sep="")]]),season==1), aes(x=as.numeric(year),y=as.numeric(biomass), group = season, color = "Increasing Pop."),size=1) +
[…]

### 2023-01-07 Manuscript/Draft1.Rmd
paper updates

@@
 ```{r setup, include=FALSE}
+
+library("knitr")
+library("kableExtra")
+library("dplyr")
+library("readr")
+library("tidyr")
+library("forcats")
+
 knitr::opts_chunk$set(echo = FALSE)
@@
 --Stratified mean vs VAST with and without covariates
+[[could use an intro sentence here to distinguish design-based (stratified mean) and model-based (VAST) estimators]] 
+[[may want to highlight the potential of VAST with covariates to address the climate change issue here as the reason we are doing these simulations]]
+**BL: Chris, please check paragraph below for intro sentence and help fill in X, Y and Z**
 
-[[could use an intro sentence here to distinguish design-based (stratified mean) and model-based (VAST) estimators]] We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
-The stratified mean is a typical survey-based approach that scales individual samples to the  strata-level by considering the area of each strata, before scaling to the region-level based on the relative size of each strata.
+Stock assessment scientists use samples from the bottom trawl survey to estimate the abundance for each stock stock.
+There are a number of approaches to obtaining abundance estimates ranging from traditional design-based estimates such as the stratified mean to model-based estimates that range in complexity.
+The stratified mean calculates the geometric mean catch per tow and does not account for spatial variation in samples or allow the inclusion of environmental covariates.
+There are a number of commonly used model-based estimates such as X, Y and Z.
+The Vector Autoregressive Spatio-Temporal (VAST) model is a spatial delta-generalized linear mixed model [@thorson2019guidance].
 VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurrence (presence/absence).
@@
 temperature).
-The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and take on the order of hours to complete.[[may want to highlight the potential of VAST with covariates to address the climate change issue here as the reason we are doing these simulations]]
+We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses.
+The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and take on the order of hours to complete.
+We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
 
@@
 Our goal is to determine indexing approaches and settings that are robust to future environmental conditions and resulting spatial biomass patterns.
-An underlying assumption in all indexing methods that individual random samples combine to accurately represent true abundance by a) sampling all strata in which the population exists and b) low enough noise level in the samples to allow for a discernible pattern.
+An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) sampling all strata in which the population exists and b) low enough noise level in the samples to allow for a discernible pattern.
 This assumption can be questioned given enough noise in the sampling process **cite?** and/or shifting spatial preferences driven by climate change causing a population to move into a previously uninhabited strata.
-To simulate the impact of noise, indexing estimates after adding noise to our samples versus those using the true sampling values.
-To evaluate the effect of populations moving into new habitat, we compare indexing estimates using samples from all strata versus those that only include a subset of the full spatial domain for each species.
-[[may need to spell the aspect of reduced spatial domain a bit more because it might be confusing to readers who are expected us to just add areas around the current ones instead of reducing the strata]]
+To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values.
+We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges bank to those that only include a subset of the full spatial domain for each species.[[may need to spell the aspect of reduced spatial domain a bit more because it might be confusing to readers who are expected us to just add areas around the current ones instead of reducing the strata]]
 
@@
 ```{r scenarios, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-library("knitr")
-library("kableExtra")
-library("dplyr")
-library("readr")
-library("tidyr")
-library("forcats")
-
-#[[note: the library calls can all be made in the first block of R code instead, it sometimes makes it easier to find them if they are all in one place, this approach of defining them within separate blocks works too, just noting a possible different approach]]
 

### 2023-01-07 Manuscript/Draft1.tex
paper updates

@@
 --Stratified mean vs VAST with and without covariates
-
-{[}{[}could use an intro sentence here to distinguish design-based (stratified mean) and model-based (VAST) estimators{]}{]} We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
-The stratified mean is a typical survey-based approach that scales individual samples to the strata-level by considering the area of each strata, before scaling to the region-level based on the relative size of each strata.
+{[}{[}could use an intro sentence here to distinguish design-based (stratified mean) and model-based (VAST) estimators{]}{]}
+{[}{[}may want to highlight the potential of VAST with covariates to address the climate change issue here as the reason we are doing these simulations{]}{]}
+\textbf{BL: Chris, please check paragraph below for intro sentence and help fill in X, Y and Z}
+
+Stock assessment scientists use samples from the bottom trawl survey to estimate the abundance for each stock stock.
+There are a number of approaches to obtaining abundance estimates ranging from traditional design-based estimates such as the stratified mean to model-based estimates that range in complexity.
+The stratified mean calculates the geometric mean catch per tow and does not account for spatial variation in samples or allow the inclusion of environmental covariates.
+There are a number of commonly used model-based estimates such as X, Y and Z.
+The Vector Autoregressive Spatio-Temporal (VAST) model is a spatial delta-generalized linear mixed model (Thorson 2019).
 VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurrence (presence/absence).
@@
 temperature).
-The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and take on the order of hours to complete.{[}{[}may want to highlight the potential of VAST with covariates to address the climate change issue here as the reason we are doing these simulations{]}{]}
+We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses.
+The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and take on the order of hours to complete.
+We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
 
@@
 Our goal is to determine indexing approaches and settings that are robust to future environmental conditions and resulting spatial biomass patterns.
-An underlying assumption in all indexing methods that individual random samples combine to accurately represent true abundance by a) sampling all strata in which the population exists and b) low enough noise level in the samples to allow for a discernible pattern.
+An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) sampling all strata in which the population exists and b) low enough noise level in the samples to allow for a discernible pattern.
 This assumption can be questioned given enough noise in the sampling process \textbf{cite?} and/or shifting spatial preferences driven by climate change causing a population to move into a previously uninhabited strata.
-To simulate the impact of noise, indexing estimates after adding noise to our samples versus those using the true sampling values.
-To evaluate the effect of populations moving into new habitat, we compare indexing estimates using samples from all strata versus those that only include a subset of the full spatial domain for each species.
-{[}{[}may need to spell the aspect of reduced spatial domain a bit more because it might be confusing to readers who are expected us to just add areas around the current ones instead of reducing the strata{]}{]}
+To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values.
+We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges bank to those that only include a subset of the full spatial domain for each species.{[}{[}may need to spell the aspect of reduced spatial domain a bit more because it might be confusing to readers who are expected us to just add areas around the current ones instead of reducing the strata{]}{]}
 

### 2023-01-09 Manuscript/Draft1.Rmd
paper updates

@@
 #read in results tables
-YT_results<-read.csv("Yellowtail_results.csv")
-Cod_results<-read.csv("Cod_results.csv")
-Had_results<-read.csv("Haddock_results.csv")
+YT_results<-read.csv("CSVs/Yellowtail_results.csv")
+Cod_results<-read.csv("CSVs/Cod_results.csv")
+Had_results<-read.csv("CSVs/Haddock_results.csv")
 ```
@@
 Our goal is to determine indexing approaches and settings that are robust to future environmental conditions and resulting spatial biomass patterns.
-An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) sampling all strata in which the population exists and b) low enough noise level in the samples to allow for a discernible pattern.
-This assumption can be questioned given enough noise in the sampling process **cite?** and/or shifting spatial preferences driven by climate change causing a population to move into a previously uninhabited strata.
+An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists.
+These assumptions can be questioned given enough noise in the sampling process **cite?** and/or climate change causing a population to move into previously uninhabited strata.
 To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values.
-We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges bank to those that only include a subset of the full spatial domain for each species.[[may need to spell the aspect of reduced spatial domain a bit more because it might be confusing to readers who are expected us to just add areas around the current ones instead of reducing the strata]]
+**BL: Help with correct notation for adding noise**.
+We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges bank to those that only include a subset of the full spatial domain for each species.
+We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from.
+Tabl XXX lists all strata inhabited by each species, those that are removed from certain calculations, and the explanation of why these strata were removed.
+
+**BL: I tried to provid emore detail above and also added another table**
+[[may need to spell the aspect of reduced spatial domain a bit more because it might be confusing to readers who are expected us to just add areas around the current ones instead of reducing the strata]]
 
 When combining population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, covariates) there are a large number of scenario combinations to consider.
-The columns in Table COMBOS show the choices that define each scenario.
+The columns in Table \@ref(tab:scenarios) show the choices that define each scenario.
 
-<div align="center">Table COMBOS Each index estimate chooses one condition from each of the following columns.
-There are $3*3*2*2*2*2*2=288$ VAST model combinations and $3*3*2*2*2=144$ stratified mean estimates.</div>
+```{r scenarios,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
 
-```{r scenarios, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
+options(knitr.kable.NA = '')
 
-tabl <- "
-|	Species	|	Population Trend | Temperature Scenario | Strata Included | Noise Added | Covariates (VAST) | Season
-|:--	|:-- |:--- |:-- |:-- |:-- |:-
-|Yellowtail | Increasing | Repeating | All strata | No| No Covariates | Spring
-|Cod | Constant | Increasing 5$^{\\circ}$| Subset | Yes| Temp + Habitat | Fall
-|Haddock | Decreasing | | | | |
-"
+combos_table = read_csv("CSVs/scenario_combos.csv")
 
-#knitr::kable(tabl, vline = "|")
+summary_tab <- combos_table %>% 
+   mutate_if(grepl('5deg',.), ~replace(., grepl('5deg', .), paste("Increasing 5","\U00B0","C",sep="")))
 
-cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
-
+# Create table to present the data]
+ kable(summary_tab, "simple",
+                     caption = "Each index estimate chooses one condition from each of the following
+225 columns. There are 3∗3∗2∗2∗2∗2∗2 = 288 VAST model combinations and 3∗3∗2∗2∗2 = 144
+226 stratified mean estimates.",
+          format = "latex", booktabs = TRUE) %>%
+          kable_styling(latex_options = "scale_down")
 ```
@@
 ## Data and Code Availability
-All data and code used in this work are available at https://github.com/cmlegault/IBMWG. [[need to update this]]
+All data and code used in this work are available at https://github.com/Blevy2/READ-PDB-blevy2-MFS2.
 
@@
 
-param_table1 = read_csv("Parameters1.csv")
+param_table1 = read_csv("CSVs/Parameters1.csv")
 
@@
 
-param_table = read_csv("Parameters.csv")
+param_table = read_csv("CSVs/Parameters.csv")
 
@@
 
-result_table = read_csv("Yellowtail_results_final.csv")
+result_table = read_csv("CSVs/Yellowtail_results_final.csv")
 
@@
 
-param_table1 = read_csv("Cod_results.csv")
+param_table1 = read_csv("CSVs/Cod_results.csv")
 

### 2023-01-09 Manuscript/Draft1.tex
paper updates

@@
 Our goal is to determine indexing approaches and settings that are robust to future environmental conditions and resulting spatial biomass patterns.
-An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) sampling all strata in which the population exists and b) low enough noise level in the samples to allow for a discernible pattern.
-This assumption can be questioned given enough noise in the sampling process \textbf{cite?} and/or shifting spatial preferences driven by climate change causing a population to move into a previously uninhabited strata.
+An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists.
+These assumptions can be questioned given enough noise in the sampling process \textbf{cite?} and/or climate change causing a population to move into previously uninhabited strata.
 To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values.
-We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges bank to those that only include a subset of the full spatial domain for each species.{[}{[}may need to spell the aspect of reduced spatial domain a bit more because it might be confusing to readers who are expected us to just add areas around the current ones instead of reducing the strata{]}{]}
+\textbf{BL: Help with correct notation for adding noise}.
+We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges bank to those that only include a subset of the full spatial domain for each species.
+We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from.
+Tabl XXX lists all strata inhabited by each species, those that are removed from certain calculations, and the explanation of why these strata were removed.
+
+\textbf{BL: I tried to provid emore detail above and also added another table}
+{[}{[}may need to spell the aspect of reduced spatial domain a bit more because it might be confusing to readers who are expected us to just add areas around the current ones instead of reducing the strata{]}{]}
 
 When combining population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, covariates) there are a large number of scenario combinations to consider.
-The columns in Table COMBOS show the choices that define each scenario.
-
-Table COMBOS Each index estimate chooses one condition from each of the following columns. There are \(3*3*2*2*2*2*2=288\) VAST model combinations and \(3*3*2*2*2=144\) stratified mean estimates.
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.14}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.14}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.19}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.14}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.14}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.14}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.10}}@{}}
+The columns in Table \ref{tab:scenarios} show the choices that define each scenario.
+
+\begin{table}
+
+\caption{\label{tab:scenarios}Each index estimate chooses one condition from each of the following
+225 columns. There are 3*3*2*2*2*2*2 = 288 VAST model combinations and 3*3*2*2*2 = 144
+226 stratified mean estimates.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lllllll}
 \toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-Species
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Population Trend
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Temperature Scenario
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Strata Included
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Noise Added
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Covariates (VAST)
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Season
-\end{minipage} \\
+Species & Population Trend & Temperature Scenario & Strata Included & Noise Added & Covariates (VAST) & Season\\
 \midrule
-\endhead
-Yellowtail & Increasing & Repeating & All strata & No & No Covariates & Spring \\
-Cod & Constant & Increasing 5\(^{\circ}\) & Subset & Yes & Temp + Habitat & Fall \\
+Yellowtail & Increasing & Repeating & All strata & No & Spring & \\
+Cod & Constant & Increasing 5°C & Subset & Yes & Fall & \\
 Haddock & Decreasing &  &  &  &  & \\
 \bottomrule
-\end{longtable}
+\end{tabular}}
+\end{table}
 
@@
 \hline
-const & all & no & spring & 0.49 & 0.18 & 0.18 & NA & \\
-\hline
-const & all & yes & spring & 0.73 & 0.43 & 0.21 & NA & Haddock\\
-\hline
-const & all & no & fall & 0.28 & 0.05 & 0.26 & NA & Increasing Population\\
-\hline
-const & all & yes & fall & 0.41 & 0.06 & 0.27 & NA & \\
-\hline
-const & reduced & no & spring & 0.34 & 0.35 & 0.45 & NA & \\
-\hline
-const & reduced & yes & spring & 0.30 & 0.33 & 0.46 & NA & \\
-\hline
-const & reduced & no & fall & 0.36 & 0.48 & 0.54 & NA & \\
-\hline
-const & reduced & yes & fall & 0.33 & 0.46 & 0.52 & NA & \\
-\hline
-increasing & all & no & spring & 0.25 & 0.05 & 0.26 & NA & \\
-\hline
-increasing & all & yes & spring & 0.30 & 0.06 & 0.31 & NA & \\
-\hline
-increasing & all & no & fall & 0.89 & 0.23 & 0.40 & NA & \\
-\hline
-increasing & all & yes & fall & 1.04 & 0.35 & 0.42 & NA & \\
-\hline
-increasing & reduced & no & spring & 0.32 & 0.40 & 0.44 & NA & \\
-\hline
-increasing & reduced & yes & spring & 0.38 & 0.37 & 0.37 & NA & \\
-\hline
-increasing & reduced & no & fall & 0.44 & 0.64 & 0.72 & NA & \\
-\hline
-increasing & reduced & yes & fall & 0.42 & 0.62 & 0.70 & NA & \\
+const & all & no & spring & 0.49 & 0.18 & 0.18 &  & \\
+\hline
+const & all & yes & spring & 0.73 & 0.43 & 0.21 &  & Haddock\\
+\hline
+const & all & no & fall & 0.28 & 0.05 & 0.26 &  & Increasing Population\\
+\hline
+const & all & yes & fall & 0.41 & 0.06 & 0.27 &  & \\
+\hline
+const & reduced & no & spring & 0.34 & 0.35 & 0.45 &  & \\
+\hline
+const & reduced & yes & spring & 0.30 & 0.33 & 0.46 &  & \\
+\hline
+const & reduced & no & fall & 0.36 & 0.48 & 0.54 &  & \\
+\hline
+const & reduced & yes & fall & 0.33 & 0.46 & 0.52 &  & \\
+\hline
+increasing & all & no & spring & 0.25 & 0.05 & 0.26 &  & \\
+\hline
+increasing & all & yes & spring & 0.30 & 0.06 & 0.31 &  & \\
+\hline
+increasing & all & no & fall & 0.89 & 0.23 & 0.40 &  & \\
+\hline
+increasing & all & yes & fall & 1.04 & 0.35 & 0.42 &  & \\
+\hline
+increasing & reduced & no & spring & 0.32 & 0.40 & 0.44 &  & \\
+\hline
+increasing & reduced & yes & spring & 0.38 & 0.37 & 0.37 &  & \\
+\hline
+increasing & reduced & no & fall & 0.44 & 0.64 & 0.72 &  & \\
+\hline
+increasing & reduced & yes & fall & 0.42 & 0.62 & 0.70 &  & \\
 \hline
@@
 
-All data and code used in this work are available at \url{https://github.com/cmlegault/IBMWG}. {[}{[}need to update this{]}{]}
+All data and code used in this work are available at \url{https://github.com/Blevy2/READ-PDB-blevy2-MFS2}.
 
@@
  \midrule
- <U+03C1> & Ford's growth coefficient & 1/wk & 4.48 & 4.43 & 4.49 & NA\\
- M & Natural mortality & 1/wk & 0.2064 & 0.2728 & 0.334 & NA\\
- W1 & Weight of fully recruited fish & kg & 0.39 & 2.95 & 1.12 & NA\\
- W2 & Weight of pre-recruit fish & kg & 0.13 & 0.39 & 0.19 & NA\\
- sigma & Variance in recruited fish & kg*kg & 0.55 & 0.55 & 0.55 & NA\\
+ <U+03C1> & Ford's growth coefficient & 1/wk & 4.48 & 4.43 & 4.49 & \\
+ M & Natural mortality & 1/wk & 0.2064 & 0.2728 & 0.334 & \\
+ W1 & Weight of fully recruited fish & kg & 0.39 & 2.95 & 1.12 & \\
+ W2 & Weight of pre-recruit fish & kg & 0.13 & 0.39 & 0.19 & \\
+ sigma & Variance in recruited fish & kg*kg & 0.55 & 0.55 & 0.55 & \\
  \addlinespace
- lambda & Decay rate for movement & - & 0.7 & 0.7 & 0.7 & NA\\
- spwn & Spawning weeks for species s & wks & 9-12 & 8-13 & 11-14 & NA\\
- rec & Recruitment weeks for species s & wks & 9-12 & 8-13 & 11-14 & NA\\
+ lambda & Decay rate for movement & - & 0.7 & 0.7 & 0.7 & \\
+ spwn & Spawning weeks for species s & wks & 9-12 & 8-13 & 11-14 & \\
+ rec & Recruitment weeks for species s & wks & 9-12 & 8-13 & 11-14 & \\
  \bottomrule

### 2023-01-11 Manuscript/Draft1.Rmd
Updates

@@
                      caption = "Each index estimate chooses one condition from each of the following
-225 columns. There are 3∗3∗2∗2∗2∗2∗2 = 288 VAST model combinations and 3∗3∗2∗2∗2 = 144
-226 stratified mean estimates.",
+7 columns. There are 3∗3∗2∗2∗2∗2∗2 = 288 VAST model combinations and 3∗3∗2∗2∗2∗2 = 144 stratified mean estimates.",
           format = "latex", booktabs = TRUE) %>%

### 2023-01-11 Manuscript/Draft1.tex
Updates

@@
 \caption{\label{tab:scenarios}Each index estimate chooses one condition from each of the following
-225 columns. There are 3*3*2*2*2*2*2 = 288 VAST model combinations and 3*3*2*2*2 = 144
-226 stratified mean estimates.}
+7 columns. There are 3*3*2*2*2*2*2 = 288 VAST model combinations and 3*3*2*2*2*2 = 144 stratified mean estimates.}
 \centering
@@
 \toprule
-Species & Population Trend & Temperature Scenario & Strata Included & Noise Added & Covariates (VAST) & Season\\
+Species & Population Trend & Temperature Scenario & Strata Included & Noise Added & Season & Covariates (VAST)\\
 \midrule
-Yellowtail & Increasing & Repeating & All strata & No & Spring & \\
-Cod & Constant & Increasing 5°C & Subset & Yes & Fall & \\
+Yellowtail & Increasing & Repeating & All strata & No & Spring & No\\
+Cod & Constant & Increasing 5°C & Subset & Yes & Fall & Yes\\
 Haddock & Decreasing &  &  &  &  & \\

### 2023-01-11 VAST/Test_3Species_GeorgesBank_Cod_Only.R

@@ -438,7 +438,7 @@ for(j in 1:6){
   
   
   FC1 = c("Omega1" = 0, "Epsilon1" =0, "Omega2" = 1, "Epsilon2" = 1) 
-  RhoConfig = c("Beta1" = 3, "Beta2" = 3, "Epsilon1" = 0, "Epsilon2" = 4)
+  RhoConfig = c("Beta1" = 3, "Beta2" = 0, "Epsilon1" = 0, "Epsilon2" = 4)
   
   #FieldConfig = c("Omega1"=0, "Epsilon1"=0, "Omega2"="IID", "Epsilon2"=0
   

### 2023-01-12 Manuscript/Draft1.Rmd
fairly minor paper updates

@@
 knitr::include_graphics(c("Images/Strata.png"))
+knitr::include_graphics(c("Images/Excluded_strata.png"))
 ```

### 2023-01-12 TestScripts/GB_habitat_create.R

@@ -1795,7 +1795,7 @@ max(temp_max[21:40,1])
 ######################################################################################
 
 
-#5) PLOTTING HABITAT FOR PUBLICATION
+#5a) PLOTTING HABITAT FOR PUBLICATION
 
 
 spp_names <- c("Yellowtail Flounder","Cod","Haddock")
@@ -1883,6 +1883,105 @@ for(s in seq(3)) {
     # ggtitle("Cod Habitat")+  
     # theme(plot.title = element_text(hjust = 0.5,size=50),legend.position="none" )
 
+  
+  
+  
+  
+  
+  ######################################################################################
+  
+  
+  #5b) PLOTTING SPECIES-SPECIFIC STRATA FOR PUBLICATION
+  
+    library(rgdal)
+    library(gridExtra)
+  
+  #load stratas for clipping etc
+  strata.dir <- "C:\\Users\\benjamin.levy\\Desktop\\NOAA\\GIS_Stuff\\" # strata shape files in this directory
+
+  # get the shapefiles
+  strata.areas <- readOGR(paste(strata.dir,"Survey_strata", sep="")) #readShapePoly is deprecated; use rgdal::readOGR or sf::st_read 
+  
+  #define georges bank for YELLOWTAIL
+  GB_strata_num <- c("01130","01140","01150","01160","01170","01180","01190","01200","01210")
+  #pull out indices corresponding to GB strata
+  GB_strata_idx <- match(GB_strata_num,strata.areas@data[["STRATUMA"]])
+  #plot them
+  #plot(strata.areas[GB_strata_idx,])
+  #define GB strata as own object
+  GB_strata_YT <- strata.areas[GB_strata_idx,]
+  
+  #define georges bank for COD  
+ GB_strata_num <- c("01130","01140","01150","01160","01170","01180","01190","01200","01210","01220","01230","01240","01250")
+  #pull out indices corresponding to GB strata
+  GB_strata_idx <- match(GB_strata_num,strata.areas@data[["STRATUMA"]])
+  #plot them
+  #plot(strata.areas[GB_strata_idx,])
+  #define GB strata as own object
+  GB_strata_Cod <- strata.areas[GB_strata_idx,]
+  
+  #define georges bank for HADDOCK  
[…]

### 2023-01-12 Manuscript/Draft1.Rmd
Paper and VAST code updates

@@
 [[may want to highlight the potential of VAST with covariates to address the climate change issue here as the reason we are doing these simulations]]
-**BL: Chris, please check paragraph below for intro sentence and help fill in X, Y and Z**
 
@@
 The stratified mean calculates the geometric mean catch per tow and does not account for spatial variation in samples or allow the inclusion of environmental covariates.
-There are a number of commonly used model-based estimates such as X, Y and Z.
+Model-based abundance estimates are often derived from model framewroks such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM) that can incorporate both random and mixed effects.
 The Vector Autoregressive Spatio-Temporal (VAST) model is a spatial delta-generalized linear mixed model [@thorson2019guidance].

### 2023-01-12 Manuscript/Draft1.tex
fairly minor paper updates
Paper and VAST code updates

@@
 
-{\centering \includegraphics[width=0.95\linewidth]{Images/Strata}
+{\centering \includegraphics[width=0.95\linewidth]{Images/Strata} \includegraphics[width=0.95\linewidth]{Images/Excluded_strata}
 
@@
 {[}{[}may want to highlight the potential of VAST with covariates to address the climate change issue here as the reason we are doing these simulations{]}{]}
-\textbf{BL: Chris, please check paragraph below for intro sentence and help fill in X, Y and Z}
 
@@
 The stratified mean calculates the geometric mean catch per tow and does not account for spatial variation in samples or allow the inclusion of environmental covariates.
-There are a number of commonly used model-based estimates such as X, Y and Z.
+Model-based abundance estimates are often derived from model framewroks such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM) that can incorporate both random and mixed effects.
 The Vector Autoregressive Spatio-Temporal (VAST) model is a spatial delta-generalized linear mixed model (Thorson 2019).

### 2023-01-12 VAST/Test_3Species_GeorgesBank_Cod_Only.R

@@ -691,7 +691,7 @@ for(j in 1:6){
   
   
   FC2 = c("Omega1" = 0, "Epsilon1" =0, "Omega2" = 1, "Epsilon2" = 1) 
-  RhoConfig = c("Beta1" = 3, "Beta2" = 3, "Epsilon1" = 0, "Epsilon2" =4)
+  RhoConfig = c("Beta1" = 3, "Beta2" = 0, "Epsilon1" = 0, "Epsilon2" = 4)
   
   settings <- make_settings(n_x = 500,
                             Region=example$Region,

### 2023-01-12 VAST/Test_3Species_GeorgesBank_Had_Only.R

@@ -451,7 +451,7 @@ for(j in 1:6){
   
   
   FC1 = c("Omega1" = 0, "Epsilon1" =0, "Omega2" = 1, "Epsilon2" = 1) 
-  RhoConfig = c("Beta1" = 3, "Beta2" = 3, "Epsilon1" = 0, "Epsilon2" = 0) #tried 2 #was 4 for most runs
+  RhoConfig = c("Beta1" = 3, "Beta2" = 0, "Epsilon1" = 0, "Epsilon2" = 4) #tried 2 #was 4 for most runs
   
   #FieldConfig = c("Omega1"=0, "Epsilon1"=0, "Omega2"="IID", "Epsilon2"=0
   
@@ -709,7 +709,7 @@ for(j in 1:6){
   #FC2 = c("Omega1" = 1, "Epsilon1" = 1, "Omega2" = 1, "Epsilon2" = 1) 
   
   FC2 = c("Omega1" = 0, "Epsilon1" =0, "Omega2" = 1, "Epsilon2" = 1) 
-  RhoConfig = c("Beta1" = 3, "Beta2" = 3, "Epsilon1" = 0, "Epsilon2" =2) #used to be 4
+  RhoConfig = c("Beta1" = 3, "Beta2" = 0, "Epsilon1" = 0, "Epsilon2" = 4) #used to be 4
   
   settings <- make_settings(n_x = 500,
                             Region=example$Region,

### 2023-01-12 VAST/Test_3Species_GeorgesBank_YT_Only.R

@@ -483,7 +483,7 @@ for(j in 1:6){
   # FC1[2,2] <-0
   
   FC1 = c("Omega1" = 0, "Epsilon1" =0, "Omega2" = 1, "Epsilon2" = 1) 
-  RhoConfig = c("Beta1" = 3, "Beta2" = 3, "Epsilon1" = 0, "Epsilon2" = 4)
+  RhoConfig = c("Beta1" = 3, "Beta2" = 0, "Epsilon1" = 0, "Epsilon2" = 4)
   
   #FieldConfig = c("Omega1"=0, "Epsilon1"=0, "Omega2"="IID", "Epsilon2"=0
   
@@ -732,7 +732,7 @@ for(j in 1:6){
   
   
   FC2 = c("Omega1" = 0, "Epsilon1" =0, "Omega2" = 1, "Epsilon2" = 1) 
-  RhoConfig = c("Beta1" = 3, "Beta2" = 3, "Epsilon1" = 0, "Epsilon2" =4)
+  RhoConfig = c("Beta1" = 3, "Beta2" = 0, "Epsilon1" = 0, "Epsilon2" = 4)
   
   settings <- make_settings(n_x = 500,
                             Region=example$Region,

### 2023-01-12 VAST/Test_3Species_GeorgesBank_Cod_Only.R
vast update

@@ -21,7 +21,7 @@ surv_random_sample <- as.matrix(surv_random_sample,ncol= 12)
 
 scenario1 <- "DecPop_IncTemp"
 #survey results without noise
-list_all <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Results\\",scenario1,"\\list_all_",scenario1,".RDS",sep=""))
+list_all <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Simulation_Results\\",scenario1,"\\list_all_",scenario1,".RDS",sep=""))
 
 exclude_strata <- FALSE
 

### 2023-01-12 VAST/Test_3Species_GeorgesBank_Had_Only.R

@@ -20,11 +20,11 @@ orig.dir <- getwd()
 
 scenario1 <- "IncPop_ConTemp"
 #survey results without noise
-list_all <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Results\\",scenario1,"\\list_all_",scenario1,".RDS",sep=""))
+list_all <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Simulation_Results\\",scenario1,"\\list_all_",scenario1,".RDS",sep=""))
 
 exclude_strata <- FALSE
 
-with_noise <- TRUE
+with_noise <- FALSE
 
 covariates <- FALSE
 

### 2023-01-12 VAST/Test_3Species_GeorgesBank_YT_Only.R

@@ -20,7 +20,7 @@ orig.dir <- getwd()
 
 scenario1 <- "DecPop_ConTemp"
 #survey results without noise
-list_all <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Results\\",scenario1,"\\list_all_",scenario1,".RDS",sep=""))
+list_all <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Simulation_Results\\",scenario1,"\\list_all_",scenario1,".RDS",sep=""))
 
 exclude_strata <- FALSE
 

### 2023-01-19 Manuscript/Draft1cml.tex

--- 2023-01-09 Manuscript/Draft1.tex
@@
 \linenumbers
-\usepackage[belowskip=0pt,aboveskip=0pt]{caption} \usepackage{array} \usepackage{caption} \usepackage{graphicx} \usepackage{siunitx} \usepackage{colortbl} \usepackage{multirow} \usepackage{hhline} \usepackage{calc} \usepackage{tabularx} \usepackage{tabulary} \usepackage{threeparttable} \usepackage{wrapfig}
+\usepackage[belowskip=0pt,aboveskip=0pt]{caption}
 \usepackage{booktabs}
@@
 
-{[}{[}Chris comments in double square brackets, search for them to see comments{]}{]}
-
 \newpage
@@
 
-much of below is from \url{https://apps-nefsc.fisheries.noaa.gov/nefsc/ecosystem-ecology/}
-or
-\url{https://www.fisheries.noaa.gov/data-tools/fisheries-economics-united-states-data-and-visualizations}
-
-\begin{itemize}
-\tightlist
-\item
-  The eastern continental shelf is ecologically diverse and economically important
-\end{itemize}
-
-The Northeast United States continental shelf spans from the Outer Banks of North Carolina to the Gulf of Maine.
-The region covers over 250,000 km\(^2\) of ocean, extending over 200 km from shore in the largest areas in New England to just 30 km off shore in the southern regions.
-This ecologically diverse region contains approximately 18,000 vertebrate marine species.
-Commercial fisheries have been an important part of local economies for centuries.
-In 2019, New England fisheries produced \$22 billion in sales, which sustained over 200,000 jobs.
-Maintaining a healthy ecosystem is therefore vital to sustained ecological health and economic prosperity of the region.
-
-\begin{itemize}
-\tightlist
-\item
-  Bottom trawl survey is important for monitoring population trends
-\end{itemize}
-
-Fish stocks in this highly productive and economically important region are managed by the National Oceanic and Atmospheric Administration's (NOAA) Northeast Fisheries Science Center (NEFSC) in Woods Hole, Massachusetts.
-Federal scientists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (Azarovitz 1981; Politis et al.
-2014).
-The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf.
-The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits \textbf{just listed a few uses of survey.
-change/add others?}.
-
-The survey takes place twice each year- once in the spring and again in the fall.
-Since most spatial analyses and projections of future distributions typically assume a constant survey catchability and/or availability over time, NOAA's survey design includes sampling during approximately the same 2-3 week time period in each season.
-
-\begin{itemize}
-\tightlist
-\item
-  Climate change is happening
-\end{itemize}
-
-Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades, resulting in a shift in spatial distributions of many species.
-Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments.
-More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey will change, altering the relationship between the index and the true population.
-We are therefore interested in analyzing the impact of climate change on the accuracy of future stock assessment models as measured by NOAA's ongoing bottom-trawl survey along the East coast.
-
-\textbf{use more info from initial proposal}
-
-\begin{itemize}
-\tightlist
-\item
-  Briefly describe our study to test this
-\end{itemize}
-
-To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depend on temperature preferences.
-We can then consider the impact of climate change by simulating scenarios with repeating temperature patterns and those where temperature increases on average over time.
-In both cases we analyze the ability of stratified random sampling to track population trends.
-
 \section{Methods}
-
-\begin{itemize}
-\tightlist
-\item
-  Describe simulation study
-\end{itemize}
-
-We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on Georges Bank, where movement of each species combine static species-specific habitat preferences with temperature preferences.
-Model dynamics are driven by a time series of temperature gradients that were estimated from data to create simulated data sets for each population where the true biomass is known.
-Using temperature gradients that repeat each year creates data sets with predictable, repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time, approximating the future distribution in the region under climate change.
-We conducting stratified random sampling on our simulation output to mimic the bottom trawl survey and compare the ability of contemporary indexing methods to track population trends.
-
-\subsection{Population Model Formulation}
 
@@
 
-We use the R package \emph{MixFishSim} (MFS) to model our populations (Dolder et al.
-2020).
+We use the R package \emph{MixFishSim} (MFS) to model our populations (\textbf{dolder2020highly?}).
 MFS is a discrete spatiotemporal simulation tool where users can model multiple species under varying environmental conditions.
@@
 We formulate the following inputs for the MFS package to address our research question.
-{[}{[}perhaps for discussion, but we should mention somewhere that MFS only tracks biomass not age structure, the latter is important for stock assessments but not considered in this paper{]}{]}
-
-\emph{Study Area}
-
-We obtained a shapefile for the 15 strata that comprise Georges Bank, where strata were partitioned based primarily on depth and secondarily by latitude (Politis et al.
-2014).
-The region was discritized into a raster with 88 rows and 144 columns to use as our modeling environment.
-A fish stock is considered to be a subpopulation of a species that has similar intrinsic parameters.
-Each of the species being modeled has multiple distinct stocks along the Atlantic coast resulting from local environmental conditions.
-Biological differences between species results in each stock inhabiting a different number of strata on Georges Bank.
-Haddock inhabit all 15 strata in the domain, Cod inhabit 13 strata, and Yellowtail exist in 9 strata.
-Figure \ref{fig:strata-plot} shows the regions used in our models.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{Images/Strata}
-
-}
-
-\caption{Strata inhabited by each species in our population models.}\label{fig:strata-plot}
-\end{figure}
-
-\emph{Population Dynamics and Recruitment}
-
-The time step for our models is one week.
-MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass in each cell in our study area (Dolder et al.
-2020).
-Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits.
-{[}{[}not sure we need to highlight the recruitment model, might be able to just use the table because Bev-Holt formulation is well known in fisheries{]}{]}\textbf{BL: MFS has 2 recruitment options and I edited the package to use previous years biomass} We chose to represent recruitment in the model using a Beverton-Holt formulation \textbf{cite}.
-Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.
-Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics.
-A full list of parameters used in our model can be seen below in Tables \ref{paramsALL} and \ref{tab:paramsSCENARIOS}.
-
-\emph{Movement}
-
-The package was designed to generate theoretical habitat preferences using Gaussian Random Fields that combine with hypothetical temperature gradients to drive the probability of movement from cell \(I\) to cell \(J\) using the formulation
-
-\begin{align}
-Pr(C_{wk+1}=J|C_{wk}=I) = \frac{e^{-\lambda \cdot d_{I,J}}\cdot(Hab^2_{J,s} \cdot Tol_{J,s,wk})}{\sum^C_{c=1}e^{-\lambda \cdot d} \cdot (Hab^2_{c,s} \cdot Tol_{c,s,wk})},
-\label{moveP}
-\end{align}
-
-where
-
-\(e^{-\lambda \cdot d_{I,J}}\) accounts for distance between cells \(I\) and \(J\),
-
-\(Hab^2_{J,s}\) is the static habitat value for species \(s\) in cell \(J\), and
-
-\(Tol_{c,s,wk}\) is the value from normally distributed temperature tolerance for species \(s\) in cell \(c\) in week \(wk\).
-
-The following sections describe how we formulated the habitat and temperature components to model real species on the northeast continental shelf.
-
-\emph{Habitat Input}
-
-Species-specific habitat preferences were derived using the \emph{lrren} tool from the R package \emph{envi} (Buller 2022) to create a niche model for each species.
-The \emph{lrren} tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors.
-We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow {[}{[}I don't think we need to cite the trawl data because we said we were using trawl data above{]}{]} \textbf{BL: Does trawl data ever get cited?
-It was not cited above.
-cite trawl data?}.
-We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences.
-Depth and mean sediment size were used as our covariate predictors.
-Estimated depth for the region was obtained from FVCOM (Chen et al.
-2006).
-The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method \textbf{cite arcmap} using point data collected by the United States Geologic Survey (USGS) (McMullen et al.
-n.d.).
-Since the values in \(Hab^2_{J,s}\) are required to be between 0 and 1, we transform the spatial estimates from \emph{lrren} to fall between these bounds.
-See Figure \ref{fig:hab-plot1} for a visual representation of this process being applied to Cod.
-Figure \ref{fig:hab-plot} depicts habitat preferences \(Hab^2_{J,s}\) for each species.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{Images/hab_snip3}
-
-}
-
-\caption{Visual representation of niche model for Cod.}\label{fig:hab-plot1}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{Images/Habitat_3species}
-
-}
-
-\caption{Static habitat preferences for each species in our population models (Yellotwtail, Cod, Haddock).}\label{fig:hab-plot}
-\end{figure}
-
-\emph{Temperature Input}
-
-Each species is assumed to have normally distributed temperature preferences (\(N(\mu,\sigma)\)).
-We assume Yellowtail Flounder's preferences are \(N(8.75,4.25)\), while Haddock and Cod have preferences \(N(9,4)\).
-We chose these values by combining information in the literature with temperatures recorded in the bottom trawl survey.
-Weekly estimated temperature data for the region for 2012 was obtained from FVCOM (Chen et al.
-2006).
-The 2012 data was repeated yearly in the constant temperature scenario because displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values.
-We chose to repeat estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data.
-The 2012 temperature data was also transform to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation.
-We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation.
-Figure \ref{fig:temp-scenarios} depicts mean trends for the temperature scenarios used in our models.
-\textbf{dont forget to include gif in final submission}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{Images/TempScenarios}
-
-}
-
-\caption{Mean trends of temperature data used in our model.}\label{fig:temp-scenarios}
-\end{figure}
-
---Describe difference between increasing and constant temperature scenarios (images?)
-
-In equation (1), \(Hab^2_{J,s}\) is constant for the duration of the simulation, while \(Tol_{c,s,wk}\) changes each week.
-Using a temperature gradient that repeats every 52 weeks produces the same spatial preferences in a given week each year, resulting in consistent spatial biomass patterns.
-Scenarios where the temperature increases over time creates spatial preferences that evolve as the water warms, producing spatial biomass patterns that shift in a given week over the duration of the simulation.
-Thus, stratified random samples in scenarios with a repeating temperature pattern will have constant survey catchability and availability over time, which may not be true for increasing temperature scenarios due to evolving spatial preferences.
-{[}{[}\textbf{BL: included in above paragraph} may want to mention here or in the discussion that we used repeating 2012 to simplify the simulations and reduce the number of factors explored that would have occurred with different annual cycles{]}{]}
-
--- Describe each scenario that is considered
-
-We consider 20 year simulations under three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time.
-Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius {[}{[}\textbf{BL: included in above paragraph} need some justification for this large change, wanted to be extreme because if the extreme doesn't make a difference, then more reasonable increases wouldn't make a difference{]}{]} over the duration of the 20 year simulation.
-We therefore simulate 6 scenarios for each population.
-
-\emph{Simulating Bottom Trawl Survey and Population Indexing}
-
--Describe post hoc sampling process and how data is used
-
-After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year.
-We sample each strata in the same weeks in which the Spring and Fall surveys take place and the number of the samples taken reflect true target values for each strata.
-Most strata contain enough cells that we were able to generate a list of sampling locations without replacement, allowing us to sample a unique random cell in each survey over the duration of the simulation.
-For strata that contain small numbers of cells, we generate a random sampling locations with replacement of cells previously sampled.
-We then use the biomass collected from our samples in contemporary population indexing methods to estimate population trends.
-Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.
-\textbf{BL: added both to above paragraph}
-{[}{[}should we mention here or later that we did both with and without measurement error analyses?{]}{]}
-{[}{[}need to make sure that we don't lose sight of random sampling with the following sentences{]}{]}
-
---Stratified mean vs VAST with and without covariates
-{[}{[}could use an intro sentence here to distinguish design-based (stratified mean) and model-based (VAST) estimators{]}{]}
-{[}{[}may want to highlight the potential of VAST with covariates to address the climate change issue here as the reason we are doing these simulations{]}{]}
-\textbf{BL: Chris, please check paragraph below for intro sentence and help fill in X, Y and Z}
-
-Stock assessment scientists use samples from the bottom trawl survey to estimate the abundance for each stock stock.
-There are a number of approaches to obtaining abundance estimates ranging from traditional design-based estimates such as the stratified mean to model-based estimates that range in complexity.
-The stratified mean calculates the geometric mean catch per tow and does not account for spatial variation in samples or allow the inclusion of environmental covariates.
-There are a number of commonly used model-based estimates such as X, Y and Z.
-The Vector Autoregressive Spatio-Temporal (VAST) model is a spatial delta-generalized linear mixed model (Thorson 2019).
-VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurrence (presence/absence).
-If desired, VAST also allows users to include covariate data to better inform the model.
-Covariates can be static (eg.
-habitat preferences) or dynamics (eg.
-temperature).
-We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses.
-The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and take on the order of hours to complete.
-We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
-
-We follow the advice given in (Thorson 2019) to build VAST models to estimate biomass on Georges Bank using stratified mean samples from our model output.
-In addition to exploring different link functions and assumed distributions, our VAST model-building process included testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
-We also carried out the same process both including covariates in our model as well as running models without covariate information.
-We considered covariates in the form of dynamic temperature values and/or static habitat values from our population model.
-When using covariates we ultimately decided to provide the most information to the model by including both covariates for both linear predictors.
-Since we know the true population values in our models we calculate the absolute error of each VAST estimate to compare between potential settings.
-Through this process, and in consultation with the VAST package creator, we determined setting that allowed VAST models to converge for all of our scenarios while also providing the lowest absolute error values.
-Settings for our VAST models can be seen in Table ???.
-
-Our goal is to determine indexing approaches and settings that are robust to future environmental conditions and resulting spatial biomass patterns.
-An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists.
-These assumptions can be questioned given enough noise in the sampling process \textbf{cite?} and/or climate change causing a population to move into previously uninhabited strata.
-To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values.
-\textbf{BL: Help with correct notation for adding noise}.
-We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges bank to those that only include a subset of the full spatial domain for each species.
-We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from.
-Tabl XXX lists all strata inhabited by each species, those that are removed from certain calculations, and the explanation of why these strata were removed.
-
-\textbf{BL: I tried to provid emore detail above and also added another table}
-{[}{[}may need to spell the aspect of reduced spatial domain a bit more because it might be confusing to readers who are expected us to just add areas around the current ones instead of reducing the strata{]}{]}
-
-When combining population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, covariates) there are a large number of scenario combinations to consider.
-The columns in Table \ref{tab:scenarios} show the choices that define each scenario.
-
-\begin{table}
-
-\caption{\label{tab:scenarios}Each index estimate chooses one condition from each of the following
-225 columns. There are 3*3*2*2*2*2*2 = 288 VAST model combinations and 3*3*2*2*2 = 144
-226 stratified mean estimates.}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lllllll}
-\toprule
-Species & Population Trend & Temperature Scenario & Strata Included & Noise Added & Covariates (VAST) & Season\\
-\midrule
-Yellowtail & Increasing & Repeating & All strata & No & Spring & \\
-Cod & Constant & Increasing 5°C & Subset & Yes & Fall & \\
-Haddock & Decreasing &  &  &  &  & \\
-\bottomrule
-\end{tabular}}
-\end{table}
-
-\section{Results}
-
-The goal of our project was to analyze how well contemporary population indexing methods can track population trends under a host of conditions, as depicted in Table COMBOS.
-Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year {[}{[}can cite the 2022 management track assessments, see \url{https://apps-nefsc.fisheries.noaa.gov/saw/sari.php} for when the document becomes available{]}{]} \textbf{cite}.
-For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod (see Table ???) and increasing population scenarios for Haddock (see Table ???).
-To provide a comprehensive analysis of population indexing methods we consider all possible scenario combinations for Yellowtail Flounder (see Table ???).
-The specific population trends used in our analyses can be see in Figure \ref{fig:pop-scenarios}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.99\linewidth]{Images/Population_scenarios}
-
-}
-
-\caption{True population trends used in indexing analyses. Spring biomass plots are shown with fall values being very similar.}\label{fig:pop-scenarios}
-\end{figure}
-
-General themes that exist in Tables X, Y and Z are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors.
-We also see individual cases where the stratified mean produced the lowest absolut error and instances where including covarites in VAST models actually increase the absolute error.
-When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error {[}{[}as expected{]}{]}, though there are some scenarios where the impact is minimal.
-
-In the Cod results, when using reduced strata adding covariates produces worst VAST results (though still better than stratified mean).
-VAST without covariates much worse than stratified mean in fall with increasing temperature and all strata, but adding covariates corrects this.
-
-For Haddock, VAST has a particularly hard time in spring regularly producing larger errors than the stratified mean with added covariates only improving to the level of the stratified mean.
-VAST shows improved results in fall relative to the stratified mean with added covaraites producing extremely low errors in some cases.
-
-In considering the Yellowtail Flounder results in Table XXX, we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
-However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in the Table.
-These errors are corrected by including covariate information allowing for an improved VAST estimate that are significantly lower than their stratified mean counterparts.
-
-{[}{[}We'll want to expand the results section to focus on the impact of each factor one at at time within each species.
-This will be a bit dull, but it is important to walk through the results in words to ensure that readers get the message.
-They can of course examine the tables in detail and draw their own conclusions, but we should put our interpretation down on paper.{]}{]}
-
-\begin{table}
-
-\caption{\label{tab:unnamed-chunk-1}Yellowtail error results}
-\centering
-\begin{tabular}[t]{l|l|l|l|l|l|l|l|l|l|l}
-\hline
-X & X.1 & X.2 & X.3 & X.4 & Constant.Population & X.5 & Increasing.Population & X.6 & Decreasing.Population & X.7\\
-\hline
-Temp & Covariate & Strata & Noise & season & Stratified Mean & VAST Estimate & Stratified Mean & VAST Estimate & Stratified Mean & VAST Estimate\\
-\hline
-const & no cov & all & no & spring & 0.21 & 0.11 & 0.16 & 0.13 & 0.23 & 0.08\\
-\hline
-const & no cov & all & yes & spring & 0.25 & 0.16 & 0.22 & 0.21 & 0.27 & 0.11\\
-\hline
-const & w/ cov & all & no & spring & 0.21 & 0.07 & 0.16 & 0.06 & 0.23 & 0.06\\
-\hline
-const & w/ cov & all & yes & spring & 0.25 & 0.08 & 0.22 & 0.07 & 0.27 & 0.07\\
-\hline
-const & no cov & all & no & fall & 0.32 & 0.68 & 0.34 & 0.36 & 0.41 & 0.81\\
-\hline
-const & no cov & all & yes & fall & 0.31 & 0.77 & 0.46 & 0.44 & 0.44 & 1.09\\
-\hline
-const & w/ cov & all & no & fall & 0.32 & 0.08 & 0.34 & 0.08 & 0.41 & 0.08\\
-\hline
-const & w/ cov & all & yes & fall & 0.31 & 0.11 & 0.46 & 0.17 & 0.44 & 0.18\\
-\hline
-const & no cov & reduced & no & spring & 0.27 & 0.19 & 0.2 & 0.15 & 0.25 & 0.19\\
-\hline
-const & no cov & reduced & yes & spring & 0.26 & 0.15 & 0.22 & 0.11 & 0.3 & 0.16\\
-\hline
-const & w/ cov & reduced & no & spring & 0.27 & 0.19 & 0.2 & 0.17 & 0.25 & 0.19\\
-\hline
-const & w/ cov & reduced & yes & spring & 0.26 & 0.14 & 0.22 & 0.13 & 0.3 & 0.16\\
-\hline
-const & no cov & reduced & no & fall & 0.47 & 0.25 & 0.41 & 0.11 & 0.55 & 0.24\\
-\hline
-const & no cov & reduced & yes & fall & 0.49 & 0.36 & 0.46 & 0.21 & 0.53 & 0.36\\
-\hline
-const & w/ cov & reduced & no & fall & 0.47 & 0.19 & 0.41 & 0.22 & 0.55 & 0.23\\
-\hline
-const & w/ cov & reduced & yes & fall & 0.49 & 0.17 & 0.46 & 0.19 & 0.53 & 0.24\\
-\hline
-increasing & no cov & all & no & spring & 0.28 & 0.11 & 0.32 & 0.13 & 0.22 & 0.15\\
-\hline
-increasing & no cov & all & yes & spring & 0.28 & 0.15 & 0.34 & 0.16 & 0.26 & 0.17\\
-\hline
-increasing & w/ cov & all & no & spring & 0.28 & 0.06 & 0.32 & 0.07 & 0.22 & 0.07\\
-\hline
-increasing & w/ cov & all & yes & spring & 0.28 & 0.12 & 0.34 & 0.1 & 0.26 & 0.1\\
-\hline
-increasing & no cov & all & no & fall & 0.51 & 1.26 & 0.3 & 0.66 & 0.28 & 1.06\\
-\hline
-increasing & no cov & all & yes & fall & 0.5 & 1.38 & 0.39 & 0.71 & 0.25 & 1.1\\
-\hline
-increasing & w/ cov & all & no & fall & 0.51 & 0.23 & 0.3 & 0.21 & 0.28 & 0.15\\
-\hline
-increasing & w/ cov & all & yes & fall & 0.5 & 0.28 & 0.3 & 0.37 & 0.25 & 0.2\\
-\hline
-increasing & no cov & reduced & no & spring & 0.31 & 0.17 & 0.4 & 0.33 & 0.32 & 0.14\\
-\hline
-increasing & no cov & reduced & yes & spring & 0.29 & 0.19 & 0.41 & 0.26 & 0.29 & 0.15\\
-\hline
-increasing & w/ cov & reduced & no & spring & 0.31 & 0.17 & 0.4 & 0.3 & 0.32 & 0.22\\
-\hline
-increasing & w/ cov & reduced & yes & spring & 0.29 & 0.15 & 0.41 & 0.25 & 0.29 & 0.21\\
-\hline
-increasing & no cov & reduced & no & fall & 0.64 & 0.75 & 0.7 & 0.49 & 0.54 & 0.6\\
-\hline
-increasing & no cov & reduced & yes & fall & 0.66 & 0.89 & 0.69 & 0.48 & 0.53 & 0.62\\
-\hline
-increasing & w/ cov & reduced & no & fall & 0.64 & 0.2 & 0.7 & 0.53 & 0.54 & 0.31\\
-\hline
-increasing & w/ cov & reduced & yes & fall & 0.66 & 0.13 & 0.69 & 0.5 & 0.53 & 0.32\\
-\hline
-\end{tabular}
-\end{table}
-
-\begin{table}
-
-\caption{\label{tab:unnamed-chunk-1}Cod error results}
-\centering
-\begin{tabular}[t]{l|l|l|l|r|r|r}
-\hline
-Temp & Strata & Noise & season & VAST.No.Cov & VAST.w..Cov & Stratified.Mean\\
-\hline
-constant & all & no & spring & 0.11 & 0.12 & 0.36\\
-\hline
-constant & all & yes & spring & 0.12 & 0.15 & 0.35\\
-\hline
-constant & all & no & fall & 0.19 & 0.05 & 0.49\\
-\hline
-constant & all & yes & fall & 0.30 & 0.23 & 0.41\\
-\hline
-constant & reduced & no & spring & 0.17 & 0.24 & 0.41\\
-\hline
-constant & reduced & yes & spring & 0.20 & 0.23 & 0.46\\
-\hline
-constant & reduced & no & fall & 0.21 & 0.33 & 0.60\\
-\hline
-constant & reduced & yes & fall & 0.18 & 0.31 & 0.58\\
-\hline
-increasing & all & no & spring & 0.12 & 0.15 & 0.25\\
-\hline
-increasing & all & yes & spring & 0.19 & 0.19 & 0.27\\
-\hline
-increasing & all & no & fall & 0.76 & 0.13 & 0.45\\
-\hline
-increasing & all & yes & fall & 0.89 & 0.33 & 0.44\\
-\hline
-increasing & reduced & no & spring & 0.14 & 0.22 & 0.32\\
-\hline
-increasing & reduced & yes & spring & 0.15 & 0.21 & 0.29\\
-\hline
-increasing & reduced & no & fall & 0.60 & 0.31 & 0.54\\
-\hline
-increasing & reduced & yes & fall & 0.62 & 0.32 & 0.53\\
-\hline
-\end{tabular}
-\end{table}
-
-\begin{table}
-
-\caption{\label{tab:unnamed-chunk-1}Haddock error results}
-\centering
-\begin{tabular}[t]{l|l|l|l|r|r|r|l|l}
-\hline
-Temp & Strata & Noise & season & VAST.No.Cov & VAST.w..Cov & Stratified.Mean & X & X.1\\
-\hline
-const & all & no & spring & 0.49 & 0.18 & 0.18 &  & \\
-\hline
-const & all & yes & spring & 0.73 & 0.43 & 0.21 &  & Haddock\\
-\hline
-const & all & no & fall & 0.28 & 0.05 & 0.26 &  & Increasing Population\\
-\hline
-const & all & yes & fall & 0.41 & 0.06 & 0.27 &  & \\
-\hline
-const & reduced & no & spring & 0.34 & 0.35 & 0.45 &  & \\
-\hline
-const & reduced & yes & spring & 0.30 & 0.33 & 0.46 &  & \\
-\hline
-const & reduced & no & fall & 0.36 & 0.48 & 0.54 &  & \\
-\hline
-const & reduced & yes & fall & 0.33 & 0.46 & 0.52 &  & \\
-\hline
-increasing & all & no & spring & 0.25 & 0.05 & 0.26 &  & \\
-\hline
-increasing & all & yes & spring & 0.30 & 0.06 & 0.31 &  & \\
-\hline
-increasing & all & no & fall & 0.89 & 0.23 & 0.40 &  & \\
-\hline
-increasing & all & yes & fall & 1.04 & 0.35 & 0.42 &  & \\
-\hline
-increasing & reduced & no & spring & 0.32 & 0.40 & 0.44 &  & \\
-\hline
-increasing & reduced & yes & spring & 0.38 & 0.37 & 0.37 &  & \\
-\hline
-increasing & reduced & no & fall & 0.44 & 0.64 & 0.72 &  & \\
-\hline
-increasing & reduced & yes & fall & 0.42 & 0.62 & 0.70 &  & \\
-\hline
-\end{tabular}
-\end{table}
-
-\section{Discussion}
-
-{[}{[} some things that I think we'll need to address, in addition to those mentioned above are (in no particular order):
-number of simulations done for each scenario - time required to do them and analyses
-pros and cons of using VAST with covariates in a changing environment
-implications for current practice of using stratified mean (not horrible, but might be able to do better with model-based using covariates if enviro changing, note we don't want to say stratified mean should not be used because that will draw a lot of tomato throwing)
-future work - some of the ideas we've discussed already could be mentioned
-limitations to our study
-perhaps for discussion, but we should mention somewhere that MFS only tracks biomass not age structure, the latter is important for stock assessments but not considered in this paper{]}{]}
-
-\section{Acknowledgements}
-
-{[}{[}we should thank Jim (obviously), but also those who helped you with the habitat data (robyn, david Chev.) and others?
-also should note that funding provided by Climate-Groundfish source (I'll dig up the official name of the funding source){]}{]}
-
-\hypertarget{data-and-code-availability}{%
-\subsection{Data and Code Availability}\label{data-and-code-availability}}
-
-All data and code used in this work are available at \url{https://github.com/Blevy2/READ-PDB-blevy2-MFS2}.
 
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-azarovitz1981brief}{}}%
-Azarovitz, T. 1981. A brief historical review of the woods hole laboratory trawl survey time series. Bottom trawl surveys.
-
-\leavevmode\vadjust pre{\hypertarget{ref-envi}{}}%
-Buller, I.D.
-2022.
-Envi: Environmental interpolation using spatial kernel density estimation.
-The Comprehensive R Archive Network.
-doi:\href{https://doi.org/10.5281/zenodo.5347826}{10.5281/zenodo.5347826}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-chen2006unstructured}{}}%
-Chen, C., Beardsley, R.C., Cowles, G., Qi, J., Lai, Z., Gao, G., and others. 2006. An unstructured grid, finite-volume coastal ocean model: FVCOM user manual. SMAST/UMASSD: 6--8.
-
-\leavevmode\vadjust pre{\hypertarget{ref-dolder2020highly}{}}%
-Dolder, P.J., Minto, C., Guarini, J.-M., and Poos, J.J. 2020. Highly resolved spatiotemporal simulations for exploring mixed fishery dynamics. Ecological Modelling \textbf{424}: 109000. Elsevier.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mcmullen2005usgs}{}}%
-McMullen, K., Paskevich, V., and Poppe, L. (n.d.). USGS east-coast sediment analysis: Procedures, database, and GIS data. US Geological Survey Open-File Report \textbf{2005}: 1001.
-
-\leavevmode\vadjust pre{\hypertarget{ref-politis2014northeast}{}}%
-Politis, P.J., Galbraith, J.K., Kostovick, P., and Brown, R.W. 2014. Northeast fisheries science center bottom trawl survey protocols for the NOAA ship henry b. bigelow.
-
-\leavevmode\vadjust pre{\hypertarget{ref-thorson2019guidance}{}}%
-Thorson, J.T.
-2019.
-Guidance for decisions using the vector autoregressive spatio-temporal (VAST) package in stock, ecosystem, habitat and climate assessments.
-Fisheries Research \textbf{210}: 143--161.
-Elsevier.
-
+\begin{CSLReferences}{0}{0}
 \end{CSLReferences}
@@
 
-\pagebreak
-
-Table 1. Parameters used in all population models.
-
-\begin{longtable}[]{@{}lllllll@{}}
-\toprule
-Parameter & Description & Unit & Yellowtail & Cod & Haddock & Source \\
-\midrule
-\endhead
-\(\rho\) & Ford's growth coefficient & wk\(^{-1}\) & 4.48 & 4.43 & 4.49 & \\
-\(M\) & Natural Mortality & wk\(^{-1}\) & 0.2064 & 0.2728 & 0.3340 & \\
-\(W_R\) & Weight of fully recruited fish & kg & 0.39 & 2.95 & 1.12 & \\
-\(W_{R-1}\) & Weight of pre-recruit fish & kg & 0.13 & 0.39 & 0.19 & \\
-\(\sigma^2\) & Variance in recruited fish & kg\(^2\) & 0.55 & 0.55 & 0.55 & \\
-\(\lambda\) & Decay rate for movement & - & 0.7 & 0.7 & 0.7 & \\
-\(Spwn_s\) & Spawning weeks for species \(s\) & wk & 9-12 & 8-13 & 11-14 & \\
-\(Rec_s\) & Recruitment weeks for species \(s\) & wk & 9-12 & 8-13 & 11-14 & \\
-\bottomrule
-\end{longtable}
-
-\begin{table}
-
- \caption{\label{tab:paramsALL2}Parameters used in all population models.}
- \centering
- \fontsize{10}{12}\selectfont
- \begin{tabular}[t]{lllllll}
- \toprule
- Parameter & Description & Unit & Yellowtail & Cod & Haddock & Source\\
- \midrule
- <U+03C1> & Ford's growth coefficient & 1/wk & 4.48 & 4.43 & 4.49 & \\
- M & Natural mortality & 1/wk & 0.2064 & 0.2728 & 0.334 & \\
- W1 & Weight of fully recruited fish & kg & 0.39 & 2.95 & 1.12 & \\
- W2 & Weight of pre-recruit fish & kg & 0.13 & 0.39 & 0.19 & \\
- sigma & Variance in recruited fish & kg*kg & 0.55 & 0.55 & 0.55 & \\
- \addlinespace
- lambda & Decay rate for movement & - & 0.7 & 0.7 & 0.7 & \\
- spwn & Spawning weeks for species s & wks & 9-12 & 8-13 & 11-14 & \\
- rec & Recruitment weeks for species s & wks & 9-12 & 8-13 & 11-14 & \\
- \bottomrule
- \end{tabular}
- \end{table}
-
-\begin{table}
-
-\caption{\label{tab:paramsSCENARIOS}Parameters used in population models for each scenario.}
-\centering
-\fontsize{10}{12}\selectfont
-\begin{tabular}[t]{llllll}
-\toprule
-Parameter & Description & Unit & Yellowtail & Cod & Haddock\\
-\midrule
-\addlinespace[0.3em]
-\multicolumn{1}{l}{\textbf{Constant Population}}\\
-\hspace{1em}M+F & Adjusted Mortality (Natural + Fishing) & 1/wk & 0.764 & 0.83 & 0.309\\
-\hspace{1em}P0 & Initial Biomass & kg & 3190 & 21500 & \vphantom{1} 180000\\
-\hspace{1em}a & Max recruitment rate & kg & 30400 & 27900 & 73600\\
-\hspace{1em}ß & Recruitment half saturation value & kg & 4300 & 10500 & 40500\\
-\addlinespace[0.3em]
-\multicolumn{1}{l}{\textbf{Decreasing Population}}\\
-\hspace{1em}M+F & Adjusted Mortality (Natural + Fishing) & 1/wk & 0.764 & 0.623 & 0.334\\
-\hspace{1em}P0 & Initial Biomass & kg & 50000 & 21500 & 180000\\
-\hspace{1em}a & Max recruitment rate & kg & 1.07e+12 & 3.89e+08 & 4.97e+08\\
-\hspace{1em}ß & Recruitment half saturation value & kg & 2.3e+12 & 9.8e+08 & 2.08e+09\\
-\addlinespace[0.3em]
-\multicolumn{1}{l}{\textbf{Increasing Population}}\\
-\hspace{1em}M+F & Adjusted Mortality (Natural + Fishing) & 1/wk & 0.564 & 0.372 & 0.134\\
-\hspace{1em}P0 & Initial Biomass & kg & 3190 & 21500 & 180000\\
-\hspace{1em}a & Max recruitment rate & kg & 40000 & 45000 & 1e+05\\
-\hspace{1em}ß & Recruitment half saturation value & kg & 43000 & 62800 & 405000\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-Table XX. Parameters used in all VAST models.
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.13}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.53}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.33}}@{}}
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-Parameter
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Description
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Input/Value
-\end{minipage} \\
-\midrule
-\endhead
-\emph{ObsModel} & Link function and assumed distribution & c(10,2) \\
-FieldCOnfig & Specified spatial and/or spatio-temporal variation in predictors & c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1) \\
-RhoConfig & Specifying whether intercepts or spatio-temporal variation is structured among time intervals & c(Beta1=3, Beta2=3, Epsilon1=0, Epsilon2=4) \\
-X1\_formula & Right-sided formula affecting the 1st linear predictor & X1\_formula = \textasciitilde{} poly(Temp, degree=2 ) \\
-X2\_formula & Right-sided formula affecting the 2nd linear predictor & X2\_formula = \textasciitilde{} poly(Temp, degree=2 ) + poly(Habitat, degree=2 ) \\
-\bottomrule
-\end{longtable}
-
-\begin{table}
-\centering\begingroup\fontsize{10}{12}\selectfont
-
-\begin{tabular}{l|l|l|l|l|l|r|r}
-\hline
-Temp & Covariate & Strata & Noise & Season & Scenario & Stratified Mean & VAST Estimate\\
-\hline
-\multicolumn{1}{l}{\textbf{all}}\\
-\hline
-\multicolumn{1}{l}{\textbf{Constant Population}}\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & all & no & spring & Constant Population & 0.21 & 0.11\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & all & yes & spring & Constant Population & 0.25 & 0.16\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & no & spring & Constant Population & 0.21 & 0.07\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & yes & spring & Constant Population & 0.25 & 0.08\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & all & no & fall & Constant Population & 0.32 & 0.68\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & all & yes & fall & Constant Population & 0.31 & 0.77\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & no & fall & Constant Population & 0.32 & 0.08\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & yes & fall & Constant Population & 0.31 & 0.11\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & no & spring & Constant Population & 0.27 & 0.19\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & yes & spring & Constant Population & 0.26 & 0.15\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & no & spring & Constant Population & 0.27 & 0.19\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & yes & spring & Constant Population & 0.26 & 0.14\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & no & fall & Constant Population & 0.47 & 0.25\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & yes & fall & Constant Population & 0.49 & 0.36\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & no & fall & Constant Population & 0.47 & 0.19\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & yes & fall & Constant Population & 0.49 & 0.17\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & no & spring & Constant Population & 0.28 & 0.11\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & yes & spring & Constant Population & 0.28 & 0.15\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & no & spring & Constant Population & 0.28 & 0.06\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & yes & spring & Constant Population & 0.28 & 0.12\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & no & fall & Constant Population & 0.51 & 1.26\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & yes & fall & Constant Population & 0.50 & 1.38\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & no & fall & Constant Population & 0.51 & 0.23\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & yes & fall & Constant Population & 0.50 & 0.28\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & no & spring & Constant Population & 0.31 & 0.17\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & yes & spring & Constant Population & 0.29 & 0.19\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & no & spring & Constant Population & 0.31 & 0.17\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & yes & spring & Constant Population & 0.29 & 0.15\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & no & fall & Constant Population & 0.64 & 0.75\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & yes & fall & Constant Population & 0.66 & 0.89\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & no & fall & Constant Population & 0.64 & 0.20\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & yes & fall & Constant Population & 0.66 & 0.13\\
-\hline
-\multicolumn{1}{l}{\textbf{Increasing Population}}\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & all & no & spring & Increasing Population & 0.16 & 0.13\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & all & yes & spring & Increasing Population & 0.22 & 0.21\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & no & spring & Increasing Population & 0.16 & 0.06\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & yes & spring & Increasing Population & 0.22 & 0.07\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & all & no & fall & Increasing Population & 0.34 & 0.36\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & all & yes & fall & Increasing Population & 0.46 & 0.44\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & no & fall & Increasing Population & 0.34 & 0.08\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & yes & fall & Increasing Population & 0.46 & 0.17\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & no & spring & Increasing Population & 0.20 & 0.15\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & yes & spring & Increasing Population & 0.22 & 0.11\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & no & spring & Increasing Population & 0.20 & 0.17\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & yes & spring & Increasing Population & 0.22 & 0.13\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & no & fall & Increasing Population & 0.41 & 0.11\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & yes & fall & Increasing Population & 0.46 & 0.21\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & no & fall & Increasing Population & 0.41 & 0.22\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & yes & fall & Increasing Population & 0.46 & 0.19\\
-\hline
-\multicolumn{1}{l}{\textbf{reduced}}\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & no & spring & Increasing Population & 0.32 & 0.13\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & yes & spring & Increasing Population & 0.34 & 0.16\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & no & spring & Increasing Population & 0.32 & 0.07\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & yes & spring & Increasing Population & 0.34 & 0.10\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & no & fall & Increasing Population & 0.30 & 0.66\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & yes & fall & Increasing Population & 0.39 & 0.71\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & no & fall & Increasing Population & 0.30 & 0.21\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & yes & fall & Increasing Population & 0.30 & 0.37\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & no & spring & Increasing Population & 0.40 & 0.33\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & yes & spring & Increasing Population & 0.41 & 0.26\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & no & spring & Increasing Population & 0.40 & 0.30\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & yes & spring & Increasing Population & 0.41 & 0.25\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & no & fall & Increasing Population & 0.70 & 0.49\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & yes & fall & Increasing Population & 0.69 & 0.48\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & no & fall & Increasing Population & 0.70 & 0.53\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & yes & fall & Increasing Population & 0.69 & 0.50\\
-\hline
-\multicolumn{1}{l}{\textbf{Decreasing Population}}\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & all & no & spring & Decreasing Population & 0.23 & 0.08\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & all & yes & spring & Decreasing Population & 0.27 & 0.11\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & no & spring & Decreasing Population & 0.23 & 0.06\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & yes & spring & Decreasing Population & 0.27 & 0.07\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & all & no & fall & Decreasing Population & 0.41 & 0.81\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & all & yes & fall & Decreasing Population & 0.44 & 1.09\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & no & fall & Decreasing Population & 0.41 & 0.08\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & yes & fall & Decreasing Population & 0.44 & 0.18\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & no & spring & Decreasing Population & 0.25 & 0.19\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & yes & spring & Decreasing Population & 0.30 & 0.16\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & no & spring & Decreasing Population & 0.25 & 0.19\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & yes & spring & Decreasing Population & 0.30 & 0.16\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & no & fall & Decreasing Population & 0.55 & 0.24\\
-\hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & yes & fall & Decreasing Population & 0.53 & 0.36\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & no & fall & Decreasing Population & 0.55 & 0.23\\
-\hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & yes & fall & Decreasing Population & 0.53 & 0.24\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & no & spring & Decreasing Population & 0.22 & 0.15\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & yes & spring & Decreasing Population & 0.26 & 0.17\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & no & spring & Decreasing Population & 0.22 & 0.07\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & yes & spring & Decreasing Population & 0.26 & 0.10\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & no & fall & Decreasing Population & 0.28 & 1.06\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & yes & fall & Decreasing Population & 0.25 & 1.10\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & no & fall & Decreasing Population & 0.28 & 0.15\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & yes & fall & Decreasing Population & 0.25 & 0.20\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & no & spring & Decreasing Population & 0.32 & 0.14\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & yes & spring & Decreasing Population & 0.29 & 0.15\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & no & spring & Decreasing Population & 0.32 & 0.22\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & yes & spring & Decreasing Population & 0.29 & 0.21\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & no & fall & Decreasing Population & 0.54 & 0.60\\
-\hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & yes & fall & Decreasing Population & 0.53 & 0.62\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & no & fall & Decreasing Population & 0.54 & 0.31\\
-\hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & yes & fall & Decreasing Population & 0.53 & 0.32\\
-\hline
-\end{tabular}
-\endgroup{}
-\end{table}
-
-\begin{table}
-
-\caption{\label{tab:CodERROR}Cod error results.}
-\centering
-\fontsize{10}{12}\selectfont
-\begin{tabular}[t]{lllrrr}
-\toprule
-Strata & Noise & season & VAST No Cov & VAST w/ Cov & Stratified Mean\\
-\midrule
-\addlinespace[0.3em]
-\multicolumn{1}{l}{\textbf{constant}}\\
-\hspace{1em}all & no & spring & 0.11 & 0.12 & 0.36\\
-\hspace{1em}all & yes & spring & 0.12 & 0.15 & 0.35\\
-\hspace{1em}all & no & fall & 0.19 & 0.05 & 0.49\\
-\hspace{1em}all & yes & fall & 0.30 & 0.23 & 0.41\\
-\hspace{1em}reduced & no & spring & 0.17 & 0.24 & 0.41\\
-\hspace{1em}reduced & yes & spring & 0.20 & 0.23 & 0.46\\
-\hspace{1em}reduced & no & fall & 0.21 & 0.33 & 0.60\\
-\hspace{1em}reduced & yes & fall & 0.18 & 0.31 & 0.58\\
-\addlinespace[0.3em]
-\multicolumn{1}{l}{\textbf{increasing}}\\
-\hspace{1em}all & no & spring & 0.12 & 0.15 & 0.25\\
-\hspace{1em}all & yes & spring & 0.19 & 0.19 & 0.27\\
-\hspace{1em}all & no & fall & 0.76 & 0.13 & 0.45\\
-\hspace{1em}all & yes & fall & 0.89 & 0.33 & 0.44\\
-\hspace{1em}reduced & no & spring & 0.14 & 0.22 & 0.32\\
-\hspace{1em}reduced & yes & spring & 0.15 & 0.21 & 0.29\\
-\hspace{1em}reduced & no & fall & 0.60 & 0.31 & 0.54\\
-\hspace{1em}reduced & yes & fall & 0.62 & 0.32 & 0.53\\
-\bottomrule
-\end{tabular}
-\end{table}
-
 \end{document}

### 2023-01-24 Manuscript/Draft1.Rmd
get Unicode to work for Chris (see https://bookdown.org/yihui/rmarkdown/pdf-document.html#latex-engine)

@@
   bookdown::pdf_document2:
+    latex_engine: xelatex
     toc: no

### 2023-01-24 Manuscript/Draft1.Rmd
add Chris comments in Intro

@@
 change/add others?**.
+[[I think these are fine as they are the ones we are concerned with in this paper]]
 
@@
 Since most spatial analyses and  projections of future distributions typically assume a constant survey catchability and/or availability over time, NOAA's survey design includes sampling during approximately the same 2-3 week time period in each season.
+[[the survey for the entire region takes ~8 weeks to complete each season, the GB strata get covered in a similar time period each season because the survey starts in the south and moves north.
+But there are times when logistics, e.g., storms, Covid, cause a change in the timing of any particular stratum being sampled.
+This timing issue has come up in the past and we've looked at it using the dates of sampled stations within a particular stock definition.
+See for example GBYT_survey_timing_cdf.png taken from the 2018 assessment of Georges Bank yellowtail flounder that I copied into a newly created "miscellaneous" folder on the Google Drive.
+Given all this, I suggest modifying the wording to something like "NOAA's survey design includes sampling Georges Bank during approximately the same 3-4 week time period in each season"]]
 

### 2023-01-24 Manuscript/Draft1.Rmd
add a few Chris comments in methods

@@
 The stratified mean calculates the geometric mean catch per tow and does not account for spatial variation in samples or allow the inclusion of environmental covariates.
-Model-based abundance estimates are often derived from model framewroks such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM) that can incorporate both random and mixed effects.
+Model-based abundance estimates are often derived from model framewroks [[frameworks]] such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM) that can incorporate both random and mixed effects.
 The Vector Autoregressive Spatio-Temporal (VAST) model is a spatial delta-generalized linear mixed model [@thorson2019guidance].
@@
 We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses.
-The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and take on the order of hours to complete.
+The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and [[can?]] take on the order of hours to complete.
 We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
@@
 **BL: Help with correct notation for adding noise**.
-We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges bank to those that only include a subset of the full spatial domain for each species.
+We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges bank [[Bank]] to those that only include a subset of the full spatial domain for each species.
 We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from.
-Tabl XXX lists all strata inhabited by each species, those that are removed from certain calculations, and the explanation of why these strata were removed.
+Tabl XXX [[needs to be added]] lists all strata inhabited by each species, those that are removed from certain calculations, and the explanation of why these strata were removed.
 
@@
 General themes that exist in Tables X, Y and Z are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors.
-We also see individual cases where the stratified mean produced the lowest absolut error and instances where including covarites in VAST models actually increase the absolute error.
+We also see individual cases where the stratified mean produced the lowest absolut [[absolute]] error and instances where including covarites in VAST models actually increase the absolute error.
 When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error [[as expected]], though there are some scenarios where the impact is minimal.

### 2023-02-21 Manuscript/Draft1.Rmd
manuscript update and percent area occ. update

@@
 The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf.
-The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits **just listed a few uses of survey.
-change/add others?**.
-[[I think these are fine as they are the ones we are concerned with in this paper]]
+The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits
 
@@
 
-The goal of our project was to analyze how well contemporary population indexing methods can track population trends under a host of conditions, as depicted in Table COMBOS.
-Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year [[can cite the 2022 management track assessments, see https://apps-nefsc.fisheries.noaa.gov/saw/sari.php for when the document becomes available]] **cite**.
+The goal of our project was to analyze how well contemporary population indexing methods can track population trends under a variety of conditions, as depicted in Table \@ref(tab:scenarios) Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year [[can cite the 2022 management track assessments, see https://apps-nefsc.fisheries.noaa.gov/saw/sari.php for when the document becomes available]] **cite**.
 For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod (see Table ???) and increasing population scenarios for Haddock (see Table ???).
@@
 General themes that exist in Tables X, Y and Z are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors.
-We also see individual cases where the stratified mean produced the lowest absolut [[absolute]] error and instances where including covarites in VAST models actually increase the absolute error.
-When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error [[as expected]], though there are some scenarios where the impact is minimal.
+We also see individual cases where the stratified mean produced the lowest absolute error and instances where including covarites in VAST models actually increase the absolute error.
+When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.
 
@@
 
-In considering the Yellowtail Flounder results in Table XXX, we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
+In considering the Yellowtail Flounder results in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata), we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
 However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in the Table.
 These errors are corrected by including covariate information allowing for an improved VAST estimate that are significantly lower than their stratified mean counterparts.
+
+Notes on differences between VAST settings: For YTF, 29/96 ~ 30% of VAST runs with new settings were better (70% were worse).
+80/96 ~ 83% of scenarios had a VAST run with a better error than the stratified mean.
+15/16 of the times the stratified mean was better VAST was not using covariates.
+The 1 time that VAST was using covariates and was still worse than the stratified mean was IncPop_IncTemp_Allstrata_WCov_WNoise in the Fall (VAST had strong overestimate in the fall during IncPOP_IncTemP_allstrata for Had and YT).
+The overestimate in the fall only seems to be related to the way the population shifts between season.
+Looking at the percent shift plots, the population is shifting out of strata 16 in both seasons (large east strata).
+In the spring the population is shifting mostly into strata 17 (thin strata adjacent to 16), but in the fall they are shifting into both 17 and 18 (18 thin one adjacent to 17).
+Thus it seems like with much of the population concentrated in the really small outer most strata, VAST produces an underestimate, even with covariates
+
+For Haddock (IncPop), 18/32 ~ 57% of VAST runs better with new settings.
+24/32 = 75% of scenarios had a VAST run with a better error than the stratified mean.
+14/16 VAST were better with covariates (the two that were worse were essentially the same) while only 10/16 were better without covariates.
+Stratified mean tended to perform better when all strata were included in the calculation, while VAST tended to perform better when strata are removed.
+
+For Cod (DecPop), only 11/32 ~ 34% of VAST runs were better with new settings.
+28/32~ 88% of scenarios had a VAST run with a better estimate than the stratified mean.
+All 16/16 = 100% of VAST runs with a covariate were better than the stratified mean.
+12/16 VAST were better without covariates.
+The 4 runs without covariates that were worse were all extremely large errors and all from increasing temp in the fall (with and without all strata and noise).
+In the fall the population is moving out of strata 16 (large eastern strata) and 21 (north large) and into a number of strata (22 (tiny north), 24 (large north), 18 (tiny north)).
+Might be having trouble tracking in to smaller strata again.
 
@@
 
-```{r}
-
-knitr::kable(YT_results, caption = "Yellowtail error results")
-
-knitr::kable(Cod_results, caption = "Cod error results")
-
-knitr::kable(Had_results, caption = "Haddock error results")
-```
+<!-- # ```{r} -->
+<!-- # -->
+<!-- # -->
+<!-- # knitr::kable(YT_results, caption = "Yellowtail error results") -->
+<!-- # -->
+<!-- # knitr::kable(Cod_results, caption = "Cod error results") -->
+<!-- # -->
+<!-- # knitr::kable(Had_results, caption = "Haddock error results") -->
+<!-- # ``` -->
 
 \section{Discussion}
+
+Estimation methods that have large, inaccurate swings (stratified mean) can lead to changes in quotas that do not correspond to the true population trend, which could have a compounding effect (can lead to quotas that are too high/low given an incorrect assumption of increase/decrease in biomass).
+Our model has a constant assumed mortality that accounts for fishing and natural death and will not account for impacts of these decisions.
+VAST may provide more consistent biomass predictions(?)
+
 [[ some things that I think we'll need to address, in addition to those mentioned above are (in no particular order):
@@
 
-```{r YT_results,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
+<!-- ```{r YT_results,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'} -->
+<!-- library("knitr") -->
+<!-- library("kableExtra") -->
+<!-- library("dplyr") -->
+<!-- library("readr") -->
+<!-- library("tidyr") -->
+<!-- library("forcats") -->
+
+<!-- result_table = read_csv("CSVs/Yellowtail_results_final.csv") -->
+
+<!-- summary_tab <- result_table %>%  -->
+
+<!-- # Group the table according to Parameter, Scenario -->
+<!--   group_by(Scenario,Strata) -->
+
+<!-- # Create table to present the data] -->
+<!-- Fig_5 <- kable(summary_tab) %>% -->
+<!--   kable_styling(font_size = 10)  %>% -->
+<!--   pack_rows(tab_kable, colnum = c(1,4), -->
+<!--     index = table(fct_inorder(summary_tab$Strata), useNA = "no")) %>% -->
+<!--   pack_rows(tab_kable, colnum = 1, -->
+<!--     index = table(fct_inorder(summary_tab$Scenario), useNA = "no"))  -->
+
+<!-- ``` -->
+<!-- ```{r, results='asis'} -->
+<!-- Fig_5 -->
+<!-- ``` -->
+
+```{r YTresultsallstrata,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
 library("knitr")
@@
 
-result_table = read_csv("CSVs/Yellowtail_results_final.csv")
+#result_table = read_csv("CSVs/Yellowtail_results_final.csv")
+result_table = read_csv("CSVs/Yellowtail_results_final_newsettings_forpaper.csv")
 
-summary_tab <- result_table %>%
+summary_tab_all <- result_table %>% 
+  subset(Strata == "all")%>%
+ # subset(Scenario=="Constant Population")%>%
 
-# Group the table according to Parameter, Scenario
-  group_by(Scenario,Strata)
+#   group_by(`Temperature Scenario`)%>%
+# # 
+# #   group_by(Scenario)%>%
+#   mutate(`Temperature Scenario` = factor(`Temperature Scenario`))%>%
+#   
+#   arrange(`Temperature Scenario`) %>%
+  select(Scenario,`Temperature Scenario`,Season,Covariate,Noise,`VAST A`,`VAST B`, `Stratified Mean`)
+
+#COLOR INDIVIDUAL ROWS BY CATEGORY
+#define colors
+
+# summary_tab_all <- result_table %>% 
+# # Group the table according to Parameter, Scenario
+#   group_by(Scenario)%>%
+#   mutate(Scenario = factor(Scenario)) %>%
+#   arrange(Scenario) %>%
+#   select(Scenario, everything())
 
 # Create table to present the data]
-Fig_5 <- kable(summary_tab) %>%
+Fig_5a <- kable(summary_tab_all[, -1],
+      align = c("l", "l", "l", "l", "c", "c", "c"),
+  caption = "Yellowtail flounder error results with all strata included in calculations. Row colors correspond to the same settings applied in different seasons.") %>%
+  row_spec(0,bold = TRUE)%>% #bold column titles
   kable_styling(font_size = 10)  %>%
-  pack_rows(tab_kable, colnum = c(1,4),
-    index = table(fct_inorder(summary_tab$Strata), useNA = "no")) %>%
-  pack_rows(tab_kable, colnum = 1,
-    index = table(fct_inorder(summary_tab$Scenario), useNA = "no"))
+  pack_rows(tab_kable(), colnum = 1,
+    index = table(fct_inorder(summary_tab_all$Scenario), useNA = "no"))%>%
+  #color every 4th row a different color
+  row_spec(seq(1,96,4), bold = T, color = "white", background = "black")%>%
+  row_spec(seq(2,96,4), bold = T, color = "white", background = "#5C5E60")%>%
+  row_spec(seq(3,96,4), bold = T, color = "white", background = "#A4A4A4")%>%
+  row_spec(seq(4,96,4), bold = T, color = "white", background = "#D2D2D2")
 
 ```
+
+```{r YTresultsreducedstrata,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
+library("knitr")
+library("kableExtra")
+library("dplyr")
+library("readr")
+library("tidyr")
+library("forcats")
+
+#result_table = read_csv("CSVs/Yellowtail_results_final.csv")
+result_table = read_csv("CSVs/Yellowtail_results_final_newsettings_forpaper.csv")
+
+summary_tab_exclude <- result_table %>% 
+  subset(Strata == "reduced")%>%
+ # subset(Scenario=="Constant Population")%>%
+
+#   group_by(`Temperature Scenario`)%>%
+# # 
+# #   group_by(Scenario)%>%
+#   mutate(`Temperature Scenario` = factor(`Temperature Scenario`))%>%
+#   arrange(`Temperature Scenario`) %>%
+  select(Scenario,`Temperature Scenario`,Season,Covariate,Noise,`VAST A`,`VAST B`, `Stratified Mean`)
+
+#COLOR INDIVIDUAL ROWS BY CATEGORY
+#define colors
+
+Fig_5b <- kable(summary_tab_exclude[, -1],
+      align = c("l", "l", "l", "l", "c", "c", "c"),
+  caption = "Yellowtail flounder error results with certain strata excluded from calculations. Row colors correspond to the same settings applied in different seasons.") %>%
+  row_spec(0,bold = TRUE)%>% #bold column titles
+  kable_styling(font_size = 10)  %>%
+  pack_rows(tab_kable(), colnum = 1,
+    index = table(fct_inorder(summary_tab_exclude$Scenario), useNA = "no"))%>%
+  #color every 4th row a different color
+   row_spec(seq(1,96,4), bold = T, color = "white", background = "black")%>%
+  row_spec(seq(2,96,4), bold = T, color = "white", background = "#5C5E60")%>%
+  row_spec(seq(3,96,4), bold = T, color = "white", background = "#A4A4A4")%>%
+  row_spec(seq(4,96,4), bold = T, color = "white", background = "#D2D2D2");
+
+```
+
 ```{r, results='asis'}
-Fig_5
+Fig_5a
+```
+```{r, results='asis'}
+Fig_5b
 ```
@@
 
-param_table1 = read_csv("CSVs/Cod_results.csv")
+#param_table1 = read_csv("CSVs/Cod_results.csv")
+param_table1 = read_csv("CSVs/Cod_results_newsettings_forpaper.csv")
 
@@
 
+```{r HadERROR,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
+library("knitr")
+library("kableExtra")
+library("dplyr")
+library("readr")
+library("tidyr")
+library("forcats")
+
+#param_table1 = read_csv("CSVs/Cod_results.csv")
+param_table1 = read_csv("CSVs/Haddock_results_newsettings_forpaper.csv")
+
+summary_tab1 <- param_table1 %>% 
+# Group the table according to Parameter, Scenario
+  group_by(Temp)%>%
+  mutate(Temp = factor(Temp)) %>%
+  arrange(Temp) %>%
+  select(Temp, everything())
+
+# Create table to present the data]
+Fig_3b <-  kable(summary_tab1[, -1], 
+  caption = "Haddock error results.",
+    format = "latex", booktabs = TRUE) %>%
+  kable_styling(font_size = 10) %>%
+  pack_rows(tab_kable(), colnum = 1,
+    index = table(fct_inorder(summary_tab1$Temp), useNA = "no"))
+```
+
+```{r, results='asis'}
+Fig_3b
+```
+
 <!-- ```{r PMsallscenarios, echo=FALSE, message=FALSE, warnings=FALSE, out.width='7in', fig.cap=figcap1} -->

### 2023-02-21 Manuscript/Draft1.tex
manuscript update and percent area occ. update

@@
 The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf.
-The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits \textbf{just listed a few uses of survey.
-change/add others?}.
+The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits
 
@@
 Since most spatial analyses and projections of future distributions typically assume a constant survey catchability and/or availability over time, NOAA's survey design includes sampling during approximately the same 2-3 week time period in each season.
+{[}{[}the survey for the entire region takes \textasciitilde8 weeks to complete each season, the GB strata get covered in a similar time period each season because the survey starts in the south and moves north.
+But there are times when logistics, e.g., storms, Covid, cause a change in the timing of any particular stratum being sampled.
+This timing issue has come up in the past and we've looked at it using the dates of sampled stations within a particular stock definition.
+See for example GBYT\_survey\_timing\_cdf.png taken from the 2018 assessment of Georges Bank yellowtail flounder that I copied into a newly created ``miscellaneous'' folder on the Google Drive.
+Given all this, I suggest modifying the wording to something like ``NOAA's survey design includes sampling Georges Bank during approximately the same 3-4 week time period in each season''{]}{]}
 
@@
 The stratified mean calculates the geometric mean catch per tow and does not account for spatial variation in samples or allow the inclusion of environmental covariates.
-Model-based abundance estimates are often derived from model framewroks such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM) that can incorporate both random and mixed effects.
+Model-based abundance estimates are often derived from model framewroks {[}{[}frameworks{]}{]} such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM) that can incorporate both random and mixed effects.
 The Vector Autoregressive Spatio-Temporal (VAST) model is a spatial delta-generalized linear mixed model (Thorson 2019).
@@
 We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses.
-The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and take on the order of hours to complete.
+The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and {[}{[}can?{]}{]} take on the order of hours to complete.
 We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
@@
 \textbf{BL: Help with correct notation for adding noise}.
-We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges bank to those that only include a subset of the full spatial domain for each species.
+We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges bank {[}{[}Bank{]}{]} to those that only include a subset of the full spatial domain for each species.
 We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from.
-Tabl XXX lists all strata inhabited by each species, those that are removed from certain calculations, and the explanation of why these strata were removed.
+Tabl XXX {[}{[}needs to be added{]}{]} lists all strata inhabited by each species, those that are removed from certain calculations, and the explanation of why these strata were removed.
 
@@
 
-The goal of our project was to analyze how well contemporary population indexing methods can track population trends under a host of conditions, as depicted in Table COMBOS.
-Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year {[}{[}can cite the 2022 management track assessments, see \url{https://apps-nefsc.fisheries.noaa.gov/saw/sari.php} for when the document becomes available{]}{]} \textbf{cite}.
+The goal of our project was to analyze how well contemporary population indexing methods can track population trends under a variety of conditions, as depicted in Table \ref{tab:scenarios} Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year {[}{[}can cite the 2022 management track assessments, see \url{https://apps-nefsc.fisheries.noaa.gov/saw/sari.php} for when the document becomes available{]}{]} \textbf{cite}.
 For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod (see Table ???) and increasing population scenarios for Haddock (see Table ???).
@@
 General themes that exist in Tables X, Y and Z are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors.
-We also see individual cases where the stratified mean produced the lowest absolut error and instances where including covarites in VAST models actually increase the absolute error.
-When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error {[}{[}as expected{]}{]}, though there are some scenarios where the impact is minimal.
+We also see individual cases where the stratified mean produced the lowest absolute error and instances where including covarites in VAST models actually increase the absolute error.
+When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.
 
@@
 
-In considering the Yellowtail Flounder results in Table XXX, we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
+In considering the Yellowtail Flounder results in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}, we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
 However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in the Table.
 These errors are corrected by including covariate information allowing for an improved VAST estimate that are significantly lower than their stratified mean counterparts.
+
+Notes on differences between VAST settings: For YTF, 29/96 \textasciitilde{} 30\% of VAST runs with new settings were better (70\% were worse).
+80/96 \textasciitilde{} 83\% of scenarios had a VAST run with a better error than the stratified mean.
+15/16 of the times the stratified mean was better VAST was not using covariates.
+The 1 time that VAST was using covariates and was still worse than the stratified mean was IncPop\_IncTemp\_Allstrata\_WCov\_WNoise in the Fall (VAST had strong overestimate in the fall during IncPOP\_IncTemP\_allstrata for Had and YT).
+The overestimate in the fall only seems to be related to the way the population shifts between season.
+Looking at the percent shift plots, the population is shifting out of strata 16 in both seasons (large east strata).
+In the spring the population is shifting mostly into strata 17 (thin strata adjacent to 16), but in the fall they are shifting into both 17 and 18 (18 thin one adjacent to 17).
+Thus it seems like with much of the population concentrated in the really small outer most strata, VAST produces an underestimate, even with covariates
+
+For Haddock (IncPop), 18/32 \textasciitilde{} 57\% of VAST runs better with new settings.
+24/32 = 75\% of scenarios had a VAST run with a better error than the stratified mean.
+14/16 VAST were better with covariates (the two that were worse were essentially the same) while only 10/16 were better without covariates.
+Stratified mean tended to perform better when all strata were included in the calculation, while VAST tended to perform better when strata are removed.
+
+For Cod (DecPop), only 11/32 \textasciitilde{} 34\% of VAST runs were better with new settings.
+28/32\textasciitilde{} 88\% of scenarios had a VAST run with a better estimate than the stratified mean.
+All 16/16 = 100\% of VAST runs with a covariate were better than the stratified mean.
+12/16 VAST were better without covariates.
+The 4 runs without covariates that were worse were all extremely large errors and all from increasing temp in the fall (with and without all strata and noise).
+In the fall the population is moving out of strata 16 (large eastern strata) and 21 (north large) and into a number of strata (22 (tiny north), 24 (large north), 18 (tiny north)).
+Might be having trouble tracking in to smaller strata again.
 
@@
 
-\begin{table}
+\section{Discussion}
 
-\caption{\label{tab:unnamed-chunk-1}Yellowtail error results}
-\centering
-\begin{tabular}[t]{l|l|l|l|l|l|l|l|l|l|l}
-\hline
-X & X.1 & X.2 & X.3 & X.4 & Constant.Population & X.5 & Increasing.Population & X.6 & Decreasing.Population & X.7\\
-\hline
-Temp & Covariate & Strata & Noise & season & Stratified Mean & VAST Estimate & Stratified Mean & VAST Estimate & Stratified Mean & VAST Estimate\\
-\hline
-const & no cov & all & no & spring & 0.21 & 0.11 & 0.16 & 0.13 & 0.23 & 0.08\\
-\hline
-const & no cov & all & yes & spring & 0.25 & 0.16 & 0.22 & 0.21 & 0.27 & 0.11\\
-\hline
-const & w/ cov & all & no & spring & 0.21 & 0.07 & 0.16 & 0.06 & 0.23 & 0.06\\
-\hline
-const & w/ cov & all & yes & spring & 0.25 & 0.08 & 0.22 & 0.07 & 0.27 & 0.07\\
-\hline
-const & no cov & all & no & fall & 0.32 & 0.68 & 0.34 & 0.36 & 0.41 & 0.81\\
-\hline
-const & no cov & all & yes & fall & 0.31 & 0.77 & 0.46 & 0.44 & 0.44 & 1.09\\
-\hline
-const & w/ cov & all & no & fall & 0.32 & 0.08 & 0.34 & 0.08 & 0.41 & 0.08\\
-\hline
-const & w/ cov & all & yes & fall & 0.31 & 0.11 & 0.46 & 0.17 & 0.44 & 0.18\\
-\hline
-const & no cov & reduced & no & spring & 0.27 & 0.19 & 0.2 & 0.15 & 0.25 & 0.19\\
-\hline
-const & no cov & reduced & yes & spring & 0.26 & 0.15 & 0.22 & 0.11 & 0.3 & 0.16\\
-\hline
-const & w/ cov & reduced & no & spring & 0.27 & 0.19 & 0.2 & 0.17 & 0.25 & 0.19\\
-\hline
-const & w/ cov & reduced & yes & spring & 0.26 & 0.14 & 0.22 & 0.13 & 0.3 & 0.16\\
-\hline
-const & no cov & reduced & no & fall & 0.47 & 0.25 & 0.41 & 0.11 & 0.55 & 0.24\\
-\hline
-const & no cov & reduced & yes & fall & 0.49 & 0.36 & 0.46 & 0.21 & 0.53 & 0.36\\
-\hline
-const & w/ cov & reduced & no & fall & 0.47 & 0.19 & 0.41 & 0.22 & 0.55 & 0.23\\
-\hline
-const & w/ cov & reduced & yes & fall & 0.49 & 0.17 & 0.46 & 0.19 & 0.53 & 0.24\\
-\hline
-increasing & no cov & all & no & spring & 0.28 & 0.11 & 0.32 & 0.13 & 0.22 & 0.15\\
-\hline
-increasing & no cov & all & yes & spring & 0.28 & 0.15 & 0.34 & 0.16 & 0.26 & 0.17\\
-\hline
-increasing & w/ cov & all & no & spring & 0.28 & 0.06 & 0.32 & 0.07 & 0.22 & 0.07\\
-\hline
-increasing & w/ cov & all & yes & spring & 0.28 & 0.12 & 0.34 & 0.1 & 0.26 & 0.1\\
-\hline
-increasing & no cov & all & no & fall & 0.51 & 1.26 & 0.3 & 0.66 & 0.28 & 1.06\\
-\hline
-increasing & no cov & all & yes & fall & 0.5 & 1.38 & 0.39 & 0.71 & 0.25 & 1.1\\
-\hline
-increasing & w/ cov & all & no & fall & 0.51 & 0.23 & 0.3 & 0.21 & 0.28 & 0.15\\
-\hline
-increasing & w/ cov & all & yes & fall & 0.5 & 0.28 & 0.3 & 0.37 & 0.25 & 0.2\\
-\hline
-increasing & no cov & reduced & no & spring & 0.31 & 0.17 & 0.4 & 0.33 & 0.32 & 0.14\\
-\hline
-increasing & no cov & reduced & yes & spring & 0.29 & 0.19 & 0.41 & 0.26 & 0.29 & 0.15\\
-\hline
-increasing & w/ cov & reduced & no & spring & 0.31 & 0.17 & 0.4 & 0.3 & 0.32 & 0.22\\
-\hline
-increasing & w/ cov & reduced & yes & spring & 0.29 & 0.15 & 0.41 & 0.25 & 0.29 & 0.21\\
-\hline
-increasing & no cov & reduced & no & fall & 0.64 & 0.75 & 0.7 & 0.49 & 0.54 & 0.6\\
-\hline
-increasing & no cov & reduced & yes & fall & 0.66 & 0.89 & 0.69 & 0.48 & 0.53 & 0.62\\
-\hline
-increasing & w/ cov & reduced & no & fall & 0.64 & 0.2 & 0.7 & 0.53 & 0.54 & 0.31\\
-\hline
-increasing & w/ cov & reduced & yes & fall & 0.66 & 0.13 & 0.69 & 0.5 & 0.53 & 0.32\\
-\hline
-\end{tabular}
-\end{table}
-
-\begin{table}
-
-\caption{\label{tab:unnamed-chunk-1}Cod error results}
-\centering
-\begin{tabular}[t]{l|l|l|l|r|r|r}
-\hline
-Temp & Strata & Noise & season & VAST.No.Cov & VAST.w..Cov & Stratified.Mean\\
-\hline
-constant & all & no & spring & 0.11 & 0.12 & 0.36\\
-\hline
-constant & all & yes & spring & 0.12 & 0.15 & 0.35\\
-\hline
-constant & all & no & fall & 0.19 & 0.05 & 0.49\\
-\hline
-constant & all & yes & fall & 0.30 & 0.23 & 0.41\\
-\hline
-constant & reduced & no & spring & 0.17 & 0.24 & 0.41\\
-\hline
-constant & reduced & yes & spring & 0.20 & 0.23 & 0.46\\
-\hline
-constant & reduced & no & fall & 0.21 & 0.33 & 0.60\\
-\hline
-constant & reduced & yes & fall & 0.18 & 0.31 & 0.58\\
-\hline
-increasing & all & no & spring & 0.12 & 0.15 & 0.25\\
-\hline
-increasing & all & yes & spring & 0.19 & 0.19 & 0.27\\
-\hline
-increasing & all & no & fall & 0.76 & 0.13 & 0.45\\
-\hline
-increasing & all & yes & fall & 0.89 & 0.33 & 0.44\\
-\hline
-increasing & reduced & no & spring & 0.14 & 0.22 & 0.32\\
-\hline
-increasing & reduced & yes & spring & 0.15 & 0.21 & 0.29\\
-\hline
-increasing & reduced & no & fall & 0.60 & 0.31 & 0.54\\
-\hline
-increasing & reduced & yes & fall & 0.62 & 0.32 & 0.53\\
-\hline
-\end{tabular}
-\end{table}
-
-\begin{table}
-
-\caption{\label{tab:unnamed-chunk-1}Haddock error results}
-\centering
-\begin{tabular}[t]{l|l|l|l|r|r|r|l|l}
-\hline
-Temp & Strata & Noise & season & VAST.No.Cov & VAST.w..Cov & Stratified.Mean & X & X.1\\
-\hline
-const & all & no & spring & 0.49 & 0.18 & 0.18 &  & \\
-\hline
-const & all & yes & spring & 0.73 & 0.43 & 0.21 &  & Haddock\\
-\hline
-const & all & no & fall & 0.28 & 0.05 & 0.26 &  & Increasing Population\\
-\hline
-const & all & yes & fall & 0.41 & 0.06 & 0.27 &  & \\
-\hline
-const & reduced & no & spring & 0.34 & 0.35 & 0.45 &  & \\
-\hline
-const & reduced & yes & spring & 0.30 & 0.33 & 0.46 &  & \\
-\hline
-const & reduced & no & fall & 0.36 & 0.48 & 0.54 &  & \\
-\hline
-const & reduced & yes & fall & 0.33 & 0.46 & 0.52 &  & \\
-\hline
-increasing & all & no & spring & 0.25 & 0.05 & 0.26 &  & \\
-\hline
-increasing & all & yes & spring & 0.30 & 0.06 & 0.31 &  & \\
-\hline
-increasing & all & no & fall & 0.89 & 0.23 & 0.40 &  & \\
-\hline
-increasing & all & yes & fall & 1.04 & 0.35 & 0.42 &  & \\
-\hline
-increasing & reduced & no & spring & 0.32 & 0.40 & 0.44 &  & \\
-\hline
-increasing & reduced & yes & spring & 0.38 & 0.37 & 0.37 &  & \\
-\hline
-increasing & reduced & no & fall & 0.44 & 0.64 & 0.72 &  & \\
-\hline
-increasing & reduced & yes & fall & 0.42 & 0.62 & 0.70 &  & \\
-\hline
-\end{tabular}
-\end{table}
-
-\section{Discussion}
+Estimation methods that have large, inaccurate swings (stratified mean) can lead to changes in quotas that do not correspond to the true population trend, which could have a compounding effect (can lead to quotas that are too high/low given an incorrect assumption of increase/decrease in biomass).
+Our model has a constant assumed mortality that accounts for fishing and natural death and will not account for impacts of these decisions.
+VAST may provide more consistent biomass predictions(?)
 
@@
 \begin{table}
-\centering\begingroup\fontsize{10}{12}\selectfont
 
-\begin{tabular}{l|l|l|l|l|l|r|r}
+\caption{\label{tab:YTresultsallstrata}Yellowtail flounder error results with all strata included in calculations. Row colors correspond to the same settings applied in different seasons.}
+\centering
+\fontsize{10}{12}\selectfont
+\begin{tabular}[t]{l|l|l|l|c|c|c}
 \hline
-Temp & Covariate & Strata & Noise & Season & Scenario & Stratified Mean & VAST Estimate\\
-\hline
-\multicolumn{1}{l}{\textbf{all}}\\
+\textbf{Temperature Scenario} & \textbf{Season} & \textbf{Covariate} & \textbf{Noise} & \textbf{VAST A} & \textbf{VAST B} & \textbf{Stratified Mean}\\
 \hline
@@
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & all & no & spring & Constant Population & 0.21 & 0.11\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor{black}{\textcolor{white}{\textbf{spring}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.13}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.11}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.21}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & all & yes & spring & Constant Population & 0.25 & 0.16\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.14}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.16}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.25}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & no & spring & Constant Population & 0.21 & 0.07\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.07}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.07}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.21}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & yes & spring & Constant Population & 0.25 & 0.08\\
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.08}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.08}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.25}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & all & no & fall & Constant Population & 0.32 & 0.68\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor{black}{\textcolor{white}{\textbf{fall}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.63}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.68}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.32}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & all & yes & fall & Constant Population & 0.31 & 0.77\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.80}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.77}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.31}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & no & fall & Constant Population & 0.32 & 0.08\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.14}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.08}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.32}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & yes & fall & Constant Population & 0.31 & 0.11\\
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.11}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.31}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & no & spring & Constant Population & 0.27 & 0.19\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor{black}{\textcolor{white}{\textbf{spring}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.14}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.11}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.28}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & yes & spring & Constant Population & 0.26 & 0.15\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.18}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.15}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.28}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & no & spring & Constant Population & 0.27 & 0.19\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.05}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.06}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.28}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & yes & spring & Constant Population & 0.26 & 0.14\\
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.10}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.12}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.28}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & no & fall & Constant Population & 0.47 & 0.25\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor{black}{\textcolor{white}{\textbf{fall}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{1.46}}} & \cellcolor{black}{\textcolor{white}{\textbf{1.26}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.51}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & yes & fall & Constant Population & 0.49 & 0.36\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{1.40}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{1.38}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.50}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & no & fall & Constant Population & 0.47 & 0.19\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.21}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.23}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.51}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & yes & fall & Constant Population & 0.49 & 0.17\\
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.30}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.28}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.50}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & no & spring & Constant Population & 0.28 & 0.11\\
+\multicolumn{1}{l}{\textbf{Decreasing Population}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & yes & spring & Constant Population & 0.28 & 0.15\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor{black}{\textcolor{white}{\textbf{spring}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.11}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.08}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.23}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & no & spring & Constant Population & 0.28 & 0.06\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.12}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.11}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.27}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & yes & spring & Constant Population & 0.28 & 0.12\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.07}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.06}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.23}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & no & fall & Constant Population & 0.51 & 1.26\\
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.11}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.07}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.27}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & yes & fall & Constant Population & 0.50 & 1.38\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor{black}{\textcolor{white}{\textbf{fall}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.97}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.81}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.41}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & no & fall & Constant Population & 0.51 & 0.23\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.99}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{1.09}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.37}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & yes & fall & Constant Population & 0.50 & 0.28\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.16}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.08}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.41}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & no & spring & Constant Population & 0.31 & 0.17\\
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.29}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.18}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.37}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & yes & spring & Constant Population & 0.29 & 0.19\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor{black}{\textcolor{white}{\textbf{spring}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.17}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.15}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.22}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & no & spring & Constant Population & 0.31 & 0.17\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.15}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.26}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & yes & spring & Constant Population & 0.29 & 0.15\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.08}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.07}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.22}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & no & fall & Constant Population & 0.64 & 0.75\\
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.16}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.10}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.26}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & yes & fall & Constant Population & 0.66 & 0.89\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor{black}{\textcolor{white}{\textbf{fall}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{1.17}}} & \cellcolor{black}{\textcolor{white}{\textbf{1.06}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.28}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & no & fall & Constant Population & 0.64 & 0.20\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{1.14}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{1.10}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.25}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & yes & fall & Constant Population & 0.66 & 0.13\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.40}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.15}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.28}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.40}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.20}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.25}}}\\
 \hline
@@
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & all & no & spring & Increasing Population & 0.16 & 0.13\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor{black}{\textcolor{white}{\textbf{spring}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.46}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.13}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.16}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & all & yes & spring & Increasing Population & 0.22 & 0.21\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.43}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.21}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.22}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & no & spring & Increasing Population & 0.16 & 0.06\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.06}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.06}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.16}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & yes & spring & Increasing Population & 0.22 & 0.07\\
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.08}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.07}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.22}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & all & no & fall & Increasing Population & 0.34 & 0.36\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor{black}{\textcolor{white}{\textbf{fall}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.40}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.36}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.34}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & all & yes & fall & Increasing Population & 0.46 & 0.44\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.38}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.44}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.46}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & no & fall & Increasing Population & 0.34 & 0.08\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.11}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.08}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.34}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & yes & fall & Increasing Population & 0.46 & 0.17\\
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.24}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.46}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & no & spring & Increasing Population & 0.20 & 0.15\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor{black}{\textcolor{white}{\textbf{spring}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.16}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.13}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.32}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & yes & spring & Increasing Population & 0.22 & 0.11\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.21}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.16}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.32}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & no & spring & Increasing Population & 0.20 & 0.17\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.06}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.07}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.32}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & yes & spring & Increasing Population & 0.22 & 0.13\\
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.12}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.10}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.32}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & no & fall & Increasing Population & 0.41 & 0.11\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor{black}{\textcolor{white}{\textbf{fall}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.71}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.66}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.30}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & yes & fall & Increasing Population & 0.46 & 0.21\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{1.03}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.71}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.39}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & no & fall & Increasing Population & 0.41 & 0.22\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.43}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.21}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.30}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & yes & fall & Increasing Population & 0.46 & 0.19\\
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.51}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.37}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.30}}}\\
 \hline
-\multicolumn{1}{l}{\textbf{reduced}}\\
+\end{tabular}
+\end{table}
+\begin{table}
+
+\caption{\label{tab:YTresultsreducedstrata}Yellowtail flounder error results with certain strata excluded from calculations. Row colors correspond to the same settings applied in different seasons.}
+\centering
+\fontsize{10}{12}\selectfont
+\begin{tabular}[t]{l|l|l|l|c|c|c}
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & no & spring & Increasing Population & 0.32 & 0.13\\
+\textbf{Temperature Scenario} & \textbf{Season} & \textbf{Covariate} & \textbf{Noise} & \textbf{VAST A} & \textbf{VAST B} & \textbf{Stratified Mean}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & yes & spring & Increasing Population & 0.34 & 0.16\\
+\multicolumn{1}{l}{\textbf{Constant Population}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & no & spring & Increasing Population & 0.32 & 0.07\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor{black}{\textcolor{white}{\textbf{spring}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.24}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.19}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.27}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & yes & spring & Increasing Population & 0.34 & 0.10\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.16}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.15}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.22}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & no & fall & Increasing Population & 0.30 & 0.66\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.19}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.19}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.27}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & yes & fall & Increasing Population & 0.39 & 0.71\\
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.12}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.14}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.22}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & no & fall & Increasing Population & 0.30 & 0.21\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor{black}{\textcolor{white}{\textbf{fall}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.30}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.25}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.47}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & yes & fall & Increasing Population & 0.30 & 0.37\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.78}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.36}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.44}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & no & spring & Increasing Population & 0.40 & 0.33\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.22}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.19}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.47}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & yes & spring & Increasing Population & 0.41 & 0.26\\
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.24}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.44}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & no & spring & Increasing Population & 0.40 & 0.30\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor{black}{\textcolor{white}{\textbf{spring}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.23}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.17}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.31}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & yes & spring & Increasing Population & 0.41 & 0.25\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.25}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.19}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.29}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & no & fall & Increasing Population & 0.70 & 0.49\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.31}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & yes & fall & Increasing Population & 0.69 & 0.48\\
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.62}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.15}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.29}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & no & fall & Increasing Population & 0.70 & 0.53\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor{black}{\textcolor{white}{\textbf{fall}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{2.22}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.75}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.64}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & yes & fall & Increasing Population & 0.69 & 0.50\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{1.75}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.89}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.59}}}\\
+\hline
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.24}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.20}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.64}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.59}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.13}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.59}}}\\
 \hline
@@
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & all & no & spring & Decreasing Population & 0.23 & 0.08\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor{black}{\textcolor{white}{\textbf{spring}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.31}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.19}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.25}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & all & yes & spring & Decreasing Population & 0.27 & 0.11\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.27}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.16}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.27}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & no & spring & Decreasing Population & 0.23 & 0.06\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.19}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.19}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.25}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & yes & spring & Decreasing Population & 0.27 & 0.07\\
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.15}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.16}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.27}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & all & no & fall & Decreasing Population & 0.41 & 0.81\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor{black}{\textcolor{white}{\textbf{fall}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.53}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.24}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.55}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & all & yes & fall & Decreasing Population & 0.44 & 1.09\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.53}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.36}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.53}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & no & fall & Decreasing Population & 0.41 & 0.08\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.18}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.23}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.55}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & all & yes & fall & Decreasing Population & 0.44 & 0.18\\
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.16}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.24}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.53}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & no & spring & Decreasing Population & 0.25 & 0.19\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor{black}{\textcolor{white}{\textbf{spring}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.18}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.14}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.32}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & yes & spring & Decreasing Population & 0.30 & 0.16\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.37}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.15}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.29}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & no & spring & Decreasing Population & 0.25 & 0.19\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.21}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.22}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.32}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & yes & spring & Decreasing Population & 0.30 & 0.16\\
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.16}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.21}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.29}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & no & fall & Decreasing Population & 0.55 & 0.24\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor{black}{\textcolor{white}{\textbf{fall}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.90}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.60}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.54}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & no cov & reduced & yes & fall & Decreasing Population & 0.53 & 0.36\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.84}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.62}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.48}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & no & fall & Decreasing Population & 0.55 & 0.23\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.32}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.31}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.54}}}\\
 \hline
-\hspace{1em}\hspace{1em}constant & w/ cov & reduced & yes & fall & Decreasing Population & 0.53 & 0.24\\
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.36}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.32}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.48}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & no & spring & Decreasing Population & 0.22 & 0.15\\
+\multicolumn{1}{l}{\textbf{Increasing Population}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & yes & spring & Decreasing Population & 0.26 & 0.17\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor{black}{\textcolor{white}{\textbf{spring}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.22}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.15}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.20}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & no & spring & Decreasing Population & 0.22 & 0.07\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.19}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.11}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.22}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & yes & spring & Decreasing Population & 0.26 & 0.10\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.20}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & no & fall & Decreasing Population & 0.28 & 1.06\\
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.11}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.13}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.22}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & all & yes & fall & Decreasing Population & 0.25 & 1.10\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor{black}{\textcolor{white}{\textbf{fall}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.19}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.11}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.41}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & no & fall & Decreasing Population & 0.28 & 0.15\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.26}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.21}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.46}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & all & yes & fall & Decreasing Population & 0.25 & 0.20\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.21}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.22}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.41}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & no & spring & Decreasing Population & 0.32 & 0.14\\
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.19}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.46}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & yes & spring & Decreasing Population & 0.29 & 0.15\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor{black}{\textcolor{white}{\textbf{spring}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.31}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.33}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.40}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & no & spring & Decreasing Population & 0.32 & 0.22\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.30}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.26}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.38}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & yes & spring & Decreasing Population & 0.29 & 0.21\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.30}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.30}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.40}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & no & fall & Decreasing Population & 0.54 & 0.60\\
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.31}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.25}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.38}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & no cov & reduced & yes & fall & Decreasing Population & 0.53 & 0.62\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor{black}{\textcolor{white}{\textbf{fall}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.56}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.49}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.70}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & no & fall & Decreasing Population & 0.54 & 0.31\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.58}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.48}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.69}}}\\
 \hline
-\hspace{1em}\hspace{1em}increasing & w/ cov & reduced & yes & fall & Decreasing Population & 0.53 & 0.32\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.48}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.53}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.70}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.47}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.50}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.69}}}\\
 \hline
 \end{tabular}
-\endgroup{}
 \end{table}
@@
 \fontsize{10}{12}\selectfont
-\begin{tabular}[t]{lllrrr}
+\begin{tabular}[t]{lllrrrrr}
 \toprule
-Strata & Noise & season & VAST No Cov & VAST w/ Cov & Stratified Mean\\
+Strata & Noise & Season & VAST NC  A & VAST NC  B & VAST WC A & VAST WC B & Stratified Mean\\
 \midrule
 \addlinespace[0.3em]
-\multicolumn{1}{l}{\textbf{constant}}\\
-\hspace{1em}all & no & spring & 0.11 & 0.12 & 0.36\\
-\hspace{1em}all & yes & spring & 0.12 & 0.15 & 0.35\\
-\hspace{1em}all & no & fall & 0.19 & 0.05 & 0.49\\
-\hspace{1em}all & yes & fall & 0.30 & 0.23 & 0.41\\
-\hspace{1em}reduced & no & spring & 0.17 & 0.24 & 0.41\\
-\hspace{1em}reduced & yes & spring & 0.20 & 0.23 & 0.46\\
-\hspace{1em}reduced & no & fall & 0.21 & 0.33 & 0.60\\
-\hspace{1em}reduced & yes & fall & 0.18 & 0.31 & 0.58\\
+\multicolumn{1}{l}{\textbf{Constant Temp.}}\\
+\hspace{1em}all & no & spring & 0.11 & 0.11 & 0.13 & 0.12 & 0.36\\
+\hspace{1em}all & yes & spring & 0.14 & 0.12 & 0.09 & 0.15 & 0.35\\
+\hspace{1em}all & no & fall & 0.23 & 0.19 & 0.09 & 0.05 & 0.49\\
+\hspace{1em}all & yes & fall & 0.34 & 0.30 & 0.20 & 0.23 & 0.41\\
+\hspace{1em}reduced & no & spring & 0.25 & 0.17 & 0.22 & 0.24 & 0.41\\
+\hspace{1em}reduced & yes & spring & 0.25 & 0.20 & 0.14 & 0.23 & 0.46\\
+\hspace{1em}reduced & no & fall & 0.16 & 0.21 & 0.26 & 0.33 & 0.60\\
+\hspace{1em}reduced & yes & fall & 0.16 & 0.18 & 0.26 & 0.31 & 0.58\\
 \addlinespace[0.3em]
-\multicolumn{1}{l}{\textbf{increasing}}\\
-\hspace{1em}all & no & spring & 0.12 & 0.15 & 0.25\\
-\hspace{1em}all & yes & spring & 0.19 & 0.19 & 0.27\\
-\hspace{1em}all & no & fall & 0.76 & 0.13 & 0.45\\
-\hspace{1em}all & yes & fall & 0.89 & 0.33 & 0.44\\
-\hspace{1em}reduced & no & spring & 0.14 & 0.22 & 0.32\\
-\hspace{1em}reduced & yes & spring & 0.15 & 0.21 & 0.29\\
-\hspace{1em}reduced & no & fall & 0.60 & 0.31 & 0.54\\
-\hspace{1em}reduced & yes & fall & 0.62 & 0.32 & 0.53\\
+\multicolumn{1}{l}{\textbf{Increasing Temp.}}\\
+\hspace{1em}all & no & spring & 0.12 & 0.12 & 0.16 & 0.15 & 0.25\\
+\hspace{1em}all & yes & spring & 0.16 & 0.19 & 0.23 & 0.19 & 0.27\\
+\hspace{1em}all & no & fall & 0.86 & 0.76 & 0.47 & 0.13 & 0.45\\
+\hspace{1em}all & yes & fall & 1.13 & 0.89 & 0.55 & 0.33 & 0.44\\
+\hspace{1em}reduced & no & spring & 0.29 & 0.26 & 0.22 & 0.21 & 0.34\\
+\hspace{1em}reduced & yes & spring & 0.32 & 0.19 & 0.19 & 0.11 & 0.33\\
+\hspace{1em}reduced & no & fall & 1.41 & 0.79 & 0.37 & 0.26 & 0.62\\
+\hspace{1em}reduced & yes & fall & 2.09 & 1.37 & 0.40 & 0.26 & 0.57\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}
+
+\caption{\label{tab:HadERROR}Haddock error results.}
+\centering
+\fontsize{10}{12}\selectfont
+\begin{tabular}[t]{lllrrrrr}
+\toprule
+Strata & Noise & Season & VAST NC  A & VAST NC  B & VAST WC A & VAST WC B & Stratified Mean\\
+\midrule
+\addlinespace[0.3em]
+\multicolumn{1}{l}{\textbf{Constant Temp.}}\\
+\hspace{1em}all & no & spring & 0.45 & 0.49 & 0.13 & 0.18 & 0.18\\
+\hspace{1em}all & yes & spring & 0.55 & 0.73 & 0.18 & 0.43 & 0.14\\
+\hspace{1em}all & no & fall & 0.31 & 0.28 & 0.05 & 0.05 & 0.26\\
+\hspace{1em}all & yes & fall & 0.45 & 0.41 & 0.15 & 0.06 & 0.27\\
+\hspace{1em}reduced & no & spring & 0.34 & 0.34 & 0.30 & 0.35 & 0.45\\
+\hspace{1em}reduced & yes & spring & 0.31 & 0.30 & 0.45 & 0.33 & 0.44\\
+\hspace{1em}reduced & no & fall & 0.34 & 0.36 & 0.46 & 0.48 & 0.54\\
+\hspace{1em}reduced & yes & fall & 0.29 & 0.33 & 0.41 & 0.46 & 0.50\\
+\addlinespace[0.3em]
+\multicolumn{1}{l}{\textbf{Increasing Temp.}}\\
+\hspace{1em}all & no & spring & 0.28 & 0.25 & 0.11 & 0.05 & 0.26\\
+\hspace{1em}all & yes & spring & 0.35 & 0.30 & 0.11 & 0.06 & 0.31\\
+\hspace{1em}all & no & fall & 0.82 & 0.89 & 0.23 & 0.23 & 0.40\\
+\hspace{1em}all & yes & fall & 1.01 & 1.04 & 0.29 & 0.35 & 0.39\\
+\hspace{1em}reduced & no & spring & 0.35 & 0.32 & 0.41 & 0.40 & 0.44\\
+\hspace{1em}reduced & yes & spring & 0.33 & 0.38 & 0.39 & 0.37 & 0.36\\
+\hspace{1em}reduced & no & fall & 0.48 & 0.44 & 0.61 & 0.64 & 0.72\\
+\hspace{1em}reduced & yes & fall & 0.49 & 0.42 & 0.60 & 0.62 & 0.70\\
 \bottomrule

### 2023-02-21 TestScripts/percent_area_occupied.R

@@ -29,9 +29,11 @@ strata_species[["Had"]] <- c(13,14,15,16,17,18,19,20,21,22,23,24,25,29,30)
 #number of extra years in simulation. subtract this value from year in log matrix
 years_cut <- 2
 
+scenario <- "ConPop_IncTemp"
+
 #FIRST READ IN RES WHICH CONTAINS SPATIAL OUTPUT FOR GIVEN SCENARIO
 
-res <- readRDS("E:\\READ-PDB-blevy2-MFS2\\GB_Results\\ConPop_ConTemp\\res_ConPop_ConTemp.RDS")
+res <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Simulation_Results\\",scenario,"\\res_",scenario,".RDS",sep=""))
 
 #read in habitat
 hab <- readRDS(file="hab_GB_3species.RDS") #courser resolution
@@ -181,7 +183,7 @@ for(s in short_names){
 
 
 
-
+pdf(file=paste0('testfolder/',scenario,'_Spatial_pop_plots','.pdf'))
 
 #################################################################################
 # PLOTS
@@ -223,7 +225,7 @@ for(s in short_names){
 
 
 for(s in short_names){
-  par( mfrow = c(1,2), mar = c(1, 1, 1, 1))
+  par( mfrow = c(2,1), mar = c(1, 1, 1, 1))
   for(wk in plot_wks){
   
     #WITHOUT USING BREAKS
@@ -261,7 +263,7 @@ for(s in short_names){
 
 
 
-pdf(file=paste0('testfolder/Spatial_pop_plots','.pdf'))
+
 
 for(s in short_names){
 
@@ -277,7 +279,7 @@ for(s in short_names){
     text(0.5, 0.98, labels = paste( s, 'Week', wk, "Year", yr), cex = 1)
     
     }
-        par( mfrow = c(1,2), mar = c(1,1,1,1))
+        par( mfrow = c(2,1), mar = c(1,1,1,1))
     #plot first and last years for easy comparison
     fields::image.plot(rotate(spatial_pop[[s]][[wk]][[1]]), breaks = Qbreaks_list[[s]][[wk]]$brks, nlevel = (length( Qbreaks_list[[s]][[wk]]$brks)-1) )
     text(0.5, 0.98, labels = paste( s, 'Week', wk, "Year", 1), cex = 1)
[…]

### 2023-03-01 Manuscript/Draft1.Rmd
update manuscript

@@
 The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf.
-The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits
-
+The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits.
 The survey takes place twice each year- once in the spring and again in the fall.
-Since most spatial analyses and  projections of future distributions typically assume a constant survey catchability and/or availability over time, NOAA's survey design includes sampling during approximately the same 2-3 week time period in each season.
-[[the survey for the entire region takes ~8 weeks to complete each season, the GB strata get covered in a similar time period each season because the survey starts in the south and moves north.
-But there are times when logistics, e.g., storms, Covid, cause a change in the timing of any particular stratum being sampled.
-This timing issue has come up in the past and we've looked at it using the dates of sampled stations within a particular stock definition.
-See for example GBYT_survey_timing_cdf.png taken from the 2018 assessment of Georges Bank yellowtail flounder that I copied into a newly created "miscellaneous" folder on the Google Drive.
-Given all this, I suggest modifying the wording to something like "NOAA's survey design includes sampling Georges Bank during approximately the same 3-4 week time period in each season"]]
+Since most spatial analyses and  projections of future distributions typically assume a constant survey catchability and/or availability over time, NOAA's survey design includes sampling Georges Bank during approximately the same 3-4 week time period in each season.
 

### 2023-03-01 Manuscript/Draft1.Rmd
manuscript updates

@@
 
--- Used MixFishSim. Describe edits made to package
+-- Used MixFishSim.
 
@@
 We formulate the following inputs for the MFS package to address our research question.
-[[perhaps for discussion, but we should mention somewhere that MFS only tracks biomass not age structure, the latter is important for stock assessments but not considered in this paper]]
 
@@
 Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits.
-[[not sure we need to highlight the recruitment model, might be able to just use the table because Bev-Holt formulation is well known in fisheries]]**BL: MFS has 2 recruitment options and I edited the package to use previous years biomass** We chose to represent recruitment in the model using a Beverton-Holt formulation **cite**.
+We chose to represent recruitment in the model using a Beverton-Holt formulation **cite**.
 Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.
@@
 Weekly estimated temperature data for the region for 2012 was obtained from FVCOM [@chen2006unstructured].
-The 2012 data was repeated yearly in the constant temperature scenario because displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values.
 We chose to repeat estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data.
-The 2012 temperature data was also transform to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation.
+The 2012 data was chosen because it displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values, allowing for a smooth repeating yearly temperature pattern for the constant temperature scenario.
+The 2012 temperature data was also transformed to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation.
 We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation.
@@
 Thus, stratified random samples in scenarios with a repeating temperature pattern will have constant survey catchability and availability over time, which may not be true for increasing temperature scenarios due to evolving spatial preferences.
-[[**BL: included in above paragraph** may want to mention here or in the discussion that we used repeating 2012 to simplify the simulations and reduce the number of factors explored that would have occurred with different annual cycles]]
 
@@
 We consider 20 year simulations under three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time.
-Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius [[**BL: included in above paragraph** need some justification for this large change, wanted to be extreme because if the extreme doesn't make a difference, then more reasonable increases wouldn't make a difference]] over the duration of the 20 year simulation.
+Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius over the duration of the 20 year simulation.
 We therefore simulate 6 scenarios for each population.
@@
 After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year.
-We sample each strata in the same weeks in which the Spring and Fall surveys take place and the number of the samples taken reflect true target values for each strata.
+We sample each strata in the same weeks in which the Spring and Fall surveys take place, and the number of the samples taken reflect true target values for each strata.
 Most strata contain enough cells that we were able to generate a list of sampling locations without replacement, allowing us to sample a unique random cell in each survey over the duration of the simulation.
@@
 Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.
-**BL: added both to above paragraph**
-[[should we mention here or later that we did both with and without measurement error analyses?]]
-[[need to make sure that we don't lose sight of random sampling with the following sentences]]
 
 --Stratified mean vs VAST with and without covariates
-[[could use an intro sentence here to distinguish design-based (stratified mean) and model-based (VAST) estimators]] 
-[[may want to highlight the potential of VAST with covariates to address the climate change issue here as the reason we are doing these simulations]]
 
-Stock assessment scientists use samples from the bottom trawl survey to estimate the abundance for each stock stock.
-There are a number of approaches to obtaining abundance estimates ranging from traditional design-based estimates such as the stratified mean to model-based estimates that range in complexity.
-The stratified mean calculates the geometric mean catch per tow and does not account for spatial variation in samples or allow the inclusion of environmental covariates.
-Model-based abundance estimates are often derived from model framewroks [[frameworks]] such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM) that can incorporate both random and mixed effects.
-The Vector Autoregressive Spatio-Temporal (VAST) model is a spatial delta-generalized linear mixed model [@thorson2019guidance].
-VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurrence (presence/absence).
+Stock assessment scientists use samples from the bottom trawl survey to estimate the abundance for each fish stock.
+There are a number of approaches to obtain abundance estimates including traditional design-based estimates to model-based estimates that range in complexity.
+Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest.
+These methods do not account for spatial variation in samples or allow the inclusion of environmental influences.
+Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors).
+Model-based estimators, such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM), help account for complex relationships between variables and can help overcome problems with sampling design.
+
+We compare abundance estimates from stratified mean calculations to those derived from the Vector Autoregressive Spatio-Temporal (VAST) model.
+The stratified mean is a design-based approach that calculates the geometric mean catch per tow and has traditionally been used with stratified random sample designs.
+VAST is a spatial delta-generalized linear mixed model that estimates both abundance (biomass) and probability of occurrence (presence/absence) [@thorson2019guidance].
 If desired, VAST also allows users to include covariate data to better inform the model.
@@
 We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses.
-The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and [[can?]] take on the order of hours to complete.
+The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and can take on the order of hours to complete.
 We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.

### 2023-03-01 Manuscript/Draft1.tex
manuscript updates

@@
 The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf.
-The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits
-
+The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits.
 The survey takes place twice each year- once in the spring and again in the fall.
-Since most spatial analyses and projections of future distributions typically assume a constant survey catchability and/or availability over time, NOAA's survey design includes sampling during approximately the same 2-3 week time period in each season.
-{[}{[}the survey for the entire region takes \textasciitilde8 weeks to complete each season, the GB strata get covered in a similar time period each season because the survey starts in the south and moves north.
-But there are times when logistics, e.g., storms, Covid, cause a change in the timing of any particular stratum being sampled.
-This timing issue has come up in the past and we've looked at it using the dates of sampled stations within a particular stock definition.
-See for example GBYT\_survey\_timing\_cdf.png taken from the 2018 assessment of Georges Bank yellowtail flounder that I copied into a newly created ``miscellaneous'' folder on the Google Drive.
-Given all this, I suggest modifying the wording to something like ``NOAA's survey design includes sampling Georges Bank during approximately the same 3-4 week time period in each season''{]}{]}
+Since most spatial analyses and projections of future distributions typically assume a constant survey catchability and/or availability over time, NOAA's survey design includes sampling Georges Bank during approximately the same 3-4 week time period in each season.
 
@@
 We formulate the following inputs for the MFS package to address our research question.
-{[}{[}perhaps for discussion, but we should mention somewhere that MFS only tracks biomass not age structure, the latter is important for stock assessments but not considered in this paper{]}{]}
 
@@
 Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits.
-{[}{[}not sure we need to highlight the recruitment model, might be able to just use the table because Bev-Holt formulation is well known in fisheries{]}{]}\textbf{BL: MFS has 2 recruitment options and I edited the package to use previous years biomass} We chose to represent recruitment in the model using a Beverton-Holt formulation \textbf{cite}.
+We chose to represent recruitment in the model using a Beverton-Holt formulation \textbf{cite}.
 Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.
@@
 2006).
-The 2012 data was repeated yearly in the constant temperature scenario because displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values.
 We chose to repeat estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data.
-The 2012 temperature data was also transform to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation.
+The 2012 data was chosen because it displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values, allowing for a smooth repeating yearly temperature pattern for the constant temperature scenario.
+The 2012 temperature data was also transformed to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation.
 We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation.
@@
 Thus, stratified random samples in scenarios with a repeating temperature pattern will have constant survey catchability and availability over time, which may not be true for increasing temperature scenarios due to evolving spatial preferences.
-{[}{[}\textbf{BL: included in above paragraph} may want to mention here or in the discussion that we used repeating 2012 to simplify the simulations and reduce the number of factors explored that would have occurred with different annual cycles{]}{]}
 
@@
 We consider 20 year simulations under three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time.
-Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius {[}{[}\textbf{BL: included in above paragraph} need some justification for this large change, wanted to be extreme because if the extreme doesn't make a difference, then more reasonable increases wouldn't make a difference{]}{]} over the duration of the 20 year simulation.
+Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius over the duration of the 20 year simulation.
 We therefore simulate 6 scenarios for each population.
@@
 Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.
-\textbf{BL: added both to above paragraph}
-{[}{[}should we mention here or later that we did both with and without measurement error analyses?{]}{]}
-{[}{[}need to make sure that we don't lose sight of random sampling with the following sentences{]}{]}
 
 --Stratified mean vs VAST with and without covariates
-{[}{[}could use an intro sentence here to distinguish design-based (stratified mean) and model-based (VAST) estimators{]}{]}
-{[}{[}may want to highlight the potential of VAST with covariates to address the climate change issue here as the reason we are doing these simulations{]}{]}
 
-Stock assessment scientists use samples from the bottom trawl survey to estimate the abundance for each stock stock.
-There are a number of approaches to obtaining abundance estimates ranging from traditional design-based estimates such as the stratified mean to model-based estimates that range in complexity.
-The stratified mean calculates the geometric mean catch per tow and does not account for spatial variation in samples or allow the inclusion of environmental covariates.
-Model-based abundance estimates are often derived from model framewroks {[}{[}frameworks{]}{]} such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM) that can incorporate both random and mixed effects.
-The Vector Autoregressive Spatio-Temporal (VAST) model is a spatial delta-generalized linear mixed model (Thorson 2019).
-VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurrence (presence/absence).
+Stock assessment scientists use samples from the bottom trawl survey to estimate the abundance for each fish stock.
+There are a number of approaches to obtain abundance estimates including traditional design-based estimates to model-based estimates that range in complexity.
+Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest.
+These methods do not account for spatial variation in samples or allow the inclusion of environmental influences.
+Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors).
+Model-based estimators, such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM), help account for complex relationships between variables and can help overcome problems with sampling design.
+
+We compare abundance estimates from stratified mean calculations to those derived from the Vector Autoregressive Spatio-Temporal (VAST) model.
+The stratified mean is a design-based approach that calculates the geometric mean catch per tow and has traditionally been used with stratified random sample designs.
+VAST is a spatial delta-generalized linear mixed model that estimates both abundance (biomass) and probability of occurrence (presence/absence) (Thorson 2019).
 If desired, VAST also allows users to include covariate data to better inform the model.
@@
 We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses.
-The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and {[}{[}can?{]}{]} take on the order of hours to complete.
+The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and can take on the order of hours to complete.
 We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.

### 2023-03-03 Manuscript/Draft1.Rmd
manuscript update

@@
 
-We compare abundance estimates from stratified mean calculations to those derived from the Vector Autoregressive Spatio-Temporal (VAST) model.
+We compare yearly abundance estimates obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
 The stratified mean is a design-based approach that calculates the geometric mean catch per tow and has traditionally been used with stratified random sample designs.
@@
 The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and can take on the order of hours to complete.
-We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
 
-We follow the advice given in [@thorson2019guidance] to build VAST models to estimate biomass on Georges Bank using stratified mean samples from our model output.
-In addition to exploring different link functions and assumed distributions, our VAST model-building process included testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
-We also carried out the same process both including covariates in our model as well as running models without covariate information.
-We considered covariates in the form of dynamic temperature values and/or static habitat values from our population model.
-When using covariates we ultimately decided to provide the most information to the model by including both covariates for both linear predictors.
-Since we know the true population values in our models we calculate the absolute error of each VAST estimate to compare between potential settings.
-Through this process, and in consultation with the VAST package creator, we determined setting that allowed VAST models to converge for all of our scenarios while also providing the lowest absolute error values.
-Settings for our VAST models can be seen in Table ???.
+We follow the advice given in [@thorson2019guidance] to build VAST models to estimate biomass on Georges Bank using stratified random samples from our model output.
+In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
+We also carried out the same process running models without covariate information as well as including covariates in our model.
+We considered covariates in the form of dynamic temperature values and/or static habitat values ($Hab_{J.s}$) from our population model.
+When using covariates we ultimately decided to provide the most information to the model by including both temperature and habitat covariates for both linear predictors.
+Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings.
+Through this process, and in consultation with the VAST package creator, we ultimately compared the performance of two sets of settings in our VAST models, which can be seen in Table\@ref(tab:VASTsettings).
 
@@
 **BL: Help with correct notation for adding noise**.
-We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges bank [[Bank]] to those that only include a subset of the full spatial domain for each species.
+We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species.
 We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from.
 Tabl XXX [[needs to be added]] lists all strata inhabited by each species, those that are removed from certain calculations, and the explanation of why these strata were removed.
-
-**BL: I tried to provid emore detail above and also added another table**
-[[may need to spell the aspect of reduced spatial domain a bit more because it might be confusing to readers who are expected us to just add areas around the current ones instead of reducing the strata]]
 
@@
 
-The goal of our project was to analyze how well contemporary population indexing methods can track population trends under a variety of conditions, as depicted in Table \@ref(tab:scenarios) Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year [[can cite the 2022 management track assessments, see https://apps-nefsc.fisheries.noaa.gov/saw/sari.php for when the document becomes available]] **cite**.
-For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod (see Table ???) and increasing population scenarios for Haddock (see Table ???).
-To provide a comprehensive analysis of population indexing methods we consider all possible scenario combinations for Yellowtail Flounder (see Table ???).
-The specific population trends used in our analyses can be see in Figure \@ref(fig:pop-scenarios)
+The goal of our project was to analyze the ability of contemporary population indexing methods to track population trends under a variety of conditions, as depicted in Table \@ref(tab:scenarios).
+Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year [[can cite the 2022 management track assessments, see https://apps-nefsc.fisheries.noaa.gov/saw/sari.php for when the document becomes available]] **cite**.
+For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
+To provide a comprehensive analysis of population indexing methods we consider all possible scenario combinations for Yellowtail Flounder.
+The specific population trends used in our analyses can be see in Figure \@ref(fig:pop-scenarios).
 
@@
 
-General themes that exist in Tables X, Y and Z are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors.
-We also see individual cases where the stratified mean produced the lowest absolute error and instances where including covarites in VAST models actually increase the absolute error.
+Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the results of our comparison between the absolute error of abundance estimates and model output.
+General themes are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors.
+We also see individual cases where the stratified mean produced the lowest absolute error and instances where including covariates in VAST models actually increase the absolute error.
 When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.
 
-In the Cod results, when using reduced strata adding covariates produces worst VAST results (though still better than stratified mean).
+(see Table \@ref(tab:CodERROR)) In the Cod results, when using reduced strata adding covariates produces worst VAST results (though still better than stratified mean).
 VAST without covariates much worse than stratified mean in fall with increasing temperature and all strata, but adding covariates corrects this.
 
-For Haddock, VAST has a particularly hard time in spring regularly producing larger errors than the stratified mean with added covariates only improving to the level of the stratified mean.
-VAST shows improved results in fall relative to the stratified mean with added covaraites producing extremely low errors in some cases.
+(see Table \@ref(tab:HadERROR)) For Haddock, VAST has a particularly hard time in spring regularly producing larger errors than the stratified mean with added covariates only improving to the level of the stratified mean.
+VAST shows improved results in fall relative to the stratified mean with added covariates producing extremely low errors in some cases.
 
@@
 
+<font size="5">*Cod Results*</font>
+
+Of the scenarios without covariates, $12/16=75$\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
+Including detailed covariate information in our Cod VAST estimates resulted in all $16/16=100\%$ having a lower relative error than the corresponding stratified mean estimate.
+The 4 scenarios without covariates that had a higher VAST error were each overestimates for the fall season that involved an increasing temperature gradient.
+As seen in Figure XXX, Cod are shifting their spatial distribution much more in the fall than they are in the spring.
+While adding covariates to these fits allowed for an improved error relative to the stratified mean, the VAST abundance still display an overestimate early in the time series before covariates correct the trend, allowing for a lower overall absolute error **(show these plots?
+IncTemp fall scenarios)**.
+
+<font size="5">*Haddock Results*</font>
+
+$10/16\approx63$\% of the scenarios without covariates had a VAST fit with a lower relative error compared to the stratified mean estimate.
+The 6 covariate-free VAST estimates that resulted in a higher error than the stratified mean spanned all scenarios and seasons, with several fall errors being especially large (significant overestimates).
+Adding covariate information resulted in $14/16\approx88$\% of VAST estimates having improved error compared to the stratified mean.
+The 2 scenarios that procedure worse error spanned temperature scenarios, but were both in the spring season when the proportion of the population in each strata remained constant in each scenario.**why??
+doesnt make sense.
+the values with cov are about the same as SM** Qualitatively, including covariate information has a larger impact on decreasing the relative error compared to the stratified mean compared to failing to include covariates.
+
+<font size="5">*Yellowtail Flounder Results*</font>
+
+Of the scenarios without covariates, $33/48\approx68$\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
+All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the Fall season.
+These 15 Fall estimates span all other scenario variations and had a common theme of producing abundance estimates that are well above the true model value.
+Including detailed covariate information made a noticeable difference in our Yellowtail Flounder VAST estimates.
+Of the VAST estimates that included covariates information, $47/48\approx98$\% had a lower relative error than the corresponding stratified mean estimate.
+**For discussion: clearly our perfect covariate information made a big difference here, but the question is could we achieve similar results without "perfect" information?
+Which covariate is best?
+Good question for future research**
+
+A more careful visual analysis of all error plots for each species reveals that VAST estimates tend to provide error estimates that are too high, while the stratified mean estimates are, on average, too low.
+Additionally, model/estimate ratio plots for all VAST estimates remain much closer to the desired value of 1 compared to stratified mean estimates, which can range as high as 10 and exhibit large yearly changes.
+
 \section{Discussion}
@@
 \pagebreak
+
+```{r YTpct, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="add caption"}
+knitr::include_graphics(c("Images/PercentPlots_ConPopConTemp_YT.png","Images/PercentPlots_ConPopIncTemp_YT.png"))
+```
 
@@
 tabl <- "
-|	Parameter	|	Description | Input/Value
-|:-	|:------- |:----
-| *ObsModel* | Link function and assumed distribution | c(10,2)
-|FieldCOnfig | Specified spatial and/or spatio-temporal variation in predictors| c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1)
-| RhoConfig | Specifying whether intercepts or spatio-temporal variation is structured among time intervals | c(Beta1=3, Beta2=3, Epsilon1=0, Epsilon2=4)
-| X1_formula | Right-sided formula affecting the 1st linear predictor  | X1_formula = ~ poly(Temp, degree=2 )
-| X2_formula | Right-sided formula affecting the 2nd linear predictor | X2_formula = ~ poly(Temp, degree=2 ) + poly(Habitat, degree=2 )
+|	Parameter	|	Description | Settings A | Settings B
+|:--	|:------- |:---- |:----
+| *ObsModel* | Link function and assumed distribution | c(10,2) | c(10,2)
+|FieldConfig | Specified spatial and/or spatio-temporal variation in predictors| c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1) | c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1)
+| RhoConfig | Specifying whether intercepts or spatio-temporal variation is structured among time intervals | c(Beta1=3, Beta2=, Epsilon1=0, Epsilon2=4) | c(Beta1=3, Beta2=3, Epsilon1=0, Epsilon2=4)
+| X1_formula | Right-sided formula affecting the 1st linear predictor  | X1_formula = ~ poly(Temp, degree=2 ) | N/A
+| X2_formula | Right-sided formula affecting the 2nd linear predictor | X2_formula = ~ poly(Temp, degree=2 ) + poly(Habitat, degree=2 ) | X2_formula = ~ poly(Temp, degree=2 ) + poly(Habitat, degree=2 )
 

### 2023-03-03 Manuscript/Draft1.tex
manuscript update

@@
 The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and can take on the order of hours to complete.
-We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
 
-We follow the advice given in (Thorson 2019) to build VAST models to estimate biomass on Georges Bank using stratified mean samples from our model output.
-In addition to exploring different link functions and assumed distributions, our VAST model-building process included testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
-We also carried out the same process both including covariates in our model as well as running models without covariate information.
-We considered covariates in the form of dynamic temperature values and/or static habitat values from our population model.
-When using covariates we ultimately decided to provide the most information to the model by including both covariates for both linear predictors.
-Since we know the true population values in our models we calculate the absolute error of each VAST estimate to compare between potential settings.
-Through this process, and in consultation with the VAST package creator, we determined setting that allowed VAST models to converge for all of our scenarios while also providing the lowest absolute error values.
-Settings for our VAST models can be seen in Table ???.
+We follow the advice given in (Thorson 2019) to build VAST models to estimate biomass on Georges Bank using stratified random samples from our model output.
+In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
+We also carried out the same process running models without covariate information as well as including covariates in our model.
+We considered covariates in the form of dynamic temperature values and/or static habitat values (\(Hab_{J.s}\)) from our population model.
+When using covariates we ultimately decided to provide the most information to the model by including both temperature and habitat covariates for both linear predictors.
+Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings.
+Through this process, and in consultation with the VAST package creator, we ultimately compared the performance of two sets of settings in our VAST models, which can be seen in Table\ref{tab:VASTsettings}.
 
@@
 \textbf{BL: Help with correct notation for adding noise}.
-We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges bank {[}{[}Bank{]}{]} to those that only include a subset of the full spatial domain for each species.
+We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species.
 We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from.
 Tabl XXX {[}{[}needs to be added{]}{]} lists all strata inhabited by each species, those that are removed from certain calculations, and the explanation of why these strata were removed.
-
-\textbf{BL: I tried to provid emore detail above and also added another table}
-{[}{[}may need to spell the aspect of reduced spatial domain a bit more because it might be confusing to readers who are expected us to just add areas around the current ones instead of reducing the strata{]}{]}
 
@@
 
-The goal of our project was to analyze how well contemporary population indexing methods can track population trends under a variety of conditions, as depicted in Table \ref{tab:scenarios} Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year {[}{[}can cite the 2022 management track assessments, see \url{https://apps-nefsc.fisheries.noaa.gov/saw/sari.php} for when the document becomes available{]}{]} \textbf{cite}.
-For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod (see Table ???) and increasing population scenarios for Haddock (see Table ???).
-To provide a comprehensive analysis of population indexing methods we consider all possible scenario combinations for Yellowtail Flounder (see Table ???).
-The specific population trends used in our analyses can be see in Figure \ref{fig:pop-scenarios}
+The goal of our project was to analyze the ability of contemporary population indexing methods to track population trends under a variety of conditions, as depicted in Table \ref{tab:scenarios}.
+Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year {[}{[}can cite the 2022 management track assessments, see \url{https://apps-nefsc.fisheries.noaa.gov/saw/sari.php} for when the document becomes available{]}{]} \textbf{cite}.
+For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
+To provide a comprehensive analysis of population indexing methods we consider all possible scenario combinations for Yellowtail Flounder.
+The specific population trends used in our analyses can be see in Figure \ref{fig:pop-scenarios}.
 
@@
 
-General themes that exist in Tables X, Y and Z are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors.
-We also see individual cases where the stratified mean produced the lowest absolute error and instances where including covarites in VAST models actually increase the absolute error.
+Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the results of our comparison between the absolute error of abundance estimates and model output.
+General themes are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors.
+We also see individual cases where the stratified mean produced the lowest absolute error and instances where including covariates in VAST models actually increase the absolute error.
 When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.
 
-In the Cod results, when using reduced strata adding covariates produces worst VAST results (though still better than stratified mean).
+(see Table \ref{tab:CodERROR}) In the Cod results, when using reduced strata adding covariates produces worst VAST results (though still better than stratified mean).
 VAST without covariates much worse than stratified mean in fall with increasing temperature and all strata, but adding covariates corrects this.
 
-For Haddock, VAST has a particularly hard time in spring regularly producing larger errors than the stratified mean with added covariates only improving to the level of the stratified mean.
-VAST shows improved results in fall relative to the stratified mean with added covaraites producing extremely low errors in some cases.
+(see Table \ref{tab:HadERROR}) For Haddock, VAST has a particularly hard time in spring regularly producing larger errors than the stratified mean with added covariates only improving to the level of the stratified mean.
+VAST shows improved results in fall relative to the stratified mean with added covariates producing extremely low errors in some cases.
 
@@
 They can of course examine the tables in detail and draw their own conclusions, but we should put our interpretation down on paper.{]}{]}
+
+\emph{Cod Results}
+
+Of the scenarios without covariates, \(12/16=75\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
+Including detailed covariate information in our Cod VAST estimates resulted in all \(16/16=100\%\) having a lower relative error than the corresponding stratified mean estimate.
+The 4 scenarios without covariates that had a higher VAST error were each overestimates for the fall season that involved an increasing temperature gradient.
+As seen in Figure XXX, Cod are shifting their spatial distribution much more in the fall than they are in the spring.
+While adding covariates to these fits allowed for an improved error relative to the stratified mean, the VAST abundance still display an overestimate early in the time series before covariates correct the trend, allowing for a lower overall absolute error \textbf{(show these plots?
+IncTemp fall scenarios)}.
+
+\emph{Haddock Results}
+
+\(10/16\approx63\)\% of the scenarios without covariates had a VAST fit with a lower relative error compared to the stratified mean estimate.
+The 6 covariate-free VAST estimates that resulted in a higher error than the stratified mean spanned all scenarios and seasons, with several fall errors being especially large (significant overestimates).
+Adding covariate information resulted in \(14/16\approx88\)\% of VAST estimates having improved error compared to the stratified mean.
+The 2 scenarios that procedure worse error spanned temperature scenarios, but were both in the spring season when the proportion of the population in each strata remained constant in each scenario.\textbf{why??
+doesnt make sense.
+the values with cov are about the same as SM} Qualitatively, including covariate information has a larger impact on decreasing the relative error compared to the stratified mean compared to failing to include covariates.
+
+\emph{Yellowtail Flounder Results}
+
+Of the scenarios without covariates, \(33/48\approx68\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
+All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the Fall season.
+These 15 Fall estimates span all other scenario variations and had a common theme of producing abundance estimates that are well above the true model value.
+Including detailed covariate information made a noticeable difference in our Yellowtail Flounder VAST estimates.
+Of the VAST estimates that included covariates information, \(47/48\approx98\)\% had a lower relative error than the corresponding stratified mean estimate.
+\textbf{For discussion: clearly our perfect covariate information made a big difference here, but the question is could we achieve similar results without ``perfect'' information?
+Which covariate is best?
+Good question for future research}
+
+A more careful visual analysis of all error plots for each species reveals that VAST estimates tend to provide error estimates that are too high, while the stratified mean estimates are, on average, too low.
+Additionally, model/estimate ratio plots for all VAST estimates remain much closer to the desired value of 1 compared to stratified mean estimates, which can range as high as 10 and exhibit large yearly changes.
 
@@
 \pagebreak
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{Images/PercentPlots_ConPopConTemp_YT} \includegraphics[width=0.95\linewidth]{Images/PercentPlots_ConPopIncTemp_YT}
+
+}
+
+\caption{add caption}\label{fig:YTpct}
+\end{figure}
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.13}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.53}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.33}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.14}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.38}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.24}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.24}}@{}}
 \toprule
@@
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Input/Value
+Settings A
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Settings B
 \end{minipage} \\
@@
 \endhead
-\emph{ObsModel} & Link function and assumed distribution & c(10,2) \\
-FieldCOnfig & Specified spatial and/or spatio-temporal variation in predictors & c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1) \\
-RhoConfig & Specifying whether intercepts or spatio-temporal variation is structured among time intervals & c(Beta1=3, Beta2=3, Epsilon1=0, Epsilon2=4) \\
-X1\_formula & Right-sided formula affecting the 1st linear predictor & X1\_formula = \textasciitilde{} poly(Temp, degree=2 ) \\
-X2\_formula & Right-sided formula affecting the 2nd linear predictor & X2\_formula = \textasciitilde{} poly(Temp, degree=2 ) + poly(Habitat, degree=2 ) \\
+\emph{ObsModel} & Link function and assumed distribution & c(10,2) & c(10,2) \\
+FieldConfig & Specified spatial and/or spatio-temporal variation in predictors & c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1) & c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1) \\
+RhoConfig & Specifying whether intercepts or spatio-temporal variation is structured among time intervals & c(Beta1=3, Beta2=, Epsilon1=0, Epsilon2=4) & c(Beta1=3, Beta2=3, Epsilon1=0, Epsilon2=4) \\
+X1\_formula & Right-sided formula affecting the 1st linear predictor & X1\_formula = \textasciitilde{} poly(Temp, degree=2 ) & N/A \\
+X2\_formula & Right-sided formula affecting the 2nd linear predictor & X2\_formula = \textasciitilde{} poly(Temp, degree=2 ) + poly(Habitat, degree=2 ) & X2\_formula = \textasciitilde{} poly(Temp, degree=2 ) + poly(Habitat, degree=2 ) \\
 \bottomrule

### 2023-03-03 Manuscript/Draft1.Rmd
update manuscript

@@
 To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depend on temperature preferences.
-We can then consider the impact of climate change by simulating scenarios with repeating temperature patterns and those where temperature increases on average over time.
+We consider the impact of climate change by comparing simulations that use a repeating temperature pattern to those where temperature increases on average over time.
 In both cases we analyze the ability of stratified random sampling to track population trends.
@@
 Model dynamics are driven by a time series of temperature gradients that were estimated from data to create simulated data sets for each population where the true biomass is known.
-Using temperature gradients that repeat each year creates data sets with predictable, repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time, approximating the future distribution in the region under climate change.
-We conducting stratified random sampling on our simulation output to mimic the bottom trawl survey and compare the ability of contemporary indexing methods to track population trends.
+Using temperature gradients that repeat each year creates data sets with predictable, repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time.
+We conduct stratified random sampling on our simulation output to mimic the bottom trawl survey and use the samples to compare the ability of contemporary indexing methods to track population trends.
 
@@
 
+<font size="5">*Yellowtail Flounder Results*</font>
+
+Figure \@ref(fig:PopPct) depicts the spatial shifting that occurs in our population models in each of the survey weeks.
+
+Of the scenarios without covariates, $33/48\approx68$\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
+All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the Fall season.
+These 15 Fall estimates span all other scenario variations and had a common theme of producing abundance estimates that are well above the true model value.
+Including detailed covariate information made a noticeable difference in our Yellowtail Flounder VAST estimates.
+Of the VAST estimates that included covariates information, $47/48\approx98$\% had a lower relative error than the corresponding stratified mean estimate.
+**For discussion: clearly our perfect covariate information made a big difference here, but the question is could we achieve similar results without "perfect" information?
+Which covariate is best?
+Good question for future research**
+
 <font size="5">*Cod Results*</font>
 
-Of the scenarios without covariates, $12/16=75$\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
+As seen in Figure \@ref(fig:PopPct), in the constant temperature scenario the Cod population is decreasing its presence in stratas 19 and 20 over the duration of the simulation in both seasons, while simultaneously increasing their presence in strata 16.
+However, an increase in average temperature had a seasonal impact on Cod's presence in strata 16 in our simulations.
+In the spring the population followed a similar trend of leaving 19 and 20 while simultaneously entering 16.
+However, in the fall the increasing temperature drives the population out of strata 16 and 21, and into 18, 22, and 24.
+**check the impact of this difference on estiamtes**
+
+Of the abundance estimates without covariates, $12/16=75$\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
 Including detailed covariate information in our Cod VAST estimates resulted in all $16/16=100\%$ having a lower relative error than the corresponding stratified mean estimate.
@@
 the values with cov are about the same as SM** Qualitatively, including covariate information has a larger impact on decreasing the relative error compared to the stratified mean compared to failing to include covariates.
-
-<font size="5">*Yellowtail Flounder Results*</font>
-
-Of the scenarios without covariates, $33/48\approx68$\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
-All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the Fall season.
-These 15 Fall estimates span all other scenario variations and had a common theme of producing abundance estimates that are well above the true model value.
-Including detailed covariate information made a noticeable difference in our Yellowtail Flounder VAST estimates.
-Of the VAST estimates that included covariates information, $47/48\approx98$\% had a lower relative error than the corresponding stratified mean estimate.
-**For discussion: clearly our perfect covariate information made a big difference here, but the question is could we achieve similar results without "perfect" information?
-Which covariate is best?
-Good question for future research**
 
@@
 
-```{r YTpct, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="add caption"}
+```{r PopPct, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Percent of each species in each strata for during survey weeks in our spatial simulations.
+All constant temperature scenarios follow the patterns on the left while increasing temperature scenarios follow the patterns on the right.",out.width="47.5%"}
 knitr::include_graphics(c("Images/PercentPlots_ConPopConTemp_YT.png","Images/PercentPlots_ConPopIncTemp_YT.png"))
+knitr::include_graphics(c("Images/PercentPlots_DecPopConTemp_Cod.png","Images/PercentPlots_DecPopIncTemp_Cod.png"))
+knitr::include_graphics(c("Images/PercentPlots_IncPopConTemp_Had.png","Images/PercentPlots_IncPopIncTemp_Had.png"))
 ```

### 2023-03-03 Manuscript/Draft1.tex
update manuscript

@@
 
+\emph{Yellowtail Flounder Results}
+
+Figure \ref{fig:PopPct} depicts the spatial shifting that occurs in our population models in each of the survey weeks.
+
+Of the scenarios without covariates, \(33/48\approx68\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
+All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the Fall season.
+These 15 Fall estimates span all other scenario variations and had a common theme of producing abundance estimates that are well above the true model value.
+Including detailed covariate information made a noticeable difference in our Yellowtail Flounder VAST estimates.
+Of the VAST estimates that included covariates information, \(47/48\approx98\)\% had a lower relative error than the corresponding stratified mean estimate.
+\textbf{For discussion: clearly our perfect covariate information made a big difference here, but the question is could we achieve similar results without ``perfect'' information?
+Which covariate is best?
+Good question for future research}
+
 \emph{Cod Results}
 
-Of the scenarios without covariates, \(12/16=75\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
+As seen in Figure \ref{fig:PopPct}, in the constant temperature scenario the Cod population is decreasing its presence in stratas 19 and 20 over the duration of the simulation in both seasons, while simultaneously increasing their presence in strata 16.
+However, an increase in average temperature had a seasonal impact on Cod's presence in strata 16 in our simulations.
+In the spring the population followed a similar trend of leaving 19 and 20 while simultaneously entering 16.
+However, in the fall the increasing temperature drives the population out of strata 16 and 21, and into 18, 22, and 24.
+\textbf{check the impact of this difference on estiamtes}
+
+Of the abundance estimates without covariates, \(12/16=75\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
 Including detailed covariate information in our Cod VAST estimates resulted in all \(16/16=100\%\) having a lower relative error than the corresponding stratified mean estimate.
@@
 the values with cov are about the same as SM} Qualitatively, including covariate information has a larger impact on decreasing the relative error compared to the stratified mean compared to failing to include covariates.
-
-\emph{Yellowtail Flounder Results}
-
-Of the scenarios without covariates, \(33/48\approx68\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
-All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the Fall season.
-These 15 Fall estimates span all other scenario variations and had a common theme of producing abundance estimates that are well above the true model value.
-Including detailed covariate information made a noticeable difference in our Yellowtail Flounder VAST estimates.
-Of the VAST estimates that included covariates information, \(47/48\approx98\)\% had a lower relative error than the corresponding stratified mean estimate.
-\textbf{For discussion: clearly our perfect covariate information made a big difference here, but the question is could we achieve similar results without ``perfect'' information?
-Which covariate is best?
-Good question for future research}
 
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{Images/PercentPlots_ConPopConTemp_YT} \includegraphics[width=0.95\linewidth]{Images/PercentPlots_ConPopIncTemp_YT}
+{\centering \includegraphics[width=0.475\linewidth]{Images/PercentPlots_ConPopConTemp_YT} \includegraphics[width=0.475\linewidth]{Images/PercentPlots_ConPopIncTemp_YT} \includegraphics[width=0.475\linewidth]{Images/PercentPlots_DecPopConTemp_Cod} \includegraphics[width=0.475\linewidth]{Images/PercentPlots_DecPopIncTemp_Cod} \includegraphics[width=0.475\linewidth]{Images/PercentPlots_IncPopConTemp_Had} \includegraphics[width=0.475\linewidth]{Images/PercentPlots_IncPopIncTemp_Had}
 
@@
 
-\caption{add caption}\label{fig:YTpct}
+\caption{Percent of each species in each strata for during survey weeks in our spatial simulations.
+All constant temperature scenarios follow the patterns on the left while increasing temperature scenarios follow the patterns on the right.}\label{fig:PopPct}
 \end{figure}

### 2023-03-03 TestScripts/percent_area_occupied.R

@@ -29,7 +29,16 @@ strata_species[["Had"]] <- c(13,14,15,16,17,18,19,20,21,22,23,24,25,29,30)
 #number of extra years in simulation. subtract this value from year in log matrix
 years_cut <- 2
 
-scenario <- "ConPop_IncTemp"
+scenario <- "DecPop_IncTemp"
+
+
+#define for labeling later
+if(scenario=="ConPop_ConTemp"){temp_scenario <- " (Constant Temperature)"}
+if(scenario=="ConPop_IncTemp"){temp_scenario <- " (Increasing Temperature)"}
+if(scenario=="IncPop_ConTemp"){temp_scenario <- " (Constant Temperature)"}
+if(scenario=="IncPop_IncTemp"){temp_scenario <- " (Increasing Temperature)"}
+if(scenario=="DecPop_ConTemp"){temp_scenario <- " (Constant Temperature)"}
+if(scenario=="DecPop_IncTemp"){temp_scenario <- " (Increasing Temperature)"}
 
 #FIRST READ IN RES WHICH CONTAINS SPATIAL OUTPUT FOR GIVEN SCENARIO
 
@@ -183,7 +192,7 @@ for(s in short_names){
 
 
 
-pdf(file=paste0('testfolder/',scenario,'_Spatial_pop_plots','.pdf'))
+#pdf(file=paste0('testfolder/',scenario,'_Spatial_pop_plots','.pdf'))
 
 #################################################################################
 # PLOTS
@@ -304,14 +313,20 @@ long_names <- c("Yellowtail Flounder","Cod","Haddock")
 idx<-1
 for(s in short_names){
 p<- ggplot() +
-  geom_line(data=species_pct2[[s]], aes(x=as.numeric(year), y=as.numeric(`% Pop`), group=week, color=week))+
-  geom_point(data=species_pct2[[s]], aes(x=as.numeric(year), y=as.numeric(`% Pop`), group=week, color=week))+
+  geom_line(data=subset(species_pct2[[s]],stratum %in% strata_species[[s]]), aes(x=as.numeric(year), y=as.numeric(`% Pop`), group=week, color=week))+
+  geom_point(data=subset(species_pct2[[s]],stratum %in% strata_species[[s]]), aes(x=as.numeric(year), y=as.numeric(`% Pop`), group=week, color=week))+
   
-  geom_hline(data = species_pct2[[s]],aes( yintercept = as.numeric(`pct area`), linetype = "")) +
[…]

### 2023-03-06 Manuscript/Draft1.Rmd
update manuscript

@@
 ```{r strata-plot, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Strata inhabited by each species in our population models."}
-knitr::include_graphics(c("Images/Strata.png"))
-knitr::include_graphics(c("Images/Excluded_strata.png"))
+#knitr::include_graphics(c("Images/Strata.png"))
+#knitr::include_graphics(c("Images/Excluded_strata.png"))
+knitr::include_graphics(c("Images/Strata2.png"))
+knitr::include_graphics(c("Images/Excluded_strata2.png"))
 ```
@@
 
+Figure \@ref(fig:PopPct) depicts the spatial shifting that occurs in each strata within our population models, specifically during the bottom trawl survey weeks in the spring (13 and 14) and fall (37 and 28).
+See Figure \@ref(fig:strata-plot) for a spatial reference of the Georges Bank strata and Figure \@ref(fig:hab-plot) for the static habitat preferences for each species ($Hab^2_{J,s}$) used in our population models.
+
 <font size="5">*Yellowtail Flounder Results*</font>
 
-Figure \@ref(fig:PopPct) depicts the spatial shifting that occurs in our population models in each of the survey weeks.
+The top two panels in Figure \@ref(fig:PopPct) depict the results for Yellowtail Flounder with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature).
+In both temperature scenarios we see the percent of Yellowtail Flounder in strata 13 decrease over the course of the time series in both seasons.
+The spring population in stratum 19 also decreases in both temperature scenarios as well.
+The percent of the population increases in stratum 16 in the spring over the duration of both time series, which implies the flow out of strata 13 and 19 in the spring are going into stratum 16.
+These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail Flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12.
+These spring dynamics are therefore related to the static habitat values in our model rather than the dynamics temperature preferences.
+While we observe similar dynamics in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \@ref(fig:PopPct)).
+These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow exterior strata 17 and 18.
+One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how ~10\% of the population exists in the narrow exterior strata 17 in the fall, while seemingly none of the population exists in any of the exterior strata (14, 15, 17, 18) in the spring.
 
-Of the scenarios without covariates, $33/48\approx68$\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
-All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the Fall season.
-These 15 Fall estimates span all other scenario variations and had a common theme of producing abundance estimates that are well above the true model value.
-Including detailed covariate information made a noticeable difference in our Yellowtail Flounder VAST estimates.
-Of the VAST estimates that included covariates information, $47/48\approx98$\% had a lower relative error than the corresponding stratified mean estimate.
+Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata) contain the absolute error between our abundance estimates for Yellowtail Flounder comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B).
+Of all scenarios without covariates, $33/48\approx68$\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
+All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the fall season and had a common theme of producing abundance estimates that are above the true model value.
+These 15 fall estimates span all other scenario variations.
+The implication of this is that our model-based approach without covariates struggles with the primary seasonal difference for Yellowtail Flounder, which is that a larger percentage of the population exists in the narrow exterior strata (18 and/or 17).
+This theory is further supported by the fact that the absolute error in the increasing temperature scenarios increased dramatically in the fall season, when the combined percentage of the population in the outer strata 17 and 18 increased to over 40\% by the end of the simulation.
+
+Including detailed covariate information made a noticeable difference in our Yellowtail Flounder model-based VAST estimates.
+All of the VAST estimates that included covariate information produced a lower relative error than the corresponding VAST model that did not include covariates.
+The largest improvements were seen in the increasing temperature scenarios.
+As a result, of the VAST estimates that included covariates information, $47/48\approx98$\% had a lower relative error than the corresponding stratified mean estimate.
+This implies that the covariate information helped our model-based estimate account for the design-based issues related to small strata that contain a large percentage of the population, which can become exacerbated in the increasing temperature scenario.
+
+The performance of estimates tend to decrease under increasing temperature, with the most dramatic changes seen in our VAST estimates without covariates.
+One exception to this is when we use a stratified mean approach while the population is decreasing.
+Since our analyses have found that the stratified mean tends to under estimate the true abundance, as the Yellowtail population decreases towards zero the difference between the estimate and the true value also decrease.
+That is, if the population is low enough, failing to appropriately sample the population produces the same result as appropriately sampling.
+
 **For discussion: clearly our perfect covariate information made a big difference here, but the question is could we achieve similar results without "perfect" information?
@@
 
-As seen in Figure \@ref(fig:PopPct), in the constant temperature scenario the Cod population is decreasing its presence in stratas 19 and 20 over the duration of the simulation in both seasons, while simultaneously increasing their presence in strata 16.
+As seen in Figure \@ref(fig:PopPct), in the constant temperature scenario the Cod population is decreasing its presence in strata 19 and 20 over the duration of the simulation in both seasons, while simultaneously increasing their presence in strata 16.
 However, an increase in average temperature had a seasonal impact on Cod's presence in strata 16 in our simulations.
@@
 However, in the fall the increasing temperature drives the population out of strata 16 and 21, and into 18, 22, and 24.
-**check the impact of this difference on estiamtes**
+**check the impact of this difference on estimates**
 

### 2023-03-06 Manuscript/Draft1.tex
update manuscript

@@
 
-{\centering \includegraphics[width=0.95\linewidth]{Images/Strata} \includegraphics[width=0.95\linewidth]{Images/Excluded_strata}
+{\centering \includegraphics[width=0.95\linewidth]{Images/Strata2} \includegraphics[width=0.95\linewidth]{Images/Excluded_strata2}
 
@@
 
+Figure \ref{fig:PopPct} depicts the spatial shifting that occurs in each strata within our population models, specifically during the bottom trawl survey weeks in the spring (13 and 14) and fall (37 and 28).
+See Figure \ref{fig:strata-plot} for a spatial reference of the Georges Bank strata and Figure \ref{fig:hab-plot} for the static habitat preferences for each species (\(Hab^2_{J,s}\)) used in our population models.
+
 \emph{Yellowtail Flounder Results}
 
-Figure \ref{fig:PopPct} depicts the spatial shifting that occurs in our population models in each of the survey weeks.
+The top two panels in Figure \ref{fig:PopPct} depict the results for Yellowtail Flounder with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature).
+In both temperature scenarios we see the percent of Yellowtail Flounder in strata 13 decrease over the course of the time series in both seasons.
+The spring population in stratum 19 also decreases in both temperature scenarios as well.
+The percent of the population increases in stratum 16 in the spring over the duration of both time series, which implies the flow out of strata 13 and 19 in the spring are going into stratum 16.
+These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail Flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12.
+These spring dynamics are therefore related to the static habitat values in our model rather than the dynamics temperature preferences.
+While we observe similar dynamics in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \ref{fig:PopPct}).
+These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow exterior strata 17 and 18.
+One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how \textasciitilde10\% of the population exists in the narrow exterior strata 17 in the fall, while seemingly none of the population exists in any of the exterior strata (14, 15, 17, 18) in the spring.
 
-Of the scenarios without covariates, \(33/48\approx68\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
-All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the Fall season.
-These 15 Fall estimates span all other scenario variations and had a common theme of producing abundance estimates that are well above the true model value.
-Including detailed covariate information made a noticeable difference in our Yellowtail Flounder VAST estimates.
-Of the VAST estimates that included covariates information, \(47/48\approx98\)\% had a lower relative error than the corresponding stratified mean estimate.
+Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata} contain the absolute error between our abundance estimates for Yellowtail Flounder comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B).
+Of all scenarios without covariates, \(33/48\approx68\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
+All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the fall season and had a common theme of producing abundance estimates that are above the true model value.
+These 15 fall estimates span all other scenario variations.
+The implication of this is that our model-based approach without covariates struggles with the primary seasonal difference for Yellowtail Flounder, which is that a larger percentage of the population exists in the narrow exterior strata (18 and/or 17).
+This theory is further supported by the fact that the absolute error in the increasing temperature scenarios increased dramatically in the fall season, when the combined percentage of the population in the outer strata 17 and 18 increased to over 40\% by the end of the simulation.
+
+Including detailed covariate information made a noticeable difference in our Yellowtail Flounder model-based VAST estimates.
+All of the VAST estimates that included covariate information produced a lower relative error than the corresponding VAST model that did not include covariates.
+The largest improvements were seen in the increasing temperature scenarios.
+As a result, of the VAST estimates that included covariates information, \(47/48\approx98\)\% had a lower relative error than the corresponding stratified mean estimate.
+This implies that the covariate information helped our model-based estimate account for the design-based issues related to small strata that contain a large percentage of the population, which can become exacerbated in the increasing temperature scenario.
+
+The performance of estimates tend to decrease under increasing temperature, with the most dramatic changes seen in our VAST estimates without covariates.
+One exception to this is when we use a stratified mean approach while the population is decreasing.
+Since our analyses have found that the stratified mean tends to under estimate the true abundance, as the Yellowtail population decreases towards zero the difference between the estimate and the true value also decrease.
+That is, if the population is low enough, failing to appropriately sample the population produces the same result as appropriately sampling.
+
 \textbf{For discussion: clearly our perfect covariate information made a big difference here, but the question is could we achieve similar results without ``perfect'' information?
@@
 
-As seen in Figure \ref{fig:PopPct}, in the constant temperature scenario the Cod population is decreasing its presence in stratas 19 and 20 over the duration of the simulation in both seasons, while simultaneously increasing their presence in strata 16.
+As seen in Figure \ref{fig:PopPct}, in the constant temperature scenario the Cod population is decreasing its presence in strata 19 and 20 over the duration of the simulation in both seasons, while simultaneously increasing their presence in strata 16.
 However, an increase in average temperature had a seasonal impact on Cod's presence in strata 16 in our simulations.
@@
 However, in the fall the increasing temperature drives the population out of strata 16 and 21, and into 18, 22, and 24.
-\textbf{check the impact of this difference on estiamtes}
+\textbf{check the impact of this difference on estimates}
 

### 2023-03-06 Manuscript/Draft1.Rmd
manuscript update

@@
 
-The performance of estimates tend to decrease under increasing temperature, with the most dramatic changes seen in our VAST estimates without covariates.
+The performance of our abundance estimates tend to decrease under increasing temperature, with the most dramatic changes seen in our VAST estimates without covariates.
 One exception to this is when we use a stratified mean approach while the population is decreasing.
 Since our analyses have found that the stratified mean tends to under estimate the true abundance, as the Yellowtail population decreases towards zero the difference between the estimate and the true value also decrease.
-That is, if the population is low enough, failing to appropriately sample the population produces the same result as appropriately sampling.
+That is, if the population is low enough, failing to appropriately sample the population in a design-based method produces the same result as appropriately sampling.
 
@@
 
-As seen in Figure \@ref(fig:PopPct), in the constant temperature scenario the Cod population is decreasing its presence in strata 19 and 20 over the duration of the simulation in both seasons, while simultaneously increasing their presence in strata 16.
+As seen in Figure \@ref(fig:PopPct), in the constant temperature scenario the Cod population is decreasing its presence in strata 19 and 20 over the duration of the simulation in both seasons, while simultaneously increasing presence in stratum 16.
 However, an increase in average temperature had a seasonal impact on Cod's presence in strata 16 in our simulations.
-In the spring the population followed a similar trend of leaving 19 and 20 while simultaneously entering 16.
-However, in the fall the increasing temperature drives the population out of strata 16 and 21, and into 18, 22, and 24.
+In the spring the population followed a similar trend of leaving 19 and 20 while simultaneously entering 16, but in the fall the increasing temperature drives the population out of strata 16 and 21, and into 18, 22, and 24.
 **check the impact of this difference on estimates**
@@
 <font size="5">*Haddock Results*</font>
+
+Figure \@ref(fig:PopPct) reveals some subtle seasonal differences in the percent of haddock in each strata.
+In the constant temperature scenarios, the spring shows a decrease in strata 19 and 20 that correspond to increases in 16, 29 and 24.
+While strata 24 and 29 also increase in the fall under constant temperature, the corresponding decrease is primarily from strata 16 and 13.
+While similar results can be seen in the spring for the increasing temperature scenario, much more dramatic results exist in the fall under increasing temperature as we see a significant decreases in strata 13, 16, 21, and 22 that leads to the most noticable increases in strata 17, 18, and 29.
+This shift represents movement from the less deep and more centrally located strata towards exterior deeper strata.
 

### 2023-03-06 Manuscript/Draft1.tex
manuscript update

@@
 
-As seen in Figure \ref{fig:PopPct}, in the constant temperature scenario the Cod population is decreasing its presence in strata 19 and 20 over the duration of the simulation in both seasons, while simultaneously increasing their presence in strata 16.
+As seen in Figure \ref{fig:PopPct}, in the constant temperature scenario the Cod population is decreasing its presence in strata 19 and 20 over the duration of the simulation in both seasons, while simultaneously increasing presence in stratum 16.
 However, an increase in average temperature had a seasonal impact on Cod's presence in strata 16 in our simulations.
-In the spring the population followed a similar trend of leaving 19 and 20 while simultaneously entering 16.
-However, in the fall the increasing temperature drives the population out of strata 16 and 21, and into 18, 22, and 24.
+In the spring the population followed a similar trend of leaving 19 and 20 while simultaneously entering 16, but in the fall the increasing temperature drives the population out of strata 16 and 21, and into 18, 22, and 24.
 \textbf{check the impact of this difference on estimates}
@@
 \emph{Haddock Results}
+
+Figure \ref{fig:PopPct} reveals some subtle seasonal differences in the percent of haddock in each strata.
+In the constant temperature scenarios, the spring shows a decrease in strata 19 and 20 that correspond to increases in 16, 29 and 24.
+While strata 24 and 29 also increase in the fall under constant temperature, the corresponding decrease is primarily from strata 16 and 13.
+While similar results can be seen in the spring for the increasing temperature scenario, much more dramatic results exist in the fall under increasing temperature as we see a significant decreases in strata 13, 16, 21, and 22 that leads to the most noticable increases in strata 17, 18, and 29.
+This shift represents movement from the less deep and more centrally located strata towards exterior deeper strata.
 

### 2023-03-09 Manuscript/Draft1.Rmd
manuscript update

@@
 One exception to this is when we use a stratified mean approach while the population is decreasing.
-Since our analyses have found that the stratified mean tends to under estimate the true abundance, as the Yellowtail population decreases towards zero the difference between the estimate and the true value also decrease.
+Our analyses have found that the stratified mean tends to under estimate the true abundance and since these estimates are bounded below by zero, as the Yellowtail population decreases towards zero the difference between the estimate and the true value also decrease.
 That is, if the population is low enough, failing to appropriately sample the population in a design-based method produces the same result as appropriately sampling.
@@
 
-As seen in Figure \@ref(fig:PopPct), in the constant temperature scenario the Cod population is decreasing its presence in strata 19 and 20 over the duration of the simulation in both seasons, while simultaneously increasing presence in stratum 16.
-However, an increase in average temperature had a seasonal impact on Cod's presence in strata 16 in our simulations.
-In the spring the population followed a similar trend of leaving 19 and 20 while simultaneously entering 16, but in the fall the increasing temperature drives the population out of strata 16 and 21, and into 18, 22, and 24.
-**check the impact of this difference on estimates**
+We see a seasonal impact of an increasing average temperature in stratum 16 in the Cod results in Figure \@ref(fig:PopPct).
+In the constant temperature scenario for Cod, in both seasons the population is decreasing its presence in strata 19 and 20 over the duration of the simulation, while simultaneously increasing presence in stratum 16.
+However, an increase in average temperature in the fall drives the population out of strata 16 and 21, and into 18, 22, and 24 (see Figure \@ref(fig:hab-plot)).
+Similar to the Yellowtail Flounder population, stratum the habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler.
+When the temperature increases over time the fall population begins to prefer the adjacent strata that are deeper and/or more northern than stratum 16.
 
@@
 The 4 scenarios without covariates that had a higher VAST error were each overestimates for the fall season that involved an increasing temperature gradient.
-As seen in Figure XXX, Cod are shifting their spatial distribution much more in the fall than they are in the spring.
+Similar to Yellowtail Flounder results, Cod are shifting their spatial distribution into the smaller exterior strata in these conditions, which is decreasing the accuracy of the covariate-free VAST estimate.
 While adding covariates to these fits allowed for an improved error relative to the stratified mean, the VAST abundance still display an overestimate early in the time series before covariates correct the trend, allowing for a lower overall absolute error **(show these plots?
@@
 Figure \@ref(fig:PopPct) reveals some subtle seasonal differences in the percent of haddock in each strata.
-In the constant temperature scenarios, the spring shows a decrease in strata 19 and 20 that correspond to increases in 16, 29 and 24.
+In the constant temperature scenarios, the spring shows a decrease in strata 19 and 20 that correspond to increases in 16, 24 and 29.
+This change represents a northward movement between larger centrally located strata.
 While strata 24 and 29 also increase in the fall under constant temperature, the corresponding decrease is primarily from strata 16 and 13.
-While similar results can be seen in the spring for the increasing temperature scenario, much more dramatic results exist in the fall under increasing temperature as we see a significant decreases in strata 13, 16, 21, and 22 that leads to the most noticable increases in strata 17, 18, and 29.
-This shift represents movement from the less deep and more centrally located strata towards exterior deeper strata.
+While similar results can be seen in the spring for the increasing temperature scenario, much more dramatic results exist in the fall under increasing temperature as we see a significant decreases in strata 13, 16, 21, and 22 that leads to the most noticeable increases in strata 17, 18, and 29.
+This shift represents movement from the shallower and more centrally located strata towards exterior deeper strata.
+Since stratum 16 contains very favorable habitat including much of the species' spawning ground, the strong shift out of 16 and into the northern most strata of 29 in the fall demonstrates a climate-driven change in movement preference.
 

### 2023-03-09 Manuscript/Draft1.tex
manuscript update

@@
 
-As seen in Figure \ref{fig:PopPct}, in the constant temperature scenario the Cod population is decreasing its presence in strata 19 and 20 over the duration of the simulation in both seasons, while simultaneously increasing presence in stratum 16.
-However, an increase in average temperature had a seasonal impact on Cod's presence in strata 16 in our simulations.
-In the spring the population followed a similar trend of leaving 19 and 20 while simultaneously entering 16, but in the fall the increasing temperature drives the population out of strata 16 and 21, and into 18, 22, and 24.
-\textbf{check the impact of this difference on estimates}
+We see a seasonal impact of an increasing average temperature in stratum 16 in the Cod results in Figure \ref{fig:PopPct}.
+In the constant temperature scenario for Cod, in both seasons the population is decreasing its presence in strata 19 and 20 over the duration of the simulation, while simultaneously increasing presence in stratum 16.
+However, an increase in average temperature in the fall drives the population out of strata 16 and 21, and into 18, 22, and 24 (see Figure \ref{fig:hab-plot}).
+Similar to the Yellowtail Flounder population, stratum the habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler.
+When the temperature increases over time the fall population begins to prefer the adjacent strata that are deeper and/or more northern than stratum 16.
 
@@
 The 4 scenarios without covariates that had a higher VAST error were each overestimates for the fall season that involved an increasing temperature gradient.
-As seen in Figure XXX, Cod are shifting their spatial distribution much more in the fall than they are in the spring.
+Similar to Yellowtail Flounder results, Cod are shifting their spatial distribution into the smaller exterior strata in these conditions, which is decreasing the accuracy of the covariate-free VAST estimate.
 While adding covariates to these fits allowed for an improved error relative to the stratified mean, the VAST abundance still display an overestimate early in the time series before covariates correct the trend, allowing for a lower overall absolute error \textbf{(show these plots?
@@
 Figure \ref{fig:PopPct} reveals some subtle seasonal differences in the percent of haddock in each strata.
-In the constant temperature scenarios, the spring shows a decrease in strata 19 and 20 that correspond to increases in 16, 29 and 24.
+In the constant temperature scenarios, the spring shows a decrease in strata 19 and 20 that correspond to increases in 16, 24 and 29.
+This change represents a northward movement between larger centrally located strata.
 While strata 24 and 29 also increase in the fall under constant temperature, the corresponding decrease is primarily from strata 16 and 13.
-While similar results can be seen in the spring for the increasing temperature scenario, much more dramatic results exist in the fall under increasing temperature as we see a significant decreases in strata 13, 16, 21, and 22 that leads to the most noticable increases in strata 17, 18, and 29.
-This shift represents movement from the less deep and more centrally located strata towards exterior deeper strata.
+While similar results can be seen in the spring for the increasing temperature scenario, much more dramatic results exist in the fall under increasing temperature as we see a significant decreases in strata 13, 16, 21, and 22 that leads to the most noticeable increases in strata 17, 18, and 29.
+This shift represents movement from the shallower and more centrally located strata towards exterior deeper strata.
+Since stratum 16 contains very favorable habitat including much of the species' spawning ground, the strong shift out of 16 and into the northern most strata of 29 in the fall demonstrates a climate-driven change in movement preference.
 

### 2023-03-10 Manuscript/Draft1.Rmd
updates

@@
 
-(see Table \@ref(tab:CodERROR)) In the Cod results, when using reduced strata adding covariates produces worst VAST results (though still better than stratified mean).
-VAST without covariates much worse than stratified mean in fall with increasing temperature and all strata, but adding covariates corrects this.
+(see Table \@ref(tab:CodERROR)) In the Cod results, when using reduced strata in constant temperature, adding covariates produces worse VAST results (though still better than stratified mean).
+This is similar to what we see with all of Haddock results (adding covariates makes it worse with reduced strata).
+This is NOT seen with Yellowtail (adding covariates always helps).
+VAST without covariates much worse than stratified mean in fall with increasing temperature, but adding covariates corrects this and makes VAST estimates better than stratified mean.
 
@@
 VAST shows improved results in fall relative to the stratified mean with added covariates producing extremely low errors in some cases.
+In Haddock results, adding covariates improves estimates only when all strata are included (when reduced strata, adding covariates actually makes VAST worse).
+Since this occurs in all scenarios, it seems to again be related to failing to accurately monitor strata 17.
 
@@
 These errors are corrected by including covariate information allowing for an improved VAST estimate that are significantly lower than their stratified mean counterparts.
+Also for Yellowtail, with all strata included adding covariates greatly improved estimates.
+With strata excluded there was still an improvement with covariates, but much less dramatic than with all strata.
 
@@
 
-A more careful visual analysis of all error plots for each species reveals that VAST estimates tend to provide error estimates that are too high, while the stratified mean estimates are, on average, too low.
-Additionally, model/estimate ratio plots for all VAST estimates remain much closer to the desired value of 1 compared to stratified mean estimates, which can range as high as 10 and exhibit large yearly changes.
+<font size="5">*Model / Estimate Ratio Results*</font>
+
+A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the model value, while the stratified mean estimates are, on average, too low.
+This can be further examined by viewing model/estimate ratio plots, in which VAST estimates remain much closer to the desired value of 1 compared to stratified mean estimates, which can range as high as 10 and exhibit large yearly changes.
+
+A more careful analysis of individual yearly model/estimate ratios when all strata are included revealed that about 26\% of all VAST ratios were above 1 (84\% less than 1), with an average of 85\% with a standard deviation of 0.09.
+On the other hand, 67\% of stratified mean estimates were above 1 with an average of 1.51 and a standard deviation of 0.31.
+There were seasonal differences in estimate ratios with spring VAST ratios producing (0.96,0.07), fall VAST ratios being (0.74,0.14), stratified mean spring ratios resulting in (1.31,0.19), and fall values of (1.78,0.51).
+The breakdown for individual species followed a similar pattern.
+
+When the entire domain is sampled, adding covariates brings the estimate ratio closer to 1.
+However, with a reduced domain adding covariates can decrease the accuracy of VAST estimates.
+When including all strata, the stratified mean produced (1.54,0.32), while adding covariates improved the VAST estimate ratio from (0.79,0.06) to (0.92,0.06).
+With a reduced number of strata, the stratified mean resulted in our worst summary values of (3.48,1.82), and adding covariates to our VAST models decreased the ratio results from (1.33,0.31 ) to (1.47,0.34).
 

### 2023-03-10 Manuscript/Draft1.tex
updates

@@
 
-(see Table \ref{tab:CodERROR}) In the Cod results, when using reduced strata adding covariates produces worst VAST results (though still better than stratified mean).
-VAST without covariates much worse than stratified mean in fall with increasing temperature and all strata, but adding covariates corrects this.
+(see Table \ref{tab:CodERROR}) In the Cod results, when using reduced strata in constant temperature, adding covariates produces worse VAST results (though still better than stratified mean).
+This is similar to what we see with all of Haddock results (adding covariates makes it worse with reduced strata).
+This is NOT seen with Yellowtail (adding covariates always helps).
+VAST without covariates much worse than stratified mean in fall with increasing temperature, but adding covariates corrects this and makes VAST estimates better than stratified mean.
 

### 2023-03-10 TestScripts/Plots_for_paper_presentation.R

@@ -1,0 +1,274 @@
+
+
+#RUN BELOW AFTER RUNNING VAST_STRATMEAN_ERROR_COMPARISON_ALL_FORPAPER FIRST TO GET THE CORRECT VARIABLES
+
+
+#YELLOWTAIL ONLY CONSTNATPOPULATION INCREASING TEMPERATURE PLOT
+
+model_line_size = 1.5
+model_point_size = 3.5
+estimate_line_size = 1.25
+estimate_point_size = 3
+
+season_ = 2
+
+ggplot() +
+  
+  
+  #this way plots data by season
+  geom_point(data = subset(as.data.frame(pop_by_season[[s]]),season==season_), aes(x=as.numeric(year),y=biomass, group = season, color = "Model"),size=model_point_size) +
+  geom_line(data = subset(as.data.frame(pop_by_season[[s]]),season==season_), aes(x=as.numeric(year),y=biomass, group =season, color = "Model"),size=model_line_size) +
+  
+  
+  
+  #plot VAST estimate without covariates with NO noise
+  geom_errorbar(data=subset(VAST_data[[s]][["_NoCovs_"]][["NoNoise_"]][[folder]],season==season_),aes(x=Year,y=Estimate,group=season,ymin=Estimate-(1.96*Std..Error.for.Estimate), ymax=Estimate+(1.96*Std..Error.for.Estimate), color = "VAST No Cov No Noise"),width=.3) +
+  #geom_linerange(data=subset(VAST_data[[s]][[folder]],season==2),aes(x=Year,y=Estimate,group=season,ymin=Estimate-(1.96*Std..Error.for.Estimate), ymax=Estimate+(1.96*Std..Error.for.Estimate), color = "VAST Estimate")) +
+  geom_point(data=subset(VAST_data[[s]][["_NoCovs_"]][["NoNoise_"]][[folder]],season==season_),aes(x=Year,y=Estimate,group=season, color = "VAST No Cov No Noise"), size=estimate_point_size)+
+  geom_line(data=subset(VAST_data[[s]][["_NoCovs_"]][["NoNoise_"]][[folder]],season==season_),aes(x=Year,y=Estimate,group=season, color = "VAST No Cov No Noise"),size=estimate_line_size)+
+  # 
+  # #plot VAST estimate without covariates with WITH noise
[…]

### 2023-03-10 TestScripts/VAST_StratMean_Error_Comparison_All_withCovs.R

@@ -1,0 +1,834 @@
+#THIS SCRIPT IS WAY TO GO BEYOND AIC AND MEASURE ACTUAL ERROR BETWEEN MODEL SELECTION OPTIONS AND 
+#TRUE MODEL OUTPUT
+
+
+
+##################################################################################################
+#THINGS WE NEED
+##################################################################################################
+scenario <- "ConPop_ConTemp_7_29"
+
+#original project directory so we can switch back to it
+orig.dir <- getwd()
+
+n_spp <- 3
+
+years_sim <- 22
+
+years_cut <- 2
+
+#survey results without noise
+list_all_temp <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Results\\",scenario,"\\list_all_",scenario,".RDS",sep=""))
+
+#simulation results
+memory.limit(45000)
+result <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Results\\",scenario,"\\result_",scenario,".RDS",sep=""))
+
+#load existing result_goodones, if it exists
+result <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Results\\",scenario,"\\result_goodones_",scenario,".RDS",sep=""))
+
+#1- single set of random survey locations used in stratified mean analysis
+surv_random <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Results\\",scenario,"\\surv_random_",scenario,".RDS",sep=""))
+
+#pick specific simulation
+#For YT:
+#for Conpop_ConTemp (also used for IncTemp) run 1 shows decent constant value with single spike early on
+#for Incpop_ConTemp (also used for IncTemp) run 77 shows strong increase for yellowtail
+#for DecPop_ConTemp run 25 shows clear decrease with small values towards the end
+#for DecPop_IncTemp run 13 shows clear decrease with small values towards the end
+
+#For Cod:
+#for ConPop_ConTemp iteration 13 is pretty good
+#for ConPop_IncTemp iteration 1 is pretty good
+#for IncPop_ConTemp iteration 63 shows clear increase with some variation towards end
+#for IncPop_IncTemp iteration 44 shows clear increase with some variation towards end
[…]

### 2023-03-10 TestScripts/VAST_StratMean_Error_Comparison_All_withCovs_ForPaper.R

@@ -1,0 +1,1180 @@
+#THIS SCRIPT IS WAY TO GO BEYOND AIC AND MEASURE ACTUAL ERROR BETWEEN MODEL SELECTION OPTIONS AND 
+#TRUE MODEL OUTPUT
+
+#original project directory so we can switch back to it
+orig.dir <- getwd()
+
+
+
+##################################################################################################
+#THINGS WE NEED
+##################################################################################################
+scenario1 <- "ConPop_ConTemp" #the folder name
+
+#spp1 spp2 spp3
+#short_names <- c("YT","Cod","Had")   #fixed above
+short_names <- c("YT")#,"Had")
+
+exclude_strata <- TRUE
+
+#for getting into correct subfolder
+ifelse(exclude_strata==TRUE, 
+       {str_dir <- "ExcludeStrata"},
+       {str_dir <- "AllStrata"})
+
+###################################################################################################
+
+n_spp <- 3
+
+years_sim <- 22
+
+years_cut <- 2
+
+#survey results without noise
+list_all_temp <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Simulation_Results\\",scenario1,"\\list_all_",scenario1,".RDS",sep=""))
+
+#simulation results
+#memory.limit(45000)
+#result <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Results\\",scenario1,"\\result_",scenario1,".RDS",sep=""))
+
+#load existing result_goodones, if it exists
+result <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Simulation_Results\\",scenario1,"\\result_goodones_",scenario1,".RDS",sep=""))
+
+#1- single set of random survey locations used in stratified mean analysis
+surv_random <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Simulation_Results\\",scenario1,"\\surv_random_",scenario1,".RDS",sep=""))
+
+#pick specific simulation
+#For YT:
+#for Conpop_ConTemp (also used for IncTemp) run 1 shows decent constant value with single spike early on
+#for Incpop_ConTemp (also used for IncTemp) run 77 shows strong increase for yellowtail
[…]

### 2023-03-15 Manuscript/Draft1.Rmd
updated manuscript and ratio plots

@@
 
-Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the results of our comparison between the absolute error of abundance estimates and model output.
-General themes are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors.
-We also see individual cases where the stratified mean produced the lowest absolute error and instances where including covariates in VAST models actually increase the absolute error.
-When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.
+Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the results of our comparison between the absolute error between abundance estimates and model output.
+General themes are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors and standard deviations.
+On the contrary, we also see individual cases where the stratified mean produced the lowest absolute error.
+When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.
+Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain.
+
+In considering the Yellowtail Flounder results in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata), we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
+When all strata are included on sampling, adding covariates greatly improved estimates in all scenarios.
+We still see an improvement in VAST estimates when certain strata are excluded from sampling, but the improvement was much less dramatic than with all strata included.
+However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata).
+These errors are corrected by including covariate information resulting in improved VAST estimates that are significantly lower than their stratified mean counterparts.
 
@@
 
-In considering the Yellowtail Flounder results in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata), we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
-However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in the Table.
-These errors are corrected by including covariate information allowing for an improved VAST estimate that are significantly lower than their stratified mean counterparts.
-Also for Yellowtail, with all strata included adding covariates greatly improved estimates.
-With strata excluded there was still an improvement with covariates, but much less dramatic than with all strata.
+<font size="5">*Differences Between VAST Settings*</font>
 
-Notes on differences between VAST settings: For YTF, 29/96 ~ 30% of VAST runs with new settings were better (70% were worse).
+For YTF, 29/96 ~ 30% of VAST runs with new settings were better (70% were worse).
 80/96 ~ 83% of scenarios had a VAST run with a better error than the stratified mean.
@@
 
-A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the model value, while the stratified mean estimates are, on average, too low.
-This can be further examined by viewing model/estimate ratio plots, in which VAST estimates remain much closer to the desired value of 1 compared to stratified mean estimates, which can range as high as 10 and exhibit large yearly changes.
+A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, too low.
+This can be further examined by viewing estimate/ratio ratio plots, where we divide the yearly abundance estimates by the true model value.
+In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can range as low as 0 and exhibit large yearly changes.
 
-A more careful analysis of individual yearly model/estimate ratios when all strata are included revealed that about 26\% of all VAST ratios were above 1 (84\% less than 1), with an average of 85\% with a standard deviation of 0.09.
-On the other hand, 67\% of stratified mean estimates were above 1 with an average of 1.51 and a standard deviation of 0.31.
-There were seasonal differences in estimate ratios with spring VAST ratios producing (0.96,0.07), fall VAST ratios being (0.74,0.14), stratified mean spring ratios resulting in (1.31,0.19), and fall values of (1.78,0.51).
+A more careful analysis of individual yearly model/estimate ratios when all strata are included revealed that about 73\% of all VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21.
+On the other hand, 33\% of stratified mean estimates were above 1 with an average of 0.874 and a standard deviation of 0.12.
+There were seasonal differences in estimate ratios with spring VAST ratios producing (1.08,0.12), fall VAST ratios being (1.50,0.38), stratified mean spring ratios resulting in (0.91,0.10), and fall values of (0.84,0.16).
 The breakdown for individual species followed a similar pattern.
@@
 When the entire domain is sampled, adding covariates brings the estimate ratio closer to 1.
-However, with a reduced domain adding covariates can decrease the accuracy of VAST estimates.
-When including all strata, the stratified mean produced (1.54,0.32), while adding covariates improved the VAST estimate ratio from (0.79,0.06) to (0.92,0.06).
-With a reduced number of strata, the stratified mean resulted in our worst summary values of (3.48,1.82), and adding covariates to our VAST models decreased the ratio results from (1.33,0.31 ) to (1.47,0.34).
+However, when failing to sample the entire domain, adding covariates can decrease the accuracy of VAST estimates.
+When including all strata, the stratified mean produced (0.87,0.12), while adding covariates improved the VAST estimate ratio from (1.45,0.17) to (1.13,0.09).
+With a reduced number of strata, the stratified mean resulted in our worst summary values of (0.59,0.10), and adding covariates to our VAST models decreased the average ratio results from (1.04,0.33 ) to (0.78,0.19).
 

### 2023-03-15 Manuscript/Draft1.tex
updated manuscript and ratio plots

@@
 
-Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the results of our comparison between the absolute error of abundance estimates and model output.
-General themes are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors.
-We also see individual cases where the stratified mean produced the lowest absolute error and instances where including covariates in VAST models actually increase the absolute error.
-When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.
+Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the results of our comparison between the absolute error between abundance estimates and model output.
+General themes are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors and standard deviations.
+On the contrary, we also see individual cases where the stratified mean produced the lowest absolute error.
+When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.
+Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain.
+
+In considering the Yellowtail Flounder results in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}, we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
+When all strata are included on sampling, adding covariates greatly improved estimates in all scenarios.
+We still see an improvement in VAST estimates when certain strata are excluded from sampling, but the improvement was much less dramatic than with all strata included.
+However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}.
+These errors are corrected by including covariate information resulting in improved VAST estimates that are significantly lower than their stratified mean counterparts.
 
@@
 
-In considering the Yellowtail Flounder results in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}, we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
-However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in the Table.
-These errors are corrected by including covariate information allowing for an improved VAST estimate that are significantly lower than their stratified mean counterparts.
-Also for Yellowtail, with all strata included adding covariates greatly improved estimates.
-With strata excluded there was still an improvement with covariates, but much less dramatic than with all strata.
-
-Notes on differences between VAST settings: For YTF, 29/96 \textasciitilde{} 30\% of VAST runs with new settings were better (70\% were worse).
+\emph{Differences Between VAST Settings}
+
+For YTF, 29/96 \textasciitilde{} 30\% of VAST runs with new settings were better (70\% were worse).
 80/96 \textasciitilde{} 83\% of scenarios had a VAST run with a better error than the stratified mean.
@@
 
-A more careful visual analysis of all error plots for each species reveals that VAST estimates tend to provide error estimates that are too high, while the stratified mean estimates are, on average, too low.
-Additionally, model/estimate ratio plots for all VAST estimates remain much closer to the desired value of 1 compared to stratified mean estimates, which can range as high as 10 and exhibit large yearly changes.
+\emph{Model / Estimate Ratio Results}
+
+A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, too low.
+This can be further examined by viewing estimate/ratio ratio plots, where we divide the yearly abundance estimates by the true model value.
+In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can range as low as 0 and exhibit large yearly changes.
+
+A more careful analysis of individual yearly model/estimate ratios when all strata are included revealed that about 73\% of all VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21.
+On the other hand, 33\% of stratified mean estimates were above 1 with an average of 0.874 and a standard deviation of 0.12.
+There were seasonal differences in estimate ratios with spring VAST ratios producing (1.08,0.12), fall VAST ratios being (1.50,0.38), stratified mean spring ratios resulting in (0.91,0.10), and fall values of (0.84,0.16).
+The breakdown for individual species followed a similar pattern.
+
+When the entire domain is sampled, adding covariates brings the estimate ratio closer to 1.
+However, when failing to sample the entire domain, adding covariates can decrease the accuracy of VAST estimates.
+When including all strata, the stratified mean produced (0.87,0.12), while adding covariates improved the VAST estimate ratio from (1.45,0.17) to (1.13,0.09).
+With a reduced number of strata, the stratified mean resulted in our worst summary values of (0.59,0.10), and adding covariates to our VAST models decreased the average ratio results from (1.04,0.33 ) to (0.78,0.19).
 

### 2023-03-15 TestScripts/Estimate_Ratio_summary.R

@@ -1,0 +1,130 @@
+#THIS SCRIPT READS IN CSV RELATED TO RATIO PLOTS AND CALCULATED SUMMARY INFORMATION
+
+
+ratio_data <- read.csv(file = paste(getwd(),"/VAST/ForPaper/All_Ratio_summary_info_new.csv",sep=""))
+
+short_names <- c("YTF","Cod","Had")
+
+summary_info <- list()
+
+
+#THESE DONT MAKE SENSE BECAUSE THEY MIX ALL STRATA WITH REDUCED STRATA
+# #mean and std dev of all vast
+# summary_info[["mean_std_all_VAST"]] <- c( mean(ratio_data[ratio_data[,"Estimate"]=="VAST","Mean_overall_finite"]) , sd(ratio_data[ratio_data[,"Estimate"]=="VAST","Mean_overall_finite"]))
+# #mean and std dev of all sm
+# summary_info[["mean_std_all_SM"]] <-c( mean(ratio_data[ratio_data[,"Estimate"]=="Strat Mean","Mean_overall_finite"]) , sd(ratio_data[ratio_data[,"Estimate"]=="Strat Mean","Mean_overall_finite"]))
+# 
+
+
+
+
+
+#USING ALL STRATA...
+#MEAN AND STD OF ALL VAST AND SM BY SEASON
+
+  for(E in c("VAST","Strat Mean")){
+    d <- ratio_data %>%
+      filter(Strata == "All", Estimate == E)
+    
+    ifelse(E=="Strat Mean", EE <- "SM", EE <- "VAST")
+    
+    summary_info[[paste("mean_std_ALL_spring_",EE,sep="")]] <- c(mean(d[,"Mean_Spring"]),sd(d[,"Mean_Spring"]))
+    summary_info[[paste("mean_std_ALL_fall_",EE,sep="")]] <- c(mean(d[,"Mean_Fall"]),sd(d[,"Mean_Fall"]))
+    
+  }
+
+
+#USING ALL STRATA...
+#MEAN AND STD OF ALL VAST AND SM BY SEASON AND SPECIES
+for(s in short_names){
+  for(E in c("VAST","Strat Mean")){
+d <- ratio_data %>%
+    filter(Species == s, Strata == "All", Estimate == E)
+
+ifelse(E=="Strat Mean", EE <- "SM", EE <- "VAST")
+
[…]

### 2023-03-15 TestScripts/VAST_StratMean_Error_Comparison_All_withCovs_ForPaper.R

@@ -9,13 +9,13 @@ orig.dir <- getwd()
 ##################################################################################################
 #THINGS WE NEED
 ##################################################################################################
-scenario1 <- "ConPop_ConTemp" #the folder name
+scenario1 <- "DecPop_IncTemp" #the folder name
 
 #spp1 spp2 spp3
 #short_names <- c("YT","Cod","Had")   #fixed above
-short_names <- c("YT")#,"Had")
+short_names <- c("YT","Cod")
 
-exclude_strata <- TRUE
+exclude_strata <- FALSE
 
 #for getting into correct subfolder
 ifelse(exclude_strata==TRUE, 
@@ -386,7 +386,7 @@ dir.create( paste0(getwd(),"/VAST/",scenario)) #create folder to store upcoming 
 
 ifelse(exclude_strata==TRUE, strat_ex <- "excludestrata", strat_ex <- "allstrata")
 
-saveRDS(strat_mean_all,paste0(getwd(),"/VAST/",scenario,"/strat_mean_all_",scenario1,"_",strat_ex,".RDS"))
+#saveRDS(strat_mean_all,paste0(getwd(),"/VAST/",scenario,"/strat_mean_all_",scenario1,"_",strat_ex,".RDS"))
 
 
 
@@ -548,7 +548,7 @@ for(folder in model_types[[s]]){
   colnames(FC_settings[[s]][[cov_direct]][[noise]][["fall"]]) <- c("Omega1","Epsilon1","Omega2","Epsilon2")
   }
 
-write.csv(FC_settings_DF,file = paste(getwd(),"/",scenario1,"_",str_dir,"_FC_settings.csv",sep=""))
+#write.csv(FC_settings_DF,file = paste(getwd(),"/",scenario1,"_",str_dir,"_FC_settings.csv",sep=""))
 
 
 
@@ -564,7 +564,7 @@ write.csv(FC_settings_DF,file = paste(getwd(),"/",scenario1,"_",str_dir,"_FC_set
 
 #All (potentially) species
 
-pdf(file=paste(getwd(),"/",scenario1,"_",str_dir,".pdf",sep=""))
[…]

### 2023-03-17 Manuscript/Draft1.Rmd
trying comment feature

@@
 
-[[Chris comments in double square brackets, search for them to see comments]]
+<!-- [[Chris comments in double square brackets, search for them to see comments]] -->
 
@@
 
-much of below is from https://apps-nefsc.fisheries.noaa.gov/nefsc/ecosystem-ecology/
-or
-https://www.fisheries.noaa.gov/data-tools/fisheries-economics-united-states-data-and-visualizations
+<!-- much of below is from https://apps-nefsc.fisheries.noaa.gov/nefsc/ecosystem-ecology/ -->
+<!-- or -->
+<!-- https://www.fisheries.noaa.gov/data-tools/fisheries-economics-united-states-data-and-visualizations -->
 
-- The eastern continental shelf is ecologically diverse and economically important
+<!-- - The eastern continental shelf is ecologically diverse and economically important -->
 
@@
 
-- Bottom trawl survey is important for monitoring population trends
+<!-- - Bottom trawl survey is important for monitoring population trends -->
 
@@
 
-- Climate change is happening
+<!-- - Climate change is happening -->
 
@@
 
-- Briefly describe our study to test this
+<!-- - Briefly describe our study to test this -->
 
@@
 
-- Describe simulation study
+<!-- - Describe simulation study -->
 
@@
 
--- Used MixFishSim.
+<!-- -- Used MixFishSim. -->
 
@@
 The *lrren* tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors.
-We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow [[I don't think we need to cite the trawl data because we said we were using trawl data above]] **BL: Does trawl data ever get cited?
-It was not cited above.
-cite trawl data?**.
-We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences.
+We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow  We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences.
 Depth and mean sediment size were used as our covariate predictors.
@@
 Figure \@ref(fig:hab-plot) depicts habitat preferences $Hab^2_{J,s}$ for each species.
+
+<!-- [[I don't think we need to cite the trawl data because we said we were using trawl data above]] **BL: Does trawl data ever get cited?
+It was not cited above.
+cite trawl data?**.
+Trawl data is now cited above, suggest removing this comment.
+-->
 
@@
 Figure \@ref(fig:temp-scenarios) depicts mean trends for the temperature scenarios used in our models.
-**dont forget to include gif in final submission**
+**dont forget to include animated gif in final submission**
 
@@
 
---Describe difference between increasing and constant temperature scenarios (images?)
+<!-- --Describe difference between increasing and constant temperature scenarios (images?) -->
 
@@
 
--- Describe each scenario that is considered
+<!-- -- Describe each scenario that is considered -->
 
@@
 
--Describe post hoc sampling process and how data is used
+<!-- -Describe post hoc sampling process and how data is used -->
 
@@
 
---Stratified mean vs VAST with and without covariates
+<!-- --Stratified mean vs VAST with and without covariates -->
 
@@
 
-[[ some things that I think we'll need to address, in addition to those mentioned above are (in no particular order):
-number of simulations done for each scenario - time required to do them and analyses
-pros and cons of using VAST with covariates in a changing environment
-implications for current practice of using stratified mean (not horrible, but might be able to do better with model-based using covariates if enviro changing, note we don't want to say stratified mean should not be used because that will draw a lot of tomato throwing)
-future work - some of the ideas we've discussed already could be mentioned
-limitations to our study
-perhaps for discussion, but we should mention somewhere that MFS only tracks biomass not age structure, the latter is important for stock assessments but not considered in this paper
-]]
+<!-- [[ some things that I think we'll need to address, in addition to those mentioned above are (in no particular order): -->
+<!-- number of simulations done for each scenario - time required to do them and analyses -->
+<!-- pros and cons of using VAST with covariates in a changing environment -->
+<!-- implications for current practice of using stratified mean (not horrible, but might be able to do better with model-based using covariates if enviro changing, note we don't want to say stratified mean should not be used because that will draw a lot of tomato throwing) -->
+<!-- future work - some of the ideas we've discussed already could be mentioned -->
+<!-- limitations to our study -->
+<!-- perhaps for discussion, but we should mention somewhere that MFS only tracks biomass not age structure, the latter is important for stock assessments but not considered in this paper -->
+<!-- ]] -->
 

### 2023-03-19 Manuscript/Draft1.Rmd
small updates

@@
 Figure \@ref(fig:hab-plot) depicts habitat preferences $Hab^2_{J,s}$ for each species.
-
-<!-- [[I don't think we need to cite the trawl data because we said we were using trawl data above]] **BL: Does trawl data ever get cited?
-It was not cited above.
-cite trawl data?**.
-Trawl data is now cited above, suggest removing this comment.
--->
 
@@
 ```{r paramsALL, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-library("knitr")
-library("kableExtra")
-library("dplyr")
-library("readr")
-library("tidyr")
-library("forcats")
 
@@
 ```{r paramsALL2,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-library("knitr")
-library("kableExtra")
-library("dplyr")
-library("readr")
-library("tidyr")
-library("forcats")
 
@@
 <!-- ```{r,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'} -->
-<!-- library("knitr") -->
-<!-- library("kableExtra") -->
-<!-- library("dplyr") -->
-<!-- library("readr") -->
-<!-- library("tidyr") -->
-<!-- library("forcats") -->
 
@@
 ```{r paramsSCENARIOS,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-library("knitr")
-library("kableExtra")
-library("dplyr")
-library("readr")
-library("tidyr")
-library("forcats")
 
@@
 <!-- ```{r YT_results,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'} -->
-<!-- library("knitr") -->
-<!-- library("kableExtra") -->
-<!-- library("dplyr") -->
-<!-- library("readr") -->
-<!-- library("tidyr") -->
-<!-- library("forcats") -->
 
@@
 ```{r YTresultsallstrata,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-library("knitr")
-library("kableExtra")
-library("dplyr")
-library("readr")
-library("tidyr")
-library("forcats")
 
@@
 ```{r YTresultsreducedstrata,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-library("knitr")
-library("kableExtra")
-library("dplyr")
-library("readr")
-library("tidyr")
-library("forcats")
 
@@
 ```{r CodERROR,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-library("knitr")
-library("kableExtra")
-library("dplyr")
-library("readr")
-library("tidyr")
-library("forcats")
 
@@
 ```{r HadERROR,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-library("knitr")
-library("kableExtra")
-library("dplyr")
-library("readr")
-library("tidyr")
-library("forcats")
 

### 2023-03-19 Manuscript/Draft1.tex
small updates

@@
 \section{Introduction}
-
-much of below is from \url{https://apps-nefsc.fisheries.noaa.gov/nefsc/ecosystem-ecology/}
-or
-\url{https://www.fisheries.noaa.gov/data-tools/fisheries-economics-united-states-data-and-visualizations}
-
-\begin{itemize}
-\tightlist
-\item
-  The eastern continental shelf is ecologically diverse and economically important
-\end{itemize}
 
@@
 Maintaining a healthy ecosystem is therefore vital to sustained ecological health and economic prosperity of the region.
-
-\begin{itemize}
-\tightlist
-\item
-  Bottom trawl survey is important for monitoring population trends
-\end{itemize}
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Climate change is happening
-\end{itemize}
-
 Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades, resulting in a shift in spatial distributions of many species.
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Briefly describe our study to test this
-\end{itemize}
-
 To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depend on temperature preferences.
@@
 \section{Methods}
-
-\begin{itemize}
-\tightlist
-\item
-  Describe simulation study
-\end{itemize}
 
@@
 \subsection{Population Model Formulation}
-
--- Used MixFishSim.
 
@@
 The \emph{lrren} tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors.
-We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow {[}{[}I don't think we need to cite the trawl data because we said we were using trawl data above{]}{]} \textbf{BL: Does trawl data ever get cited?
-It was not cited above.
-cite trawl data?}.
-We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences.
+We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences.
 Depth and mean sediment size were used as our covariate predictors.
@@
 Figure \ref{fig:temp-scenarios} depicts mean trends for the temperature scenarios used in our models.
-\textbf{dont forget to include gif in final submission}
+\textbf{dont forget to include animated gif in final submission}
 
@@
 \end{figure}
-
---Describe difference between increasing and constant temperature scenarios (images?)
 
@@
 
--- Describe each scenario that is considered
-
 We consider 20 year simulations under three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time.
@@
 \emph{Simulating Bottom Trawl Survey and Population Indexing}
-
--Describe post hoc sampling process and how data is used
 
@@
 Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.
-
---Stratified mean vs VAST with and without covariates
 
@@
 VAST may provide more consistent biomass predictions(?)
-
-{[}{[} some things that I think we'll need to address, in addition to those mentioned above are (in no particular order):
-number of simulations done for each scenario - time required to do them and analyses
-pros and cons of using VAST with covariates in a changing environment
-implications for current practice of using stratified mean (not horrible, but might be able to do better with model-based using covariates if enviro changing, note we don't want to say stratified mean should not be used because that will draw a lot of tomato throwing)
-future work - some of the ideas we've discussed already could be mentioned
-limitations to our study
-perhaps for discussion, but we should mention somewhere that MFS only tracks biomass not age structure, the latter is important for stock assessments but not considered in this paper{]}{]}
 

### 2023-03-20 Manuscript/Draft1.Rmd
add comments

@@
 
+<!-- Paragraph above: Stocks in this region are not managed by NEFSC, they are assessed by us.
+The management is done by the Councils and GARFO.
+NEFSC actually only assesses the federally managed stocks (not the state managed stocks like tautog).
+We also assess some of the invertebrate stocks, such as squid, scallops, ocean quahogs, etc., in addition to fish stocks.
+Perhaps easier to focus on the bottom trawl itself and replace the first two sentences with something like The NEFSC has conducted a bottom trawl survey since 1963 to support assessment and management of the fish and invertebrate populations in the region (azaorivitz and politis refs).
+-->
+
 <!-- - Climate change is happening -->
@@
 
+<!-- Agree that could bulk up the above paragraph with references from the initial proposal, along with more recent refs like https://doi.org/10.1016/j.fishres.2023.106652, DOI 10.1007/s11160-016-9444-z, https://doi.org/10.1093/icesjms/fsac140, etc.
+Don't want to lose connection with the next paragraph, so don't get lost in the weeds.
+-->
+
 <!-- - Briefly describe our study to test this -->
@@
 In both cases we analyze the ability of stratified random sampling to track population trends.
+
+<!-- General note: we should do a sweep through entire document and change everything to past tense.
+This is the typical verb tense in fisheries papers (I know present tense is used in other fields).
+In above paragraph, need to add VAST and VAST with covariates.
+-->
 
@@
 
+<!-- In the second sentence above, can we give a general size of the space occupied by each cell in our study?
+I know it varies by latitude, but I think an approximate size would be useful for readers to demonstrate that we are dealing with relatively small scale phenomena in the simulations.
+I stumbled over inhabit, inhabit, exist in the final sentence.
+I don't have a good suggestion for fixing it, just noting that the different final term caught my eye and made me ask if yellowtail were different from cod and haddock somehow.
+-->
+
 ```{r strata-plot, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Strata inhabited by each species in our population models."}
@@
 A full list of parameters used in our model can be seen below in Tables \@ref(paramsALL) and \@ref(tab:paramsSCENARIOS).
+
+<!-- I think we can delete the sentence about using Beverton-Holt recruitment because it appear in the table.  -->
 
@@
 
+<!-- I don't think this paragraph accurately represents what we did because we did the full factorial for yellowtail, but only some for cod and haddock.
+Need to think about how to restructure this paragraph to better reflect what we did.
+-->
+
 <font size="5">*Simulating Bottom Trawl Survey and Population Indexing*</font>
@@
 Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.
+
+<!-- Can we simplify the third and fourth sentences to say something like "Randomly selected cells were sampled each year."?
+I realize the current text is more accurate, but worry that we may be losing the readers in the details.
+-->
 
@@
 
+<!-- Note the Table XXX needs to be added comment above. -->
 When combining population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, covariates) there are a large number of scenario combinations to consider.
@@
 
-[[We'll want to expand the results section to focus on the impact of each factor one at at time within each species.
+<!-- [[We'll want to expand the results section to focus on the impact of each factor one at at time within each species.
 This will be a bit dull, but it is important to walk through the results in words to ensure that readers get the message.
-They can of course examine the tables in detail and draw their own conclusions, but we should put our interpretation down on paper.]]
+They can of course examine the tables in detail and draw their own conclusions, but we should put our interpretation down on paper.]] I think the above paragraphs are addressing this comment, but wanted to be sure they are not just "notes to self" before providing comments on them.
+-->
 
@@
 also should note that funding provided by Climate-Groundfish source (I'll dig up the official name of the funding source)]]
+
+<!-- Funding source is "Northeast Fisheries and Climate program at the NEFSC" -->
 
@@
 #param_table1 = read_csv("CSVs/Cod_results.csv")
-param_table1 = read_csv("CSVs/Cod_results_newsettings_forpaper.csv")
+param_table1 = read_csv("CSVs/Cod_results_newsettings_forpaper.csv") # not in repo
 
@@
 
-#param_table1 = read_csv("CSVs/Cod_results.csv")
-param_table1 = read_csv("CSVs/Haddock_results_newsettings_forpaper.csv")
+#param_table1 = read_csv("CSVs/Haddock_results.csv")
+param_table1 = read_csv("CSVs/Haddock_results_newsettings_forpaper.csv") # not in repo
 

### 2023-03-21 Manuscript/Draft1.tex
small updates to manuscript and improved survey plots

@@
 Might be having trouble tracking in to smaller strata again.
-
-{[}{[}We'll want to expand the results section to focus on the impact of each factor one at at time within each species.
-This will be a bit dull, but it is important to walk through the results in words to ensure that readers get the message.
-They can of course examine the tables in detail and draw their own conclusions, but we should put our interpretation down on paper.{]}{]}
 

### 2023-03-21 TestScripts/GB_habitat_create.R

@@ -1558,14 +1558,15 @@ Had_strata_ras <- raster::rasterize(GB_strata_Had,Had_ras,'STR2')
 GB_strata_ras <- list(YT_strata_ras,Cod_strata_ras,Had_strata_ras)
 names(GB_strata_ras) <- spp_names_short
 
-
+GB_strata_sp <- list(GB_strata_YT,GB_strata_Cod,GB_strata_Had)
+names(GB_strata_sp) <- spp_names_short
 
 
 
 
 
 #survey results without noise
-list_all_temp2 <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Simulation_Results\\",scenario,"\\list_all_more_",scenario,".RDS",sep=""))
+list_all_temp <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Simulation_Results\\",scenario,"\\list_all_",scenario,".RDS",sep=""))
 
 list_all <- list()
 list_all[["YT"]] <- list_all_temp[[good_iter[1]]]
@@ -1612,8 +1613,8 @@ for(i in seq(length(list_all[[s]][,1]))){
   survey_points[[s]][[yr]][[wk]][idx,"x"] <- as.numeric(list_all[[s]][i,"x"])
   survey_points[[s]][[yr]][[wk]][idx,"y"] <- as.numeric(list_all[[s]][i,"y"])
   survey_points[[s]][[yr]][[wk]][idx,"survey"] <- as.numeric(list_all[[s]][i,"stratum"])
-  survey_points[[s]][[yr]][[wk]][idx,"lon"] <- xFromCol(hab_ras, col = as.numeric(list_all[[s]][i,"y"]))
-  survey_points[[s]][[yr]][[wk]][idx,"lat"] <- yFromRow(hab_ras, row = as.numeric(list_all[[s]][i,"x"]))
+  survey_points[[s]][[yr]][[wk]][idx,"lon"] <- xFromCol(GB_strata_ras[[s]], col = as.numeric(list_all[[s]][i,"y"]))
+  survey_points[[s]][[yr]][[wk]][idx,"lat"] <- yFromRow(GB_strata_ras[[s]], row = as.numeric(list_all[[s]][i,"x"]))
   idx=idx+1
   }
   }
@@ -1696,7 +1697,7 @@ temp_max <- matrix(nrow=40,ncol=3)
 temp_color_min <- 0 
 
 
[…]

### 2023-03-24 Manuscript/Draft1.Rmd
small manuscript changes, many diagnostic changes

@@
 
-The package was designed to generate theoretical habitat preferences using Gaussian Random Fields that combine with hypothetical temperature gradients to drive the probability of movement from cell $I$ to cell $J$ using the formulation
+The package combines species-specific temperature tolerances with habitat preferences to drive the probability of movement from cell $I$ to cell $J$ using the formulation
 
@@
 
-The following sections describe how we formulated the habitat and temperature components to model real species on the northeast continental shelf.
+The package was designed to generate hypothetical temperature gradients and theoretical habitat preferences using Gaussian Random Fields.
+The following sections describe how we formulated the habitat and temperature components to model real species in the western Atlantic Ocean.
 
@@
 
-Species-specific habitat preferences were derived using the *lrren* tool from the R package *envi* [@envi] to create a niche model for each species.
+Species-specific habitat preferences were derived from niche model for each species using the *lrren* tool from the R package *envi* [@envi].
 The *lrren* tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors.
-We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow  We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences.
+We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow.
+We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences.
 Depth and mean sediment size were used as our covariate predictors.
@@
 The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method **cite arcmap** using point data collected by the United States Geologic Survey (USGS) [@mcmullen2005usgs].
-Since the values in $Hab^2_{J,s}$ are required to be between 0 and 1, we transform the spatial estimates from *lrren* to fall between these bounds.
+Since the values in $Hab_{J,s}$ are required to be between 0 and 1, we transform the spatial estimates from *lrren* to fall between these bounds.
 See Figure \@ref(fig:hab-plot1) for a visual representation of this process being applied to Cod.
-Figure \@ref(fig:hab-plot) depicts habitat preferences $Hab^2_{J,s}$ for each species.
+Figure \@ref(fig:hab-plot) depicts habitat preferences $Hab_{J,s}$ for each species.
 
@@
 Each species is assumed to have normally distributed temperature preferences ($N(\mu,\sigma)$).
+Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey.
 We assume Yellowtail Flounder's preferences are $N(8.75,4.25)$, while Haddock and Cod have preferences $N(9,4)$.
-We chose these values by combining information in the literature with temperatures recorded in the bottom trawl survey.
 Weekly estimated temperature data for the region for 2012 was obtained from FVCOM [@chen2006unstructured].

### 2023-03-24 Manuscript/Draft1.tex
small manuscript changes, many diagnostic changes

@@
 
-The package was designed to generate theoretical habitat preferences using Gaussian Random Fields that combine with hypothetical temperature gradients to drive the probability of movement from cell \(I\) to cell \(J\) using the formulation
+The package combines species-specific temperature tolerances with habitat preferences to drive the probability of movement from cell \(I\) to cell \(J\) using the formulation
 
@@
 
-Species-specific habitat preferences were derived using the \emph{lrren} tool from the R package \emph{envi} (Buller 2022) to create a niche model for each species.
+Species-specific habitat preferences were derived from niche model for each species using the \emph{lrren} tool from the R package \emph{envi} (Buller 2022).
 The \emph{lrren} tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors.
-We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences.
+We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow.
+We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences.
 Depth and mean sediment size were used as our covariate predictors.
@@
 n.d.).
-Since the values in \(Hab^2_{J,s}\) are required to be between 0 and 1, we transform the spatial estimates from \emph{lrren} to fall between these bounds.
+Since the values in \(Hab_{J,s}\) are required to be between 0 and 1, we transform the spatial estimates from \emph{lrren} to fall between these bounds.
 See Figure \ref{fig:hab-plot1} for a visual representation of this process being applied to Cod.
-Figure \ref{fig:hab-plot} depicts habitat preferences \(Hab^2_{J,s}\) for each species.
+Figure \ref{fig:hab-plot} depicts habitat preferences \(Hab_{J,s}\) for each species.
 
@@
 Each species is assumed to have normally distributed temperature preferences (\(N(\mu,\sigma)\)).
+Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey.
 We assume Yellowtail Flounder's preferences are \(N(8.75,4.25)\), while Haddock and Cod have preferences \(N(9,4)\).
-We chose these values by combining information in the literature with temperatures recorded in the bottom trawl survey.
 Weekly estimated temperature data for the region for 2012 was obtained from FVCOM (Chen et al.

### 2023-03-24 TestScripts/VAST_StratMean_Error_Comparison_All_withCovs_ForPaper.R

@@ -10,11 +10,11 @@ setwd(orig.dir)
 ##################################################################################################
 #THINGS WE NEED
 ##################################################################################################
-scenario1 <- "ConPop_ConTemp" #the folder name
+scenario1 <- "IncPop_IncTemp" #the folder name
 
 #spp1 spp2 spp3
 #short_names <- c("YT","Cod","Had")   #fixed above
-short_names <- c("YT")#,"Had")
+short_names <- c("YT","Had")
 
 exclude_strata <- FALSE
 
@@ -565,7 +565,7 @@ for(folder in model_types[[s]]){
 
 #All (potentially) species
 
-pdf(file=paste(getwd(),"/",scenario1,"_",str_dir,"_new.pdf",sep=""))
+pdf(file=paste(getwd(),"/",scenario1,"_",str_dir,"_newTEST.pdf",sep=""))
 
 #Just YT
 #pdf(file=paste(getwd(),"/YT/",scenario1,"_",str_dir,"_YT.pdf",sep=""))
@@ -582,7 +582,9 @@ Obsmodel_plot_log <- list()
 Est_ratio <- list()
 SRS_ratio <- list()
 Est_ratio_plot <- list()
+Est_ratio_plot2 <- list()
 Est_ratio_plot_log <- list()
+SRS_VAST_ratio <- list()
 
 Ratio_summary_info <- matrix(nrow=8*length(short_names),ncol=13)
 ratio_idx <- 1
@@ -644,12 +646,17 @@ for(folder in model_types[[s]]){
   
   oldnames = names(VAST_est[[s]][[cov_direct]][[noise]][[folder]][["spring"]])
   #add Est_ratio to VAST_est to plot later
-  VAST_est[[s]][[cov_direct]][[noise]][[folder]][["spring"]] <- cbind(VAST_est[[s]][[cov_direct]][[noise]][[folder]][["spring"]],Est_ratio[[s]][[cov_direct]][[noise]][[folder]][["spring"]])
-  names(VAST_est[[s]][[cov_direct]][[noise]][[folder]][["spring"]]) <- c(oldnames,"Est_ratio") 
+  VAST_est[[s]][[cov_direct]][[noise]][[folder]][["spring"]] <- cbind(VAST_est[[s]][[cov_direct]][[noise]][[folder]][["spring"]],
+                                                                      Est_ratio[[s]][[cov_direct]][[noise]][[folder]][["spring"]],
[…]

### 2023-03-27 Manuscript/Draft1.Rmd
small updates

@@
 
-Fish stocks in this highly productive and economically important region are managed by the National Oceanic and Atmospheric Administration’s (NOAA) Northeast Fisheries Science Center (NEFSC) in Woods Hole, Massachusetts.
-Federal scientists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 [@azarovitz1981brief; @politis2014northeast].
+The NEFSC has conducted a bottom trawl survey since 1963 to support assessment and management of the fish and invertebrate populations in the region [@azarovitz1981brief; @politis2014northeast].
 The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf.
@@
 Since most spatial analyses and  projections of future distributions typically assume a constant survey catchability and/or availability over time, NOAA's survey design includes sampling Georges Bank during approximately the same 3-4 week time period in each season.
-
-<!-- Paragraph above: Stocks in this region are not managed by NEFSC, they are assessed by us.
-The management is done by the Councils and GARFO.
-NEFSC actually only assesses the federally managed stocks (not the state managed stocks like tautog).
-We also assess some of the invertebrate stocks, such as squid, scallops, ocean quahogs, etc., in addition to fish stocks.
-Perhaps easier to focus on the bottom trawl itself and replace the first two sentences with something like The NEFSC has conducted a bottom trawl survey since 1963 to support assessment and management of the fish and invertebrate populations in the region (azaorivitz and politis refs).
--->
 

### 2023-03-27 Manuscript/Draft1.tex
small updates

@@
 
-Fish stocks in this highly productive and economically important region are managed by the National Oceanic and Atmospheric Administration's (NOAA) Northeast Fisheries Science Center (NEFSC) in Woods Hole, Massachusetts.
-Federal scientists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (Azarovitz 1981; Politis et al.
+The NEFSC has conducted a bottom trawl survey since 1963 to support assessment and management of the fish and invertebrate populations in the region (Azarovitz 1981; Politis et al.
 2014).

### 2023-03-28 Manuscript/Draft1.Rmd
small changes

@@
 
-Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades, resulting in a shift in spatial distributions of many species.
+Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades.
+The changes have resulted in a shift in spatial distributions of many species (cite Nye).
+Existing research has focused on temperature as the driver of such changes (cite klein, smith and kritzer) and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges (kerr, barajas, wiedenmann).
 Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments.
@@
 We consider the impact of climate change by comparing simulations that use a repeating temperature pattern to those where temperature increases on average over time.
-In both cases we analyze the ability of stratified random sampling to track population trends.
-
-<!-- General note: we should do a sweep through entire document and change everything to past tense.
-This is the typical verb tense in fisheries papers (I know present tense is used in other fields).
-In above paragraph, need to add VAST and VAST with covariates.
--->
+In both cases we analyze the ability of stratified random sampling to track population trends through a design-based approach (stratified mean) compared to a model-based approach that allows for the inclusion of covariates (Vector Autoregressive Spatiotemporal Model).
 
@@
 The region was discritized into a raster with 88 rows and 144 columns to use as our modeling environment.
+Each cell in our simulation domain represents approximately 8.7 km$^2$.
 A fish stock is considered to be a subpopulation of a species that has similar intrinsic parameters.
@@
 Biological differences between species results in each stock inhabiting a different number of strata on Georges Bank.
-Haddock inhabit all 15 strata in the domain, Cod inhabit 13 strata, and Yellowtail exist in 9 strata.
+Haddock inhabit all 15 strata in the domain, Cod populate 13 strata, and Yellowtail can be found in 9 strata.
 Figure \@ref(fig:strata-plot) shows the regions used in our models.
-
-<!-- In the second sentence above, can we give a general size of the space occupied by each cell in our study?
-I know it varies by latitude, but I think an approximate size would be useful for readers to demonstrate that we are dealing with relatively small scale phenomena in the simulations.
-I stumbled over inhabit, inhabit, exist in the final sentence.
-I don't have a good suggestion for fixing it, just noting that the different final term caught my eye and made me ask if yellowtail were different from cod and haddock somehow.
--->
 
@@
 Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits.
-We chose to represent recruitment in the model using a Beverton-Holt formulation **cite**.
 Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.
@@
 A full list of parameters used in our model can be seen below in Tables \@ref(paramsALL) and \@ref(tab:paramsSCENARIOS).
-
-<!-- I think we can delete the sentence about using Beverton-Holt recruitment because it appear in the table.  -->
 
@@
 
-We consider 20 year simulations under three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time.
-Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius over the duration of the 20 year simulation.
-We therefore simulate 6 scenarios for each population.
+We carry out 20 year simulations for each of our three species under various population scenarios.
+Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year [[can cite the 2022 management track assessments, see https://apps-nefsc.fisheries.noaa.gov/saw/sari.php for when the document becomes available]] **cite**.
+For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
+To provide a comprehensive analysis of population indexing methods we consider all possible scenario combinations for Yellowtail Flounder.
+Each of these scenarios is simulated twice: first with with an oscillating temperature gradient that repeats and second with a temperature gradient that increases roughly 5 degrees Celsius over the duration of the 20 year simulation, for a total of 10 simulated spatial datasets.
+The specific population trends used in our analyses can be see in Figure \@ref(fig:pop-scenarios).
 
-<!-- I don't think this paragraph accurately represents what we did because we did the full factorial for yellowtail, but only some for cod and haddock.
-Need to think about how to restructure this paragraph to better reflect what we did.
--->
+```{r pop-scenarios, echo=FALSE,out.width="99%",fig.show='hold', fig.align='center', fig.cap="True population trends used in indexing analyses.
+Spring biomass plots are shown with fall values being very similar."}
+knitr::include_graphics(c("Images/Population_scenarios.png"))
+```
 
@@
 We sample each strata in the same weeks in which the Spring and Fall surveys take place, and the number of the samples taken reflect true target values for each strata.
-Most strata contain enough cells that we were able to generate a list of sampling locations without replacement, allowing us to sample a unique random cell in each survey over the duration of the simulation.
-For strata that contain small numbers of cells, we generate a random sampling locations with replacement of cells previously sampled.
+Randomly selected cells were samples each year.
 We then use the biomass collected from our samples in contemporary population indexing methods to estimate population trends.
 Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.
-
-<!-- Can we simplify the third and fourth sentences to say something like "Randomly selected cells were sampled each year."?
-I realize the current text is more accurate, but worry that we may be losing the readers in the details.
--->
 
@@
 The goal of our project was to analyze the ability of contemporary population indexing methods to track population trends under a variety of conditions, as depicted in Table \@ref(tab:scenarios).
-Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year [[can cite the 2022 management track assessments, see https://apps-nefsc.fisheries.noaa.gov/saw/sari.php for when the document becomes available]] **cite**.
-For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
-To provide a comprehensive analysis of population indexing methods we consider all possible scenario combinations for Yellowtail Flounder.
-The specific population trends used in our analyses can be see in Figure \@ref(fig:pop-scenarios).
-
-```{r pop-scenarios, echo=FALSE,out.width="99%",fig.show='hold', fig.align='center', fig.cap="True population trends used in indexing analyses.
-Spring biomass plots are shown with fall values being very similar."}
-knitr::include_graphics(c("Images/Population_scenarios.png"))
-```
 
@@
 [[we should thank Jim (obviously), but also those who helped you with the habitat data (robyn, david Chev.) and others?
-also should note that funding provided by Climate-Groundfish source (I'll dig up the official name of the funding source)]]
-
-<!-- Funding source is "Northeast Fisheries and Climate program at the NEFSC" -->
+also should note that funding provided by Northeast Fisheries and Climate program at the NEFSC source (I'll dig up the official name of the funding source)]]
 

### 2023-03-28 Manuscript/Draft1.tex
small changes

@@
 The region was discritized into a raster with 88 rows and 144 columns to use as our modeling environment.
+Each cell in our simulation domain represents approximately 8.7 km\(^2\).
 A fish stock is considered to be a subpopulation of a species that has similar intrinsic parameters.
@@
 Biological differences between species results in each stock inhabiting a different number of strata on Georges Bank.
-Haddock inhabit all 15 strata in the domain, Cod inhabit 13 strata, and Yellowtail exist in 9 strata.
+Haddock inhabit all 15 strata in the domain, Cod populate 13 strata, and Yellowtail can be found in 9 strata.
 Figure \ref{fig:strata-plot} shows the regions used in our models.
@@
 Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits.
-We chose to represent recruitment in the model using a Beverton-Holt formulation \textbf{cite}.
 Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.
@@
 
-We consider 20 year simulations under three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time.
-Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius over the duration of the 20 year simulation.
-We therefore simulate 6 scenarios for each population.
+We carry out 20 year simulations for each of our three species under various population scenarios.
+Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year {[}{[}can cite the 2022 management track assessments, see \url{https://apps-nefsc.fisheries.noaa.gov/saw/sari.php} for when the document becomes available{]}{]} \textbf{cite}.
+For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
+To provide a comprehensive analysis of population indexing methods we consider all possible scenario combinations for Yellowtail Flounder.
+Each of these scenarios is simulated twice: first with with an oscillating temperature gradient that repeats and second with a temperature gradient that increases roughly 5 degrees Celsius over the duration of the 20 year simulation, for a total of 10 simulated spatial datasets.
+The specific population trends used in our analyses can be see in Figure \ref{fig:pop-scenarios}.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.99\linewidth]{Images/Population_scenarios}
+
+}
+
+\caption{True population trends used in indexing analyses. Spring biomass plots are shown with fall values being very similar.}\label{fig:pop-scenarios}
+\end{figure}
 
@@
 We sample each strata in the same weeks in which the Spring and Fall surveys take place, and the number of the samples taken reflect true target values for each strata.
-Most strata contain enough cells that we were able to generate a list of sampling locations without replacement, allowing us to sample a unique random cell in each survey over the duration of the simulation.
-For strata that contain small numbers of cells, we generate a random sampling locations with replacement of cells previously sampled.
+Randomly selected cells were samples each year.
 We then use the biomass collected from our samples in contemporary population indexing methods to estimate population trends.
@@
 The goal of our project was to analyze the ability of contemporary population indexing methods to track population trends under a variety of conditions, as depicted in Table \ref{tab:scenarios}.
-Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year {[}{[}can cite the 2022 management track assessments, see \url{https://apps-nefsc.fisheries.noaa.gov/saw/sari.php} for when the document becomes available{]}{]} \textbf{cite}.
-For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
-To provide a comprehensive analysis of population indexing methods we consider all possible scenario combinations for Yellowtail Flounder.
-The specific population trends used in our analyses can be see in Figure \ref{fig:pop-scenarios}.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.99\linewidth]{Images/Population_scenarios}
-
-}
-
-\caption{True population trends used in indexing analyses. Spring biomass plots are shown with fall values being very similar.}\label{fig:pop-scenarios}
-\end{figure}
 
@@
 {[}{[}we should thank Jim (obviously), but also those who helped you with the habitat data (robyn, david Chev.) and others?
-also should note that funding provided by Climate-Groundfish source (I'll dig up the official name of the funding source){]}{]}
+also should note that funding provided by Northeast Fisheries and Climate program at the NEFSC source (I'll dig up the official name of the funding source){]}{]}
 

### 2023-04-18 Manuscript/Draft1.tex
small manuscript updates

@@
 We consider the impact of climate change by comparing simulations that use a repeating temperature pattern to those where temperature increases on average over time.
-In both cases we analyze the ability of stratified random sampling to track population trends.
+In both cases we analyze the ability of stratified random sampling to track population trends through a design-based approach (stratified mean) compared to a model-based approach that allows for the inclusion of covariates (Vector Autoregressive Spatiotemporal Model).
 

### 2023-04-18 TestScripts/Calc_strat_mean/fn_srs_survey_BENS.R
updates to plotting files and stratified mean by stratum

@@ -43,7 +43,7 @@ srs_survey <- function(df, sa, str, ta=1, sppname = NULL  )  {
     dplyr::summarise(Total=sum(STRATUM_AREA))
   
   #DAY COLUMN IN DF IS CAUSING ISSUES AND NOT NEEDED
-  df <- subset(df,select = -c(day))
+  if(!is.null(df$day)){df <- subset(df,select = -c(day))}
 
   tmp.tibble <- df %>%
   #  filter(substr(SURVEY, 1, 4) =="NMFS") %>% #STILL DONT THINK I NEED THIS
@@ -72,7 +72,17 @@ surv.ind.yr <- surv.ind.str %>%
   #remove chracter Season so we can summarize with Reduce
   surv.ind.yr <- subset(surv.ind.yr,select = -c(Season))
 
-return(surv.ind.yr)
+  #calculate estimate by strata
+  mean.yr.strrr <- surv.ind.str %>%
+    left_join(sa,by="stratum") %>%
+    left_join(tmp.total.area,by="Season") %>%
+    mutate(mean.yr.str = (mean.str*STRATUM_AREA) ) %>%
+    group_by(year, Season) %>% #INDEX GOING DOWN TO YEAR?
+    dplyr::summarise(mean.yr.strr = mean.yr.str, stratum=stratum,season = ifelse(Season=="SPRING",1,2))
+  
+  out_put <- list(surv.ind.yr,surv.ind.str, mean.yr.strrr)
+  names(out_put) <- c("surv.ind.yr","surv.ind.str", "mean.yr.strrr")
+return(out_put)
 
 } # end function srs_nefsc
 

### 2023-04-18 TestScripts/Plots_for_paper_presentation.R

@@ -10,9 +10,11 @@ model_point_size = 3.5
 estimate_line_size = 1.25
 estimate_point_size = 3
 
-season_ = 2
+for(season_ in c(1,2)){
 
-ggplot() +
+ ssnn = ifelse(season_==1, "Spring Estimate", "Fall Estimate")
+  
+print(ggplot() +
   
   
   #this way plots data by season
@@ -70,20 +72,20 @@ ggplot() +
   #                                          "  FC=", toString(FC_fall),
   #                                          "  FeSM=",round(SRS_Model_error[[s]][[folder]][["fall"]],digits=2),sep=""), color ="" )
   
-  labs(x="year",y="Biomass", title ="Yellowtail Flounder", color ="" )+
+  labs(x="year",y="Biomass", title =paste("Yellowtail Flounder ",ssnn,sep=""), color ="" )+
   
   theme(axis.text=element_text(size=12),
         axis.title=element_text(size=12),
         title=element_text(size=14))
+)
+
+}
 
 
 
 
 
-
-
-
-
+#
 
 
 
@@ -219,40 +221,42 @@ model_point_size = 3.5
 estimate_line_size = 1.25
 estimate_point_size = 3
 
-season_ = 2
+for(season_ in c(1,2)){
+  
+  ssnn = ifelse(season_==1, "Spring", "Fall")
 
-
+print(
 ggplot() +
   
   
   #this way plots data by season divided by itself so it equals 1
-  geom_point(data = subset(as.data.frame(pop_by_season[[s]]),season==season_), aes(x=as.numeric(year),y=Est_ratio, group = season, color = "Model"),size=estimate_point_size) +
-  geom_line(data = subset(as.data.frame(pop_by_season[[s]]),season==season_), aes(x=as.numeric(year),y=Est_ratio, group =season, color = "Model"),size=estimate_line_size ) +
+  geom_point(data = subset(as.data.frame(pop_by_season[[s]]),season==season_), aes(x=as.numeric(year),y=log(Est_ratio), group = season, color = "Model"),size=estimate_point_size) +
+  geom_line(data = subset(as.data.frame(pop_by_season[[s]]),season==season_), aes(x=as.numeric(year),y=log(Est_ratio), group =season, color = "Model"),size=estimate_line_size ) +
   
[…]

### 2023-04-18 TestScripts/VAST_StratMean_Error_Comparison_All_withCovs_ForPaper.R

@@ -10,7 +10,7 @@ setwd(orig.dir)
 ##################################################################################################
 #THINGS WE NEED
 ##################################################################################################
-scenario1 <- "IncPop_IncTemp" #the folder name
+scenario1 <- "IncPop_ConTemp" #the folder name
 
 #spp1 spp2 spp3
 #short_names <- c("YT","Cod","Had")   #fixed above
@@ -127,6 +127,7 @@ library(TMB)
 library(VAST)
 library(dplyr)
 library(ggplot2)
+library(rgdal)
 
 #read in habitat matrix
 hab <- readRDS(file="hab_GB_3species.RDS") #courser resolution
@@ -139,7 +140,36 @@ Had_ras <- readRDS(file="TestScripts/Habitat_plots/Haddock/Had_Weighted_AdaptFal
 #translate habitat matrix back into raster
 hab_ras <-raster(hab$hab$spp3)
 extent(hab_ras) <- extent(Had_ras)
+crs(hab_ras) <- crs(Had_ras)
 #plot(hab_ras)
+
+#read in species-specific polygons
+#load stratas for clipping etc
+strata.dir <- "C:\\Users\\benjamin.levy\\Desktop\\NOAA\\GIS_Stuff\\" # strata shape files in this directory
+# get the shapefiles
+strata.areas <- readOGR(paste(strata.dir,"Survey_strata", sep="")) #readShapePoly is deprecated; use rgdal::readOGR or sf::st_read 
+#define georges bank for YELLOWTAIL
+GB_strata_num <- c("01130","01140","01150","01160","01170","01180","01190","01200","01210")
+#pull out indices corresponding to GB strata
+GB_strata_idx <- match(GB_strata_num,strata.areas@data[["STRATUMA"]])
+GB_strata_YT <- strata.areas[GB_strata_idx,]
+#define georges bank for COD  
+GB_strata_num <- c("01130","01140","01150","01160","01170","01180","01190","01200","01210","01220","01230","01240","01250")
+#pull out indices corresponding to GB strata
[…]

### 2023-04-19 Manuscript/Draft1.Rmd
manuscipt updates

@@
 The goal of our project was to analyze the ability of contemporary population indexing methods to track population trends under a variety of conditions, as depicted in Table \@ref(tab:scenarios).
+See Figure \@ref(fig:strata-plot) for a spatial reference of the Georges Bank strata and Figure \@ref(fig:hab-plot) for the static habitat preferences for each species ($Hab^2_{J,s}$) used in our population models.
+
+Figure \@ref(fig:PopPct) depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey weeks in the spring (13 and 14) and fall (37 and 28).
+The left column depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern.
+In these scenarios we notice a small amount of shifting between successive years as the population aggregates on hot spots in the domain.
+On the other hand, in the right column where the temperature is increasing over time we note significant shifting taking places in certain strata.
+This figure shows how we were able to mimic shifting distributions due to climate change by running simulations with a temperature gradient that increased on average over time.
 
 Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the results of our comparison between the absolute error between abundance estimates and model output.
-General themes are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors and standard deviations.
-On the contrary, we also see individual cases where the stratified mean produced the lowest absolute error.
+While the model-based results tended to provide a lower relative error to the stratified mean with a mean relative error of 0.34 compared to 0.38, the model-based estimates were very sensitive to the settings used.
+As a result, although $132/160=82.5\%$ of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the variance for the VAST errors was much larger than the stratified mean (0.10 compared to 0.02).
+VAST models that include covariate information provided the lowest overall errors and standard deviations.
 When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.
 Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain.
-
-In considering the Yellowtail Flounder results in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata), we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
-When all strata are included on sampling, adding covariates greatly improved estimates in all scenarios.
-We still see an improvement in VAST estimates when certain strata are excluded from sampling, but the improvement was much less dramatic than with all strata included.
-However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata).
-These errors are corrected by including covariate information resulting in improved VAST estimates that are significantly lower than their stratified mean counterparts.
-
-(see Table \@ref(tab:CodERROR)) In the Cod results, when using reduced strata in constant temperature, adding covariates produces worse VAST results (though still better than stratified mean).
-This is similar to what we see with all of Haddock results (adding covariates makes it worse with reduced strata).
-This is NOT seen with Yellowtail (adding covariates always helps).
-VAST without covariates much worse than stratified mean in fall with increasing temperature, but adding covariates corrects this and makes VAST estimates better than stratified mean.
-
-(see Table \@ref(tab:HadERROR)) For Haddock, VAST has a particularly hard time in spring regularly producing larger errors than the stratified mean with added covariates only improving to the level of the stratified mean.
-VAST shows improved results in fall relative to the stratified mean with added covariates producing extremely low errors in some cases.
-In Haddock results, adding covariates improves estimates only when all strata are included (when reduced strata, adding covariates actually makes VAST worse).
-Since this occurs in all scenarios, it seems to again be related to failing to accurately monitor strata 17.
 
@@
 
-Figure \@ref(fig:PopPct) depicts the spatial shifting that occurs in each strata within our population models, specifically during the bottom trawl survey weeks in the spring (13 and 14) and fall (37 and 28).
-See Figure \@ref(fig:strata-plot) for a spatial reference of the Georges Bank strata and Figure \@ref(fig:hab-plot) for the static habitat preferences for each species ($Hab^2_{J,s}$) used in our population models.
-
 <font size="5">*Yellowtail Flounder Results*</font>
@@
 These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail Flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12.
-These spring dynamics are therefore related to the static habitat values in our model rather than the dynamics temperature preferences.
+These spring dynamics are therefore related to the static habitat values in our model rather than the dynamics temperature preferences, wihch is why we seen the same dynamics with constant and increasing temperature.
 While we observe similar dynamics in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \@ref(fig:PopPct)).
@@
 One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how ~10\% of the population exists in the narrow exterior strata 17 in the fall, while seemingly none of the population exists in any of the exterior strata (14, 15, 17, 18) in the spring.
+
+In considering the Yellowtail Flounder results in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata), we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
+When all strata are sampled, adding covariates greatly improved estimates in all scenarios.
+We still see an improvement in VAST estimates when certain strata are excluded from sampling, but the improvement was much less dramatic than with all strata included.
+However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata).
+Each of these instances are These errors are corrected by including covariate information resulting in improved VAST estimates that are significantly lower than their stratified mean counterparts.
 
@@
 
+(see Table \@ref(tab:CodERROR)) In the Cod results, when using reduced strata in constant temperature, adding covariates produces worse VAST results (though still better than stratified mean).
+This is similar to what we see with all of Haddock results (adding covariates makes it worse with reduced strata).
+This is NOT seen with Yellowtail (adding covariates always helps).
+VAST without covariates much worse than stratified mean in fall with increasing temperature, but adding covariates corrects this and makes VAST estimates better than stratified mean.
+
 Of the abundance estimates without covariates, $12/16=75$\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
@@
 Since stratum 16 contains very favorable habitat including much of the species' spawning ground, the strong shift out of 16 and into the northern most strata of 29 in the fall demonstrates a climate-driven change in movement preference.
+
+(see Table \@ref(tab:HadERROR)) For Haddock, VAST has a particularly hard time in spring regularly producing larger errors than the stratified mean with added covariates only improving to the level of the stratified mean.
+VAST shows improved results in fall relative to the stratified mean with added covariates producing extremely low errors in some cases.
+In Haddock results, adding covariates improves estimates only when all strata are included (when reduced strata, adding covariates actually makes VAST worse).
+Since this occurs in all scenarios, it seems to again be related to failing to accurately monitor strata 17.
 
@@
 
-<div align="center">Table 1. Parameters used in all population models.</div>
+<div align="center">Table 1. Parameters used in all population models. SAW 1: [@nefsc1218], SAW 2: [@nefsc1311], SAW 3: ??? </div>
 
@@
 |:---	|:------- |:-- |:--- |:-- |:-- |:--
-| $\\rho$ | Ford's growth coefficient | wk$^{-1}$  | 4.48 | 4.43 | 4.49 | |
-| $M$ | Natural Mortality| wk$^{-1}$  | 0.2064 | 0.2728 | 0.3340 | |
-| $W_R$ |Weight of fully recruited fish| kg | 0.39 | 2.95 | 1.12 | |
-| $W_{R-1}$ | Weight of pre-recruit fish| kg | 0.13 | 0.39 | 0.19 | |
-| $\\sigma^2$ | Variance in recruited fish | kg$^2$ | 0.55 | 0.55 | 0.55 ||
-| $\\lambda$ | Decay rate for movement | - | 0.7 | 0.7 | 0.7 ||
-| $Spwn_s$ | Spawning weeks for species $s$ | wk | 9-12 | 8-13 | 11-14 ||
-| $Rec_s$ | Recruitment weeks for species $s$ | wk | 9-12 | 8-13 | 11-14 ||
+| $\\rho$ | Ford's growth coefficient | wk$^{-1}$  | 4.48 | 4.43 | 4.49 | [@thorson2020predicting] |
+| $M$ | Natural Mortality| wk$^{-1}$  | 0.2064 | 0.2728 | 0.3340 | [@thorson2020predicting]|
+| $F$ | Fishing Mortality| wk$^{-1}$  | 0.358 | 0.511 | 0.45 | SAW 1, 2, 3 |
+| $W_R$ |Weight of fully recruited fish| kg | 0.39 | 2.95 | 1.12 | SAW 1, 2, 3 |
+| $W_{R-1}$ | Weight of pre-recruit fish| kg | 0.13 | 0.39 | 0.19 | SAW 1, 2, 3 |
+| $\\sigma^2$ | Variance in recruited fish | kg$^2$ | 0.55 | 0.55 | 0.55 |assumed|
+| $\\lambda$ | Decay rate for movement | - | 0.7 | 0.7 | 0.7 |assumed|
+| $Spwn_s$ | Spawning weeks for species $s$ | wk | 9-12 | 8-13 | 11-14 |SAW 1, 2, 3|
+| $Rec_s$ | Recruitment weeks for species $s$ | wk | 9-12 | 8-13 | 11-14 |SAW 1, 2, 3|
 "

### 2023-04-19 Manuscript/Draft1.tex
manuscipt updates

@@
 The goal of our project was to analyze the ability of contemporary population indexing methods to track population trends under a variety of conditions, as depicted in Table \ref{tab:scenarios}.
+See Figure \ref{fig:strata-plot} for a spatial reference of the Georges Bank strata and Figure \ref{fig:hab-plot} for the static habitat preferences for each species (\(Hab^2_{J,s}\)) used in our population models.
+
+Figure \ref{fig:PopPct} depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey weeks in the spring (13 and 14) and fall (37 and 28).
+The left column depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern.
+In these scenarios we notice a small amount of shifting between successive years as the population aggregates on hot spots in the domain.
+On the other hand, in the right column where the temperature is increasing over time we note significant shifting taking places in certain strata.
+This figure shows how we were able to mimic shifting distributions due to climate change by running simulations with a temperature gradient that increased on average over time.
 
 Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the results of our comparison between the absolute error between abundance estimates and model output.
-General themes are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors and standard deviations.
-On the contrary, we also see individual cases where the stratified mean produced the lowest absolute error.
+While the model-based results tended to provide a lower relative error to the stratified mean with a mean relative error of 0.34 compared to 0.38, the model-based estimates were very sensitive to the settings used.
+As a result, although \(132/160=82.5\%\) of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the variance for the VAST errors was much larger than the stratified mean (0.10 compared to 0.02).
+VAST models that include covariate information provided the lowest overall errors and standard deviations.
 When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.
 Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain.
-
-In considering the Yellowtail Flounder results in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}, we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
-When all strata are included on sampling, adding covariates greatly improved estimates in all scenarios.
-We still see an improvement in VAST estimates when certain strata are excluded from sampling, but the improvement was much less dramatic than with all strata included.
-However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}.
-These errors are corrected by including covariate information resulting in improved VAST estimates that are significantly lower than their stratified mean counterparts.
-
-(see Table \ref{tab:CodERROR}) In the Cod results, when using reduced strata in constant temperature, adding covariates produces worse VAST results (though still better than stratified mean).
-This is similar to what we see with all of Haddock results (adding covariates makes it worse with reduced strata).
-This is NOT seen with Yellowtail (adding covariates always helps).
-VAST without covariates much worse than stratified mean in fall with increasing temperature, but adding covariates corrects this and makes VAST estimates better than stratified mean.
-
-(see Table \ref{tab:HadERROR}) For Haddock, VAST has a particularly hard time in spring regularly producing larger errors than the stratified mean with added covariates only improving to the level of the stratified mean.
-VAST shows improved results in fall relative to the stratified mean with added covariates producing extremely low errors in some cases.
-In Haddock results, adding covariates improves estimates only when all strata are included (when reduced strata, adding covariates actually makes VAST worse).
-Since this occurs in all scenarios, it seems to again be related to failing to accurately monitor strata 17.
 
@@
 
-Figure \ref{fig:PopPct} depicts the spatial shifting that occurs in each strata within our population models, specifically during the bottom trawl survey weeks in the spring (13 and 14) and fall (37 and 28).
-See Figure \ref{fig:strata-plot} for a spatial reference of the Georges Bank strata and Figure \ref{fig:hab-plot} for the static habitat preferences for each species (\(Hab^2_{J,s}\)) used in our population models.
-
 \emph{Yellowtail Flounder Results}
@@
 These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail Flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12.
-These spring dynamics are therefore related to the static habitat values in our model rather than the dynamics temperature preferences.
+These spring dynamics are therefore related to the static habitat values in our model rather than the dynamics temperature preferences, wihch is why we seen the same dynamics with constant and increasing temperature.
 While we observe similar dynamics in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \ref{fig:PopPct}).
@@
 One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how \textasciitilde10\% of the population exists in the narrow exterior strata 17 in the fall, while seemingly none of the population exists in any of the exterior strata (14, 15, 17, 18) in the spring.
+
+In considering the Yellowtail Flounder results in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}, we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
+When all strata are sampled, adding covariates greatly improved estimates in all scenarios.
+We still see an improvement in VAST estimates when certain strata are excluded from sampling, but the improvement was much less dramatic than with all strata included.
+However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}.
+Each of these instances are These errors are corrected by including covariate information resulting in improved VAST estimates that are significantly lower than their stratified mean counterparts.
 
@@
 
+(see Table \ref{tab:CodERROR}) In the Cod results, when using reduced strata in constant temperature, adding covariates produces worse VAST results (though still better than stratified mean).
+This is similar to what we see with all of Haddock results (adding covariates makes it worse with reduced strata).
+This is NOT seen with Yellowtail (adding covariates always helps).
+VAST without covariates much worse than stratified mean in fall with increasing temperature, but adding covariates corrects this and makes VAST estimates better than stratified mean.
+
 Of the abundance estimates without covariates, \(12/16=75\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
@@
 Since stratum 16 contains very favorable habitat including much of the species' spawning ground, the strong shift out of 16 and into the northern most strata of 29 in the fall demonstrates a climate-driven change in movement preference.
+
+(see Table \ref{tab:HadERROR}) For Haddock, VAST has a particularly hard time in spring regularly producing larger errors than the stratified mean with added covariates only improving to the level of the stratified mean.
+VAST shows improved results in fall relative to the stratified mean with added covariates producing extremely low errors in some cases.
+In Haddock results, adding covariates improves estimates only when all strata are included (when reduced strata, adding covariates actually makes VAST worse).
+Since this occurs in all scenarios, it seems to again be related to failing to accurately monitor strata 17.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-nefsc1218}{}}%
+NEFSC.
+2012.
+{54th Northeast Regional Stock Assessment Workshop (54th SAW) Assessment Report}.
+{Northeast Fish.
+Sci.
+Cent.
+Ref.
+Doc.
+12-18; 604 p.
+Available from: National Marine Fisheries Service, 166 Water Street, Woods Hole, MA 02543-1026, or online at http://www.nefsc.noaa.gov/nefsc/publications/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nefsc1311}{}}%
+NEFSC.
+2013.
+{55th Northeast Regional Stock Assessment Workshop (55th SAW) Assessment Report}.
+{Northeast Fish.
+Sci.
+Cent.
+Ref.
+Doc.
+13-11; 849 p.
+Available from: National Marine Fisheries Service, 166 Water Street, Woods Hole, MA 02543-1026, or online at http://www.nefsc.noaa.gov/nefsc/publications/}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-politis2014northeast}{}}%
@@
 Elsevier.
+
+\leavevmode\vadjust pre{\hypertarget{ref-thorson2020predicting}{}}%
+Thorson, J.T.
+2020.
+Predicting recruitment density dependence and intrinsic growth rate for all fishes worldwide using a data-integrated life-history model.
+Fish and Fisheries \textbf{21}(2): 237--251.
+Wiley Online Library.
 
@@
 
-Table 1. Parameters used in all population models.
+Table 1. Parameters used in all population models. SAW 1: (NEFSC 2012), SAW 2: (NEFSC 2013), SAW 3: ???
 
@@
 \endhead
-\(\rho\) & Ford's growth coefficient & wk\(^{-1}\) & 4.48 & 4.43 & 4.49 & \\
-\(M\) & Natural Mortality & wk\(^{-1}\) & 0.2064 & 0.2728 & 0.3340 & \\
-\(W_R\) & Weight of fully recruited fish & kg & 0.39 & 2.95 & 1.12 & \\
-\(W_{R-1}\) & Weight of pre-recruit fish & kg & 0.13 & 0.39 & 0.19 & \\
-\(\sigma^2\) & Variance in recruited fish & kg\(^2\) & 0.55 & 0.55 & 0.55 & \\
-\(\lambda\) & Decay rate for movement & - & 0.7 & 0.7 & 0.7 & \\
-\(Spwn_s\) & Spawning weeks for species \(s\) & wk & 9-12 & 8-13 & 11-14 & \\
-\(Rec_s\) & Recruitment weeks for species \(s\) & wk & 9-12 & 8-13 & 11-14 & \\
+\(\rho\) & Ford's growth coefficient & wk\(^{-1}\) & 4.48 & 4.43 & 4.49 & (Thorson 2020) \\
+\(M\) & Natural Mortality & wk\(^{-1}\) & 0.2064 & 0.2728 & 0.3340 & (Thorson 2020) \\
+\(F\) & Fishing Mortality & wk\(^{-1}\) & 0.358 & 0.511 & 0.45 & SAW 1, 2, 3 \\
+\(W_R\) & Weight of fully recruited fish & kg & 0.39 & 2.95 & 1.12 & SAW 1, 2, 3 \\
+\(W_{R-1}\) & Weight of pre-recruit fish & kg & 0.13 & 0.39 & 0.19 & SAW 1, 2, 3 \\
+\(\sigma^2\) & Variance in recruited fish & kg\(^2\) & 0.55 & 0.55 & 0.55 & assumed \\
+\(\lambda\) & Decay rate for movement & - & 0.7 & 0.7 & 0.7 & assumed \\
+\(Spwn_s\) & Spawning weeks for species \(s\) & wk & 9-12 & 8-13 & 11-14 & SAW 1, 2, 3 \\
+\(Rec_s\) & Recruitment weeks for species \(s\) & wk & 9-12 & 8-13 & 11-14 & SAW 1, 2, 3 \\
 \bottomrule

### 2023-04-21 Manuscript/Draft1.Rmd
update manuscript

@@
 Randomly selected cells were samples each year.
-We then use the biomass collected from our samples in contemporary population indexing methods to estimate population trends.
+We then use the biomass collected from our samples in contemporary abundance indexing methods to estimate population trends.
 Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.
@@
 The left column depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern.
-In these scenarios we notice a small amount of shifting between successive years as the population aggregates on hot spots in the domain.
-On the other hand, in the right column where the temperature is increasing over time we note significant shifting taking places in certain strata.
-This figure shows how we were able to mimic shifting distributions due to climate change by running simulations with a temperature gradient that increased on average over time.
+In these scenarios we notice a small amount of shifting between successive years as the population aggregates on especially suitable habitat in the domain.
+More exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time, as seen right column of Figure \@ref(fig:PopPct).
+This figure demonstrates how we were able to mimic shifting distributions due to climate change by running simulations with a temperature gradient that increased on average over time.
 
@@
 While the model-based results tended to provide a lower relative error to the stratified mean with a mean relative error of 0.34 compared to 0.38, the model-based estimates were very sensitive to the settings used.
-As a result, although $132/160=82.5\%$ of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the variance for the VAST errors was much larger than the stratified mean (0.10 compared to 0.02).
+As a result, although $132/160=82.5\%$ of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the variance for the VAST relative errors was much larger than the stratified mean (0.10 compared to 0.02).
 VAST models that include covariate information provided the lowest overall errors and standard deviations.
@@
 Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain.
-
-<font size="5">*Differences Between VAST Settings*</font>
-
-For YTF, 29/96 ~ 30% of VAST runs with new settings were better (70% were worse).
-80/96 ~ 83% of scenarios had a VAST run with a better error than the stratified mean.
-15/16 of the times the stratified mean was better VAST was not using covariates.
-The 1 time that VAST was using covariates and was still worse than the stratified mean was IncPop_IncTemp_Allstrata_WCov_WNoise in the Fall (VAST had strong overestimate in the fall during IncPOP_IncTemP_allstrata for Had and YT).
-The overestimate in the fall only seems to be related to the way the population shifts between season.
-Looking at the percent shift plots, the population is shifting out of strata 16 in both seasons (large east strata).
-In the spring the population is shifting mostly into strata 17 (thin strata adjacent to 16), but in the fall they are shifting into both 17 and 18 (18 thin one adjacent to 17).
-Thus it seems like with much of the population concentrated in the really small outer most strata, VAST produces an underestimate, even with covariates
-
-For Haddock (IncPop), 18/32 ~ 57% of VAST runs better with new settings.
-24/32 = 75% of scenarios had a VAST run with a better error than the stratified mean.
-14/16 VAST were better with covariates (the two that were worse were essentially the same) while only 10/16 were better without covariates.
-Stratified mean tended to perform better when all strata were included in the calculation, while VAST tended to perform better when strata are removed.
-
-For Cod (DecPop), only 11/32 ~ 34% of VAST runs were better with new settings.
-28/32~ 88% of scenarios had a VAST run with a better estimate than the stratified mean.
-All 16/16 = 100% of VAST runs with a covariate were better than the stratified mean.
-12/16 VAST were better without covariates.
-The 4 runs without covariates that were worse were all extremely large errors and all from increasing temp in the fall (with and without all strata and noise).
-In the fall the population is moving out of strata 16 (large eastern strata) and 21 (north large) and into a number of strata (22 (tiny north), 24 (large north), 18 (tiny north)).
-Might be having trouble tracking in to smaller strata again.
-
-<!-- [[We'll want to expand the results section to focus on the impact of each factor one at at time within each species.
-This will be a bit dull, but it is important to walk through the results in words to ensure that readers get the message.
-They can of course examine the tables in detail and draw their own conclusions, but we should put our interpretation down on paper.]] I think the above paragraphs are addressing this comment, but wanted to be sure they are not just "notes to self" before providing comments on them.
--->
 
@@
 However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata).
-Each of these instances are These errors are corrected by including covariate information resulting in improved VAST estimates that are significantly lower than their stratified mean counterparts.
+Including covariate information to each VAST model produced estimates with significantly lower error than their stratified mean counterparts.
 
@@
 
-Including detailed covariate information made a noticeable difference in our Yellowtail Flounder model-based VAST estimates.
+Including covariate information made a noticeable difference in our Yellowtail Flounder model-based VAST estimates.
 All of the VAST estimates that included covariate information produced a lower relative error than the corresponding VAST model that did not include covariates.
@@
 As a result, of the VAST estimates that included covariates information, $47/48\approx98$\% had a lower relative error than the corresponding stratified mean estimate.
-This implies that the covariate information helped our model-based estimate account for the design-based issues related to small strata that contain a large percentage of the population, which can become exacerbated in the increasing temperature scenario.
+This implies that the covariate information helped our model-based estimate account for the design-based issues related to an increasing percentage of the population entering smaller strata, which can become exacerbated in the increasing temperature simulations.
 
-The performance of our abundance estimates tend to decrease under increasing temperature, with the most dramatic changes seen in our VAST estimates without covariates.
+We see diminished performance of our abundance estimates for Yellowtail under increasing temperature, with the most dramatic changes seen in our VAST estimates without covariates.
 One exception to this is when we use a stratified mean approach while the population is decreasing.
@@
 
-**For discussion: clearly our perfect covariate information made a big difference here, but the question is could we achieve similar results without "perfect" information?
-Which covariate is best?
-Good question for future research**
-
 <font size="5">*Cod Results*</font>
 
-We see a seasonal impact of an increasing average temperature in stratum 16 in the Cod results in Figure \@ref(fig:PopPct).
-In the constant temperature scenario for Cod, in both seasons the population is decreasing its presence in strata 19 and 20 over the duration of the simulation, while simultaneously increasing presence in stratum 16.
-However, an increase in average temperature in the fall drives the population out of strata 16 and 21, and into 18, 22, and 24 (see Figure \@ref(fig:hab-plot)).
-Similar to the Yellowtail Flounder population, stratum the habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler.
-When the temperature increases over time the fall population begins to prefer the adjacent strata that are deeper and/or more northern than stratum 16.
+In the constant temperature scenario for Cod, in both seasons the population decreases its presence in strata 19 and 20 over the duration of the simulation, while simultaneously increasing presence in stratum 16.
+However, we see a seasonal impact in stratum 16 during the increasing average temperature simulations where in the fall the population decrease presence in strata 16 and 21, and increase presence in 18, 22, and 24 (see Figure \@ref(fig:hab-plot)).
+Similar to the Yellowtail Flounder population, the favorable habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler.
+When the temperature increases over time, the fall population compensates by shifting their preference to the adjacent strata that are deeper and/or further north than stratum 16.
 
-(see Table \@ref(tab:CodERROR)) In the Cod results, when using reduced strata in constant temperature, adding covariates produces worse VAST results (though still better than stratified mean).
-This is similar to what we see with all of Haddock results (adding covariates makes it worse with reduced strata).
-This is NOT seen with Yellowtail (adding covariates always helps).
-VAST without covariates much worse than stratified mean in fall with increasing temperature, but adding covariates corrects this and makes VAST estimates better than stratified mean.
+Table \@ref(tab:CodERROR) contains the absolute error between our abundance estimates for Cod and the true model values.
+Of the abundance estimates without covariates, $12/16=75$\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
+VAST produced higher relative error compared to the stratified mean in scenarios that involved increasing temperature in the fall season without covariates.
+Similar to the Yellowtail results, this implies the model-based approach had seasonal trouble with populations shifting into smaller strata where fewer samples take place.
+Providing covariate information in these cases once again helped the model-based approach to provide improved absolute error estimates relative to the stratified mean.
+However, we see that adding covariates to VAST in the fall produces higher absolute error values when sampling reduced strata in the constant temperature scenario.
 
-Of the abundance estimates without covariates, $12/16=75$\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
-Including detailed covariate information in our Cod VAST estimates resulted in all $16/16=100\%$ having a lower relative error than the corresponding stratified mean estimate.
-The 4 scenarios without covariates that had a higher VAST error were each overestimates for the fall season that involved an increasing temperature gradient.
-Similar to Yellowtail Flounder results, Cod are shifting their spatial distribution into the smaller exterior strata in these conditions, which is decreasing the accuracy of the covariate-free VAST estimate.
+<!-- Similar to Yellowtail Flounder results, Cod are shifting their spatial distribution into the smaller exterior strata in these conditions, which is decreasing the accuracy of the covariate-free VAST estimate.
 While adding covariates to these fits allowed for an improved error relative to the stratified mean, the VAST abundance still display an overestimate early in the time series before covariates correct the trend, allowing for a lower overall absolute error **(show these plots?
 IncTemp fall scenarios)**.
+-->
 
@@
 
-(see Table \@ref(tab:HadERROR)) For Haddock, VAST has a particularly hard time in spring regularly producing larger errors than the stratified mean with added covariates only improving to the level of the stratified mean.
+Table \@ref(tab:HadERROR) contains the absolute error between our abundance estimates for Haddock and the true model values.
+We notice that the model-based VAST produced particularly large errors in spring compared to the stratified mean, with added covariates only improving to the level of the stratified mean.
 VAST shows improved results in fall relative to the stratified mean with added covariates producing extremely low errors in some cases.
-In Haddock results, adding covariates improves estimates only when all strata are included (when reduced strata, adding covariates actually makes VAST worse).
-Since this occurs in all scenarios, it seems to again be related to failing to accurately monitor strata 17.
+For Haddock results, adding covariates improves estimates only when all strata are included.
+That is, similar to Cod, when sampling a reduced domain adding covariates actually decreases VAST's accuracy.
+Since this occurs in all scenarios, it seems to again be related to failing to accurately monitor smaller exterior stratum 17.
 
-$10/16\approx63$\% of the scenarios without covariates had a VAST fit with a lower relative error compared to the stratified mean estimate.
+Of the scenarios without covariates, $10/16\approx63$\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
 The 6 covariate-free VAST estimates that resulted in a higher error than the stratified mean spanned all scenarios and seasons, with several fall errors being especially large (significant overestimates).
 Adding covariate information resulted in $14/16\approx88$\% of VAST estimates having improved error compared to the stratified mean.
-The 2 scenarios that procedure worse error spanned temperature scenarios, but were both in the spring season when the proportion of the population in each strata remained constant in each scenario.**why??
-doesnt make sense.
-the values with cov are about the same as SM** Qualitatively, including covariate information has a larger impact on decreasing the relative error compared to the stratified mean compared to failing to include covariates.
+The 2 scenarios that procedure worse error spanned temperature scenarios, but were both in the spring season when the proportion of the population in each strata remained constant in each scenario.
+Qualitatively, including covariate information has a larger impact on decreasing the relative error compared to the stratified mean compared to failing to include covariates.
+
+<font size="5">*Impact of Covariates*</font>
+
+Generally help, but dependent on appropriate covariate response.
+Yellowtail cases always had appropriate response and thus adding covariates always improved estimates.
+VAST showed decreased performance with Cod and Haddock when covariates added in fall because covariate response was not fully developed.
+(show plot comparing good vs bad response)
 
@@
 
+<font size="5">*Differences Between VAST Settings*</font>
+
+For YTF, 29/96 ~ 30% of VAST runs with new settings were better (70% were worse).
+80/96 ~ 83% of scenarios had a VAST run with a better error than the stratified mean.
+15/16 of the times the stratified mean was better VAST was not using covariates.
+The 1 time that VAST was using covariates and was still worse than the stratified mean was IncPop_IncTemp_Allstrata_WCov_WNoise in the Fall (VAST had strong overestimate in the fall during IncPOP_IncTemP_allstrata for Had and YT).
+The overestimate in the fall only seems to be related to the way the population shifts between season.
+Looking at the percent shift plots, the population is shifting out of strata 16 in both seasons (large east strata).
+In the spring the population is shifting mostly into strata 17 (thin strata adjacent to 16), but in the fall they are shifting into both 17 and 18 (18 thin one adjacent to 17).
+Thus it seems like with much of the population concentrated in the really small outer most strata, VAST produces an underestimate, even with covariates
+
+For Haddock (IncPop), 18/32 ~ 57% of VAST runs better with new settings.
+24/32 = 75% of scenarios had a VAST run with a better error than the stratified mean.
+14/16 VAST were better with covariates (the two that were worse were essentially the same) while only 10/16 were better without covariates.
+Stratified mean tended to perform better when all strata were included in the calculation, while VAST tended to perform better when strata are removed.
+
+For Cod (DecPop), only 11/32 ~ 34% of VAST runs were better with new settings.
+28/32~ 88% of scenarios had a VAST run with a better estimate than the stratified mean.
+All 16/16 = 100% of VAST runs with a covariate were better than the stratified mean.
+12/16 VAST were better without covariates.
+The 4 runs without covariates that were worse were all extremely large errors and all from increasing temp in the fall (with and without all strata and noise).
+In the fall the population is moving out of strata 16 (large eastern strata) and 21 (north large) and into a number of strata (22 (tiny north), 24 (large north), 18 (tiny north)).
+Might be having trouble tracking in to smaller strata again.
+
 \section{Discussion}
+
+**For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without "perfect" information?
+Which covariate is best?
+Good question for future research**
 
@@
 
-<div align="center">Table 1. Parameters used in all population models. SAW 1: [@nefsc1218], SAW 2: [@nefsc1311], SAW 3: ??? </div>
+<div align="center">Table 1. Parameters used in all population models. SAW 1: [@nefsc1218], SAW 2: [@nefsc1311], SAW 3: not out yet??? </div>
 
@@
 #result_table = read_csv("CSVs/Yellowtail_results_final.csv")
-result_table = read_csv("CSVs/Yellowtail_results_final_newsettings_forpaper.csv")
+result_table = read_csv("CSVs/Yellowtail_results_final_newsettings_forpaper_na.csv")
 

### 2023-04-21 Manuscript/Draft1.tex
update manuscript

@@
 The left column depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern.
-In these scenarios we notice a small amount of shifting between successive years as the population aggregates on hot spots in the domain.
-On the other hand, in the right column where the temperature is increasing over time we note significant shifting taking places in certain strata.
-This figure shows how we were able to mimic shifting distributions due to climate change by running simulations with a temperature gradient that increased on average over time.
+In these scenarios we notice a small amount of shifting between successive years as the population aggregates on especially suitable habitat in the domain.
+More exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time, as seen right column of Figure \ref{fig:PopPct}.
+This figure demonstrates how we were able to mimic shifting distributions due to climate change by running simulations with a temperature gradient that increased on average over time.
 
@@
 While the model-based results tended to provide a lower relative error to the stratified mean with a mean relative error of 0.34 compared to 0.38, the model-based estimates were very sensitive to the settings used.
-As a result, although \(132/160=82.5\%\) of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the variance for the VAST errors was much larger than the stratified mean (0.10 compared to 0.02).
+As a result, although \(132/160=82.5\%\) of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the variance for the VAST relative errors was much larger than the stratified mean (0.10 compared to 0.02).
 VAST models that include covariate information provided the lowest overall errors and standard deviations.
@@
 Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain.
-
-\emph{Differences Between VAST Settings}
-
-For YTF, 29/96 \textasciitilde{} 30\% of VAST runs with new settings were better (70\% were worse).
-80/96 \textasciitilde{} 83\% of scenarios had a VAST run with a better error than the stratified mean.
-15/16 of the times the stratified mean was better VAST was not using covariates.
-The 1 time that VAST was using covariates and was still worse than the stratified mean was IncPop\_IncTemp\_Allstrata\_WCov\_WNoise in the Fall (VAST had strong overestimate in the fall during IncPOP\_IncTemP\_allstrata for Had and YT).
-The overestimate in the fall only seems to be related to the way the population shifts between season.
-Looking at the percent shift plots, the population is shifting out of strata 16 in both seasons (large east strata).
-In the spring the population is shifting mostly into strata 17 (thin strata adjacent to 16), but in the fall they are shifting into both 17 and 18 (18 thin one adjacent to 17).
-Thus it seems like with much of the population concentrated in the really small outer most strata, VAST produces an underestimate, even with covariates
-
-For Haddock (IncPop), 18/32 \textasciitilde{} 57\% of VAST runs better with new settings.
-24/32 = 75\% of scenarios had a VAST run with a better error than the stratified mean.
-14/16 VAST were better with covariates (the two that were worse were essentially the same) while only 10/16 were better without covariates.
-Stratified mean tended to perform better when all strata were included in the calculation, while VAST tended to perform better when strata are removed.
-
-For Cod (DecPop), only 11/32 \textasciitilde{} 34\% of VAST runs were better with new settings.
-28/32\textasciitilde{} 88\% of scenarios had a VAST run with a better estimate than the stratified mean.
-All 16/16 = 100\% of VAST runs with a covariate were better than the stratified mean.
-12/16 VAST were better without covariates.
-The 4 runs without covariates that were worse were all extremely large errors and all from increasing temp in the fall (with and without all strata and noise).
-In the fall the population is moving out of strata 16 (large eastern strata) and 21 (north large) and into a number of strata (22 (tiny north), 24 (large north), 18 (tiny north)).
-Might be having trouble tracking in to smaller strata again.
 
@@
 However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}.
-Each of these instances are These errors are corrected by including covariate information resulting in improved VAST estimates that are significantly lower than their stratified mean counterparts.
+Including covariate information to each VAST model produced estimates with significantly lower error than their stratified mean counterparts.
 
@@
 As a result, of the VAST estimates that included covariates information, \(47/48\approx98\)\% had a lower relative error than the corresponding stratified mean estimate.
-This implies that the covariate information helped our model-based estimate account for the design-based issues related to small strata that contain a large percentage of the population, which can become exacerbated in the increasing temperature scenario.
-
-The performance of our abundance estimates tend to decrease under increasing temperature, with the most dramatic changes seen in our VAST estimates without covariates.
+This implies that the covariate information helped our model-based estimate account for the design-based issues related to an increasing percentage of the population entering smaller strata, which can become exacerbated in the increasing temperature simulations.
+
+We see diminished performance of our abundance estimates for Yellowtail under increasing temperature, with the most dramatic changes seen in our VAST estimates without covariates.
 One exception to this is when we use a stratified mean approach while the population is decreasing.
@@
 
-\textbf{For discussion: clearly our perfect covariate information made a big difference here, but the question is could we achieve similar results without ``perfect'' information?
-Which covariate is best?
-Good question for future research}
-
 \emph{Cod Results}
 
-We see a seasonal impact of an increasing average temperature in stratum 16 in the Cod results in Figure \ref{fig:PopPct}.
-In the constant temperature scenario for Cod, in both seasons the population is decreasing its presence in strata 19 and 20 over the duration of the simulation, while simultaneously increasing presence in stratum 16.
-However, an increase in average temperature in the fall drives the population out of strata 16 and 21, and into 18, 22, and 24 (see Figure \ref{fig:hab-plot}).
-Similar to the Yellowtail Flounder population, stratum the habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler.
-When the temperature increases over time the fall population begins to prefer the adjacent strata that are deeper and/or more northern than stratum 16.
-
-(see Table \ref{tab:CodERROR}) In the Cod results, when using reduced strata in constant temperature, adding covariates produces worse VAST results (though still better than stratified mean).
-This is similar to what we see with all of Haddock results (adding covariates makes it worse with reduced strata).
-This is NOT seen with Yellowtail (adding covariates always helps).
-VAST without covariates much worse than stratified mean in fall with increasing temperature, but adding covariates corrects this and makes VAST estimates better than stratified mean.
-
+In the constant temperature scenario for Cod, in both seasons the population decreases its presence in strata 19 and 20 over the duration of the simulation, while simultaneously increasing presence in stratum 16.
+However, we see a seasonal impact in stratum 16 during the increasing average temperature simulations where in the fall the population decrease presence in strata 16 and 21, and increase presence in 18, 22, and 24 (see Figure \ref{fig:hab-plot}).
+Similar to the Yellowtail Flounder population, the favorable habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler.
+When the temperature increases over time, the fall population compensates by shifting their preference to the adjacent strata that are deeper and/or further north than stratum 16.
+
+Table \ref{tab:CodERROR} contains the absolute error between our abundance estimates for Cod and the true model values.
 Of the abundance estimates without covariates, \(12/16=75\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
-Including detailed covariate information in our Cod VAST estimates resulted in all \(16/16=100\%\) having a lower relative error than the corresponding stratified mean estimate.
-The 4 scenarios without covariates that had a higher VAST error were each overestimates for the fall season that involved an increasing temperature gradient.
-Similar to Yellowtail Flounder results, Cod are shifting their spatial distribution into the smaller exterior strata in these conditions, which is decreasing the accuracy of the covariate-free VAST estimate.
-While adding covariates to these fits allowed for an improved error relative to the stratified mean, the VAST abundance still display an overestimate early in the time series before covariates correct the trend, allowing for a lower overall absolute error \textbf{(show these plots?
-IncTemp fall scenarios)}.
+VAST produced higher relative error compared to the stratified mean in scenarios that involved increasing temperature in the fall season without covariates.
+Similar to the Yellowtail results, this implies the model-based approach had seasonal trouble with populations shifting into smaller strata where fewer samples take place.
+Providing covariate information in these cases once again helped the model-based approach to provide improved absolute error estimates relative to the stratified mean.
+However, we see that adding covariates to VAST in the fall produces higher absolute error values when sampling reduced strata in the constant temperature scenario.
 
@@
 
-(see Table \ref{tab:HadERROR}) For Haddock, VAST has a particularly hard time in spring regularly producing larger errors than the stratified mean with added covariates only improving to the level of the stratified mean.
+Table \ref{tab:HadERROR} contains the absolute error between our abundance estimates for Haddock and the true model values.
+We notice that the model-based VAST produced particularly large errors in spring compared to the stratified mean, with added covariates only improving to the level of the stratified mean.
 VAST shows improved results in fall relative to the stratified mean with added covariates producing extremely low errors in some cases.
-In Haddock results, adding covariates improves estimates only when all strata are included (when reduced strata, adding covariates actually makes VAST worse).
-Since this occurs in all scenarios, it seems to again be related to failing to accurately monitor strata 17.
-
-\(10/16\approx63\)\% of the scenarios without covariates had a VAST fit with a lower relative error compared to the stratified mean estimate.
+For Haddock results, adding covariates improves estimates only when all strata are included.
+That is, similar to Cod, when sampling a reduced domain adding covariates actually decreases VAST's accuracy.
+Since this occurs in all scenarios, it seems to again be related to failing to accurately monitor smaller exterior stratum 17.
+
+Of the scenarios without covariates, \(10/16\approx63\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
 The 6 covariate-free VAST estimates that resulted in a higher error than the stratified mean spanned all scenarios and seasons, with several fall errors being especially large (significant overestimates).
 Adding covariate information resulted in \(14/16\approx88\)\% of VAST estimates having improved error compared to the stratified mean.
-The 2 scenarios that procedure worse error spanned temperature scenarios, but were both in the spring season when the proportion of the population in each strata remained constant in each scenario.\textbf{why??
-doesnt make sense.
-the values with cov are about the same as SM} Qualitatively, including covariate information has a larger impact on decreasing the relative error compared to the stratified mean compared to failing to include covariates.
+The 2 scenarios that procedure worse error spanned temperature scenarios, but were both in the spring season when the proportion of the population in each strata remained constant in each scenario.
+Qualitatively, including covariate information has a larger impact on decreasing the relative error compared to the stratified mean compared to failing to include covariates.
+
+\emph{Impact of Covariates}
+
+Generally help, but dependent on appropriate covariate response.
+Yellowtail cases always had appropriate response and thus adding covariates always improved estimates.
+VAST showed decreased performance with Cod and Haddock when covariates added in fall because covariate response was not fully developed.
+(show plot comparing good vs bad response)
 
@@
 
+\emph{Differences Between VAST Settings}
+
+For YTF, 29/96 \textasciitilde{} 30\% of VAST runs with new settings were better (70\% were worse).
+80/96 \textasciitilde{} 83\% of scenarios had a VAST run with a better error than the stratified mean.
+15/16 of the times the stratified mean was better VAST was not using covariates.
+The 1 time that VAST was using covariates and was still worse than the stratified mean was IncPop\_IncTemp\_Allstrata\_WCov\_WNoise in the Fall (VAST had strong overestimate in the fall during IncPOP\_IncTemP\_allstrata for Had and YT).
+The overestimate in the fall only seems to be related to the way the population shifts between season.
+Looking at the percent shift plots, the population is shifting out of strata 16 in both seasons (large east strata).
+In the spring the population is shifting mostly into strata 17 (thin strata adjacent to 16), but in the fall they are shifting into both 17 and 18 (18 thin one adjacent to 17).
+Thus it seems like with much of the population concentrated in the really small outer most strata, VAST produces an underestimate, even with covariates
+
+For Haddock (IncPop), 18/32 \textasciitilde{} 57\% of VAST runs better with new settings.
+24/32 = 75\% of scenarios had a VAST run with a better error than the stratified mean.
+14/16 VAST were better with covariates (the two that were worse were essentially the same) while only 10/16 were better without covariates.
+Stratified mean tended to perform better when all strata were included in the calculation, while VAST tended to perform better when strata are removed.
+
+For Cod (DecPop), only 11/32 \textasciitilde{} 34\% of VAST runs were better with new settings.
+28/32\textasciitilde{} 88\% of scenarios had a VAST run with a better estimate than the stratified mean.
+All 16/16 = 100\% of VAST runs with a covariate were better than the stratified mean.
+12/16 VAST were better without covariates.
+The 4 runs without covariates that were worse were all extremely large errors and all from increasing temp in the fall (with and without all strata and noise).
+In the fall the population is moving out of strata 16 (large eastern strata) and 21 (north large) and into a number of strata (22 (tiny north), 24 (large north), 18 (tiny north)).
+Might be having trouble tracking in to smaller strata again.
+
 \section{Discussion}
+
+\textbf{For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without ``perfect'' information?
+Which covariate is best?
+Good question for future research}
 
@@
 
-Table 1. Parameters used in all population models. SAW 1: (NEFSC 2012), SAW 2: (NEFSC 2013), SAW 3: ???
+Table 1. Parameters used in all population models. SAW 1: (NEFSC 2012), SAW 2: (NEFSC 2013), SAW 3: not out yet???
 
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.07}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.07}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.21}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.08}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.08}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.25}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.07}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.07}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.08}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.08}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.14}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.08}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.32}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.11}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.31}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.14}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.08}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.11}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.05}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.06}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.28}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.10}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.12}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.28}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.05}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.06}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.10}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.12}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{1.40}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{1.38}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.50}}}\\
-\hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.21}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.23}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.51}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.30}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.28}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.50}}}\\
+\cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{no cov}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{1.40}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{1.38}}} & \cellcolor[HTML]{5C5E60}{\textcolor{white}{\textbf{0.5}}}\\
+\hline
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.21}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.23}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.30}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.28}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.07}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.06}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.23}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.11}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.07}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.27}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.07}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.06}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.11}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.07}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.16}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.08}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.41}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.29}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.18}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.37}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.16}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.08}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.29}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.18}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.08}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.07}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.22}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.16}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.10}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.26}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.08}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.07}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.16}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.10}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.40}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.15}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.28}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.40}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.20}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.25}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.40}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.15}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.40}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.20}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.06}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.06}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.16}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.08}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.07}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.22}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.06}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.06}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.08}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.07}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.11}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.08}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.34}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.24}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.46}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.11}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.08}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.24}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.06}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.07}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.32}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.12}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.10}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.32}}}\\
-\hline
-\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor{black}{\textcolor{white}{\textbf{fall}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.71}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.66}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.30}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.06}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.07}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.12}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.10}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor{black}{\textcolor{white}{\textbf{fall}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.71}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.66}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.3}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.43}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.21}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.30}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.51}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.37}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.30}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.43}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.21}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.51}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.37}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.19}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.19}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.27}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.12}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.14}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.22}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.19}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.19}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\vphantom{1} n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.12}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.14}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.22}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.19}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.47}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.24}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.44}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.22}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.19}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.24}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.31}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.62}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.15}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.29}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.62}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.15}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.24}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.20}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.64}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.59}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.13}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.59}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.24}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.20}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.59}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.13}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.19}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.19}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.25}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.15}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.16}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.27}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.19}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.19}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.15}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.16}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.18}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.23}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.55}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.16}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.24}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.53}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.18}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.23}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.16}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.24}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.21}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.22}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.32}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.16}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.21}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.29}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.21}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.22}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.16}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.21}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.32}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.31}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.54}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.36}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.32}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.48}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.32}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.31}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.36}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.32}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
 \hline
@@
 \hline
-\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor{black}{\textcolor{white}{\textbf{spring}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.22}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.15}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.20}}}\\
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor{black}{\textcolor{white}{\textbf{spring}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.22}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.15}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.2}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.20}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.11}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.13}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.22}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.11}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.13}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.21}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.22}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.41}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.19}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.46}}}\\
-\hline
-\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor{black}{\textcolor{white}{\textbf{spring}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.31}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.33}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.40}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.21}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.22}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Constant}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.17}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.19}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor{black}{\textcolor{white}{\textbf{spring}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.31}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.33}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.4}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.30}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.30}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.40}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.31}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.25}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.38}}}\\
-\hline
-\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor{black}{\textcolor{white}{\textbf{fall}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.56}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.49}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.70}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.30}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.30}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{spring}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.31}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.25}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor{black}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor{black}{\textcolor{white}{\textbf{fall}}} & \cellcolor{black}{\textcolor{white}{\textbf{no cov}}} & \cellcolor{black}{\textcolor{white}{\textbf{no}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.56}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.49}}} & \cellcolor{black}{\textcolor{white}{\textbf{0.7}}}\\
 \hline
@@
 \hline
-\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.48}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.53}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.70}}}\\
-\hline
-\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.47}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.50}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.69}}}\\
+\cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{no}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.48}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{0.53}}} & \cellcolor[HTML]{A4A4A4}{\textcolor{white}{\textbf{n/a}}}\\
+\hline
+\cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{\hspace{1em}Increasing}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{fall}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{w/ cov}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{yes}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.47}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{0.50}}} & \cellcolor[HTML]{D2D2D2}{\textcolor{white}{\textbf{n/a}}}\\
 \hline

### 2023-04-25 Manuscript/Draft1.Rmd
a few manuscript updates

@@
 While the model-based results tended to provide a lower relative error to the stratified mean with a mean relative error of 0.34 compared to 0.38, the model-based estimates were very sensitive to the settings used.
-As a result, although $132/160=82.5\%$ of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the variance for the VAST relative errors was much larger than the stratified mean (0.10 compared to 0.02).
+As a result, although $77/80\approx 96\%$ of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the variance for the VAST relative errors was much larger than the stratified mean (0.10 compared to 0.02).
 VAST models that include covariate information provided the lowest overall errors and standard deviations.
@@
 That is, if the population is low enough, failing to appropriately sample the population in a design-based method produces the same result as appropriately sampling.
+In comparing abundances estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a larger impact on the model-based estimates compared to the design-based method.
+Specifically, when comparing the error of abundance estimates from the constant temperature scenario to the corresponding increasing temperature scenario, VAST increased by an average factor of 1.75 (both with and without covariates) while the  stratified mean error increased by an average factor of 1.25.
 
@@
 However, we see that adding covariates to VAST in the fall produces higher absolute error values when sampling reduced strata in the constant temperature scenario.
+
+In comparing abundances estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a large negative impact on the model-based estimates.
+Specifically, when comparing the error of abundance estimates from the constant temperature scenario to the corresponding increasing temperature scenario for Cod, VAST increased by an average factor of 3.57 without covariates and 1.67 with covariates.
+In considering the average change in the error of stratified mean estimates between constant temperature scenarios and increasing temperature scenarios we surprisingly find an improvement in the error of abundance estimates, with an average decrease by a factir of 0.87.
 
@@
 Adding covariate information resulted in $14/16\approx88$\% of VAST estimates having improved error compared to the stratified mean.
-The 2 scenarios that procedure worse error spanned temperature scenarios, but were both in the spring season when the proportion of the population in each strata remained constant in each scenario.
-Qualitatively, including covariate information has a larger impact on decreasing the relative error compared to the stratified mean compared to failing to include covariates.
-
-<font size="5">*Impact of Covariates*</font>
-
-Generally help, but dependent on appropriate covariate response.
-Yellowtail cases always had appropriate response and thus adding covariates always improved estimates.
-VAST showed decreased performance with Cod and Haddock when covariates added in fall because covariate response was not fully developed.
-(show plot comparing good vs bad response)
+The 2 scenarios that produced worse error spanned temperature scenarios, but were both in the spring season when the proportion of the population in each strata remained constant in each scenario.
+The average change in the error of abundance estimates between constant temperature scenarios and increasing temperature scenarios were 1.42 for VAST estimates without covariates, 1.80 for VAST estimates that included covariates, and 1.37 for stratified mean estimates.
 
@@
 
+<font size="5">*Impact of Covariates*</font>
+
+As see in Table **cite VAST setting table**, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
+While using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates with $77/80\approx 96\%$ of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate, there is a great deal more to explore with respect to the impact of covariates.
+For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates.
+Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information.
+One could also test the impact of assuming the wrong covarite response function (linear vs normally distributed etc).
 **For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without "perfect" information?
@@
 Good question for future research**
+
+As noted in results second, adding covariate information to our Yellowtail Flounder VAST models always decreased the error in our abundance estimates, while covariate information sometimes lead to an increase in the absolute error between the abundance estimate and true model abundance for Cod and Haddock.
+More specifically, when a reduced number of strata are sampled adding covariate information leads to a decrease in performance for fall estimates of Haddock and Cod.
+The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, which leads to an incomplete estimation of the covariate response.
+Generally help, but dependent on appropriate covariate response.
+Yellowtail cases always had appropriate response and thus adding covariates always improved estimates.
+VAST showed decreased performance with Cod and Haddock when covariates added in fall because covariate response was not fully developed.
+(show plot comparing good vs bad response)
+
+increasing temperature scenarios resulted in a larger increase in average error for VAST estimates compared to stratified mean estimates.
+When including covarites the increase is less than not including them, but stratified mean saw a lower increase in the average error seen in constand an dincreasing temperature.
+Shouldnt model based be more stable with respect to increasing temperature?
 
@@
 
-<div align="center">Table XX. Parameters used in all VAST models.</div>
+<div align="center">Table XX. Settings for our two VAST models.</div>
 
@@
 |FieldConfig | Specified spatial and/or spatio-temporal variation in predictors| c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1) | c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1)
-| RhoConfig | Specifying whether intercepts or spatio-temporal variation is structured among time intervals | c(Beta1=3, Beta2=, Epsilon1=0, Epsilon2=4) | c(Beta1=3, Beta2=3, Epsilon1=0, Epsilon2=4)
-| X1_formula | Right-sided formula affecting the 1st linear predictor  | X1_formula = ~ poly(Temp, degree=2 ) | N/A
+| RhoConfig | Specifying whether intercepts or spatio-temporal variation is structured among time intervals | c(Beta1=3, Beta2=0, Epsilon1=0, Epsilon2=4) | c(Beta1=3, Beta2=3, Epsilon1=0, Epsilon2=4)
+| X1_formula | Right-sided formula affecting the 1st linear predictor  |N/A  | X1_formula = ~ poly(Temp, degree=2 )
 | X2_formula | Right-sided formula affecting the 2nd linear predictor | X2_formula = ~ poly(Temp, degree=2 ) + poly(Habitat, degree=2 ) | X2_formula = ~ poly(Temp, degree=2 ) + poly(Habitat, degree=2 )

### 2023-04-25 Manuscript/Draft1.tex
a few manuscript updates

@@
 While the model-based results tended to provide a lower relative error to the stratified mean with a mean relative error of 0.34 compared to 0.38, the model-based estimates were very sensitive to the settings used.
-As a result, although \(132/160=82.5\%\) of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the variance for the VAST relative errors was much larger than the stratified mean (0.10 compared to 0.02).
+As a result, although \(77/80\approx 96\%\) of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the variance for the VAST relative errors was much larger than the stratified mean (0.10 compared to 0.02).
 VAST models that include covariate information provided the lowest overall errors and standard deviations.
@@
 That is, if the population is low enough, failing to appropriately sample the population in a design-based method produces the same result as appropriately sampling.
+In comparing abundances estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a larger impact on the model-based estimates compared to the design-based method.
+Specifically, when comparing the error of abundance estimates from the constant temperature scenario to the corresponding increasing temperature scenario, VAST increased by an average factor of 1.75 (both with and without covariates) while the stratified mean error increased by an average factor of 1.25.
 
@@
 Adding covariate information resulted in \(14/16\approx88\)\% of VAST estimates having improved error compared to the stratified mean.
-The 2 scenarios that procedure worse error spanned temperature scenarios, but were both in the spring season when the proportion of the population in each strata remained constant in each scenario.
-Qualitatively, including covariate information has a larger impact on decreasing the relative error compared to the stratified mean compared to failing to include covariates.
-
-\emph{Impact of Covariates}
-
-Generally help, but dependent on appropriate covariate response.
-Yellowtail cases always had appropriate response and thus adding covariates always improved estimates.
-VAST showed decreased performance with Cod and Haddock when covariates added in fall because covariate response was not fully developed.
-(show plot comparing good vs bad response)
+The 2 scenarios that produced worse error spanned temperature scenarios, but were both in the spring season when the proportion of the population in each strata remained constant in each scenario.
+The average change in the error of abundance estimates between constant temperature scenarios and increasing temperature scenarios were 1.42 for VAST estimates without covariates, 1.80 for VAST estimates that included covariates, and 1.37 for stratified mean estimates.
 
@@
 
+\emph{Impact of Covariates}
+
+As see in Table \textbf{cite VAST setting table}, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
+While using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates with \(77/80\approx 96\%\) of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate, there is a great deal more to explore with respect to the impact of covariates.
+For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates.
+Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information.
+One could also test the impact of assuming the wrong covarite response function (linear vs normally distributed etc).
 \textbf{For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without ``perfect'' information?
@@
 Good question for future research}
+
+As noted in results second, adding covariate information to our Yellowtail Flounder VAST models always decreased the error in our abundance estimates, while covariate information sometimes lead to an increase in the absolute error between the abundance estimate and true model abundance for Cod and Haddock.
+More specifically, when a reduced number of strata are sampled adding covariate information leads to a decrease in performance for fall estimates of Haddock and Cod.
+The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, which leads to an incomplete estimation of the covariate response.
+Generally help, but dependent on appropriate covariate response.
+Yellowtail cases always had appropriate response and thus adding covariates always improved estimates.
+VAST showed decreased performance with Cod and Haddock when covariates added in fall because covariate response was not fully developed.
+(show plot comparing good vs bad response)
+
+increasing temperature scenarios resulted in a larger increase in average error for VAST estimates compared to stratified mean estimates.
+When including covarites the increase is less than not including them, but stratified mean saw a lower increase in the average error seen in constand an dincreasing temperature.
+Shouldnt model based be more stable with respect to increasing temperature?
 
@@
 
-Table XX. Parameters used in all VAST models.
+Table XX. Settings for our two VAST models.
 
@@
 FieldConfig & Specified spatial and/or spatio-temporal variation in predictors & c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1) & c(Omega1=0, Epsilon1=0, Omega2=1, Epsilon2=1) \\
-RhoConfig & Specifying whether intercepts or spatio-temporal variation is structured among time intervals & c(Beta1=3, Beta2=, Epsilon1=0, Epsilon2=4) & c(Beta1=3, Beta2=3, Epsilon1=0, Epsilon2=4) \\
-X1\_formula & Right-sided formula affecting the 1st linear predictor & X1\_formula = \textasciitilde{} poly(Temp, degree=2 ) & N/A \\
+RhoConfig & Specifying whether intercepts or spatio-temporal variation is structured among time intervals & c(Beta1=3, Beta2=0, Epsilon1=0, Epsilon2=4) & c(Beta1=3, Beta2=3, Epsilon1=0, Epsilon2=4) \\
+X1\_formula & Right-sided formula affecting the 1st linear predictor & N/A & X1\_formula = \textasciitilde{} poly(Temp, degree=2 ) \\
 X2\_formula & Right-sided formula affecting the 2nd linear predictor & X2\_formula = \textasciitilde{} poly(Temp, degree=2 ) + poly(Habitat, degree=2 ) & X2\_formula = \textasciitilde{} poly(Temp, degree=2 ) + poly(Habitat, degree=2 ) \\

### 2023-04-26 Manuscript/Draft1.Rmd
manuscript update with a few more results from presentation

@@
 A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, too low.
-This can be further examined by viewing estimate/ratio ratio plots, where we divide the yearly abundance estimates by the true model value.
+This can be further examined by viewing estimate/model ratio plots, where we divide the yearly abundance estimates by the true model value and plot them on the log scale.
 In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can range as low as 0 and exhibit large yearly changes.
+A representative example of this trend is shown in Figure \@ref(fig:RatioTrend)
 
@@
 
-As noted in results second, adding covariate information to our Yellowtail Flounder VAST models always decreased the error in our abundance estimates, while covariate information sometimes lead to an increase in the absolute error between the abundance estimate and true model abundance for Cod and Haddock.
-More specifically, when a reduced number of strata are sampled adding covariate information leads to a decrease in performance for fall estimates of Haddock and Cod.
-The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, which leads to an incomplete estimation of the covariate response.
+As noted in results second, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates, while adding covariate information leads to a decrease in performance for fall estimates of Haddock and Cod when a reduced number of strata are sampled.
+The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response.
+Figure \@ref(fig:strata-plot) shows the strata excluded from sampling and Figure \@ref(fig:CovResponse) depicts the resulting covariate response for Yellowtail Flounder and Haddock for the same population scenario.
+Although strata were excluded for both species, the samples for Yellowtail Flounder ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata).
+As a result, when projecting into those strata as seen in the lower portion of Figure \@ref(fig:CovResponse), Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend at the wrong magnitude.
+
 Generally help, but dependent on appropriate covariate response.
@@
 increasing temperature scenarios resulted in a larger increase in average error for VAST estimates compared to stratified mean estimates.
-When including covarites the increase is less than not including them, but stratified mean saw a lower increase in the average error seen in constand an dincreasing temperature.
-Shouldnt model based be more stable with respect to increasing temperature?
+When including covariates the increase is less than not including them, but stratified mean saw a lower increase in the average error seen in constant an decreasing temperature.
+Shouldn't model based be more stable with respect to increasing temperature?
 
@@
 
-```{r PopPct, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Percent of each species in each strata for during survey weeks in our spatial simulations.
+```{r PopPct, echo=FALSE,fig.show='hold', fig.align='center', fig.cap="Percent of each species in each strata for during survey weeks in our spatial simulations.
 All constant temperature scenarios follow the patterns on the left while increasing temperature scenarios follow the patterns on the right.",out.width="47.5%"}
@@
 knitr::include_graphics(c("Images/PercentPlots_IncPopConTemp_Had.png","Images/PercentPlots_IncPopIncTemp_Had.png"))
+```
+
+```{r CovResponse, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Temperature covariate response plots and resulting population estimate for Yellowtail Flounder on the left and Haddock on the right.
+In both cases the spatial simulations that were surveyed had an increasing population over time, increasing average temperature, and certain strata excluded from sampling, as shown in Figure 1**check**.",out.width="90%"}
+# knitr::include_graphics(c("Images/covariate_good.png","Images/covariate_bad.png"))
+# knitr::include_graphics(c("Images/covariate_est.png"))
+
+knitr::include_graphics(c("Images/covariate_plot.png"))
+
+```
+
+```{r RatioTrend, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Representative example of typical ratio trend as shown on a log scale.",out.width="90%"}
+# knitr::include_graphics(c("Images/covariate_good.png","Images/covariate_bad.png"))
+# knitr::include_graphics(c("Images/covariate_est.png"))
+
+knitr::include_graphics(c("Images/ratio_trend.png"))
+
 ```

### 2023-04-26 Manuscript/Draft1.tex
manuscript update with a few more results from presentation

@@
 A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, too low.
-This can be further examined by viewing estimate/ratio ratio plots, where we divide the yearly abundance estimates by the true model value.
+This can be further examined by viewing estimate/model ratio plots, where we divide the yearly abundance estimates by the true model value and plot them on the log scale.
 In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can range as low as 0 and exhibit large yearly changes.
+A representative example of this trend is shown in Figure \ref{fig:RatioTrend}
 
@@
 
-As noted in results second, adding covariate information to our Yellowtail Flounder VAST models always decreased the error in our abundance estimates, while covariate information sometimes lead to an increase in the absolute error between the abundance estimate and true model abundance for Cod and Haddock.
-More specifically, when a reduced number of strata are sampled adding covariate information leads to a decrease in performance for fall estimates of Haddock and Cod.
-The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, which leads to an incomplete estimation of the covariate response.
+As noted in results second, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates, while adding covariate information leads to a decrease in performance for fall estimates of Haddock and Cod when a reduced number of strata are sampled.
+The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response.
+Figure \ref{fig:strata-plot} shows the strata excluded from sampling and Figure \ref{fig:CovResponse} depicts the resulting covariate response for Yellowtail Flounder and Haddock for the same population scenario.
+Although strata were excluded for both species, the samples for Yellowtail Flounder ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata).
+As a result, when projecting into those strata as seen in the lower portion of Figure \ref{fig:CovResponse}, Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend at the wrong magnitude.
+
 Generally help, but dependent on appropriate covariate response.
@@
 All constant temperature scenarios follow the patterns on the left while increasing temperature scenarios follow the patterns on the right.}\label{fig:PopPct}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{Images/covariate_plot}
+
+}
+
+\caption{Temperature covariate response plots and resulting population estimate for Yellowtail Flounder on the left and Haddock on the right.
+In both cases the spatial simulations that were surveyed had an increasing population over time, increasing average temperature, and certain strata excluded from sampling, as shown in Figure 1**check**.}\label{fig:CovResponse}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{Images/ratio_trend}
+
+}
+
+\caption{Representative example of typical ratio trend as shown on a log scale.}\label{fig:RatioTrend}
 \end{figure}

### 2023-04-27 Manuscript/Draft1.Rmd
Manuscript update

@@
 The survey takes place twice each year- once in the spring and again in the fall.
-Since most spatial analyses and  projections of future distributions typically assume a constant survey catchability and/or availability over time, NOAA's survey design includes sampling Georges Bank during approximately the same 3-4 week time period in each season.
+Most spatial analyses and  projections of future distributions typically assume a constant survey catchability and/or availability over time.
+For this reason, NOAA's survey design includes sampling each strata in approximately the same 3-4 week time period in each season.
 
@@
 Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades.
-The changes have resulted in a shift in spatial distributions of many species (cite Nye).
-Existing research has focused on temperature as the driver of such changes (cite klein, smith and kritzer) and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges (kerr, barajas, wiedenmann).
+The changes have resulted in a shift in spatial distributions of many species [@nye2009changing; @kleisner2017marine; @henderson2017effects].
 Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments.
 More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey will change, altering the relationship between the index and the true population.
-We are therefore interested in analyzing the impact of climate change on the accuracy of future stock assessment models as measured by NOAA’s ongoing bottom-trawl survey along the East coast.
+A species shifting its range beyond the survey area is an additional compounding factor to consider.
+Existing research has focused on temperature as the driver of such changes [@klein2017effects] and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges [@kerr2022coherence].
+In these scenarios, management strategy evaluations have shown that unintended overfishing can occur resulting from misconceptions of stock status, which can ultimately have detrimental ecologic and economic impacts [@mazur2023consequences].
+We are therefore interested in analyzing the impact of climate change on the accuracy of abundance estimates derived from NOAA’s ongoing bottom-trawl survey along the East coast.
 
@@
 To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depend on temperature preferences.
-We consider the impact of climate change by comparing simulations that use a repeating temperature pattern to those where temperature increases on average over time.
+We consider the impact of climate change by comparing simulations that use a repeating water temperature pattern to those where temperature increases on average over time.
 In both cases we analyze the ability of stratified random sampling to track population trends through a design-based approach (stratified mean) compared to a model-based approach that allows for the inclusion of covariates (Vector Autoregressive Spatiotemporal Model).
@@
 
-We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on Georges Bank, where movement of each species combine static species-specific habitat preferences with temperature preferences.
-Model dynamics are driven by a time series of temperature gradients that were estimated from data to create simulated data sets for each population where the true biomass is known.
-Using temperature gradients that repeat each year creates data sets with predictable, repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time.
+We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on Georges Bank, where movement of each species combine static species-specific habitat preferences with biologically-based temperature preferences.
+Model dynamics are driven by a time series of temperature gradients to create simulated data sets for each population where the true biomass is known.
+Using temperature gradients that repeat each year creates data sets with repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time.
 We conduct stratified random sampling on our simulation output to mimic the bottom trawl survey and use the samples to compare the ability of contemporary indexing methods to track population trends.
@@
 MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass in each cell in our study area [@dolder2020highly].
-Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits.
+Individual terms in the formulation account for growth of mature adults, mortality (natural and fishing), and the addition of new recruits.
 Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.
@@
 We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow.
-We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences.
+We combined data from both the fall and spring surveys to obscure the influence of temperature to allow the niche model to instead infer habitat preferences.
 Depth and mean sediment size were used as our covariate predictors.
 Estimated depth for the region was obtained from FVCOM [@chen2006unstructured].
-The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method **cite arcmap** using point data collected by the United States Geologic Survey (USGS) [@mcmullen2005usgs].
-Since the values in $Hab_{J,s}$ are required to be between 0 and 1, we transform the spatial estimates from *lrren* to fall between these bounds.
+The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method using point data collected by the United States Geologic Survey (USGS) [@mcmullen2005usgs].
+Since the values in $Hab_{J,s}$ are required to be between 0 and 1, we rescaled the spatial estimates from *lrren* to fall between these bounds.
 See Figure \@ref(fig:hab-plot1) for a visual representation of this process being applied to Cod.
@@
 
-```{r hab-plot, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Static habitat preferences for each species in our population models (Yellotwtail, Cod, Haddock)."}
+```{r hab-plot, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Static habitat preferences for each species in our population models.
+From left to right: Yellotwtail Flounder, Cod, and Haddock."}
 knitr::include_graphics(c("Images/Habitat_3species.png"))
@@
 Weekly estimated temperature data for the region for 2012 was obtained from FVCOM [@chen2006unstructured].
-We chose to repeat estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data.
+We chose to repeat temperature estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data.
 The 2012 data was chosen because it displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values, allowing for a smooth repeating yearly temperature pattern for the constant temperature scenario.
@@
 
-In equation (1), $Hab^2_{J,s}$ is constant for the duration of the simulation, while $Tol_{c,s,wk}$ changes each week.
+In equation (1), $Hab^2_{J,s}$ remains constant for each species for the duration of the simulation, while $Tol_{c,s,wk}$ changes each week with temperature fluctuations.
 Using a temperature gradient that repeats every 52 weeks produces the same spatial preferences in a given week each year, resulting in consistent spatial biomass patterns.
@@
 After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year.
-We sample each strata in the same weeks in which the Spring and Fall surveys take place, and the number of the samples taken reflect true target values for each strata.
-Randomly selected cells were samples each year.
+We sample each strata in the same weeks in which the Spring and Fall surveys take place (weeks 13 & 14 in the Spring and 37 & 38 in the Fall).
+The number of the samples taken reflect true target values for each strata and sampling cells were randomly selected.
 We then use the biomass collected from our samples in contemporary abundance indexing methods to estimate population trends.
-Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.
+Knowing the true population values in our simulations allows us to compute and compare the absolute error calculated from each estimation method.
 
@@
 
-Stock assessment scientists use samples from the bottom trawl survey to estimate the abundance for each fish stock.
-There are a number of approaches to obtain abundance estimates including traditional design-based estimates to model-based estimates that range in complexity.
+Stock assessment scientists choose from a number of approaches to obtain abundance estimates that are derived from the survey data.
+Methods range from traditional design-based estimates to model-based estimates that vary in complexity.
 Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest.
-These methods do not account for spatial variation in samples or allow the inclusion of environmental influences.
+These methods do not account for spatial variation in samples and are not able to account for environmental influences on survey values.
 Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors).
@@
 
-We follow the advice given in [@thorson2019guidance] to build VAST models to estimate biomass on Georges Bank using stratified random samples from our model output.
+We follow the advice given in [@thorson2019guidance] to build VAST models to estimate biomass in our Georges Bank population models using stratified random samples from our model output.
 In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
-We also carried out the same process running models without covariate information as well as including covariates in our model.
-We considered covariates in the form of dynamic temperature values and/or static habitat values ($Hab_{J.s}$) from our population model.
-When using covariates we ultimately decided to provide the most information to the model by including both temperature and habitat covariates for both linear predictors.
+We carried out the same model-building process running models without covariate information as well as including covariates in our model.
+We considered covariates in the form of dynamic temperature values and/or static habitat values ($Hab_{J.s}$) from our population models.
+When using covariates we ultimately decided to consider a best-case scenario by including both temperature and habitat covariates for both linear predictors in order to provide the most information to the model.
 Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings.
@@
 
-Our goal is to determine indexing approaches and settings that are robust to future environmental conditions and resulting spatial biomass patterns.
+Our goal is to determine indexing approaches and settings that are robust to changing environmental conditions and resulting spatial biomass patterns.
 An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists.
@@
 **BL: Help with correct notation for adding noise**.
-We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species.
+The impication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population.
+We therefore simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species.
 We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from.
 Tabl XXX [[needs to be added]] lists all strata inhabited by each species, those that are removed from certain calculations, and the explanation of why these strata were removed.
+The yellow regions in Figure \@ref(fig:strata-plot) depict the strata that were removed from certain stratified random surveys for each species.
 
 <!-- Note the Table XXX needs to be added comment above. -->
-When combining population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, covariates) there are a large number of scenario combinations to consider.
+Each scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, covariates), resulting in a large number of scenario combinations to consider.
 The columns in Table \@ref(tab:scenarios) show the choices that define each scenario.
@@
 
-The goal of our project was to analyze the ability of contemporary population indexing methods to track population trends under a variety of conditions, as depicted in Table \@ref(tab:scenarios).
-See Figure \@ref(fig:strata-plot) for a spatial reference of the Georges Bank strata and Figure \@ref(fig:hab-plot) for the static habitat preferences for each species ($Hab^2_{J,s}$) used in our population models.
+The goal of our project was to analyze the ability of contemporary population indexing methods to track population trends under the variety of conditions shown in Table \@ref(tab:scenarios).
+and Figure \@ref(fig:hab-plot) for the static habitat preferences for each species ($Hab^2_{J,s}$) used in our population models.
 
 Figure \@ref(fig:PopPct) depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey weeks in the spring (13 and 14) and fall (37 and 28).
-The left column depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern.
+See Figure \@ref(fig:strata-plot) for a spatial reference of the Georges Bank strata.
+The left column in Figure \@ref(fig:PopPct) depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern.
 In these scenarios we notice a small amount of shifting between successive years as the population aggregates on especially suitable habitat in the domain.
 More exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time, as seen right column of Figure \@ref(fig:PopPct).
-This figure demonstrates how we were able to mimic shifting distributions due to climate change by running simulations with a temperature gradient that increased on average over time.
+This demonstrates how we were able to model shifting distributions due to climate change by running simulations with a temperature gradient that increased on average over time.
 
-Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the results of our comparison between the absolute error between abundance estimates and model output.
-While the model-based results tended to provide a lower relative error to the stratified mean with a mean relative error of 0.34 compared to 0.38, the model-based estimates were very sensitive to the settings used.
-As a result, although $77/80\approx 96\%$ of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the variance for the VAST relative errors was much larger than the stratified mean (0.10 compared to 0.02).
+Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the absolute error between abundance estimates and model output for each of our abundance estimates.
+While the model-based results tended to provide a slightly lower mean absolute error compared to the stratified mean with a mean relative error of 0.34 compared to 0.38, the variance seen in VAST relative errors was much larger than the stratified mean (0.10 compared to 0.02).
 VAST models that include covariate information provided the lowest overall errors and standard deviations.
+As a result, while $77/80\approx 96\%$ of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the high variance in error values indicate how sensitive model-based estimates were to the settings used.
 When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.
@@
 These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail Flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12.
-These spring dynamics are therefore related to the static habitat values in our model rather than the dynamics temperature preferences, wihch is why we seen the same dynamics with constant and increasing temperature.
+These spring dynamics are therefore related to the static habitat values in our model rather than the dynamics temperature preferences, which is why we seen the same dynamics with constant and increasing temperature.
 While we observe similar dynamics in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \@ref(fig:PopPct)).
@@
 
-In considering the Yellowtail Flounder results in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata), we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
+Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata) contain the absolute error between our abundance estimates for Yellowtail Flounder comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B).
+In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
 When all strata are sampled, adding covariates greatly improved estimates in all scenarios.
 We still see an improvement in VAST estimates when certain strata are excluded from sampling, but the improvement was much less dramatic than with all strata included.
-However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata).
-Including covariate information to each VAST model produced estimates with significantly lower error than their stratified mean counterparts.
+There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata).
+However, Including covariate information to each VAST model produced estimates with significantly lower error than their stratified mean counterparts.
 
-Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata) contain the absolute error between our abundance estimates for Yellowtail Flounder comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B).
 Of all scenarios without covariates, $33/48\approx68$\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
@@
 
-<font size="5">*Model / Estimate Ratio Results*</font>
+<font size="5">*Abundance Estimate Ratio Results*</font>
 
-A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, too low.
-This can be further examined by viewing estimate/model ratio plots, where we divide the yearly abundance estimates by the true model value and plot them on the log scale.
+A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values.
+This can be further examined by viewing yearly estimate/model ratio plots, where we divide the yearly abundance estimates by the true model value and plot them on the log scale.
 In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can range as low as 0 and exhibit large yearly changes.
-A representative example of this trend is shown in Figure \@ref(fig:RatioTrend)
+A representative example of this trend is shown in Figure \@ref(fig:RatioTrend).
 
-A more careful analysis of individual yearly model/estimate ratios when all strata are included revealed that about 73\% of all VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21.
+Analysis of individual yearly model/estimate ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21.
 On the other hand, 33\% of stratified mean estimates were above 1 with an average of 0.874 and a standard deviation of 0.12.
-There were seasonal differences in estimate ratios with spring VAST ratios producing (1.08,0.12), fall VAST ratios being (1.50,0.38), stratified mean spring ratios resulting in (0.91,0.10), and fall values of (0.84,0.16).
+There were seasonal differences in estimate ratios for VAST with spring VAST ratios producing a mean value of 1.08 with a standard deviation of 0.12 while fall VAST ratios were larger with a mean of 1.50 and standard deviation of 0.38.
+Stratified mean ratios were more consistent between seasons with spring ratios resulting in a mean value of 0.91 with a standard deviation of0.10, and fall values of providing a mean ratio of 0.84 with a standard deviation of 0.16.
 The breakdown for individual species followed a similar pattern.
@@
 When the entire domain is sampled, adding covariates brings the estimate ratio closer to 1.
+Specifically, when including all strata, the stratified mean produced an average yearly ration of 0.87 with a standard deviation of 0.12, while adding covariates improved the VAST estimate ratio from a mean of 1.45 and standard deviation of 0.17 to a mean of 1.13 and standard deviation of 0.09.
 However, when failing to sample the entire domain, adding covariates can decrease the accuracy of VAST estimates.
-When including all strata, the stratified mean produced (0.87,0.12), while adding covariates improved the VAST estimate ratio from (1.45,0.17) to (1.13,0.09).
-With a reduced number of strata, the stratified mean resulted in our worst summary values of (0.59,0.10), and adding covariates to our VAST models decreased the average ratio results from (1.04,0.33 ) to (0.78,0.19).
+With a reduced number of strata, the stratified mean resulted in our worst ratio estimate with a mean value of 0.59 and standard deviation of 0.10, and adding covariates to our VAST models decreased the average ratio results from a mean of 1.04 and standard devnation of 0.33 to a mean of 0.78 and standard deviation of 0.19.
 
@@
 As see in Table **cite VAST setting table**, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
-While using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates with $77/80\approx 96\%$ of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate, there is a great deal more to explore with respect to the impact of covariates.
+While using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates with $77/80\approx 96\%$ of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate, there is more to explore with respect to the impact of covariates.
 For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates.
@@
 One could also test the impact of assuming the wrong covarite response function (linear vs normally distributed etc).
-**For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without "perfect" information?
+<!-- **For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without "perfect" information?
 Which covariate is best?
-Good question for future research**
+Good question for future research** -->
 
-As noted in results second, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates, while adding covariate information leads to a decrease in performance for fall estimates of Haddock and Cod when a reduced number of strata are sampled.
+As noted in our results, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates, when a reduced number of strata are sampled adding covariate information leads to a decrease in performance for fall estimates of Haddock and Cod.
 The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response.
 Figure \@ref(fig:strata-plot) shows the strata excluded from sampling and Figure \@ref(fig:CovResponse) depicts the resulting covariate response for Yellowtail Flounder and Haddock for the same population scenario.
-Although strata were excluded for both species, the samples for Yellowtail Flounder ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata).
-As a result, when projecting into those strata as seen in the lower portion of Figure \@ref(fig:CovResponse), Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend at the wrong magnitude.
-
-Generally help, but dependent on appropriate covariate response.
-Yellowtail cases always had appropriate response and thus adding covariates always improved estimates.
-VAST showed decreased performance with Cod and Haddock when covariates added in fall because covariate response was not fully developed.
-(show plot comparing good vs bad response)
-
-increasing temperature scenarios resulted in a larger increase in average error for VAST estimates compared to stratified mean estimates.
-When including covariates the increase is less than not including them, but stratified mean saw a lower increase in the average error seen in constant an decreasing temperature.
-Shouldn't model based be more stable with respect to increasing temperature?
+Although strata were excluded for both species, the survey samples for Yellowtail Flounder still ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata).
+As a result, when projecting into those strata as seen in the lower portion of Figure \@ref(fig:CovResponse), Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend and estimate the wrong magnitude of biomass.
 
@@
 
-```{r VASTsettings, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
+```{r VASTsettings, echo=FALSE, id="VASTsettings",message=FALSE, warnings=FALSE, results='asis'}
 
@@
 "
+
 cat(tabl) # output the table in a format good for HTML/PDF/docx conversion

### 2023-04-27 Manuscript/Draft1.tex
Manuscript update

@@
 The survey takes place twice each year- once in the spring and again in the fall.
-Since most spatial analyses and projections of future distributions typically assume a constant survey catchability and/or availability over time, NOAA's survey design includes sampling Georges Bank during approximately the same 3-4 week time period in each season.
+Most spatial analyses and projections of future distributions typically assume a constant survey catchability and/or availability over time.
+For this reason, NOAA's survey design includes sampling each strata in approximately the same 3-4 week time period in each season.
 
 Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades.
-The changes have resulted in a shift in spatial distributions of many species (cite Nye).
-Existing research has focused on temperature as the driver of such changes (cite klein, smith and kritzer) and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges (kerr, barajas, wiedenmann).
+The changes have resulted in a shift in spatial distributions of many species (Nye et al.
+2009; Henderson et al.
+2017; Kleisner et al.
+2017).
 Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments.
 More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey will change, altering the relationship between the index and the true population.
-We are therefore interested in analyzing the impact of climate change on the accuracy of future stock assessment models as measured by NOAA's ongoing bottom-trawl survey along the East coast.
+A species shifting its range beyond the survey area is an additional compounding factor to consider.
+Existing research has focused on temperature as the driver of such changes (Klein et al.
+2017) and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges (Kerr et al.
+2022).
+In these scenarios, management strategy evaluations have shown that unintended overfishing can occur resulting from misconceptions of stock status, which can ultimately have detrimental ecologic and economic impacts (Mazur et al.
+2023).
+We are therefore interested in analyzing the impact of climate change on the accuracy of abundance estimates derived from NOAA's ongoing bottom-trawl survey along the East coast.
 
@@
 2020).
-Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits.
+Individual terms in the formulation account for growth of mature adults, mortality (natural and fishing), and the addition of new recruits.
 Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.
@@
 We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow.
-We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences.
+We combined data from both the fall and spring surveys to obscure the influence of temperature to allow the niche model to instead infer habitat preferences.
 Depth and mean sediment size were used as our covariate predictors.
@@
 2006).
-The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method \textbf{cite arcmap} using point data collected by the United States Geologic Survey (USGS) (McMullen et al.
+The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method using point data collected by the United States Geologic Survey (USGS) (McMullen et al.
 n.d.).
-Since the values in \(Hab_{J,s}\) are required to be between 0 and 1, we transform the spatial estimates from \emph{lrren} to fall between these bounds.
+Since the values in \(Hab_{J,s}\) are required to be between 0 and 1, we rescaled the spatial estimates from \emph{lrren} to fall between these bounds.
 See Figure \ref{fig:hab-plot1} for a visual representation of this process being applied to Cod.
@@
 
-\caption{Static habitat preferences for each species in our population models (Yellotwtail, Cod, Haddock).}\label{fig:hab-plot}
+\caption{Static habitat preferences for each species in our population models. From left to right: Yellotwtail Flounder, Cod, and Haddock.}\label{fig:hab-plot}
 \end{figure}
@@
 2006).
-We chose to repeat estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data.
+We chose to repeat temperature estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data.
 The 2012 data was chosen because it displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values, allowing for a smooth repeating yearly temperature pattern for the constant temperature scenario.
@@
 
-In equation (1), \(Hab^2_{J,s}\) is constant for the duration of the simulation, while \(Tol_{c,s,wk}\) changes each week.
+In equation (1), \(Hab^2_{J,s}\) remains constant for each species for the duration of the simulation, while \(Tol_{c,s,wk}\) changes each week with temperature fluctuations.
 Using a temperature gradient that repeats every 52 weeks produces the same spatial preferences in a given week each year, resulting in consistent spatial biomass patterns.
@@
 After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year.
-We sample each strata in the same weeks in which the Spring and Fall surveys take place, and the number of the samples taken reflect true target values for each strata.
-Randomly selected cells were samples each year.
+We sample each strata in the same weeks in which the Spring and Fall surveys take place (weeks 13 \& 14 in the Spring and 37 \& 38 in the Fall).
+The number of the samples taken reflect true target values for each strata and sampling cells were randomly selected.
 We then use the biomass collected from our samples in contemporary abundance indexing methods to estimate population trends.
-Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.
-
-Stock assessment scientists use samples from the bottom trawl survey to estimate the abundance for each fish stock.
-There are a number of approaches to obtain abundance estimates including traditional design-based estimates to model-based estimates that range in complexity.
+Knowing the true population values in our simulations allows us to compute and compare the absolute error calculated from each estimation method.
+
+Stock assessment scientists choose from a number of approaches to obtain abundance estimates that are derived from the survey data.
+Methods range from traditional design-based estimates to model-based estimates that vary in complexity.
 Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest.
-These methods do not account for spatial variation in samples or allow the inclusion of environmental influences.
+These methods do not account for spatial variation in samples and are not able to account for environmental influences on survey values.
 Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors).
@@
 
-We follow the advice given in (Thorson 2019) to build VAST models to estimate biomass on Georges Bank using stratified random samples from our model output.
+We follow the advice given in (Thorson 2019) to build VAST models to estimate biomass in our Georges Bank population models using stratified random samples from our model output.
 In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
-We also carried out the same process running models without covariate information as well as including covariates in our model.
-We considered covariates in the form of dynamic temperature values and/or static habitat values (\(Hab_{J.s}\)) from our population model.
-When using covariates we ultimately decided to provide the most information to the model by including both temperature and habitat covariates for both linear predictors.
+We carried out the same model-building process running models without covariate information as well as including covariates in our model.
+We considered covariates in the form of dynamic temperature values and/or static habitat values (\(Hab_{J.s}\)) from our population models.
+When using covariates we ultimately decided to consider a best-case scenario by including both temperature and habitat covariates for both linear predictors in order to provide the most information to the model.
 Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings.
@@
 \textbf{BL: Help with correct notation for adding noise}.
-We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species.
+The impication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population.
+We therefore simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species.
 We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from.
 Tabl XXX {[}{[}needs to be added{]}{]} lists all strata inhabited by each species, those that are removed from certain calculations, and the explanation of why these strata were removed.
-
-When combining population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, covariates) there are a large number of scenario combinations to consider.
+The yellow regions in Figure \ref{fig:strata-plot} depict the strata that were removed from certain stratified random surveys for each species.
+
+Each scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, covariates), resulting in a large number of scenario combinations to consider.
 The columns in Table \ref{tab:scenarios} show the choices that define each scenario.
@@
 
-The goal of our project was to analyze the ability of contemporary population indexing methods to track population trends under a variety of conditions, as depicted in Table \ref{tab:scenarios}.
-See Figure \ref{fig:strata-plot} for a spatial reference of the Georges Bank strata and Figure \ref{fig:hab-plot} for the static habitat preferences for each species (\(Hab^2_{J,s}\)) used in our population models.
+The goal of our project was to analyze the ability of contemporary population indexing methods to track population trends under the variety of conditions shown in Table \ref{tab:scenarios}.
+and Figure \ref{fig:hab-plot} for the static habitat preferences for each species (\(Hab^2_{J,s}\)) used in our population models.
 
 Figure \ref{fig:PopPct} depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey weeks in the spring (13 and 14) and fall (37 and 28).
-The left column depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern.
+See Figure \ref{fig:strata-plot} for a spatial reference of the Georges Bank strata.
+The left column in Figure \ref{fig:PopPct} depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern.
 In these scenarios we notice a small amount of shifting between successive years as the population aggregates on especially suitable habitat in the domain.
 More exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time, as seen right column of Figure \ref{fig:PopPct}.
-This figure demonstrates how we were able to mimic shifting distributions due to climate change by running simulations with a temperature gradient that increased on average over time.
-
-Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the results of our comparison between the absolute error between abundance estimates and model output.
-While the model-based results tended to provide a lower relative error to the stratified mean with a mean relative error of 0.34 compared to 0.38, the model-based estimates were very sensitive to the settings used.
-As a result, although \(77/80\approx 96\%\) of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the variance for the VAST relative errors was much larger than the stratified mean (0.10 compared to 0.02).
+This demonstrates how we were able to model shifting distributions due to climate change by running simulations with a temperature gradient that increased on average over time.
+
+Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the absolute error between abundance estimates and model output for each of our abundance estimates.
+While the model-based results tended to provide a slightly lower mean absolute error compared to the stratified mean with a mean relative error of 0.34 compared to 0.38, the variance seen in VAST relative errors was much larger than the stratified mean (0.10 compared to 0.02).
 VAST models that include covariate information provided the lowest overall errors and standard deviations.
+As a result, while \(77/80\approx 96\%\) of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the high variance in error values indicate how sensitive model-based estimates were to the settings used.
 When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.
@@
 These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail Flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12.
-These spring dynamics are therefore related to the static habitat values in our model rather than the dynamics temperature preferences, wihch is why we seen the same dynamics with constant and increasing temperature.
+These spring dynamics are therefore related to the static habitat values in our model rather than the dynamics temperature preferences, which is why we seen the same dynamics with constant and increasing temperature.
 While we observe similar dynamics in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \ref{fig:PopPct}).
@@
 
-In considering the Yellowtail Flounder results in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}, we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
+Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata} contain the absolute error between our abundance estimates for Yellowtail Flounder comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B).
+In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
 When all strata are sampled, adding covariates greatly improved estimates in all scenarios.
 We still see an improvement in VAST estimates when certain strata are excluded from sampling, but the improvement was much less dramatic than with all strata included.
-However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}.
-Including covariate information to each VAST model produced estimates with significantly lower error than their stratified mean counterparts.
-
-Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata} contain the absolute error between our abundance estimates for Yellowtail Flounder comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B).
+There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}.
+However, Including covariate information to each VAST model produced estimates with significantly lower error than their stratified mean counterparts.
+
 Of all scenarios without covariates, \(33/48\approx68\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
@@
 
-\emph{Model / Estimate Ratio Results}
-
-A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, too low.
-This can be further examined by viewing estimate/model ratio plots, where we divide the yearly abundance estimates by the true model value and plot them on the log scale.
+\emph{Abundance Estimate Ratio Results}
+
+A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values.
+This can be further examined by viewing yearly estimate/model ratio plots, where we divide the yearly abundance estimates by the true model value and plot them on the log scale.
 In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can range as low as 0 and exhibit large yearly changes.
-A representative example of this trend is shown in Figure \ref{fig:RatioTrend}
-
-A more careful analysis of individual yearly model/estimate ratios when all strata are included revealed that about 73\% of all VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21.
+A representative example of this trend is shown in Figure \ref{fig:RatioTrend}.
+
+Analysis of individual yearly model/estimate ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21.
 On the other hand, 33\% of stratified mean estimates were above 1 with an average of 0.874 and a standard deviation of 0.12.
-There were seasonal differences in estimate ratios with spring VAST ratios producing (1.08,0.12), fall VAST ratios being (1.50,0.38), stratified mean spring ratios resulting in (0.91,0.10), and fall values of (0.84,0.16).
+There were seasonal differences in estimate ratios for VAST with spring VAST ratios producing a mean value of 1.08 with a standard deviation of 0.12 while fall VAST ratios were larger with a mean of 1.50 and standard deviation of 0.38.
+Stratified mean ratios were more consistent between seasons with spring ratios resulting in a mean value of 0.91 with a standard deviation of0.10, and fall values of providing a mean ratio of 0.84 with a standard deviation of 0.16.
 The breakdown for individual species followed a similar pattern.
@@
 As see in Table \textbf{cite VAST setting table}, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
-While using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates with \(77/80\approx 96\%\) of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate, there is a great deal more to explore with respect to the impact of covariates.
+While using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates with \(77/80\approx 96\%\) of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate, there is more to explore with respect to the impact of covariates.
 For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates.
@@
 One could also test the impact of assuming the wrong covarite response function (linear vs normally distributed etc).
-\textbf{For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without ``perfect'' information?
-Which covariate is best?
-Good question for future research}
-
-As noted in results second, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates, while adding covariate information leads to a decrease in performance for fall estimates of Haddock and Cod when a reduced number of strata are sampled.
+
+As noted in our results, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates, when a reduced number of strata are sampled adding covariate information leads to a decrease in performance for fall estimates of Haddock and Cod.
 The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response.
 Figure \ref{fig:strata-plot} shows the strata excluded from sampling and Figure \ref{fig:CovResponse} depicts the resulting covariate response for Yellowtail Flounder and Haddock for the same population scenario.
-Although strata were excluded for both species, the samples for Yellowtail Flounder ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata).
-As a result, when projecting into those strata as seen in the lower portion of Figure \ref{fig:CovResponse}, Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend at the wrong magnitude.
-
-Generally help, but dependent on appropriate covariate response.
-Yellowtail cases always had appropriate response and thus adding covariates always improved estimates.
-VAST showed decreased performance with Cod and Haddock when covariates added in fall because covariate response was not fully developed.
-(show plot comparing good vs bad response)
-
-increasing temperature scenarios resulted in a larger increase in average error for VAST estimates compared to stratified mean estimates.
-When including covariates the increase is less than not including them, but stratified mean saw a lower increase in the average error seen in constant an decreasing temperature.
-Shouldn't model based be more stable with respect to increasing temperature?
+Although strata were excluded for both species, the survey samples for Yellowtail Flounder still ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata).
+As a result, when projecting into those strata as seen in the lower portion of Figure \ref{fig:CovResponse}, Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend and estimate the wrong magnitude of biomass.
 
@@
 Dolder, P.J., Minto, C., Guarini, J.-M., and Poos, J.J. 2020. Highly resolved spatiotemporal simulations for exploring mixed fishery dynamics. Ecological Modelling \textbf{424}: 109000. Elsevier.
+
+\leavevmode\vadjust pre{\hypertarget{ref-henderson2017effects}{}}%
+Henderson, M.E., Mills, K.E., Thomas, A.C., Pershing, A.J., and Nye, J.A.
+2017.
+Effects of spring onset and summer duration on fish species distribution and biomass along the northeast united states continental shelf.
+Reviews in Fish Biology and Fisheries \textbf{27}: 411--424.
+Springer.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kerr2022coherence}{}}%
+Kerr, L., Barajas, M., and Wiedenmann, J.
+2022.
+Coherence and potential drivers of stock assessment uncertainty in northeast US groundfish stocks.
+ICES Journal of Marine Science \textbf{79}(8): 2217--2230.
+Oxford University Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-klein2017effects}{}}%
+Klein, E.S., Smith, S.L., and Kritzer, J.P. 2017. Effects of climate change on four new england groundfish species. Reviews in Fish Biology and Fisheries \textbf{27}: 317--338. Springer.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kleisner2017marine}{}}%
+Kleisner, K.M., Fogarty, M.J., McGee, S., Hare, J.A., Moret, S., Perretti, C.T., and Saba, V.S.
+2017.
+Marine species distribution shifts on the US northeast continental shelf under continued ocean warming.
+Progress in Oceanography \textbf{153}: 24--36.
+Elsevier.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mazur2023consequences}{}}%
+Mazur, M.D., Jesse, J., Cadrin, S.X., Truesdell, S., and Kerr, L.
+2023.
+Consequences of ignoring climate impacts on new england groundfish stock assessment and management.
+Fisheries Research \textbf{262}: 106652.
+Elsevier.
 
@@
 Available from: National Marine Fisheries Service, 166 Water Street, Woods Hole, MA 02543-1026, or online at http://www.nefsc.noaa.gov/nefsc/publications/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nye2009changing}{}}%
+Nye, J.A., Link, J.S., Hare, J.A., and Overholtz, W.J.
+2009.
+Changing spatial distribution of fish stocks in relation to climate and population size on the northeast united states continental shelf.
+Marine Ecology Progress Series \textbf{393}: 111--129.
 

### 2023-05-03 Manuscript/Draft1.Rmd
manuscript update

@@
 
+<font size="5">*Abundance Estimate Ratio Results*</font>
+
+A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values.
+This can be further examined by examining yearly estimate:model ratio values, where we divide the yearly abundance estimates by the true model value.
+In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can estimate yearly abundance values as low as 0 at times and exhibit large yearly changes.
+A representative example of this trend for each species is shown on the log scale in Figure \@ref(fig:RatioTrend).
+
+Analysis of individual yearly model:estimate ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21.
+On the other hand, just 33\% of stratified mean estimate ratios were above 1 with an average of 0.874 and a standard deviation of 0.12.
+There were seasonal differences in estimate ratios for VAST with spring VAST ratios producing a mean value of 1.08 with a standard deviation of 0.12 while fall VAST ratios were larger with a mean of 1.50 and standard deviation of 0.38.
+Stratified mean ratios were more consistent between seasons with spring ratios resulting in a mean value of 0.91 with a standard deviation of 0.10, and fall values of providing a mean ratio of 0.84 with a standard deviation of 0.16.
+The breakdown for individual species followed a similar pattern.
+
+When the entire domain is sampled, the stratified mean produced an average yearly ratio of 0.87 with a standard deviation of 0.12.
+Adding covariates in these scenarios brings the VAST estimate ratio closer to 1.
+Specifically, when all strata are sampled, adding covariates improved the VAST estimate ratio from a mean of 1.45 and standard deviation of 0.17 to a mean of 1.13 and standard deviation of 0.09.
+Failing to sample the entire domain predictably decreases each individual yearly estimate as the entire population is not being accounted for in the sampling process, which in tern decreases the corresponding estimate ratio.
+For example, failing to sample the entire domain decreased the estimate ratio for the stratified mean from 0.87 to 0.59 (standard deviation of 0.10).
+With a reduced number of strata, adding covariates to our VAST models decreased the average ratio results from a mean of 1.04 and standard deviation of 0.33 to a mean of 0.78 and standard deviation of 0.19.
+As discussed later, when we fail to sample the entire domain in VAST models adding covariates sometimes decreased the accuracy of VAST estimates, typically in the form of additional yearly overestimation.
+
 <font size="5">*Yellowtail Flounder Results*</font>
@@
 
-<font size="5">*Abundance Estimate Ratio Results*</font>
-
-A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values.
-This can be further examined by viewing yearly estimate/model ratio plots, where we divide the yearly abundance estimates by the true model value and plot them on the log scale.
-In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can range as low as 0 and exhibit large yearly changes.
-A representative example of this trend is shown in Figure \@ref(fig:RatioTrend).
-
-Analysis of individual yearly model/estimate ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21.
-On the other hand, 33\% of stratified mean estimates were above 1 with an average of 0.874 and a standard deviation of 0.12.
-There were seasonal differences in estimate ratios for VAST with spring VAST ratios producing a mean value of 1.08 with a standard deviation of 0.12 while fall VAST ratios were larger with a mean of 1.50 and standard deviation of 0.38.
-Stratified mean ratios were more consistent between seasons with spring ratios resulting in a mean value of 0.91 with a standard deviation of0.10, and fall values of providing a mean ratio of 0.84 with a standard deviation of 0.16.
-The breakdown for individual species followed a similar pattern.
-
-When the entire domain is sampled, adding covariates brings the estimate ratio closer to 1.
-Specifically, when including all strata, the stratified mean produced an average yearly ration of 0.87 with a standard deviation of 0.12, while adding covariates improved the VAST estimate ratio from a mean of 1.45 and standard deviation of 0.17 to a mean of 1.13 and standard deviation of 0.09.
-However, when failing to sample the entire domain, adding covariates can decrease the accuracy of VAST estimates.
-With a reduced number of strata, the stratified mean resulted in our worst ratio estimate with a mean value of 0.59 and standard deviation of 0.10, and adding covariates to our VAST models decreased the average ratio results from a mean of 1.04 and standard devnation of 0.33 to a mean of 0.78 and standard deviation of 0.19.
-
-<font size="5">*Differences Between VAST Settings*</font>
+<font size="5">*Differences Between VAST Settings (mostly notes, not fully formed)*</font>
 
@@
 As see in Table **cite VAST setting table**, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
-While using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates with $77/80\approx 96\%$ of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate, there is more to explore with respect to the impact of covariates.
+Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with $77/80\approx 96\%$ of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate.
+**count how many w/ cov better than w/ out cov?** There is more to explore with respect to the impact of covariates.
 For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates.
 Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information.
-One could also test the impact of assuming the wrong covarite response function (linear vs normally distributed etc).
+One could also test the impact of assuming the wrong covariate response function (linear vs normally distributed etc).
 <!-- **For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without "perfect" information?
@@
 
-As noted in our results, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates, when a reduced number of strata are sampled adding covariate information leads to a decrease in performance for fall estimates of Haddock and Cod.
+As noted in our results, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates.
+However, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates.
 The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response.
@@
 
-```{r RatioTrend, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Representative example of typical ratio trend as shown on a log scale.",out.width="90%"}
+```{r RatioTrend, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Representative example of typical ratio trend for each species, as shown on a log scale.",out.width="90%"}
 # knitr::include_graphics(c("Images/covariate_good.png","Images/covariate_bad.png"))
 # knitr::include_graphics(c("Images/covariate_est.png"))
+#Yellowtail estimates are for the constant population, constant temperature, all strata used in estimates.
+#Cod are Decpop_ConTemp_allStrata_
+#Haddock are IncPop_IncTemp_all strata_
 
-knitr::include_graphics(c("Images/ratio_trend.png"))
-
+knitr::include_graphics(c("Images/ratio_trend_YTF.PNG"))
+knitr::include_graphics(c("Images/ratio_trend_cod.PNG"))
+knitr::include_graphics(c("Images/ratio_trend_had.PNG"))
 ```

### 2023-05-03 Manuscript/Draft1.tex
manuscript update

@@
 
+\emph{Abundance Estimate Ratio Results}
+
+A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values.
+This can be further examined by examining yearly estimate:model ratio values, where we divide the yearly abundance estimates by the true model value.
+In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can estimate yearly abundance values as low as 0 at times and exhibit large yearly changes.
+A representative example of this trend for each species is shown on the log scale in Figure \ref{fig:RatioTrend}.
+
+Analysis of individual yearly model:estimate ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21.
+On the other hand, just 33\% of stratified mean estimate ratios were above 1 with an average of 0.874 and a standard deviation of 0.12.
+There were seasonal differences in estimate ratios for VAST with spring VAST ratios producing a mean value of 1.08 with a standard deviation of 0.12 while fall VAST ratios were larger with a mean of 1.50 and standard deviation of 0.38.
+Stratified mean ratios were more consistent between seasons with spring ratios resulting in a mean value of 0.91 with a standard deviation of 0.10, and fall values of providing a mean ratio of 0.84 with a standard deviation of 0.16.
+The breakdown for individual species followed a similar pattern.
+
+When the entire domain is sampled, the stratified mean produced an average yearly ratio of 0.87 with a standard deviation of 0.12.
+Adding covariates in these scenarios brings the VAST estimate ratio closer to 1.
+Specifically, when all strata are sampled, adding covariates improved the VAST estimate ratio from a mean of 1.45 and standard deviation of 0.17 to a mean of 1.13 and standard deviation of 0.09.
+Failing to sample the entire domain predictably decreases each individual yearly estimate as the entire population is not being accounted for in the sampling process, which in tern decreases the corresponding estimate ratio.
+For example, failing to sample the entire domain decreased the estimate ratio for the stratified mean from 0.87 to 0.59 (standard deviation of 0.10).
+With a reduced number of strata, adding covariates to our VAST models decreased the average ratio results from a mean of 1.04 and standard deviation of 0.33 to a mean of 0.78 and standard deviation of 0.19.
+As discussed later, when we fail to sample the entire domain in VAST models adding covariates sometimes decreased the accuracy of VAST estimates, typically in the form of additional yearly overestimation.
+
 \emph{Yellowtail Flounder Results}
@@
 
-\emph{Abundance Estimate Ratio Results}
-
-A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values.
-This can be further examined by viewing yearly estimate/model ratio plots, where we divide the yearly abundance estimates by the true model value and plot them on the log scale.
-In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can range as low as 0 and exhibit large yearly changes.
-A representative example of this trend is shown in Figure \ref{fig:RatioTrend}.
-
-Analysis of individual yearly model/estimate ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21.
-On the other hand, 33\% of stratified mean estimates were above 1 with an average of 0.874 and a standard deviation of 0.12.
-There were seasonal differences in estimate ratios for VAST with spring VAST ratios producing a mean value of 1.08 with a standard deviation of 0.12 while fall VAST ratios were larger with a mean of 1.50 and standard deviation of 0.38.
-Stratified mean ratios were more consistent between seasons with spring ratios resulting in a mean value of 0.91 with a standard deviation of0.10, and fall values of providing a mean ratio of 0.84 with a standard deviation of 0.16.
-The breakdown for individual species followed a similar pattern.
-
-When the entire domain is sampled, adding covariates brings the estimate ratio closer to 1.
-Specifically, when including all strata, the stratified mean produced an average yearly ration of 0.87 with a standard deviation of 0.12, while adding covariates improved the VAST estimate ratio from a mean of 1.45 and standard deviation of 0.17 to a mean of 1.13 and standard deviation of 0.09.
-However, when failing to sample the entire domain, adding covariates can decrease the accuracy of VAST estimates.
-With a reduced number of strata, the stratified mean resulted in our worst ratio estimate with a mean value of 0.59 and standard deviation of 0.10, and adding covariates to our VAST models decreased the average ratio results from a mean of 1.04 and standard devnation of 0.33 to a mean of 0.78 and standard deviation of 0.19.
-
-\emph{Differences Between VAST Settings}
+\emph{Differences Between VAST Settings (mostly notes, not fully formed)}
 
@@
 As see in Table \textbf{cite VAST setting table}, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
-While using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates with \(77/80\approx 96\%\) of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate, there is more to explore with respect to the impact of covariates.
+Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with \(77/80\approx 96\%\) of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate.
+\textbf{count how many w/ cov better than w/ out cov?} There is more to explore with respect to the impact of covariates.
 For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates.
 Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information.
-One could also test the impact of assuming the wrong covarite response function (linear vs normally distributed etc).
-
-As noted in our results, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates, when a reduced number of strata are sampled adding covariate information leads to a decrease in performance for fall estimates of Haddock and Cod.
+One could also test the impact of assuming the wrong covariate response function (linear vs normally distributed etc).
+
+As noted in our results, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates.
+However, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates.
 The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response.
@@
 
-{\centering \includegraphics[width=0.9\linewidth]{Images/ratio_trend}
+{\centering \includegraphics[width=0.9\linewidth]{Images/ratio_trend_YTF} \includegraphics[width=0.9\linewidth]{Images/ratio_trend_cod} \includegraphics[width=0.9\linewidth]{Images/ratio_trend_had}
 
@@
 
-\caption{Representative example of typical ratio trend as shown on a log scale.}\label{fig:RatioTrend}
+\caption{Representative example of typical ratio trend for each species, as shown on a log scale.}\label{fig:RatioTrend}
 \end{figure}

### 2023-05-05 Manuscript/Draft1.Rmd
manuscript update

@@
 As see in Table **cite VAST setting table**, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
-Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with $77/80\approx 96\%$ of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate.
-**count how many w/ cov better than w/ out cov?** There is more to explore with respect to the impact of covariates.
+Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73% had a lower absolute error compared to the corresponding estimate without covariates ($117/160\approx 73\%$), and $77/80\approx 96\%$ of VAST models with covariates provided lower absolute error than the corresponding stratified mean estimate.
+There is more to explore with respect to the impact of covariates.
 For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates.
 Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information.
-One could also test the impact of assuming the wrong covariate response function (linear vs normally distributed etc).
+One could also test the impact of assuming the wrong covariate response function (linear vs polynomial etc).
 <!-- **For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without "perfect" information?
@@
 
-As noted in our results, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates.
+ Of the ~27% of VAST models where covariates did not improve the estimate, most provided a comparable error (for ex, 0.11 vs 0.13).
+The instances where including covariates provided a significant different in error took place scenarios that included either an increasing temperature and/or reduced sampling domain.
+More specifically, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates.
 However, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates.
@@
 
-Estimation methods that have large, inaccurate swings (stratified mean) can lead to changes in quotas that do not correspond to the true population trend, which could have a compounding effect (can lead to quotas that are too high/low given an incorrect assumption of increase/decrease in biomass).
-Our model has a constant assumed mortality that accounts for fishing and natural death and will not account for impacts of these decisions.
+Estimation methods that produce large variation between yearly estimates as displayed by the stratified mean can potentially lead to changes in catch limits that do not correspond to the true population trend, which could have a compounding effect.
+For example, a large, increasing biomass estimate when the population has actually decreased and is fairly low could potentially lead to a windfall catch limit that further reduces the total biomass available the following year.
+A second overestimate the following year could then have a detrimental impact by reducing the population even further.
+Our population model has assumed a constant total mortality that accounts for both fishing and natural death, and therefore will not account for impacts of such management decisions.
+This type of question can be best explored with a management strategy evaluation.
+
 VAST may provide more consistent biomass predictions(?)

### 2023-05-05 Manuscript/Draft1.tex
manuscript update

@@
 As see in Table \textbf{cite VAST setting table}, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
-Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with \(77/80\approx 96\%\) of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate.
-\textbf{count how many w/ cov better than w/ out cov?} There is more to explore with respect to the impact of covariates.
+Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73\% had a lower absolute error compared to the corresponding estimate without covariates (\(117/160\approx 73\%\)), and \(77/80\approx 96\%\) of VAST models with covariates provided lower absolute error than the corresponding stratified mean estimate.
+There is more to explore with respect to the impact of covariates.
 For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates.
 Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information.
-One could also test the impact of assuming the wrong covariate response function (linear vs normally distributed etc).
-
-As noted in our results, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates.
+One could also test the impact of assuming the wrong covariate response function (linear vs polynomial etc).
+
+Of the \textasciitilde27\% of VAST models where covariates did not improve the estimate, most provided a comparable error (for ex, 0.11 vs 0.13).
+The instances where including covariates provided a significant different in error took place scenarios that included either an increasing temperature and/or reduced sampling domain.
+More specifically, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates.
 However, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates.

### 2023-05-23 Manuscript/Draft1.Rmd
manuscript update

@@
 Commercial fisheries have been an important part of local economies for centuries.
-In 2019, New England fisheries produced $22 billion in sales, which sustained over 200,000 jobs.
+In 2019, New England fisheries produced $22 billion in sales, which sustained over 200,000 jobs [@Econ].
 Maintaining a healthy ecosystem is therefore vital to sustained ecological health and economic prosperity of the region.
@@
 The survey takes place twice each year- once in the spring and again in the fall.
-Most spatial analyses and  projections of future distributions typically assume a constant survey catchability and/or availability over time.
+Most spatial analyses and  projections of future distributions typically assume a constant survey catchability and availability over time.
 For this reason, NOAA's survey design includes sampling each strata in approximately the same 3-4 week time period in each season.
@@
 Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments.
-More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey will change, altering the relationship between the index and the true population.
+More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey will change, altering the relationship between the index and the true population [@arreguin1996catchability; @langan2021climate].
 A species shifting its range beyond the survey area is an additional compounding factor to consider.
@@
 We are therefore interested in analyzing the impact of climate change on the accuracy of abundance estimates derived from NOAA’s ongoing bottom-trawl survey along the East coast.
-
-**use more info from initial proposal**
 
@@
 
-```{r strata-plot, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Strata inhabited by each species in our population models."}
+```{r strata-plot, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Strata inhabited by each species in our population models.
+All strata on the eastern US contintental shelf are shown in the first image with Georges Bank highlighted in orange.
+Stratum numbers used by the NEFSC bottom trawl survey are shown in the middle row.
+Strata that are excluded from certain stratified random sampling are shown in the bottom row in yellow."}
 #knitr::include_graphics(c("Images/Strata.png"))
 #knitr::include_graphics(c("Images/Excluded_strata.png"))
+knitr::include_graphics(c("Images/GB_strata.jpg"))
 knitr::include_graphics(c("Images/Strata2.png"))
@@
 Spring biomass plots are shown with fall values being very similar."}
-knitr::include_graphics(c("Images/Population_scenarios.png"))
+knitr::include_graphics(c("Images/Population_scenarios2.png"))
 ```
@@
 We compare yearly abundance estimates obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
-The stratified mean is a design-based approach that calculates the geometric mean catch per tow and has traditionally been used with stratified random sample designs.
+The stratified mean is a design-based approach that calculates the arithmetic mean catch per tow and has traditionally been used with stratified random sample designs.
 VAST is a spatial delta-generalized linear mixed model that estimates both abundance (biomass) and probability of occurrence (presence/absence) [@thorson2019guidance].
@@
 An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists.
-These assumptions can be questioned given enough noise in the sampling process **cite?** and/or climate change causing a population to move into previously uninhabited strata.
+These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata.
 To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values.
-**BL: Help with correct notation for adding noise**.
-The impication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population.
+Annual survey observations were simulated as log-normal deviations from the underlying “true” survey catches with a CV of 0.3 in the spring.
+The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population.
 We therefore simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species.
 We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from.
-Tabl XXX [[needs to be added]] lists all strata inhabited by each species, those that are removed from certain calculations, and the explanation of why these strata were removed.
+Figure \ref(fig:strata-plot) shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \ref(fig:PopPct).
 The yellow regions in Figure \@ref(fig:strata-plot) depict the strata that were removed from certain stratified random surveys for each species.
@@
  kable(summary_tab, "simple",
-                     caption = "Each index estimate chooses one condition from each of the following
-7 columns. There are 3∗3∗2∗2∗2∗2∗2 = 288 VAST model combinations and 3∗3∗2∗2∗2∗2 = 144 stratified mean estimates.",
+                     caption = "Each index estimate chooses one condition from each of the following 7 columns.",
           format = "latex", booktabs = TRUE) %>%
@@
 
-The goal of our project was to analyze the ability of contemporary population indexing methods to track population trends under the variety of conditions shown in Table \@ref(tab:scenarios).
+<!-- The goal of our project was to analyze the ability of contemporary population indexing methods to track population trends under the variety of conditions shown in Table \@ref(tab:scenarios).
 and Figure \@ref(fig:hab-plot) for the static habitat preferences for each species ($Hab^2_{J,s}$) used in our population models.
+-->
 
-Figure \@ref(fig:PopPct) depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey weeks in the spring (13 and 14) and fall (37 and 28).
-See Figure \@ref(fig:strata-plot) for a spatial reference of the Georges Bank strata.
+Figure \@ref(fig:PopPct) depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey in the spring (weeks 13 and 14) and fall (weeks 37 and 28).
 The left column in Figure \@ref(fig:PopPct) depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern.
@@
 Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the absolute error between abundance estimates and model output for each of our abundance estimates.
-While the model-based results tended to provide a slightly lower mean absolute error compared to the stratified mean with a mean relative error of 0.34 compared to 0.38, the variance seen in VAST relative errors was much larger than the stratified mean (0.10 compared to 0.02).
-VAST models that include covariate information provided the lowest overall errors and standard deviations.
-As a result, while $77/80\approx 96\%$ of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the high variance in error values indicate how sensitive model-based estimates were to the settings used.
+While the model-based results provided a slightly lower mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02) highlighting the sensitivity of model-based estimates to the settings being used.
+When considering individual scenarios, $77/80\approx 96\%$ had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, with VAST models that include covariate information providing the lowest overall errors and standard deviations.
+Setting misspecification and/or a poor covariate response contribute to the large variance in VAST errors as illustrated by the fact that $\approx 63\%$ of individual scenarios contain a VAST estimate with a worst error than the corresponding stratified mean estimate.
+VAST models that included covariates had an average absolute error of 0.22, models that did not include covariate information had an average error of 0.46, and stratified mean estimates produces an average error of 0.39.
 When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.
-Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain.
+Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain (e.g.
+in Table \@ref(tab:HadERROR) rows 8 and 9, VAST models without covariates show lower errors than when covariates are included).
+We analyze these results further in the sections that follow.
 
@@
 Specifically, when all strata are sampled, adding covariates improved the VAST estimate ratio from a mean of 1.45 and standard deviation of 0.17 to a mean of 1.13 and standard deviation of 0.09.
-Failing to sample the entire domain predictably decreases each individual yearly estimate as the entire population is not being accounted for in the sampling process, which in tern decreases the corresponding estimate ratio.
+Failing to sample the entire domain predictably decreases each individual yearly estimate as the entire population is not being accounted for in the sampling process, which in turn decreases the corresponding estimate ratio.
 For example, failing to sample the entire domain decreased the estimate ratio for the stratified mean from 0.87 to 0.59 (standard deviation of 0.10).
@@
 These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail Flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12.
-These spring dynamics are therefore related to the static habitat values in our model rather than the dynamics temperature preferences, which is why we seen the same dynamics with constant and increasing temperature.
-While we observe similar dynamics in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \@ref(fig:PopPct)).
+These spring changes are therefore related to the static habitat values in our model rather than the temperature preferences, which is why we seen the same dynamics with constant and increasing temperature.
+While we observe similar changes in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in the population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \@ref(fig:PopPct)).
 These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow exterior strata 17 and 18.
@@
 The largest improvements were seen in the increasing temperature scenarios.
-As a result, of the VAST estimates that included covariates information, $47/48\approx98$\% had a lower relative error than the corresponding stratified mean estimate.
+When comparing to the design-based estimates, $47/48\approx98$\% of the VAST estimates that included covariates information had a lower relative error than the corresponding stratified mean estimate.
 This implies that the covariate information helped our model-based estimate account for the design-based issues related to an increasing percentage of the population entering smaller strata, which can become exacerbated in the increasing temperature simulations.
@@
 Specifically, when comparing the error of abundance estimates from the constant temperature scenario to the corresponding increasing temperature scenario for Cod, VAST increased by an average factor of 3.57 without covariates and 1.67 with covariates.
-In considering the average change in the error of stratified mean estimates between constant temperature scenarios and increasing temperature scenarios we surprisingly find an improvement in the error of abundance estimates, with an average decrease by a factir of 0.87.
+In considering the average change in the error of stratified mean estimates between constant temperature scenarios and increasing temperature scenarios we surprisingly find an improvement in the error of abundance estimates, with an average decrease by a factor of 0.87.
 
@@
 
-As see in Table **cite VAST setting table**, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
+As seen in Table **cite VAST setting table**, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
 Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73% had a lower absolute error compared to the corresponding estimate without covariates ($117/160\approx 73\%$), and $77/80\approx 96\%$ of VAST models with covariates provided lower absolute error than the corresponding stratified mean estimate.
@@
 ```{r PopPct, echo=FALSE,fig.show='hold', fig.align='center', fig.cap="Percent of each species in each strata for during survey weeks in our spatial simulations.
-All constant temperature scenarios follow the patterns on the left while increasing temperature scenarios follow the patterns on the right.",out.width="47.5%"}
+All constant temperature scenarios follow the patterns on the left while increasing temperature scenarios follow the patterns on the right.
+See Figure 1??
+for a spatial reference of the Georges Bank strata.",out.width="47.5%"}
 knitr::include_graphics(c("Images/PercentPlots_ConPopConTemp_YT.png","Images/PercentPlots_ConPopIncTemp_YT.png"))

### 2023-05-23 Manuscript/Draft1.tex
manuscript update

@@
 Commercial fisheries have been an important part of local economies for centuries.
-In 2019, New England fisheries produced \$22 billion in sales, which sustained over 200,000 jobs.
+In 2019, New England fisheries produced \$22 billion in sales, which sustained over 200,000 jobs ({``{Fisheries Economics of the United States: Data and Visualizations}''} n.d.).
 Maintaining a healthy ecosystem is therefore vital to sustained ecological health and economic prosperity of the region.
@@
 The survey takes place twice each year- once in the spring and again in the fall.
-Most spatial analyses and projections of future distributions typically assume a constant survey catchability and/or availability over time.
+Most spatial analyses and projections of future distributions typically assume a constant survey catchability and availability over time.
 For this reason, NOAA's survey design includes sampling each strata in approximately the same 3-4 week time period in each season.
@@
 Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments.
-More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey will change, altering the relationship between the index and the true population.
+More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey will change, altering the relationship between the index and the true population (Arreguı́n-Sánchez 1996; Langan et al.
+2021).
 A species shifting its range beyond the survey area is an additional compounding factor to consider.
@@
 We are therefore interested in analyzing the impact of climate change on the accuracy of abundance estimates derived from NOAA's ongoing bottom-trawl survey along the East coast.
-
-\textbf{use more info from initial proposal}
 
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{Images/Strata2} \includegraphics[width=0.95\linewidth]{Images/Excluded_strata2}
+{\centering \includegraphics[width=0.95\linewidth]{Images/GB_strata} \includegraphics[width=0.95\linewidth]{Images/Strata2} \includegraphics[width=0.95\linewidth]{Images/Excluded_strata2}
 
@@
 
-\caption{Strata inhabited by each species in our population models.}\label{fig:strata-plot}
+\caption{Strata inhabited by each species in our population models.
+All strata on the eastern US contintental shelf are shown in the first image with Georges Bank highlighted in orange.
+Stratum numbers used by the NEFSC bottom trawl survey are shown in the middle row.
+Strata that are excluded from certain stratified random sampling are shown in the bottom row in yellow.}\label{fig:strata-plot}
 \end{figure}
@@
 
-{\centering \includegraphics[width=0.99\linewidth]{Images/Population_scenarios}
+{\centering \includegraphics[width=0.99\linewidth]{Images/Population_scenarios2}
 
@@
 We compare yearly abundance estimates obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
-The stratified mean is a design-based approach that calculates the geometric mean catch per tow and has traditionally been used with stratified random sample designs.
+The stratified mean is a design-based approach that calculates the arithmetic mean catch per tow and has traditionally been used with stratified random sample designs.
 VAST is a spatial delta-generalized linear mixed model that estimates both abundance (biomass) and probability of occurrence (presence/absence) (Thorson 2019).
@@
 An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists.
-These assumptions can be questioned given enough noise in the sampling process \textbf{cite?} and/or climate change causing a population to move into previously uninhabited strata.
+These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata.
 To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values.
-\textbf{BL: Help with correct notation for adding noise}.
-The impication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population.
+Annual survey observations were simulated as log-normal deviations from the underlying ``true'' survey catches with a CV of 0.3 in the spring.
+The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population.
 We therefore simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species.
 We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from.
-Tabl XXX {[}{[}needs to be added{]}{]} lists all strata inhabited by each species, those that are removed from certain calculations, and the explanation of why these strata were removed.
+Figure \ref(fig:strata-plot) shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \ref(fig:PopPct).
 The yellow regions in Figure \ref{fig:strata-plot} depict the strata that were removed from certain stratified random surveys for each species.
@@
 
-\caption{\label{tab:scenarios}Each index estimate chooses one condition from each of the following
-7 columns. There are 3*3*2*2*2*2*2 = 288 VAST model combinations and 3*3*2*2*2*2 = 144 stratified mean estimates.}
+\caption{\label{tab:scenarios}Each index estimate chooses one condition from each of the following 7 columns.}
 \centering
@@
 
-The goal of our project was to analyze the ability of contemporary population indexing methods to track population trends under the variety of conditions shown in Table \ref{tab:scenarios}.
-and Figure \ref{fig:hab-plot} for the static habitat preferences for each species (\(Hab^2_{J,s}\)) used in our population models.
-
-Figure \ref{fig:PopPct} depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey weeks in the spring (13 and 14) and fall (37 and 28).
-See Figure \ref{fig:strata-plot} for a spatial reference of the Georges Bank strata.
+Figure \ref{fig:PopPct} depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey in the spring (weeks 13 and 14) and fall (weeks 37 and 28).
 The left column in Figure \ref{fig:PopPct} depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern.
@@
 Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the absolute error between abundance estimates and model output for each of our abundance estimates.
-While the model-based results tended to provide a slightly lower mean absolute error compared to the stratified mean with a mean relative error of 0.34 compared to 0.38, the variance seen in VAST relative errors was much larger than the stratified mean (0.10 compared to 0.02).
-VAST models that include covariate information provided the lowest overall errors and standard deviations.
-As a result, while \(77/80\approx 96\%\) of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the high variance in error values indicate how sensitive model-based estimates were to the settings used.
+While the model-based results provided a slightly lower mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02) highlighting the sensitivity of model-based estimates to the settings being used.
+When considering individual scenarios, \(77/80\approx 96\%\) had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, with VAST models that include covariate information providing the lowest overall errors and standard deviations.
+Setting misspecification and/or a poor covariate response contribute to the large variance in VAST errors as illustrated by the fact that \(\approx 63\%\) of individual scenarios contain a VAST estimate with a worst error than the corresponding stratified mean estimate.
+VAST models that included covariates had an average absolute error of 0.22, models that did not include covariate information had an average error of 0.46, and stratified mean estimates produces an average error of 0.39.
 When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.
-Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain.
+Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain (e.g.~in Table \ref{tab:HadERROR} rows 8 and 9, VAST models without covariates show lower errors than when covariates are included).
+We analyze these results further in the sections that follow.
 
@@
 These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail Flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12.
-These spring dynamics are therefore related to the static habitat values in our model rather than the dynamics temperature preferences, which is why we seen the same dynamics with constant and increasing temperature.
-While we observe similar dynamics in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \ref{fig:PopPct}).
+These spring changes are therefore related to the static habitat values in our model rather than the temperature preferences, which is why we seen the same dynamics with constant and increasing temperature.
+While we observe similar changes in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in the population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \ref{fig:PopPct}).
 These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow exterior strata 17 and 18.
@@
 The largest improvements were seen in the increasing temperature scenarios.
-As a result, of the VAST estimates that included covariates information, \(47/48\approx98\)\% had a lower relative error than the corresponding stratified mean estimate.
+When comparing to the design-based estimates, \(47/48\approx98\)\% of the VAST estimates that included covariates information had a lower relative error than the corresponding stratified mean estimate.
 This implies that the covariate information helped our model-based estimate account for the design-based issues related to an increasing percentage of the population entering smaller strata, which can become exacerbated in the increasing temperature simulations.
@@
 
-As see in Table \textbf{cite VAST setting table}, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
+As seen in Table \textbf{cite VAST setting table}, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
 Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73\% had a lower absolute error compared to the corresponding estimate without covariates (\(117/160\approx 73\%\)), and \(77/80\approx 96\%\) of VAST models with covariates provided lower absolute error than the corresponding stratified mean estimate.
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-arreguin1996catchability}{}}%
+Arreguı́n-Sánchez, F. 1996. Catchability: A key parameter for fish stock assessment. Reviews in fish biology and fisheries \textbf{6}: 221--242. Springer.
+
 \leavevmode\vadjust pre{\hypertarget{ref-azarovitz1981brief}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Econ}{}}%
+{Fisheries Economics of the United States: Data and Visualizations}. (n.d.). \url{https://www.fisheries.noaa.gov/data-tools/fisheries-economics-united-states-data-and-visualizations\#}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-henderson2017effects}{}}%
@@
 Elsevier.
+
+\leavevmode\vadjust pre{\hypertarget{ref-langan2021climate}{}}%
+Langan, J.A., Puggioni, G., Oviatt, C.A., Henderson, M.E., and Collie, J.S. 2021. Climate alters the migration phenology of coastal marine species. Marine Ecology Progress Series \textbf{660}: 1--18.
 
@@
 \caption{Percent of each species in each strata for during survey weeks in our spatial simulations.
-All constant temperature scenarios follow the patterns on the left while increasing temperature scenarios follow the patterns on the right.}\label{fig:PopPct}
+All constant temperature scenarios follow the patterns on the left while increasing temperature scenarios follow the patterns on the right.
+See Figure 1??
+for a spatial reference of the Georges Bank strata.}\label{fig:PopPct}
 \end{figure}

### 2023-05-23 TestScripts/GB_habitat_create.R

@@ -1575,7 +1575,7 @@ list_all[["Had"]] <- list_all_temp[[good_iter[3]]]
 
 #simulation results (LOAD JUST ONE OF THE FOLLOWING)
 #memory.limit(45000) #this one for all results
-#result <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Results\\",scenario,"\\result_",scenario,".RDS",sep=""))
+#result <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Simulation_Results\\",scenario,"\\result_",scenario,".RDS",sep=""))
 #load existing result_goodones, if it exists
 result <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Simulation_Results\\",scenario,"\\result_goodones_",scenario,".RDS",sep=""))
 
@@ -2262,8 +2262,8 @@ spplot(GB_strata[[s]],zcol="Exclude", col.regions = c("yellow","green"), par.set
       
       scenario1 <- paste(PopScen,"_",TempScen,sep="")
       
-      result1[[PopScen]][[TempScen]] <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Results\\",scenario1,"\\result_goodones_",scenario1,".RDS",sep=""))
-      list_all_temp[[PopScen]][[TempScen]] <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Results\\",scenario1,"\\list_all_",scenario1,".RDS",sep=""))
+      result1[[PopScen]][[TempScen]] <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Simulation_Results\\",scenario1,"\\result_goodones_",scenario1,".RDS",sep=""))
+      list_all_temp[[PopScen]][[TempScen]] <- readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Simulation_Results\\",scenario1,"\\list_all_",scenario1,".RDS",sep=""))
       
       
       
@@ -2419,6 +2419,9 @@ spplot(GB_strata[[s]],zcol="Exclude", col.regions = c("yellow","green"), par.set
         
         labs(x=NULL,y=NULL, title = "Yellowtail Flounder") + #
         
+        #start y axis at 0
+        #expand_limits(y=0) +
+        
         theme(legend.position="bottom")  +
         
         theme(axis.text=element_text(size=12),
[…]

### 2023-05-23 TestScripts/Plots_for_paper_presentation.R

@@ -212,9 +212,10 @@ labs(x="year",y="Biomass", title ="Haddock Fall Estimate", color ="" )+
 
 
 
-
-#COD DECPOP_CONTEMP ESTIMATE RATIO
-
+# ESTIMATE RATIO PLOTS
+#Yellowtail estimates are for the constant population, constant temperature, all strata
+#COD DECPOP_CONTEMP
+#Haddock are IncPop_IncTemp_all strata_
 
 model_line_size = 1.5
 model_point_size = 3.5
@@ -267,7 +268,7 @@ ggplot() +
   #                                          "  FC=", toString(FC_fall),
   #                                          "  FeSM=",round(SRS_Model_error[[s]][[folder]][["fall"]],digits=2),sep=""), color ="" )
   
-  labs(x="year",y="log(Estimate / Model)", title = paste("Cod Ratio Trend- ", ssnn,sep=""), color ="" )+
+  labs(x="year",y="log(Estimate / Model)", title = paste("Haddock Ratio Trend- ", ssnn,sep=""), color ="" )+
   
   theme(axis.text=element_text(size=18),
         axis.title=element_text(size=16),

### 2023-05-25 Manuscript/Draft1.Rmd
small manuscript update

@@
 
+Stock assessment scientists choose from a number of approaches to obtain abundance estimates that are derived from the survey data.
+Methods range from traditional design-based estimates to model-based estimates that vary in complexity.
+Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest.
+These methods do not account for spatial variation in samples and are not able to account for environmental influences on survey values.
+Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors).
+Model-based estimators, such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM), help account for complex relationships between variables and can help overcome problems with sampling design.
+
 <!-- - Climate change is happening -->
@@
 We consider the impact of climate change by comparing simulations that use a repeating water temperature pattern to those where temperature increases on average over time.
-In both cases we analyze the ability of stratified random sampling to track population trends through a design-based approach (stratified mean) compared to a model-based approach that allows for the inclusion of covariates (Vector Autoregressive Spatiotemporal Model).
+In both cases we conduct stratified random sampling on model output and analyze the ability of the samples to track population trends.
+We compare yearly abundance estimates obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
+The stratified mean is a design-based approach that calculates the arithmetic mean catch per tow and has traditionally been used with stratified random sample designs.
+VAST is a spatial delta-generalized linear mixed model that estimates both abundance (biomass) and probability of occurrence (presence/absence) [@thorson2019guidance].
+If desired, VAST also allows users to include covariate data to better inform the model.
+Covariates can be static (eg.
+habitat preferences) or dynamics (eg.
+temperature).
+We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses.
 
@@
 
-<!-- --Stratified mean vs VAST with and without covariates -->
+The design-based method we used is the stratified mean, which divides the inhabited domain into $N$ disjoint strata based on relevant geographic and environmental information such as depth and latitude/longitude.
+The number of samples taken from each stratum $S_j$ for $j=1...N$ is relative to the given area.
+The stratified mean biomass $SM_{s,y}$ for a given season $s$ and year $y$ can then be calculated by through the weighted average
 
-Stock assessment scientists choose from a number of approaches to obtain abundance estimates that are derived from the survey data.
-Methods range from traditional design-based estimates to model-based estimates that vary in complexity.
-Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest.
-These methods do not account for spatial variation in samples and are not able to account for environmental influences on survey values.
-Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors).
-Model-based estimators, such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM), help account for complex relationships between variables and can help overcome problems with sampling design.
+\[
+  \makebox[\linewidth]{$SM_{s,y} = \sum_{i=1}^N W_i \frac{\sum_{k=1}^{S_i}y_{k,i}}{S_i}$},
+\]
 
-We compare yearly abundance estimates obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
-The stratified mean is a design-based approach that calculates the arithmetic mean catch per tow and has traditionally been used with stratified random sample designs.
-VAST is a spatial delta-generalized linear mixed model that estimates both abundance (biomass) and probability of occurrence (presence/absence) [@thorson2019guidance].
-If desired, VAST also allows users to include covariate data to better inform the model.
-Covariates can be static (eg.
-habitat preferences) or dynamics (eg.
-temperature).
-We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses.
-The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and can take on the order of hours to complete.
+where $W_i=\frac{S_i}{\sum_{j=1}^N S_j}$ and $y_{k,i}$ represents the biomass in sample $k$ of stratum $i$ in season $s$ and year $y$.
+These tow-dependent calculations are quick and easy to calculate, especially relative to the model-based VAST estimates.
+
+We compare stratified mean estimates to those derived from the model-based VAST approach.
+VAST models both biomass $p_1(i)$ and presence/absence $p_2(i)$ for each observation $i$ as linear predictors using a spatial delta-generalized linear mixed model.
+The equation representing the biomass predictor can be represented by
+
+\[
+  \makebox[\linewidth]{$p_1(i)=\text{temporal variation} + \text{spatial variation} + \text{spatio-temporal variation} + \text{vessel effects} + \text{habitat covariates} + \text{catchability covariates}$},
+\]
+
+where more specific functional forms of each component are further described in [@thorson2019guidance].
+$p_2(i)$ takes on a similar form as $p_1(i)$.
+VAST models require numerous user inputs to determine how the linear predictors will be conditioned and solved.
+As a result, VAST models take on the order of hours to complete.
 
@@
 Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings.
-Through this process, and in consultation with the VAST package creator, we ultimately compared the performance of two sets of settings in our VAST models, which can be seen in Table\@ref(tab:VASTsettings).
+Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models, which can be seen in Table\@ref(tab:VASTsettings).
+VAST Settings B were chosen after reviewing the information provided in the "Seasonal Model" section of VAST's online tutorial https://james-thorson-noaa.github.io/docs/tutorials/seasonal-model/.
+After running models for each of our scenarios using these settings we shared some of our results with the VAST package creator, James (Jim) Thorson.
+Jim reviewed results and suggested we also consider the options in Settings A.
+Both settings use a Tweedie model and turn off the first linear predictor $p_1$ so that only biomass is being modeled.
+The difference between the two settings is in the Rho Configuration setting where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A.
+Intercept for the biomass predictor is a fixed effect (beta2=0) vs constant among years as a fixed effect (beta2=3)
 
@@
 While we observe similar changes in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in the population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \@ref(fig:PopPct)).
-These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow exterior strata 17 and 18.
-One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how ~10\% of the population exists in the narrow exterior strata 17 in the fall, while seemingly none of the population exists in any of the exterior strata (14, 15, 17, 18) in the spring.
+These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow outer strata 17 and 18.
+One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how ~10\% of the population exists in the narrow outer strata 17 in the fall, while seemingly none of the population exists in any of the strata near the edge of the domain (14, 15, 17, 18) in the spring.
 
@@
 In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
+The settings used in VAST mattered with Settings B outperforming Settings A in $67/96\approx 70\%$ of scenarios.
 When all strata are sampled, adding covariates greatly improved estimates in all scenarios.
@@
 These 15 fall estimates span all other scenario variations.
-The implication of this is that our model-based approach without covariates struggles with the primary seasonal difference for Yellowtail Flounder, which is that a larger percentage of the population exists in the narrow exterior strata (18 and/or 17).
+The implication of this is that our model-based approach without covariates struggles with the primary seasonal difference for Yellowtail Flounder, which is that a larger percentage of the population exists in the narrow strata toward the edge of the domain (18 and/or 17).
 This theory is further supported by the fact that the absolute error in the increasing temperature scenarios increased dramatically in the fall season, when the combined percentage of the population in the outer strata 17 and 18 increased to over 40\% by the end of the simulation.
@@
 Table \@ref(tab:CodERROR) contains the absolute error between our abundance estimates for Cod and the true model values.
+The VAST settings performed comparably to each other with $18/32\approx57%$ of estimates performing better with VAST Settings A.
 Of the abundance estimates without covariates, $12/16=75$\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
@@
 While similar results can be seen in the spring for the increasing temperature scenario, much more dramatic results exist in the fall under increasing temperature as we see a significant decreases in strata 13, 16, 21, and 22 that leads to the most noticeable increases in strata 17, 18, and 29.
-This shift represents movement from the shallower and more centrally located strata towards exterior deeper strata.
+This shift represents movement from the shallower and more centrally located strata towards deeper strata located near the edge of the domain.
 Since stratum 16 contains very favorable habitat including much of the species' spawning ground, the strong shift out of 16 and into the northern most strata of 29 in the fall demonstrates a climate-driven change in movement preference.
@@
 Table \@ref(tab:HadERROR) contains the absolute error between our abundance estimates for Haddock and the true model values.
+Absolute error values tended to be lower with VAST Settings B about $21/32\approx 66\%$ of scenarios.
 We notice that the model-based VAST produced particularly large errors in spring compared to the stratified mean, with added covariates only improving to the level of the stratified mean.
@@
 That is, similar to Cod, when sampling a reduced domain adding covariates actually decreases VAST's accuracy.
-Since this occurs in all scenarios, it seems to again be related to failing to accurately monitor smaller exterior stratum 17.
+Since this occurs in all scenarios, it seems to again be related to failing to accurately monitor stratum 17 near the edge of the domain.
 
@@
 
-<font size="5">*Differences Between VAST Settings (mostly notes, not fully formed)*</font>
+<!-- <font size="5">*Differences Between VAST Settings (mostly notes, not fully formed)*</font> -->
 
-For YTF, 29/96 ~ 30% of VAST runs with new settings were better (70% were worse).
+<!-- For YTF, 29/96 ~ 30% of VAST runs with new settings were better (70% were worse).
 80/96 ~ 83% of scenarios had a VAST run with a better error than the stratified mean.
@@
 In the spring the population is shifting mostly into strata 17 (thin strata adjacent to 16), but in the fall they are shifting into both 17 and 18 (18 thin one adjacent to 17).
-Thus it seems like with much of the population concentrated in the really small outer most strata, VAST produces an underestimate, even with covariates
+Thus it seems like with much of the population concentrated in the really small outer most strata, VAST produces an underestimate, even with covariates   -->
 
-For Haddock (IncPop), 18/32 ~ 57% of VAST runs better with new settings.
+<!-- For Haddock (IncPop), 18/32 ~ 57% of VAST runs better with new settings.
 24/32 = 75% of scenarios had a VAST run with a better error than the stratified mean.
@@
 Stratified mean tended to perform better when all strata were included in the calculation, while VAST tended to perform better when strata are removed.
+-->
 
-For Cod (DecPop), only 11/32 ~ 34% of VAST runs were better with new settings.
+<!-- For Cod (DecPop), only 11/32 ~ 34% of VAST runs were better with new settings.
 28/32~ 88% of scenarios had a VAST run with a better estimate than the stratified mean.
@@
 Might be having trouble tracking in to smaller strata again.
+-->
 
@@
 A second overestimate the following year could then have a detrimental impact by reducing the population even further.
+Conversely, an overly smoothed estimator could miss true signals of change in the population and delay needed management response to either sudden increases or decreases in the population.
 Our population model has assumed a constant total mortality that accounts for both fishing and natural death, and therefore will not account for impacts of such management decisions.
 This type of question can be best explored with a management strategy evaluation.
-
-VAST may provide more consistent biomass predictions(?)
 
@@
 [[we should thank Jim (obviously), but also those who helped you with the habitat data (robyn, david Chev.) and others?
-also should note that funding provided by Northeast Fisheries and Climate program at the NEFSC source (I'll dig up the official name of the funding source)]]
+also should note that funding provided by Northeast Fisheries and Climate program at the Northeast Fisheries Science Center ]]  Also thank Joe Langan for helping develop the proposal that funded this work.
 

### 2023-05-25 Manuscript/Draft1.tex
small manuscript update

@@
 
+Stock assessment scientists choose from a number of approaches to obtain abundance estimates that are derived from the survey data.
+Methods range from traditional design-based estimates to model-based estimates that vary in complexity.
+Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest.
+These methods do not account for spatial variation in samples and are not able to account for environmental influences on survey values.
+Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors).
+Model-based estimators, such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM), help account for complex relationships between variables and can help overcome problems with sampling design.
+
 Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades.
@@
 We consider the impact of climate change by comparing simulations that use a repeating water temperature pattern to those where temperature increases on average over time.
-In both cases we analyze the ability of stratified random sampling to track population trends through a design-based approach (stratified mean) compared to a model-based approach that allows for the inclusion of covariates (Vector Autoregressive Spatiotemporal Model).
+In both cases we conduct stratified random sampling on model output and analyze the ability of the samples to track population trends.
+We compare yearly abundance estimates obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
+The stratified mean is a design-based approach that calculates the arithmetic mean catch per tow and has traditionally been used with stratified random sample designs.
+VAST is a spatial delta-generalized linear mixed model that estimates both abundance (biomass) and probability of occurrence (presence/absence) (Thorson 2019).
+If desired, VAST also allows users to include covariate data to better inform the model.
+Covariates can be static (eg.
+habitat preferences) or dynamics (eg.
+temperature).
+We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses.
 
@@
 
-Stock assessment scientists choose from a number of approaches to obtain abundance estimates that are derived from the survey data.
-Methods range from traditional design-based estimates to model-based estimates that vary in complexity.
-Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest.
-These methods do not account for spatial variation in samples and are not able to account for environmental influences on survey values.
-Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors).
-Model-based estimators, such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM), help account for complex relationships between variables and can help overcome problems with sampling design.
-
-We compare yearly abundance estimates obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
-The stratified mean is a design-based approach that calculates the arithmetic mean catch per tow and has traditionally been used with stratified random sample designs.
-VAST is a spatial delta-generalized linear mixed model that estimates both abundance (biomass) and probability of occurrence (presence/absence) (Thorson 2019).
-If desired, VAST also allows users to include covariate data to better inform the model.
-Covariates can be static (eg.
-habitat preferences) or dynamics (eg.
-temperature).
-We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses.
-The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and can take on the order of hours to complete.
+The design-based method we used is the stratified mean, which divides the inhabited domain into \(N\) disjoint strata based on relevant geographic and environmental information such as depth and latitude/longitude.
+The number of samples taken from each stratum \(S_j\) for \(j=1...N\) is relative to the given area.
+The stratified mean biomass \(SM_{s,y}\) for a given season \(s\) and year \(y\) can then be calculated by through the weighted average
+
+\[
+  \makebox[\linewidth]{$SM_{s,y} = \sum_{i=1}^N W_i \frac{\sum_{k=1}^{S_i}y_{k,i}}{S_i}$},
+\]
+
+where \(W_i=\frac{S_i}{\sum_{j=1}^N S_j}\) and \(y_{k,i}\) represents the biomass in sample \(k\) of stratum \(i\) in season \(s\) and year \(y\).
+These tow-dependent calculations are quick and easy to calculate, especially relative to the model-based VAST estimates.
+
+We compare stratified mean estimates to those derived from the model-based VAST approach.
+VAST models both biomass \(p_1(i)\) and presence/absence \(p_2(i)\) for each observation \(i\) as linear predictors using a spatial delta-generalized linear mixed model.
+The equation representing the biomass predictor can be represented by
+
+\[
+  \makebox[\linewidth]{$p_1(i)=\text{temporal variation} + \text{spatial variation} + \text{spatio-temporal variation} + \text{vessel effects} + \text{habitat covariates} + \text{catchability covariates}$},
+\]
+
+where more specific functional forms of each component are further described in (Thorson 2019).
+\(p_2(i)\) takes on a similar form as \(p_1(i)\).
+VAST models require numerous user inputs to determine how the linear predictors will be conditioned and solved.
+As a result, VAST models take on the order of hours to complete.
 
@@
 Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings.
-Through this process, and in consultation with the VAST package creator, we ultimately compared the performance of two sets of settings in our VAST models, which can be seen in Table\ref{tab:VASTsettings}.
+Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models, which can be seen in Table\ref{tab:VASTsettings}.
+VAST Settings B were chosen after reviewing the information provided in the ``Seasonal Model'' section of VAST's online tutorial \url{https://james-thorson-noaa.github.io/docs/tutorials/seasonal-model/}.
+After running models for each of our scenarios using these settings we shared some of our results with the VAST package creator, James (Jim) Thorson.
+Jim reviewed results and suggested we also consider the options in Settings A.
+Both settings use a Tweedie model and turn off the first linear predictor \(p_1\) so that only biomass is being modeled.
+The difference between the two settings is in the Rho Configuration setting where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A.
+Intercept for the biomass predictor is a fixed effect (beta2=0) vs constant among years as a fixed effect (beta2=3)
 
@@
 While we observe similar changes in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in the population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \ref{fig:PopPct}).
-These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow exterior strata 17 and 18.
-One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how \textasciitilde10\% of the population exists in the narrow exterior strata 17 in the fall, while seemingly none of the population exists in any of the exterior strata (14, 15, 17, 18) in the spring.
+These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow outer strata 17 and 18.
+One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how \textasciitilde10\% of the population exists in the narrow outer strata 17 in the fall, while seemingly none of the population exists in any of the strata near the edge of the domain (14, 15, 17, 18) in the spring.
 
@@
 In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
+The settings used in VAST mattered with Settings B outperforming Settings A in \(67/96\approx 70\%\) of scenarios.
 When all strata are sampled, adding covariates greatly improved estimates in all scenarios.
@@
 
-Table \ref{tab:CodERROR} contains the absolute error between our abundance estimates for Cod and the true model values.
+Table \ref{tab:CodERROR} contains the absolute error between our abundance estimates for Cod and the true model values. The VAST settings performed comparably to each other with \(18/32\approx57%
+\) of estimates performing better with VAST Settings A.
 Of the abundance estimates without covariates, \(12/16=75\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
@@
 Table \ref{tab:HadERROR} contains the absolute error between our abundance estimates for Haddock and the true model values.
+Absolute error values tended to be lower with VAST Settings B about \(21/32\approx 66\%\) of scenarios.
 We notice that the model-based VAST produced particularly large errors in spring compared to the stratified mean, with added covariates only improving to the level of the stratified mean.
@@
 The average change in the error of abundance estimates between constant temperature scenarios and increasing temperature scenarios were 1.42 for VAST estimates without covariates, 1.80 for VAST estimates that included covariates, and 1.37 for stratified mean estimates.
-
-\emph{Differences Between VAST Settings (mostly notes, not fully formed)}
-
-For YTF, 29/96 \textasciitilde{} 30\% of VAST runs with new settings were better (70\% were worse).
-80/96 \textasciitilde{} 83\% of scenarios had a VAST run with a better error than the stratified mean.
-15/16 of the times the stratified mean was better VAST was not using covariates.
-The 1 time that VAST was using covariates and was still worse than the stratified mean was IncPop\_IncTemp\_Allstrata\_WCov\_WNoise in the Fall (VAST had strong overestimate in the fall during IncPOP\_IncTemP\_allstrata for Had and YT).
-The overestimate in the fall only seems to be related to the way the population shifts between season.
-Looking at the percent shift plots, the population is shifting out of strata 16 in both seasons (large east strata).
-In the spring the population is shifting mostly into strata 17 (thin strata adjacent to 16), but in the fall they are shifting into both 17 and 18 (18 thin one adjacent to 17).
-Thus it seems like with much of the population concentrated in the really small outer most strata, VAST produces an underestimate, even with covariates
-
-For Haddock (IncPop), 18/32 \textasciitilde{} 57\% of VAST runs better with new settings.
-24/32 = 75\% of scenarios had a VAST run with a better error than the stratified mean.
-14/16 VAST were better with covariates (the two that were worse were essentially the same) while only 10/16 were better without covariates.
-Stratified mean tended to perform better when all strata were included in the calculation, while VAST tended to perform better when strata are removed.
-
-For Cod (DecPop), only 11/32 \textasciitilde{} 34\% of VAST runs were better with new settings.
-28/32\textasciitilde{} 88\% of scenarios had a VAST run with a better estimate than the stratified mean.
-All 16/16 = 100\% of VAST runs with a covariate were better than the stratified mean.
-12/16 VAST were better without covariates.
-The 4 runs without covariates that were worse were all extremely large errors and all from increasing temp in the fall (with and without all strata and noise).
-In the fall the population is moving out of strata 16 (large eastern strata) and 21 (north large) and into a number of strata (22 (tiny north), 24 (large north), 18 (tiny north)).
-Might be having trouble tracking in to smaller strata again.
 
@@
 A second overestimate the following year could then have a detrimental impact by reducing the population even further.
+Conversely, an overly smoothed estimator could miss true signals of change in the population and delay needed management response to either sudden increases or decreases in the population.
 Our population model has assumed a constant total mortality that accounts for both fishing and natural death, and therefore will not account for impacts of such management decisions.
@@
 
-VAST may provide more consistent biomass predictions(?)
-
 \section{Acknowledgements}
@@
 {[}{[}we should thank Jim (obviously), but also those who helped you with the habitat data (robyn, david Chev.) and others?
-also should note that funding provided by Northeast Fisheries and Climate program at the NEFSC source (I'll dig up the official name of the funding source){]}{]}
+also should note that funding provided by Northeast Fisheries and Climate program at the Northeast Fisheries Science Center {]}{]} Also thank Joe Langan for helping develop the proposal that funded this work.
 

### 2023-05-30 Manuscript/Draft1.Rmd
manuscript update

@@
 
-We follow the advice given in [@thorson2019guidance] to build VAST models to estimate biomass in our Georges Bank population models using stratified random samples from our model output.
+We follow the advice given in [@thorson2019guidance] to build VAST models that estimate biomass in our Georges Bank population models using stratified random samples from our model output.
 In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
-We carried out the same model-building process running models without covariate information as well as including covariates in our model.
-We considered covariates in the form of dynamic temperature values and/or static habitat values ($Hab_{J.s}$) from our population models.
+We carried out the same model-building process using covariate information to inform our models as well as without covariates in our models.
+The covariates we considered were the dynamic temperature values and static habitat values ($Hab_{J.s}$) from our population models.
 When using covariates we ultimately decided to consider a best-case scenario by including both temperature and habitat covariates for both linear predictors in order to provide the most information to the model.
@@
 Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models, which can be seen in Table\@ref(tab:VASTsettings).
-VAST Settings B were chosen after reviewing the information provided in the "Seasonal Model" section of VAST's online tutorial https://james-thorson-noaa.github.io/docs/tutorials/seasonal-model/.
-After running models for each of our scenarios using these settings we shared some of our results with the VAST package creator, James (Jim) Thorson.
-Jim reviewed results and suggested we also consider the options in Settings A.
+VAST Settings B were chosen after reviewing the information provided in the "Seasonal Model" section of VAST's online Github tutorial [@ThorsonSeasonal].
+After running models for each of our scenarios using these settings, we shared some of our results with the VAST package creator, James (Jim) Thorson.
+Jim reviewed our VAST model results and suggested we also consider the options shown in Settings A.
 Both settings use a Tweedie model and turn off the first linear predictor $p_1$ so that only biomass is being modeled.
-The difference between the two settings is in the Rho Configuration setting where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A.
-Intercept for the biomass predictor is a fixed effect (beta2=0) vs constant among years as a fixed effect (beta2=3)
+The difference between the two settings can be seen in the Rho Configuration, where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A.
+The difference amounts to intercepts for the biomass predictor being a fixed effect (RhoBeta2=0) versus intercepts being constant among years as a fixed effect (RhoBeta2=3).
+According to Jim, the Settings A values are representative of what stock assessment scientists typically use in their analyses.
+As a result we will focus our analysis on VAST results obtained using Settings A.
 
@@
 We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from.
-Figure \ref(fig:strata-plot) shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \ref(fig:PopPct).
+Figure \@ref(fig:strata-plot) shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \@ref(fig:PopPct).
 The yellow regions in Figure \@ref(fig:strata-plot) depict the strata that were removed from certain stratified random surveys for each species.
@@
 
+To test the impact of shifting populations on abundance estimates, we simulated known population data that included environmental forcing and compared biomass estimates derived from stratified random sampling to the true values.
+We carried out simulations for 3 species on Georges Bank that took into account different population trends, temperature scenarios, noise in survey samples, and seasonal variation.
+We compared a design-based estimation approach (stratified mean) with a model-based approach that allows for the inclusion of environmental covariates (VAST) when the entire domain was sampled as well as when a reduced number of strata were included.
+Our analysis included a comparison of absolute error values between 20-year estimates and consideration of the variation seen year-to-year displayed by each approach as well as the consistency between each 20-year estimate.
+
+Our analysis showed that both approaches are capable of providing abundance estimates that track the true abundance trend.
+While VAST estimates tended to provide lower relative error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B combined to produce over 30 abundance estimates with larger error values than the largest stratified mean error.
+This fact highlights how sensitive VAST is to the choices made regarding settings and data used in analyses.
+This problem is compounded by the lack of clear guidance regarding how to build and analyze models with the correct inputs.
+Even more concerning, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using the wrong settings **cite Chris C paper**.
+The value of model-based approaches such as VAST are clear- they allow users to tailor the model to overcome issues related to the input data by and to account for factors such as environmental change.
+However, the many consequential decisions that must be made in a given model and the lack of clear criteria for making these decisions makes it challenging to make these choices and even more difficult to determine whether the choices are correct.
+
+Spatial factors contributed to many of the scenarios that produced larger error values seen in our estimates.
+Flow from the shallow centrally located strata into the more deeper, smaller, and less-sampled eastern strata contributed to VAST failing to accurately modeling the abundance of yellowtail flounder without the addition of covariate information, the impact of which was greatly exacerbated in the increasing temperature scenarios.
+While adding covariates can help inform the model and improve the estimate as seen with the yellowtail flounder simulation results, including covariates can hinder the model in certain instances.
+For example, including covariates for haddock when the northern most strata were removed from the survey meant that the population was not being sampled in the coldest part of its domain which ultimately increased the error.
+
+something about seasonal impact seen in all estimates with fall estimates typically noticeably worse.
+We explored this through the ratio plots.
+This was mostly a problem with VAST as SM showed more consistent estimates between seasons.
+
 <font size="5">*Impact of Covariates*</font>
@@
 
- Of the ~27% of VAST models where covariates did not improve the estimate, most provided a comparable error (for ex, 0.11 vs 0.13).
+ Of the ~27% of VAST models where covariates did not improve the VAST estimate, most provided a comparable error (for ex, 0.11 vs 0.13).
 The instances where including covariates provided a significant different in error took place scenarios that included either an increasing temperature and/or reduced sampling domain.

### 2023-05-30 Manuscript/Draft1.tex
manuscript update

@@
 
-We follow the advice given in (Thorson 2019) to build VAST models to estimate biomass in our Georges Bank population models using stratified random samples from our model output.
+We follow the advice given in (Thorson 2019) to build VAST models that estimate biomass in our Georges Bank population models using stratified random samples from our model output.
 In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
-We carried out the same model-building process running models without covariate information as well as including covariates in our model.
-We considered covariates in the form of dynamic temperature values and/or static habitat values (\(Hab_{J.s}\)) from our population models.
+We carried out the same model-building process using covariate information to inform our models as well as without covariates in our models.
+The covariates we considered were the dynamic temperature values and static habitat values (\(Hab_{J.s}\)) from our population models.
 When using covariates we ultimately decided to consider a best-case scenario by including both temperature and habitat covariates for both linear predictors in order to provide the most information to the model.
@@
 Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models, which can be seen in Table\ref{tab:VASTsettings}.
-VAST Settings B were chosen after reviewing the information provided in the ``Seasonal Model'' section of VAST's online tutorial \url{https://james-thorson-noaa.github.io/docs/tutorials/seasonal-model/}.
-After running models for each of our scenarios using these settings we shared some of our results with the VAST package creator, James (Jim) Thorson.
-Jim reviewed results and suggested we also consider the options in Settings A.
+VAST Settings B were chosen after reviewing the information provided in the ``Seasonal Model'' section of VAST's online Github tutorial (Thorson, James T n.d.).
+After running models for each of our scenarios using these settings, we shared some of our results with the VAST package creator, James (Jim) Thorson.
+Jim reviewed our VAST model results and suggested we also consider the options shown in Settings A.
 Both settings use a Tweedie model and turn off the first linear predictor \(p_1\) so that only biomass is being modeled.
-The difference between the two settings is in the Rho Configuration setting where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A.
-Intercept for the biomass predictor is a fixed effect (beta2=0) vs constant among years as a fixed effect (beta2=3)
+The difference between the two settings can be seen in the Rho Configuration, where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A.
+The difference amounts to intercepts for the biomass predictor being a fixed effect (RhoBeta2=0) versus intercepts being constant among years as a fixed effect (RhoBeta2=3).
+According to Jim, the Settings A values are representative of what stock assessment scientists typically use in their analyses.
+As a result we will focus our analysis on VAST results obtained using Settings A.
 
@@
 We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from.
-Figure \ref(fig:strata-plot) shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \ref(fig:PopPct).
+Figure \ref{fig:strata-plot} shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \ref{fig:PopPct}.
 The yellow regions in Figure \ref{fig:strata-plot} depict the strata that were removed from certain stratified random surveys for each species.
@@
 
+To test the impact of shifting populations on abundance estimates, we simulated known population data that included environmental forcing and compared biomass estimates derived from stratified random sampling to the true values.
+We carried out simulations for 3 species on Georges Bank that took into account different population trends, temperature scenarios, noise in survey samples, and seasonal variation.
+We compared a design-based estimation approach (stratified mean) with a model-based approach that allows for the inclusion of environmental covariates (VAST) when the entire domain was sampled as well as when a reduced number of strata were included.
+Our analysis included a comparison of absolute error values between 20-year estimates and consideration of the variation seen year-to-year displayed by each approach as well as the consistency between each 20-year estimate.
+
+Our analysis showed that both approaches are capable of providing abundance estimates that track the true abundance trend.
+While VAST estimates tended to provide lower relative error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B combined to produce over 30 abundance estimates with larger error values than the largest stratified mean error.
+This fact highlights how sensitive VAST is to the choices made regarding settings and data used in analyses.
+This problem is compounded by the lack of clear guidance regarding how to build and analyze models with the correct inputs.
+Even more concerning, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using the wrong settings \textbf{cite Chris C paper}.
+The value of model-based approaches such as VAST are clear- they allow users to tailor the model to overcome issues related to the input data by and to account for factors such as environmental change.
+However, the many consequential decisions that must be made in a given model and the lack of clear criteria for making these decisions makes it challenging to make these choices and even more difficult to determine whether the choices are correct.
+
+Spatial factors contributed to many of the scenarios that produced larger error values seen in our estimates.
+Flow from the shallow centrally located strata into the more deeper, smaller, and less-sampled eastern strata contributed to VAST failing to accurately modeling the abundance of yellowtail flounder without the addition of covariate information, the impact of which was greatly exacerbated in the increasing temperature scenarios.
+While adding covariates can help inform the model and improve the estimate as seen with the yellowtail flounder simulation results, including covariates can hinder the model in certain instances.
+For example, including covariates for haddock when the northern most strata were removed from the survey meant that the population was not being sampled in the coldest part of its domain which ultimately increased the error.
+
+something about seasonal impact seen in all estimates with fall estimates typically noticeably worse.
+We explored this through the ratio plots.
+This was mostly a problem with VAST as SM showed more consistent estimates between seasons.
+
 \emph{Impact of Covariates}
@@
 
-Of the \textasciitilde27\% of VAST models where covariates did not improve the estimate, most provided a comparable error (for ex, 0.11 vs 0.13).
+Of the \textasciitilde27\% of VAST models where covariates did not improve the VAST estimate, most provided a comparable error (for ex, 0.11 vs 0.13).
 The instances where including covariates provided a significant different in error took place scenarios that included either an increasing temperature and/or reduced sampling domain.
@@
 Politis, P.J., Galbraith, J.K., Kostovick, P., and Brown, R.W. 2014. Northeast fisheries science center bottom trawl survey protocols for the NOAA ship henry b. bigelow.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ThorsonSeasonal}{}}%
+Thorson, James T. (n.d.). {Seasonal and Interannual Dynamics Github Documentation}. {https://james-thorson-noaa.github.io/docs/tutorials/seasonal-model/}.
 

### 2023-05-30 Manuscript/Draft1.Rmd
manuscript update

@@
 Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics.
-A full list of parameters used in our model can be seen below in Tables \@ref(paramsALL) and \@ref(tab:paramsSCENARIOS).
+A full list of parameters used in our model can be seen below in Tables 1 and \@ref(tab:paramsSCENARIOS).
 
@@
 We compare stratified mean estimates to those derived from the model-based VAST approach.
-VAST models both biomass $p_1(i)$ and presence/absence $p_2(i)$ for each observation $i$ as linear predictors using a spatial delta-generalized linear mixed model.
-The equation representing the biomass predictor can be represented by
+VAST models both biomass $p_1(i)$ and presence/absence $p_2(i)$ for each observation $i$ as linear predictors using a spatial delta-generalized linear mixed model that can be represented by
 
 \[
-  \makebox[\linewidth]{$p_1(i)=\text{temporal variation} + \text{spatial variation} + \text{spatio-temporal variation} + \text{vessel effects} + \text{habitat covariates} + \text{catchability covariates}$},
+  \makebox[\linewidth]{$p_1(i)=\text{temporal variation} + \text{spatial variation} + \text{spatio-temporal variation}  + \text{habitat covariates} + \text{catchability covariates}$},
+\]
+
+\[
+  \makebox[\linewidth]{$p_*(i)= \beta_*(t_i) + L_{\omega_*}\omega_*(s_i) + L_* \epsilon_*(c_i,t_i) + \upsilon(s_i,t_i) + \xi(i)$},
 \]
@@
 Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings.
-Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models, which can be seen in Table\@ref(tab:VASTsettings).
+Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models, which can be seen in Table 3.
 VAST Settings B were chosen after reviewing the information provided in the "Seasonal Model" section of VAST's online Github tutorial [@ThorsonSeasonal].
@@
 
-```{r scenarios,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
-
-options(knitr.kable.NA = '')
-
-combos_table = read_csv("CSVs/scenario_combos.csv")
-
-summary_tab <- combos_table %>% 
-   mutate_if(grepl('5deg',.), ~replace(., grepl('5deg', .), paste("Increasing 5","\U00B0","C",sep="")))
-
-# Create table to present the data]
- kable(summary_tab, "simple",
-                     caption = "Each index estimate chooses one condition from each of the following 7 columns.",
-          format = "latex", booktabs = TRUE) %>%
-          kable_styling(latex_options = "scale_down")
-```
-
 \section{Results}
@@
 
-Figure \@ref(fig:PopPct) depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey in the spring (weeks 13 and 14) and fall (weeks 37 and 28).
+Figure \@ref(fig:PopPct) depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey in the spring (weeks 13 and 14) and fall (weeks 37 and 38).
 The left column in Figure \@ref(fig:PopPct) depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern.
@@
 We compared a design-based estimation approach (stratified mean) with a model-based approach that allows for the inclusion of environmental covariates (VAST) when the entire domain was sampled as well as when a reduced number of strata were included.
-Our analysis included a comparison of absolute error values between 20-year estimates and consideration of the variation seen year-to-year displayed by each approach as well as the consistency between each 20-year estimate.
+Our analysis included a comparison of absolute error values between 20-year estimates and consideration of the year-to-year variation displayed by each approach as well as the variation between each 20-year estimate.
 
-Our analysis showed that both approaches are capable of providing abundance estimates that track the true abundance trend.
+Our analysis showed that the design-based and model-based approaches are each capable of providing biomass estimates that track the true abundance trend.
 While VAST estimates tended to provide lower relative error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B combined to produce over 30 abundance estimates with larger error values than the largest stratified mean error.
@@
 This problem is compounded by the lack of clear guidance regarding how to build and analyze models with the correct inputs.
-Even more concerning, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using the wrong settings **cite Chris C paper**.
+Even more concerning, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using settings that decrease the accuracy of the model **cite Chris C paper**.
 The value of model-based approaches such as VAST are clear- they allow users to tailor the model to overcome issues related to the input data by and to account for factors such as environmental change.
@@
 While adding covariates can help inform the model and improve the estimate as seen with the yellowtail flounder simulation results, including covariates can hinder the model in certain instances.
-For example, including covariates for haddock when the northern most strata were removed from the survey meant that the population was not being sampled in the coldest part of its domain which ultimately increased the error.
+For example, including covariates for haddock when the northern most strata were removed from the survey meant that the population was not being sampled in the coldest part of its domain which led to an incorrect covariate response that ultimately increased the error between the VAST estimate and the true model abundance.
 
+Seasonal differences in spatial abundance patterns in our population models led to seasonal differences in abundance estimates, with most estimates producing a higher error in the Fall 
+spawn weeks 9-14
 something about seasonal impact seen in all estimates with fall estimates typically noticeably worse.
@@
 
-<font size="5">*Impact of Covariates*</font>
+<!-- <font size="5">*Impact of Covariates*</font> -->
 
@@
 
-```{r PopPct, echo=FALSE,fig.show='hold', fig.align='center', fig.cap="Percent of each species in each strata for during survey weeks in our spatial simulations.
-All constant temperature scenarios follow the patterns on the left while increasing temperature scenarios follow the patterns on the right.
-See Figure 1??
-for a spatial reference of the Georges Bank strata.",out.width="47.5%"}
-knitr::include_graphics(c("Images/PercentPlots_ConPopConTemp_YT.png","Images/PercentPlots_ConPopIncTemp_YT.png"))
-knitr::include_graphics(c("Images/PercentPlots_DecPopConTemp_Cod.png","Images/PercentPlots_DecPopIncTemp_Cod.png"))
-knitr::include_graphics(c("Images/PercentPlots_IncPopConTemp_Had.png","Images/PercentPlots_IncPopIncTemp_Had.png"))
-```
-
-```{r CovResponse, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Temperature covariate response plots and resulting population estimate for Yellowtail Flounder on the left and Haddock on the right.
-In both cases the spatial simulations that were surveyed had an increasing population over time, increasing average temperature, and certain strata excluded from sampling, as shown in Figure 1**check**.",out.width="90%"}
-# knitr::include_graphics(c("Images/covariate_good.png","Images/covariate_bad.png"))
-# knitr::include_graphics(c("Images/covariate_est.png"))
-
-knitr::include_graphics(c("Images/covariate_plot.png"))
-
-```
-
-```{r RatioTrend, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Representative example of typical ratio trend for each species, as shown on a log scale.",out.width="90%"}
-# knitr::include_graphics(c("Images/covariate_good.png","Images/covariate_bad.png"))
-# knitr::include_graphics(c("Images/covariate_est.png"))
-#Yellowtail estimates are for the constant population, constant temperature, all strata used in estimates.
-#Cod are Decpop_ConTemp_allStrata_
-#Haddock are IncPop_IncTemp_all strata_
-
-knitr::include_graphics(c("Images/ratio_trend_YTF.PNG"))
-knitr::include_graphics(c("Images/ratio_trend_cod.PNG"))
-knitr::include_graphics(c("Images/ratio_trend_had.PNG"))
-```
-
 <div align="center">Table 1. Parameters used in all population models. SAW 1: [@nefsc1218], SAW 2: [@nefsc1311], SAW 3: not out yet??? </div>
@@
 
-```{r paramsALL2,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
+<!-- ```{r paramsALL2,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'} -->
 
-param_table1 = read_csv("CSVs/Parameters1.csv")
+<!-- param_table1 = read_csv("CSVs/Parameters1.csv") -->
 
-summary_tab1 <- param_table1 %>% 
- # dplyr::mutate_if(is.numeric, list(~as.character(signif(., 3))))%>%
+<!-- summary_tab1 <- param_table1 %>%  -->
+<!--  # dplyr::mutate_if(is.numeric, list(~as.character(signif(., 3))))%>% -->
 
-  mutate_if(grepl('rrr',.), ~replace(., grepl('rrr', .), paste("\U03C1")))#should be replacing r with the rho symbol
- # mutate_if(grepl('W1',.), ~replace(., grepl('W1', .), "WW")) #should be replacing beta with the symbol
- #    mutate_if(grepl('wk',.), ~replace(., grepl('wk', .), "1/wk")) %>% #should be replacing beta with the symbol
- #
+<!--   mutate_if(grepl('rrr',.), ~replace(., grepl('rrr', .), paste("\U03C1")))#should be replacing r with the rho symbol -->
+<!--  # mutate_if(grepl('W1',.), ~replace(., grepl('W1', .), "WW")) #should be replacing beta with the symbol -->
+<!--  #    mutate_if(grepl('wk',.), ~replace(., grepl('wk', .), "1/wk")) %>% #should be replacing beta with the symbol -->
+<!--  #   -->
 
-# Change the orientation of the table so that Species become the variable headings
+<!-- # Change the orientation of the table so that Species become the variable headings -->
 
-# Create table to present the data]
-Fig_1 <- kable(summary_tab1, 
-  caption = "Parameters used in all population models.",
-    format = "latex", booktabs = TRUE) %>%
-  kable_styling(font_size = 10)
+<!-- # Create table to present the data] -->
+<!-- Fig_1 <- kable(summary_tab1,  -->
+<!--   caption = "Parameters used in all population models.", -->
+<!--     format = "latex", booktabs = TRUE) %>% -->
+<!--   kable_styling(font_size = 10) -->
 
-```
+<!-- ``` -->
 
- ```{r, results='asis'}
+<!--  ```{r, results='asis'} -->
 
- Fig_1
+<!--  Fig_1 -->
 
- ```
+<!--  ``` -->
 
@@
 
-<div align="center">Table XX. Settings for our two VAST models.</div>
+<div align="center">Table 3. Settings for our two VAST models.</div>
 
@@
 
+```
+
+```{r scenarios,echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
+
+options(knitr.kable.NA = '')
+
+combos_table = read_csv("CSVs/scenario_combos.csv")
+
+summary_tab <- combos_table %>% 
+   mutate_if(grepl('5deg',.), ~replace(., grepl('5deg', .), paste("Increasing 5","\U00B0","C",sep="")))
+
+# Create table to present the data]
+ kable(summary_tab, "simple",
+                     caption = "Each index estimate chooses one condition from each of the following 7 columns.",
+          format = "latex", booktabs = TRUE) %>%
+          kable_styling(latex_options = "scale_down")
+```
+
+```{r PopPct, echo=FALSE,fig.show='hold', fig.align='center', fig.cap="Percent of each species in each strata for during survey weeks in our spatial simulations.
+All constant temperature scenarios follow the patterns on the left while increasing temperature scenarios follow the patterns on the right.
+See Figure 1 for a spatial reference of the Georges Bank strata.",out.width="47.5%"}
+knitr::include_graphics(c("Images/PercentPlots_ConPopConTemp_YT.png","Images/PercentPlots_ConPopIncTemp_YT.png"))
+knitr::include_graphics(c("Images/PercentPlots_DecPopConTemp_Cod.png","Images/PercentPlots_DecPopIncTemp_Cod.png"))
+knitr::include_graphics(c("Images/PercentPlots_IncPopConTemp_Had.png","Images/PercentPlots_IncPopIncTemp_Had.png"))
+```
+
+```{r CovResponse, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Temperature covariate response plots and resulting population estimate for Yellowtail Flounder on the left and Haddock on the right.
+In both cases the spatial simulations that were surveyed had an increasing population over time, increasing average temperature, and certain strata excluded from sampling, as shown in Figure 1**check**.",out.width="90%"}
+# knitr::include_graphics(c("Images/covariate_good.png","Images/covariate_bad.png"))
+# knitr::include_graphics(c("Images/covariate_est.png"))
+
+knitr::include_graphics(c("Images/covariate_plot.png"))
+
+```
+
+```{r RatioTrend, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Representative example of typical ratio trend for each species, as shown on a log scale.",out.width="90%"}
+# knitr::include_graphics(c("Images/covariate_good.png","Images/covariate_bad.png"))
+# knitr::include_graphics(c("Images/covariate_est.png"))
+#Yellowtail estimates are for the constant population, constant temperature, all strata used in estimates.
+#Cod are Decpop_ConTemp_allStrata_
+#Haddock are IncPop_IncTemp_all strata_
+
+knitr::include_graphics(c("Images/ratio_trend_YTF.PNG"))
+knitr::include_graphics(c("Images/ratio_trend_cod.PNG"))
+knitr::include_graphics(c("Images/ratio_trend_had.PNG"))
 ```

### 2023-05-30 Manuscript/Draft1.tex
manuscript update

@@
 Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics.
-A full list of parameters used in our model can be seen below in Tables \ref{paramsALL} and \ref{tab:paramsSCENARIOS}.
+A full list of parameters used in our model can be seen below in Tables 1 and \ref{tab:paramsSCENARIOS}.
 
@@
 We compare stratified mean estimates to those derived from the model-based VAST approach.
-VAST models both biomass \(p_1(i)\) and presence/absence \(p_2(i)\) for each observation \(i\) as linear predictors using a spatial delta-generalized linear mixed model.
-The equation representing the biomass predictor can be represented by
+VAST models both biomass \(p_1(i)\) and presence/absence \(p_2(i)\) for each observation \(i\) as linear predictors using a spatial delta-generalized linear mixed model that can be represented by
 
 \[
-  \makebox[\linewidth]{$p_1(i)=\text{temporal variation} + \text{spatial variation} + \text{spatio-temporal variation} + \text{vessel effects} + \text{habitat covariates} + \text{catchability covariates}$},
+  \makebox[\linewidth]{$p_1(i)=\text{temporal variation} + \text{spatial variation} + \text{spatio-temporal variation}  + \text{habitat covariates} + \text{catchability covariates}$},
+\]
+
+\[
+  \makebox[\linewidth]{$p_*(i)= \beta_*(t_i) + L_{\omega_*}\omega_*(s_i) + L_* \epsilon_*(c_i,t_i) + \upsilon(s_i,t_i) + \xi(i)$},
 \]
@@
 Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings.
-Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models, which can be seen in Table\ref{tab:VASTsettings}.
+Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models, which can be seen in Table 3.
 VAST Settings B were chosen after reviewing the information provided in the ``Seasonal Model'' section of VAST's online Github tutorial (Thorson, James T n.d.).
@@
 
-\begin{table}
-
-\caption{\label{tab:scenarios}Each index estimate chooses one condition from each of the following 7 columns.}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lllllll}
-\toprule
-Species & Population Trend & Temperature Scenario & Strata Included & Noise Added & Season & Covariates (VAST)\\
-\midrule
-Yellowtail & Increasing & Repeating & All strata & No & Spring & No\\
-Cod & Constant & Increasing 5°C & Subset & Yes & Fall & Yes\\
-Haddock & Decreasing &  &  &  &  & \\
-\bottomrule
-\end{tabular}}
-\end{table}
-
 \section{Results}
 
-Figure \ref{fig:PopPct} depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey in the spring (weeks 13 and 14) and fall (weeks 37 and 28).
+Figure \ref{fig:PopPct} depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey in the spring (weeks 13 and 14) and fall (weeks 37 and 38).
 The left column in Figure \ref{fig:PopPct} depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern.
@@
 We compared a design-based estimation approach (stratified mean) with a model-based approach that allows for the inclusion of environmental covariates (VAST) when the entire domain was sampled as well as when a reduced number of strata were included.
-Our analysis included a comparison of absolute error values between 20-year estimates and consideration of the variation seen year-to-year displayed by each approach as well as the consistency between each 20-year estimate.
-
-Our analysis showed that both approaches are capable of providing abundance estimates that track the true abundance trend.
+Our analysis included a comparison of absolute error values between 20-year estimates and consideration of the year-to-year variation displayed by each approach as well as the variation between each 20-year estimate.
+
+Our analysis showed that the design-based and model-based approaches are each capable of providing biomass estimates that track the true abundance trend.
 While VAST estimates tended to provide lower relative error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B combined to produce over 30 abundance estimates with larger error values than the largest stratified mean error.
@@
 This problem is compounded by the lack of clear guidance regarding how to build and analyze models with the correct inputs.
-Even more concerning, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using the wrong settings \textbf{cite Chris C paper}.
+Even more concerning, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using settings that decrease the accuracy of the model \textbf{cite Chris C paper}.
 The value of model-based approaches such as VAST are clear- they allow users to tailor the model to overcome issues related to the input data by and to account for factors such as environmental change.
@@
 While adding covariates can help inform the model and improve the estimate as seen with the yellowtail flounder simulation results, including covariates can hinder the model in certain instances.
-For example, including covariates for haddock when the northern most strata were removed from the survey meant that the population was not being sampled in the coldest part of its domain which ultimately increased the error.
-
+For example, including covariates for haddock when the northern most strata were removed from the survey meant that the population was not being sampled in the coldest part of its domain which led to an incorrect covariate response that ultimately increased the error between the VAST estimate and the true model abundance.
+
+Seasonal differences in spatial abundance patterns in our population models led to seasonal differences in abundance estimates, with most estimates producing a higher error in the Fall
+spawn weeks 9-14
 something about seasonal impact seen in all estimates with fall estimates typically noticeably worse.
@@
 This was mostly a problem with VAST as SM showed more consistent estimates between seasons.
-
-\emph{Impact of Covariates}
 
@@
 \pagebreak
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.475\linewidth]{Images/PercentPlots_ConPopConTemp_YT} \includegraphics[width=0.475\linewidth]{Images/PercentPlots_ConPopIncTemp_YT} \includegraphics[width=0.475\linewidth]{Images/PercentPlots_DecPopConTemp_Cod} \includegraphics[width=0.475\linewidth]{Images/PercentPlots_DecPopIncTemp_Cod} \includegraphics[width=0.475\linewidth]{Images/PercentPlots_IncPopConTemp_Had} \includegraphics[width=0.475\linewidth]{Images/PercentPlots_IncPopIncTemp_Had}
-
-}
-
-\caption{Percent of each species in each strata for during survey weeks in our spatial simulations.
-All constant temperature scenarios follow the patterns on the left while increasing temperature scenarios follow the patterns on the right.
-See Figure 1??
-for a spatial reference of the Georges Bank strata.}\label{fig:PopPct}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{Images/covariate_plot}
-
-}
-
-\caption{Temperature covariate response plots and resulting population estimate for Yellowtail Flounder on the left and Haddock on the right.
-In both cases the spatial simulations that were surveyed had an increasing population over time, increasing average temperature, and certain strata excluded from sampling, as shown in Figure 1**check**.}\label{fig:CovResponse}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{Images/ratio_trend_YTF} \includegraphics[width=0.9\linewidth]{Images/ratio_trend_cod} \includegraphics[width=0.9\linewidth]{Images/ratio_trend_had}
-
-}
-
-\caption{Representative example of typical ratio trend for each species, as shown on a log scale.}\label{fig:RatioTrend}
-\end{figure}
 
@@
 \end{longtable}
-
-\begin{table}
-
- \caption{\label{tab:paramsALL2}Parameters used in all population models.}
- \centering
- \fontsize{10}{12}\selectfont
- \begin{tabular}[t]{lllllll}
- \toprule
- Parameter & Description & Unit & Yellowtail & Cod & Haddock & Source\\
- \midrule
- <U+03C1> & Ford's growth coefficient & 1/wk & 4.48 & 4.43 & 4.49 & \\
- M & Natural mortality & 1/wk & 0.2064 & 0.2728 & 0.334 & \\
- W1 & Weight of fully recruited fish & kg & 0.39 & 2.95 & 1.12 & \\
- W2 & Weight of pre-recruit fish & kg & 0.13 & 0.39 & 0.19 & \\
- sigma & Variance in recruited fish & kg*kg & 0.55 & 0.55 & 0.55 & \\
- \addlinespace
- lambda & Decay rate for movement & - & 0.7 & 0.7 & 0.7 & \\
- spwn & Spawning weeks for species s & wks & 9-12 & 8-13 & 11-14 & \\
- rec & Recruitment weeks for species s & wks & 9-12 & 8-13 & 11-14 & \\
- \bottomrule
- \end{tabular}
- \end{table}
 
@@
 
-Table XX. Settings for our two VAST models.
+Table 3. Settings for our two VAST models.
 
@@
 \end{longtable}
+
+\begin{table}
+
+\caption{\label{tab:scenarios}Each index estimate chooses one condition from each of the following 7 columns.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lllllll}
+\toprule
+Species & Population Trend & Temperature Scenario & Strata Included & Noise Added & Season & Covariates (VAST)\\
+\midrule
+Yellowtail & Increasing & Repeating & All strata & No & Spring & No\\
+Cod & Constant & Increasing 5°C & Subset & Yes & Fall & Yes\\
+Haddock & Decreasing &  &  &  &  & \\
+\bottomrule
+\end{tabular}}
+\end{table}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.475\linewidth]{Images/PercentPlots_ConPopConTemp_YT} \includegraphics[width=0.475\linewidth]{Images/PercentPlots_ConPopIncTemp_YT} \includegraphics[width=0.475\linewidth]{Images/PercentPlots_DecPopConTemp_Cod} \includegraphics[width=0.475\linewidth]{Images/PercentPlots_DecPopIncTemp_Cod} \includegraphics[width=0.475\linewidth]{Images/PercentPlots_IncPopConTemp_Had} \includegraphics[width=0.475\linewidth]{Images/PercentPlots_IncPopIncTemp_Had}
+
+}
+
+\caption{Percent of each species in each strata for during survey weeks in our spatial simulations.
+All constant temperature scenarios follow the patterns on the left while increasing temperature scenarios follow the patterns on the right.
+See Figure 1 for a spatial reference of the Georges Bank strata.}\label{fig:PopPct}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{Images/covariate_plot}
+
+}
+
+\caption{Temperature covariate response plots and resulting population estimate for Yellowtail Flounder on the left and Haddock on the right.
+In both cases the spatial simulations that were surveyed had an increasing population over time, increasing average temperature, and certain strata excluded from sampling, as shown in Figure 1**check**.}\label{fig:CovResponse}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{Images/ratio_trend_YTF} \includegraphics[width=0.9\linewidth]{Images/ratio_trend_cod} \includegraphics[width=0.9\linewidth]{Images/ratio_trend_had}
+
+}
+
+\caption{Representative example of typical ratio trend for each species, as shown on a log scale.}\label{fig:RatioTrend}
+\end{figure}
 

### 2023-06-01 Manuscript/Draft1.Rmd
manuscript update

@@
 In 2019, New England fisheries produced $22 billion in sales, which sustained over 200,000 jobs [@Econ].
-Maintaining a healthy ecosystem is therefore vital to sustained ecological health and economic prosperity of the region.
+Maintaining a healthy ecosystem is therefore vital to sustained ecosystem health and economic prosperity of the region.
 
@@
 
-The NEFSC has conducted a bottom trawl survey since 1963 to support assessment and management of the fish and invertebrate populations in the region [@azarovitz1981brief; @politis2014northeast].
+The Northeast Fisheries Science Center (NEFSC) has conducted a bottom trawl survey since 1963 to support assessment and management of the fish and invertebrate populations in the region [@azarovitz1981brief; @politis2014northeast].
 The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf.
-The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits.
+The survey has created a rich time series data set with many uses including species-specific habitat identification, age and life history information, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments, and ultimately fishing quota limits.
 The survey takes place twice each year- once in the spring and again in the fall.
@@
 
-Stock assessment scientists choose from a number of approaches to obtain abundance estimates that are derived from the survey data.
-Methods range from traditional design-based estimates to model-based estimates that vary in complexity.
+One component of a stock assessment is an estimation of yearly abundance that is often derived from the bottom trawl survey, when applicable.
+Stock assessment scientists choose from a number of approaches to obtain abundance estimates ranging from traditional design-based estimates to model-based estimates that vary in complexity.
 Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest.
@@
 Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors).
-Model-based estimators, such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM), help account for complex relationships between variables and can help overcome problems with sampling design.
+Model-based estimators help account for complex relationships between variables and can help overcome problems with sampling design.
+Common model-based approaches include General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM).
 
@@
 Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments.
-More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey will change, altering the relationship between the index and the true population [@arreguin1996catchability; @langan2021climate].
+More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey could change, altering the relationship between the index and the true population [@arreguin1996catchability; @langan2021climate].
 A species shifting its range beyond the survey area is an additional compounding factor to consider.
 Existing research has focused on temperature as the driver of such changes [@klein2017effects] and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges [@kerr2022coherence].
-In these scenarios, management strategy evaluations have shown that unintended overfishing can occur resulting from misconceptions of stock status, which can ultimately have detrimental ecologic and economic impacts [@mazur2023consequences].
+In these scenarios, management strategy evaluations have shown that misconceptions of stock status can lead to unintended overfishing, which can ultimately have detrimental ecologic and economic impacts [@mazur2023consequences].
 We are therefore interested in analyzing the impact of climate change on the accuracy of abundance estimates derived from NOAA’s ongoing bottom-trawl survey along the East coast.
@@
 We use the R package *MixFishSim* (MFS) to model our populations [@dolder2020highly].
-MFS is a discrete spatiotemporal simulation tool where users can model multiple species under varying environmental conditions.
+MFS is a discrete spatiotemporal simulation tool that allows users to model multiple species under varying environmental conditions.
 The package uses a delay-difference population model with discrete processes for growth, death, and recruitment of the population.
@@
 
-We obtained a shapefile for the 15 strata that comprise Georges Bank, where strata were partitioned based primarily on depth and secondarily by latitude [@politis2014northeast].
+We obtained a shapefile for the 15 strata that comprise Georges Bank, where strata are partitioned based primarily on depth and secondarily by latitude [@politis2014northeast].
 The region was discritized into a raster with 88 rows and 144 columns to use as our modeling environment.
@@
 A fish stock is considered to be a subpopulation of a species that has similar intrinsic parameters.
-Each of the species being modeled has multiple distinct stocks along the Atlantic coast resulting from local environmental conditions.
-Biological differences between species results in each stock inhabiting a different number of strata on Georges Bank.
+Each of the species being modeled has multiple biologically distinct stocks along the Atlantic coast resulting from local environmental conditions.
+As a result, each stock inhabits a different number of strata on Georges Bank.
 Haddock inhabit all 15 strata in the domain, Cod populate 13 strata, and Yellowtail can be found in 9 strata.
-Figure \@ref(fig:strata-plot) shows the regions used in our models.
+Figure \@ref(fig:strata-plot) shows the strata used in our model for each species.
 
@@
 Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics.
-A full list of parameters used in our model can be seen below in Tables 1 and \@ref(tab:paramsSCENARIOS).
+A full list of parameters used in our model can be seen in Tables 1 and \@ref(tab:paramsSCENARIOS).
 
@@
 The package was designed to generate hypothetical temperature gradients and theoretical habitat preferences using Gaussian Random Fields.
-The following sections describe how we formulated the habitat and temperature components to model real species in the western Atlantic Ocean.
+The following sections describe how we formulated the habitat and temperature components to model our 3 real species on Georges Bank.
 
@@
 The *lrren* tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors.
-We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow.
-We combined data from both the fall and spring surveys to obscure the influence of temperature to allow the niche model to instead infer habitat preferences.
+We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and by representing a successful catch by the biomass of the given tow.
+We combined data from both the fall and spring surveys to obscure the influence of temperature to allow the niche model to instead infer static habitat preferences.
 Depth and mean sediment size were used as our covariate predictors.
@@
 
-Each species is assumed to have normally distributed temperature preferences ($N(\mu,\sigma)$).
+Each species is assumed to have normally distributed temperature preferences $N(\mu,\sigma)$, with mean $\mu$ and standard deviation $\sigma$.
 Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey.
@@
 For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
-To provide a comprehensive analysis of population indexing methods we consider all possible scenario combinations for Yellowtail Flounder.
+To provide a comprehensive analysis of population indexing methods we consider all possible population trend combinations for Yellowtail Flounder.
+The specific population trends used in our analyses can be see in Figure \@ref(fig:pop-scenarios).
 Each of these scenarios is simulated twice: first with with an oscillating temperature gradient that repeats and second with a temperature gradient that increases roughly 5 degrees Celsius over the duration of the 20 year simulation, for a total of 10 simulated spatial datasets.
-The specific population trends used in our analyses can be see in Figure \@ref(fig:pop-scenarios).
 
@@
 The number of the samples taken reflect true target values for each strata and sampling cells were randomly selected.
+An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists.
+These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata.
+To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values.
+Annual survey observations were simulated as log-normal deviations from the underlying “true” survey catches with a CV of 0.3 in the spring.
+The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population.
+We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species.
+The strata to exclude for each species were chosen by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from.
+Figure \@ref(fig:strata-plot) shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \@ref(fig:PopPct).
+The yellow regions in Figure \@ref(fig:strata-plot) depict the strata that were removed from certain stratified random surveys for each species.
 We then use the biomass collected from our samples in contemporary abundance indexing methods to estimate population trends.
@@
 \[
-  \makebox[\linewidth]{$SM_{s,y} = \sum_{i=1}^N W_i \frac{\sum_{k=1}^{S_i}y_{k,i}}{S_i}$},
+  \makebox[\linewidth]{$SM_{s,y} = \sum_{i=1}^N W_i \frac{\sum_{k=1}^{S_i}y_{k,i}}{S_i},$}
 \]
@@
 
+<!-- \[ -->
+<!--   \makebox[\linewidth]{$p_1(i)=\text{temporal variation} + \text{spatial variation} + \text{spatio-temporal variation}  + \text{habitat covariates} + \text{catchability covariates}$}, -->
+<!-- \] -->
+
 \[
-  \makebox[\linewidth]{$p_1(i)=\text{temporal variation} + \text{spatial variation} + \text{spatio-temporal variation}  + \text{habitat covariates} + \text{catchability covariates}$},
+  \makebox[\linewidth]{$p_*(i)= \beta_*(t_i) + L_{\omega_*}\omega_*(s_i) + L_* \epsilon_*(c_i,t_i) + \upsilon(s_i,t_i) + \xi(i),$}
 \]
 
-\[
-  \makebox[\linewidth]{$p_*(i)= \beta_*(t_i) + L_{\omega_*}\omega_*(s_i) + L_* \epsilon_*(c_i,t_i) + \upsilon(s_i,t_i) + \xi(i)$},
-\]
-
-where more specific functional forms of each component are further described in [@thorson2019guidance].
-$p_2(i)$ takes on a similar form as $p_1(i)$.
+for $*=1,2$, where more specific functional forms of each component are further described in [@thorson2019guidance].
 VAST models require numerous user inputs to determine how the linear predictors will be conditioned and solved.
@@
 
-We follow the advice given in [@thorson2019guidance] to build VAST models that estimate biomass in our Georges Bank population models using stratified random samples from our model output.
+We follow the advice given in [@thorson2019guidance] and [@VASTwiki] to build VAST models that estimate biomass in our Georges Bank population models using stratified random samples from our model output.
 In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
 We carried out the same model-building process using covariate information to inform our models as well as without covariates in our models.
-The covariates we considered were the dynamic temperature values and static habitat values ($Hab_{J.s}$) from our population models.
-When using covariates we ultimately decided to consider a best-case scenario by including both temperature and habitat covariates for both linear predictors in order to provide the most information to the model.
+The covariates we considered were the driving factors of movement in our population models- the dynamic temperature values and static habitat values ($Hab_{J.s}$).
+When using covariates we ultimately decided to provide the most information to the model by including both temperature and habitat covariates to consider a best-case scenario.
 Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings.
-Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models, which can be seen in Table 3.
+Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models as shown in Table 3.
 VAST Settings B were chosen after reviewing the information provided in the "Seasonal Model" section of VAST's online Github tutorial [@ThorsonSeasonal].
@@
 
-Our goal is to determine indexing approaches and settings that are robust to changing environmental conditions and resulting spatial biomass patterns.
-An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists.
-These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata.
-To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values.
-Annual survey observations were simulated as log-normal deviations from the underlying “true” survey catches with a CV of 0.3 in the spring.
-The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population.
-We therefore simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species.
-We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from.
-Figure \@ref(fig:strata-plot) shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \@ref(fig:PopPct).
-The yellow regions in Figure \@ref(fig:strata-plot) depict the strata that were removed from certain stratified random surveys for each species.
-
-<!-- Note the Table XXX needs to be added comment above. -->
 Each scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, covariates), resulting in a large number of scenario combinations to consider.
@@
 The left column in Figure \@ref(fig:PopPct) depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern.
-In these scenarios we notice a small amount of shifting between successive years as the population aggregates on especially suitable habitat in the domain.
-More exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time, as seen right column of Figure \@ref(fig:PopPct).
-This demonstrates how we were able to model shifting distributions due to climate change by running simulations with a temperature gradient that increased on average over time.
+We notice a small amount of shifting between successive years in these constant temperature scenarios as the population aggregates on especially suitable habitat in the domain.
+As seen right column of Figure \@ref(fig:PopPct), more exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time.
+By running simulations in *MixFishSim* with a temperature gradient that increased on average over time we were able to create spatial datasets with shifting biomass distributions.
 
-Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the absolute error between abundance estimates and model output for each of our abundance estimates.
-While the model-based results provided a slightly lower mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02) highlighting the sensitivity of model-based estimates to the settings being used.
-When considering individual scenarios, $77/80\approx 96\%$ had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, with VAST models that include covariate information providing the lowest overall errors and standard deviations.
-Setting misspecification and/or a poor covariate response contribute to the large variance in VAST errors as illustrated by the fact that $\approx 63\%$ of individual scenarios contain a VAST estimate with a worst error than the corresponding stratified mean estimate.
-VAST models that included covariates had an average absolute error of 0.22, models that did not include covariate information had an average error of 0.46, and stratified mean estimates produces an average error of 0.39.
+Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the absolute error between biomass estimates and model output for each of our abundance estimates.
+While the model-based results provided a slightly lower overall mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02).
+When considering individual scenarios, while $77/80\approx 96\%$ had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, $63\%$ of individual scenarios contained a VAST estimate with a worst error than the corresponding stratified mean estimate.
+VAST models that included covariate information provided the lowest overall errors and standard deviations.
+Models with covariates had an average absolute error of 0.22, compared to models that did not include covariate information having an average error of 0.46, and stratified mean estimates produces an average error of 0.39.
+These findings demonstrate the ability of spatio-temporal models to account for spatial variation in the species being sampled under the right conditions, but also highlight volatility/sensitivity of model estimates to the settings being used and/or the covariate response detected in the data.
+
 When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.
@@
 A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values.
-This can be further examined by examining yearly estimate:model ratio values, where we divide the yearly abundance estimates by the true model value.
-In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can estimate yearly abundance values as low as 0 at times and exhibit large yearly changes.
+This can be further examined through yearly estimate:model ratio values, where we divide the yearly abundance estimates by the true model value.
+In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can result in yearly abundance values as low as 0 at times and exhibit large yearly changes.
 A representative example of this trend for each species is shown on the log scale in Figure \@ref(fig:RatioTrend).
@@
 Stratified mean ratios were more consistent between seasons with spring ratios resulting in a mean value of 0.91 with a standard deviation of 0.10, and fall values of providing a mean ratio of 0.84 with a standard deviation of 0.16.
-The breakdown for individual species followed a similar pattern.
+The breakdown for individual species followed similar patterns.
 
@@
 The settings used in VAST mattered with Settings B outperforming Settings A in $67/96\approx 70\%$ of scenarios.
-When all strata are sampled, adding covariates greatly improved estimates in all scenarios.
-We still see an improvement in VAST estimates when certain strata are excluded from sampling, but the improvement was much less dramatic than with all strata included.
-There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata).
-However, Including covariate information to each VAST model produced estimates with significantly lower error than their stratified mean counterparts.
+When all strata are sampled, adding covariates improved estimates in all scenarios.
+Including covariates still improve VAST estimates when certain strata are excluded from sampling, but the change in error was smaller than with all strata included.
+There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata), but including covariate information to each VAST model produced estimates with significantly lower error than the corresponding stratified mean estimate
 
@@
 
-In the constant temperature scenario for Cod, in both seasons the population decreases its presence in strata 19 and 20 over the duration of the simulation, while simultaneously increasing presence in stratum 16.
-However, we see a seasonal impact in stratum 16 during the increasing average temperature simulations where in the fall the population decrease presence in strata 16 and 21, and increase presence in 18, 22, and 24 (see Figure \@ref(fig:hab-plot)).
+In the constant temperature scenario for Cod, the population decreases its presence in strata 19 and 20 in both seasons over the duration of the simulation, while simultaneously increasing presence in stratum 16.
+However, we see a seasonal impact in stratum 16 during the increasing average temperature simulations where in the fall the population decrease presence in strata 16 and 21, and increase presence in 18, 22, and 24 (see Figures \@ref(fig:strata-plot) and \@ref(fig:PopPct)).
 Similar to the Yellowtail Flounder population, the favorable habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler.
@@
 Table \@ref(tab:CodERROR) contains the absolute error between our abundance estimates for Cod and the true model values.
-The VAST settings performed comparably to each other with $18/32\approx57%$ of estimates performing better with VAST Settings A.
+The VAST different VAST settings produced similar error values with $18/32\approx57%$ of estimates performing better with VAST Settings A.
 Of the abundance estimates without covariates, $12/16=75$\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
@@
 Table \@ref(tab:HadERROR) contains the absolute error between our abundance estimates for Haddock and the true model values.
-Absolute error values tended to be lower with VAST Settings B about $21/32\approx 66\%$ of scenarios.
+Absolute error values were lower with VAST Settings B in $21/32\approx 66\%$ of scenarios.
 We notice that the model-based VAST produced particularly large errors in spring compared to the stratified mean, with added covariates only improving to the level of the stratified mean.
-VAST shows improved results in fall relative to the stratified mean with added covariates producing extremely low errors in some cases.
+VAST shows improved results in fall relative to the stratified mean with incluidng covariates producing extremely low errors in some cases.
 For Haddock results, adding covariates improves estimates only when all strata are included.
@@
 Adding covariate information resulted in $14/16\approx88$\% of VAST estimates having improved error compared to the stratified mean.
-The 2 scenarios that produced worse error spanned temperature scenarios, but were both in the spring season when the proportion of the population in each strata remained constant in each scenario.
+The 2 scenarios that produced worse error with covarites spanned temperature scenarios, but were both in the spring season when the proportion of the population in each strata remained constant in each scenario.
 The average change in the error of abundance estimates between constant temperature scenarios and increasing temperature scenarios were 1.42 for VAST estimates without covariates, 1.80 for VAST estimates that included covariates, and 1.37 for stratified mean estimates.
@@
 
-Our analysis showed that the design-based and model-based approaches are each capable of providing biomass estimates that track the true abundance trend.
+Our analysis showed that both the design-based and model-based approaches are each capable of providing biomass estimates that track the true abundance trend.
 While VAST estimates tended to provide lower relative error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B combined to produce over 30 abundance estimates with larger error values than the largest stratified mean error.
-This fact highlights how sensitive VAST is to the choices made regarding settings and data used in analyses.
-This problem is compounded by the lack of clear guidance regarding how to build and analyze models with the correct inputs.
+These facts highlight the power of model-based estimates as well as how sensitive they are to the choices made regarding settings and data used in analyses.
+This problem can be compounded by the lack of clear guidance regarding how to build and analyze models with the correct inputs.
 Even more concerning, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using settings that decrease the accuracy of the model **cite Chris C paper**.
@@
 
-Seasonal differences in spatial abundance patterns in our population models led to seasonal differences in abundance estimates, with most estimates producing a higher error in the Fall 
+Seasonal differences in spatial abundance patterns in our population models led to seasonal differences in abundance estimates, with most estimates producing a higher error in the fall.  
 spawn weeks 9-14
@@
 
- Of the ~27% of VAST models where covariates did not improve the VAST estimate, most provided a comparable error (for ex, 0.11 vs 0.13).
-The instances where including covariates provided a significant different in error took place scenarios that included either an increasing temperature and/or reduced sampling domain.
+ Of the ~27% of VAST models where including covariates did not improve the VAST estimate, most provided a comparable error (for ex, 0.11 vs 0.13).
+The instances where including covariates provided a significant different in error took place in scenarios that included either an increasing temperature and/or reduced sampling domain.
 More specifically, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates.
@@
 
-```{r CovResponse, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Temperature covariate response plots and resulting population estimate for Yellowtail Flounder on the left and Haddock on the right.
-In both cases the spatial simulations that were surveyed had an increasing population over time, increasing average temperature, and certain strata excluded from sampling, as shown in Figure 1**check**.",out.width="90%"}
-# knitr::include_graphics(c("Images/covariate_good.png","Images/covariate_bad.png"))
-# knitr::include_graphics(c("Images/covariate_est.png"))
-
-knitr::include_graphics(c("Images/covariate_plot.png"))
-
-```
-
 ```{r RatioTrend, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Representative example of typical ratio trend for each species, as shown on a log scale.",out.width="90%"}
@@
 knitr::include_graphics(c("Images/ratio_trend_had.PNG"))
+```
+
+```{r CovResponse, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Temperature covariate response plots and resulting population estimate for Yellowtail Flounder on the left and Haddock on the right.
+In both cases the spatial simulations that were surveyed had an increasing population over time, increasing average temperature, and certain strata excluded from sampling, as shown in Figure 1**check**.",out.width="90%"}
+# knitr::include_graphics(c("Images/covariate_good.png","Images/covariate_bad.png"))
+# knitr::include_graphics(c("Images/covariate_est.png"))
+
+knitr::include_graphics(c("Images/covariate_plot.png"))
+
 ```

### 2023-06-01 Manuscript/Draft1.tex
manuscript update

@@
 In 2019, New England fisheries produced \$22 billion in sales, which sustained over 200,000 jobs ({``{Fisheries Economics of the United States: Data and Visualizations}''} n.d.).
-Maintaining a healthy ecosystem is therefore vital to sustained ecological health and economic prosperity of the region.
-
-The NEFSC has conducted a bottom trawl survey since 1963 to support assessment and management of the fish and invertebrate populations in the region (Azarovitz 1981; Politis et al.
+Maintaining a healthy ecosystem is therefore vital to sustained ecosystem health and economic prosperity of the region.
+
+The Northeast Fisheries Science Center (NEFSC) has conducted a bottom trawl survey since 1963 to support assessment and management of the fish and invertebrate populations in the region (Azarovitz 1981; Politis et al.
 2014).
 The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf.
-The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits.
+The survey has created a rich time series data set with many uses including species-specific habitat identification, age and life history information, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments, and ultimately fishing quota limits.
 The survey takes place twice each year- once in the spring and again in the fall.
@@
 Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments.
-More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey will change, altering the relationship between the index and the true population (Arreguı́n-Sánchez 1996; Langan et al.
+More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey could change, altering the relationship between the index and the true population (Arreguı́n-Sánchez 1996; Langan et al.
 2021).
@@
 2022).
-In these scenarios, management strategy evaluations have shown that unintended overfishing can occur resulting from misconceptions of stock status, which can ultimately have detrimental ecologic and economic impacts (Mazur et al.
+In these scenarios, management strategy evaluations have shown that misconceptions of stock status can lead to unintended overfishing, which can ultimately have detrimental ecologic and economic impacts (Mazur et al.
 2023).
@@
 2020).
-MFS is a discrete spatiotemporal simulation tool where users can model multiple species under varying environmental conditions.
+MFS is a discrete spatiotemporal simulation tool that allows users to model multiple species under varying environmental conditions.
 The package uses a delay-difference population model with discrete processes for growth, death, and recruitment of the population.
@@
 
-We obtained a shapefile for the 15 strata that comprise Georges Bank, where strata were partitioned based primarily on depth and secondarily by latitude (Politis et al.
+We obtained a shapefile for the 15 strata that comprise Georges Bank, where strata are partitioned based primarily on depth and secondarily by latitude (Politis et al.
 2014).
@@
 A fish stock is considered to be a subpopulation of a species that has similar intrinsic parameters.
-Each of the species being modeled has multiple distinct stocks along the Atlantic coast resulting from local environmental conditions.
-Biological differences between species results in each stock inhabiting a different number of strata on Georges Bank.
+Each of the species being modeled has multiple biologically distinct stocks along the Atlantic coast resulting from local environmental conditions.
+As a result, each stock inhabits a different number of strata on Georges Bank.
 Haddock inhabit all 15 strata in the domain, Cod populate 13 strata, and Yellowtail can be found in 9 strata.
-Figure \ref{fig:strata-plot} shows the regions used in our models.
+Figure \ref{fig:strata-plot} shows the strata used in our model for each species.
 
@@
 Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics.
-A full list of parameters used in our model can be seen below in Tables 1 and \ref{tab:paramsSCENARIOS}.
+A full list of parameters used in our model can be seen in Tables 1 and \ref{tab:paramsSCENARIOS}.
 
@@
 The \emph{lrren} tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors.
-We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow.
-We combined data from both the fall and spring surveys to obscure the influence of temperature to allow the niche model to instead infer habitat preferences.
+We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and by representing a successful catch by the biomass of the given tow.
+We combined data from both the fall and spring surveys to obscure the influence of temperature to allow the niche model to instead infer static habitat preferences.
 Depth and mean sediment size were used as our covariate predictors.
@@
 
-Each species is assumed to have normally distributed temperature preferences (\(N(\mu,\sigma)\)).
+Each species is assumed to have normally distributed temperature preferences \(N(\mu,\sigma)\), with mean \(\mu\) and standard deviation \(\sigma\).
 Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey.
@@
 For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
-To provide a comprehensive analysis of population indexing methods we consider all possible scenario combinations for Yellowtail Flounder.
+To provide a comprehensive analysis of population indexing methods we consider all possible population trend combinations for Yellowtail Flounder.
+The specific population trends used in our analyses can be see in Figure \ref{fig:pop-scenarios}.
 Each of these scenarios is simulated twice: first with with an oscillating temperature gradient that repeats and second with a temperature gradient that increases roughly 5 degrees Celsius over the duration of the 20 year simulation, for a total of 10 simulated spatial datasets.
-The specific population trends used in our analyses can be see in Figure \ref{fig:pop-scenarios}.
 
@@
 The number of the samples taken reflect true target values for each strata and sampling cells were randomly selected.
+An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists.
+These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata.
+To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values.
+Annual survey observations were simulated as log-normal deviations from the underlying ``true'' survey catches with a CV of 0.3 in the spring.
+The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population.
+We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species.
+The strata to exclude for each species were chosen by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from.
+Figure \ref{fig:strata-plot} shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \ref{fig:PopPct}.
+The yellow regions in Figure \ref{fig:strata-plot} depict the strata that were removed from certain stratified random surveys for each species.
 We then use the biomass collected from our samples in contemporary abundance indexing methods to estimate population trends.
@@
 \[
-  \makebox[\linewidth]{$p_1(i)=\text{temporal variation} + \text{spatial variation} + \text{spatio-temporal variation}  + \text{habitat covariates} + \text{catchability covariates}$},
+  \makebox[\linewidth]{$p_*(i)= \beta_*(t_i) + L_{\omega_*}\omega_*(s_i) + L_* \epsilon_*(c_i,t_i) + \upsilon(s_i,t_i) + \xi(i),$}
 \]
 
-\[
-  \makebox[\linewidth]{$p_*(i)= \beta_*(t_i) + L_{\omega_*}\omega_*(s_i) + L_* \epsilon_*(c_i,t_i) + \upsilon(s_i,t_i) + \xi(i)$},
-\]
-
-where more specific functional forms of each component are further described in (Thorson 2019).
-\(p_2(i)\) takes on a similar form as \(p_1(i)\).
+for \(*=1,2\), where more specific functional forms of each component are further described in (Thorson 2019).
 VAST models require numerous user inputs to determine how the linear predictors will be conditioned and solved.
@@
 
-We follow the advice given in (Thorson 2019) to build VAST models that estimate biomass in our Georges Bank population models using stratified random samples from our model output.
+We follow the advice given in (Thorson 2019) and (Thorson, James T n.d.a) to build VAST models that estimate biomass in our Georges Bank population models using stratified random samples from our model output.
 In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
 We carried out the same model-building process using covariate information to inform our models as well as without covariates in our models.
-The covariates we considered were the dynamic temperature values and static habitat values (\(Hab_{J.s}\)) from our population models.
-When using covariates we ultimately decided to consider a best-case scenario by including both temperature and habitat covariates for both linear predictors in order to provide the most information to the model.
+The covariates we considered were the driving factors of movement in our population models- the dynamic temperature values and static habitat values (\(Hab_{J.s}\)).
+When using covariates we ultimately decided to provide the most information to the model by including both temperature and habitat covariates to consider a best-case scenario.
 Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings.
-Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models, which can be seen in Table 3.
-VAST Settings B were chosen after reviewing the information provided in the ``Seasonal Model'' section of VAST's online Github tutorial (Thorson, James T n.d.).
+Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models as shown in Table 3.
+VAST Settings B were chosen after reviewing the information provided in the ``Seasonal Model'' section of VAST's online Github tutorial (Thorson, James T n.d.b).
 After running models for each of our scenarios using these settings, we shared some of our results with the VAST package creator, James (Jim) Thorson.
@@
 
-Our goal is to determine indexing approaches and settings that are robust to changing environmental conditions and resulting spatial biomass patterns.
-An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists.
-These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata.
-To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values.
-Annual survey observations were simulated as log-normal deviations from the underlying ``true'' survey catches with a CV of 0.3 in the spring.
-The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population.
-We therefore simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species.
-We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from.
-Figure \ref{fig:strata-plot} shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \ref{fig:PopPct}.
-The yellow regions in Figure \ref{fig:strata-plot} depict the strata that were removed from certain stratified random surveys for each species.
-
 Each scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, covariates), resulting in a large number of scenario combinations to consider.
@@
 The left column in Figure \ref{fig:PopPct} depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern.
-In these scenarios we notice a small amount of shifting between successive years as the population aggregates on especially suitable habitat in the domain.
-More exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time, as seen right column of Figure \ref{fig:PopPct}.
-This demonstrates how we were able to model shifting distributions due to climate change by running simulations with a temperature gradient that increased on average over time.
-
-Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the absolute error between abundance estimates and model output for each of our abundance estimates.
-While the model-based results provided a slightly lower mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02) highlighting the sensitivity of model-based estimates to the settings being used.
-When considering individual scenarios, \(77/80\approx 96\%\) had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, with VAST models that include covariate information providing the lowest overall errors and standard deviations.
-Setting misspecification and/or a poor covariate response contribute to the large variance in VAST errors as illustrated by the fact that \(\approx 63\%\) of individual scenarios contain a VAST estimate with a worst error than the corresponding stratified mean estimate.
-VAST models that included covariates had an average absolute error of 0.22, models that did not include covariate information had an average error of 0.46, and stratified mean estimates produces an average error of 0.39.
+We notice a small amount of shifting between successive years in these constant temperature scenarios as the population aggregates on especially suitable habitat in the domain.
+As seen right column of Figure \ref{fig:PopPct}, more exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time.
+By running simulations in \emph{MixFishSim} with a temperature gradient that increased on average over time we were able to create spatial datasets with shifting biomass distributions.
+
+Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the absolute error between biomass estimates and model output for each of our abundance estimates.
+While the model-based results provided a slightly lower overall mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02).
+When considering individual scenarios, while \(77/80\approx 96\%\) had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, \(63\%\) of individual scenarios contained a VAST estimate with a worst error than the corresponding stratified mean estimate.
+VAST models that included covariate information provided the lowest overall errors and standard deviations.
+Models with covariates had an average absolute error of 0.22, compared to models that did not include covariate information having an average error of 0.46, and stratified mean estimates produces an average error of 0.39.
+These findings demonstrate the ability of spatio-temporal models to account for spatial variation in the species being sampled under the right conditions, but also highlight volatility/sensitivity of model estimates to the settings being used and/or the covariate response detected in the data.
+
 When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.
@@
 A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values.
-This can be further examined by examining yearly estimate:model ratio values, where we divide the yearly abundance estimates by the true model value.
-In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can estimate yearly abundance values as low as 0 at times and exhibit large yearly changes.
+This can be further examined through yearly estimate:model ratio values, where we divide the yearly abundance estimates by the true model value.
+In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can result in yearly abundance values as low as 0 at times and exhibit large yearly changes.
 A representative example of this trend for each species is shown on the log scale in Figure \ref{fig:RatioTrend}.
@@
 The settings used in VAST mattered with Settings B outperforming Settings A in \(67/96\approx 70\%\) of scenarios.
-When all strata are sampled, adding covariates greatly improved estimates in all scenarios.
-We still see an improvement in VAST estimates when certain strata are excluded from sampling, but the improvement was much less dramatic than with all strata included.
-There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}.
-However, Including covariate information to each VAST model produced estimates with significantly lower error than their stratified mean counterparts.
+When all strata are sampled, adding covariates improved estimates in all scenarios.
+Including covariates still improve VAST estimates when certain strata are excluded from sampling, but the change in error was smaller than with all strata included.
+There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}, but including covariate information to each VAST model produced estimates with significantly lower error than the corresponding stratified mean estimate
 
@@
 
-In the constant temperature scenario for Cod, in both seasons the population decreases its presence in strata 19 and 20 over the duration of the simulation, while simultaneously increasing presence in stratum 16.
-However, we see a seasonal impact in stratum 16 during the increasing average temperature simulations where in the fall the population decrease presence in strata 16 and 21, and increase presence in 18, 22, and 24 (see Figure \ref{fig:hab-plot}).
+In the constant temperature scenario for Cod, the population decreases its presence in strata 19 and 20 in both seasons over the duration of the simulation, while simultaneously increasing presence in stratum 16.
+However, we see a seasonal impact in stratum 16 during the increasing average temperature simulations where in the fall the population decrease presence in strata 16 and 21, and increase presence in 18, 22, and 24 (see Figures \ref{fig:strata-plot} and \ref{fig:PopPct}).
 Similar to the Yellowtail Flounder population, the favorable habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler.
@@
 
-Table \ref{tab:CodERROR} contains the absolute error between our abundance estimates for Cod and the true model values. The VAST settings performed comparably to each other with \(18/32\approx57%
+Table \ref{tab:CodERROR} contains the absolute error between our abundance estimates for Cod and the true model values.
+The VAST different VAST settings produced similar error values with \(18/32\approx57%
 \) of estimates performing better with VAST Settings A.
@@
 Table \ref{tab:HadERROR} contains the absolute error between our abundance estimates for Haddock and the true model values.
-Absolute error values tended to be lower with VAST Settings B about \(21/32\approx 66\%\) of scenarios.
+Absolute error values were lower with VAST Settings B in \(21/32\approx 66\%\) of scenarios.
 We notice that the model-based VAST produced particularly large errors in spring compared to the stratified mean, with added covariates only improving to the level of the stratified mean.
-VAST shows improved results in fall relative to the stratified mean with added covariates producing extremely low errors in some cases.
+VAST shows improved results in fall relative to the stratified mean with incluidng covariates producing extremely low errors in some cases.
 For Haddock results, adding covariates improves estimates only when all strata are included.
@@
 Adding covariate information resulted in \(14/16\approx88\)\% of VAST estimates having improved error compared to the stratified mean.
-The 2 scenarios that produced worse error spanned temperature scenarios, but were both in the spring season when the proportion of the population in each strata remained constant in each scenario.
+The 2 scenarios that produced worse error with covarites spanned temperature scenarios, but were both in the spring season when the proportion of the population in each strata remained constant in each scenario.
 The average change in the error of abundance estimates between constant temperature scenarios and increasing temperature scenarios were 1.42 for VAST estimates without covariates, 1.80 for VAST estimates that included covariates, and 1.37 for stratified mean estimates.
@@
 
-Our analysis showed that the design-based and model-based approaches are each capable of providing biomass estimates that track the true abundance trend.
+Our analysis showed that both the design-based and model-based approaches are each capable of providing biomass estimates that track the true abundance trend.
 While VAST estimates tended to provide lower relative error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B combined to produce over 30 abundance estimates with larger error values than the largest stratified mean error.
-This fact highlights how sensitive VAST is to the choices made regarding settings and data used in analyses.
-This problem is compounded by the lack of clear guidance regarding how to build and analyze models with the correct inputs.
+These facts highlight the power of model-based estimates as well as how sensitive they are to the choices made regarding settings and data used in analyses.
+This problem can be compounded by the lack of clear guidance regarding how to build and analyze models with the correct inputs.
 Even more concerning, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using settings that decrease the accuracy of the model \textbf{cite Chris C paper}.
@@
 
-Seasonal differences in spatial abundance patterns in our population models led to seasonal differences in abundance estimates, with most estimates producing a higher error in the Fall
+Seasonal differences in spatial abundance patterns in our population models led to seasonal differences in abundance estimates, with most estimates producing a higher error in the fall.\\
 spawn weeks 9-14
@@
 
-Of the \textasciitilde27\% of VAST models where covariates did not improve the VAST estimate, most provided a comparable error (for ex, 0.11 vs 0.13).
-The instances where including covariates provided a significant different in error took place scenarios that included either an increasing temperature and/or reduced sampling domain.
+Of the \textasciitilde27\% of VAST models where including covariates did not improve the VAST estimate, most provided a comparable error (for ex, 0.11 vs 0.13).
+The instances where including covariates provided a significant different in error took place in scenarios that included either an increasing temperature and/or reduced sampling domain.
 More specifically, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-ThorsonSeasonal}{}}%
-Thorson, James T. (n.d.). {Seasonal and Interannual Dynamics Github Documentation}. {https://james-thorson-noaa.github.io/docs/tutorials/seasonal-model/}.
+Thorson, James T. (n.d.b). {Seasonal and Interannual Dynamics Github Documentation}. {https://james-thorson-noaa.github.io/docs/tutorials/seasonal-model/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-VASTwiki}{}}%
+Thorson, James T. (n.d.a). {Vector Autoregressive Spatio-Temporal (VAST) model wiki page.} {https://github.com/James-Thorson-NOAA/VAST/wiki}.
 
@@
 
+{\centering \includegraphics[width=0.9\linewidth]{Images/ratio_trend_YTF} \includegraphics[width=0.9\linewidth]{Images/ratio_trend_cod} \includegraphics[width=0.9\linewidth]{Images/ratio_trend_had}
+
+}
+
+\caption{Representative example of typical ratio trend for each species, as shown on a log scale.}\label{fig:RatioTrend}
+\end{figure}
+
+\begin{figure}
+
 {\centering \includegraphics[width=0.9\linewidth]{Images/covariate_plot}
@@
 In both cases the spatial simulations that were surveyed had an increasing population over time, increasing average temperature, and certain strata excluded from sampling, as shown in Figure 1**check**.}\label{fig:CovResponse}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{Images/ratio_trend_YTF} \includegraphics[width=0.9\linewidth]{Images/ratio_trend_cod} \includegraphics[width=0.9\linewidth]{Images/ratio_trend_had}
-
-}
-
-\caption{Representative example of typical ratio trend for each species, as shown on a log scale.}\label{fig:RatioTrend}
 \end{figure}

### 2023-06-09 Manuscript/Draft1.Rmd
Manuscript update including representative plots

@@
 
-We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on Georges Bank, where movement of each species combine static species-specific habitat preferences with biologically-based temperature preferences.
+We construct spatial models for yellowtail flounder, Atlantic Cod, and Haddock on Georges Bank, where movement of each species combine static species-specific habitat preferences with biologically-based temperature preferences.
 Model dynamics are driven by a time series of temperature gradients to create simulated data sets for each population where the true biomass is known.
@@
 Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey.
-We assume Yellowtail Flounder's preferences are $N(8.75,4.25)$, while Haddock and Cod have preferences $N(9,4)$.
+We assume yellowtail flounder's preferences are $N(8.75,4.25)$, while Haddock and Cod have preferences $N(9,4)$.
 Weekly estimated temperature data for the region for 2012 was obtained from FVCOM [@chen2006unstructured].
@@
 For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
-To provide a comprehensive analysis of population indexing methods we consider all possible population trend combinations for Yellowtail Flounder.
+To provide a comprehensive analysis of population indexing methods we consider all possible population trend combinations for yellowtail flounder.
 The specific population trends used in our analyses can be see in Figure \@ref(fig:pop-scenarios).
@@
 
-The top two panels in Figure \@ref(fig:PopPct) depict the results for Yellowtail Flounder with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature).
-In both temperature scenarios we see the percent of Yellowtail Flounder in strata 13 decrease over the course of the time series in both seasons.
+The top two panels in Figure \@ref(fig:PopPct) depict the results for yellowtail flounder with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature).
+In both temperature scenarios we see the percent of yellowtail flounder in strata 13 decrease over the course of the time series in both seasons.
 The spring population in stratum 19 also decreases in both temperature scenarios as well.
 The percent of the population increases in stratum 16 in the spring over the duration of both time series, which implies the flow out of strata 13 and 19 in the spring are going into stratum 16.
-These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail Flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12.
+These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for yellowtail flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12.
 These spring changes are therefore related to the static habitat values in our model rather than the temperature preferences, which is why we seen the same dynamics with constant and increasing temperature.
@@
 
-Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata) contain the absolute error between our abundance estimates for Yellowtail Flounder comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B).
+Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata) contain the absolute error between our abundance estimates for yellowtail flounder comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B).
 In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
@@
 Including covariates still improve VAST estimates when certain strata are excluded from sampling, but the change in error was smaller than with all strata included.
-There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata), but including covariate information to each VAST model produced estimates with significantly lower error than the corresponding stratified mean estimate
+There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata), but including covariate information to each VAST model produced estimates with significantly lower error than the corresponding stratified mean estimate.
 
@@
 These 15 fall estimates span all other scenario variations.
-The implication of this is that our model-based approach without covariates struggles with the primary seasonal difference for Yellowtail Flounder, which is that a larger percentage of the population exists in the narrow strata toward the edge of the domain (18 and/or 17).
+The implication of this is that our model-based approach without covariates struggles with the primary seasonal difference for yellowtail flounder, which is that a larger percentage of the population exists in the narrow strata toward the edge of the domain (18 and/or 17).
 This theory is further supported by the fact that the absolute error in the increasing temperature scenarios increased dramatically in the fall season, when the combined percentage of the population in the outer strata 17 and 18 increased to over 40\% by the end of the simulation.
 
-Including covariate information made a noticeable difference in our Yellowtail Flounder model-based VAST estimates.
+Including covariate information made a noticeable difference in our yellowtail flounder model-based VAST estimates.
 All of the VAST estimates that included covariate information produced a lower relative error than the corresponding VAST model that did not include covariates.
@@
 One exception to this is when we use a stratified mean approach while the population is decreasing.
+A representative example of yellowtail abundance estimates is shown in Figure \@ref(fig:YTEstEx), where one can visualize the impact on biomass estimates between constant and increasing temperature, and how including covariates can improve the estimate.
 Our analyses have found that the stratified mean tends to under estimate the true abundance and since these estimates are bounded below by zero, as the Yellowtail population decreases towards zero the difference between the estimate and the true value also decrease.
@@
 However, we see a seasonal impact in stratum 16 during the increasing average temperature simulations where in the fall the population decrease presence in strata 16 and 21, and increase presence in 18, 22, and 24 (see Figures \@ref(fig:strata-plot) and \@ref(fig:PopPct)).
-Similar to the Yellowtail Flounder population, the favorable habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler.
+Similar to the yellowtail flounder population, the favorable habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler.
 When the temperature increases over time, the fall population compensates by shifting their preference to the adjacent strata that are deeper and/or further north than stratum 16.
@@
 Table \@ref(tab:CodERROR) contains the absolute error between our abundance estimates for Cod and the true model values.
-The VAST different VAST settings produced similar error values with $18/32\approx57%$ of estimates performing better with VAST Settings A.
+The different VAST settings produced similar error values with $18/32\approx57\%$ of estimates performing better with VAST Settings A.
 Of the abundance estimates without covariates, $12/16=75$\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
@@
 Similar to the Yellowtail results, this implies the model-based approach had seasonal trouble with populations shifting into smaller strata where fewer samples take place.
+These impacts are shown in the plots of representative biomass estimates shown in Figure \@ref(fig:CodEstEx).
 Providing covariate information in these cases once again helped the model-based approach to provide improved absolute error estimates relative to the stratified mean.
@@
 
-<!-- Similar to Yellowtail Flounder results, Cod are shifting their spatial distribution into the smaller exterior strata in these conditions, which is decreasing the accuracy of the covariate-free VAST estimate.
+<!-- Similar to yellowtail flounder results, Cod are shifting their spatial distribution into the smaller exterior strata in these conditions, which is decreasing the accuracy of the covariate-free VAST estimate.
 While adding covariates to these fits allowed for an improved error relative to the stratified mean, the VAST abundance still display an overestimate early in the time series before covariates correct the trend, allowing for a lower overall absolute error **(show these plots?
@@
 
-Table \@ref(tab:HadERROR) contains the absolute error between our abundance estimates for Haddock and the true model values.
+Table \@ref(tab:HadERROR) contains the absolute error between our abundance estimates for haddock and the true model values.
 Absolute error values were lower with VAST Settings B in $21/32\approx 66\%$ of scenarios.
 We notice that the model-based VAST produced particularly large errors in spring compared to the stratified mean, with added covariates only improving to the level of the stratified mean.
-VAST shows improved results in fall relative to the stratified mean with incluidng covariates producing extremely low errors in some cases.
-For Haddock results, adding covariates improves estimates only when all strata are included.
+VAST shows improved results in fall relative to the stratified mean with including covariates producing extremely low errors in some cases.
+For haddock results, adding covariates improves estimates only when all strata are included.
 That is, similar to Cod, when sampling a reduced domain adding covariates actually decreases VAST's accuracy.
 Since this occurs in all scenarios, it seems to again be related to failing to accurately monitor stratum 17 near the edge of the domain.
+Some representative biomass estimates for Haddock that demonstrate these results can be seen in Figure \@ref(fig:HadEstEx).
 
@@
 Our analysis showed that both the design-based and model-based approaches are each capable of providing biomass estimates that track the true abundance trend.
-While VAST estimates tended to provide lower relative error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B combined to produce over 30 abundance estimates with larger error values than the largest stratified mean error.
-These facts highlight the power of model-based estimates as well as how sensitive they are to the choices made regarding settings and data used in analyses.
-This problem can be compounded by the lack of clear guidance regarding how to build and analyze models with the correct inputs.
-Even more concerning, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using settings that decrease the accuracy of the model **cite Chris C paper**.
-The value of model-based approaches such as VAST are clear- they allow users to tailor the model to overcome issues related to the input data by and to account for factors such as environmental change.
+While VAST estimates tended to provide lower relative error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B also combined to produce over 30 abundance estimates with larger error values than the largest stratified mean error.
+These facts demonstrate the power of model-based estimates while also highlighting how sensitive they are to the choices made regarding settings and data used in analyses.
+Users often rely on traditional model-building processes and diagnostic techniques such as AIC to guide their choices, but successfully navigating the many decisions depends on the background of the given user and care should be taken to which diagnostic tools are used.
+For example, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using settings that decrease the accuracy of the model **cite Chris C paper**.
+The value of model-based approaches such as VAST are clear- they allow users to overcome issues related to the input data by and to account for factors such as environmental change.
 However, the many consequential decisions that must be made in a given model and the lack of clear criteria for making these decisions makes it challenging to make these choices and even more difficult to determine whether the choices are correct.
 
+Seasonal differences in spatial abundance patterns in our population models led to seasonal differences in abundance estimates, with most estimates producing a higher error in the fall.
+This may have to do with the fact that our species spawn during months 9-14, which coincides with the spring sampling weeks 13 & 14.
+Spawning concentrates much of each population in the spawning grounds towards the center of the domain where more samples occur.
+The populations are more spread out during the fall survey in weeks 37 & 38 and since the water temperature is warmer during these months biomass shifts towards the deeper strata, which tend to be less sampled.
+We explored this through abundance estimate ratios which revealed how the model-based estimates had a larger change in average ratio between seasonal estimates compared to the design-based estimates.
+
 Spatial factors contributed to many of the scenarios that produced larger error values seen in our estimates.
-Flow from the shallow centrally located strata into the more deeper, smaller, and less-sampled eastern strata contributed to VAST failing to accurately modeling the abundance of yellowtail flounder without the addition of covariate information, the impact of which was greatly exacerbated in the increasing temperature scenarios.
-While adding covariates can help inform the model and improve the estimate as seen with the yellowtail flounder simulation results, including covariates can hinder the model in certain instances.
-For example, including covariates for haddock when the northern most strata were removed from the survey meant that the population was not being sampled in the coldest part of its domain which led to an incorrect covariate response that ultimately increased the error between the VAST estimate and the true model abundance.
-
-Seasonal differences in spatial abundance patterns in our population models led to seasonal differences in abundance estimates, with most estimates producing a higher error in the fall.  
-spawn weeks 9-14
-something about seasonal impact seen in all estimates with fall estimates typically noticeably worse.
-We explored this through the ratio plots.
-This was mostly a problem with VAST as SM showed more consistent estimates between seasons.
+Flow from the shallow centrally located strata into the deeper, smaller, and less-sampled eastern strata contributed to VAST failing to accurately modeling the abundance of yellowtail flounder without the addition of covariate information, the impact of which was exacerbated in the increasing temperature scenarios.
+While adding covariates can help inform the model and improve the estimate as seen with the yellowtail flounder simulation results, covariates can hinder the model in certain instances.
 
 <!-- <font size="5">*Impact of Covariates*</font> -->
-
-As seen in Table **cite VAST setting table**, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
-Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73% had a lower absolute error compared to the corresponding estimate without covariates ($117/160\approx 73\%$), and $77/80\approx 96\%$ of VAST models with covariates provided lower absolute error than the corresponding stratified mean estimate.
+<!-- ($117/160\approx 73\%$) -->
+<!-- $77/80\approx 96\%$ -->
+
+As seen in Table 3, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
+Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73% had a lower absolute error compared to the corresponding estimate without covariates , and $\approx 96\, and $77/80\approx 96\%$ of VAST models with covariates provided lower absolute error than the corresponding stratified mean estimate.
 There is more to explore with respect to the impact of covariates.
@@
 Good question for future research** -->
-
+As seen in Table 3, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
+Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73% had a lower absolute error compared to the corresponding estimate without covariates As seen in Table 3, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
+Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73% had a lower absolute error compared to the corresponding estimate without covariates
 Of the ~27% of VAST models where including covariates did not improve the VAST estimate, most provided a comparable error (for ex, 0.11 vs 0.13).
-The instances where including covariates provided a significant different in error took place in scenarios that included either an increasing temperature and/or reduced sampling domain.
-More specifically, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates.
-However, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates.
-The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response.
-Figure \@ref(fig:strata-plot) shows the strata excluded from sampling and Figure \@ref(fig:CovResponse) depicts the resulting covariate response for Yellowtail Flounder and Haddock for the same population scenario.
-Although strata were excluded for both species, the survey samples for Yellowtail Flounder still ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata).
-As a result, when projecting into those strata as seen in the lower portion of Figure \@ref(fig:CovResponse), Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend and estimate the wrong magnitude of biomass.
+The instances where including covariates provided a noticeable change in error took place in scenarios that included either an increasing temperature and/or reduced sampling domain.
+More specifically, when a reduced number of strata are sampled for haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates.
+This is in contrast to our yellowtail flounder results, where adding covariate information to VAST models always decreased the absolute error in the resulting abundance estimates.
+Although strata were excluded for both species, the decline in performance for just Cod and haddock can be explained by a failure to sample the full spectrum of temperature values where the species exists.
+We notice in Figure \@ref(fig:strata-plot) that the strata excluded from sampling for yellowtail flounder still allowed the survey to range across the species preferred values and thus provided a covariate response that improved the estimate.
+On the other hand, the specific strata that were excluded from sampling for haddock were located in the northern region of the domain, which means the covariate response for haddock was not informed by any data from the coldest temperature range.
+This could explain the failure to form a complete covariate response that not only degrades the estimate in the regions being sampled, but also  confounds the ability to project estimates into an unsampled region.
+Figure \@ref(fig:CovResponse) depicts the impact on biomass estimates of including a covariate response for Haddock when both the average temperature and biomass were increasing over time.
+When projecting into the unsampled strata as seen Figure \@ref(fig:CovResponse), projecting biomass into the unsampled strata provides the wrong magnitude of biomass and also a different trend.
 
@@
 \section{Acknowledgements}
-[[we should thank Jim (obviously), but also those who helped you with the habitat data (robyn, david Chev.) and others?
-also should note that funding provided by Northeast Fisheries and Climate program at the Northeast Fisheries Science Center ]]  Also thank Joe Langan for helping develop the proposal that funded this work.
+The authors would like to thank Joe Langan for helping to develop the proposal that funded this work.
+Funding was provided by the Northeast Fisheries and Climate program at the Northeast Fisheries Science Center.
+Thank you to Jim Thorson for consulting with us about VAST settings and implementation.
+We appreciate the help from Robyn Linner for helping to extract FVCOM temperature data and are thankful for David Chevrier for meeting with us to clarify questions related to some of our GIS data.
 
@@
 
-```{r CovResponse, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Temperature covariate response plots and resulting population estimate for Yellowtail Flounder on the left and Haddock on the right.
-In both cases the spatial simulations that were surveyed had an increasing population over time, increasing average temperature, and certain strata excluded from sampling, as shown in Figure 1**check**.",out.width="90%"}
+```{r YTEstEx, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Representative example of bioamass estimate trend for yellowtail flounder.
+Both scenarios have a relatively constant population with all strata being surveyed.
+The main difference is the left column has a repeating temperature gradient while the right column has increasing average temperature.
+",out.width="90%"}
+
+knitr::include_graphics(c("Images/Yellowtail_Estimate_Example.PNG"))
+
+```
+
+```{r CodEstEx, echo=FALSE,out.width="100%",fig.show='hold', fig.align='center', fig.cap="Representative example of bioamass estimate trend for atlantic cod.
+Both scenarios have a decreasing population trend with all strata being surveyed.
+The main difference is the left column has a repeating temperature gradient while the right column has increasing average temperature",out.width="90%"}
+
+knitr::include_graphics(c("Images/Cod_Estimate_Example.PNG"))
+
+```
+
+```{r HadEstEx, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Representative example of bioamass estimate trend for haddock.
+Both scenarios have an increasing population trend and a repeating temperature gradient.
+The main difference is the left column has all strata being surveyed while the right column has certain strata excluded from sampling",out.width="90%"}
+
+knitr::include_graphics(c("Images/Haddock_Estimate_Example.PNG"))
+
+```
+
+```{r CovResponse, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Temperature covariate response plot and resulting population estimate for Haddock when the population was increasing over time, the average tempearture was increasing, and certain strata shown in Figure 1 were excluded from sampling.",out.width="90%"}
 # knitr::include_graphics(c("Images/covariate_good.png","Images/covariate_bad.png"))
@@
 
-knitr::include_graphics(c("Images/covariate_plot.png"))
+knitr::include_graphics(c("Images/covariate_plot3.png"))
 

### 2023-06-09 Manuscript/Draft1.tex
Manuscript update including representative plots

@@
 Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey.
-We assume Yellowtail Flounder's preferences are \(N(8.75,4.25)\), while Haddock and Cod have preferences \(N(9,4)\).
+We assume yellowtail flounder's preferences are \(N(8.75,4.25)\), while Haddock and Cod have preferences \(N(9,4)\).
 Weekly estimated temperature data for the region for 2012 was obtained from FVCOM (Chen et al.
@@
 For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
-To provide a comprehensive analysis of population indexing methods we consider all possible population trend combinations for Yellowtail Flounder.
+To provide a comprehensive analysis of population indexing methods we consider all possible population trend combinations for yellowtail flounder.
 The specific population trends used in our analyses can be see in Figure \ref{fig:pop-scenarios}.
@@
 
-The top two panels in Figure \ref{fig:PopPct} depict the results for Yellowtail Flounder with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature).
-In both temperature scenarios we see the percent of Yellowtail Flounder in strata 13 decrease over the course of the time series in both seasons.
+The top two panels in Figure \ref{fig:PopPct} depict the results for yellowtail flounder with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature).
+In both temperature scenarios we see the percent of yellowtail flounder in strata 13 decrease over the course of the time series in both seasons.
 The spring population in stratum 19 also decreases in both temperature scenarios as well.
 The percent of the population increases in stratum 16 in the spring over the duration of both time series, which implies the flow out of strata 13 and 19 in the spring are going into stratum 16.
-These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail Flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12.
+These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for yellowtail flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12.
 These spring changes are therefore related to the static habitat values in our model rather than the temperature preferences, which is why we seen the same dynamics with constant and increasing temperature.
@@
 
-Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata} contain the absolute error between our abundance estimates for Yellowtail Flounder comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B).
+Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata} contain the absolute error between our abundance estimates for yellowtail flounder comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B).
 In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
@@
 Including covariates still improve VAST estimates when certain strata are excluded from sampling, but the change in error was smaller than with all strata included.
-There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}, but including covariate information to each VAST model produced estimates with significantly lower error than the corresponding stratified mean estimate
+There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}, but including covariate information to each VAST model produced estimates with significantly lower error than the corresponding stratified mean estimate.
 
@@
 One exception to this is when we use a stratified mean approach while the population is decreasing.
+A representative example of yellowtail abundance estimates is shown in Figure \ref{fig:YTEstEx}, where one can visualize the impact on biomass estimates between constant and increasing temperature, and how including covariates can improve the estimate.
 Our analyses have found that the stratified mean tends to under estimate the true abundance and since these estimates are bounded below by zero, as the Yellowtail population decreases towards zero the difference between the estimate and the true value also decrease.
@@
 However, we see a seasonal impact in stratum 16 during the increasing average temperature simulations where in the fall the population decrease presence in strata 16 and 21, and increase presence in 18, 22, and 24 (see Figures \ref{fig:strata-plot} and \ref{fig:PopPct}).
-Similar to the Yellowtail Flounder population, the favorable habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler.
+Similar to the yellowtail flounder population, the favorable habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler.
 When the temperature increases over time, the fall population compensates by shifting their preference to the adjacent strata that are deeper and/or further north than stratum 16.
@@
 Table \ref{tab:CodERROR} contains the absolute error between our abundance estimates for Cod and the true model values.
-The VAST different VAST settings produced similar error values with \(18/32\approx57%
-\) of estimates performing better with VAST Settings A.
+The different VAST settings produced similar error values with \(18/32\approx57\%\) of estimates performing better with VAST Settings A.
 Of the abundance estimates without covariates, \(12/16=75\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
@@
 Similar to the Yellowtail results, this implies the model-based approach had seasonal trouble with populations shifting into smaller strata where fewer samples take place.
+These impacts are shown in the plots of representative biomass estimates shown in Figure \ref{fig:CodEstEx}.
 Providing covariate information in these cases once again helped the model-based approach to provide improved absolute error estimates relative to the stratified mean.
@@
 
-Table \ref{tab:HadERROR} contains the absolute error between our abundance estimates for Haddock and the true model values.
+Table \ref{tab:HadERROR} contains the absolute error between our abundance estimates for haddock and the true model values.
 Absolute error values were lower with VAST Settings B in \(21/32\approx 66\%\) of scenarios.
 We notice that the model-based VAST produced particularly large errors in spring compared to the stratified mean, with added covariates only improving to the level of the stratified mean.
-VAST shows improved results in fall relative to the stratified mean with incluidng covariates producing extremely low errors in some cases.
-For Haddock results, adding covariates improves estimates only when all strata are included.
+VAST shows improved results in fall relative to the stratified mean with including covariates producing extremely low errors in some cases.
+For haddock results, adding covariates improves estimates only when all strata are included.
 That is, similar to Cod, when sampling a reduced domain adding covariates actually decreases VAST's accuracy.
 Since this occurs in all scenarios, it seems to again be related to failing to accurately monitor stratum 17 near the edge of the domain.
+Some representative biomass estimates for Haddock that demonstrate these results can be seen in Figure \ref{fig:HadEstEx}.
 
@@
 Our analysis showed that both the design-based and model-based approaches are each capable of providing biomass estimates that track the true abundance trend.
-While VAST estimates tended to provide lower relative error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B combined to produce over 30 abundance estimates with larger error values than the largest stratified mean error.
-These facts highlight the power of model-based estimates as well as how sensitive they are to the choices made regarding settings and data used in analyses.
-This problem can be compounded by the lack of clear guidance regarding how to build and analyze models with the correct inputs.
-Even more concerning, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using settings that decrease the accuracy of the model \textbf{cite Chris C paper}.
-The value of model-based approaches such as VAST are clear- they allow users to tailor the model to overcome issues related to the input data by and to account for factors such as environmental change.
+While VAST estimates tended to provide lower relative error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B also combined to produce over 30 abundance estimates with larger error values than the largest stratified mean error.
+These facts demonstrate the power of model-based estimates while also highlighting how sensitive they are to the choices made regarding settings and data used in analyses.
+Users often rely on traditional model-building processes and diagnostic techniques such as AIC to guide their choices, but successfully navigating the many decisions depends on the background of the given user and care should be taken to which diagnostic tools are used.
+For example, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using settings that decrease the accuracy of the model \textbf{cite Chris C paper}.
+The value of model-based approaches such as VAST are clear- they allow users to overcome issues related to the input data by and to account for factors such as environmental change.
 However, the many consequential decisions that must be made in a given model and the lack of clear criteria for making these decisions makes it challenging to make these choices and even more difficult to determine whether the choices are correct.
 
+Seasonal differences in spatial abundance patterns in our population models led to seasonal differences in abundance estimates, with most estimates producing a higher error in the fall.
+This may have to do with the fact that our species spawn during months 9-14, which coincides with the spring sampling weeks 13 \& 14.
+Spawning concentrates much of each population in the spawning grounds towards the center of the domain where more samples occur.
+The populations are more spread out during the fall survey in weeks 37 \& 38 and since the water temperature is warmer during these months biomass shifts towards the deeper strata, which tend to be less sampled.
+We explored this through abundance estimate ratios which revealed how the model-based estimates had a larger change in average ratio between seasonal estimates compared to the design-based estimates.
+
 Spatial factors contributed to many of the scenarios that produced larger error values seen in our estimates.
-Flow from the shallow centrally located strata into the more deeper, smaller, and less-sampled eastern strata contributed to VAST failing to accurately modeling the abundance of yellowtail flounder without the addition of covariate information, the impact of which was greatly exacerbated in the increasing temperature scenarios.
-While adding covariates can help inform the model and improve the estimate as seen with the yellowtail flounder simulation results, including covariates can hinder the model in certain instances.
-For example, including covariates for haddock when the northern most strata were removed from the survey meant that the population was not being sampled in the coldest part of its domain which led to an incorrect covariate response that ultimately increased the error between the VAST estimate and the true model abundance.
-
-Seasonal differences in spatial abundance patterns in our population models led to seasonal differences in abundance estimates, with most estimates producing a higher error in the fall.\\
-spawn weeks 9-14
-something about seasonal impact seen in all estimates with fall estimates typically noticeably worse.
-We explored this through the ratio plots.
-This was mostly a problem with VAST as SM showed more consistent estimates between seasons.
-
-As seen in Table \textbf{cite VAST setting table}, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
-Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73\% had a lower absolute error compared to the corresponding estimate without covariates (\(117/160\approx 73\%\)), and \(77/80\approx 96\%\) of VAST models with covariates provided lower absolute error than the corresponding stratified mean estimate.
+Flow from the shallow centrally located strata into the deeper, smaller, and less-sampled eastern strata contributed to VAST failing to accurately modeling the abundance of yellowtail flounder without the addition of covariate information, the impact of which was exacerbated in the increasing temperature scenarios.
+While adding covariates can help inform the model and improve the estimate as seen with the yellowtail flounder simulation results, covariates can hinder the model in certain instances.
+
+As seen in Table 3, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
+Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73\% had a lower absolute error compared to the corresponding estimate without covariates , and \$\approx 96, and \(77/80\approx 96\%\) of VAST models with covariates provided lower absolute error than the corresponding stratified mean estimate.
 There is more to explore with respect to the impact of covariates.
@@
 One could also test the impact of assuming the wrong covariate response function (linear vs polynomial etc).
-
+As seen in Table 3, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
+Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73\% had a lower absolute error compared to the corresponding estimate without covariates As seen in Table 3, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
+Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73\% had a lower absolute error compared to the corresponding estimate without covariates
 Of the \textasciitilde27\% of VAST models where including covariates did not improve the VAST estimate, most provided a comparable error (for ex, 0.11 vs 0.13).
-The instances where including covariates provided a significant different in error took place in scenarios that included either an increasing temperature and/or reduced sampling domain.
-More specifically, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates.
-However, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates.
-The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response.
-Figure \ref{fig:strata-plot} shows the strata excluded from sampling and Figure \ref{fig:CovResponse} depicts the resulting covariate response for Yellowtail Flounder and Haddock for the same population scenario.
-Although strata were excluded for both species, the survey samples for Yellowtail Flounder still ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata).
-As a result, when projecting into those strata as seen in the lower portion of Figure \ref{fig:CovResponse}, Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend and estimate the wrong magnitude of biomass.
+The instances where including covariates provided a noticeable change in error took place in scenarios that included either an increasing temperature and/or reduced sampling domain.
+More specifically, when a reduced number of strata are sampled for haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates.
+This is in contrast to our yellowtail flounder results, where adding covariate information to VAST models always decreased the absolute error in the resulting abundance estimates.
+Although strata were excluded for both species, the decline in performance for just Cod and haddock can be explained by a failure to sample the full spectrum of temperature values where the species exists.
+We notice in Figure \ref{fig:strata-plot} that the strata excluded from sampling for yellowtail flounder still allowed the survey to range across the species preferred values and thus provided a covariate response that improved the estimate.
+On the other hand, the specific strata that were excluded from sampling for haddock were located in the northern region of the domain, which means the covariate response for haddock was not informed by any data from the coldest temperature range.
+This could explain the failure to form a complete covariate response that not only degrades the estimate in the regions being sampled, but also confounds the ability to project estimates into an unsampled region.
+Figure \ref{fig:CovResponse} depicts the impact on biomass estimates of including a covariate response for Haddock when both the average temperature and biomass were increasing over time.
+When projecting into the unsampled strata as seen Figure \ref{fig:CovResponse}, projecting biomass into the unsampled strata provides the wrong magnitude of biomass and also a different trend.
 
@@
 
-{[}{[}we should thank Jim (obviously), but also those who helped you with the habitat data (robyn, david Chev.) and others?
-also should note that funding provided by Northeast Fisheries and Climate program at the Northeast Fisheries Science Center {]}{]} Also thank Joe Langan for helping develop the proposal that funded this work.
+The authors would like to thank Joe Langan for helping to develop the proposal that funded this work.
+Funding was provided by the Northeast Fisheries and Climate program at the Northeast Fisheries Science Center.
+Thank you to Jim Thorson for consulting with us about VAST settings and implementation.
+We appreciate the help from Robyn Linner for helping to extract FVCOM temperature data and are thankful for David Chevrier for meeting with us to clarify questions related to some of our GIS data.
 
@@
 
-{\centering \includegraphics[width=0.9\linewidth]{Images/covariate_plot}
+{\centering \includegraphics[width=0.9\linewidth]{Images/Yellowtail_Estimate_Example}
 
@@
 
-\caption{Temperature covariate response plots and resulting population estimate for Yellowtail Flounder on the left and Haddock on the right.
-In both cases the spatial simulations that were surveyed had an increasing population over time, increasing average temperature, and certain strata excluded from sampling, as shown in Figure 1**check**.}\label{fig:CovResponse}
+\caption{Representative example of bioamass estimate trend for yellowtail flounder.
+Both scenarios have a relatively constant population with all strata being surveyed.
+The main difference is the left column has a repeating temperature gradient while the right column has increasing average temperature.
+}\label{fig:YTEstEx}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{Images/Cod_Estimate_Example}
+
+}
+
+\caption{Representative example of bioamass estimate trend for atlantic cod.
+Both scenarios have a decreasing population trend with all strata being surveyed.
+The main difference is the left column has a repeating temperature gradient while the right column has increasing average temperature}\label{fig:CodEstEx}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{Images/Haddock_Estimate_Example}
+
+}
+
+\caption{Representative example of bioamass estimate trend for haddock.
+Both scenarios have an increasing population trend and a repeating temperature gradient.
+The main difference is the left column has all strata being surveyed while the right column has certain strata excluded from sampling}\label{fig:HadEstEx}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{Images/covariate_plot3}
+
+}
+
+\caption{Temperature covariate response plot and resulting population estimate for Haddock when the population was increasing over time, the average tempearture was increasing, and certain strata shown in Figure 1 were excluded from sampling.}\label{fig:CovResponse}
 \end{figure}

### 2023-06-09 TestScripts/Plots_for_paper_presentation.R

@@ -212,6 +212,99 @@ labs(x="year",y="Biomass", title ="Haddock Fall Estimate", color ="" )+
 
 
 
+#COD DEC POP CON TEMP ALL STRATA AND DEC POP INC TEMP ALL STRATA
+
+model_line_size = 1.5
+model_point_size = 3.5
+estimate_line_size = 1.25
+estimate_point_size = 3
+
+for(season_ in c(1,2)){
+  
+  ssnn = ifelse(season_==1, "Spring Estimate", "Fall Estimate")
+  
+  print(ggplot() +
+          
+          
+          #this way plots data by season
+          geom_point(data = subset(as.data.frame(pop_by_season[[s]]),season==season_), aes(x=as.numeric(year),y=biomass, group = season, color = "Model"),size=model_point_size) +
+          geom_line(data = subset(as.data.frame(pop_by_season[[s]]),season==season_), aes(x=as.numeric(year),y=biomass, group =season, color = "Model"),size=model_line_size) +
+          
+          
+          
+          #plot VAST estimate without covariates with NO noise
+          geom_errorbar(data=subset(VAST_data[[s]][["_NoCovs_"]][["NoNoise_"]][[folder]],season==season_),aes(x=Year,y=Estimate,group=season,ymin=Estimate-(1.96*Std..Error.for.Estimate), ymax=Estimate+(1.96*Std..Error.for.Estimate), color = "VAST No Cov No Noise"),width=.3) +
+          #geom_linerange(data=subset(VAST_data[[s]][[folder]],season==2),aes(x=Year,y=Estimate,group=season,ymin=Estimate-(1.96*Std..Error.for.Estimate), ymax=Estimate+(1.96*Std..Error.for.Estimate), color = "VAST Estimate")) +
+          geom_point(data=subset(VAST_data[[s]][["_NoCovs_"]][["NoNoise_"]][[folder]],season==season_),aes(x=Year,y=Estimate,group=season, color = "VAST No Cov No Noise"), size=estimate_point_size)+
+          geom_line(data=subset(VAST_data[[s]][["_NoCovs_"]][["NoNoise_"]][[folder]],season==season_),aes(x=Year,y=Estimate,group=season, color = "VAST No Cov No Noise"),size=estimate_line_size)+
+          # 
+          # #plot VAST estimate without covariates with WITH noise
[…]

### 2023-06-09 TestScripts/VAST_StratMean_Error_Comparison_All_withCovs_ForPaper.R

@@ -14,9 +14,9 @@ scenario1 <- "IncPop_ConTemp" #the folder name
 
 #spp1 spp2 spp3
 #short_names <- c("YT","Cod","Had")   #fixed above
-short_names <- c("YT","Had")
+short_names <- c("Had")
 
-exclude_strata <- FALSE
+exclude_strata <- TRUE
 
 #for getting into correct subfolder
 ifelse(exclude_strata==TRUE, 

### 2023-06-14 Manuscript/Draft1.Rmd
update manuscript

@@
 fontsize: 12pt
+indent: true
 header-includes:
@@
 #read in results tables
-YT_results<-read.csv("CSVs/Yellowtail_results.csv")
-Cod_results<-read.csv("CSVs/Cod_results.csv")
-Had_results<-read.csv("CSVs/Haddock_results.csv")
+#YT_results<-read.csv("CSVs/Yellowtail_results.csv")
+#Cod_results<-read.csv("CSVs/Cod_results.csv")
+#Had_results<-read.csv("CSVs/Haddock_results.csv")
 ```
@@
 
-An Abstract
+Many stock assessments in the United States use abundance estimates derived from stratified random bottom trawl data.
+To accurately represent true abundance, catches of a species must contain a low enough noise level to allow for a discernible pattern and the proportion of the population that is sampled should be consistent over time.
+These assumptions could be violated given enough noise in the sampling process and/or climate change causing a population to increase its abundance in strata that are not sampled.
+Using the R package MixFishSim, we have developed data-driven spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock in the western Atlantic Ocean to allow examination of these assumptions through simulation.
+Movement rates combine species-specific static habitat preferences with temperature tolerances.
+Habitat preferences were derived from niche models relating bottom trawl catches to environmental covariates.
+A repeating yearly temperature pattern produces repeating spatial biomass distributions in a given week, while a temperature gradient that increases on average over time results in spatial preferences that evolve throughout a given simulation.
+We created simulated spatial time series datasets for each species for several temperature scenarios and population trends.
+Using stratified random sampling on model output we were able to compare abundance estimates derived from the design-based stratified mean to those using a spatio-temporal model-based approach that allows inclusion of environmental covariates (VAST).
+Our focus was on the ability of contemporary indexing methods to track population trends under shifting spatial preferences.
+
+<!-- We found that in general, both index calculation methods are able to produce similar abundance trends to the model output, with model-based estimates providing more consistent year-to-year estimates.
+The design-based approach tended to provide abundance estimates that were below the true values in our models, while model-based estimates tended to overestimate true abundance.
+We highlight scenarios where index calculation methods produced misleading estimates as species shifted their spatial densities over time within their assumed stock boundaries.
+While including environmental covariates can improve model-based estimates, the outcome is sensitive to the multitude of inputs and settings.
+The impact of shifting spatial densities became magnified when only a portion of a species’ spatial distribution was sampled.
+Counter-intuitively, including covariates can sometimes degrade the estimates in this situation.
+-->
 
@@
 
-keyword 1, keyword 2
+survey, design-based, model-based, VAST, population simulation, climate change
 
@@
 
-The Northeast United States continental shelf spans from the Outer Banks of North Carolina to the Gulf of Maine.
+The continental shelf off the coast of the Northeast United States spans from the Outer Banks of North Carolina to the Gulf of Maine.
 The region covers over 250,000 km$^2$ of ocean, extending over 200 km from shore in the largest areas in New England to just 30 km off shore in the southern regions.
-This ecologically diverse region contains approximately 18,000 vertebrate marine species.
-Commercial fisheries have been an important part of local economies for centuries.
-In 2019, New England fisheries produced $22 billion in sales, which sustained over 200,000 jobs [@Econ].
+This ecologically diverse region contains approximately 18,000 vertebrate marine species, making commercial fisheries an important part of local economies for centuries.
+In 2019, New England fisheries produced $22 billion in sales, which supported over 200,000 jobs [@Econ].
 Maintaining a healthy ecosystem is therefore vital to sustained ecosystem health and economic prosperity of the region.
@@
 Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest.
-These methods do not account for spatial variation in samples and are not able to account for environmental influences on survey values.
+These methods do not account for spatial variation in samples and are not able to relate environmental influences with survey values.
 Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors).
-Model-based estimators help account for complex relationships between variables and can help overcome problems with sampling design.
+Model-based estimators help account for complex relationships between variables and can help overcome problems related to the sampling process.
 Common model-based approaches include General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM).
@@
 In these scenarios, management strategy evaluations have shown that misconceptions of stock status can lead to unintended overfishing, which can ultimately have detrimental ecologic and economic impacts [@mazur2023consequences].
-We are therefore interested in analyzing the impact of climate change on the accuracy of abundance estimates derived from NOAA’s ongoing bottom-trawl survey along the East coast.
+We are therefore interested in analyzing the impact of climate change on the accuracy of abundance estimates derived from NOAA’s ongoing bottom-trawl survey along the east coast.
 
@@
 
-We construct spatial models for yellowtail flounder, Atlantic Cod, and Haddock on Georges Bank, where movement of each species combine static species-specific habitat preferences with biologically-based temperature preferences.
-Model dynamics are driven by a time series of temperature gradients to create simulated data sets for each population where the true biomass is known.
+We construct spatial population models for Yellowtail Flounder (*Limanda ferruginea*), Atlantic Cod (*Gadus morhua*), and Haddock (*Melanogrammus aeglefinus*) on Georges Bank, hereafter referred to as Yellowtail, Cod, and Haddock.
+Movement of each species combine static species-specific habitat preferences with biologically-based temperature preferences.
+Model dynamics are driven by a time series of temperature gradients to create simulated spatial data sets for each population where the true biomass is known.
 Using temperature gradients that repeat each year creates data sets with repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time.
@@
 The package uses a delay-difference population model with discrete processes for growth, death, and recruitment of the population.
-We formulate the following inputs for the MFS package to address our research question.
+To address our research question we formulate the following inputs for the MFS package.
 
@@
 ```{r strata-plot, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Strata inhabited by each species in our population models.
-All strata on the eastern US contintental shelf are shown in the first image with Georges Bank highlighted in orange.
+All strata on the eastern US contintental shelf are shown in the first image with Georges Bank highlighted in orange [@walsh2015long] **fix cite**.
 Stratum numbers used by the NEFSC bottom trawl survey are shown in the middle row.
@@
 knitr::include_graphics(c("Images/GB_strata.jpg"))
+
 knitr::include_graphics(c("Images/Strata2.png"))
@@
 
-The package combines species-specific temperature tolerances with habitat preferences to drive the probability of movement from cell $I$ to cell $J$ using the formulation
+The package combines species-specific temperature tolerances with habitat preferences to drive the probability of movement from cell $I$ to cell $J$ in week $wk+1$ using the formulation
 
@@
 The package was designed to generate hypothetical temperature gradients and theoretical habitat preferences using Gaussian Random Fields.
-The following sections describe how we formulated the habitat and temperature components to model our 3 real species on Georges Bank.
+The following sections describe how we formulated the habitat and temperature components to model our three real species on Georges Bank.
 
@@
 The *lrren* tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors.
-We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and by representing a successful catch by the biomass of the given tow.
-We combined data from both the fall and spring surveys to obscure the influence of temperature to allow the niche model to instead infer static habitat preferences.
+We used bottom trawl point data from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and by representing a successful catch by the biomass of the given tow.
+We combined data from both the fall and spring surveys to obscure seasonal influences to allow the niche model to instead infer static habitat preferences independent of temperature.
 Depth and mean sediment size were used as our covariate predictors.
@@
 
-```{r hab-plot1, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Visual representation of niche model for Cod."}
+```{r hab-plot1, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Visual representation of niche model for Atlantic Cod."}
 knitr::include_graphics(c("Images/hab_snip3.png"))
@@
 ```{r hab-plot, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Static habitat preferences for each species in our population models.
-From left to right: Yellotwtail Flounder, Cod, and Haddock."}
+From left to right: Yellotwtail Flounder, Atlantic Cod, and Haddock."}
 knitr::include_graphics(c("Images/Habitat_3species.png"))
@@
 Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey.
-We assume yellowtail flounder's preferences are $N(8.75,4.25)$, while Haddock and Cod have preferences $N(9,4)$.
+We assume Yellowtail's preferences are $N(8.75,4.25)$, while Haddock and Cod have preferences $N(9,4)$.
 Weekly estimated temperature data for the region for 2012 was obtained from FVCOM [@chen2006unstructured].
@@
 The 2012 temperature data was also transformed to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation.
-We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation.
+We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on spatial patterns while remaining within reasonable computational limits in terms of the length of the simulation.
 Figure \@ref(fig:temp-scenarios) depicts mean trends for the temperature scenarios used in our models.
@@
 In equation (1), $Hab^2_{J,s}$ remains constant for each species for the duration of the simulation, while $Tol_{c,s,wk}$ changes each week with temperature fluctuations.
-Using a temperature gradient that repeats every 52 weeks produces the same spatial preferences in a given week each year, resulting in consistent spatial biomass patterns.
+Using a temperature gradient that repeats every 52 weeks produces the same spatial preferences in a given week each year, resulting in relatively consistent spatial biomass patterns.
 Scenarios where the temperature increases over time creates spatial preferences that evolve as the water warms, producing spatial biomass patterns that shift in a given week over the duration of the simulation.
-Thus, stratified random samples in scenarios with a repeating temperature pattern will have constant survey catchability and availability over time, which may not be true for increasing temperature scenarios due to evolving spatial preferences.
 
@@
 We carry out 20 year simulations for each of our three species under various population scenarios.
-Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year [[can cite the 2022 management track assessments, see https://apps-nefsc.fisheries.noaa.gov/saw/sari.php for when the document becomes available]] **cite**.
+Historically, Cod has seen significant decline over the last 50**??** years while Haddock has increased in abundance in recent years [[can cite the 2022 management track assessments, see https://apps-nefsc.fisheries.noaa.gov/saw/sari.php for when the document becomes available]] **just cite 2021 assessments?**.
 For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
-To provide a comprehensive analysis of population indexing methods we consider all possible population trend combinations for yellowtail flounder.
+We consider increasing, decreasing, and constant population trends for Yellowtail for completeness.
 The specific population trends used in our analyses can be see in Figure \@ref(fig:pop-scenarios).
@@
 
-After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year.
-We sample each strata in the same weeks in which the Spring and Fall surveys take place (weeks 13 & 14 in the Spring and 37 & 38 in the Fall).
+After completing all simulations, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year.
+We sample each strata in the same weeks in which the spring and fall surveys take place (weeks 13 & 14 in the spring and 37 & 38 in the fall).
 The number of the samples taken reflect true target values for each strata and sampling cells were randomly selected.
+
 An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists.
@@
 To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values.
-Annual survey observations were simulated as log-normal deviations from the underlying “true” survey catches with a CV of 0.3 in the spring.
+Annual survey observations were simulated as log-normal deviations from the underlying “true” survey catches with a CV of 0.3.
 The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population.
@@
 
-We follow the advice given in [@thorson2019guidance] and [@VASTwiki] to build VAST models that estimate biomass in our Georges Bank population models using stratified random samples from our model output.
+We follow the advice given in [@thorson2019guidance] and [@VASTwiki] to build VAST estimates of biomass in our Georges Bank spatial population models using stratified random samples from our simulation output.
 In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
 We carried out the same model-building process using covariate information to inform our models as well as without covariates in our models.
-The covariates we considered were the driving factors of movement in our population models- the dynamic temperature values and static habitat values ($Hab_{J.s}$).
-When using covariates we ultimately decided to provide the most information to the model by including both temperature and habitat covariates to consider a best-case scenario.
+We considered the driving factors of movement in our population models as covariates- the dynamic temperature values and static habitat values ($Hab_{J.s}$).
+We ultimately decided to provide the most information to the model by including exact values for both temperature and habitat covariates to consider a best-case scenario.
+
 Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings.
@@
 According to Jim, the Settings A values are representative of what stock assessment scientists typically use in their analyses.
-As a result we will focus our analysis on VAST results obtained using Settings A.
-
-Each scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, covariates), resulting in a large number of scenario combinations to consider.
+As a result we will focus our analysis on VAST results obtained using Settings A, but present Settings B results to demonstrate how the settings used in model-based approaches impact model performance.
+
+Each indexing scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, & covariates), resulting in a large number of scenario combinations to consider.
 The columns in Table \@ref(tab:scenarios) show the choices that define each scenario.
@@
 We notice a small amount of shifting between successive years in these constant temperature scenarios as the population aggregates on especially suitable habitat in the domain.
-As seen right column of Figure \@ref(fig:PopPct), more exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time.
 By running simulations in *MixFishSim* with a temperature gradient that increased on average over time we were able to create spatial datasets with shifting biomass distributions.
+As seen in the right column of Figure \@ref(fig:PopPct), more exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time.
 
@@
 While the model-based results provided a slightly lower overall mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02).
-When considering individual scenarios, while $77/80\approx 96\%$ had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, $63\%$ of individual scenarios contained a VAST estimate with a worst error than the corresponding stratified mean estimate.
+When considering individual scenarios, while about $96\%$ had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, $63\%$ of individual scenarios contained a VAST estimate with a worse error than the corresponding stratified mean estimate.
 VAST models that included covariate information provided the lowest overall errors and standard deviations.
@@
 
-The top two panels in Figure \@ref(fig:PopPct) depict the results for yellowtail flounder with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature).
-In both temperature scenarios we see the percent of yellowtail flounder in strata 13 decrease over the course of the time series in both seasons.
+The top two panels in Figure \@ref(fig:PopPct) depict the results for Yellowtail with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature).
+In both temperature scenarios we see the percent of Yellowtail in stratum 13 decrease over the course of the time series in both seasons.
 The spring population in stratum 19 also decreases in both temperature scenarios as well.
 The percent of the population increases in stratum 16 in the spring over the duration of both time series, which implies the flow out of strata 13 and 19 in the spring are going into stratum 16.
-These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for yellowtail flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12.
+These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12.
 These spring changes are therefore related to the static habitat values in our model rather than the temperature preferences, which is why we seen the same dynamics with constant and increasing temperature.
@@
 
-Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata) contain the absolute error between our abundance estimates for yellowtail flounder comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B).
+Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata) contain the absolute error between our abundance estimates for Yellowtail comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B).
 In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
@@
 Including covariates still improve VAST estimates when certain strata are excluded from sampling, but the change in error was smaller than with all strata included.
-There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata), but including covariate information to each VAST model produced estimates with significantly lower error than the corresponding stratified mean estimate.
+There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata).
+However, including covariate information to each VAST model produced estimates with significantly lower error than the corresponding stratified mean estimate.
 
@@
 These 15 fall estimates span all other scenario variations.
-The implication of this is that our model-based approach without covariates struggles with the primary seasonal difference for yellowtail flounder, which is that a larger percentage of the population exists in the narrow strata toward the edge of the domain (18 and/or 17).
+The implication of this is that our model-based approach without covariates struggles with the primary seasonal difference for Yellowtail, which is that a larger percentage of the population exists in the narrow strata toward the edge of the domain (18 and/or 17).
 This theory is further supported by the fact that the absolute error in the increasing temperature scenarios increased dramatically in the fall season, when the combined percentage of the population in the outer strata 17 and 18 increased to over 40\% by the end of the simulation.
 
-Including covariate information made a noticeable difference in our yellowtail flounder model-based VAST estimates.
+Including covariate information made a noticeable difference in our Yellowtail model-based VAST estimates.
 All of the VAST estimates that included covariate information produced a lower relative error than the corresponding VAST model that did not include covariates.
@@
 This implies that the covariate information helped our model-based estimate account for the design-based issues related to an increasing percentage of the population entering smaller strata, which can become exacerbated in the increasing temperature simulations.
+A representative example of Yellowtail abundance estimates is shown in Figure \@ref(fig:YTEstEx), where one can visualize the impact on biomass estimates between constant and increasing temperature, and how including covariates can improve the estimate.
 
@@
 One exception to this is when we use a stratified mean approach while the population is decreasing.
-A representative example of yellowtail abundance estimates is shown in Figure \@ref(fig:YTEstEx), where one can visualize the impact on biomass estimates between constant and increasing temperature, and how including covariates can improve the estimate.
 Our analyses have found that the stratified mean tends to under estimate the true abundance and since these estimates are bounded below by zero, as the Yellowtail population decreases towards zero the difference between the estimate and the true value also decrease.
@@
 However, we see a seasonal impact in stratum 16 during the increasing average temperature simulations where in the fall the population decrease presence in strata 16 and 21, and increase presence in 18, 22, and 24 (see Figures \@ref(fig:strata-plot) and \@ref(fig:PopPct)).
-Similar to the yellowtail flounder population, the favorable habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler.
+Similar to the Yellowtail population, the favorable habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler.
 When the temperature increases over time, the fall population compensates by shifting their preference to the adjacent strata that are deeper and/or further north than stratum 16.
@@
 Similar to the Yellowtail results, this implies the model-based approach had seasonal trouble with populations shifting into smaller strata where fewer samples take place.
-These impacts are shown in the plots of representative biomass estimates shown in Figure \@ref(fig:CodEstEx).
+These impacts are shown in the plots of representative biomass estimates for Cod shown in Figure \@ref(fig:CodEstEx).
 Providing covariate information in these cases once again helped the model-based approach to provide improved absolute error estimates relative to the stratified mean.
@@
 
-In comparing abundances estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a large negative impact on the model-based estimates.
+In comparing abundances estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a noticable negative impact on the model-based estimates.
 Specifically, when comparing the error of abundance estimates from the constant temperature scenario to the corresponding increasing temperature scenario for Cod, VAST increased by an average factor of 3.57 without covariates and 1.67 with covariates.
-In considering the average change in the error of stratified mean estimates between constant temperature scenarios and increasing temperature scenarios we surprisingly find an improvement in the error of abundance estimates, with an average decrease by a factor of 0.87.
-
-<!-- Similar to yellowtail flounder results, Cod are shifting their spatial distribution into the smaller exterior strata in these conditions, which is decreasing the accuracy of the covariate-free VAST estimate.
+In considering the average change in the error of stratified mean estimates between constant temperature scenarios and increasing temperature scenarios we surprisingly find an improvement in the error of abundance estimates, decreasing by an average factor of 0.87.
+
+<!-- Similar to Yellowtail results, Cod are shifting their spatial distribution into the smaller exterior strata in these conditions, which is decreasing the accuracy of the covariate-free VAST estimate.
 While adding covariates to these fits allowed for an improved error relative to the stratified mean, the VAST abundance still display an overestimate early in the time series before covariates correct the trend, allowing for a lower overall absolute error **(show these plots?
@@
 
-Figure \@ref(fig:PopPct) reveals some subtle seasonal differences in the percent of haddock in each strata.
+Figure \@ref(fig:PopPct) reveals some subtle seasonal differences in the percent of Haddock in each strata.
 In the constant temperature scenarios, the spring shows a decrease in strata 19 and 20 that correspond to increases in 16, 24 and 29.
@@
 
-Table \@ref(tab:HadERROR) contains the absolute error between our abundance estimates for haddock and the true model values.
+Table \@ref(tab:HadERROR) contains the absolute error between our abundance estimates for Haddock and the true model values.
 Absolute error values were lower with VAST Settings B in $21/32\approx 66\%$ of scenarios.
-We notice that the model-based VAST produced particularly large errors in spring compared to the stratified mean, with added covariates only improving to the level of the stratified mean.
+We notice that VAST produced particularly large errors in spring compared to the stratified mean, with added covariates only improving to the level of the stratified mean.
 VAST shows improved results in fall relative to the stratified mean with including covariates producing extremely low errors in some cases.
-For haddock results, adding covariates improves estimates only when all strata are included.
+For Haddock results, adding covariates improves estimates only when all strata are included.
 That is, similar to Cod, when sampling a reduced domain adding covariates actually decreases VAST's accuracy.
@@
 15/16 of the times the stratified mean was better VAST was not using covariates.
-The 1 time that VAST was using covariates and was still worse than the stratified mean was IncPop_IncTemp_Allstrata_WCov_WNoise in the Fall (VAST had strong overestimate in the fall during IncPOP_IncTemP_allstrata for Had and YT).
+The 1 time that VAST was using covariates and was still worse than the stratified mean was IncPop_IncTemp_Allstrata_WCov_WNoise in the fall (VAST had strong overestimate in the fall during IncPOP_IncTemP_allstrata for Had and YT).
 The overestimate in the fall only seems to be related to the way the population shifts between season.
@@
 
-To test the impact of shifting populations on abundance estimates, we simulated known population data that included environmental forcing and compared biomass estimates derived from stratified random sampling to the true values.
+To test the impact of shifting populations on abundance estimates, we simulated population data that included environmental forcing and compared biomass estimates derived from stratified random sampling to the known true values.
 We carried out simulations for 3 species on Georges Bank that took into account different population trends, temperature scenarios, noise in survey samples, and seasonal variation.
-We compared a design-based estimation approach (stratified mean) with a model-based approach that allows for the inclusion of environmental covariates (VAST) when the entire domain was sampled as well as when a reduced number of strata were included.
+We compared a design-based estimation approach (stratified mean) with a model-based approach that allows for the inclusion of environmental covariates (VAST) when the entire domain was sampled as well as when a reduced number of strata were included to simulate our species shifting into a new habitat.
 Our analysis included a comparison of absolute error values between 20-year estimates and consideration of the year-to-year variation displayed by each approach as well as the variation between each 20-year estimate.
@@
 For example, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using settings that decrease the accuracy of the model **cite Chris C paper**.
-The value of model-based approaches such as VAST are clear- they allow users to overcome issues related to the input data by and to account for factors such as environmental change.
+The value of model-based approaches such as VAST are clear- they allow users to overcome issues related to the input data to account for factors such as environmental change.
 However, the many consequential decisions that must be made in a given model and the lack of clear criteria for making these decisions makes it challenging to make these choices and even more difficult to determine whether the choices are correct.
@@
 This may have to do with the fact that our species spawn during months 9-14, which coincides with the spring sampling weeks 13 & 14.
-Spawning concentrates much of each population in the spawning grounds towards the center of the domain where more samples occur.
+Spawning in our population models concentrates much of each species in the spawning grounds towards the center of the domain where more samples occur.
 The populations are more spread out during the fall survey in weeks 37 & 38 and since the water temperature is warmer during these months biomass shifts towards the deeper strata, which tend to be less sampled.
@@
 Spatial factors contributed to many of the scenarios that produced larger error values seen in our estimates.
-Flow from the shallow centrally located strata into the deeper, smaller, and less-sampled eastern strata contributed to VAST failing to accurately modeling the abundance of yellowtail flounder without the addition of covariate information, the impact of which was exacerbated in the increasing temperature scenarios.
-While adding covariates can help inform the model and improve the estimate as seen with the yellowtail flounder simulation results, covariates can hinder the model in certain instances.
+Flow from the shallow centrally located strata into the deeper, smaller, and less-sampled eastern strata contributed to VAST failing to accurately modeling the abundance of Yellowtail without the addition of covariate information, the impact of which was exacerbated in the increasing temperature scenarios.
+While adding covariates can help inform the model and improve the estimate as seen with the Yellowtail simulation results, covariates can hinder the model in certain instances with Cod and Haddock.
 
@@
 As seen in Table 3, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
-Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73% had a lower absolute error compared to the corresponding estimate without covariates , and $\approx 96\, and $77/80\approx 96\%$ of VAST models with covariates provided lower absolute error than the corresponding stratified mean estimate.
-There is more to explore with respect to the impact of covariates.
-For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates.
-Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information.
-One could also test the impact of assuming the wrong covariate response function (linear vs polynomial etc).
-<!-- **For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without "perfect" information?
+We considered a best-case scenario by providing covariate information to VAST in the form of the exact habitat and temperature values that governed movement from our population models.
+This covariate information typically resulted in improved estimates, with 73% of VAST estimates with covariates providing a lower absolute error compared to the corresponding estimate without covariates, and 96% of VAST models with covariates resulting in a lower absolute error than the corresponding stratified mean estimate.
+In reality the covariates that influence a given species would be less clear and the collected values would contain sampling noise.
+One could test the impact these differences by assuming the wrong covariate response function (linear vs polynomial etc) or by including each of the covariates individually.
+Including noise in the covariate data would further explore how robust the model-based estimates are to additional uncertainty in the covariate information.
+
+<!-- **For discussion: clearly our perfect covariate information made a big difference for Yellowtail (and others), but the question is could we achieve similar results without "perfect" information?
 Which covariate is best?
 Good question for future research** -->
-As seen in Table 3, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
-Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73% had a lower absolute error compared to the corresponding estimate without covariates As seen in Table 3, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
-Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73% had a lower absolute error compared to the corresponding estimate without covariates
+
 Of the ~27% of VAST models where including covariates did not improve the VAST estimate, most provided a comparable error (for ex, 0.11 vs 0.13).
 The instances where including covariates provided a noticeable change in error took place in scenarios that included either an increasing temperature and/or reduced sampling domain.
-More specifically, when a reduced number of strata are sampled for haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates.
-This is in contrast to our yellowtail flounder results, where adding covariate information to VAST models always decreased the absolute error in the resulting abundance estimates.
-Although strata were excluded for both species, the decline in performance for just Cod and haddock can be explained by a failure to sample the full spectrum of temperature values where the species exists.
-We notice in Figure \@ref(fig:strata-plot) that the strata excluded from sampling for yellowtail flounder still allowed the survey to range across the species preferred values and thus provided a covariate response that improved the estimate.
-On the other hand, the specific strata that were excluded from sampling for haddock were located in the northern region of the domain, which means the covariate response for haddock was not informed by any data from the coldest temperature range.
+More specifically, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates.
+This is in contrast to our Yellowtail results, where adding covariate information to VAST models always decreased the absolute error in the resulting abundance estimates.
+Although strata were excluded for both species, the decline in performance for just Cod and Haddock can be explained by a failure to sample the full spectrum of temperature values where the species exists.
+We notice in Figure \@ref(fig:strata-plot) that the strata excluded from sampling for Yellowtail still allowed the survey to range across the species preferred values and thus provided a covariate response that improved the estimate.
+On the other hand, the specific strata that were excluded from sampling for Haddock were located in the northern region of the domain, which means the covariate response for Haddock was not informed by any data from the coldest temperature range.
 This could explain the failure to form a complete covariate response that not only degrades the estimate in the regions being sampled, but also  confounds the ability to project estimates into an unsampled region.
@@
 
-```{r YTEstEx, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Representative example of bioamass estimate trend for yellowtail flounder.
+```{r YTEstEx, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Representative example of bioamass estimate trend for Yellowtail.
 Both scenarios have a relatively constant population with all strata being surveyed.
@@
 
-```{r CodEstEx, echo=FALSE,out.width="100%",fig.show='hold', fig.align='center', fig.cap="Representative example of bioamass estimate trend for atlantic cod.
+```{r CodEstEx, echo=FALSE,out.width="100%",fig.show='hold', fig.align='center', fig.cap="Representative example of bioamass estimate trend for Atlantic Cod.
 Both scenarios have a decreasing population trend with all strata being surveyed.
@@
 
-```{r HadEstEx, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Representative example of bioamass estimate trend for haddock.
+```{r HadEstEx, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Representative example of bioamass estimate trend for Haddock.
 Both scenarios have an increasing population trend and a repeating temperature gradient.
@@
       align = c("l", "l", "l", "l", "c", "c", "c"),
-  caption = "Yellowtail flounder error results with all strata included in calculations. Row colors correspond to the same settings applied in different seasons.") %>%
+  caption = "Yellowtail Flounder error results with all strata included in calculations. Row colors correspond to the same settings applied in different seasons.") %>%
   row_spec(0,bold = TRUE)%>% #bold column titles
@@
       align = c("l", "l", "l", "l", "c", "c", "c"),
-  caption = "Yellowtail flounder error results with certain strata excluded from calculations. Row colors correspond to the same settings applied in different seasons.") %>%
+  caption = "Yellowtail Flounder error results with certain strata excluded from calculations. Row colors correspond to the same settings applied in different seasons.") %>%
   row_spec(0,bold = TRUE)%>% #bold column titles

### 2023-06-14 Manuscript/Draft1.tex
update manuscript

@@
 }{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
 \usepackage{xcolor}
@@
 
-An Abstract
+Many stock assessments in the United States use abundance estimates derived from stratified random bottom trawl data.
+To accurately represent true abundance, catches of a species must contain a low enough noise level to allow for a discernible pattern and the proportion of the population that is sampled should be consistent over time.
+These assumptions could be violated given enough noise in the sampling process and/or climate change causing a population to increase its abundance in strata that are not sampled.
+Using the R package MixFishSim, we have developed data-driven spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock in the western Atlantic Ocean to allow examination of these assumptions through simulation.
+Movement rates combine species-specific static habitat preferences with temperature tolerances.
+Habitat preferences were derived from niche models relating bottom trawl catches to environmental covariates.
+A repeating yearly temperature pattern produces repeating spatial biomass distributions in a given week, while a temperature gradient that increases on average over time results in spatial preferences that evolve throughout a given simulation.
+We created simulated spatial time series datasets for each species for several temperature scenarios and population trends.
+Using stratified random sampling on model output we were able to compare abundance estimates derived from the design-based stratified mean to those using a spatio-temporal model-based approach that allows inclusion of environmental covariates (VAST).
+Our focus was on the ability of contemporary indexing methods to track population trends under shifting spatial preferences.
 
@@
 
-The Northeast United States continental shelf spans from the Outer Banks of North Carolina to the Gulf of Maine.
+The continental shelf off the coast of the Northeast United States spans from the Outer Banks of North Carolina to the Gulf of Maine.
 The region covers over 250,000 km\(^2\) of ocean, extending over 200 km from shore in the largest areas in New England to just 30 km off shore in the southern regions.
-This ecologically diverse region contains approximately 18,000 vertebrate marine species.
-Commercial fisheries have been an important part of local economies for centuries.
-In 2019, New England fisheries produced \$22 billion in sales, which sustained over 200,000 jobs ({``{Fisheries Economics of the United States: Data and Visualizations}''} n.d.).
+This ecologically diverse region contains approximately 18,000 vertebrate marine species, making commercial fisheries an important part of local economies for centuries.
+In 2019, New England fisheries produced \$22 billion in sales, which supported over 200,000 jobs ({``{Fisheries Economics of the United States: Data and Visualizations}''} n.d.).
 Maintaining a healthy ecosystem is therefore vital to sustained ecosystem health and economic prosperity of the region.
@@
 2023).
-We are therefore interested in analyzing the impact of climate change on the accuracy of abundance estimates derived from NOAA's ongoing bottom-trawl survey along the East coast.
+We are therefore interested in analyzing the impact of climate change on the accuracy of abundance estimates derived from NOAA's ongoing bottom-trawl survey along the east coast.
 
@@
 
-We construct spatial models for yellowtail flounder, Atlantic Cod, and Haddock on Georges Bank, where movement of each species combine static species-specific habitat preferences with biologically-based temperature preferences.
-Model dynamics are driven by a time series of temperature gradients to create simulated data sets for each population where the true biomass is known.
+We construct spatial population models for Yellowtail Flounder (\emph{Limanda ferruginea}), Atlantic Cod (\emph{Gadus morhua}), and Haddock (\emph{Melanogrammus aeglefinus}) on Georges Bank, hereafter referred to as Yellowtail, Cod, and Haddock.
+Movement of each species combine static species-specific habitat preferences with biologically-based temperature preferences.
+Model dynamics are driven by a time series of temperature gradients to create simulated spatial data sets for each population where the true biomass is known.
 Using temperature gradients that repeat each year creates data sets with repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time.
@@
 \caption{Strata inhabited by each species in our population models.
-All strata on the eastern US contintental shelf are shown in the first image with Georges Bank highlighted in orange.
+All strata on the eastern US contintental shelf are shown in the first image with Georges Bank highlighted in orange [@walsh2015long] **fix cite**.
 Stratum numbers used by the NEFSC bottom trawl survey are shown in the middle row.
@@
 
-The package combines species-specific temperature tolerances with habitat preferences to drive the probability of movement from cell \(I\) to cell \(J\) using the formulation
+The package combines species-specific temperature tolerances with habitat preferences to drive the probability of movement from cell \(I\) to cell \(J\) in week \(wk+1\) using the formulation
 
@@
 The \emph{lrren} tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors.
-We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and by representing a successful catch by the biomass of the given tow.
-We combined data from both the fall and spring surveys to obscure the influence of temperature to allow the niche model to instead infer static habitat preferences.
+We used bottom trawl point data from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and by representing a successful catch by the biomass of the given tow.
+We combined data from both the fall and spring surveys to obscure seasonal influences to allow the niche model to instead infer static habitat preferences independent of temperature.
 Depth and mean sediment size were used as our covariate predictors.
@@
 
-\caption{Visual representation of niche model for Cod.}\label{fig:hab-plot1}
+\caption{Visual representation of niche model for Atlantic Cod.}\label{fig:hab-plot1}
 \end{figure}
@@
 
-\caption{Static habitat preferences for each species in our population models. From left to right: Yellotwtail Flounder, Cod, and Haddock.}\label{fig:hab-plot}
+\caption{Static habitat preferences for each species in our population models. From left to right: Yellotwtail Flounder, Atlantic Cod, and Haddock.}\label{fig:hab-plot}
 \end{figure}
@@
 Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey.
-We assume yellowtail flounder's preferences are \(N(8.75,4.25)\), while Haddock and Cod have preferences \(N(9,4)\).
+We assume Yellowtail's preferences are \(N(8.75,4.25)\), while Haddock and Cod have preferences \(N(9,4)\).
 Weekly estimated temperature data for the region for 2012 was obtained from FVCOM (Chen et al.
@@
 The 2012 temperature data was also transformed to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation.
-We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation.
+We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on spatial patterns while remaining within reasonable computational limits in terms of the length of the simulation.
 Figure \ref{fig:temp-scenarios} depicts mean trends for the temperature scenarios used in our models.
@@
 In equation (1), \(Hab^2_{J,s}\) remains constant for each species for the duration of the simulation, while \(Tol_{c,s,wk}\) changes each week with temperature fluctuations.
-Using a temperature gradient that repeats every 52 weeks produces the same spatial preferences in a given week each year, resulting in consistent spatial biomass patterns.
+Using a temperature gradient that repeats every 52 weeks produces the same spatial preferences in a given week each year, resulting in relatively consistent spatial biomass patterns.
 Scenarios where the temperature increases over time creates spatial preferences that evolve as the water warms, producing spatial biomass patterns that shift in a given week over the duration of the simulation.
-Thus, stratified random samples in scenarios with a repeating temperature pattern will have constant survey catchability and availability over time, which may not be true for increasing temperature scenarios due to evolving spatial preferences.
 
 We carry out 20 year simulations for each of our three species under various population scenarios.
-Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year {[}{[}can cite the 2022 management track assessments, see \url{https://apps-nefsc.fisheries.noaa.gov/saw/sari.php} for when the document becomes available{]}{]} \textbf{cite}.
+Historically, Cod has seen significant decline over the last 50\textbf{??} years while Haddock has increased in abundance in recent years {[}{[}can cite the 2022 management track assessments, see \url{https://apps-nefsc.fisheries.noaa.gov/saw/sari.php} for when the document becomes available{]}{]} \textbf{just cite 2021 assessments?}.
 For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
-To provide a comprehensive analysis of population indexing methods we consider all possible population trend combinations for yellowtail flounder.
+We consider increasing, decreasing, and constant population trends for Yellowtail for completeness.
 The specific population trends used in our analyses can be see in Figure \ref{fig:pop-scenarios}.
@@
 
-After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year.
-We sample each strata in the same weeks in which the Spring and Fall surveys take place (weeks 13 \& 14 in the Spring and 37 \& 38 in the Fall).
+After completing all simulations, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year.
+We sample each strata in the same weeks in which the spring and fall surveys take place (weeks 13 \& 14 in the spring and 37 \& 38 in the fall).
 The number of the samples taken reflect true target values for each strata and sampling cells were randomly selected.
+
 An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists.
@@
 To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values.
-Annual survey observations were simulated as log-normal deviations from the underlying ``true'' survey catches with a CV of 0.3 in the spring.
+Annual survey observations were simulated as log-normal deviations from the underlying ``true'' survey catches with a CV of 0.3.
 The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population.
@@
 
-We follow the advice given in (Thorson 2019) and (Thorson, James T n.d.a) to build VAST models that estimate biomass in our Georges Bank population models using stratified random samples from our model output.
+We follow the advice given in (Thorson 2019) and (Thorson, James T n.d.a) to build VAST estimates of biomass in our Georges Bank spatial population models using stratified random samples from our simulation output.
 In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
 We carried out the same model-building process using covariate information to inform our models as well as without covariates in our models.
-The covariates we considered were the driving factors of movement in our population models- the dynamic temperature values and static habitat values (\(Hab_{J.s}\)).
-When using covariates we ultimately decided to provide the most information to the model by including both temperature and habitat covariates to consider a best-case scenario.
+We considered the driving factors of movement in our population models as covariates- the dynamic temperature values and static habitat values (\(Hab_{J.s}\)).
+We ultimately decided to provide the most information to the model by including exact values for both temperature and habitat covariates to consider a best-case scenario.
+
 Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings.
@@
 According to Jim, the Settings A values are representative of what stock assessment scientists typically use in their analyses.
-As a result we will focus our analysis on VAST results obtained using Settings A.
-
-Each scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, covariates), resulting in a large number of scenario combinations to consider.
+As a result we will focus our analysis on VAST results obtained using Settings A, but present Settings B results to demonstrate how the settings used in model-based approaches impact model performance.
+
+Each indexing scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, \& covariates), resulting in a large number of scenario combinations to consider.
 The columns in Table \ref{tab:scenarios} show the choices that define each scenario.
@@
 We notice a small amount of shifting between successive years in these constant temperature scenarios as the population aggregates on especially suitable habitat in the domain.
-As seen right column of Figure \ref{fig:PopPct}, more exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time.
 By running simulations in \emph{MixFishSim} with a temperature gradient that increased on average over time we were able to create spatial datasets with shifting biomass distributions.
+As seen in the right column of Figure \ref{fig:PopPct}, more exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time.
 
@@
 While the model-based results provided a slightly lower overall mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02).
-When considering individual scenarios, while \(77/80\approx 96\%\) had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, \(63\%\) of individual scenarios contained a VAST estimate with a worst error than the corresponding stratified mean estimate.
+When considering individual scenarios, while about \(96\%\) had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, \(63\%\) of individual scenarios contained a VAST estimate with a worse error than the corresponding stratified mean estimate.
 VAST models that included covariate information provided the lowest overall errors and standard deviations.
@@
 
-The top two panels in Figure \ref{fig:PopPct} depict the results for yellowtail flounder with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature).
-In both temperature scenarios we see the percent of yellowtail flounder in strata 13 decrease over the course of the time series in both seasons.
+The top two panels in Figure \ref{fig:PopPct} depict the results for Yellowtail with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature).
+In both temperature scenarios we see the percent of Yellowtail in stratum 13 decrease over the course of the time series in both seasons.
 The spring population in stratum 19 also decreases in both temperature scenarios as well.
 The percent of the population increases in stratum 16 in the spring over the duration of both time series, which implies the flow out of strata 13 and 19 in the spring are going into stratum 16.
-These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for yellowtail flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12.
+These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12.
 These spring changes are therefore related to the static habitat values in our model rather than the temperature preferences, which is why we seen the same dynamics with constant and increasing temperature.
@@
 
-Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata} contain the absolute error between our abundance estimates for yellowtail flounder comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B).
+Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata} contain the absolute error between our abundance estimates for Yellowtail comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B).
 In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
@@
 Including covariates still improve VAST estimates when certain strata are excluded from sampling, but the change in error was smaller than with all strata included.
-There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}, but including covariate information to each VAST model produced estimates with significantly lower error than the corresponding stratified mean estimate.
+There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}.
+However, including covariate information to each VAST model produced estimates with significantly lower error than the corresponding stratified mean estimate.
 
@@
 This implies that the covariate information helped our model-based estimate account for the design-based issues related to an increasing percentage of the population entering smaller strata, which can become exacerbated in the increasing temperature simulations.
+A representative example of Yellowtail abundance estimates is shown in Figure \ref{fig:YTEstEx}, where one can visualize the impact on biomass estimates between constant and increasing temperature, and how including covariates can improve the estimate.
 
@@
 One exception to this is when we use a stratified mean approach while the population is decreasing.
-A representative example of yellowtail abundance estimates is shown in Figure \ref{fig:YTEstEx}, where one can visualize the impact on biomass estimates between constant and increasing temperature, and how including covariates can improve the estimate.
 Our analyses have found that the stratified mean tends to under estimate the true abundance and since these estimates are bounded below by zero, as the Yellowtail population decreases towards zero the difference between the estimate and the true value also decrease.
@@
 However, we see a seasonal impact in stratum 16 during the increasing average temperature simulations where in the fall the population decrease presence in strata 16 and 21, and increase presence in 18, 22, and 24 (see Figures \ref{fig:strata-plot} and \ref{fig:PopPct}).
-Similar to the yellowtail flounder population, the favorable habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler.
+Similar to the Yellowtail population, the favorable habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler.
 When the temperature increases over time, the fall population compensates by shifting their preference to the adjacent strata that are deeper and/or further north than stratum 16.
@@
 Similar to the Yellowtail results, this implies the model-based approach had seasonal trouble with populations shifting into smaller strata where fewer samples take place.
-These impacts are shown in the plots of representative biomass estimates shown in Figure \ref{fig:CodEstEx}.
+These impacts are shown in the plots of representative biomass estimates for Cod shown in Figure \ref{fig:CodEstEx}.
 Providing covariate information in these cases once again helped the model-based approach to provide improved absolute error estimates relative to the stratified mean.
@@
 
-In comparing abundances estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a large negative impact on the model-based estimates.
+In comparing abundances estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a noticable negative impact on the model-based estimates.
 Specifically, when comparing the error of abundance estimates from the constant temperature scenario to the corresponding increasing temperature scenario for Cod, VAST increased by an average factor of 3.57 without covariates and 1.67 with covariates.
-In considering the average change in the error of stratified mean estimates between constant temperature scenarios and increasing temperature scenarios we surprisingly find an improvement in the error of abundance estimates, with an average decrease by a factor of 0.87.
+In considering the average change in the error of stratified mean estimates between constant temperature scenarios and increasing temperature scenarios we surprisingly find an improvement in the error of abundance estimates, decreasing by an average factor of 0.87.
 
@@
 
-Figure \ref{fig:PopPct} reveals some subtle seasonal differences in the percent of haddock in each strata.
+Figure \ref{fig:PopPct} reveals some subtle seasonal differences in the percent of Haddock in each strata.
 In the constant temperature scenarios, the spring shows a decrease in strata 19 and 20 that correspond to increases in 16, 24 and 29.
@@
 
-Table \ref{tab:HadERROR} contains the absolute error between our abundance estimates for haddock and the true model values.
+Table \ref{tab:HadERROR} contains the absolute error between our abundance estimates for Haddock and the true model values.
 Absolute error values were lower with VAST Settings B in \(21/32\approx 66\%\) of scenarios.
-We notice that the model-based VAST produced particularly large errors in spring compared to the stratified mean, with added covariates only improving to the level of the stratified mean.
+We notice that VAST produced particularly large errors in spring compared to the stratified mean, with added covariates only improving to the level of the stratified mean.
 VAST shows improved results in fall relative to the stratified mean with including covariates producing extremely low errors in some cases.
-For haddock results, adding covariates improves estimates only when all strata are included.
+For Haddock results, adding covariates improves estimates only when all strata are included.
 That is, similar to Cod, when sampling a reduced domain adding covariates actually decreases VAST's accuracy.
@@
 For example, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using settings that decrease the accuracy of the model \textbf{cite Chris C paper}.
-The value of model-based approaches such as VAST are clear- they allow users to overcome issues related to the input data by and to account for factors such as environmental change.
+The value of model-based approaches such as VAST are clear- they allow users to overcome issues related to the input data to account for factors such as environmental change.
 However, the many consequential decisions that must be made in a given model and the lack of clear criteria for making these decisions makes it challenging to make these choices and even more difficult to determine whether the choices are correct.
@@
 This may have to do with the fact that our species spawn during months 9-14, which coincides with the spring sampling weeks 13 \& 14.
-Spawning concentrates much of each population in the spawning grounds towards the center of the domain where more samples occur.
+Spawning in our population models concentrates much of each species in the spawning grounds towards the center of the domain where more samples occur.
 The populations are more spread out during the fall survey in weeks 37 \& 38 and since the water temperature is warmer during these months biomass shifts towards the deeper strata, which tend to be less sampled.
@@
 Spatial factors contributed to many of the scenarios that produced larger error values seen in our estimates.
-Flow from the shallow centrally located strata into the deeper, smaller, and less-sampled eastern strata contributed to VAST failing to accurately modeling the abundance of yellowtail flounder without the addition of covariate information, the impact of which was exacerbated in the increasing temperature scenarios.
-While adding covariates can help inform the model and improve the estimate as seen with the yellowtail flounder simulation results, covariates can hinder the model in certain instances.
+Flow from the shallow centrally located strata into the deeper, smaller, and less-sampled eastern strata contributed to VAST failing to accurately modeling the abundance of Yellowtail without the addition of covariate information, the impact of which was exacerbated in the increasing temperature scenarios.
+While adding covariates can help inform the model and improve the estimate as seen with the Yellowtail simulation results, covariates can hinder the model in certain instances with Cod and Haddock.
 
 As seen in Table 3, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
-Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73\% had a lower absolute error compared to the corresponding estimate without covariates , and \$\approx 96, and \(77/80\approx 96\%\) of VAST models with covariates provided lower absolute error than the corresponding stratified mean estimate.
-There is more to explore with respect to the impact of covariates.
-For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates.
-Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information.
-One could also test the impact of assuming the wrong covariate response function (linear vs polynomial etc).
-As seen in Table 3, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
-Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73\% had a lower absolute error compared to the corresponding estimate without covariates As seen in Table 3, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses.
-Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73\% had a lower absolute error compared to the corresponding estimate without covariates
+We considered a best-case scenario by providing covariate information to VAST in the form of the exact habitat and temperature values that governed movement from our population models.
+This covariate information typically resulted in improved estimates, with 73\% of VAST estimates with covariates providing a lower absolute error compared to the corresponding estimate without covariates, and 96\% of VAST models with covariates resulting in a lower absolute error than the corresponding stratified mean estimate.
+In reality the covariates that influence a given species would be less clear and the collected values would contain sampling noise.
+One could test the impact these differences by assuming the wrong covariate response function (linear vs polynomial etc) or by including each of the covariates individually.
+Including noise in the covariate data would further explore how robust the model-based estimates are to additional uncertainty in the covariate information.
+
 Of the \textasciitilde27\% of VAST models where including covariates did not improve the VAST estimate, most provided a comparable error (for ex, 0.11 vs 0.13).
 The instances where including covariates provided a noticeable change in error took place in scenarios that included either an increasing temperature and/or reduced sampling domain.
-More specifically, when a reduced number of strata are sampled for haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates.
-This is in contrast to our yellowtail flounder results, where adding covariate information to VAST models always decreased the absolute error in the resulting abundance estimates.
-Although strata were excluded for both species, the decline in performance for just Cod and haddock can be explained by a failure to sample the full spectrum of temperature values where the species exists.
-We notice in Figure \ref{fig:strata-plot} that the strata excluded from sampling for yellowtail flounder still allowed the survey to range across the species preferred values and thus provided a covariate response that improved the estimate.
-On the other hand, the specific strata that were excluded from sampling for haddock were located in the northern region of the domain, which means the covariate response for haddock was not informed by any data from the coldest temperature range.
+More specifically, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates.
+This is in contrast to our Yellowtail results, where adding covariate information to VAST models always decreased the absolute error in the resulting abundance estimates.
+Although strata were excluded for both species, the decline in performance for just Cod and Haddock can be explained by a failure to sample the full spectrum of temperature values where the species exists.
+We notice in Figure \ref{fig:strata-plot} that the strata excluded from sampling for Yellowtail still allowed the survey to range across the species preferred values and thus provided a covariate response that improved the estimate.
+On the other hand, the specific strata that were excluded from sampling for Haddock were located in the northern region of the domain, which means the covariate response for Haddock was not informed by any data from the coldest temperature range.
 This could explain the failure to form a complete covariate response that not only degrades the estimate in the regions being sampled, but also confounds the ability to project estimates into an unsampled region.
@@
 
-\caption{Representative example of bioamass estimate trend for yellowtail flounder.
+\caption{Representative example of bioamass estimate trend for Yellowtail.
 Both scenarios have a relatively constant population with all strata being surveyed.
@@
 
-\caption{Representative example of bioamass estimate trend for atlantic cod.
+\caption{Representative example of bioamass estimate trend for Atlantic Cod.
 Both scenarios have a decreasing population trend with all strata being surveyed.
@@
 
-\caption{Representative example of bioamass estimate trend for haddock.
+\caption{Representative example of bioamass estimate trend for Haddock.
 Both scenarios have an increasing population trend and a repeating temperature gradient.
@@
 
-\caption{\label{tab:YTresultsallstrata}Yellowtail flounder error results with all strata included in calculations. Row colors correspond to the same settings applied in different seasons.}
+\caption{\label{tab:YTresultsallstrata}Yellowtail Flounder error results with all strata included in calculations. Row colors correspond to the same settings applied in different seasons.}
 \centering
@@
 
-\caption{\label{tab:YTresultsreducedstrata}Yellowtail flounder error results with certain strata excluded from calculations. Row colors correspond to the same settings applied in different seasons.}
+\caption{\label{tab:YTresultsreducedstrata}Yellowtail Flounder error results with certain strata excluded from calculations. Row colors correspond to the same settings applied in different seasons.}
 \centering

### 2023-08-01 Manuscript/Draft2.Rmd
incorporating liz and tim's edits

--- 2023-06-14 Manuscript/Draft1.Rmd
@@
 Many stock assessments in the United States use abundance estimates derived from stratified random bottom trawl data.
-To accurately represent true abundance, catches of a species must contain a low enough noise level to allow for a discernible pattern and the proportion of the population that is sampled should be consistent over time.
-These assumptions could be violated given enough noise in the sampling process and/or climate change causing a population to increase its abundance in strata that are not sampled.
-Using the R package MixFishSim, we have developed data-driven spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock in the western Atlantic Ocean to allow examination of these assumptions through simulation.
+To accurately track changes in true abundance of a species, catch observations must contain low noise and the proportion of the population that is sampled should be consistent over time.
+One increasingly documented phenomenon that violates the latter assumption is climate-induces changes in relative abundance between sampled and unsampled portions of the stock area.
+Using the R package MixFishSim, we have developed data-driven spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock in the western Atlantic Ocean to evaluate two methods of estimating population trend given temperature driven movement.
 Movement rates combine species-specific static habitat preferences with temperature tolerances.
@@
 A repeating yearly temperature pattern produces repeating spatial biomass distributions in a given week, while a temperature gradient that increases on average over time results in spatial preferences that evolve throughout a given simulation.
-We created simulated spatial time series datasets for each species for several temperature scenarios and population trends.
+We simulated spatial time series datasets for each species for several temperature scenarios and population trends.
 Using stratified random sampling on model output we were able to compare abundance estimates derived from the design-based stratified mean to those using a spatio-temporal model-based approach that allows inclusion of environmental covariates (VAST).
-Our focus was on the ability of contemporary indexing methods to track population trends under shifting spatial preferences.
+Our focus was on the ability of contemporary index standardization methods to track population trends under shifting spatial preferences.
 
@@
 
-survey, design-based, model-based, VAST, population simulation, climate change
+survey, design-based, model-based, climate change, population simulation, VAST
 
@@
 In 2019, New England fisheries produced $22 billion in sales, which supported over 200,000 jobs [@Econ].
-Maintaining a healthy ecosystem is therefore vital to sustained ecosystem health and economic prosperity of the region.
+Maintaining a healthy ecosystem is therefore vital to the economic prosperity of the region.
 
@@
 The Northeast Fisheries Science Center (NEFSC) has conducted a bottom trawl survey since 1963 to support assessment and management of the fish and invertebrate populations in the region [@azarovitz1981brief; @politis2014northeast].
-The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf.
+The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf, where depth and latitude define the strata.
 The survey has created a rich time series data set with many uses including species-specific habitat identification, age and life history information, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments, and ultimately fishing quota limits.
@@
 Most spatial analyses and  projections of future distributions typically assume a constant survey catchability and availability over time.
-For this reason, NOAA's survey design includes sampling each strata in approximately the same 3-4 week time period in each season.
+For this reason, NOAA's survey design includes sampling each stratum in approximately the same 6 week time period in each season.
 
@@
 Model-based estimators help account for complex relationships between variables and can help overcome problems related to the sampling process.
-Common model-based approaches include General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM).
+Common model-based approaches include Generalized Linear Models (GLM), Generalized Additive Model (GAM), and General Linear Mixed Models (GLMM).
 
@@
 Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades.
-The changes have resulted in a shift in spatial distributions of many species [@nye2009changing; @kleisner2017marine; @henderson2017effects].
+At the same time a number of recent studies have found a shift in spatial distributions of many species [@nye2009changing; @kleisner2017marine; @henderson2017effects].
 Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments.
 More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey could change, altering the relationship between the index and the true population [@arreguin1996catchability; @langan2021climate].
-A species shifting its range beyond the survey area is an additional compounding factor to consider.
+A species shifting its range beyond the survey area or outside of the assumed stock boundary are additional compounding factor to consider.
 Existing research has focused on temperature as the driver of such changes [@klein2017effects] and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges [@kerr2022coherence].
 In these scenarios, management strategy evaluations have shown that misconceptions of stock status can lead to unintended overfishing, which can ultimately have detrimental ecologic and economic impacts [@mazur2023consequences].
-We are therefore interested in analyzing the impact of climate change on the accuracy of abundance estimates derived from NOAA’s ongoing bottom-trawl survey along the east coast.
+We are therefore interested in analyzing the impact of climate driven movement and distribution changes on the accuracy of abundance estimates derived from NOAA’s ongoing bottom-trawl survey along the east coast.
 
@@
 
-To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depend on temperature preferences.
+To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depends on temperature preferences.
 We consider the impact of climate change by comparing simulations that use a repeating water temperature pattern to those where temperature increases on average over time.
@@
 We compare yearly abundance estimates obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
-The stratified mean is a design-based approach that calculates the arithmetic mean catch per tow and has traditionally been used with stratified random sample designs.
+The stratified mean is a design-based approach that calculates the stratum specific mean catch per tow, which are then weighted by stratum area to obtain an overall mean.
 VAST is a spatial delta-generalized linear mixed model that estimates both abundance (biomass) and probability of occurrence (presence/absence) [@thorson2019guidance].
-If desired, VAST also allows users to include covariate data to better inform the model.
-Covariates can be static (eg.
-habitat preferences) or dynamics (eg.
+VAST also allows users to include covariate data to better inform the model.
+Covariates can be static (e.g.
+habitat preferences) or dynamics (e.g.
 temperature).
-We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses.
+We explore whether including environmental predictors leads to improved abundance estimates, which is particularly relevant as climate change progresses.
 
@@
 We construct spatial population models for Yellowtail Flounder (*Limanda ferruginea*), Atlantic Cod (*Gadus morhua*), and Haddock (*Melanogrammus aeglefinus*) on Georges Bank, hereafter referred to as Yellowtail, Cod, and Haddock.
-Movement of each species combine static species-specific habitat preferences with biologically-based temperature preferences.
+Movement of each species combines static species-specific habitat preferences with biologically-based temperature preferences.
 Model dynamics are driven by a time series of temperature gradients to create simulated spatial data sets for each population where the true biomass is known.
 Using temperature gradients that repeat each year creates data sets with repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time.
-We conduct stratified random sampling on our simulation output to mimic the bottom trawl survey and use the samples to compare the ability of contemporary indexing methods to track population trends.
+We conduct stratified random sampling on our simulation data to mimic the bottom trawl survey and use the samples to compare the ability of contemporary index standardization methods to track population trends.
 
@@
 The package uses a delay-difference population model with discrete processes for growth, death, and recruitment of the population.
-To address our research question we formulate the following inputs for the MFS package.
+To address our research question we specify the following inputs for the MFS package.
 
@@
 
-We obtained a shapefile for the 15 strata that comprise Georges Bank, where strata are partitioned based primarily on depth and secondarily by latitude [@politis2014northeast].
-The region was discritized into a raster with 88 rows and 144 columns to use as our modeling environment.
+A shapefile for the 15 strata that comprise Georges Bank was discretized into a raster with 88 rows and 144 columns to use as our modeling environment.
 Each cell in our simulation domain represents approximately 8.7 km$^2$.
@@
 As a result, each stock inhabits a different number of strata on Georges Bank.
-Haddock inhabit all 15 strata in the domain, Cod populate 13 strata, and Yellowtail can be found in 9 strata.
-Figure \@ref(fig:strata-plot) shows the strata used in our model for each species.
+Haddock inhabit all 15 strata in the domain, Cod populate 13 strata, and Yellowtail can be found in 9 strata (Figure \@ref(fig:strata-plot)).
 
@@
 MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass in each cell in our study area [@dolder2020highly].
-Individual terms in the formulation account for growth of mature adults, mortality (natural and fishing), and the addition of new recruits.
+Parameters in the population model account for growth of mature adults, mortality (natural and fishing), and the addition of new recruits.
 Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.
-Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics.
+Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics, and therefore, these results cannot be compared directly with stock assessment results.
 A full list of parameters used in our model can be seen in Tables 1 and \@ref(tab:paramsSCENARIOS).
@@
 
-The package was designed to generate hypothetical temperature gradients and theoretical habitat preferences using Gaussian Random Fields.
-The following sections describe how we formulated the habitat and temperature components to model our three real species on Georges Bank.
+The MFS package was designed to generate hypothetical temperature gradients and theoretical habitat preferences using Gaussian Random Fields.
+The following sections describe how we configured the habitat and temperature components to model our three species on Georges Bank.
 
@@
 
-Species-specific habitat preferences were derived from niche model for each species using the *lrren* tool from the R package *envi* [@envi].
+The habitat input for each species in MFS should represent spatial habitat preferences for each species that are time-invariant and are instead related to static geographic features.
+Species-specific habitat preferences were derived from niche models for each species using the *lrren* tool from the R package *envi* [@envi].
 The *lrren* tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors.
@@
 
-```{r hab-plot1, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Visual representation of niche model for Atlantic Cod."}
+```{r hab-plot1, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Visual representation of niche model for Atlantic Cod.
+In the left most plot, tows with positive Cod catch are shown in red while tows without Cod are in blue.
+Positive tows are weighted by the biomass of the given catch.
+The middle two plots show our covariate predictors used in the *lrren* niche model- depth and median sediment size.
+The plot on the right shows the final niche model that has been rescaled between 0 and 1, where blue values are closer to 0 and red values are closer to 1."}
 knitr::include_graphics(c("Images/hab_snip3.png"))
@@
 Historically, Cod has seen significant decline over the last 50**??** years while Haddock has increased in abundance in recent years [[can cite the 2022 management track assessments, see https://apps-nefsc.fisheries.noaa.gov/saw/sari.php for when the document becomes available]] **just cite 2021 assessments?**.
-For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
+For this reason we compare survey index estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
 We consider increasing, decreasing, and constant population trends for Yellowtail for completeness.
@@
 
-```{r pop-scenarios, echo=FALSE,out.width="99%",fig.show='hold', fig.align='center', fig.cap="True population trends used in indexing analyses.
+```{r pop-scenarios, echo=FALSE,out.width="99%",fig.show='hold', fig.align='center', fig.cap="True population trends used in index analyses.
 Spring biomass plots are shown with fall values being very similar."}
@@
 
-<font size="5">*Simulating Bottom Trawl Survey and Population Indexing*</font>
+<font size="5">*Simulating Bottom Trawl Survey and Population Index*</font>
 
@@
 
-After completing all simulations, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year.
-We sample each strata in the same weeks in which the spring and fall surveys take place (weeks 13 & 14 in the spring and 37 & 38 in the fall).
-The number of the samples taken reflect true target values for each strata and sampling cells were randomly selected.
-
-An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists.
+After distributing each species in its stock area with simulated population trajectories, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited stratum twice each year.
+We sample each stratum in the same weeks in which the spring and fall surveys have historically sampled Georges Bank (weeks 13 & 14 in the spring and 37 & 38 in the fall).
+The number of simulated tows made reflects true target values for each stratum and sample cells where "tows" are made were randomly selected.
+
+An underlying assumption in all index methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) consistently sampling the same proportion of the population.
 These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata.
-To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values.
+To simulate the impact of noise, we compare index estimates after adding noise to our samples versus those using the true sampling values.
 Annual survey observations were simulated as log-normal deviations from the underlying “true” survey catches with a CV of 0.3.
-The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population.
-We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species.
+The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extent of the population.
+We simulate the effect of populations moving into new habitat by comparing index estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species.
 The strata to exclude for each species were chosen by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from.
 Figure \@ref(fig:strata-plot) shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \@ref(fig:PopPct).
-The yellow regions in Figure \@ref(fig:strata-plot) depict the strata that were removed from certain stratified random surveys for each species.
-We then use the biomass collected from our samples in contemporary abundance indexing methods to estimate population trends.
-Knowing the true population values in our simulations allows us to compute and compare the absolute error calculated from each estimation method.
+The yellow regions in Figure \@ref(fig:strata-plot) depict the strata that were not sampled for each species.
+We then use the biomass collected from our samples in contemporary abundance index methods to estimate yearly seasonal population trends.
 
 The design-based method we used is the stratified mean, which divides the inhabited domain into $N$ disjoint strata based on relevant geographic and environmental information such as depth and latitude/longitude.
-The number of samples taken from each stratum $S_j$ for $j=1...N$ is relative to the given area.
-The stratified mean biomass $SM_{s,y}$ for a given season $s$ and year $y$ can then be calculated by through the weighted average
-
-\[
-  \makebox[\linewidth]{$SM_{s,y} = \sum_{i=1}^N W_i \frac{\sum_{k=1}^{S_i}y_{k,i}}{S_i},$}
-\]
+The number of samples taken from each stratum $S_j$ for $j=1...N$ is proportional to the given area.
+The stratified mean biomass $SM_{s,y}$ for a given season $s$ and year $y$ can then be calculated as the weighted average
+
+\begin{equation*}
+ SM_{s,y} = \sum_{i=1}^N W_i \frac{\sum_{k=1}^{S_i}y_{k,i}}{S_i},
+\end{equation*}
 
@@
 We compare stratified mean estimates to those derived from the model-based VAST approach.
-VAST models both biomass $p_1(i)$ and presence/absence $p_2(i)$ for each observation $i$ as linear predictors using a spatial delta-generalized linear mixed model that can be represented by
+VAST models both encounter probability $p_1(i)$ and positive catch rate $p_2(i)$ for each observation $i$ as linear predictors using a spatial delta-generalized linear mixed model that can be represented by
 
@@
 \[
-  \makebox[\linewidth]{$p_*(i)= \beta_*(t_i) + L_{\omega_*}\omega_*(s_i) + L_* \epsilon_*(c_i,t_i) + \upsilon(s_i,t_i) + \xi(i),$}
+  \makebox[\linewidth]{$f(p_*(i))= \beta_*(t_i) + L_{\omega_*}\omega_*(s_i) + L_* \epsilon_*(c_i,t_i) + \upsilon(s_i,t_i) + \xi(i),$}
 \]
 
-for $*=1,2$, where more specific functional forms of each component are further described in [@thorson2019guidance].
+for $*=1,2$, where more specific functional forms of each component are further described in @thorson2019guidance.
 VAST models require numerous user inputs to determine how the linear predictors will be conditioned and solved.
@@
 
-Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings.
-Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models as shown in Table 3.
+Knowing the true population values in our simulations allows us to compute the absolute error in a given season $E_s$ from each estimation method and scenario as given by
+
+\begin{equation}
+E_s = \frac{\sum_{y=1}^{20}|M_{i,s}-I_{i,s}|}{\sum_{y=1}^{20}M_{i,s}},
+\end{equation}
+
+where $M_{i,s}$ is the model biomass in year $i$ and season $s$ and $I_{i,s}$ is the biomass estimate provided by either the stratified mean or VAST.
+
+Through this process, and in consultation with the VAST package creator, we compared the performance of two model setting configurations in our VAST models as shown in Table 3.
 VAST Settings B were chosen after reviewing the information provided in the "Seasonal Model" section of VAST's online Github tutorial [@ThorsonSeasonal].
-After running models for each of our scenarios using these settings, we shared some of our results with the VAST package creator, James (Jim) Thorson.
-Jim reviewed our VAST model results and suggested we also consider the options shown in Settings A.
+After running models for each of our scenarios using these settings, we received feedback to also consider configurations for Settings A [@ThorsonChat].
 Both settings use a Tweedie model and turn off the first linear predictor $p_1$ so that only biomass is being modeled.
@@
 The difference amounts to intercepts for the biomass predictor being a fixed effect (RhoBeta2=0) versus intercepts being constant among years as a fixed effect (RhoBeta2=3).
-According to Jim, the Settings A values are representative of what stock assessment scientists typically use in their analyses.
-As a result we will focus our analysis on VAST results obtained using Settings A, but present Settings B results to demonstrate how the settings used in model-based approaches impact model performance.
-
-Each indexing scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, & covariates), resulting in a large number of scenario combinations to consider.
+The specifications for Settings A are more  typical for stock assessment application [@ThorsonChat], and we therefore focus our discussion on these VAST results.
+Results for Settings B are shown to demonstrate how the settings used in model-based approaches impact model performance.
+
+Each abundance index scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, & covariates), resulting in a large number of scenario combinations to consider.
 The columns in Table \@ref(tab:scenarios) show the choices that define each scenario.
@@
 
-<!-- The goal of our project was to analyze the ability of contemporary population indexing methods to track population trends under the variety of conditions shown in Table \@ref(tab:scenarios).
+<!-- The goal of our project was to analyze the ability of contemporary population index methods to track population trends under the variety of conditions shown in Table \@ref(tab:scenarios).
 and Figure \@ref(fig:hab-plot) for the static habitat preferences for each species ($Hab^2_{J,s}$) used in our population models.
@@
 
-Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the absolute error between biomass estimates and model output for each of our abundance estimates.
+Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the absolute error between biomass estimates and model output for each of our 20-year abundance estimates.
 While the model-based results provided a slightly lower overall mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02).
-When considering individual scenarios, while about $96\%$ had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, $63\%$ of individual scenarios contained a VAST estimate with a worse error than the corresponding stratified mean estimate.
-VAST models that included covariate information provided the lowest overall errors and standard deviations.
+When considering individual scenarios, while about $96\%$ had a VAST estimate with a lower absolute error than the corresponding stratified mean estimate, $63\%$ of individual scenarios contained a VAST estimate with a worse error than the corresponding stratified mean estimate.
+VAST models that included covariate information provided the lowest overall absolute errors and standard deviations.
 Models with covariates had an average absolute error of 0.22, compared to models that did not include covariate information having an average error of 0.46, and stratified mean estimates produces an average error of 0.39.
@@
 
-When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.
+When we reduce the number of strata that are included in index calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.
 Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain (e.g.
-in Table \@ref(tab:HadERROR) rows 8 and 9, VAST models without covariates show lower errors than when covariates are included).
+in Table \@ref(tab:HadERROR) rows 8 and 9, VAST models without covariates show lower absolute errors than when covariates are included).
 We analyze these results further in the sections that follow.
@@
 This can be further examined through yearly estimate:model ratio values, where we divide the yearly abundance estimates by the true model value.
-In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can result in yearly abundance values as low as 0 at times and exhibit large yearly changes.
-A representative example of this trend for each species is shown on the log scale in Figure \@ref(fig:RatioTrend).
-
-Analysis of individual yearly model:estimate ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21.
+In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can result in yearly abundance values as low as 0 at times and exhibit large yearly changes (Figure \@ref(fig:RatioTrend)).
+
+Analysis of individual yearly estimate:model ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21.
 On the other hand, just 33\% of stratified mean estimate ratios were above 1 with an average of 0.874 and a standard deviation of 0.12.
@@
 
-When the entire domain is sampled, the stratified mean produced an average yearly ratio of 0.87 with a standard deviation of 0.12.
+When the entire domain is sampled, the stratified mean produced an average yearly estimate:model ratio of 0.87 with a standard deviation of 0.12.
 Adding covariates in these scenarios brings the VAST estimate ratio closer to 1.
@@
 The percent of the population increases in stratum 16 in the spring over the duration of both time series, which implies the flow out of strata 13 and 19 in the spring are going into stratum 16.
-These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12.
-These spring changes are therefore related to the static habitat values in our model rather than the temperature preferences, which is why we seen the same dynamics with constant and increasing temperature.
+These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail that coincides with most of the areas we have designated as the species' spawning ground and spawning takes place immediately before spring sampling in weeks 9-12.
+These spring changes are therefore related to the static habitat values in our model rather than the temperature preferences, which is why we see the same dynamics with constant and increasing temperature.
 While we observe similar changes in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in the population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \@ref(fig:PopPct)).
-These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow outer strata 17 and 18.
+These dynamics imply that an increase in temperature results in the more shallow stratum 16 becoming less desirable than the deeper and more narrow outer strata 17 and 18.
 One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how ~10\% of the population exists in the narrow outer strata 17 in the fall, while seemingly none of the population exists in any of the strata near the edge of the domain (14, 15, 17, 18) in the spring.
@@
 
-Of all scenarios without covariates, $33/48\approx68$\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
+Of all scenarios without covariates, $33/48\approx68$\% had a VAST fit with a lower absolute error compared to the stratified mean estimate.
 All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the fall season and had a common theme of producing abundance estimates that are above the true model value.
@@
 Including covariate information made a noticeable difference in our Yellowtail model-based VAST estimates.
-All of the VAST estimates that included covariate information produced a lower relative error than the corresponding VAST model that did not include covariates.
+All of the VAST estimates that included covariate information produced a lower absolute error than the corresponding VAST model that did not include covariates.
 The largest improvements were seen in the increasing temperature scenarios.
-When comparing to the design-based estimates, $47/48\approx98$\% of the VAST estimates that included covariates information had a lower relative error than the corresponding stratified mean estimate.
+When comparing to the design-based estimates, $47/48\approx98$\% of the VAST estimates that included covariates information had a lower absolute error than the corresponding stratified mean estimate.
 This implies that the covariate information helped our model-based estimate account for the design-based issues related to an increasing percentage of the population entering smaller strata, which can become exacerbated in the increasing temperature simulations.
@@
 The different VAST settings produced similar error values with $18/32\approx57\%$ of estimates performing better with VAST Settings A.
-Of the abundance estimates without covariates, $12/16=75$\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
-VAST produced higher relative error compared to the stratified mean in scenarios that involved increasing temperature in the fall season without covariates.
+Of the abundance estimates without covariates, $12/16=75$\% had a VAST fit with a lower absolute error compared to the stratified mean estimate.
+VAST produced higher absolute error compared to the stratified mean in scenarios that involved increasing temperature in the fall season without covariates.
 Similar to the Yellowtail results, this implies the model-based approach had seasonal trouble with populations shifting into smaller strata where fewer samples take place.
@@
 
-Figure \@ref(fig:PopPct) reveals some subtle seasonal differences in the percent of Haddock in each strata.
+Figure \@ref(fig:PopPct) reveals some subtle seasonal differences in the percent of Haddock in each stratum.
 In the constant temperature scenarios, the spring shows a decrease in strata 19 and 20 that correspond to increases in 16, 24 and 29.
@@
 This shift represents movement from the shallower and more centrally located strata towards deeper strata located near the edge of the domain.
-Since stratum 16 contains very favorable habitat including much of the species' spawning ground, the strong shift out of 16 and into the northern most strata of 29 in the fall demonstrates a climate-driven change in movement preference.
+Since stratum 16 contains very favorable habitat including much of the species' spawning ground, the strong shift out of 16 and into the northern most stratum 29 in the fall demonstrates a climate-driven change in movement preference.
 
@@
 
-Of the scenarios without covariates, $10/16\approx63$\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
+Of the scenarios without covariates, $10/16\approx63$\% had a VAST fit with a lower absolute error compared to the stratified mean estimate.
 The 6 covariate-free VAST estimates that resulted in a higher error than the stratified mean spanned all scenarios and seasons, with several fall errors being especially large (significant overestimates).
 Adding covariate information resulted in $14/16\approx88$\% of VAST estimates having improved error compared to the stratified mean.
-The 2 scenarios that produced worse error with covarites spanned temperature scenarios, but were both in the spring season when the proportion of the population in each strata remained constant in each scenario.
+The 2 scenarios that produced worse error with covarites spanned temperature scenarios, but were both in the spring season when the proportion of the population in each stratum remained constant in each scenario.
 The average change in the error of abundance estimates between constant temperature scenarios and increasing temperature scenarios were 1.42 for VAST estimates without covariates, 1.80 for VAST estimates that included covariates, and 1.37 for stratified mean estimates.
@@
 Our analysis showed that both the design-based and model-based approaches are each capable of providing biomass estimates that track the true abundance trend.
-While VAST estimates tended to provide lower relative error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B also combined to produce over 30 abundance estimates with larger error values than the largest stratified mean error.
+While VAST estimates tended to provide lower absolute error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B also combined to produce over 30 abundance estimates with larger error values than the largest stratified mean error.
 These facts demonstrate the power of model-based estimates while also highlighting how sensitive they are to the choices made regarding settings and data used in analyses.
-Users often rely on traditional model-building processes and diagnostic techniques such as AIC to guide their choices, but successfully navigating the many decisions depends on the background of the given user and care should be taken to which diagnostic tools are used.
+Users often rely on traditional model-building processes and diagnostic techniques such as AIC to guide their choices, but successfully navigating the many decisions depends on the background of the given user and care should be taken in determining which diagnostic tools are used.
 For example, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using settings that decrease the accuracy of the model **cite Chris C paper**.
@@
 Spatial factors contributed to many of the scenarios that produced larger error values seen in our estimates.
-Flow from the shallow centrally located strata into the deeper, smaller, and less-sampled eastern strata contributed to VAST failing to accurately modeling the abundance of Yellowtail without the addition of covariate information, the impact of which was exacerbated in the increasing temperature scenarios.
+Flow from the shallow centrally located strata into the deeper, smaller, and less-sampled eastern strata contributed to VAST failing to accurately model the abundance of Yellowtail without the addition of covariate information, the impact of which was exacerbated in the increasing temperature scenarios.
 While adding covariates can help inform the model and improve the estimate as seen with the Yellowtail simulation results, covariates can hinder the model in certain instances with Cod and Haddock.
@@
 In reality the covariates that influence a given species would be less clear and the collected values would contain sampling noise.
-One could test the impact these differences by assuming the wrong covariate response function (linear vs polynomial etc) or by including each of the covariates individually.
+One could test the impact of these differences by assuming the wrong covariate response function (linear vs polynomial etc) or by including each of the covariates individually.
 Including noise in the covariate data would further explore how robust the model-based estimates are to additional uncertainty in the covariate information.
@@
 Thank you to Jim Thorson for consulting with us about VAST settings and implementation.
+We are thanksful for Chris Cacciapaglia and Charles Adams for the time they spent with us sharing their experiences using VAST.
 We appreciate the help from Robyn Linner for helping to extract FVCOM temperature data and are thankful for David Chevrier for meeting with us to clarify questions related to some of our GIS data.
@@
 
-```{r PopPct, echo=FALSE,fig.show='hold', fig.align='center', fig.cap="Percent of each species in each strata for during survey weeks in our spatial simulations.
+```{r PopPct, echo=FALSE,fig.show='hold', fig.align='center', fig.cap="Percent of each species in each stratum for during survey weeks in our spatial simulations.
 All constant temperature scenarios follow the patterns on the left while increasing temperature scenarios follow the patterns on the right.
@@
 
-```{r YTEstEx, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Representative example of bioamass estimate trend for Yellowtail.
+```{r YTEstEx, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Representative example of biomass estimate trend for Yellowtail.
 Both scenarios have a relatively constant population with all strata being surveyed.
@@
 
-```{r CodEstEx, echo=FALSE,out.width="100%",fig.show='hold', fig.align='center', fig.cap="Representative example of bioamass estimate trend for Atlantic Cod.
+```{r CodEstEx, echo=FALSE,out.width="100%",fig.show='hold', fig.align='center', fig.cap="Representative example of biomass estimate trend for Atlantic Cod.
 Both scenarios have a decreasing population trend with all strata being surveyed.
@@
 
-```{r HadEstEx, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Representative example of bioamass estimate trend for Haddock.
+```{r HadEstEx, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Representative example of biomass estimate trend for Haddock.
 Both scenarios have an increasing population trend and a repeating temperature gradient.
@@
       align = c("l", "l", "l", "l", "c", "c", "c"),
-  caption = "Yellowtail Flounder error results with all strata included in calculations. Row colors correspond to the same settings applied in different seasons.") %>%
+  caption = "Yellowtail Flounder absolute error results with all strata included in calculations. Row colors correspond to the same settings applied in different seasons.") %>%
   row_spec(0,bold = TRUE)%>% #bold column titles
@@
       align = c("l", "l", "l", "l", "c", "c", "c"),
-  caption = "Yellowtail Flounder error results with certain strata excluded from calculations. Row colors correspond to the same settings applied in different seasons.") %>%
+  caption = "Yellowtail Flounder absoloute error results with certain strata excluded from calculations. Row colors correspond to the same settings applied in different seasons.") %>%
   row_spec(0,bold = TRUE)%>% #bold column titles
@@
 Fig_3a <-  kable(summary_tab1[, -1], 
-  caption = "Cod error results.",
+  caption = "Cod absolute error results.",
     format = "latex", booktabs = TRUE) %>%
@@
 Fig_3b <-  kable(summary_tab1[, -1], 
-  caption = "Haddock error results.",
+  caption = "Haddock absolute error results.",
     format = "latex", booktabs = TRUE) %>%

### 2023-08-01 Manuscript/Draft2.tex
incorporating liz and tim's edits

--- 2023-06-14 Manuscript/Draft1.tex
@@
 In 2019, New England fisheries produced \$22 billion in sales, which supported over 200,000 jobs ({``{Fisheries Economics of the United States: Data and Visualizations}''} n.d.).
-Maintaining a healthy ecosystem is therefore vital to sustained ecosystem health and economic prosperity of the region.
+Maintaining a healthy ecosystem is therefore vital to the economic prosperity of the region.
 
@@
 2014).
-The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf.
+The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf, where depth and latitude define the strata.
 The survey has created a rich time series data set with many uses including species-specific habitat identification, age and life history information, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments, and ultimately fishing quota limits.
@@
 Most spatial analyses and projections of future distributions typically assume a constant survey catchability and availability over time.
-For this reason, NOAA's survey design includes sampling each strata in approximately the same 3-4 week time period in each season.
+For this reason, NOAA's survey design includes sampling each stratum in approximately the same 6 week time period in each season.
 
@@
 Model-based estimators help account for complex relationships between variables and can help overcome problems related to the sampling process.
-Common model-based approaches include General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM).
+Common model-based approaches include Generalized Linear Models (GLM), Generalized Additive Model (GAM), and General Linear Mixed Models (GLMM).
 
 Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades.
-The changes have resulted in a shift in spatial distributions of many species (Nye et al.
+At the same time a number of recent studies have found a shift in spatial distributions of many species (Nye et al.
 2009; Henderson et al.
@@
 2021).
-A species shifting its range beyond the survey area is an additional compounding factor to consider.
+A species shifting its range beyond the survey area or outside of the assumed stock boundary are additional compounding factor to consider.
 Existing research has focused on temperature as the driver of such changes (Klein et al.
@@
 2023).
-We are therefore interested in analyzing the impact of climate change on the accuracy of abundance estimates derived from NOAA's ongoing bottom-trawl survey along the east coast.
-
-To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depend on temperature preferences.
+We are therefore interested in analyzing the impact of climate driven movement and distribution changes on the accuracy of abundance estimates derived from NOAA's ongoing bottom-trawl survey along the east coast.
+
+To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depends on temperature preferences.
 We consider the impact of climate change by comparing simulations that use a repeating water temperature pattern to those where temperature increases on average over time.
@@
 We compare yearly abundance estimates obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
-The stratified mean is a design-based approach that calculates the arithmetic mean catch per tow and has traditionally been used with stratified random sample designs.
+The stratified mean is a design-based approach that calculates the stratum specific mean catch per tow, which are then weighted by stratum area to obtain an overall mean.
 VAST is a spatial delta-generalized linear mixed model that estimates both abundance (biomass) and probability of occurrence (presence/absence) (Thorson 2019).
-If desired, VAST also allows users to include covariate data to better inform the model.
-Covariates can be static (eg.
-habitat preferences) or dynamics (eg.
-temperature).
-We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses.
+VAST also allows users to include covariate data to better inform the model.
+Covariates can be static (e.g.~habitat preferences) or dynamics (e.g.~temperature).
+We explore whether including environmental predictors leads to improved abundance estimates, which is particularly relevant as climate change progresses.
 
@@
 We construct spatial population models for Yellowtail Flounder (\emph{Limanda ferruginea}), Atlantic Cod (\emph{Gadus morhua}), and Haddock (\emph{Melanogrammus aeglefinus}) on Georges Bank, hereafter referred to as Yellowtail, Cod, and Haddock.
-Movement of each species combine static species-specific habitat preferences with biologically-based temperature preferences.
+Movement of each species combines static species-specific habitat preferences with biologically-based temperature preferences.
 Model dynamics are driven by a time series of temperature gradients to create simulated spatial data sets for each population where the true biomass is known.
 Using temperature gradients that repeat each year creates data sets with repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time.
-We conduct stratified random sampling on our simulation output to mimic the bottom trawl survey and use the samples to compare the ability of contemporary indexing methods to track population trends.
+We conduct stratified random sampling on our simulation data to mimic the bottom trawl survey and use the samples to compare the ability of contemporary index standardization methods to track population trends.
 
@@
 
-We obtained a shapefile for the 15 strata that comprise Georges Bank, where strata are partitioned based primarily on depth and secondarily by latitude (Politis et al.
-2014).
-The region was discritized into a raster with 88 rows and 144 columns to use as our modeling environment.
+A shapefile for the 15 strata that comprise Georges Bank was discretized into a raster with 88 rows and 144 columns to use as our modeling environment.
 Each cell in our simulation domain represents approximately 8.7 km\(^2\).
@@
 As a result, each stock inhabits a different number of strata on Georges Bank.
-Haddock inhabit all 15 strata in the domain, Cod populate 13 strata, and Yellowtail can be found in 9 strata.
-Figure \ref{fig:strata-plot} shows the strata used in our model for each species.
+Haddock inhabit all 15 strata in the domain, Cod populate 13 strata, and Yellowtail can be found in 9 strata (Figure \ref{fig:strata-plot}).
 
@@
 2020).
-Individual terms in the formulation account for growth of mature adults, mortality (natural and fishing), and the addition of new recruits.
+Parameters in the population model account for growth of mature adults, mortality (natural and fishing), and the addition of new recruits.
 Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period.
-Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics.
+Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics, and therefore, these results cannot be compared directly with stock assessment results.
 A full list of parameters used in our model can be seen in Tables 1 and \ref{tab:paramsSCENARIOS}.
@@
 
-Species-specific habitat preferences were derived from niche model for each species using the \emph{lrren} tool from the R package \emph{envi} (Buller 2022).
+The habitat input for each species in MFS should represent spatial habitat preferences for each species that are time-invariant and are instead related to static geographic features.
+Species-specific habitat preferences were derived from niche models for each species using the \emph{lrren} tool from the R package \emph{envi} (Buller 2022).
 The \emph{lrren} tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors.
@@
 
-\caption{Visual representation of niche model for Atlantic Cod.}\label{fig:hab-plot1}
+\caption{Visual representation of niche model for Atlantic Cod.
+In the left most plot, tows with positive Cod catch are shown in red while tows without Cod are in blue.
+Positive tows are weighted by the biomass of the given catch.
+The middle two plots show our covariate predictors used in the *lrren* niche model- depth and median sediment size.
+The plot on the right shows the final niche model that has been rescaled between 0 and 1, where blue values are closer to 0 and red values are closer to 1.}\label{fig:hab-plot1}
 \end{figure}
@@
 Historically, Cod has seen significant decline over the last 50\textbf{??} years while Haddock has increased in abundance in recent years {[}{[}can cite the 2022 management track assessments, see \url{https://apps-nefsc.fisheries.noaa.gov/saw/sari.php} for when the document becomes available{]}{]} \textbf{just cite 2021 assessments?}.
-For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
+For this reason we compare survey index estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
 We consider increasing, decreasing, and constant population trends for Yellowtail for completeness.
@@
 
-\caption{True population trends used in indexing analyses. Spring biomass plots are shown with fall values being very similar.}\label{fig:pop-scenarios}
+\caption{True population trends used in index analyses. Spring biomass plots are shown with fall values being very similar.}\label{fig:pop-scenarios}
 \end{figure}
 
-\emph{Simulating Bottom Trawl Survey and Population Indexing}
-
-After completing all simulations, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year.
-We sample each strata in the same weeks in which the spring and fall surveys take place (weeks 13 \& 14 in the spring and 37 \& 38 in the fall).
-The number of the samples taken reflect true target values for each strata and sampling cells were randomly selected.
-
-An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists.
+\emph{Simulating Bottom Trawl Survey and Population Index}
+
+After distributing each species in its stock area with simulated population trajectories, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited stratum twice each year.
+We sample each stratum in the same weeks in which the spring and fall surveys have historically sampled Georges Bank (weeks 13 \& 14 in the spring and 37 \& 38 in the fall).
+The number of simulated tows made reflects true target values for each stratum and sample cells where ``tows'' are made were randomly selected.
+
+An underlying assumption in all index methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) consistently sampling the same proportion of the population.
 These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata.
-To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values.
+To simulate the impact of noise, we compare index estimates after adding noise to our samples versus those using the true sampling values.
 Annual survey observations were simulated as log-normal deviations from the underlying ``true'' survey catches with a CV of 0.3.
-The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population.
-We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species.
+The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extent of the population.
+We simulate the effect of populations moving into new habitat by comparing index estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species.
 The strata to exclude for each species were chosen by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from.
 Figure \ref{fig:strata-plot} shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \ref{fig:PopPct}.
-The yellow regions in Figure \ref{fig:strata-plot} depict the strata that were removed from certain stratified random surveys for each species.
-We then use the biomass collected from our samples in contemporary abundance indexing methods to estimate population trends.
-Knowing the true population values in our simulations allows us to compute and compare the absolute error calculated from each estimation method.
+The yellow regions in Figure \ref{fig:strata-plot} depict the strata that were not sampled for each species.
+We then use the biomass collected from our samples in contemporary abundance index methods to estimate yearly seasonal population trends.
 
 The design-based method we used is the stratified mean, which divides the inhabited domain into \(N\) disjoint strata based on relevant geographic and environmental information such as depth and latitude/longitude.
-The number of samples taken from each stratum \(S_j\) for \(j=1...N\) is relative to the given area.
-The stratified mean biomass \(SM_{s,y}\) for a given season \(s\) and year \(y\) can then be calculated by through the weighted average
-
-\[
-  \makebox[\linewidth]{$SM_{s,y} = \sum_{i=1}^N W_i \frac{\sum_{k=1}^{S_i}y_{k,i}}{S_i},$}
-\]
+The number of samples taken from each stratum \(S_j\) for \(j=1...N\) is proportional to the given area.
+The stratified mean biomass \(SM_{s,y}\) for a given season \(s\) and year \(y\) can then be calculated as the weighted average
+
+\begin{equation*}
+ SM_{s,y} = \sum_{i=1}^N W_i \frac{\sum_{k=1}^{S_i}y_{k,i}}{S_i},
+\end{equation*}
 
@@
 We compare stratified mean estimates to those derived from the model-based VAST approach.
-VAST models both biomass \(p_1(i)\) and presence/absence \(p_2(i)\) for each observation \(i\) as linear predictors using a spatial delta-generalized linear mixed model that can be represented by
+VAST models both encounter probability \(p_1(i)\) and positive catch rate \(p_2(i)\) for each observation \(i\) as linear predictors using a spatial delta-generalized linear mixed model that can be represented by
 
 \[
-  \makebox[\linewidth]{$p_*(i)= \beta_*(t_i) + L_{\omega_*}\omega_*(s_i) + L_* \epsilon_*(c_i,t_i) + \upsilon(s_i,t_i) + \xi(i),$}
+  \makebox[\linewidth]{$f(p_*(i))= \beta_*(t_i) + L_{\omega_*}\omega_*(s_i) + L_* \epsilon_*(c_i,t_i) + \upsilon(s_i,t_i) + \xi(i),$}
 \]
 
-for \(*=1,2\), where more specific functional forms of each component are further described in (Thorson 2019).
+for \(*=1,2\), where more specific functional forms of each component are further described in Thorson (2019).
 VAST models require numerous user inputs to determine how the linear predictors will be conditioned and solved.
@@
 
-Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings.
-Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models as shown in Table 3.
+Knowing the true population values in our simulations allows us to compute the absolute error in a given season \(E_s\) from each estimation method and scenario as given by
+
+\begin{equation}
+E_s = \frac{\sum_{y=1}^{20}|M_{i,s}-I_{i,s}|}{\sum_{y=1}^{20}M_{i,s}},
+\end{equation}
+
+where \(M_{i,s}\) is the model biomass in year \(i\) and season \(s\) and \(I_{i,s}\) is the biomass estimate provided by either the stratified mean or VAST.
+
+Through this process, and in consultation with the VAST package creator, we compared the performance of two model setting configurations in our VAST models as shown in Table 3.
 VAST Settings B were chosen after reviewing the information provided in the ``Seasonal Model'' section of VAST's online Github tutorial (Thorson, James T n.d.b).
-After running models for each of our scenarios using these settings, we shared some of our results with the VAST package creator, James (Jim) Thorson.
-Jim reviewed our VAST model results and suggested we also consider the options shown in Settings A.
+After running models for each of our scenarios using these settings, we received feedback to also consider configurations for Settings A (Thorson, James T n.d.c).
 Both settings use a Tweedie model and turn off the first linear predictor \(p_1\) so that only biomass is being modeled.
@@
 The difference amounts to intercepts for the biomass predictor being a fixed effect (RhoBeta2=0) versus intercepts being constant among years as a fixed effect (RhoBeta2=3).
-According to Jim, the Settings A values are representative of what stock assessment scientists typically use in their analyses.
-As a result we will focus our analysis on VAST results obtained using Settings A, but present Settings B results to demonstrate how the settings used in model-based approaches impact model performance.
-
-Each indexing scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, \& covariates), resulting in a large number of scenario combinations to consider.
+The specifications for Settings A are more typical for stock assessment application (Thorson, James T n.d.c), and we therefore focus our discussion on these VAST results.
+Results for Settings B are shown to demonstrate how the settings used in model-based approaches impact model performance.
+
+Each abundance index scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, \& covariates), resulting in a large number of scenario combinations to consider.
 The columns in Table \ref{tab:scenarios} show the choices that define each scenario.
@@
 
-Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the absolute error between biomass estimates and model output for each of our abundance estimates.
+Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the absolute error between biomass estimates and model output for each of our 20-year abundance estimates.
 While the model-based results provided a slightly lower overall mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02).
-When considering individual scenarios, while about \(96\%\) had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, \(63\%\) of individual scenarios contained a VAST estimate with a worse error than the corresponding stratified mean estimate.
-VAST models that included covariate information provided the lowest overall errors and standard deviations.
+When considering individual scenarios, while about \(96\%\) had a VAST estimate with a lower absolute error than the corresponding stratified mean estimate, \(63\%\) of individual scenarios contained a VAST estimate with a worse error than the corresponding stratified mean estimate.
+VAST models that included covariate information provided the lowest overall absolute errors and standard deviations.
 Models with covariates had an average absolute error of 0.22, compared to models that did not include covariate information having an average error of 0.46, and stratified mean estimates produces an average error of 0.39.
@@
 
-When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.
-Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain (e.g.~in Table \ref{tab:HadERROR} rows 8 and 9, VAST models without covariates show lower errors than when covariates are included).
+When we reduce the number of strata that are included in index calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.
+Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain (e.g.~in Table \ref{tab:HadERROR} rows 8 and 9, VAST models without covariates show lower absolute errors than when covariates are included).
 We analyze these results further in the sections that follow.
@@
 This can be further examined through yearly estimate:model ratio values, where we divide the yearly abundance estimates by the true model value.
-In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can result in yearly abundance values as low as 0 at times and exhibit large yearly changes.
-A representative example of this trend for each species is shown on the log scale in Figure \ref{fig:RatioTrend}.
-
-Analysis of individual yearly model:estimate ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21.
+In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can result in yearly abundance values as low as 0 at times and exhibit large yearly changes (Figure \ref{fig:RatioTrend}).
+
+Analysis of individual yearly estimate:model ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21.
 On the other hand, just 33\% of stratified mean estimate ratios were above 1 with an average of 0.874 and a standard deviation of 0.12.
@@
 The percent of the population increases in stratum 16 in the spring over the duration of both time series, which implies the flow out of strata 13 and 19 in the spring are going into stratum 16.
-These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12.
-These spring changes are therefore related to the static habitat values in our model rather than the temperature preferences, which is why we seen the same dynamics with constant and increasing temperature.
+These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail that coincides with most of the areas we have designated as the species' spawning ground and spawning takes place immediately before spring sampling in weeks 9-12.
+These spring changes are therefore related to the static habitat values in our model rather than the temperature preferences, which is why we see the same dynamics with constant and increasing temperature.
 While we observe similar changes in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in the population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \ref{fig:PopPct}).
-These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow outer strata 17 and 18.
+These dynamics imply that an increase in temperature results in the more shallow stratum 16 becoming less desirable than the deeper and more narrow outer strata 17 and 18.
 One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how \textasciitilde10\% of the population exists in the narrow outer strata 17 in the fall, while seemingly none of the population exists in any of the strata near the edge of the domain (14, 15, 17, 18) in the spring.
@@
 
-Of all scenarios without covariates, \(33/48\approx68\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
+Of all scenarios without covariates, \(33/48\approx68\)\% had a VAST fit with a lower absolute error compared to the stratified mean estimate.
 All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the fall season and had a common theme of producing abundance estimates that are above the true model value.
@@
 Including covariate information made a noticeable difference in our Yellowtail model-based VAST estimates.
-All of the VAST estimates that included covariate information produced a lower relative error than the corresponding VAST model that did not include covariates.
+All of the VAST estimates that included covariate information produced a lower absolute error than the corresponding VAST model that did not include covariates.
 The largest improvements were seen in the increasing temperature scenarios.
-When comparing to the design-based estimates, \(47/48\approx98\)\% of the VAST estimates that included covariates information had a lower relative error than the corresponding stratified mean estimate.
+When comparing to the design-based estimates, \(47/48\approx98\)\% of the VAST estimates that included covariates information had a lower absolute error than the corresponding stratified mean estimate.
 This implies that the covariate information helped our model-based estimate account for the design-based issues related to an increasing percentage of the population entering smaller strata, which can become exacerbated in the increasing temperature simulations.
@@
 The different VAST settings produced similar error values with \(18/32\approx57\%\) of estimates performing better with VAST Settings A.
-Of the abundance estimates without covariates, \(12/16=75\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
-VAST produced higher relative error compared to the stratified mean in scenarios that involved increasing temperature in the fall season without covariates.
+Of the abundance estimates without covariates, \(12/16=75\)\% had a VAST fit with a lower absolute error compared to the stratified mean estimate.
+VAST produced higher absolute error compared to the stratified mean in scenarios that involved increasing temperature in the fall season without covariates.
 Similar to the Yellowtail results, this implies the model-based approach had seasonal trouble with populations shifting into smaller strata where fewer samples take place.
@@
 
-Figure \ref{fig:PopPct} reveals some subtle seasonal differences in the percent of Haddock in each strata.
+Figure \ref{fig:PopPct} reveals some subtle seasonal differences in the percent of Haddock in each stratum.
 In the constant temperature scenarios, the spring shows a decrease in strata 19 and 20 that correspond to increases in 16, 24 and 29.
@@
 
-Of the scenarios without covariates, \(10/16\approx63\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate.
+Of the scenarios without covariates, \(10/16\approx63\)\% had a VAST fit with a lower absolute error compared to the stratified mean estimate.
 The 6 covariate-free VAST estimates that resulted in a higher error than the stratified mean spanned all scenarios and seasons, with several fall errors being especially large (significant overestimates).
 Adding covariate information resulted in \(14/16\approx88\)\% of VAST estimates having improved error compared to the stratified mean.
-The 2 scenarios that produced worse error with covarites spanned temperature scenarios, but were both in the spring season when the proportion of the population in each strata remained constant in each scenario.
+The 2 scenarios that produced worse error with covarites spanned temperature scenarios, but were both in the spring season when the proportion of the population in each stratum remained constant in each scenario.
 The average change in the error of abundance estimates between constant temperature scenarios and increasing temperature scenarios were 1.42 for VAST estimates without covariates, 1.80 for VAST estimates that included covariates, and 1.37 for stratified mean estimates.
@@
 Our analysis showed that both the design-based and model-based approaches are each capable of providing biomass estimates that track the true abundance trend.
-While VAST estimates tended to provide lower relative error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B also combined to produce over 30 abundance estimates with larger error values than the largest stratified mean error.
+While VAST estimates tended to provide lower absolute error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B also combined to produce over 30 abundance estimates with larger error values than the largest stratified mean error.
 These facts demonstrate the power of model-based estimates while also highlighting how sensitive they are to the choices made regarding settings and data used in analyses.
-Users often rely on traditional model-building processes and diagnostic techniques such as AIC to guide their choices, but successfully navigating the many decisions depends on the background of the given user and care should be taken to which diagnostic tools are used.
+Users often rely on traditional model-building processes and diagnostic techniques such as AIC to guide their choices, but successfully navigating the many decisions depends on the background of the given user and care should be taken in determining which diagnostic tools are used.
 For example, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using settings that decrease the accuracy of the model \textbf{cite Chris C paper}.
@@
 Thorson, James T. (n.d.a). {Vector Autoregressive Spatio-Temporal (VAST) model wiki page.} {https://github.com/James-Thorson-NOAA/VAST/wiki}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ThorsonChat}{}}%
+Thorson, James T. (n.d.c). {Personal Communication.} {??}
 
@@
 
-\caption{Percent of each species in each strata for during survey weeks in our spatial simulations.
+\caption{Percent of each species in each stratum for during survey weeks in our spatial simulations.
 All constant temperature scenarios follow the patterns on the left while increasing temperature scenarios follow the patterns on the right.
@@
 
-\caption{Representative example of bioamass estimate trend for Yellowtail.
+\caption{Representative example of biomass estimate trend for Yellowtail.
 Both scenarios have a relatively constant population with all strata being surveyed.
@@
 
-\caption{Representative example of bioamass estimate trend for Atlantic Cod.
+\caption{Representative example of biomass estimate trend for Atlantic Cod.
 Both scenarios have a decreasing population trend with all strata being surveyed.
@@
 
-\caption{Representative example of bioamass estimate trend for Haddock.
+\caption{Representative example of biomass estimate trend for Haddock.
 Both scenarios have an increasing population trend and a repeating temperature gradient.
@@
 
-\caption{\label{tab:YTresultsallstrata}Yellowtail Flounder error results with all strata included in calculations. Row colors correspond to the same settings applied in different seasons.}
+\caption{\label{tab:YTresultsallstrata}Yellowtail Flounder absolute error results with all strata included in calculations. Row colors correspond to the same settings applied in different seasons.}
 \centering
@@
 
-\caption{\label{tab:YTresultsreducedstrata}Yellowtail Flounder error results with certain strata excluded from calculations. Row colors correspond to the same settings applied in different seasons.}
+\caption{\label{tab:YTresultsreducedstrata}Yellowtail Flounder absoloute error results with certain strata excluded from calculations.
+Row colors correspond to the same settings applied in different seasons.}
 \centering
@@
 
-\caption{\label{tab:CodERROR}Cod error results.}
+\caption{\label{tab:CodERROR}Cod absolute error results.}
 \centering
@@
 
-\caption{\label{tab:HadERROR}Haddock error results.}
+\caption{\label{tab:HadERROR}Haddock absolute error results.}
 \centering

### 2023-08-15 Manuscript/Draft2.Rmd
manuscript updates

@@
 
-# Estimating Population Trends with Stratified Random Sampling Under the Pressures of Climate Change
+# Estimating Population Trends Under the Pressures of Climate Change
 
@@
 
-Corresponding author: Ben Levy (benjamin.levy@noaa.gov)
+Corresponding author: Ben Levy (blevy6@gmail.com)
 
@@
 
-Many stock assessments in the United States use abundance estimates derived from stratified random bottom trawl data.
-To accurately track changes in true abundance of a species, catch observations must contain low noise and the proportion of the population that is sampled should be consistent over time.
-One increasingly documented phenomenon that violates the latter assumption is climate-induces changes in relative abundance between sampled and unsampled portions of the stock area.
+Many stock assessments in the United States use biomass estimates derived from stratified random bottom trawl data.
+To accurately track changes in true biomass of a species, catch observations must contain low noise and the proportion of the population that is sampled should be consistent over time.
+One increasingly documented phenomenon that violates the latter assumption is climate-induces movement leading to changes in relative biomass between sampled and unsampled portions of the stock area.
 Using the R package MixFishSim, we have developed data-driven spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock in the western Atlantic Ocean to evaluate two methods of estimating population trend given temperature driven movement.
@@
 A repeating yearly temperature pattern produces repeating spatial biomass distributions in a given week, while a temperature gradient that increases on average over time results in spatial preferences that evolve throughout a given simulation.
-We simulated spatial time series datasets for each species for several temperature scenarios and population trends.
-Using stratified random sampling on model output we were able to compare abundance estimates derived from the design-based stratified mean to those using a spatio-temporal model-based approach that allows inclusion of environmental covariates (VAST).
+We simulated spatial time series data sets for each species for several temperature scenarios and population trends.
+Using stratified random sampling on model output we were able to compare biomass estimates derived from the design-based stratified mean to those using a spatio-temporal model-based approach that allows inclusion of environmental covariates.
 Our focus was on the ability of contemporary index standardization methods to track population trends under shifting spatial preferences.
 
-<!-- We found that in general, both index calculation methods are able to produce similar abundance trends to the model output, with model-based estimates providing more consistent year-to-year estimates.
-The design-based approach tended to provide abundance estimates that were below the true values in our models, while model-based estimates tended to overestimate true abundance.
+<!-- We found that in general, both index calculation methods are able to produce similar biomass trends to the model output, with model-based estimates providing more consistent year-to-year estimates.
+The design-based approach tended to provide biomass estimates that were below the true values in our models, while model-based estimates tended to overestimate true biomass.
 We highlight scenarios where index calculation methods produced misleading estimates as species shifted their spatial densities over time within their assumed stock boundaries.
@@
 The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf, where depth and latitude define the strata.
-The survey has created a rich time series data set with many uses including species-specific habitat identification, age and life history information, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments, and ultimately fishing quota limits.
+The survey has created a rich time series data set with many uses including species-specific habitat identification, age and life history information, analysis of how environmental conditions influence species biomass, and estimating yearly species biomass trends to help inform stock assessments, and ultimately fishing quota limits.
 The survey takes place twice each year- once in the spring and again in the fall.
@@
 
-One component of a stock assessment is an estimation of yearly abundance that is often derived from the bottom trawl survey, when applicable.
-Stock assessment scientists choose from a number of approaches to obtain abundance estimates ranging from traditional design-based estimates to model-based estimates that vary in complexity.
+One component of a stock assessment is an estimation of yearly biomass that is often derived from the bottom trawl survey, when applicable.
+Stock assessment scientists choose from a number of approaches to obtain biomass estimates ranging from traditional design-based estimates to model-based estimates that vary in complexity.
 Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest.
-These methods do not account for spatial variation in samples and are not able to relate environmental influences with survey values.
-Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors).
+These methods typically account for spatial variation in samples by stratifying the sampling area.
+While it is possible to include environmental influences with design-based methods [@deville1992calibration], it is not a common practice.
+Model-based biomass estimates use statistical models to measure the relationship between response variables (such as presence or biomass) and predictor variables (such as environmental factors).
 Model-based estimators help account for complex relationships between variables and can help overcome problems related to the sampling process.
@@
 In these scenarios, management strategy evaluations have shown that misconceptions of stock status can lead to unintended overfishing, which can ultimately have detrimental ecologic and economic impacts [@mazur2023consequences].
-We are therefore interested in analyzing the impact of climate driven movement and distribution changes on the accuracy of abundance estimates derived from NOAA’s ongoing bottom-trawl survey along the east coast.
+We are therefore interested in analyzing the impact of climate driven movement and distribution changes on the accuracy of biomass estimates derived from NOAA’s ongoing bottom-trawl survey along the east coast.
 
@@
 We consider the impact of climate change by comparing simulations that use a repeating water temperature pattern to those where temperature increases on average over time.
-In both cases we conduct stratified random sampling on model output and analyze the ability of the samples to track population trends.
-We compare yearly abundance estimates obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
+In both cases we conduct stratified random sampling (SRS) on model output and analyze the ability of the samples to track population trends.
+We compare yearly biomass estimates obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
 The stratified mean is a design-based approach that calculates the stratum specific mean catch per tow, which are then weighted by stratum area to obtain an overall mean.
-VAST is a spatial delta-generalized linear mixed model that estimates both abundance (biomass) and probability of occurrence (presence/absence) [@thorson2019guidance].
+VAST is a spatial delta-generalized linear mixed model that estimates both biomass and probability of occurrence (presence/absence) [@thorson2019guidance].
 VAST also allows users to include covariate data to better inform the model.
@@
 temperature).
-We explore whether including environmental predictors leads to improved abundance estimates, which is particularly relevant as climate change progresses.
+We explore whether including environmental predictors leads to improved biomass estimates, which is particularly relevant as climate change progresses.
 
@@
 Using temperature gradients that repeat each year creates data sets with repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time.
-We conduct stratified random sampling on our simulation data to mimic the bottom trawl survey and use the samples to compare the ability of contemporary index standardization methods to track population trends.
+We conduct SRS on our simulation data to mimic the bottom trawl survey and use the samples to compare the ability of contemporary index standardization methods to track population trends.
 
@@
 Historically, Cod has seen significant decline over the last 50**??** years while Haddock has increased in abundance in recent years [[can cite the 2022 management track assessments, see https://apps-nefsc.fisheries.noaa.gov/saw/sari.php for when the document becomes available]] **just cite 2021 assessments?**.
-For this reason we compare survey index estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
-We consider increasing, decreasing, and constant population trends for Yellowtail for completeness.
+For this reason we compare survey index estimates using SRS from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
+To consider a wide range of possibilities, we include increasing, decreasing, and constant population trends for Yellowtail.
 The specific population trends used in our analyses can be see in Figure \@ref(fig:pop-scenarios).
-Each of these scenarios is simulated twice: first with with an oscillating temperature gradient that repeats and second with a temperature gradient that increases roughly 5 degrees Celsius over the duration of the 20 year simulation, for a total of 10 simulated spatial datasets.
+Each of these scenarios is simulated twice: first with with an oscillating temperature gradient that repeats and second with a temperature gradient that increases roughly 5 degrees Celsius over the duration of the 20 year simulation, for a total of 10 simulated spatial data sets.
 
@@
 
-After distributing each species in its stock area with simulated population trajectories, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited stratum twice each year.
+After distributing each species in its stock area with simulated population trajectories, we mimic the bottom trawl survey by conducting SRS in the set of strata associated with each stock twice each year.
 We sample each stratum in the same weeks in which the spring and fall surveys have historically sampled Georges Bank (weeks 13 & 14 in the spring and 37 & 38 in the fall).
@@
 
-An underlying assumption in all index methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) consistently sampling the same proportion of the population.
+An underlying assumption in all index methods is that individual random samples combine to accurately represent true biomass by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) consistently sampling the same proportion of the population.
 These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata.
 To simulate the impact of noise, we compare index estimates after adding noise to our samples versus those using the true sampling values.
-Annual survey observations were simulated as log-normal deviations from the underlying “true” survey catches with a CV of 0.3.
-The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extent of the population.
+Annual survey observations were simulated as log-normal deviations from the underlying “true” survey catches with a CV of 0.35.
+The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that SRS will fail to sample the entire geographic extent of the population.
 We simulate the effect of populations moving into new habitat by comparing index estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species.
-The strata to exclude for each species were chosen by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from.
+The strata to exclude for each species were chosen by reviewing how the distribution of each species evolved in our increasing temperature scenarios and removing strata where the proportion of a given population was changing over time.
 Figure \@ref(fig:strata-plot) shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \@ref(fig:PopPct).
 The yellow regions in Figure \@ref(fig:strata-plot) depict the strata that were not sampled for each species.
-We then use the biomass collected from our samples in contemporary abundance index methods to estimate yearly seasonal population trends.
+We then use the biomass collected from our samples in contemporary biomass index methods to estimate yearly seasonal population trends.
 
@@
 
-where $W_i=\frac{S_i}{\sum_{j=1}^N S_j}$ and $y_{k,i}$ represents the biomass in sample $k$ of stratum $i$ in season $s$ and year $y$.
+where $W_i=\frac{A_i}{\sum_{j=1}^N A_j}$, $A_i$ is the area of stratum $i$, and $y_{k,i}$ represents the biomass in sample $k$ of stratum $i$ in season $s$ and year $y$.
 These tow-dependent calculations are quick and easy to calculate, especially relative to the model-based VAST estimates.
@@
 
-We follow the advice given in [@thorson2019guidance] and [@VASTwiki] to build VAST estimates of biomass in our Georges Bank spatial population models using stratified random samples from our simulation output.
+We follow the advice given in [@thorson2019guidance] and https://github.com/James-Thorson-NOAA/VAST/wiki to build VAST estimates of biomass in our Georges Bank spatial population models using stratified random samples from our simulation output.
 In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
@@
 Through this process, and in consultation with the VAST package creator, we compared the performance of two model setting configurations in our VAST models as shown in Table 3.
-VAST Settings B were chosen after reviewing the information provided in the "Seasonal Model" section of VAST's online Github tutorial [@ThorsonSeasonal].
+VAST Settings B were chosen after reviewing the information provided in the "Seasonal Model" section of VAST's online Github tutorial found at https://james-thorson-noaa.github.io/docs/tutorials/seasonal-model/.
 After running models for each of our scenarios using these settings, we received feedback to also consider configurations for Settings A [@ThorsonChat].
@@
 
-Each abundance index scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, & covariates), resulting in a large number of scenario combinations to consider.
+Each biomass index scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, & covariates), resulting in a large number of scenario combinations to consider.
 The columns in Table \@ref(tab:scenarios) show the choices that define each scenario.
@@
 We notice a small amount of shifting between successive years in these constant temperature scenarios as the population aggregates on especially suitable habitat in the domain.
-By running simulations in *MixFishSim* with a temperature gradient that increased on average over time we were able to create spatial datasets with shifting biomass distributions.
+By running simulations in *MixFishSim* with a temperature gradient that increased on average over time we were able to create spatial data sets with shifting biomass distributions.
 As seen in the right column of Figure \@ref(fig:PopPct), more exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time.
 
-Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the absolute error between biomass estimates and model output for each of our 20-year abundance estimates.
+Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the absolute error between biomass estimates and model output for each of our 20-year biomass estimates.
 While the model-based results provided a slightly lower overall mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02).
@@
 
-<font size="5">*Abundance Estimate Ratio Results*</font>
-
-A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values.
-This can be further examined through yearly estimate:model ratio values, where we divide the yearly abundance estimates by the true model value.
-In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can result in yearly abundance values as low as 0 at times and exhibit large yearly changes (Figure \@ref(fig:RatioTrend)).
+<font size="5">*Biomass Estimate Ratio Results*</font>
+
+A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide biomass estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values.
+This can be further examined through yearly estimate:model ratio values, where we divide the yearly biomass estimates by the true model value.
+In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can result in yearly biomass values as low as 0 at times and exhibit large yearly changes (Figure \@ref(fig:RatioTrend)).
 
@@
 
-Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata) contain the absolute error between our abundance estimates for Yellowtail comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B).
+Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata) contain the absolute error between our biomass estimates for Yellowtail comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B).
 In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
@@
 Including covariates still improve VAST estimates when certain strata are excluded from sampling, but the change in error was smaller than with all strata included.
-There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata).
+There are several instances in which VAST failed to provide improved biomass estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata).
 However, including covariate information to each VAST model produced estimates with significantly lower error than the corresponding stratified mean estimate.
@@
 Of all scenarios without covariates, $33/48\approx68$\% had a VAST fit with a lower absolute error compared to the stratified mean estimate.
-All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the fall season and had a common theme of producing abundance estimates that are above the true model value.
+All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the fall season and had a common theme of producing biomass estimates that are above the true model value.
 These 15 fall estimates span all other scenario variations.
@@
 This implies that the covariate information helped our model-based estimate account for the design-based issues related to an increasing percentage of the population entering smaller strata, which can become exacerbated in the increasing temperature simulations.
-A representative example of Yellowtail abundance estimates is shown in Figure \@ref(fig:YTEstEx), where one can visualize the impact on biomass estimates between constant and increasing temperature, and how including covariates can improve the estimate.
-
-We see diminished performance of our abundance estimates for Yellowtail under increasing temperature, with the most dramatic changes seen in our VAST estimates without covariates.
+A representative example of Yellowtail biomass estimates is shown in Figure \@ref(fig:YTEstEx), where one can visualize the impact on biomass estimates between constant and increasing temperature, and how including covariates can improve the estimate.
+
+We see diminished performance of our biomass estimates for Yellowtail under increasing temperature, with the most dramatic changes seen in our VAST estimates without covariates.
 One exception to this is when we use a stratified mean approach while the population is decreasing.
-Our analyses have found that the stratified mean tends to under estimate the true abundance and since these estimates are bounded below by zero, as the Yellowtail population decreases towards zero the difference between the estimate and the true value also decrease.
+Our analyses have found that the stratified mean tends to under estimate the true biomass and since these estimates are bounded below by zero, as the Yellowtail population decreases towards zero the difference between the estimate and the true value also decrease.
 That is, if the population is low enough, failing to appropriately sample the population in a design-based method produces the same result as appropriately sampling.
-In comparing abundances estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a larger impact on the model-based estimates compared to the design-based method.
-Specifically, when comparing the error of abundance estimates from the constant temperature scenario to the corresponding increasing temperature scenario, VAST increased by an average factor of 1.75 (both with and without covariates) while the  stratified mean error increased by an average factor of 1.25.
+In comparing biomass estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a larger impact on the model-based estimates compared to the design-based method.
+Specifically, when comparing the error of biomass estimates from the constant temperature scenario to the corresponding increasing temperature scenario, VAST increased by an average factor of 1.75 (both with and without covariates) while the  stratified mean error increased by an average factor of 1.25.
 
@@
 
-Table \@ref(tab:CodERROR) contains the absolute error between our abundance estimates for Cod and the true model values.
+Table \@ref(tab:CodERROR) contains the absolute error between our biomass estimates for Cod and the true model values.
 The different VAST settings produced similar error values with $18/32\approx57\%$ of estimates performing better with VAST Settings A.
-Of the abundance estimates without covariates, $12/16=75$\% had a VAST fit with a lower absolute error compared to the stratified mean estimate.
+Of the biomass estimates without covariates, $12/16=75$\% had a VAST fit with a lower absolute error compared to the stratified mean estimate.
 VAST produced higher absolute error compared to the stratified mean in scenarios that involved increasing temperature in the fall season without covariates.
@@
 
-In comparing abundances estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a noticable negative impact on the model-based estimates.
-Specifically, when comparing the error of abundance estimates from the constant temperature scenario to the corresponding increasing temperature scenario for Cod, VAST increased by an average factor of 3.57 without covariates and 1.67 with covariates.
-In considering the average change in the error of stratified mean estimates between constant temperature scenarios and increasing temperature scenarios we surprisingly find an improvement in the error of abundance estimates, decreasing by an average factor of 0.87.
+In comparing biomass estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a noticable negative impact on the model-based estimates.
+Specifically, when comparing the error of biomass estimates from the constant temperature scenario to the corresponding increasing temperature scenario for Cod, VAST increased by an average factor of 3.57 without covariates and 1.67 with covariates.
+In considering the average change in the error of stratified mean estimates between constant temperature scenarios and increasing temperature scenarios we surprisingly find an improvement in the error of biomass estimates, decreasing by an average factor of 0.87.
 
 <!-- Similar to Yellowtail results, Cod are shifting their spatial distribution into the smaller exterior strata in these conditions, which is decreasing the accuracy of the covariate-free VAST estimate.
-While adding covariates to these fits allowed for an improved error relative to the stratified mean, the VAST abundance still display an overestimate early in the time series before covariates correct the trend, allowing for a lower overall absolute error **(show these plots?
+While adding covariates to these fits allowed for an improved error relative to the stratified mean, the VAST biomass still display an overestimate early in the time series before covariates correct the trend, allowing for a lower overall absolute error **(show these plots?
 IncTemp fall scenarios)**.
@@
 
-Table \@ref(tab:HadERROR) contains the absolute error between our abundance estimates for Haddock and the true model values.
+Table \@ref(tab:HadERROR) contains the absolute error between our biomass estimates for Haddock and the true model values.
 Absolute error values were lower with VAST Settings B in $21/32\approx 66\%$ of scenarios.
@@
 The 2 scenarios that produced worse error with covarites spanned temperature scenarios, but were both in the spring season when the proportion of the population in each stratum remained constant in each scenario.
-The average change in the error of abundance estimates between constant temperature scenarios and increasing temperature scenarios were 1.42 for VAST estimates without covariates, 1.80 for VAST estimates that included covariates, and 1.37 for stratified mean estimates.
+The average change in the error of biomass estimates between constant temperature scenarios and increasing temperature scenarios were 1.42 for VAST estimates without covariates, 1.80 for VAST estimates that included covariates, and 1.37 for stratified mean estimates.
 
@@
 
-To test the impact of shifting populations on abundance estimates, we simulated population data that included environmental forcing and compared biomass estimates derived from stratified random sampling to the known true values.
+To test the impact of shifting populations on biomass estimates, we simulated population data that included environmental forcing and compared biomass estimates derived from SRS to the known true values.
 We carried out simulations for 3 species on Georges Bank that took into account different population trends, temperature scenarios, noise in survey samples, and seasonal variation.
@@
 
-Our analysis showed that both the design-based and model-based approaches are each capable of providing biomass estimates that track the true abundance trend.
-While VAST estimates tended to provide lower absolute error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B also combined to produce over 30 abundance estimates with larger error values than the largest stratified mean error.
+Our analysis showed that both the design-based and model-based approaches are each capable of providing biomass estimates that track the true biomass trend.
+While VAST estimates tended to provide lower absolute error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B also combined to produce over 30 biomass estimates with larger error values than the largest stratified mean error.
 These facts demonstrate the power of model-based estimates while also highlighting how sensitive they are to the choices made regarding settings and data used in analyses.
@@
 
-Seasonal differences in spatial abundance patterns in our population models led to seasonal differences in abundance estimates, with most estimates producing a higher error in the fall.
+Seasonal differences in spatial biomass patterns in our population models led to seasonal differences in biomass estimates, with most estimates producing a higher error in the fall.
 This may have to do with the fact that our species spawn during months 9-14, which coincides with the spring sampling weeks 13 & 14.
@@
 The populations are more spread out during the fall survey in weeks 37 & 38 and since the water temperature is warmer during these months biomass shifts towards the deeper strata, which tend to be less sampled.
-We explored this through abundance estimate ratios which revealed how the model-based estimates had a larger change in average ratio between seasonal estimates compared to the design-based estimates.
+We explored this through biomass estimate ratios which revealed how the model-based estimates had a larger change in average ratio between seasonal estimates compared to the design-based estimates.
 
 Spatial factors contributed to many of the scenarios that produced larger error values seen in our estimates.
-Flow from the shallow centrally located strata into the deeper, smaller, and less-sampled eastern strata contributed to VAST failing to accurately model the abundance of Yellowtail without the addition of covariate information, the impact of which was exacerbated in the increasing temperature scenarios.
+Flow from the shallow centrally located strata into the deeper, smaller, and less-sampled eastern strata contributed to VAST failing to accurately model the biomass of Yellowtail without the addition of covariate information, the impact of which was exacerbated in the increasing temperature scenarios.
 While adding covariates can help inform the model and improve the estimate as seen with the Yellowtail simulation results, covariates can hinder the model in certain instances with Cod and Haddock.
@@
 More specifically, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates.
-This is in contrast to our Yellowtail results, where adding covariate information to VAST models always decreased the absolute error in the resulting abundance estimates.
+This is in contrast to our Yellowtail results, where adding covariate information to VAST models always decreased the absolute error in the resulting biomass estimates.
 Although strata were excluded for both species, the decline in performance for just Cod and Haddock can be explained by a failure to sample the full spectrum of temperature values where the species exists.

### 2023-08-15 Manuscript/Draft2.tex
manuscript updates

@@
 
-\hypertarget{estimating-population-trends-with-stratified-random-sampling-under-the-pressures-of-climate-change}{%
-\section{Estimating Population Trends with Stratified Random Sampling Under the Pressures of Climate Change}\label{estimating-population-trends-with-stratified-random-sampling-under-the-pressures-of-climate-change}}
+\hypertarget{estimating-population-trends-under-the-pressures-of-climate-change}{%
+\section{Estimating Population Trends Under the Pressures of Climate Change}\label{estimating-population-trends-under-the-pressures-of-climate-change}}
 
@@
 
-Corresponding author: Ben Levy (\href{mailto:benjamin.levy@noaa.gov}{\nolinkurl{benjamin.levy@noaa.gov}})
+Corresponding author: Ben Levy (\href{mailto:blevy6@gmail.com}{\nolinkurl{blevy6@gmail.com}})
 
@@
 A repeating yearly temperature pattern produces repeating spatial biomass distributions in a given week, while a temperature gradient that increases on average over time results in spatial preferences that evolve throughout a given simulation.
-We simulated spatial time series datasets for each species for several temperature scenarios and population trends.
-Using stratified random sampling on model output we were able to compare abundance estimates derived from the design-based stratified mean to those using a spatio-temporal model-based approach that allows inclusion of environmental covariates (VAST).
+We simulated spatial time series data sets for each species for several temperature scenarios and population trends.
+Using stratified random sampling on model output we were able to compare biomass estimates derived from the design-based stratified mean to those using a spatio-temporal model-based approach that allows inclusion of environmental covariates.
 Our focus was on the ability of contemporary index standardization methods to track population trends under shifting spatial preferences.
@@
 
-One component of a stock assessment is an estimation of yearly abundance that is often derived from the bottom trawl survey, when applicable.
-Stock assessment scientists choose from a number of approaches to obtain abundance estimates ranging from traditional design-based estimates to model-based estimates that vary in complexity.
+One component of a stock assessment is an estimation of yearly biomass that is often derived from the bottom trawl survey, when applicable.
+Stock assessment scientists choose from a number of approaches to obtain biomass estimates ranging from traditional design-based estimates to model-based estimates that vary in complexity.
 Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest.
-These methods do not account for spatial variation in samples and are not able to relate environmental influences with survey values.
-Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors).
+These methods typically account for spatial variation in samples by stratifying the sampling area.
+While it is possible to include environmental influences with design-based methods (Deville and Särndal 1992), it is not a common practice.
+Model-based biomass estimates use statistical models to measure the relationship between response variables (such as presence or biomass) and predictor variables (such as environmental factors).
 Model-based estimators help account for complex relationships between variables and can help overcome problems related to the sampling process.
@@
 2023).
-We are therefore interested in analyzing the impact of climate driven movement and distribution changes on the accuracy of abundance estimates derived from NOAA's ongoing bottom-trawl survey along the east coast.
+We are therefore interested in analyzing the impact of climate driven movement and distribution changes on the accuracy of biomass estimates derived from NOAA's ongoing bottom-trawl survey along the east coast.
 
@@
 We consider the impact of climate change by comparing simulations that use a repeating water temperature pattern to those where temperature increases on average over time.
-In both cases we conduct stratified random sampling on model output and analyze the ability of the samples to track population trends.
-We compare yearly abundance estimates obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
+In both cases we conduct stratified random sampling (SRS) on model output and analyze the ability of the samples to track population trends.
+We compare yearly biomass estimates obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
 The stratified mean is a design-based approach that calculates the stratum specific mean catch per tow, which are then weighted by stratum area to obtain an overall mean.
-VAST is a spatial delta-generalized linear mixed model that estimates both abundance (biomass) and probability of occurrence (presence/absence) (Thorson 2019).
+VAST is a spatial delta-generalized linear mixed model that estimates both biomass and probability of occurrence (presence/absence) (Thorson 2019).
 VAST also allows users to include covariate data to better inform the model.
 Covariates can be static (e.g.~habitat preferences) or dynamics (e.g.~temperature).
-We explore whether including environmental predictors leads to improved abundance estimates, which is particularly relevant as climate change progresses.
+We explore whether including environmental predictors leads to improved biomass estimates, which is particularly relevant as climate change progresses.
 
@@
 Historically, Cod has seen significant decline over the last 50\textbf{??} years while Haddock has increased in abundance in recent years {[}{[}can cite the 2022 management track assessments, see \url{https://apps-nefsc.fisheries.noaa.gov/saw/sari.php} for when the document becomes available{]}{]} \textbf{just cite 2021 assessments?}.
-For this reason we compare survey index estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
-We consider increasing, decreasing, and constant population trends for Yellowtail for completeness.
+For this reason we compare survey index estimates using SRS from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
+To consider a wide range of possibilities, we include increasing, decreasing, and constant population trends for Yellowtail.
 The specific population trends used in our analyses can be see in Figure \ref{fig:pop-scenarios}.
-Each of these scenarios is simulated twice: first with with an oscillating temperature gradient that repeats and second with a temperature gradient that increases roughly 5 degrees Celsius over the duration of the 20 year simulation, for a total of 10 simulated spatial datasets.
+Each of these scenarios is simulated twice: first with with an oscillating temperature gradient that repeats and second with a temperature gradient that increases roughly 5 degrees Celsius over the duration of the 20 year simulation, for a total of 10 simulated spatial data sets.
 
@@
 
-After distributing each species in its stock area with simulated population trajectories, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited stratum twice each year.
+After distributing each species in its stock area with simulated population trajectories, we mimic the bottom trawl survey by conducting SRS in the set of strata associated with each stock twice each year.
 We sample each stratum in the same weeks in which the spring and fall surveys have historically sampled Georges Bank (weeks 13 \& 14 in the spring and 37 \& 38 in the fall).
@@
 
-An underlying assumption in all index methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) consistently sampling the same proportion of the population.
+An underlying assumption in all index methods is that individual random samples combine to accurately represent true biomass by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) consistently sampling the same proportion of the population.
 These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata.
 To simulate the impact of noise, we compare index estimates after adding noise to our samples versus those using the true sampling values.
-Annual survey observations were simulated as log-normal deviations from the underlying ``true'' survey catches with a CV of 0.3.
-The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extent of the population.
+Annual survey observations were simulated as log-normal deviations from the underlying ``true'' survey catches with a CV of 0.35.
+The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that SRS will fail to sample the entire geographic extent of the population.
 We simulate the effect of populations moving into new habitat by comparing index estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species.
-The strata to exclude for each species were chosen by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from.
+The strata to exclude for each species were chosen by reviewing how the distribution of each species evolved in our increasing temperature scenarios and removing strata where the proportion of a given population was changing over time.
 Figure \ref{fig:strata-plot} shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \ref{fig:PopPct}.
 The yellow regions in Figure \ref{fig:strata-plot} depict the strata that were not sampled for each species.
-We then use the biomass collected from our samples in contemporary abundance index methods to estimate yearly seasonal population trends.
+We then use the biomass collected from our samples in contemporary biomass index methods to estimate yearly seasonal population trends.
 
@@
 
-where \(W_i=\frac{S_i}{\sum_{j=1}^N S_j}\) and \(y_{k,i}\) represents the biomass in sample \(k\) of stratum \(i\) in season \(s\) and year \(y\).
+where \(W_i=\frac{A_i}{\sum_{j=1}^N A_j}\), \(A_i\) is the area of stratum \(i\), and \(y_{k,i}\) represents the biomass in sample \(k\) of stratum \(i\) in season \(s\) and year \(y\).
 These tow-dependent calculations are quick and easy to calculate, especially relative to the model-based VAST estimates.
@@
 
-We follow the advice given in (Thorson 2019) and (Thorson, James T n.d.a) to build VAST estimates of biomass in our Georges Bank spatial population models using stratified random samples from our simulation output.
+We follow the advice given in (Thorson 2019) and \url{https://github.com/James-Thorson-NOAA/VAST/wiki} to build VAST estimates of biomass in our Georges Bank spatial population models using stratified random samples from our simulation output.
 In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation.
@@
 Through this process, and in consultation with the VAST package creator, we compared the performance of two model setting configurations in our VAST models as shown in Table 3.
-VAST Settings B were chosen after reviewing the information provided in the ``Seasonal Model'' section of VAST's online Github tutorial (Thorson, James T n.d.b).
-After running models for each of our scenarios using these settings, we received feedback to also consider configurations for Settings A (Thorson, James T n.d.c).
+VAST Settings B were chosen after reviewing the information provided in the ``Seasonal Model'' section of VAST's online Github tutorial found at \url{https://james-thorson-noaa.github.io/docs/tutorials/seasonal-model/}.
+After running models for each of our scenarios using these settings, we received feedback to also consider configurations for Settings A (Thorson, James T 2022).
 Both settings use a Tweedie model and turn off the first linear predictor \(p_1\) so that only biomass is being modeled.
@@
 The difference amounts to intercepts for the biomass predictor being a fixed effect (RhoBeta2=0) versus intercepts being constant among years as a fixed effect (RhoBeta2=3).
-The specifications for Settings A are more typical for stock assessment application (Thorson, James T n.d.c), and we therefore focus our discussion on these VAST results.
+The specifications for Settings A are more typical for stock assessment application (Thorson, James T 2022), and we therefore focus our discussion on these VAST results.
 Results for Settings B are shown to demonstrate how the settings used in model-based approaches impact model performance.
 
-Each abundance index scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, \& covariates), resulting in a large number of scenario combinations to consider.
+Each biomass index scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, \& covariates), resulting in a large number of scenario combinations to consider.
 The columns in Table \ref{tab:scenarios} show the choices that define each scenario.
@@
 We notice a small amount of shifting between successive years in these constant temperature scenarios as the population aggregates on especially suitable habitat in the domain.
-By running simulations in \emph{MixFishSim} with a temperature gradient that increased on average over time we were able to create spatial datasets with shifting biomass distributions.
+By running simulations in \emph{MixFishSim} with a temperature gradient that increased on average over time we were able to create spatial data sets with shifting biomass distributions.
 As seen in the right column of Figure \ref{fig:PopPct}, more exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time.
 
-Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the absolute error between biomass estimates and model output for each of our 20-year abundance estimates.
+Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the absolute error between biomass estimates and model output for each of our 20-year biomass estimates.
 While the model-based results provided a slightly lower overall mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02).
@@
 
-\emph{Abundance Estimate Ratio Results}
-
-A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values.
-This can be further examined through yearly estimate:model ratio values, where we divide the yearly abundance estimates by the true model value.
-In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can result in yearly abundance values as low as 0 at times and exhibit large yearly changes (Figure \ref{fig:RatioTrend}).
+\emph{Biomass Estimate Ratio Results}
+
+A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide biomass estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values.
+This can be further examined through yearly estimate:model ratio values, where we divide the yearly biomass estimates by the true model value.
+In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can result in yearly biomass values as low as 0 at times and exhibit large yearly changes (Figure \ref{fig:RatioTrend}).
 
@@
 
-Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata} contain the absolute error between our abundance estimates for Yellowtail comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B).
+Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata} contain the absolute error between our biomass estimates for Yellowtail comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B).
 In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
@@
 Including covariates still improve VAST estimates when certain strata are excluded from sampling, but the change in error was smaller than with all strata included.
-There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}.
+There are several instances in which VAST failed to provide improved biomass estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}.
 However, including covariate information to each VAST model produced estimates with significantly lower error than the corresponding stratified mean estimate.
@@
 Of all scenarios without covariates, \(33/48\approx68\)\% had a VAST fit with a lower absolute error compared to the stratified mean estimate.
-All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the fall season and had a common theme of producing abundance estimates that are above the true model value.
+All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the fall season and had a common theme of producing biomass estimates that are above the true model value.
 These 15 fall estimates span all other scenario variations.
@@
 This implies that the covariate information helped our model-based estimate account for the design-based issues related to an increasing percentage of the population entering smaller strata, which can become exacerbated in the increasing temperature simulations.
-A representative example of Yellowtail abundance estimates is shown in Figure \ref{fig:YTEstEx}, where one can visualize the impact on biomass estimates between constant and increasing temperature, and how including covariates can improve the estimate.
-
-We see diminished performance of our abundance estimates for Yellowtail under increasing temperature, with the most dramatic changes seen in our VAST estimates without covariates.
+A representative example of Yellowtail biomass estimates is shown in Figure \ref{fig:YTEstEx}, where one can visualize the impact on biomass estimates between constant and increasing temperature, and how including covariates can improve the estimate.
+
+We see diminished performance of our biomass estimates for Yellowtail under increasing temperature, with the most dramatic changes seen in our VAST estimates without covariates.
 One exception to this is when we use a stratified mean approach while the population is decreasing.
-Our analyses have found that the stratified mean tends to under estimate the true abundance and since these estimates are bounded below by zero, as the Yellowtail population decreases towards zero the difference between the estimate and the true value also decrease.
+Our analyses have found that the stratified mean tends to under estimate the true biomass and since these estimates are bounded below by zero, as the Yellowtail population decreases towards zero the difference between the estimate and the true value also decrease.
 That is, if the population is low enough, failing to appropriately sample the population in a design-based method produces the same result as appropriately sampling.
-In comparing abundances estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a larger impact on the model-based estimates compared to the design-based method.
-Specifically, when comparing the error of abundance estimates from the constant temperature scenario to the corresponding increasing temperature scenario, VAST increased by an average factor of 1.75 (both with and without covariates) while the stratified mean error increased by an average factor of 1.25.
+In comparing biomass estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a larger impact on the model-based estimates compared to the design-based method.
+Specifically, when comparing the error of biomass estimates from the constant temperature scenario to the corresponding increasing temperature scenario, VAST increased by an average factor of 1.75 (both with and without covariates) while the stratified mean error increased by an average factor of 1.25.
 
@@
 
-Table \ref{tab:CodERROR} contains the absolute error between our abundance estimates for Cod and the true model values.
+Table \ref{tab:CodERROR} contains the absolute error between our biomass estimates for Cod and the true model values.
 The different VAST settings produced similar error values with \(18/32\approx57\%\) of estimates performing better with VAST Settings A.
-Of the abundance estimates without covariates, \(12/16=75\)\% had a VAST fit with a lower absolute error compared to the stratified mean estimate.
+Of the biomass estimates without covariates, \(12/16=75\)\% had a VAST fit with a lower absolute error compared to the stratified mean estimate.
 VAST produced higher absolute error compared to the stratified mean in scenarios that involved increasing temperature in the fall season without covariates.
@@
 
-In comparing abundances estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a noticable negative impact on the model-based estimates.
-Specifically, when comparing the error of abundance estimates from the constant temperature scenario to the corresponding increasing temperature scenario for Cod, VAST increased by an average factor of 3.57 without covariates and 1.67 with covariates.
-In considering the average change in the error of stratified mean estimates between constant temperature scenarios and increasing temperature scenarios we surprisingly find an improvement in the error of abundance estimates, decreasing by an average factor of 0.87.
+In comparing biomass estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a noticable negative impact on the model-based estimates.
+Specifically, when comparing the error of biomass estimates from the constant temperature scenario to the corresponding increasing temperature scenario for Cod, VAST increased by an average factor of 3.57 without covariates and 1.67 with covariates.
+In considering the average change in the error of stratified mean estimates between constant temperature scenarios and increasing temperature scenarios we surprisingly find an improvement in the error of biomass estimates, decreasing by an average factor of 0.87.
 
@@
 
-Table \ref{tab:HadERROR} contains the absolute error between our abundance estimates for Haddock and the true model values.
+Table \ref{tab:HadERROR} contains the absolute error between our biomass estimates for Haddock and the true model values.
 Absolute error values were lower with VAST Settings B in \(21/32\approx 66\%\) of scenarios.
@@
 
-Seasonal differences in spatial abundance patterns in our population models led to seasonal differences in abundance estimates, with most estimates producing a higher error in the fall.
+Seasonal differences in spatial biomass patterns in our population models led to seasonal differences in biomass estimates, with most estimates producing a higher error in the fall.
 This may have to do with the fact that our species spawn during months 9-14, which coincides with the spring sampling weeks 13 \& 14.
@@
 The populations are more spread out during the fall survey in weeks 37 \& 38 and since the water temperature is warmer during these months biomass shifts towards the deeper strata, which tend to be less sampled.
-We explored this through abundance estimate ratios which revealed how the model-based estimates had a larger change in average ratio between seasonal estimates compared to the design-based estimates.
+We explored this through biomass estimate ratios which revealed how the model-based estimates had a larger change in average ratio between seasonal estimates compared to the design-based estimates.
 
 Spatial factors contributed to many of the scenarios that produced larger error values seen in our estimates.
-Flow from the shallow centrally located strata into the deeper, smaller, and less-sampled eastern strata contributed to VAST failing to accurately model the abundance of Yellowtail without the addition of covariate information, the impact of which was exacerbated in the increasing temperature scenarios.
+Flow from the shallow centrally located strata into the deeper, smaller, and less-sampled eastern strata contributed to VAST failing to accurately model the biomass of Yellowtail without the addition of covariate information, the impact of which was exacerbated in the increasing temperature scenarios.
 While adding covariates can help inform the model and improve the estimate as seen with the Yellowtail simulation results, covariates can hinder the model in certain instances with Cod and Haddock.
@@
 Chen, C., Beardsley, R.C., Cowles, G., Qi, J., Lai, Z., Gao, G., and others. 2006. An unstructured grid, finite-volume coastal ocean model: FVCOM user manual. SMAST/UMASSD: 6--8.
+
+\leavevmode\vadjust pre{\hypertarget{ref-deville1992calibration}{}}%
+Deville, J.-C., and Särndal, C.-E. 1992. Calibration estimators in survey sampling. Journal of the American statistical Association \textbf{87}(418): 376--382. Taylor \& Francis.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ThorsonSeasonal}{}}%
-Thorson, James T. (n.d.b). {Seasonal and Interannual Dynamics Github Documentation}. {https://james-thorson-noaa.github.io/docs/tutorials/seasonal-model/}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-VASTwiki}{}}%
-Thorson, James T. (n.d.a). {Vector Autoregressive Spatio-Temporal (VAST) model wiki page.} {https://github.com/James-Thorson-NOAA/VAST/wiki}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-ThorsonChat}{}}%
-Thorson, James T. (n.d.c). {Personal Communication.} {??}
+Thorson, James T. 2022. Personal conversation.
 

### 2023-09-06 Manuscript/Draft2.Rmd
manuscipt update

@@
 While the model-based results provided a slightly lower overall mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02).
+The difference in variance is further illustrated by the median error values, with model-based estimates resulting in a median value of 0.235 while stratified mean values had a median error of 0.34.
 When considering individual scenarios, while about $96\%$ had a VAST estimate with a lower absolute error than the corresponding stratified mean estimate, $63\%$ of individual scenarios contained a VAST estimate with a worse error than the corresponding stratified mean estimate.
 VAST models that included covariate information provided the lowest overall absolute errors and standard deviations.
-Models with covariates had an average absolute error of 0.22, compared to models that did not include covariate information having an average error of 0.46, and stratified mean estimates produces an average error of 0.39.
+Models with covariates had an average absolute error of 0.22 (median 0.19), compared to models that did not include covariate information having an average error of 0.46 (median 0.31), and stratified mean estimates produced an average error of 0.38 (median 0.34).
 These findings demonstrate the ability of spatio-temporal models to account for spatial variation in the species being sampled under the right conditions, but also highlight volatility/sensitivity of model estimates to the settings being used and/or the covariate response detected in the data.

### 2023-09-22 Manuscript/Draft2.Rmd
updates to paper and a few scripts

@@
 
-^1^Ben's Institution, USA  
-^2^National Marine Fisheries Service, Northeast Fisheries Science Center, Woods Hole, MA, USA
+^1^Ocean Associates Inc. in support of ^2^National Marine Fisheries Service, Northeast Fisheries Science Center, 166 Water Street, Woods Hole, Massachusetts 02543-1026
 
@@
 More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey could change, altering the relationship between the index and the true population [@arreguin1996catchability; @langan2021climate].
-A species shifting its range beyond the survey area or outside of the assumed stock boundary are additional compounding factor to consider.
+A species shifting its range beyond the survey area or outside of the assumed stock boundary is an additional compounding factor to consider.
 Existing research has focused on temperature as the driver of such changes [@klein2017effects] and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges [@kerr2022coherence].
@@
 
+(ref:crossref) [@walsh2015long]
+
 ```{r strata-plot, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Strata inhabited by each species in our population models.
-All strata on the eastern US contintental shelf are shown in the first image with Georges Bank highlighted in orange [@walsh2015long] **fix cite**.
+All strata on the eastern US contintental shelf are shown in the first image with Georges Bank highlighted in orange (ref:crossref).
 Stratum numbers used by the NEFSC bottom trawl survey are shown in the middle row.
@@
 Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics, and therefore, these results cannot be compared directly with stock assessment results.
-A full list of parameters used in our model can be seen in Tables 1 and \@ref(tab:paramsSCENARIOS).
+A full list of parameters used in our population model can be seen in Tables 1 and \@ref(tab:paramsSCENARIOS).
 
@@
 Positive tows are weighted by the biomass of the given catch.
-The middle two plots show our covariate predictors used in the *lrren* niche model- depth and median sediment size.
+The middle two plots show our covariate predictors used in the \\textit{lrren} niche model- depth and median sediment size.
 The plot on the right shows the final niche model that has been rescaled between 0 and 1, where blue values are closer to 0 and red values are closer to 1."}
@@
 Each species is assumed to have normally distributed temperature preferences $N(\mu,\sigma)$, with mean $\mu$ and standard deviation $\sigma$.
+Temperature values are in degrees Celcius.
 Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey.
@@
 We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on spatial patterns while remaining within reasonable computational limits in terms of the length of the simulation.
-Figure \@ref(fig:temp-scenarios) depicts mean trends for the temperature scenarios used in our models.
+Figure \@ref(fig:temp-scenarios) depicts mean trends for the temperature scenarios used in our population models.
 **dont forget to include animated gif in final submission**
 
-```{r temp-scenarios, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Mean trends of temperature data used in our model."}
+```{r temp-scenarios, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Mean trends of temperature data used in our MixFishSim population model."}
 knitr::include_graphics(c("Images/TempScenarios.png"))
@@
 We carry out 20 year simulations for each of our three species under various population scenarios.
-Historically, Cod has seen significant decline over the last 50**??** years while Haddock has increased in abundance in recent years [[can cite the 2022 management track assessments, see https://apps-nefsc.fisheries.noaa.gov/saw/sari.php for when the document becomes available]] **just cite 2021 assessments?**.
+Historically, Cod has seen significant decline over the last 50 years while Haddock has increased in abundance in recent years [@ManagTrack].
 For this reason we compare survey index estimates using SRS from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
@@
 After distributing each species in its stock area with simulated population trajectories, we mimic the bottom trawl survey by conducting SRS in the set of strata associated with each stock twice each year.
-We sample each stratum in the same weeks in which the spring and fall surveys have historically sampled Georges Bank (weeks 13 & 14 in the spring and 37 & 38 in the fall).
+We sample each stratum in the same weeks in which the spring and fall surveys have historically sampled Georges Bank (weeks 13 and 14 in the spring and 37 and 38 in the fall).
 The number of simulated tows made reflects true target values for each stratum and sample cells where "tows" are made were randomly selected.
@@
 Figure \@ref(fig:strata-plot) shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \@ref(fig:PopPct).
-The yellow regions in Figure \@ref(fig:strata-plot) depict the strata that were not sampled for each species.
+The yellow regions in Figure \@ref(fig:strata-plot) depict the strata that were not sampled for each species in this scenario.
 We then use the biomass collected from our samples in contemporary biomass index methods to estimate yearly seasonal population trends.
@@
 
-Knowing the true population values in our simulations allows us to compute the absolute error in a given season $E_s$ from each estimation method and scenario as given by
+Knowing the true population values in our simulations allows us to compute the relative absolute error in a given season $E_s$ from each estimation method and scenario as given by
 
@@
 After running models for each of our scenarios using these settings, we received feedback to also consider configurations for Settings A [@ThorsonChat].
-Both settings use a Tweedie model and turn off the first linear predictor $p_1$ so that only biomass is being modeled.
-The difference between the two settings can be seen in the Rho Configuration, where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A.
+Both settings use a Tweedie model and the difference between the two settings can be seen in the Rho Configuration, where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A.
 The difference amounts to intercepts for the biomass predictor being a fixed effect (RhoBeta2=0) versus intercepts being constant among years as a fixed effect (RhoBeta2=3).
@@
 
-Each biomass index scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, & covariates), resulting in a large number of scenario combinations to consider.
+Each biomass index scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, and covariates), resulting in a large number of scenario combinations to consider.
 The columns in Table \@ref(tab:scenarios) show the choices that define each scenario.
@@
 While the model-based results provided a slightly lower overall mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02).
-The difference in variance is further illustrated by the median error values, with model-based estimates resulting in a median value of 0.235 while stratified mean values had a median error of 0.34.
+The difference in variance is further illustrated by the median error values, with model-based estimates producing a median value of 0.235 while stratified mean values had a median error of 0.34.
 When considering individual scenarios, while about $96\%$ had a VAST estimate with a lower absolute error than the corresponding stratified mean estimate, $63\%$ of individual scenarios contained a VAST estimate with a worse error than the corresponding stratified mean estimate.
@@
 
-A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide biomass estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values.
-This can be further examined through yearly estimate:model ratio values, where we divide the yearly biomass estimates by the true model value.
+A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide biomass estimates that are above the true biomass value, while the stratified mean estimates are, on average, below the true biomass values.
+This can be further examined through yearly estimate:true biomass ratio values, where we divide the yearly biomass estimates by the true biomass value.
 In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can result in yearly biomass values as low as 0 at times and exhibit large yearly changes (Figure \@ref(fig:RatioTrend)).
 
-Analysis of individual yearly estimate:model ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21.
+Analysis of individual yearly estimate:true biomass ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21.
 On the other hand, just 33\% of stratified mean estimate ratios were above 1 with an average of 0.874 and a standard deviation of 0.12.
@@
 
-When the entire domain is sampled, the stratified mean produced an average yearly estimate:model ratio of 0.87 with a standard deviation of 0.12.
+When the entire domain is sampled, the stratified mean produced an average yearly estimate:true biomass ratio of 0.87 with a standard deviation of 0.12.
+As the number of years estimated approaches infinity this value should converge to 1.
 Adding covariates in these scenarios brings the VAST estimate ratio closer to 1.
@@
 
-Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata) contain the absolute error between our biomass estimates for Yellowtail comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B).
+Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata) contain the absolute error between our biomass estimates for Yellowtail comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \and B).
 In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
@@
 
-To test the impact of shifting populations on biomass estimates, we simulated population data that included environmental forcing and compared biomass estimates derived from SRS to the known true values.
+<!-- To test the impact of shifting populations on biomass estimates, we simulated population data that included environmental forcing and compared biomass estimates derived from SRS to the known true values.
 We carried out simulations for 3 species on Georges Bank that took into account different population trends, temperature scenarios, noise in survey samples, and seasonal variation.
@@
 Our analysis included a comparison of absolute error values between 20-year estimates and consideration of the year-to-year variation displayed by each approach as well as the variation between each 20-year estimate.
+-->
 
@@
 Seasonal differences in spatial biomass patterns in our population models led to seasonal differences in biomass estimates, with most estimates producing a higher error in the fall.
-This may have to do with the fact that our species spawn during months 9-14, which coincides with the spring sampling weeks 13 & 14.
+This may have to do with the fact that our species spawn during weeks 9-14, which coincides with the spring sampling weeks 13 and 14.
 Spawning in our population models concentrates much of each species in the spawning grounds towards the center of the domain where more samples occur.
-The populations are more spread out during the fall survey in weeks 37 & 38 and since the water temperature is warmer during these months biomass shifts towards the deeper strata, which tend to be less sampled.
+The populations are more spread out during the fall survey in weeks 37 and 38 and since the water temperature is warmer during these months biomass shifts towards the deeper strata, which tend to be less sampled.
 We explored this through biomass estimate ratios which revealed how the model-based estimates had a larger change in average ratio between seasonal estimates compared to the design-based estimates.
@@
 
-Of the ~27% of VAST models where including covariates did not improve the VAST estimate, most provided a comparable error (for ex, 0.11 vs 0.13).
+Of the ~27% of VAST models where including covariates did not improve the VAST estimate, most provided a comparable error (e.g., 0.11 vs 0.13).
 The instances where including covariates provided a noticeable change in error took place in scenarios that included either an increasing temperature and/or reduced sampling domain.
@@
 Thank you to Jim Thorson for consulting with us about VAST settings and implementation.
-We are thanksful for Chris Cacciapaglia and Charles Adams for the time they spent with us sharing their experiences using VAST.
+We thank Chris Cacciapaglia and Charles Adams for the time they spent with us sharing their experiences using VAST.
 We appreciate the help from Robyn Linner for helping to extract FVCOM temperature data and are thankful for David Chevrier for meeting with us to clarify questions related to some of our GIS data.

### 2023-09-22 Manuscript/Draft2.tex
updates to paper and a few scripts

@@
 
-\textsuperscript{1}Ben's Institution, USA\\
-\textsuperscript{2}National Marine Fisheries Service, Northeast Fisheries Science Center, Woods Hole, MA, USA
+\textsuperscript{1}Ocean Associates Inc.~in support of \textsuperscript{2}National Marine Fisheries Service, Northeast Fisheries Science Center, 166 Water Street, Woods Hole, Massachusetts 02543-1026
 
@@
 2021).
-A species shifting its range beyond the survey area or outside of the assumed stock boundary are additional compounding factor to consider.
+A species shifting its range beyond the survey area or outside of the assumed stock boundary is an additional compounding factor to consider.
 Existing research has focused on temperature as the driver of such changes (Klein et al.
@@
 \caption{Strata inhabited by each species in our population models.
-All strata on the eastern US contintental shelf are shown in the first image with Georges Bank highlighted in orange [@walsh2015long] **fix cite**.
+All strata on the eastern US contintental shelf are shown in the first image with Georges Bank highlighted in orange (Walsh et al.
+2015).
 Stratum numbers used by the NEFSC bottom trawl survey are shown in the middle row.
@@
 Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics, and therefore, these results cannot be compared directly with stock assessment results.
-A full list of parameters used in our model can be seen in Tables 1 and \ref{tab:paramsSCENARIOS}.
+A full list of parameters used in our population model can be seen in Tables 1 and \ref{tab:paramsSCENARIOS}.
 
@@
 Positive tows are weighted by the biomass of the given catch.
-The middle two plots show our covariate predictors used in the *lrren* niche model- depth and median sediment size.
+The middle two plots show our covariate predictors used in the \textit{lrren} niche model- depth and median sediment size.
 The plot on the right shows the final niche model that has been rescaled between 0 and 1, where blue values are closer to 0 and red values are closer to 1.}\label{fig:hab-plot1}
@@
 Each species is assumed to have normally distributed temperature preferences \(N(\mu,\sigma)\), with mean \(\mu\) and standard deviation \(\sigma\).
+Temperature values are in degrees Celcius.
 Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey.
@@
 We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on spatial patterns while remaining within reasonable computational limits in terms of the length of the simulation.
-Figure \ref{fig:temp-scenarios} depicts mean trends for the temperature scenarios used in our models.
+Figure \ref{fig:temp-scenarios} depicts mean trends for the temperature scenarios used in our population models.
 \textbf{dont forget to include animated gif in final submission}
@@
 
-\caption{Mean trends of temperature data used in our model.}\label{fig:temp-scenarios}
+\caption{Mean trends of temperature data used in our MixFishSim population model.}\label{fig:temp-scenarios}
 \end{figure}
@@
 We carry out 20 year simulations for each of our three species under various population scenarios.
-Historically, Cod has seen significant decline over the last 50\textbf{??} years while Haddock has increased in abundance in recent years {[}{[}can cite the 2022 management track assessments, see \url{https://apps-nefsc.fisheries.noaa.gov/saw/sari.php} for when the document becomes available{]}{]} \textbf{just cite 2021 assessments?}.
+Historically, Cod has seen significant decline over the last 50 years while Haddock has increased in abundance in recent years ((NEFSC) 2022).
 For this reason we compare survey index estimates using SRS from decreasing population scenarios for Cod and increasing population scenarios for Haddock.
@@
 After distributing each species in its stock area with simulated population trajectories, we mimic the bottom trawl survey by conducting SRS in the set of strata associated with each stock twice each year.
-We sample each stratum in the same weeks in which the spring and fall surveys have historically sampled Georges Bank (weeks 13 \& 14 in the spring and 37 \& 38 in the fall).
+We sample each stratum in the same weeks in which the spring and fall surveys have historically sampled Georges Bank (weeks 13 and 14 in the spring and 37 and 38 in the fall).
 The number of simulated tows made reflects true target values for each stratum and sample cells where ``tows'' are made were randomly selected.
@@
 Figure \ref{fig:strata-plot} shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \ref{fig:PopPct}.
-The yellow regions in Figure \ref{fig:strata-plot} depict the strata that were not sampled for each species.
+The yellow regions in Figure \ref{fig:strata-plot} depict the strata that were not sampled for each species in this scenario.
 We then use the biomass collected from our samples in contemporary biomass index methods to estimate yearly seasonal population trends.
@@
 
-Knowing the true population values in our simulations allows us to compute the absolute error in a given season \(E_s\) from each estimation method and scenario as given by
+Knowing the true population values in our simulations allows us to compute the relative absolute error in a given season \(E_s\) from each estimation method and scenario as given by
 
@@
 After running models for each of our scenarios using these settings, we received feedback to also consider configurations for Settings A (Thorson, James T 2022).
-Both settings use a Tweedie model and turn off the first linear predictor \(p_1\) so that only biomass is being modeled.
-The difference between the two settings can be seen in the Rho Configuration, where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A.
+Both settings use a Tweedie model and the difference between the two settings can be seen in the Rho Configuration, where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A.
 The difference amounts to intercepts for the biomass predictor being a fixed effect (RhoBeta2=0) versus intercepts being constant among years as a fixed effect (RhoBeta2=3).
@@
 
-Each biomass index scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, \& covariates), resulting in a large number of scenario combinations to consider.
+Each biomass index scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, and covariates), resulting in a large number of scenario combinations to consider.
 The columns in Table \ref{tab:scenarios} show the choices that define each scenario.
@@
 While the model-based results provided a slightly lower overall mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02).
+The difference in variance is further illustrated by the median error values, with model-based estimates producing a median value of 0.235 while stratified mean values had a median error of 0.34.
 When considering individual scenarios, while about \(96\%\) had a VAST estimate with a lower absolute error than the corresponding stratified mean estimate, \(63\%\) of individual scenarios contained a VAST estimate with a worse error than the corresponding stratified mean estimate.
 VAST models that included covariate information provided the lowest overall absolute errors and standard deviations.
-Models with covariates had an average absolute error of 0.22, compared to models that did not include covariate information having an average error of 0.46, and stratified mean estimates produces an average error of 0.39.
+Models with covariates had an average absolute error of 0.22 (median 0.19), compared to models that did not include covariate information having an average error of 0.46 (median 0.31), and stratified mean estimates produced an average error of 0.38 (median 0.34).
 These findings demonstrate the ability of spatio-temporal models to account for spatial variation in the species being sampled under the right conditions, but also highlight volatility/sensitivity of model estimates to the settings being used and/or the covariate response detected in the data.
@@
 
-A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide biomass estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values.
-This can be further examined through yearly estimate:model ratio values, where we divide the yearly biomass estimates by the true model value.
+A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide biomass estimates that are above the true biomass value, while the stratified mean estimates are, on average, below the true biomass values.
+This can be further examined through yearly estimate:true biomass ratio values, where we divide the yearly biomass estimates by the true biomass value.
 In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can result in yearly biomass values as low as 0 at times and exhibit large yearly changes (Figure \ref{fig:RatioTrend}).
 
-Analysis of individual yearly estimate:model ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21.
+Analysis of individual yearly estimate:true biomass ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21.
 On the other hand, just 33\% of stratified mean estimate ratios were above 1 with an average of 0.874 and a standard deviation of 0.12.
@@
 
-Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata} contain the absolute error between our biomass estimates for Yellowtail comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B).
+Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata} contain the absolute error between our biomass estimates for Yellowtail comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \textbackslash and B).
 In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors.
@@
 
-To test the impact of shifting populations on biomass estimates, we simulated population data that included environmental forcing and compared biomass estimates derived from SRS to the known true values.
-We carried out simulations for 3 species on Georges Bank that took into account different population trends, temperature scenarios, noise in survey samples, and seasonal variation.
-We compared a design-based estimation approach (stratified mean) with a model-based approach that allows for the inclusion of environmental covariates (VAST) when the entire domain was sampled as well as when a reduced number of strata were included to simulate our species shifting into a new habitat.
-Our analysis included a comparison of absolute error values between 20-year estimates and consideration of the year-to-year variation displayed by each approach as well as the variation between each 20-year estimate.
-
 Our analysis showed that both the design-based and model-based approaches are each capable of providing biomass estimates that track the true biomass trend.
@@
 Seasonal differences in spatial biomass patterns in our population models led to seasonal differences in biomass estimates, with most estimates producing a higher error in the fall.
-This may have to do with the fact that our species spawn during months 9-14, which coincides with the spring sampling weeks 13 \& 14.
+This may have to do with the fact that our species spawn during weeks 9-14, which coincides with the spring sampling weeks 13 and 14.
 Spawning in our population models concentrates much of each species in the spawning grounds towards the center of the domain where more samples occur.
-The populations are more spread out during the fall survey in weeks 37 \& 38 and since the water temperature is warmer during these months biomass shifts towards the deeper strata, which tend to be less sampled.
+The populations are more spread out during the fall survey in weeks 37 and 38 and since the water temperature is warmer during these months biomass shifts towards the deeper strata, which tend to be less sampled.
 We explored this through biomass estimate ratios which revealed how the model-based estimates had a larger change in average ratio between seasonal estimates compared to the design-based estimates.
@@
 
-Of the \textasciitilde27\% of VAST models where including covariates did not improve the VAST estimate, most provided a comparable error (for ex, 0.11 vs 0.13).
+Of the \textasciitilde27\% of VAST models where including covariates did not improve the VAST estimate, most provided a comparable error (e.g., 0.11 vs 0.13).
 The instances where including covariates provided a noticeable change in error took place in scenarios that included either an increasing temperature and/or reduced sampling domain.
@@
 McMullen, K., Paskevich, V., and Poppe, L. (n.d.). USGS east-coast sediment analysis: Procedures, database, and GIS data. US Geological Survey Open-File Report \textbf{2005}: 1001.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ManagTrack}{}}%
+(NEFSC), N.F.S.C. 2022. Management track assessments fall 2022. NOAA technical memorandum NMFS-NE ; 305.
 
@@
 Wiley Online Library.
+
+\leavevmode\vadjust pre{\hypertarget{ref-walsh2015long}{}}%
+Walsh, H.J., Richardson, D.E., Marancik, K.E., and Hare, J.A.
+2015.
+Long-term changes in the distributions of larval and adult fish in the northeast US shelf ecosystem.
+PLoS One \textbf{10}(9): e0137382.
+Public Library of Science San Francisco, CA USA.
 

### 2023-09-22 TestScripts/Plots_for_paper_presentation.R

@@ -324,8 +324,8 @@ ggplot() +
   
   
   #this way plots data by season divided by itself so it equals 1
-  geom_point(data = subset(as.data.frame(pop_by_season[[s]]),season==season_), aes(x=as.numeric(year),y=log(Est_ratio), group = season, color = "Model"),size=estimate_point_size) +
-  geom_line(data = subset(as.data.frame(pop_by_season[[s]]),season==season_), aes(x=as.numeric(year),y=log(Est_ratio), group =season, color = "Model"),size=estimate_line_size ) +
+  geom_point(data = subset(as.data.frame(pop_by_season[[s]]),season==season_), aes(x=as.numeric(year),y=log(Est_ratio), group = season, color = "True Biomass"),size=estimate_point_size) +
+  geom_line(data = subset(as.data.frame(pop_by_season[[s]]),season==season_), aes(x=as.numeric(year),y=log(Est_ratio), group =season, color = "True Biomass"),size=estimate_line_size ) +
   
   #plot VAST estimate without covariates with NO noise
   geom_point(data=subset(VAST_data[[s]][["_NoCovs_"]][["NoNoise_"]][[folder]],season==season_),aes(x=Year,y=log(Est_ratio),group=season, color = "VAST No Cov No Noise"),size=model_point_size)+

### 2023-09-22 TestScripts/plot_estimates_by_strat.R

@@ -1,0 +1,164 @@
+
+
+# for(scenario1 in c("ConPop_ConTemp","ConPop_IncTemp",
+#                    "DecPop_ConTemp","DecPop_IncTemp",
+#                    "IncPop_ConTemp","IncPop_IncTemp")){
+
+
+
+
+plot_estimates_by_strat <- function(){
+
+
+
+pdf(file=paste(orig.dir,"/testfolder/est_by_strata_",scenario1,"_",str_dir,".pdf",sep=""))
+
+
+for(s in short_names){
+
+for(ssn in c("SPRING","FALL")){
+ year_min = 3
+
+  ssn1 = ifelse(ssn=="SPRING",1,2)
+
+  #model values
+  P_est = readRDS(paste("E:\\READ-PDB-blevy2-MFS2\\GB_Simulation_Results\\",scenario1,"\\pop_by_strata_",scenario1,".RDS",sep=""))
+  P_est = P_est[[s]]
+  P_est = P_est[P_est$season==ssn1,] #by season
+  #sometimes stratum is labeled statum
+  if(names(P_est)[[3]]=="statum"){P_est$stratum = P_est$statum}
+  
+  #sum by stratum to compare with pop_by_season
+  #P_est_total <- P_est %>% group_by(year,season) %>% summarise(biomass2=sum(biomass))
+
+  #Stratified mean estimates (Yes Noise(YN) and No Noise (NN))
+  SM_est_NN = strat_mean_by_strat[[s]][["NoNoise_"]]
+  SM_est_YN = strat_mean_by_strat[[s]][["WithNoise_"]]
+
+  SM_est_NN = SM_est_NN[SM_est_NN$Season==ssn,]
+  SM_est_YN = SM_est_YN[SM_est_YN$Season==ssn,]#by season
+
+  #Vast estimates (Y noise, Y covs and N versions)
+  V_est_NC_NN = vast_by_str[[s]][["_NoCovs_"]][["NoNoise_"]]
+  V_est_NC_YN = vast_by_str[[s]][["_NoCovs_"]][["WithNoise_"]]
+  V_est_YC_NN = vast_by_str[[s]][["_WithCov_"]][["NoNoise_"]]
+  V_est_YC_YN = vast_by_str[[s]][["_WithCov_"]][["WithNoise_"]]
+  
[…]

### 2023-09-22 TestScripts/vast_by_strat.R

@@ -1,0 +1,101 @@
+# 
+# x <- cbind(fit$extrapolation_list$Data_Extrap[,1],
+#            fit$extrapolation_list$Data_Extrap[,2],
+#            fit$Report$D_gct,
+#            rep(seq(3,22),each=5000))
+# 
+# #x will have 100,000 rows. 5000 for each of the 20 years. The extrapolation gris has 5000 cells
+# 
+# colnames(x) <- c("Lon","Lat","Biomass","Year")
+# 
+# x<- as.data.frame(x, xy=TRUE)
+# 
+# #rasterFromXYZ(x[,1:3][x$Year==3,],res=res(hab_ras))
+# xx<-raster::extract(GB_strata_Had,x[,2:1],layer="STR2")
+# 
+# 
+# 
+# 
+# x<-sp::SpatialPoints(coords=cbind(fit$extrapolation_list$Data_Extrap[,1],fit$extrapolation_list$Data_Extrap[,2]))
+# crs(x) <- crs(r1)
+# crs(GB_strata_Had) <-crs(r1)
+# xx[[ssn]]=sp::over(x,GB_strata_Had)
+# biomass_est <- as.numeric(fit$Report$D_gct[1:5000]*fit$extrapolation_list$Area_km2_x)
+# test <- cbind(xx[[ssn]],x,biomass_est)
+# 
+# colnames(test)[c(4,8,9,10)] <- c("stratum","Lon","Lat","Biomass_est")
+
+
+#knots lat lon used:
+#plot(sp::SpatialPoints(coords=cbind(fit$spatial_list$latlon_x[,2],fit$spatial_list$latlon_x[,1])))
+
+
+
+vast_by_strat <- function(s,str_dir,cov_directory,noise, scenario1, r1, GB_strata_Had, strata_species_s ){
+
+xx<-list()
+xxx<-data.frame()
+for(ssn in c("spring","fall")){
+print(ssn)
+  fit <- readRDS(
+    paste("C:/Users/benjamin.levy/Desktop/Github/READ-PDB-blevy2-MFS2/VAST/ForPaper/",scenario1,"/",s,"/",str_dir,cov_directory,noise,"/obsmodel7/",ssn,"/fit_",ssn,".RDS",sep="")
+  )
+
+
+ssn1 = ifelse(ssn == "spring", "SPRING","FALL")
+
+#Old way (uses estimates at samples instead of extrapolation points)
[…]

### 2023-11-20 Manuscript/Draft2.Rmd
update manuscript

@@
 library("forcats")
-
+#library("embedr")
+library("vembedr")
 knitr::opts_chunk$set(echo = FALSE)
@@
 
-# Estimating Population Trends Under the Pressures of Climate Change
+# How might climate change affect indices of abundance? A simulation study starting point.
 
@@
 
-```{r temp-scenarios, echo=FALSE,out.width="95%",fig.show='hold', fig.align='center', fig.cap="Mean trends of temperature data used in our MixFishSim population model."}
-knitr::include_graphics(c("Images/TempScenarios.png"))
-```
-
 <!-- --Describe difference between increasing and constant temperature scenarios (images?) -->
@@
 Users often rely on traditional model-building processes and diagnostic techniques such as AIC to guide their choices, but successfully navigating the many decisions depends on the background of the given user and care should be taken in determining which diagnostic tools are used.
-For example, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using settings that decrease the accuracy of the model **cite Chris C paper**.
+For example, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using settings that decrease the accuracy of the model [@ChrisC].
 The value of model-based approaches such as VAST are clear- they allow users to overcome issues related to the input data to account for factors such as environmental change.
@@
 
-<div align="center">Table 1. Parameters used in all population models. SAW 1: [@nefsc1218], SAW 2: [@nefsc1311], SAW 3: not out yet??? </div>
+<div align="center">Table 1. Parameters used in all population models. </div>
 
@@
 | $M$ | Natural Mortality| wk$^{-1}$  | 0.2064 | 0.2728 | 0.3340 | [@thorson2020predicting]|
-| $F$ | Fishing Mortality| wk$^{-1}$  | 0.358 | 0.511 | 0.45 | SAW 1, 2, 3 |
-| $W_R$ |Weight of fully recruited fish| kg | 0.39 | 2.95 | 1.12 | SAW 1, 2, 3 |
-| $W_{R-1}$ | Weight of pre-recruit fish| kg | 0.13 | 0.39 | 0.19 | SAW 1, 2, 3 |
+| $F$ | Fishing Mortality| wk$^{-1}$  | 0.358 | 0.511 | 0.45 | [@ManagTrack] |
+| $W_R$ |Weight of fully recruited fish| kg | 0.39 | 2.95 | 1.12 | [@ManagTrack] |
+| $W_{R-1}$ | Weight of pre-recruit fish| kg | 0.13 | 0.39 | 0.19 | [@ManagTrack] |
 | $\\sigma^2$ | Variance in recruited fish | kg$^2$ | 0.55 | 0.55 | 0.55 |assumed|
 | $\\lambda$ | Decay rate for movement | - | 0.7 | 0.7 | 0.7 |assumed|
-| $Spwn_s$ | Spawning weeks for species $s$ | wk | 9-12 | 8-13 | 11-14 |SAW 1, 2, 3|
-| $Rec_s$ | Recruitment weeks for species $s$ | wk | 9-12 | 8-13 | 11-14 |SAW 1, 2, 3|
+| $Spwn_s$ | Spawning weeks for species $s$ | wk | 9-12 | 8-13 | 11-14 |[@ManagTrack]|
+| $Rec_s$ | Recruitment weeks for species $s$ | wk | 9-12 | 8-13 | 11-14 |[@ManagTrack]|
 "
@@
 
+# COLORs DEFINED BY ROW_SPEC ONLY WORK FOR PDF FORMAT
 # Create table to present the data]
@@
 
+# COLORs DEFINED BY ROW_SPEC ONLY WORK FOR HTML FORMAT
+# Fig_5a <- kable(summary_tab_all[, -1],
+#       align = c("l", "l", "l", "l", "c", "c", "c"),
+#   caption = "Yellowtail Flounder absolute error results with all strata included in calculations. Row colors correspond to the same settings applied in different seasons.") %>%
+#   row_spec(0,bold = TRUE)%>% #bold column titles
+#   kable_styling(font_size = 10)  %>%
+#   pack_rows(tab_kable(), colnum = 1,
+#     index = table(fct_inorder(summary_tab_all$Scenario), useNA = "no"))
+
 ```
@@
 #define colors
+# 
+#  COLORS DEFINED BY ROW_SPEC ONLY WORK FOR PDF FORMAT
 
@@
 
+#COLORS BELOW DEFINED BY ROW_SPEC ARE FOR HTML OUTPUT
+
+# Fig_5b <- kable(summary_tab_exclude[, -1],
+#       align = c("l", "l", "l", "l", "c", "c", "c"),
+#   caption = "Yellowtail Flounder absoloute error results with certain strata excluded from calculations. Row colors correspond to the same settings applied in different seasons.") %>%
+#   row_spec(0,bold = TRUE)%>% #bold column titles
+#   kable_styling(font_size = 10)  %>%
+#   pack_rows(tab_kable(), colnum = 1,
+#     index = table(fct_inorder(summary_tab_exclude$Scenario), useNA = "no"))
+
 ```

### 2023-11-20 Manuscript/Draft2.tex
update manuscript

@@
 
-\hypertarget{estimating-population-trends-under-the-pressures-of-climate-change}{%
-\section{Estimating Population Trends Under the Pressures of Climate Change}\label{estimating-population-trends-under-the-pressures-of-climate-change}}
+\hypertarget{how-might-climate-change-affect-indices-of-abundance-a-simulation-study-starting-point.}{%
+\section{How might climate change affect indices of abundance? A simulation study starting point.}\label{how-might-climate-change-affect-indices-of-abundance-a-simulation-study-starting-point.}}
 
@@
 
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{Images/TempScenarios}
-
-}
-
-\caption{Mean trends of temperature data used in our MixFishSim population model.}\label{fig:temp-scenarios}
-\end{figure}
-
 In equation (1), \(Hab^2_{J,s}\) remains constant for each species for the duration of the simulation, while \(Tol_{c,s,wk}\) changes each week with temperature fluctuations.
@@
 Users often rely on traditional model-building processes and diagnostic techniques such as AIC to guide their choices, but successfully navigating the many decisions depends on the background of the given user and care should be taken in determining which diagnostic tools are used.
-For example, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using settings that decrease the accuracy of the model \textbf{cite Chris C paper}.
+For example, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using settings that decrease the accuracy of the model (Christopher Cacciapaglia 2023).
 The value of model-based approaches such as VAST are clear- they allow users to overcome issues related to the input data to account for factors such as environmental change.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-ChrisC}{}}%
+Christopher Cacciapaglia, C.F.A., Elizabeth N.
+Brooks.
+2023.
+Developing workflow and diagnostics for model selection of a vector autoregressive spatiotemporal (VAST) model.
+Fisheries Research \textbf{In Press}.
+Elsevier.
+
 \leavevmode\vadjust pre{\hypertarget{ref-deville1992calibration}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-nefsc1218}{}}%
-NEFSC.
-2012.
-{54th Northeast Regional Stock Assessment Workshop (54th SAW) Assessment Report}.
-{Northeast Fish.
-Sci.
-Cent.
-Ref.
-Doc.
-12-18; 604 p.
-Available from: National Marine Fisheries Service, 166 Water Street, Woods Hole, MA 02543-1026, or online at http://www.nefsc.noaa.gov/nefsc/publications/}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nefsc1311}{}}%
-NEFSC.
-2013.
-{55th Northeast Regional Stock Assessment Workshop (55th SAW) Assessment Report}.
-{Northeast Fish.
-Sci.
-Cent.
-Ref.
-Doc.
-13-11; 849 p.
-Available from: National Marine Fisheries Service, 166 Water Street, Woods Hole, MA 02543-1026, or online at http://www.nefsc.noaa.gov/nefsc/publications/}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-nye2009changing}{}}%
@@
 
-Table 1. Parameters used in all population models. SAW 1: (NEFSC 2012), SAW 2: (NEFSC 2013), SAW 3: not out yet???
+Table 1. Parameters used in all population models.
 
@@
 \(M\) & Natural Mortality & wk\(^{-1}\) & 0.2064 & 0.2728 & 0.3340 & (Thorson 2020) \\
-\(F\) & Fishing Mortality & wk\(^{-1}\) & 0.358 & 0.511 & 0.45 & SAW 1, 2, 3 \\
-\(W_R\) & Weight of fully recruited fish & kg & 0.39 & 2.95 & 1.12 & SAW 1, 2, 3 \\
-\(W_{R-1}\) & Weight of pre-recruit fish & kg & 0.13 & 0.39 & 0.19 & SAW 1, 2, 3 \\
+\(F\) & Fishing Mortality & wk\(^{-1}\) & 0.358 & 0.511 & 0.45 & ((NEFSC) 2022) \\
+\(W_R\) & Weight of fully recruited fish & kg & 0.39 & 2.95 & 1.12 & ((NEFSC) 2022) \\
+\(W_{R-1}\) & Weight of pre-recruit fish & kg & 0.13 & 0.39 & 0.19 & ((NEFSC) 2022) \\
 \(\sigma^2\) & Variance in recruited fish & kg\(^2\) & 0.55 & 0.55 & 0.55 & assumed \\
 \(\lambda\) & Decay rate for movement & - & 0.7 & 0.7 & 0.7 & assumed \\
-\(Spwn_s\) & Spawning weeks for species \(s\) & wk & 9-12 & 8-13 & 11-14 & SAW 1, 2, 3 \\
-\(Rec_s\) & Recruitment weeks for species \(s\) & wk & 9-12 & 8-13 & 11-14 & SAW 1, 2, 3 \\
+\(Spwn_s\) & Spawning weeks for species \(s\) & wk & 9-12 & 8-13 & 11-14 & ((NEFSC) 2022) \\
+\(Rec_s\) & Recruitment weeks for species \(s\) & wk & 9-12 & 8-13 & 11-14 & ((NEFSC) 2022) \\
 \bottomrule

### 2023-11-20 Manuscript/Draft3.tex
new version of paper

--- 2023-08-15 Manuscript/Draft2.tex
@@
 While the model-based results provided a slightly lower overall mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02).
+The difference in variance is further illustrated by the median error values, with model-based estimates resulting in a median value of 0.235 while stratified mean values had a median error of 0.34.
 When considering individual scenarios, while about \(96\%\) had a VAST estimate with a lower absolute error than the corresponding stratified mean estimate, \(63\%\) of individual scenarios contained a VAST estimate with a worse error than the corresponding stratified mean estimate.
 VAST models that included covariate information provided the lowest overall absolute errors and standard deviations.
-Models with covariates had an average absolute error of 0.22, compared to models that did not include covariate information having an average error of 0.46, and stratified mean estimates produces an average error of 0.39.
+Models with covariates had an average absolute error of 0.22 (median 0.19), compared to models that did not include covariate information having an average error of 0.46 (median 0.31), and stratified mean estimates produced an average error of 0.38 (median 0.34).
 These findings demonstrate the ability of spatio-temporal models to account for spatial variation in the species being sampled under the right conditions, but also highlight volatility/sensitivity of model estimates to the settings being used and/or the covariate response detected in the data.
